### 2020-04-18 text/manuscript.Rmd
initial checkin add manuscript preliminary draft

---
title: "TTmanuscript"
author: "Dennis Hazelett"
date: "4/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Title: A molecular taxonomy of tumors independent of tissue-of-origin

### Authors: Peter T. Nguyen, Simon G. Coetzee, Corona I. Rosario, and Dennis J. Hazelett

## Abstract: 
Cancer is a disease that affects metabolism and other basic cellular processes common to all cells. In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying theories elude the field. To address this challenge we reduced omics mutation and expression profiling data to pathway memberships. We found that tumors can be classified according to pathway disruption in twelve distinct classes orthogonal to a tissue of origin (including histological subtypes), number of mutations and clinical stage. Moreover, known germline risk mutations such as BRCA1/2 are equally distributed among the classes. One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR. This class is neatly subdivided into PI3K/Akt and Ras branches. Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling. Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling. Intriguingly, we identify a novel class of mutations in proteasome component proteins that may be sufficient to cause compound defects across multiple signaling pathways in this class. A third class is typified by defects in metabolism, especially of RNA and protein biosynthesis. In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors. With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found. We discuss the implications for tumor biology and treatment.

## Introduction:

Recent advances in technology have made possible the study of cancer at unprecedented levels of detail and scale. A combination of array and next-generation sequencing technologies has made it possible to assay the germline variation and somatic mutations, expression profile, methylation status, copy number variation. To take advantage of these tools, several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) endeavored to sequence large number of tumors (order of thousands) and collect data from multiple assays that could be analyzed together, with an eye towards integration and a general understanding of the mechanisms of cancer. These efforts have consisted chiefly of recognizing and quantifying the importance of individual genes in cancer etiology as the product of positive selection processes, or classification of cancers relative to common patterns of aberration in the expression of those genes in various assays.

Some considerable progress has been made in analyzing these data. Statistical analyses identify hundreds of global and tissue-specific cancer driver genes (pmid 22759861, 23884480, 24390350, 26209800, 27911828, 31287869, 31363082) using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance. Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state (pmid 26559569).

Other studies have been aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin. The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data (pmid 25109877). The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin. In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method (pmid 19759197), the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney (pmid 29625048). The authors of this study analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.

More recently, attempts have been made to consider pan-cancer mutations and other data outside the context of enrichment at the gene level. Horn et al (pmid 29200198) introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors. This method proves powerful especially when a small number (on the order of hundreds) of tumors are involved. However, the larger problem of cancer etiology remains: what mutations are required, in what order, and are there specific combinations of delinquent cellular processes sufficient to cause cancer? If there is a common logic to cancer, how is it to be detected underneath the overwhelming cell-of-origin signal?

In this study, we attempted to address this challenge by focusing on discrete molecular pathways as the unit of disruption in cancer. This operation has the virtue of stripping all cell-type-specific information from the mutation data. We use the resulting statistic to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of tissue of origin. Furthermore, tumors of varying cancer type and/or histological subtype may merely reflect the likely cell-of-origin that gave birth to them.

## Results:

### Classification of tumors by pathway disruption reveals 3 major groups

To study cancer pathways we obtained a set of 7,607 tumor samples (x donors) from GDC. We wanted to classify tumors by pathway disruption considered different types of data available from TCGA. We chose to start with somatic mutations in exome sequencing data because the affected target gene is known. Therefore, we focused on all missense, nonsense, frameshift, and splicing mutations. We selected 381 Reactome pathways (reactome.org) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories (e.g. “transcription factors”) or disease associations (e.g. “mutated in colon cancer”) and filtered our gene list on membership in these pathways. We then compiled a mutation matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations.

In order to classify tumors using this dataset, we used unsupervised clustering methods. Preliminary work with hierarchical clustering produced a set of 15 clusters (optimized with clusterProfiler(ref) was very promising, as the tumors appeared to cluster largely independent of tissue of origin (Supplementary Figure S1A). To determine whether our clusters were differentiated from random selections of tumors we plotted Kaplan Meier survival curves for each of the clusters first, and compared the results to randomized sets of tumors of equal-sized proportions (Figure S1B, C). In addition we fitted a Cox proportional hazards test model, accounting for age of diagnosis and cluster membership and found that 8 of the clusters were significantly different from a reference cluster (cluster 1), suggesting that cluster membership could in principle be rooted in biological differences affecting survival (Supplementary Table S1). These differences were small relative to differences in cancer by tissue of origin (Figure S1D).

We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP (arXiv:1802.03426). Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings (Figure 1A). We call these tentatively by roman numerals I-V. While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure 1A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method. Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14). To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only). While there is a subtle trend, mostly visible for groups I and V (violin plot, Figure 1B), mutation count does not appear to be the dominant feature. To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure 1C).

The 15 hierarchical clusters were largely free of any tissue-of-origin bias, so we suspected that our 5 groups would be similarly independent from tissue of origin. To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures 1D). Full projections for all 23 cancers are presented in Supplementary Figure S2. Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies. One hypothesis to explain the lack of bias by cancer type is that the groupings are representative of these molecular subtypes. To probe this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP. Perhaps surprisingly we did not detect any group biases for these or any other subtype annotations in GDC (see Supplementary Figure S3 for the full set of projections), suggesting that these subtypes are reflective of differences either in specialized cell-type-of-origin or other cancer risk mechanisms.

There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes. However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length. BRCA1/2 genes for example are key for DNA double-stranded break repair (ref) and confer risk for breast, prostate and ovarian cancers (refs). Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure 1E). We made similar projections for the MMR genes MSH2, MSH6, MLH, PMS1 and PMS2; BRIP1, RAD15, CHEK2 and APC. None of these genes exhibited any remarkable specificity with respect to group (except maybe APC??) (Supplementary Figure S4). To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the TET2 and TET3 genes, plus telomerase 1, 2, 3, and observed similar lack of segregation by group (Supplementary Figure S4).

### Determinants of Tumor Class

In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 381 pathways used to produce it (Supplementary Figure S5), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment. We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure 2A). Using this approach, we identified 6? pathways associated with class II, 34 pathways primarily enriched in group III tumors, and 10 pathways enriched in class IV tumors. Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively. Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure 2B). Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution. Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure 2C). Class IV had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, beta-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others. As groups III and IV were the most distinct of the 5, we chose to investigate these two groups further.

Group III contained interesting sub-patterns of enrichment, which we subdivided further into class IIIa and class IIIb sequentially from right following the global naming system. Common to classes IIIa and b were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as EGFR, IGF1R, Insulin receptor, and ERBB2 and 4. In addition to these we observed SCF-Kit, DAP12, and Tie 2 signaling pathways, altogether suggesting that Class III tumors share mutations in extracellular signaling pathways generally. Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment. We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure 3 or supplement figure?). We find….(please do this analysis)

Intriguingly, we noticed that the pathways enriched for class IIIa and b were related to RAS and PI3K/Akt signaling, respectively, which potentially represent the downstream components of the extracellular signaling pathways discussed above. Moreover, class IIIb is also enriched in several immune-associated pathways including for both innate immunity and T cell receptor signaling, suggesting an interesting correlation between the PI3K/Akt subgroup and potential loss (or hyperactivation) of immune function. To address this possibility, we isolated the PI3K/Akt, RAS and immune/TCR pathways and assessed whether the mutations tended to be biased toward likely inactivating or activating mutations ….

Class IV exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors between class III and IV on the UMAP projection that have pathway characteristics of both. The top ten distinguishing pathways of Class IV tumors were cellular response to hypoxia, regulation of mitotic cell cycle, assembly of the pre-replicative complex, metabolism of polyamines, degradation of beta-catenin by the destruction complex, tumor necrosis factor receptor 2 non-canonical NF-kB pathway, AUF1 (hnRNP D0) binds and destabilizes RNA, regulation of apoptosis and 3 pathways associated with aspects of hedgehog signaling. To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (Figure 4A?). Surprisingly, this class of tumors fell into two major categories. The first contains many mutations, in nearly all of the hallmark pathways of this cluster, and the second subgroup contains far fewer mutations, but typically at least one mutation of any one of a large number of proteasome genes that happen to be common to the other 9 pathways. Moreover, it is apparent that some of these proteasome genes have a typically tissue-of-origin (or cancer-specific) mutation signature, suggesting a) that proteasome mutations could potentially substitute for a handful of more targeted mutations to these other pathways and b) that this could represent a novel class of cancer-promoting mutations that is obscured by virtue of having an abundance of targets with tissue-specific expression.

To further investigate this latter hypothesis, we examined the expression of the proteasome genes in this cohort of tumors, relating expression to percent of mutated samples by cancer (Figure 4B). We found a significant positive association between number of mutated samples (I’m making this up now, you have to determine whether it’s true and correct the record) and mean expression in the cancers by tissue-of-origin, thus confirming our suspicion that the proteasomal genes in class IV represent a novel class of tissue-specific specific cancer-promoting genes with a potential common mechanism.

### Shared mutated genes/pathways

### Oncoprint stuff

### Determinants of Metastatic vs non-metastatic tumors

Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure 2) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure 1C). Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown). We reasoned that the clusters might be different enough to dilute cluster-specific enrichment.

We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (Table 1). A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster. For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see supplemental figure S5) but not in 9 and 10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, Table 1). This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure 2) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7. Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.

There were also a number of pathways that were enriched with lower specificity (among 4 or more clusters) which were found to be enriched in metastases. We identified 18 such pathways, and of these 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing”. Of these pathways, we identified mutations in Sumoylation for 100% of the metastatic samples of clusters 3, 6, and 7 (Table 1).

The enrichment we observe may be consistent with an interpretation in which the pathways themselves are promoting a metastatic phenotype, or at the very least that tumors bearing such mutations are more likely to turn metastatic since we don’t know the order in which they were acquired. The fact that many of the metastasis enrichments are cluster-specific and have a tendency to overlap with cluster-specific pathways from non-metastatic tumors of other clusters in a complementary fashion is suggestive that there is an order of progression with the metastatically enriched pathways representing later events.

### Correlation of microenvironment with tumor class

## Materials & Methods

## Discussion

### Classification of tumors independent of tissue-of-origin

The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples (pmid 29200198). And that is not accounting for the fact that >30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential. The bioinformatic challenge therefore is to define the problem precisely and then create a new dataset that addresses this problem. Hence, we define cancer as a malady of biochemical processes and pathways, and generated a dataset that scores each of these essential pathways as mutated or not.

It is important that we chose point mutations in exome sequencing data for this task because we know unambiguously in most cases the target gene. In addition, we reduce pathway state to a boolean value because it does not matter for our purposes whether a single tumor is “enriched” for pathway mutations, chiefly because such metrics favor larger pathways over smaller ones. The data are likely to be noisy for several reasons. First, point mutations can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in genes that are potentially either activators or repressors. In the current model, we accept this uncertainty and assume mutations are generically disruptive to the pathway. Second, low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate (pmid 24132290, 29662161). To address this issue, we identified genes with low expression average for each type of cancer and eliminated them for that cancer only. Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis (pmid 23146897). We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.

We hypothesize that cancer is the result of dysfunction of a limited number of basic cellular processes that are common to all living human cells, and thus all tumors belong on a spectrum of mutations regardless of tissue of origin. An alternative hypothesis is that every tumor belongs to one of a large number of syndromes that share some mechanisms and superficial similarity in treatment strategies. We refer to these as the “single disease” and the “multiple syndrome” hypotheses. The latter seems to have gained momentum in recent years since the publication of Encode, Roadmap and TCGA consortium papers, perhaps owing to the intractable complexity of genomic data in all its forms. The inescapable conclusion thus far is that all these data point to tissue-of-origin as the most important driver of tumor behavior. Our observations support the view of all tumors as manifestations of disease of cellular differentiation and metabolism, but by no means refute entirely the multiple syndrome hypothesis.

We find that most tumors segregate into one of (at least) three biologically distinct clusters, regardless of the cancer type. Some cancer types (e.g. Pancreatic adenocarcinomas in the RAS arm cluster IIIa, and breast cancers being excluded from the same cluster) are unevenly distributed toward a specific class, we could not identify any type of cancer that was exclusive to a tumor class. Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, etc. (see Figure 1E and Supplemental Figure S3). Indeed, it would be surprising if we did not observe bias for some cancers, as it has long been known that certain treatment regimens are likely to work for a majority of patients of a given cancer type. Nonetheless, the basis for some tumors being treatment-refractory in spite of clinical indications remains elusive. If supported by future studies, our analysis suggests general mechanisms underlying tumor evolution regardless of tissue or histological or cellular subtype and may form the basis for future separation of patients into different treatment regimens.

It could be that the classes we identified are consistent with a signature of overall disease progression, such that successive, adjacent classes on the UMAP projection exhibit increasing specificity. We identified no such trend and each class instead was associated with unique combinations of pathways. In class III--which contains mutations in extracellular signaling components--we identified subgroups (class IIIa and IIIb) which correspond to the different intracellular (downstream) components of the signaling pathways. Also consistent with the groups reflecting basic differences in biology rather than stages of tumor evolution, we found that our classifications were largely independent of the number of mutations or clinical stage (Figure 1B & C). In our view, these observations suggest that cancers have available to them at least three separate mechanisms to escape the selective pressures they are exposed to in situ.

Finally, we must remark on two of the classes, I and V, which for which we did not find distinctive associations with pathways. Class I has a relatively low level of mutated pathways overall, although it is broadly enriched in many of the same tumor-promoting pathways common to the other groups. We don’t believe this to be due to a true difference in stage of tumor evolution, as our data show clearly that this class is as likely to contain stage IV as it is to contain stage I tumors. We speculate this class represents a group of tumors with aberrations in methylation, gene expression, copy number or other genome rearrangements which we excluded from the present analysis. In the future, we hope to find ways to incorporate these other data types into a pathway-centric analysis as we have done here for point mutations and indels. Class V represents a group of hypermutated tumors that harbor so many mutations that virtually no pathway is unaffected. This group highlights one of the limitations of the approach that we’ve taken, in that it is likely that a significant fraction of the samples flagged as “mutant” for each pathway are actually unaffected, and thus if the mutation rate is high enough, those samples have low information content. Similar to class I we speculate that a future analysis of other data types may resolve these tumors into more meaningful categories.

### Risk genes and tumor class

Certain mutations in risk genes increase risk for cancer quite dramatically. For example, carriers of mutations in BRCA1 and BRCA2 have increased risk for breast, ovarian, and prostate cancers (refs). The mechanism of risk in such cases is thought to be due to loss of heterozygosity (LOH) resulting in complete or near-complete loss of function(refs). In some cases it is known that carriers have a lower rate of chemo-resistance than non-carriers(refs), begging the question of whether carriers with LOH are exposed to distinctive selective pressure resulting in a constrained evolutionary outcome. If this hypothesis were true, we should have observed somatic mutations in BRCA1/2 genes clustering within one of our classes. Instead, BRCA mutations were evenly distributed amongst the classes (as in Figure 1E), suggesting that treatments or outcomes that distinguish between carriers and the general population are specific to BRCA gene biology. BRCA1/2 encode proteins whose function is involved in double-stranded break repair, and we speculate that genes involved in this and other aspects of genome stability are therefore upstream of the later events in tumor evolution that distinguish between classes II-IV. This further provides hope for combination therapies that account for carrier status and tumor class separately.

### (Include discussions of other germline LOH)

Our findings imply that there are separate processes in the etiology of cancer that can be broadly thought of as initiation, acquisition of cancer and metastasis. Initiation processes must include factors that relate to genome stability either directly or indirectly. Acquisition genes relate to selective pressures that all cells face; a complex web of signaling pathways governs cell division and migration in normal healthy cells, s.th. any cells that turn these pathways on in the absence of signal have a selective advantage. It is no wonder then that hormone-driven cancers coincide with menopause and advancement beyond middle age. With regard to metastasis….

### Two-step and 3-strike

### 2020-04-18 text/manuscript.Rmd
add references.bib; modify main text with bibtex refs

@@
 output: html_document
+bibliography: references.bib
 ---
@@
 
-### Authors: Peter T. Nguyen, Simon G. Coetzee, Corona I. Rosario, and Dennis J. Hazelett
+#### Authors: Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett
 
@@
 Some considerable progress has been made in analyzing these data.
-Statistical analyses identify hundreds of global and tissue-specific cancer driver genes (pmid 22759861, 23884480, 24390350, 26209800, 27911828, 31287869, 31363082) using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance.
-Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state (pmid 26559569).
+Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance.
+Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
 
 Other studies have been aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin.
-The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data (pmid 25109877).
+The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014].
 The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin.
-In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method (pmid 19759197), the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney (pmid 29625048).
+In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method [@shenIntegrativeClusteringMultiple2009], the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
 The authors of this study analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.
@@
 More recently, attempts have been made to consider pan-cancer mutations and other data outside the context of enrichment at the gene level.
-Horn et al (pmid 29200198) introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors.
+Horn et al [@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors.
 This method proves powerful especially when a small number (on the order of hundreds) of tumors are involved.
@@
 
-The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples (pmid 29200198).
+The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples [@hornNetSigNetworkbasedDiscovery2018].
 And that is not accounting for the fact that >30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential.
@@
 In the current model, we accept this uncertainty and assume mutations are generically disruptive to the pathway.
-Second, low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate (pmid 24132290, 29662161).
+Second, low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018].
 To address this issue, we identified genes with low expression average for each type of cancer and eliminated them for that cancer only.
-Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis (pmid 23146897).
+Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis [@parkGenomicEvidenceElevated2012].
 We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.
@@
 ### Two-step and 3-strike
+
+## References:

### 2020-04-19 text/manuscript.Rmd
figure references added; header amended add/format TOC and style plain html

@@
 ---
-title: "TTmanuscript"
-author: "Dennis Hazelett"
-date: "4/18/2020"
-output: html_document
+title: "A molecular taxonomy of tumors independent of tissue-of-origin"
+author: "Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett"
+output: 
+  html_document:
+    fig_captions: TRUE
+    number_sections: FALSE
+    toc: TRUE
+    toc_float: TRUE
+    toc_depth: 2
+    code_folding: hide
+    theme: flatly
 bibliography: references.bib
@@
 
-# Title: A molecular taxonomy of tumors independent of tissue-of-origin
-
-#### Authors: Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett
-
-## Abstract: 
+# Abstract
 Cancer is a disease that affects metabolism and other basic cellular processes common to all cells.
@@
 
-## Introduction:
+# Introduction
 
@@
 
-## Results:
+# Results
 
-### Classification of tumors by pathway disruption reveals 3 major groups
+## Classification of tumors by pathway
 
@@
 
-We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP (arXiv:1802.03426).
+We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP [@McInnes2018].
 Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings (Figure 1A).
 We call these tentatively by roman numerals I-V.
-While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure 1A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method.
+While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure \@ref(fig:umap-global-trends)A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method.
 Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14).
@@
 While there is a subtle trend, mostly visible for groups I and V (violin plot, Figure 1B), mutation count does not appear to be the dominant feature.
-To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure 1C).
+To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure \@ref(fig:umap-global-trends)C).
+
+```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE}
+set.seed(21)
+plot(x=rnorm(1000), y=rnorm(1000))
+```
 
 The 15 hierarchical clusters were largely free of any tissue-of-origin bias, so we suspected that our 5 groups would be similarly independent from tissue of origin.
-To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures 1D).
+To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures \@ref(fig:umap-global-trends)D).
 Full projections for all 23 cancers are presented in Supplementary Figure S2.
@@
 BRCA1/2 genes for example are key for DNA double-stranded break repair (ref) and confer risk for breast, prostate and ovarian cancers (refs).
-Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure 1E).
+Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure \@ref(fig:umap-global-trends)E).
 We made similar projections for the MMR genes MSH2, MSH6, MLH, PMS1 and PMS2; BRIP1, RAD15, CHEK2 and APC.
@@
 
-### Determinants of Tumor Class
+## Determinants of Class
 
 In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 381 pathways used to produce it (Supplementary Figure S5), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment.
-We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure 2A).
+We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure \@ref(fig:group-pathway-enrichment)A).
 Using this approach, we identified 6?
@@
 Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively.
-Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure 2B).
+Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure \@ref(fig:group-pathway-enrichment)B).
 Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution.
-Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure 2C).
+Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure \@ref(fig:group-pathway-enrichment)C).
 Class IV had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, beta-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others.
 As groups III and IV were the most distinct of the 5, we chose to investigate these two groups further.
+
+```{r group-pathway-enrichment, fig.cap="Placeholder for figure 2", results=FALSE}
+set.seed(28)
+plot(density(rnorm(100)))
+```
 
@@
 Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment.
-We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure 3 or supplement figure?).
-We find….(please do this analysis)
+We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure \@ref(fig:redundant-genes) or supplement figure?).
+We find… (please do this analysis).
+
+```{r redundant-genes, fig.cap="Placeholder for figure 3", results=FALSE}
+set.seed(29)
+hist(rnorm(100))
+```
 
@@
 
-### Shared mutated genes/pathways
+## Shared mutated genes/pathways
 
-### Oncoprint stuff
+## Oncoprint stuff
 
-### Determinants of Metastatic vs non-metastatic tumors
+## Metastatic _vs._ non-metastatic tumors
 
-Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure 2) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure 1C).
+Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure \@ref(fig:group-pathway-enrichment)) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure \@ref(fig:umap-global-trends)C).
 Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown).
@@
 This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, Table 1).
-This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure 2) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7.
+This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure \@ref(fig:group-pathway-enrichment)) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7.
 Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters.
@@
 
-### Correlation of microenvironment with tumor class
+## Correlation of microenvironment with tumor class
 
-## Materials & Methods
+# Materials & Methods
 
-## Discussion
+# Discussion
 
-### Classification of tumors independent of tissue-of-origin
+## Classification of tumors independent of tissue-of-origin
 
@@
 
-### Risk genes and tumor class
+## Risk genes and tumor class
 
@@
 
-### (Include discussions of other germline LOH)
+## (Include discussions of other germline LOH)
 
@@
 
-### Two-step and 3-strike
+## Two-step and 3-strike
 
-## References:
+# References

### 2020-05-30 text/manuscript.Rmd
update manuscript and references, remove roman numeral classes

--- 2020-04-18 text/manuscript.Rmd
@@
 ---
-title: "TTmanuscript"
-author: "Dennis Hazelett"
-date: "4/18/2020"
-output: html_document
+title: "A molecular taxonomy of tumors independent of tissue-of-origin"
+author: "Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett"
+output: 
+  bookdown::html_document2:
+    fig_captions: TRUE
+    number_sections: FALSE
+    toc: TRUE
+    toc_float: TRUE
+    toc_depth: 2
+    code_folding: hide
+    theme: flatly
+    smart: true
 bibliography: references.bib
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(rstan)
+library(foreach)
+library(sqldf)
+library(tidyverse)
+library(ggthemes)
 ```
 
-# Title: A molecular taxonomy of tumors independent of tissue-of-origin
-
-#### Authors: Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett
-
-## Abstract: 
+## Abstract
 Cancer is a disease that affects metabolism and other basic cellular processes common to all cells.
@@
 One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR.
-This class is neatly subdivided into PI3K/Akt and Ras branches.
+This class is neatly subdivided into PI3K/Akt and Ras signalling branches.
 Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling.
 Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling.
-Intriguingly, we identify a novel class of mutations in proteasome component proteins that may be sufficient to cause compound defects across multiple signaling pathways in this class.
-A third class is typified by defects in metabolism, especially of RNA and protein biosynthesis.
+Intriguingly, we identify a novel class of mutations in proteasome component proteins that may cause defects across multiple key signaling pathways.
+Yet another class is typified by defects in metabolism, especially of RNA and protein biosynthesis.
 In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors.
-With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found.
-We discuss the implications for tumor biology and treatment.
+With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found in.
+We discuss implications for tumor biology and treatment.
 
-## Introduction:
+## Introduction
 
-Recent advances in technology have made possible the study of cancer at unprecedented levels of detail and scale.
+Recent advances in technology have made the study of cancer possible at unprecedented levels of detail and scale.
 A combination of array and next-generation sequencing technologies has made it possible to assay the germline variation and somatic mutations, expression profile, methylation status, copy number variation.
-To take advantage of these tools, several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) endeavored to sequence large number of tumors (order of thousands) and collect data from multiple assays that could be analyzed together, with an eye towards integration and a general understanding of the mechanisms of cancer.
-These efforts have consisted chiefly of recognizing and quantifying the importance of individual genes in cancer etiology as the product of positive selection processes, or classification of cancers relative to common patterns of aberration in the expression of those genes in various assays.
+To take advantage of these tools, several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) sequenced large numbers of tumors and collected data from multiple assays that could be analyzed together, with the goal of integration and increased general understanding of the mechanisms of cancer.
+These efforts have consisted chiefly of recognizing and quantifying the importance of genes in cancer etiology as the result of selection processes, or classification of cancers relative to common patterns of aberration in the expression of those genes in various assays.
 
-Some considerable progress has been made in analyzing these data.
+Some considerable progress has been made analyzing these data.
 Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance.
@@
 
-Other studies have been aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin.
+Other studies have aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin.
 The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014].
 The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin.
-In a second study involving 33 cancer types and a much greater number of tumors using a newer iCluster method [@shenIntegrativeClusteringMultiple2009], the same authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
-The authors of this study analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.
+In a second study involving 33 cancer types and a much greater number of tumors using a newer algorithm (iCluster) [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
+The authors analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.
 
@@
 In this study, we attempted to address this challenge by focusing on discrete molecular pathways as the unit of disruption in cancer.
-This operation has the virtue of stripping all cell-type-specific information from the mutation data.
+Our approach has the virtue of stripping all cell-type-specific information from the mutation data and generalizing gene-level information to non-tissue specific cell-biology phenotypes.
 We use the resulting statistic to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of tissue of origin.
-Furthermore, tumors of varying cancer type and/or histological subtype may merely reflect the likely cell-of-origin that gave birth to them.
 
-## Results:
+## Results
 
-### Classification of tumors by pathway disruption reveals 3 major groups
+### Classification of tumors by pathway
 
-To study cancer pathways we obtained a set of 7,607 tumor samples (x donors) from GDC.
+To study cancer pathways we obtained a set of 7,607 tumor samples from GDC.
 We wanted to classify tumors by pathway disruption considered different types of data available from TCGA.
@@
 Therefore, we focused on all missense, nonsense, frameshift, and splicing mutations.
-We selected 381 Reactome pathways (reactome.org) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories (e.g.
-“transcription factors”) or disease associations (e.g.
-“mutated in colon cancer”) and filtered our gene list on membership in these pathways.
-We then compiled a mutation matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations.
+We selected 377 Reactome pathways [reactome.org](https://reactome.org/) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on membership in these pathways.
+We then compiled a matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations.
+
+```{r data-set-creation, eval=FALSE}
+
+```
 
 In order to classify tumors using this dataset, we used unsupervised clustering methods.
-Preliminary work with hierarchical clustering produced a set of 15 clusters (optimized with clusterProfiler(ref) was very promising, as the tumors appeared to cluster largely independent of tissue of origin (Supplementary Figure S1A).
-To determine whether our clusters were differentiated from random selections of tumors we plotted Kaplan Meier survival curves for each of the clusters first, and compared the results to randomized sets of tumors of equal-sized proportions (Figure S1B, C).
-In addition we fitted a Cox proportional hazards test model, accounting for age of diagnosis and cluster membership and found that 8 of the clusters were significantly different from a reference cluster (cluster 1), suggesting that cluster membership could in principle be rooted in biological differences affecting survival (Supplementary Table S1).
-These differences were small relative to differences in cancer by tissue of origin (Figure S1D).
+Preliminary work with hierarchical clustering produced a set of 15 clusters (optimized with clusterProfiler[@clusterProfiler] was very promising, as the tumors appeared to cluster largely independent of tissue of origin (**Supplementary Figure S1 A**).
 
-We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP (arXiv:1802.03426).
-Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings (Figure 1A).
-We call these tentatively by roman numerals I-V.
-While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure 1A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method.
-Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14).
+```{r tumor-clusters}
+
+```
+
+To determine whether our clusters were differentiated from random selections of tumors we plotted Kaplan Meier survival curves for each of the clusters first, and compared the results to randomized sets of tumors of equal-sized proportions (**Supplementary Figures S1B-C**).
+In addition we fitted a Cox proportional hazards test model, accounting for age of diagnosis and cluster membership and found that 8 of the clusters were significantly different from a reference cluster (cluster 1), suggesting that cluster membership could in principle be related to biological differences affecting survival (**Supplementary Table S1**).
+These differences were modest relative to differences in cancer by tissue of origin (**Supplementary Figure S1D**).
+
+```{r preliminary-survival}
+
+```
+
+We noted intriguing similarities across different clusters, which motivated us to use the dimensional reduction analysis (DRA) tool UMAP [@McInnes2018;@art-of-tsne].
+Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings.
+To represent the spatial distribution in a more principled manner, we used hdbscan[@hdbscan] which resulted in 12 distinct new clusters projected onto the UMAP (**Figure 1A**).
+Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5 tumors from the hierarchical method predominate in cluster 1 of the DRA) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14 of the hierarchical method predominate in cluster 8 of the DRA).
 To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only).
-While there is a subtle trend, mostly visible for groups I and V (violin plot, Figure 1B), mutation count does not appear to be the dominant feature.
-To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure 1C).
+While there is a subtle trend, mostly visible for groups I and V (violin plot, **Figure 1B**), mutation count does not appear to be the dominant feature to distinguish the tumors in different classes of the DRA.
+To test whether the 5 groups represent progression by stage (which are determined clinically by size, though different criteria exist for different cancers), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every DRA cluster, about equally (**Figure \@ref(fig:umap-global-trends)C**).
 
-The 15 hierarchical clusters were largely free of any tissue-of-origin bias, so we suspected that our 5 groups would be similarly independent from tissue of origin.
-To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures 1D).
-Full projections for all 23 cancers are presented in Supplementary Figure S2.
+```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
+set.seed(21)
+plot(x=rnorm(1000), y=rnorm(1000))
+```
+
+![**Figure 1 Classification of tumors independent of tissue of origin.** _**A.** UMAP: dimensional reduction analysis and clustering of pathway disruption data, followed by kNN assigns tumors to 12 distinct clusters.
+**B.** Violin plots of total mutations by pathway cluster reveal modest differences in overall mutation rate, except for cluster 8 which contains hypermutated tumors.
+In these tumors most pathways $>90%$ have at least one gene with a mutation.
+**C.** Tumors shaded by pathological stage on the UMAP from panel A do not favor any particular cluster.
+**D.** Metastatic status projected onto the UMAP clusters from panel A.
+Metastatic tumors are equally distributed among the clusters.
+**E.** Exemplars of 4 cancer types projected onto the UMAP clusters.
+Some biases are evident, but all cancers are found in every UMAP cluster.
+**F.** BRCA mutations, which are thought to increase genome instability, are found in every UMAP cluster._](../figures/fig1A_complete.png)
+
+To determine whether the 12 DRA clusters are independent from tissue of origin, we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the original UMAP (**Figures \@ref(fig:umap-global-trends)D**).
+While some modest bias is evident in some cancers, No tissue specific cancer exclusively found in one cluster.
+Full projections for all 23 cancers are presented in **Supplementary Figure S2**.
 Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies.
-One hypothesis to explain the lack of bias by cancer type is that the groupings are representative of these molecular subtypes.
-To probe this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP.
-Perhaps surprisingly we did not detect any group biases for these or any other subtype annotations in GDC (see Supplementary Figure S3 for the full set of projections), suggesting that these subtypes are reflective of differences either in specialized cell-type-of-origin or other cancer risk mechanisms.
+One hypothesis to explain the lack of bias by cancer type in these clusters is that the groupings are representative of these molecular subtypes.
+To test this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP.
+Perhaps surprisingly we did not observe any exclusive segregation by pathway for these subtype annotations in GDC (**Figure 1E**).
+We also projected histological subtype data for the remaining cancers with at best modest biases evident (see **Supplementary Figure S3** for the full set of projections), suggesting that these subtypes are as diverse with respect to our pathway disruption analysis as the parent cancer types.
 
@@
 However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length.
-BRCA1/2 genes for example are key for DNA double-stranded break repair (ref) and confer risk for breast, prostate and ovarian cancers (refs).
-Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure 1E).
-We made similar projections for the MMR genes MSH2, MSH6, MLH, PMS1 and PMS2; BRIP1, RAD15, CHEK2 and APC.
-None of these genes exhibited any remarkable specificity with respect to group (except maybe APC??) (Supplementary Figure S4).
-To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the TET2 and TET3 genes, plus telomerase 1, 2, 3, and observed similar lack of segregation by group (Supplementary Figure S4).
+_BRCA1/2_ genes for example are key for DNA double-stranded break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for breast, prostate and ovarian cancers (refs).
+The mechanism of risk is thought to involve loss of heterozygosity [@brca-loh], so we projected the somatic mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe segregation of these mutations into specific clusters (**Figure \@ref(fig:umap-global-trends)E**).
+We made similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_, _PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_.
+None of these genes except for _APC_ exhibited any remarkable specificity with respect to cluster assignment (**Supplementary Figure S4**).
+To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and observed similar lack of segregation by cluster (**Supplementary Figure S4**).
 
-### Determinants of Tumor Class
+### Determinants of Class
 
-In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 381 pathways used to produce it (Supplementary Figure S5), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment.
-We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure 2A).
+In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 377 pathways used to produce it (**Supplementary Figure S5**), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment.
+We visualized the enrichment as a heatmap (**Figure \@ref(fig:group-pathway-enrichment)A**).
 Using this approach, we identified 6?
@@
 Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively.
-Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure 2B).
+Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (**Figure \@ref(fig:group-pathway-enrichment)B**).
 Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution.
-Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure 2C).
-Class IV had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, beta-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others.
-As groups III and IV were the most distinct of the 5, we chose to investigate these two groups further.
+Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (**Figure \@ref(fig:group-pathway-enrichment)C**).
+Cluster group 6-7 had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, $\beta$-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others.
+As cluster groups 9-12 and 6-7 were the most distinct of the 12, we chose to investigate these two groups further.
 
-Group III contained interesting sub-patterns of enrichment, which we subdivided further into class IIIa and class IIIb sequentially from right following the global naming system.
-Common to classes IIIa and b were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as EGFR, IGF1R, Insulin receptor, and ERBB2 and 4.
-In addition to these we observed SCF-Kit, DAP12, and Tie 2 signaling pathways, altogether suggesting that Class III tumors share mutations in extracellular signaling pathways generally.
+```{r group-pathway-enrichment, fig.cap="Placeholder for figure 2", eval=FALSE}
+set.seed(28)
+plot(density(rnorm(100)))
+```
+
+![**Figure 2.
+Pathway-specific enrichment within clusters.** _Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
+The annotation at left indicates the Reactome pathway group at the second level (check this)._](../figures/fig2B_short.png)
+
+Common to cluster groups 9-10 and 11-12 were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as _EGFR_, _IGF1R_, _Insulin receptor_, and _ERBB2_ and _-4_.
+In addition to these we observed _SCF-Kit_, _DAP12_, and _Tie 2_ signaling pathways, altogether suggesting that cluster group 9-12 tumors share mutations in extracellular signaling pathways generally.
+
+```{r redundant-genes, fig.cap="Placeholder for figure 3", results=FALSE}
+set.seed(29)
+hist(rnorm(100))
+```
+
 Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment.
-We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure 3 or supplement figure?).
-We find….(please do this analysis)
+Upon closer examination the split between cluster groups 9+10 and 11+12 is largely driven by mutations in _PI3K_ and its homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary Figure S4**).
+Furthermore, clusters 9 and 11 are distinguished from 10 and 12 by general lack of P53 mutations (compare **Figure 1A** with the _TP53_ panel of **Supplementary Figure S4**).
 
-Intriguingly, we noticed that the pathways enriched for class IIIa and b were related to RAS and PI3K/Akt signaling, respectively, which potentially represent the downstream components of the extracellular signaling pathways discussed above.
-Moreover, class IIIb is also enriched in several immune-associated pathways including for both innate immunity and T cell receptor signaling, suggesting an interesting correlation between the PI3K/Akt subgroup and potential loss (or hyperactivation) of immune function.
-To address this possibility, we isolated the PI3K/Akt, RAS and immune/TCR pathways and assessed whether the mutations tended to be biased toward likely inactivating or activating mutations ….
-
-Class IV exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors between class III and IV on the UMAP projection that have pathway characteristics of both.
-The top ten distinguishing pathways of Class IV tumors were cellular response to hypoxia, regulation of mitotic cell cycle, assembly of the pre-replicative complex, metabolism of polyamines, degradation of beta-catenin by the destruction complex, tumor necrosis factor receptor 2 non-canonical NF-kB pathway, AUF1 (hnRNP D0) binds and destabilizes RNA, regulation of apoptosis and 3 pathways associated with aspects of hedgehog signaling.
+In contrast to cluster group 9-12, the group represented by clusters 6 & 7 exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors spatially distributed between cluster groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics of both.
+The top ten distinguishing pathways of class 6/7 tumors were cellular response to hypoxia, regulation of mitotic cell cycle, assembly of the pre-replicative complex, metabolism of polyamines, degradation of $\beta$-catenin by the destruction complex, tumor necrosis factor receptor 2 non-canonical _NF-$\kappa$B_ pathway, _AUF1_ (hnRNP D0) binds and destabilizes RNA, regulation of apoptosis and 3 pathways associated with aspects of hedgehog signaling.
 To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (Figure 4A?).
-Surprisingly, this class of tumors fell into two major categories.
-The first contains many mutations, in nearly all of the hallmark pathways of this cluster, and the second subgroup contains far fewer mutations, but typically at least one mutation of any one of a large number of proteasome genes that happen to be common to the other 9 pathways.
-Moreover, it is apparent that some of these proteasome genes have a typically tissue-of-origin (or cancer-specific) mutation signature, suggesting a) that proteasome mutations could potentially substitute for a handful of more targeted mutations to these other pathways and b) that this could represent a novel class of cancer-promoting mutations that is obscured by virtue of having an abundance of targets with tissue-specific expression.
-
-To further investigate this latter hypothesis, we examined the expression of the proteasome genes in this cohort of tumors, relating expression to percent of mutated samples by cancer (Figure 4B).
-We found a significant positive association between number of mutated samples (I’m making this up now, you have to determine whether it’s true and correct the record) and mean expression in the cancers by tissue-of-origin, thus confirming our suspicion that the proteasomal genes in class IV represent a novel class of tissue-specific specific cancer-promoting genes with a potential common mechanism.
-
-### Shared mutated genes/pathways
+Surprisingly, this class of tumors was not predominated by 1 or more highly mutated genes across all cancers.
+Instead, it appears that there is a whole class of genes in the proteosomal pathway whose activity contributes to all the aforementioned pathways, and the majority of tumors in these classes have a single mutation in one of these genes.
+Nonetheless, a significant fraction (~1/4 <-- need numbers!) do not have proteosomal mutations but instead harbor mutations in genes representing a majority of the same pathways, suggesting that most if not all of the pathways are individually important and that proteosomal mutations may be sufficient to drive the net defect of tumors in these clusters.
+We annotated the tumors in this class according to tissue of origin to determine whether proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and indeed we could observe some clustering.
+These data could explain how tissue-specific gene mutation signatures might result in general cancer phenotypes by converging at the pathway level.
 
@@
 
-### Determinants of Metastatic vs non-metastatic tumors
+### Metastatic _vs._ non-metastatic tumors
 
-Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure 2) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure 1C).
+Following the same logic we used to investigate cluster-specific enrichment of pathways (**Figure \@ref(fig:group-pathway-enrichment)**) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (**Figure \@ref(fig:umap-global-trends)C**).
 Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown).
@@
 
-We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (Table 1).
+We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (**Table 1**).
 A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster.
-For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see supplemental figure S5) but not in 9 and 10.
-This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, Table 1).
-This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure 2) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7.
+For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see **Supplementary figure S5**) but not in 9 and 10.
+This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, **Table 1**).
+This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (**Figure \@ref(fig:group-pathway-enrichment)**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
 Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters.
@@
 We identified 18 such pathways, and of these 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing”.
-Of these pathways, we identified mutations in Sumoylation for 100% of the metastatic samples of clusters 3, 6, and 7 (Table 1).
+Of these pathways, we identified mutations in Sumoylation for 100% of the metastatic samples of clusters 3, 6, and 7 (**Table 1**).
 
@@
 
-### Correlation of microenvironment with tumor class
+### Survival Analysis
 
+In order to understand whether our cluster identifications add any information to the prognositic value embedded in cancer type, we tested models of patient survival that accounts for cancer type and age at diagnosis.
+Initial models did not properly account for the trajectory of patient survival independent of age of diagnosis but showed some promise in distinguishing among classes for one or two of the deadliest cancer types.
+Therefore, we used mortality data from the CDC to accurately model the normal trajectory of patient survival independent of disease status, including gender as a factor.
+The resulting model is shown as a comparison with the census data in **Figure 6A.**
+
+![**Figure 6A.
+Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
+The orange and green curves show the model fits._](../figures/fig6A.png)
+
+Using this base survival model, we constructed a likelihood function with parameters describing the overall lethality of each cancer type as a modifier $k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of $k$ and optimized model fit using stan[@stanref].
+The $k$ in our model can be taken as, roughly, the death rate relative to the baseline rate in **Figure 6A**.
+Thus, a value of $k=2$ means that a given cancer is twice as deadly.
+The posterior distributions for the 23 cancers are shown in **Figure 6B**.
+
+![**Figure 6B**](../figures/fig6B.png)
+
+![**Figure 6C**](../figures/fig6C.png)
 ## Materials & Methods
+
+### Classification of Tumors
+
+### Enrichment Calculations
+
+### Survival Models
 
@@
 We find that most tumors segregate into one of (at least) three biologically distinct clusters, regardless of the cancer type.
-Some cancer types (e.g.
-Pancreatic adenocarcinomas in the RAS arm cluster IIIa, and breast cancers being excluded from the same cluster) are unevenly distributed toward a specific class, we could not identify any type of cancer that was exclusive to a tumor class.
-Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, etc.
-(see Figure 1E and Supplemental Figure S3).
+Some cancer types ( _e.g._ Pancreatic adenocarcinomas in the RAS arm cluster IIIa, and breast cancers being excluded from the same cluster) are unevenly distributed toward a specific class, we could not identify any type of cancer that was exclusive to a tumor class.
+Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1E** and **Supplemental Figure S3**).
 Indeed, it would be surprising if we did not observe bias for some cancers, as it has long been known that certain treatment regimens are likely to work for a majority of patients of a given cancer type.
@@
 We identified no such trend and each class instead was associated with unique combinations of pathways.
-In class III--which contains mutations in extracellular signaling components--we identified subgroups (class IIIa and IIIb) which correspond to the different intracellular (downstream) components of the signaling pathways.
-Also consistent with the groups reflecting basic differences in biology rather than stages of tumor evolution, we found that our classifications were largely independent of the number of mutations or clinical stage (Figure 1B & C).
+In classes 9-12 --- which contain mutations in extracellular signaling components --- we identified subgroups (class 9-10 and 11-12) which correspond to different intracellular signaling pathways.
+Also consistent with the groups reflecting basic differences in biology rather than stages of tumor evolution, we found that our classifications were largely independent of the number of mutations or clinical stage (**Figures 1B-C**).
 In our view, these observations suggest that cancers have available to them at least three separate mechanisms to escape the selective pressures they are exposed to in situ.
 
-Finally, we must remark on two of the classes, I and V, which for which we did not find distinctive associations with pathways.
+Finally, we must remark on two of the classes, 1 and 8, which for which we did not find distinctive associations with pathways.
 Class I has a relatively low level of mutated pathways overall, although it is broadly enriched in many of the same tumor-promoting pathways common to the other groups.
@@
 Certain mutations in risk genes increase risk for cancer quite dramatically.
-For example, carriers of mutations in BRCA1 and BRCA2 have increased risk for breast, ovarian, and prostate cancers (refs).
-The mechanism of risk in such cases is thought to be due to loss of heterozygosity (LOH) resulting in complete or near-complete loss of function(refs).
+For example, carriers of mutations in _BRCA1_ and _BRCA2_ have increased risk for breast, ovarian, and prostate cancers (refs).
+The mechanism of risk in such cases is thought to be due to loss of heterozygosity (LOH) resulting in complete or near-complete loss of function [@brca-loh].
 In some cases it is known that carriers have a lower rate of chemo-resistance than non-carriers(refs), begging the question of whether carriers with LOH are exposed to distinctive selective pressure resulting in a constrained evolutionary outcome.
-If this hypothesis were true, we should have observed somatic mutations in BRCA1/2 genes clustering within one of our classes.
-Instead, BRCA mutations were evenly distributed amongst the classes (as in Figure 1E), suggesting that treatments or outcomes that distinguish between carriers and the general population are specific to BRCA gene biology.
-BRCA1/2 encode proteins whose function is involved in double-stranded break repair, and we speculate that genes involved in this and other aspects of genome stability are therefore upstream of the later events in tumor evolution that distinguish between classes II-IV.
+If this hypothesis were true, we should have observed somatic mutations in _BRCA1/2_ genes clustering within one of our classes.
+Instead, _BRCA_ mutations were evenly distributed amongst the classes (as in **Figure 1E**), suggesting that treatments or outcomes that distinguish between carriers and the general population are specific to _BRCA_ gene biology.
+_BRCA1/2_ encode proteins whose function is involved in double-stranded break repair, and we speculate that genes involved in this and other aspects of genome stability are therefore upstream of the later events in tumor evolution that distinguish between clusters 2-5.
 This further provides hope for combination therapies that account for carrier status and tumor class separately.
@@
 It is no wonder then that hormone-driven cancers coincide with menopause and advancement beyond middle age.
-With regard to metastasis….
+With regard to metastasis.
 
@@
 
-## References:
+## Supplementary Information
+
+![**Supplementary Figure S1A.
+K-means clustering of booleanized pathway disruptions.** _Each column represents a single tumor, each row a pathway.
+In the heatmap below, red colored fields indicate a mutation in a pathway, and gray indicates no mutations in pathway genes.
+In the annotation rows above the heatmap, annotations have been spread out by cancer type to reveal the distributions across clusters._](../figures/figureS1A.png)
+
+![**Supplementary Figure S1B-S1D.
+Kaplan Meier (KM) curves.** _B.
+KM plots of k-means clustered pathway data for GDC samples.
+C.
+KM plots of randomly permuted labels on the same samples.
+D.
+KM plots of GDC samples by cancer type._](../figures/figS1B_D.png)
+
+![**Supplementary Figure S2.
+Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._](../figures/figS2.png)
+
+![**Supplementary Figure S3.
+Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_](../figures/figS3.png)
+
+![**Supplementary Figure S4.
+Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_](../figures/figS4.png)
+
+## References

### 2020-07-01 text/manuscript.Rmd
rewrite main draft add List of figures text file update references

@@
 title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett"
+author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
 output: 
@@
     toc_float: TRUE
-    toc_depth: 2
+    toc_depth: 3
     code_folding: hide
@@
 ---
+
+###### Institutional affiliations: 1.
+The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
+2.
+Lakeland Applied Sciences LLC, Los Angeles, CA.
+3.
+Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
+
+```{r echo=FALSE, eval=FALSE}
+  bookdown::pdf_book:
+    number_sections: FALSE
+    toc: FALSE
+    keep_tex: yes
+```
 
@@
 Cancer is a disease that affects metabolism and other basic cellular processes common to all cells.
-In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying theories elude the field.
-To address this challenge we reduced omics mutation and expression profiling data to pathway memberships.
-We found that tumors can be classified according to pathway disruption in twelve distinct classes orthogonal to a tissue of origin (including histological subtypes), number of mutations and clinical stage.
-Moreover, known germline risk mutations such as BRCA1/2 are equally distributed among the classes.
-One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR.
-This class is neatly subdivided into PI3K/Akt and Ras signalling branches.
-Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling.
-Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling.
-Intriguingly, we identify a novel class of mutations in proteasome component proteins that may cause defects across multiple key signaling pathways.
-Yet another class is typified by defects in metabolism, especially of RNA and protein biosynthesis.
-In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors.
-With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found in.
-We discuss implications for tumor biology and treatment.
+In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying principles elude the field.
+To address this challenge we reduced tumor mutation data to pathway memberships, using a highly filtered set of pathways representative of basic cellular functions from metabolism and DNA replication to cell-cell communication.
+We find that tumors can be classified according to pathway disruption in twelve classes entirely distinct from tissue of origin based on these criteria, and largely free of bias from mutational burden and clinical stage.
+We also show that these tumor classes have different prognoses in different tissues using a hierarchical model of longevity data from the CDC.
+As a consequence of our pathway-centric classification, we can also identify equivalent sets of class-specific gene mutations in different tissues.
+Moreover, many known germline risk mutations involved in genome stability such as BRCA1/2 and Rb1 are equally distributed among the classes, somewhat contrary to the expectation that such genes are representative of disease-specific categories or subtypes.
+Similarly, we describe mutations that distinguish between metastatic tumors and non-metastatic tumors.
+In many cases the enriched pathways from one class in metastatic tumors are the class-specific pathways from another class, suggesting that metastasis may represent phenotypic convergence, therefore potentially identifying key pathways to focus on therapeutically for prevention of class-specific metastasis.
+In summary, our observations suggest that some genes act as drivers of genome instability while other class-specific genes confer increased fitness in combinatorial fashion.
+Thus, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on their distribution in tumors of every class.
 
@@
 Recent advances in technology have made the study of cancer possible at unprecedented levels of detail and scale.
-A combination of array and next-generation sequencing technologies has made it possible to assay the germline variation and somatic mutations, expression profile, methylation status, copy number variation.
-To take advantage of these tools, several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) sequenced large numbers of tumors and collected data from multiple assays that could be analyzed together, with the goal of integration and increased general understanding of the mechanisms of cancer.
-These efforts have consisted chiefly of recognizing and quantifying the importance of genes in cancer etiology as the result of selection processes, or classification of cancers relative to common patterns of aberration in the expression of those genes in various assays.
+Array and next-generation sequencing technologies have made it possible to assay germline and somatic mutations, expression profiles, DNA methylation, and copy number variations.
+To take advantage of these tools several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large numbers of tumors and collected data from multiple assays that could be analyzed together, with the goal of integration and increased understanding of the mechanisms of cancer.
+These efforts have focused on recognizing and quantifying the importance of genes in cancer etiology as the result of selection processes, or classification of cancers relative to common patterns of aberration in multiple assays.
 
-Some considerable progress has been made analyzing these data.
+Considerable progress has been made analyzing these data.
 Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance.
-Based on these studies, it has been proposed that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
+Based in part on studies like these, it has been estimated that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
 
-Other studies have aimed at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin.
-The first of these (the “Pan-Cancer-12” analysis) identified a clade of 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014].
-The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors (10%) could be reclassified independent of tissue-of-origin.
-In a second study involving 33 cancer types and a much greater number of tumors using a newer algorithm (iCluster) [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
-The authors analyzed these categories for pathway enrichment and noted multiple overlapping similarities between these latter categories.
+Other studies aim at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin.
+The first of these (the “Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014].
+The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors ($10\%$) could be reclassified independent of tissue-of-origin.
+In a second study involving 33 cancer types and a much greater number of tumors using a newer algorithm (iCluster) [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/ squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
 
 More recently, attempts have been made to consider pan-cancer mutations and other data outside the context of enrichment at the gene level.
-Horn et al [@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors.
-This method proves powerful especially when a small number (on the order of hundreds) of tumors are involved.
+Horn _et al._ [@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors.
+This method is powerful especially when applied to a small number (on the order of hundreds) of tumors.
 However, the larger problem of cancer etiology remains: what mutations are required, in what order, and are there specific combinations of delinquent cellular processes sufficient to cause cancer?
@@
 
-In this study, we attempted to address this challenge by focusing on discrete molecular pathways as the unit of disruption in cancer.
-Our approach has the virtue of stripping all cell-type-specific information from the mutation data and generalizing gene-level information to non-tissue specific cell-biology phenotypes.
-We use the resulting statistic to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of tissue of origin.
+In this study, we attempt to address these questions by focusing on discrete molecular pathways as the unit of disruption in cancer.
+Our approach strips all cell-type-specific information from the mutation data and generalizing gene-level information to non-tissue specific cell-biological pathway phenotypes.
+We use this rubric to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of their tissues of origin.
 
@@
 
-### Classification of tumors by pathway
+### A taxonomy of tumors based on disrupted molecular pathways
 
 To study cancer pathways we obtained a set of 7,607 tumor samples from GDC.
-We wanted to classify tumors by pathway disruption considered different types of data available from TCGA.
-We chose to start with somatic mutations in exome sequencing data because the affected target gene is known.
+We chose to start with somatic mutations in exome sequencing data because the affected gene is known almost unambiguously.
 Therefore, we focused on all missense, nonsense, frameshift, and splicing mutations.
-We selected 377 Reactome pathways [reactome.org](https://reactome.org/) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on membership in these pathways.
-We then compiled a matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations.
+In order to minimize bias from well-studied diseases and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on membership in these pathways (<mark>total of x genes;</mark> see methods for details).
+
+It does not matter for our purposes whether a single tumor is “enriched” for multiple pathway mutations, chiefly because such metrics favor larger pathways ( _i.e._ pathways with more member genes) over smaller ones, and non-expressed genes over expressed ones due to transcription coupled repair [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018; @Pervasive_lesion_segregation].
+Data selected in this manner are likely to be noisy for several reasons.
+First, point mutations can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in genes that act functionally as either activators or repressors.
+Within the scope of this study we must accept this uncertainty and assume mutations are generically disruptive to the cell regardless of whether they are activating or deactivating or whether they affect a gene that we might classify as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene").
+Second, as previously mentioned, we know that low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate due to transcription coupled repair.
+To address this issue, we identified genes with low expression in each type of cancer and eliminated them for that cancer type only (see methods for details).
+<strike>Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis [@parkGenomicEvidenceElevated2012].
+We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.</strike>
+
+After selecting our pathways and genes, we then compiled a matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations (this data is visualized in **figure 1a**).
 
@@
 
-In order to classify tumors using this dataset, we used unsupervised clustering methods.
-Preliminary work with hierarchical clustering produced a set of 15 clusters (optimized with clusterProfiler[@clusterProfiler] was very promising, as the tumors appeared to cluster largely independent of tissue of origin (**Supplementary Figure S1 A**).
+```{r kNNclusters, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE, fig.cap="**Classification of tumors by pathway disruptions.** a) Each of 377 selected Reactome pathways (columns) is classified as disrupted if one or more genes is mutated in the tumor sample (rows).
+Tumors are grouped by tissue of origin using standardized abrreviations from the TCGA project.
+b) MCA analysis was performed restricted to the 36 informative components.
+c) The resulting MCA was projected down to 3 dimensions for the purpose of density based clustering (only the first 2 dimensions are shown).
+Each dot represents a tumor d) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize the fraction of unassigned tumors.
+e) HDBSCAN clustered tumors were visualized on the UMAP projection; tumors are colored by cluster membership (see legend).
+This results in some tumors remaining unclassified (grey points).
+f) kNN was used to assign membership to the remaining tumors."}
+plot(1:10,1:10)
+```
 
-```{r tumor-clusters}
+![](../figures/FIG1_TEMP.png)
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1: Classification of tumors by pathway disruptions.** a) Each of 377 selected Reactome pathways (columns) is classified as disrupted if one or more genes is mutated in the tumor sample (rows).
+Tumors are grouped by tissue of origin using standardized abrreviations from the TCGA project.
+b) MCA analysis was performed restricted to the 36 informative components.
+c) The resulting MCA was projected down to 3 dimensions for the purpose of density based clustering (only the first 2 dimensions are shown).
+Each dot represents a tumor d) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize the fraction of unassigned tumors.
+e) HDBSCAN clustered tumors were visualized on the UMAP projection; tumors are colored by cluster membership (see legend).
+This results in some tumors remaining unclassified (grey points).
+f) kNN was used to assign membership to the remaining tumors.
+g) Violin plots of sample mutation counts for tumors in each class.
+h) Projection of cancer type identity onto UMAP.
+Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma.
+i) Projection of breast cancer subtypes onto the UMAP.
+j) projection of tumor stage onto the UMAP, regardless of cancer type.
+k) Projection of metastatic status onto the UMAP.
+Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
+
+In order to classify tumors using this dataset, we used multiple correspondence analysis (MCA) [@JSSv025i01].
+First we determined the number of dimensions containing useful information (**figure 1b**).
+Then we performed a UMAP analysis, both in order to summarize the MCA graphically, and as a preprocessing step to boost the performance of density based clustering.
+The resulting map was notable for its lobed structure, with several large, robust projections reproducible regardless of random seed setting.
+A representative version of this UMAP is shown in **figure 1c** as a 2D projection.
+Following this spatial mapping we attempted to define groupings of similar tumors within the spatial map using HDBSCAN, which performs hierarchical clustering and provides metrics of cluster stability and probabilities of cluster membership for each node.
+However, HDBSCAN is sensitive to several parameters; key for our analysis are the minimum number of tumor samples in a cluster that capture the maximum number of tumor samples, measured by probability of membership of $\geq 5\%$ in at least one class.
+Thus, we created a score metric as the fraction of classified tumors with max probability $< 5\%$ in one class and chose a cluster size of 95 to minimize the score function (**figure 1d**).
+HDBSCAN with these settings resulted in 12 distinct high-density clusters which we then projected onto the UMAP (**figure 1e**).
+This classified <mark>x out of 7,607 tumors</mark> but still resulted in a significant fraction of unclassified tumor samples.
+Since we ultimately wish to be able to classify any tumor using this scheme, we performed k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every tumor in the set and then lets the $k$ most similar tumors "vote" as to the identity of the query tumor sample based on their classification labels.
+We set $k$ to be the square root of the number of tumor samples (87).
+Using this method, we assigned cluster membership to the remaining tumors (**figure 1f**).
+
+#### Independence from tissue-of-origin.
+
+Having defined tumors in terms of their pathway disruption profile, we sought to understand whether different cancer types segregated into one or more predominant classes.
+To our surprise, most cancer types (with some notable exceptions) were not heavily biased in one class, and all well-represented cancer types had representative tumors in every class (for example **figure 1g** and full tumor profiles in **supplementary figure S1**, see also interactive media from **supplementary file 1 (html)**), suggesting that, in principle at least, our novel pathway-disruption classification method identifies molecular pathologies largely independent of tissue-of-origin considerations.
+As an example of one type of cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is distinguished  by KRAS mutations among others.
+But even PAAD comprises tumors from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that patients hosting these tumors have potentially different underlying molecular pathologies.
+
+#### Independence from molecular and histological subtype.
+
+Many cancers have molecular or histological subtypes defined based on gene expression or other -omics profiles or pathology lab results.
+These subtypes often have different standards of care owing to different overall drug sensitivity (or other factors).
+If the histological subtypes represent true molecular phenotypes, one predicts that histological subtypes should segregate with our pathway-based classes, therefore providing support for the classes as proxies for molecular pathology sub-typing.
+To our surprise, we find a similar result to the previous analysis of cancer types projected onto the UMAP of pathway disruptions.
+For example we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP.
+These are among the most heavily studied molecular subtypes in cancer, which each have different prognoses and standards of care.
+We did not observe any exclusive segregation by pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental file 1**).
+We also projected histological subtype data for the remaining cancers (see **supplementary figure S2** for the full set of projections); we find that the subtypes, though often biased towards one or more classes, are almost never exclusive to one of the classes.
+We interpret these data in aggregate to mean that our pathway disruption classes do not correspond to molecular subtypes within the parent cancer type.
+Furthermore, this result raises the possibility that, molecular and histo-pathological subtypes could represent different precursor cell origins within each cancer type.
+This interpretation still leaves open the possibility that different precursor cell types, being susceptible to different teratogens and selection pressures ( _e.g._ insulin dependence) could be biased toward one or more of the pathway disruption profiles.
+
+#### Independence from drivers of genome instability.
+
+There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes.
+However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length.
+_BRCA1/2_ genes for example are key for DNA double-stranded break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for breast, prostate and ovarian cancers (refs).
+The mechanism of risk is thought to involve loss of heterozygosity [@brca-loh], so we projected the somatic mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe segregation of these mutations into specific clusters (**figure 1j**).
+We made similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_, _PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_.
+None of these genes except for _APC_ exhibited any remarkable specificity with respect to cluster assignment (**supplementary figure S3**).
+To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and observed similar lack of segregation by cluster (**supplementary figure S3**).
+
+#### Independence of Stage, mutation count and mutation profile.
+
+Tumor staging is based on physico-pathological criteria, including diameter, which can vary greatly in importance between different tissues.
+Stage is used clinically as a proxy for advancement toward a more deadly state and metastasis.
+Given these assumptions, it is possible that more advanced tumors have common pathway disruption profiles, and the UMAP which features a series of lobe-like structures on a common backbone of tumor samples (**figure 1c**) could in principle also reflect progression through a series of stages.
+In support of this hypothesis, the backbone culminates in a group of tumors (class 8) which has nearly every pathway disrupted.
+However, outside of class 8 we don't observe an obvious trend in the mutation count across the backbone of the UMAP (**figure 1g**).
+Nonetheless, to test the hypothesis that the molecular-pathway disruption classes represent advancement through stages, we projected these data onto the UMAP.
+Similar to tissue of origin and other categories of tumor, we did not observe any bias amongst the stages into specific pathway disruption classes (**figure 1k**), suggesting that stage is not an determining factor in pathway disruption classes.
+
+Finally, as a measure of tumor advancement, metastasis is the condition in which certain phenotypic criteria are met.
+These phenotypes include (among others) loss of differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune system evasion and tissue invasiveness.
+To determine whether any of our classes correspond to an especially advanced stage of cancer accross tissue types, we projected the metastasis data onto our UMAP, and surprisingly, again, we observed an even distribution across classes (**figure 1l**).
+This final observation suggests that our pathway-disruption classification is dependent on particular combinations of gene mutations that hit different pathway profiles that can each give rise to advanced stages of disease and metastasis, regardless and independent of overall mutational burden.
+
+### Each class is defined by a specific set of pathway disruptions
+
+In order to visualize pathway disruption enrichment across all cancers (pan-cancer), we created oncoprint graphs analagous to the commonly use method, replacing genes with pathways (**figure 2**).
+A complete list of pathway disruptions with percent mutated samples and top genes is provided in **supplementary table ST1**).
+These reveal the broad importance of many well known pathways that are disrupted in cancer, including <mark>PI3K/Akt activation, DNA repair etc fix this with top 5 or so quoted pathway labels</mark>.
+The top third of pathways drop off gradually in overall mutation frequency, possibly owing to relative diversity among all tumors.
+
+![](../figures/figure2_temp.png)
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2: Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in which each column represents a tumor and each row represents a pathway.
+The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively.
+All 377 pathways are shown.
+Red signifies pathway disrupted pathways, grey signifies not disrupted.</p>
+
+To discover what pathways drive the UMAP clustering, we calculated enrichment for each pathway _in-class_ relative to all other classes combined (see methods) and ranked pathways for each class from highest to lowest enrichment.
+We visualized the enrichment as a heatmap (**figure 3**).
+Using this approach, we identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7, 0 pathways in class 8 (for which most pathways are putatively disrupted), 25 pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29 pathways in class 12.
+Some pairs of classes, in particular classes 6 & 7, classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways had enrichment.
+To explore the specific pathways marking each class, we projected disruptions for each of the 377 pathways onto the UMAP (**supplementary figure S4**).
+<mark>Class x</mark> was distinguished by metabolic pathways including RNA and protein biosynthesis (**supplementary figure S4**).
+Most class 4 enriched pathways were not unique to this class but shared in class 10 & 12, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution (as distinct from clinical ``stage'').
+Classes 9-12 have in common mutations in extracellular signaling, intracellular signaling, and immune-related pathways (see **figures 3** and **supplementary figure S4**).
+Cluster group 6-7 had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, $\beta$-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others.
+As cluster groups 9-12 and 6-7 were the most distinct of the 12, we chose to investigate these two groups further.
+
+![](../figures/fig2B_short.png)
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
+Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
+The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
+
+### Gene-Pathway redundancy contributes to cluster identity
+
+<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
+
+Since pathways like these share many common genes, we investigated whether tumors in this class  have multiple mutations across signaling pathways, or whether a small number of driver genes account for enrichment.
+The split between clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **supplementary figure S4**).
+Furthermore, clusters 9 and 11 are distinguished from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the _TP53_ panel of **Supplementary Figure S4**).
+
+In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors spatially distributed between cluster groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics of both.
+The top ten distinguishing pathways of class 6/7 tumors were "cellular response to hypoxia," "regulation of mitotic cell cycle," "assembly of the pre-replicative complex," "metabolism of polyamines," "degradation of $\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2 non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated with aspects of hedgehog signaling.
+To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (**figure 4**).
+Surprisingly, this class of tumors was not predominated by 1 or more highly mutated genes across all cancers.
+Instead, there is a whole set of genes in the proteosomal pathway whose activity contributes to all the aforementioned pathways, and the majority of tumors in these classes have a single mutation in one of these genes.
+Nonetheless, a significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have proteasomal mutations but instead harbor mutations in genes representing a majority of the same pathways, suggesting that most if not all of the pathways are individually important and that proteasomal mutations may be sufficient to drive the net effect of the other mutations in this cluster.
+We annotated the tumors in this class according to tissue of origin to determine whether proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and indeed we could observe some clustering.
+These data could explain how tissue-specific gene mutation signatures might result in general cancer phenotypes by converging at the pathway level.
+
+![](../figures/figure4_temp.png)
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
+The genes are organized according to which of the top distinguishing pathways for this class that they effect.
+Some genes affect multiple pathways, as indicated by the bars at left.
+The highlighted genes at right are all proteasomal subunits.</p>
+
+### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
+
+<mark>As discussed previously, prior efforts to extract common signatures from pan-cancer datasets met with difficulty in distinguishing tumor samples from their tissue-specific -omics data signatures.
+Given our pathway-disruption based classification, this raises the question, are tumor phenotypes driven entirely by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance thresholds), or a combination of both?
+To answer this question, we ranked the top cluster-specific pathways for each cluster, and then tabulated the most frequently mutated genes for each pathway as a function of cancer type (**Table 1** and **Supplementary tables ST2-ST13**).
+We find that in most clusters the top pathways are determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
+However in many cancers we find divergence outside these top driver mutations, wherein etc etc</mark>
+
+![](../figures/table1.png)
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
+
+### Enrichment of pathways in metastasis is class-specific
+
+Following the same logic we used to investigate cluster-specific enrichment of pathways (**figure 3**) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters ( _e.g._ **figure 1l**).
+Using all non-metastatic tumors as background, we found very low levels of enrichment, less than $10\%$, in a handful of pathways.
+We reasoned that the individual clusters might be too different to detect global metastasis enrichment signal given the small sample size (n = 215 metastatic tumor samples).
+
+Therefore, we calculated cluster-specific enrichment in metastatic tumors and found 99 enriched pathways across all clusters (**table 2**).
+A number of enrichments were found in multiple clusters represented by pathways that were already shown to be enriched in a neighboring cluster.
+For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and 10.
+This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, **table 2**).
+This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
+Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters.
+These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).
+Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors.
+Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.
+
+There were also a number of pathways that were enriched with lower specificity (among 4 or more clusters) which were found to be enriched in metastases.
+We identified 18 such pathways of which 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
+
+### Pathway disruption classes vary in short-term prognosis of survival
+
+If our molecular pathway disruption classes represent distinct biological states apart from tissue of origin, they may have different prognoses within cancer types or across all cancers.
+Kaplan-Meier plots reveal large differences between cancer types as expected, and some modest differences between our classes (not shown).
+These analyses are profoundly limited by confounding factors of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to name the obvious ones.
+To explore these ideas, we tested models of survival using public longevity data from the CDC, accounting for age, gender and ethnicity.
+To build on this relatively accurate model of baseline risk (see methods), we assume that individual cancer type captures some variation not accounted for in the longevity data ( _e.g._ all tissue-specific factors, including stage [see methods]), thus allowing us to avoid introducing uncertainty by modeling each of those factors explicitly.
+Therefore, our model estimates the effect of cancer type and class-specific cancer effects independently.
+We found that cancer types, as expected, have a range of prognoses relative to the general population, as shown in **figure 5a**.
+The least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a risk multiplier of between 15 and 22 relative to the background risk of death in the population.
+Among the deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers.
+Apart from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted death, the remaining cancer types ranged from about 1 to 5 in magnitude (**figure 5a**).
+
+![](../figures/fig6B.png)
+![](../figures/fig6C.png)
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
+a) Posterior distributions of estimated relative age-dependent survival by cancer type.
+b) Posterior distributions of estimated class-dependent survival for each cancer type.
+See methods for model details.</p>
+
+Our estimates of tissue-specific class effects, in contrast, ranged from 0-3, reflecting that some classes are either less deadly or more deadly than other classes within each cancer types, where the null hypothesis corresponds to an effect size=1 (**figure 5b**).
+For several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely indestinguishable from the prior, reflecting that either there were too few mortalities in the data to make an estimate (as expected for PRAD and THCA) or two few samples period.
+We did not observe class-specific trends that held true across cancer types, which could result from different cancers having different standards of care for example.
+In support of this interpretation, we also tested a factored model which considered both cancer type and each class independent of cancer type (**supplementary figure S7??**).
+Though we were able to successfully fit these models, they exhibited much greater uncertainty, suggesting an inferior specification.
+(<mark>Dan are there additional arguments to be made for the inferiority of the factored model, here?</mark>)
+
+```{r survival-model, eval=FALSE}
 
@@
 
-To determine whether our clusters were differentiated from random selections of tumors we plotted Kaplan Meier survival curves for each of the clusters first, and compared the results to randomized sets of tumors of equal-sized proportions (**Supplementary Figures S1B-C**).
-In addition we fitted a Cox proportional hazards test model, accounting for age of diagnosis and cluster membership and found that 8 of the clusters were significantly different from a reference cluster (cluster 1), suggesting that cluster membership could in principle be related to biological differences affecting survival (**Supplementary Table S1**).
-These differences were modest relative to differences in cancer by tissue of origin (**Supplementary Figure S1D**).
+![**Figure 6B**](../figures/fig6B.png)
 
-```{r preliminary-survival}
-
-```
-
-We noted intriguing similarities across different clusters, which motivated us to use the dimensional reduction analysis (DRA) tool UMAP [@McInnes2018;@art-of-tsne].
-Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings.
-To represent the spatial distribution in a more principled manner, we used hdbscan[@hdbscan] which resulted in 12 distinct new clusters projected onto the UMAP (**Figure 1A**).
-Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5 tumors from the hierarchical method predominate in cluster 1 of the DRA) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14 of the hierarchical method predominate in cluster 8 of the DRA).
-To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only).
-While there is a subtle trend, mostly visible for groups I and V (violin plot, **Figure 1B**), mutation count does not appear to be the dominant feature to distinguish the tumors in different classes of the DRA.
-To test whether the 5 groups represent progression by stage (which are determined clinically by size, though different criteria exist for different cancers), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every DRA cluster, about equally (**Figure \@ref(fig:umap-global-trends)C**).
+![**Figure 6C**](../figures/fig6C.png)
 
@@
 
-![**Figure 1 Classification of tumors independent of tissue of origin.** _**A.** UMAP: dimensional reduction analysis and clustering of pathway disruption data, followed by kNN assigns tumors to 12 distinct clusters.
-**B.** Violin plots of total mutations by pathway cluster reveal modest differences in overall mutation rate, except for cluster 8 which contains hypermutated tumors.
-In these tumors most pathways $>90%$ have at least one gene with a mutation.
-**C.** Tumors shaded by pathological stage on the UMAP from panel A do not favor any particular cluster.
-**D.** Metastatic status projected onto the UMAP clusters from panel A.
-Metastatic tumors are equally distributed among the clusters.
-**E.** Exemplars of 4 cancer types projected onto the UMAP clusters.
-Some biases are evident, but all cancers are found in every UMAP cluster.
-**F.** BRCA mutations, which are thought to increase genome instability, are found in every UMAP cluster._](../figures/fig1A_complete.png)
-
-To determine whether the 12 DRA clusters are independent from tissue of origin, we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the original UMAP (**Figures \@ref(fig:umap-global-trends)D**).
-While some modest bias is evident in some cancers, No tissue specific cancer exclusively found in one cluster.
-Full projections for all 23 cancers are presented in **Supplementary Figure S2**.
-Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies.
-One hypothesis to explain the lack of bias by cancer type in these clusters is that the groupings are representative of these molecular subtypes.
-To test this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP.
-Perhaps surprisingly we did not observe any exclusive segregation by pathway for these subtype annotations in GDC (**Figure 1E**).
-We also projected histological subtype data for the remaining cancers with at best modest biases evident (see **Supplementary Figure S3** for the full set of projections), suggesting that these subtypes are as diverse with respect to our pathway disruption analysis as the parent cancer types.
-
-There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes.
-However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length.
-_BRCA1/2_ genes for example are key for DNA double-stranded break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for breast, prostate and ovarian cancers (refs).
-The mechanism of risk is thought to involve loss of heterozygosity [@brca-loh], so we projected the somatic mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe segregation of these mutations into specific clusters (**Figure \@ref(fig:umap-global-trends)E**).
-We made similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_, _PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_.
-None of these genes except for _APC_ exhibited any remarkable specificity with respect to cluster assignment (**Supplementary Figure S4**).
-To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and observed similar lack of segregation by cluster (**Supplementary Figure S4**).
-
-### Determinants of Class
-
-In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 377 pathways used to produce it (**Supplementary Figure S5**), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment.
-We visualized the enrichment as a heatmap (**Figure \@ref(fig:group-pathway-enrichment)A**).
-Using this approach, we identified 6?
-pathways associated with class II, 34 pathways primarily enriched in group III tumors, and 10 pathways enriched in class IV tumors.
-Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively.
-Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (**Figure \@ref(fig:group-pathway-enrichment)B**).
-Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution.
-Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (**Figure \@ref(fig:group-pathway-enrichment)C**).
-Cluster group 6-7 had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, $\beta$-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others.
-As cluster groups 9-12 and 6-7 were the most distinct of the 12, we chose to investigate these two groups further.
-
-```{r group-pathway-enrichment, fig.cap="Placeholder for figure 2", eval=FALSE}
-set.seed(28)
-plot(density(rnorm(100)))
-```
-
-![**Figure 2.
-Pathway-specific enrichment within clusters.** _Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this)._](../figures/fig2B_short.png)
-
-Common to cluster groups 9-10 and 11-12 were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as _EGFR_, _IGF1R_, _Insulin receptor_, and _ERBB2_ and _-4_.
-In addition to these we observed _SCF-Kit_, _DAP12_, and _Tie 2_ signaling pathways, altogether suggesting that cluster group 9-12 tumors share mutations in extracellular signaling pathways generally.
-
-```{r redundant-genes, fig.cap="Placeholder for figure 3", results=FALSE}
-set.seed(29)
-hist(rnorm(100))
-```
-
-Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment.
-Upon closer examination the split between cluster groups 9+10 and 11+12 is largely driven by mutations in _PI3K_ and its homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary Figure S4**).
-Furthermore, clusters 9 and 11 are distinguished from 10 and 12 by general lack of P53 mutations (compare **Figure 1A** with the _TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to cluster group 9-12, the group represented by clusters 6 & 7 exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors spatially distributed between cluster groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics of both.
-The top ten distinguishing pathways of class 6/7 tumors were cellular response to hypoxia, regulation of mitotic cell cycle, assembly of the pre-replicative complex, metabolism of polyamines, degradation of $\beta$-catenin by the destruction complex, tumor necrosis factor receptor 2 non-canonical _NF-$\kappa$B_ pathway, _AUF1_ (hnRNP D0) binds and destabilizes RNA, regulation of apoptosis and 3 pathways associated with aspects of hedgehog signaling.
-To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (Figure 4A?).
-Surprisingly, this class of tumors was not predominated by 1 or more highly mutated genes across all cancers.
-Instead, it appears that there is a whole class of genes in the proteosomal pathway whose activity contributes to all the aforementioned pathways, and the majority of tumors in these classes have a single mutation in one of these genes.
-Nonetheless, a significant fraction (~1/4 <-- need numbers!) do not have proteosomal mutations but instead harbor mutations in genes representing a majority of the same pathways, suggesting that most if not all of the pathways are individually important and that proteosomal mutations may be sufficient to drive the net defect of tumors in these clusters.
-We annotated the tumors in this class according to tissue of origin to determine whether proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and indeed we could observe some clustering.
-These data could explain how tissue-specific gene mutation signatures might result in general cancer phenotypes by converging at the pathway level.
-
-### Oncoprint stuff
-
-### Metastatic _vs._ non-metastatic tumors
-
-Following the same logic we used to investigate cluster-specific enrichment of pathways (**Figure \@ref(fig:group-pathway-enrichment)**) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (**Figure \@ref(fig:umap-global-trends)C**).
-Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown).
-We reasoned that the clusters might be different enough to dilute cluster-specific enrichment.
-
-We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (**Table 1**).
-A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster.
-For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see **Supplementary figure S5**) but not in 9 and 10.
-This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, **Table 1**).
-This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (**Figure \@ref(fig:group-pathway-enrichment)**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters.
-These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).
-Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors.
-Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.
-
-There were also a number of pathways that were enriched with lower specificity (among 4 or more clusters) which were found to be enriched in metastases.
-We identified 18 such pathways, and of these 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing”.
-Of these pathways, we identified mutations in Sumoylation for 100% of the metastatic samples of clusters 3, 6, and 7 (**Table 1**).
-
-The enrichment we observe may be consistent with an interpretation in which the pathways themselves are promoting a metastatic phenotype, or at the very least that tumors bearing such mutations are more likely to turn metastatic since we don’t know the order in which they were acquired.
-The fact that many of the metastasis enrichments are cluster-specific and have a tendency to overlap with cluster-specific pathways from non-metastatic tumors of other clusters in a complementary fashion is suggestive that there is an order of progression with the metastatically enriched pathways representing later events.
-
-### Survival Analysis
-
-In order to understand whether our cluster identifications add any information to the prognositic value embedded in cancer type, we tested models of patient survival that accounts for cancer type and age at diagnosis.
-Initial models did not properly account for the trajectory of patient survival independent of age of diagnosis but showed some promise in distinguishing among classes for one or two of the deadliest cancer types.
-Therefore, we used mortality data from the CDC to accurately model the normal trajectory of patient survival independent of disease status, including gender as a factor.
-The resulting model is shown as a comparison with the census data in **Figure 6A.**
-
-![**Figure 6A.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-Using this base survival model, we constructed a likelihood function with parameters describing the overall lethality of each cancer type as a modifier $k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of $k$ and optimized model fit using stan[@stanref].
-The $k$ in our model can be taken as, roughly, the death rate relative to the baseline rate in **Figure 6A**.
-Thus, a value of $k=2$ means that a given cancer is twice as deadly.
-The posterior distributions for the 23 cancers are shown in **Figure 6B**.
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
 ## Materials & Methods
 
-### Classification of Tumors
+All code for producing the analyses and figures herein are included in this fully reproducible manuscript.
+In addition, the R markdown files and all other models are available from repositories in the following locations on the distributed version software hub GitHub.
 
-### Enrichment Calculations
+### <mark>Selection of pathways <- Peter to write</mark>
 
-### Survival Models
+### <mark>Filtering genes <- Peter to write</mark>
+
+### Class and stage-specific Enrichment Calculations
+
+We calculated the enrichment of pathways in one set of tumors as the relative fraction (and probability of observing) of tumors with a mutated gene in the pathway to all tumors, inclusive of the category of interest.
+To do this operation, we use the beta distribution to permute a posterior distribution on the fraction of tumors with a pathway mutated for each category based on the observed set, and compared this to the posterior obtained from the full set of tumors (all tumors) as the distribution of differences between all permuted samples.
+We considered a pathway enriched if the $95\%$ range of credible differences thus obtained excludes 0 (as a null hypothesis) and the mean credible difference was greater than or equal to $20\%$ enrichment, which excludes a large number of small differences that are not likely to be biologically relevant.
+
+### Models of survival based on population-specific longevity data
+
+We modeled baseline longevity in the U.S.
+with these different factors.
+<mark>Key assumptions?
+(Room here for further descriptions and equations)</mark> Using this base survival model, we constructed a likelihood function with parameters describing the overall lethality of each cancer type as a modifier $k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of $k$ and optimized model fit using stan[@stanref].
+The $k$ in our model can be taken as, roughly, the death rate relative to the baseline rates as depicted in **supplementary figure S5**.
+Thus, a value of $k=2$ means that a given cancer is twice as deadly.
+
+With an accurate model for baseline longevity in place, we sought to model cancer as an event based modifier of long-term survival.
+Thus any factors or events such as cancer would modify the baseline risk function as a multiplier.
+The resulting coefficients can thus be interpreted as increasing baseline risk by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for female vs.
+male caucasians, blacks and asians (hispanics are not clearly marked in the tumor data; see **supplementary figure S5** for comparisons of raw CDC data and baseline risk models).
+Thus we are able to accurately model baseline risk of death taking these factors into account.
+The data themselves encompass all other causes of death for these populations including cancer, thus freeing our model from concerns about other confounding factors ( _e.g._ Diabetes).
+For demonstration purposes we also illustrate some examples of the expected survival for a population that receives a diagnosis with doubled risk of death at ages 40, 50, 60, 70, and 80 (**supplementary figure S6**).
+
+As an aside, we did not explicitly account for stage at diagnosis in our model for the following reasons.
+In the tumor sample data, stage at diagnosis is confounded with cancer type because some cancers are screened agressively ( _e.g._ colorectal and prostate cancers) while others are diagnosed typically after they become problematic for the patient's lifestyle ( _e.g._ ovarian and pancreatic cancers).
+Secondly, such a model specification would suffer from added noise because the staging data are not well standardized across cancer types, have different criteria, and because it is unclear what the relationship between stage and advancement of disease is (for example some sub-stage 4 tumors are metastatic).
+To compound this latter issue, our tumor set is vastly under-powered given the uneven representation of stage across cancer types.
 
@@
 The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples [@hornNetSigNetworkbasedDiscovery2018].
-And that is not accounting for the fact that >30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential.
-The bioinformatic challenge therefore is to define the problem precisely and then create a new dataset that addresses this problem.
-Hence, we define cancer as a malady of biochemical processes and pathways, and generated a dataset that scores each of these essential pathways as mutated or not.
+And that is not accounting for the fact that ~30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential.
+Hence, we sought to investigate cancer as a disease of basic cellular and biochemical pathways, which we accomplished by translating gene-level mutations into pathway level disruptions.
+Importantly, we selectively removed pathways that lacked generality as cellular processes or that could be specifically assigned to differentiated cell types, while excluding others.
 
-It is important that we chose point mutations in exome sequencing data for this task because we know unambiguously in most cases the target gene.
-In addition, we reduce pathway state to a boolean value because it does not matter for our purposes whether a single tumor is “enriched” for pathway mutations, chiefly because such metrics favor larger pathways over smaller ones.
-The data are likely to be noisy for several reasons.
-First, point mutations can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in genes that are potentially either activators or repressors.
-In the current model, we accept this uncertainty and assume mutations are generically disruptive to the pathway.
-Second, low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018].
-To address this issue, we identified genes with low expression average for each type of cancer and eliminated them for that cancer only.
-Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis [@parkGenomicEvidenceElevated2012].
-We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.
+We hypothesize that cancer is the result of dysfunction from a limited number of basic cellular processes that are common to all living human cells, and thus all tumors belong on a spectrum of dysfunction on this axis regardless of tissue of origin.
+An alternative hypothesis is that every tumor belongs to one of a large number of syndromes which are unique to each tissue of origin, that share some mechanisms and superficial similarity in treatment strategies.
+The latter idea has gained momentum in recent years since the publication of Encode, Roadmap and TCGA consortium papers, perhaps owing to the intractable complexity of genomic data in all its forms.
+The inescapable conclusion thus far is that all these data point to tissue-of-origin as the most important driver of tumor behavior.
+Our observations contrast with this view, and instead support an interpretation in which all tumors manifest one of a limited number of metastable phenotypes represented by our classes (**figure 1a-k**), with increased fitness regardless of tissue of origin, which then converge on end stage cancer by accumulating complementary mutations to the other classes (**figure 1l** and **Table 2**).
 
-We hypothesize that cancer is the result of dysfunction of a limited number of basic cellular processes that are common to all living human cells, and thus all tumors belong on a spectrum of mutations regardless of tissue of origin.
-An alternative hypothesis is that every tumor belongs to one of a large number of syndromes that share some mechanisms and superficial similarity in treatment strategies.
-We refer to these as the “single disease” and the “multiple syndrome” hypotheses.
-The latter seems to have gained momentum in recent years since the publication of Encode, Roadmap and TCGA consortium papers, perhaps owing to the intractable complexity of genomic data in all its forms.
-The inescapable conclusion thus far is that all these data point to tissue-of-origin as the most important driver of tumor behavior.
-Our observations support the view of all tumors as manifestations of disease of cellular differentiation and metabolism, but by no means refute entirely the multiple syndrome hypothesis.
-
-We find that most tumors segregate into one of (at least) three biologically distinct clusters, regardless of the cancer type.
-Some cancer types ( _e.g._ Pancreatic adenocarcinomas in the RAS arm cluster IIIa, and breast cancers being excluded from the same cluster) are unevenly distributed toward a specific class, we could not identify any type of cancer that was exclusive to a tumor class.
-Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1E** and **Supplemental Figure S3**).
+We find that most tumors segregate into one of 12 biologically distinct clusters, regardless of the cancer type (**figure 1h** and **supplementary figure S1**).
+Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS arm clusters 9 and 10, and breast cancers being excluded from the same cluster, see **supplementary figure S1**) are unevenly distributed toward a specific class, we could not identify cancer type that was exclusive to a single tumor class.
+Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental Figure S2**).
 Indeed, it would be surprising if we did not observe bias for some cancers, as it has long been known that certain treatment regimens are likely to work for a majority of patients of a given cancer type.
 Nonetheless, the basis for some tumors being treatment-refractory in spite of clinical indications remains elusive.
-If supported by future studies, our analysis suggests general mechanisms underlying tumor evolution regardless of tissue or histological or cellular subtype and may form the basis for future separation of patients into different treatment regimens.
+Our analysis suggests the possibility of more general mechanisms underlying tumor evolution regardless of tissue, histological or cellular sub-type and could form a basis for future separation of patients into different precision treatment regimens.
 
 It could be that the classes we identified are consistent with a signature of overall disease progression, such that successive, adjacent classes on the UMAP projection exhibit increasing specificity.
-We identified no such trend and each class instead was associated with unique combinations of pathways.
-In classes 9-12 --- which contain mutations in extracellular signaling components --- we identified subgroups (class 9-10 and 11-12) which correspond to different intracellular signaling pathways.
-Also consistent with the groups reflecting basic differences in biology rather than stages of tumor evolution, we found that our classifications were largely independent of the number of mutations or clinical stage (**Figures 1B-C**).
-In our view, these observations suggest that cancers have available to them at least three separate mechanisms to escape the selective pressures they are exposed to in situ.
+We identified no such trend (see **figure 1k**) and each class instead was associated with unique combinations of pathways (as in **figure 3**).
+In our view, these observations suggest that cancers have available to them a limited number of stable configurations with respect to pathway disruption which are consistent with increased fitness and therefore positive selection.
 
 Finally, we must remark on two of the classes, 1 and 8, which for which we did not find distinctive associations with pathways.
-Class I has a relatively low level of mutated pathways overall, although it is broadly enriched in many of the same tumor-promoting pathways common to the other groups.
-We don’t believe this to be due to a true difference in stage of tumor evolution, as our data show clearly that this class is as likely to contain stage IV as it is to contain stage I tumors.
-We speculate this class represents a group of tumors with aberrations in methylation, gene expression, copy number or other genome rearrangements which we excluded from the present analysis.
-In the future, we hope to find ways to incorporate these other data types into a pathway-centric analysis as we have done here for point mutations and indels.
-Class V represents a group of hypermutated tumors that harbor so many mutations that virtually no pathway is unaffected.
-This group highlights one of the limitations of the approach that we’ve taken, in that it is likely that a significant fraction of the samples flagged as “mutant” for each pathway are actually unaffected, and thus if the mutation rate is high enough, those samples have low information content.
-Similar to class I we speculate that a future analysis of other data types may resolve these tumors into more meaningful categories.
+Class 1 has a relatively low level of mutated pathways overall, although it is broadly enriched in many of the same tumor-promoting pathways common to the other groups.
+This likely does not reflect a true difference in stages of tumor evolution, as our data show clearly that this class is as likely to contain stage IV as it is to contain stage I tumors.
+We speculate this class represents a group of tumors with large aberrations in methylation, copy number or other genome rearrangements which we excluded from the present analysis.
+In the future, we hope to incorporate these other data into a pathway-centric analysis as we have done here for point mutations and indels.
+Class 8 represents a group of hyper-mutated tumors that harbor so many mutations that virtually no pathway is unaffected.
+This group highlights one of the limitations of the approach that we’ve taken, in that it is likely that a significant fraction of the samples flagged as “mutant” for each pathway are actually unaffected, consistent with those samples have low information content.
+Similar to class 1 we speculate that a future analysis of other data types, including improved filtering of mutations, may resolve these tumors into more meaningful categories.
 
-### Risk genes and tumor class
+### Incompleteness of the pathway disruption data
 
-Certain mutations in risk genes increase risk for cancer quite dramatically.
-For example, carriers of mutations in _BRCA1_ and _BRCA2_ have increased risk for breast, ovarian, and prostate cancers (refs).
-The mechanism of risk in such cases is thought to be due to loss of heterozygosity (LOH) resulting in complete or near-complete loss of function [@brca-loh].
-In some cases it is known that carriers have a lower rate of chemo-resistance than non-carriers(refs), begging the question of whether carriers with LOH are exposed to distinctive selective pressure resulting in a constrained evolutionary outcome.
-If this hypothesis were true, we should have observed somatic mutations in _BRCA1/2_ genes clustering within one of our classes.
-Instead, _BRCA_ mutations were evenly distributed amongst the classes (as in **Figure 1E**), suggesting that treatments or outcomes that distinguish between carriers and the general population are specific to _BRCA_ gene biology.
-_BRCA1/2_ encode proteins whose function is involved in double-stranded break repair, and we speculate that genes involved in this and other aspects of genome stability are therefore upstream of the later events in tumor evolution that distinguish between clusters 2-5.
-This further provides hope for combination therapies that account for carrier status and tumor class separately.
+The fact that we are likely excluding many defects (under chromosomal abberations, amplifications, demethylation and so on) also raises the possibility that our classes are incomplete or may change at a future time when we can account for them.
+For now at least, we think that the point mutation data are representative of the most important pathways.
 
-### (Include discussions of other germline LOH)
+### Estimates of survival reveal pathway-dependent differences
 
-Our findings imply that there are separate processes in the etiology of cancer that can be broadly thought of as initiation, acquisition of cancer and metastasis.
-Initiation processes must include factors that relate to genome stability either directly or indirectly.
-Acquisition genes relate to selective pressures that all cells face; a complex web of signaling pathways governs cell division and migration in normal healthy cells, s.th.
-any cells that turn these pathways on in the absence of signal have a selective advantage.
-It is no wonder then that hormone-driven cancers coincide with menopause and advancement beyond middle age.
-With regard to metastasis.
+By modeling CDC longevity data as a baseline risk function we showed that each of our pathway disruption classes exhibit cancer-type specific effects on expected survival (**figure 5**).
+Contrary to what one might expect at the outset, we found that models assuming different survival based on class _and_ cancer type fit better than models in which these variables are independent factors.
+However, considering that within each cancer type there are different clinical standards of care, and even within classes of drugs the preferred treatment can very between cancers, it makes sense in hindsight that they should all be different.
+Contrast the situation with ovarian vs breast cancer, which are both hormonally driven cancers, for example.
+Ovarian cancer has but one main treatment, platinum, whereas breast cancer patients have a variety of treatment regimens based on molecular subtype and other factors.
+It is easy to see how standard of care for one cancer and class could produce different outcomes than the same class of a different cancer type under the circumstances.
+Unfortunately, given the diversity drug classes and treatments (including non-chemical treatments such as surgery and palliative care), we lack sufficient power in the current study to explore these variables in the GDC data.
+It is our hope that more focused future studies will help to distinguish between treatment-specific effects or survival given different pathway disruption classes.
 
-### Two-step and 3-strike
+### Implications for the evolution of cancer
+
+Our findings imply that there are separate processes in the etiology of cancer that can be broadly thought of as general cancer promoting, class-specific mutations and metastasis.
+General cancer promoting processes must include factors that relate to genome stability either directly or indirectly.
+Such pathways are disrupted in all or most of the defined classes (see **figures 2** and **supplementary figure S1**) and often are the result of abberations involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
+Class-specific evolution must involve the acquisition of combinatorial disruptions to pathways that may individually be harmful but together produce more specialized cancer phenotypes and increased fitness (as in **figure 4** and **tables 1** and **ST2**).
+Importantly, our work does not distinguish between the order in which these mutations should accumulate.
+However, since many of the genes in the non-specific pathways involve genome stability, it is reasonable to hypothesize that such mutations promote the acquisition of class specific defects, and natural selection produces the classes we observed.
+
+### On metastasis as a convergence of phenotypes
+
+The class-specific metastatic pathway enrichment we observe (in **table 2**) may be a result of the pathways directly promoting metastatic phenotypes, or alternatively that such mutations are more likely to turn metastatic because they acquire a stronger or more complete oncogenic phenotype, since (again) we don’t know the order in which they were acquired.
+The fact that most metastatic enrichment is cluster-specific and has a tendency to overlap with cluster-specific pathways from non-metastatic tumors of neighboring clusters favors the latter explanation, and suggests a mechanism of complementarity in which newly acquired mutations converge on a single deadly phenotype that captures most or all of the critical features of end-stage cancer.
+If this is the case, we speculate that once we discover the vulnerabilities of each cluster we might also reasonably predict which pathways are susceptible to further selection and therefore target them  therapeutically at the same time to prevent further adaptation.
 
@@
 
-![**Supplementary Figure S1A.
-K-means clustering of booleanized pathway disruptions.** _Each column represents a single tumor, each row a pathway.
-In the heatmap below, red colored fields indicate a mutation in a pathway, and gray indicates no mutations in pathway genes.
-In the annotation rows above the heatmap, annotations have been spread out by cancer type to reveal the distributions across clusters._](../figures/figureS1A.png)
-
-![**Supplementary Figure S1B-S1D.
-Kaplan Meier (KM) curves.** _B.
-KM plots of k-means clustered pathway data for GDC samples.
-C.
-KM plots of randomly permuted labels on the same samples.
-D.
-KM plots of GDC samples by cancer type._](../figures/figS1B_D.png)
+![**Supplementary figure S1.
+Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._](../figures/figS2.png)
 
 ![**Supplementary Figure S2.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._](../figures/figS2.png)
+Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_](../figures/figS3.png)
 
 ![**Supplementary Figure S3.
-Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_](../figures/figS3.png)
-
-![**Supplementary Figure S4.
 Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_](../figures/figS4.png)
+![**Supplemental figure S6.
+Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
+The orange and green curves show the model fits._](../figures/fig6A.png)
 

### 2020-07-03 text/manuscript.Rmd

@@
 ## Abstract
-Cancer is a disease that affects metabolism and other basic cellular processes common to all cells.
-In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying principles elude the field.
-To address this challenge we reduced tumor mutation data to pathway memberships, using a highly filtered set of pathways representative of basic cellular functions from metabolism and DNA replication to cell-cell communication.
-We find that tumors can be classified according to pathway disruption in twelve classes entirely distinct from tissue of origin based on these criteria, and largely free of bias from mutational burden and clinical stage.
-We also show that these tumor classes have different prognoses in different tissues using a hierarchical model of longevity data from the CDC.
-As a consequence of our pathway-centric classification, we can also identify equivalent sets of class-specific gene mutations in different tissues.
-Moreover, many known germline risk mutations involved in genome stability such as BRCA1/2 and Rb1 are equally distributed among the classes, somewhat contrary to the expectation that such genes are representative of disease-specific categories or subtypes.
-Similarly, we describe mutations that distinguish between metastatic tumors and non-metastatic tumors.
-In many cases the enriched pathways from one class in metastatic tumors are the class-specific pathways from another class, suggesting that metastasis may represent phenotypic convergence, therefore potentially identifying key pathways to focus on therapeutically for prevention of class-specific metastasis.
-In summary, our observations suggest that some genes act as drivers of genome instability while other class-specific genes confer increased fitness in combinatorial fashion.
-Thus, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on their distribution in tumors of every class.
+Cancer is a disease of cellular metabolism and other basic biochemical processes common to all cells.
+However, molecular signatures remain largely indistinguishable from tissue-of-origin, and general, unifying principles remain obscure.
+To address this challenge, we reduced mutation data to disruptions in a select set of pathways pointing to basic cell biology, from DNA replication to cellular communication.
+Using dimension reduction techniques, we assigned tumor samples into twelve classes distinct from their tissues-of-origin and largely free of bias from mutational burden or clinical stage.
+We showed that these tumor classes vary in prognosis by modeling relative risk of death in different classes and cancer types.
+We identified class-specific mutations in different tissues, demonstrating how tissue-specific signatures contribute to common cellular phenotypes.
+Moreover germline risk mutations involved in genome stability, such as BRCA1 and Rb1, are equally distributed among our classes, somewhat contrary to the expectation that such genes are hallmarks of molecular sub-types.
+We also investigated pathways that distinguish between metastatic and non-metastatic tumors, and found that most meaningful differences are class-specific.
+Most metastatic enriched pathways from one class in metastatic tumors are class-specific pathways from another class, suggesting phenotypic convergence, and potentially highlighting key pathways to focus on therapeutically for prevention of class-specific relapse.
+In summary, our observations suggest that some genes act as drivers of genome instability in all classes, while different sets of class-specific genes confer increased fitness in combinatorial fashion, eventually converging on one or more end-stage disease phenotypes.
+Thus, we present a coherent picture of global tumor biology, and explain how common cellular dysfunction might arise from tissue-specific mutations.
 

### 2020-07-03 text/manuscript.Rmd
reflow text-wrap paragraphs as lines for increased efficiency of git diffs

@@
 ## Abstract
-Cancer is a disease of cellular metabolism and other basic biochemical processes common to all cells.
-However, molecular signatures remain largely indistinguishable from tissue-of-origin, and general, unifying principles remain obscure.
-To address this challenge, we reduced mutation data to disruptions in a select set of pathways pointing to basic cell biology, from DNA replication to cellular communication.
-Using dimension reduction techniques, we assigned tumor samples into twelve classes distinct from their tissues-of-origin and largely free of bias from mutational burden or clinical stage.
-We showed that these tumor classes vary in prognosis by modeling relative risk of death in different classes and cancer types.
-We identified class-specific mutations in different tissues, demonstrating how tissue-specific signatures contribute to common cellular phenotypes.
-Moreover germline risk mutations involved in genome stability, such as BRCA1 and Rb1, are equally distributed among our classes, somewhat contrary to the expectation that such genes are hallmarks of molecular sub-types.
-We also investigated pathways that distinguish between metastatic and non-metastatic tumors, and found that most meaningful differences are class-specific.
-Most metastatic enriched pathways from one class in metastatic tumors are class-specific pathways from another class, suggesting phenotypic convergence, and potentially highlighting key pathways to focus on therapeutically for prevention of class-specific relapse.
-In summary, our observations suggest that some genes act as drivers of genome instability in all classes, while different sets of class-specific genes confer increased fitness in combinatorial fashion, eventually converging on one or more end-stage disease phenotypes.
-Thus, we present a coherent picture of global tumor biology, and explain how common cellular dysfunction might arise from tissue-specific mutations.
+Cancer is a disease of cellular metabolism and other basic biochemical processes
+common to all cells. However, molecular signatures remain largely
+indistinguishable from tissue-of-origin, and general, unifying principles remain
+obscure. To address this challenge, we reduced mutation data to disruptions in a
+select set of pathways pointing to basic cell biology, from DNA replication to
+cellular communication. Using dimension reduction techniques, we assigned tumor
+samples into twelve classes distinct from their tissues-of-origin and largely
+free of bias from mutational burden or clinical stage. We showed that these
+tumor classes vary in prognosis by modeling relative risk of death in different
+classes and cancer types. We identified class-specific mutations in different
+tissues, demonstrating how tissue-specific signatures contribute to common
+cellular phenotypes. Moreover, germline risk mutations involved in genome
+stability, such as BRCA1 and Rb1, are equally distributed among our classes,
+somewhat contrary to the expectation that such genes are hallmarks of molecular
+subtypes. We also investigated pathways that distinguish between metastatic and
+non-metastatic tumors, and found that most meaningful differences are
+class-specific. Most metastatic enriched pathways from one class in metastatic
+tumors are class-specific pathways from another class, suggesting phenotypic
+convergence, and potentially highlighting key pathways to focus on
+therapeutically for prevention of class-specific relapse. In summary, our
+observations suggest that some genes act as drivers of genome instability in all
+classes, while different sets of class-specific genes confer increased fitness
+in a combinatorial fashion, eventually converging on one or more end-stage
+disease phenotypes. Thus, we present a coherent picture of global tumor biology,
+and explain how common cellular dysfunction might arise from tissue-specific
+mutations.
 
@@
 
-Recent advances in technology have made the study of cancer possible at unprecedented levels of detail and scale.
-Array and next-generation sequencing technologies have made it possible to assay germline and somatic mutations, expression profiles, DNA methylation, and copy number variations.
-To take advantage of these tools several large consortia, including the cancer genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large numbers of tumors and collected data from multiple assays that could be analyzed together, with the goal of integration and increased understanding of the mechanisms of cancer.
-These efforts have focused on recognizing and quantifying the importance of genes in cancer etiology as the result of selection processes, or classification of cancers relative to common patterns of aberration in multiple assays.
+Recent advances in technology have made the study of cancer possible at
+unprecedented levels of detail and scale. Array and next-generation sequencing
+technologies have made it possible to assay germline and somatic mutations,
+expression profiles, DNA methylation, and copy number variations. To take
+advantage of these tools several large consortia, including the cancer genome
+atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
+numbers of tumors and collected data from multiple assays that could be analyzed
+together, with the goal of integration and increased understanding of the
+mechanisms of cancer. These efforts have focused on recognizing and quantifying
+the importance of genes in cancer etiology as the result of selection processes,
+or classification of cancers relative to common patterns of aberration in
+multiple assays.
 
-Considerable progress has been made analyzing these data.
-Statistical analyses identify hundreds of global and tissue-specific cancer driver genes [@deesMuSiC2012; @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014; @kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016; @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting when genes are mutated at a greater rate than expected due to chance.
-Based in part on studies like these, it has been estimated that on the order of three mutations in key oncogenes and/or tumor suppressors would be sufficient to transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
+Considerable progress has been made analyzing these data. Statistical analyses
+identify hundreds of global and tissue-specific cancer driver genes
+[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
+@lawrenceDiscoverySaturationAnalysis2014;
+@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
+@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
+when genes are mutated at a greater rate than expected due to chance. Based in
+part on studies like these, it has been estimated that on the order of three
+mutations in key oncogenes and/or tumor suppressors would be sufficient to
+transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
 
-Other studies aim at providing a comprehensive view of mutations, gene expression and genomic signatures, with the express goal of understanding the common themes of all cancer independent of tissue of origin.
-The first of these (the “Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types, using integrative analysis with co-equal weighting of gene expression, methylation, copy number and proteomics data [@hoadleyMultiplatformAnalysis2014].
-The principal finding was that tissue-of-origin is the predominant driving factor, though a significant proportion of tumors ($10\%$) could be reclassified independent of tissue-of-origin.
-In a second study involving 33 cancer types and a much greater number of tumors using a newer algorithm (iCluster) [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that could be further related or subdivided into organ specific groups, including pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/ squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
+Other studies aim at providing a comprehensive view of mutations, gene
+expression and genomic signatures, with the express goal of understanding the
+common themes of all cancer independent of tissue of origin. The first of these
+(the “Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
+using integrative analysis with co-equal weighting of gene expression,
+methylation, copy number and proteomics data
+[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
+tissue-of-origin is the predominant driving factor, though a significant
+proportion of tumors ($10\%$) could be reclassified independent of
+tissue-of-origin. In a second study involving 33 cancer types and a much greater
+number of tumors using a newer algorithm (iCluster)
+[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
+could be further related or subdivided into organ specific groups, including
+pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
+squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
 
-More recently, attempts have been made to consider pan-cancer mutations and other data outside the context of enrichment at the gene level.
-Horn _et al._ [@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the sparseness of mutations, allowing the potential importance of a gene to be influenced by the aggregate of positive selection in its nearest network neighbors.
-This method is powerful especially when applied to a small number (on the order of hundreds) of tumors.
-However, the larger problem of cancer etiology remains: what mutations are required, in what order, and are there specific combinations of delinquent cellular processes sufficient to cause cancer?
-If there is a common logic to cancer, how is it to be detected underneath the overwhelming cell-of-origin signal?
+More recently, attempts have been made to consider pan-cancer mutations and
+other data outside the context of enrichment at the gene level. Horn _et al._
+[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
+sparseness of mutations, allowing the potential importance of a gene to be
+influenced by the aggregate of positive selection in its nearest network
+neighbors. This method is powerful especially when applied to a small number (on
+the order of hundreds) of tumors. However, the larger problem of cancer etiology
+remains: what mutations are required, in what order, and are there specific
+combinations of delinquent cellular processes sufficient to cause cancer? If
+there is a common logic to cancer, how is it to be detected underneath the
+overwhelming cell-of-origin signal?
 
-In this study, we attempt to address these questions by focusing on discrete molecular pathways as the unit of disruption in cancer.
-Our approach strips all cell-type-specific information from the mutation data and generalizing gene-level information to non-tissue specific cell-biological pathway phenotypes.
-We use this rubric to evaluate all cancers and show that, surprisingly, tumors may have distinct phenotypes completely independent of their tissues of origin.
+In this study, we attempt to address these questions by focusing on discrete
+molecular pathways as the unit of disruption in cancer. Our approach strips all
+cell-type-specific information from the mutation data and generalizing
+gene-level information to non-tissue specific cell-biological pathway
+phenotypes. We use this rubric to evaluate all cancers and show that,
+surprisingly, tumors may have distinct phenotypes completely independent of
+their tissues of origin.
 
@@
 
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC.
-We chose to start with somatic mutations in exome sequencing data because the affected gene is known almost unambiguously.
-Therefore, we focused on all missense, nonsense, frameshift, and splicing mutations.
-In order to minimize bias from well-studied diseases and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) of interest corresponding to basic cellular processes and biochemical pathways, excluding gene sets that correspond to miscellaneous categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on membership in these pathways (<mark>total of x genes;</mark> see methods for details).
+To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
+chose to start with somatic mutations in exome sequencing data because the
+affected gene is known almost unambiguously. Therefore, we focused on all
+missense, nonsense, frameshift, and splicing mutations. In order to minimize
+bias from well-studied diseases and processes, we selected 377 Reactome pathways
+[reactome.org](https://reactome.org/) of interest corresponding to basic
+cellular processes and biochemical pathways, excluding gene sets that correspond
+to miscellaneous categories ( _e.g._ “transcription factors”) or disease
+associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
+membership in these pathways (<mark>total of x genes;</mark> see methods for
+details).
 
-It does not matter for our purposes whether a single tumor is “enriched” for multiple pathway mutations, chiefly because such metrics favor larger pathways ( _i.e._ pathways with more member genes) over smaller ones, and non-expressed genes over expressed ones due to transcription coupled repair [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018; @Pervasive_lesion_segregation].
-Data selected in this manner are likely to be noisy for several reasons.
-First, point mutations can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in genes that act functionally as either activators or repressors.
-Within the scope of this study we must accept this uncertainty and assume mutations are generically disruptive to the cell regardless of whether they are activating or deactivating or whether they affect a gene that we might classify as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene").
-Second, as previously mentioned, we know that low-expressed genes and non-expressed genes are known to accumulate mutations at a higher rate due to transcription coupled repair.
-To address this issue, we identified genes with low expression in each type of cancer and eliminated them for that cancer type only (see methods for details).
-<strike>Highly expressed genes also have high mutation rates owing to transcription induced mutagenesis [@parkGenomicEvidenceElevated2012].
-We felt that this mechanism results in cell-type-specific biases that might be biologically meaningful for predisposition to different classes of cancer however, and therefore chose to include these genes in our analysis.</strike>
+It does not matter for our purposes whether a single tumor is “enriched” for
+multiple pathway mutations, chiefly because such metrics favor larger pathways (
+_i.e._ pathways with more member genes) over smaller ones, and non-expressed
+genes over expressed ones due to transcription coupled repair
+[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
+@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
+noisy for several reasons. First, point mutations can be deleterious
+(attenuating or hypomorphic) or activating (neomorphic) in genes that act
+functionally as either activators or repressors. Within the scope of this study
+we must accept this uncertainty and assume mutations are generically disruptive
+to the cell regardless of whether they are activating or deactivating or whether
+they affect a gene that we might classify as one thing or another ( _e.g._
+"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
+low-expressed genes and non-expressed genes are known to accumulate mutations at
+a higher rate due to transcription coupled repair. To address this issue, we
+identified genes with low expression in each type of cancer and eliminated them
+for that cancer type only (see methods for details). <strike>Highly expressed
+genes also have high mutation rates owing to transcription induced mutagenesis
+[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
+cell-type-specific biases that might be biologically meaningful for
+predisposition to different classes of cancer however, and therefore chose to
+include these genes in our analysis.</strike>
 
-After selecting our pathways and genes, we then compiled a matrix of the pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene carries a mutation) to the pathway regardless of the number of genes carrying mutations (this data is visualized in **figure 1a**).
+After selecting our pathways and genes, we then compiled a matrix of the
+pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
+member gene carries a mutation) to the pathway regardless of the number of genes
+carrying mutations (this data is visualized in **figure 1a**).
 
@@
 ![](../figures/FIG1_TEMP.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1: Classification of tumors by pathway disruptions.** a) Each of 377 selected Reactome pathways (columns) is classified as disrupted if one or more genes is mutated in the tumor sample (rows).
-Tumors are grouped by tissue of origin using standardized abrreviations from the TCGA project.
-b) MCA analysis was performed restricted to the 36 informative components.
-c) The resulting MCA was projected down to 3 dimensions for the purpose of density based clustering (only the first 2 dimensions are shown).
-Each dot represents a tumor d) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize the fraction of unassigned tumors.
-e) HDBSCAN clustered tumors were visualized on the UMAP projection; tumors are colored by cluster membership (see legend).
-This results in some tumors remaining unclassified (grey points).
-f) kNN was used to assign membership to the remaining tumors.
-g) Violin plots of sample mutation counts for tumors in each class.
-h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma.
-i) Projection of breast cancer subtypes onto the UMAP.
-j) projection of tumor stage onto the UMAP, regardless of cancer type.
-k) Projection of metastatic status onto the UMAP.
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
+Classification of tumors by pathway disruptions.** a) Each of 377 selected
+Reactome pathways (columns) is classified as disrupted if one or more genes is
+mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
+standardized abrreviations from the TCGA project. b) MCA analysis was performed
+restricted to the 36 informative components. c) The resulting MCA was projected
+down to 3 dimensions for the purpose of density based clustering (only the first
+2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
+perform hierarchical density based clustering tuned to minimize the fraction of
+unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
+projection; tumors are colored by cluster membership (see legend). This results
+in some tumors remaining unclassified (grey points). f) kNN was used to assign
+membership to the remaining tumors. g) Violin plots of sample mutation counts
+for tumors in each class. h) Projection of cancer type identity onto UMAP.
+Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
+Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
+cancer subtypes onto the UMAP. j) projection of tumor stage onto the UMAP,
+regardless of cancer type. k) Projection of metastatic status onto the UMAP.
 Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
 
-In order to classify tumors using this dataset, we used multiple correspondence analysis (MCA) [@JSSv025i01].
-First we determined the number of dimensions containing useful information (**figure 1b**).
-Then we performed a UMAP analysis, both in order to summarize the MCA graphically, and as a preprocessing step to boost the performance of density based clustering.
-The resulting map was notable for its lobed structure, with several large, robust projections reproducible regardless of random seed setting.
-A representative version of this UMAP is shown in **figure 1c** as a 2D projection.
-Following this spatial mapping we attempted to define groupings of similar tumors within the spatial map using HDBSCAN, which performs hierarchical clustering and provides metrics of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are the minimum number of tumor samples in a cluster that capture the maximum number of tumor samples, measured by probability of membership of $\geq 5\%$ in at least one class.
-Thus, we created a score metric as the fraction of classified tumors with max probability $< 5\%$ in one class and chose a cluster size of 95 to minimize the score function (**figure 1d**).
-HDBSCAN with these settings resulted in 12 distinct high-density clusters which we then projected onto the UMAP (**figure 1e**).
-This classified <mark>x out of 7,607 tumors</mark> but still resulted in a significant fraction of unclassified tumor samples.
-Since we ultimately wish to be able to classify any tumor using this scheme, we performed k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every tumor in the set and then lets the $k$ most similar tumors "vote" as to the identity of the query tumor sample based on their classification labels.
-We set $k$ to be the square root of the number of tumor samples (87).
-Using this method, we assigned cluster membership to the remaining tumors (**figure 1f**).
+In order to classify tumors using this dataset, we used multiple correspondence
+analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
+containing useful information (**figure 1b**). Then we performed a UMAP
+analysis, both in order to summarize the MCA graphically, and as a preprocessing
+step to boost the performance of density based clustering. The resulting map was
+notable for its lobed structure, with several large, robust projections
+reproducible regardless of random seed setting. A representative version of this
+UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
+mapping we attempted to define groupings of similar tumors within the spatial
+map using HDBSCAN, which performs hierarchical clustering and provides metrics
+of cluster stability and probabilities of cluster membership for each node.
+However, HDBSCAN is sensitive to several parameters; key for our analysis are
+the minimum number of tumor samples in a cluster that capture the maximum number
+of tumor samples, measured by probability of membership of $\geq 5\%$ in at
+least one class. Thus, we created a score metric as the fraction of classified
+tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
+to minimize the score function (**figure 1d**). HDBSCAN with these settings
+resulted in 12 distinct high-density clusters which we then projected onto the
+UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
+still resulted in a significant fraction of unclassified tumor samples. Since we
+ultimately wish to be able to classify any tumor using this scheme, we performed
+k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
+tumor in the set and then lets the $k$ most similar tumors "vote" as to the
+identity of the query tumor sample based on their classification labels. We set
+$k$ to be the square root of the number of tumor samples (87). Using this
+method, we assigned cluster membership to the remaining tumors (**figure 1f**).
 
@@
 
-Having defined tumors in terms of their pathway disruption profile, we sought to understand whether different cancer types segregated into one or more predominant classes.
-To our surprise, most cancer types (with some notable exceptions) were not heavily biased in one class, and all well-represented cancer types had representative tumors in every class (for example **figure 1g** and full tumor profiles in **supplementary figure S1**, see also interactive media from **supplementary file 1 (html)**), suggesting that, in principle at least, our novel pathway-disruption classification method identifies molecular pathologies largely independent of tissue-of-origin considerations.
-As an example of one type of cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is distinguished  by KRAS mutations among others.
-But even PAAD comprises tumors from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that patients hosting these tumors have potentially different underlying molecular pathologies.
+Having defined tumors in terms of their pathway disruption profile, we sought to
+understand whether different cancer types segregated into one or more
+predominant classes. To our surprise, most cancer types (with some notable
+exceptions) were not heavily biased in one class, and all well-represented
+cancer types had representative tumors in every class (for example **figure 1g**
+and full tumor profiles in **supplementary figure S1**, see also interactive
+media from **supplementary file 1 (html)**), suggesting that, in principle at
+least, our novel pathway-disruption classification method identifies molecular
+pathologies largely independent of tissue-of-origin considerations. As an
+example of one type of cancer that does have a biased pathway profile,
+pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
+distinguished  by KRAS mutations among others. But even PAAD comprises tumors
+from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
+patients hosting these tumors have potentially different underlying molecular
+pathologies.
 
@@
 
-Many cancers have molecular or histological subtypes defined based on gene expression or other -omics profiles or pathology lab results.
-These subtypes often have different standards of care owing to different overall drug sensitivity (or other factors).
-If the histological subtypes represent true molecular phenotypes, one predicts that histological subtypes should segregate with our pathway-based classes, therefore providing support for the classes as proxies for molecular pathology sub-typing.
-To our surprise, we find a similar result to the previous analysis of cancer types projected onto the UMAP of pathway disruptions.
-For example we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP.
-These are among the most heavily studied molecular subtypes in cancer, which each have different prognoses and standards of care.
-We did not observe any exclusive segregation by pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental file 1**).
-We also projected histological subtype data for the remaining cancers (see **supplementary figure S2** for the full set of projections); we find that the subtypes, though often biased towards one or more classes, are almost never exclusive to one of the classes.
-We interpret these data in aggregate to mean that our pathway disruption classes do not correspond to molecular subtypes within the parent cancer type.
-Furthermore, this result raises the possibility that, molecular and histo-pathological subtypes could represent different precursor cell origins within each cancer type.
-This interpretation still leaves open the possibility that different precursor cell types, being susceptible to different teratogens and selection pressures ( _e.g._ insulin dependence) could be biased toward one or more of the pathway disruption profiles.
+Many cancers have molecular or histological subtypes defined based on gene
+expression or other -omics profiles or pathology lab results. These subtypes
+often have different standards of care owing to different overall drug
+sensitivity (or other factors). If the histological subtypes represent true
+molecular phenotypes, one predicts that histological subtypes should segregate
+with our pathway-based classes, therefore providing support for the classes as
+proxies for molecular pathology sub-typing. To our surprise, we find a similar
+result to the previous analysis of cancer types projected onto the UMAP of
+pathway disruptions. For example we projected annotations for each of the breast
+cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
+normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
+most heavily studied molecular subtypes in cancer, which each have different
+prognoses and standards of care. We did not observe any exclusive segregation by
+pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
+file 1**). We also projected histological subtype data for the remaining cancers
+(see **supplementary figure S2** for the full set of projections); we find that
+the subtypes, though often biased towards one or more classes, are almost never
+exclusive to one of the classes. We interpret these data in aggregate to mean
+that our pathway disruption classes do not correspond to molecular subtypes
+within the parent cancer type. Furthermore, this result raises the possibility
+that, molecular and histo-pathological subtypes could represent different
+precursor cell origins within each cancer type. This interpretation still leaves
+open the possibility that different precursor cell types, being susceptible to
+different teratogens and selection pressures ( _e.g._ insulin dependence) could
+be biased toward one or more of the pathway disruption profiles.
 
@@
 
-There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes.
-However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length.
-_BRCA1/2_ genes for example are key for DNA double-stranded break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for breast, prostate and ovarian cancers (refs).
-The mechanism of risk is thought to involve loss of heterozygosity [@brca-loh], so we projected the somatic mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe segregation of these mutations into specific clusters (**figure 1j**).
-We made similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_, _PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_.
-None of these genes except for _APC_ exhibited any remarkable specificity with respect to cluster assignment (**supplementary figure S3**).
-To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and observed similar lack of segregation by cluster (**supplementary figure S3**).
+There are several well-known familial cancer-causing mutations that have been
+interrogated extensively for differences in basic biology, survival and
+treatment outcomes. However, the functions of these genes are related to risk
+factors such as genome stability, proof-reading and DNA damage repair, and
+telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
+break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
+breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
+involve loss of heterozygosity [@brca-loh], so we projected the somatic
+mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
+segregation of these mutations into specific clusters (**figure 1j**). We made
+similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
+_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
+except for _APC_ exhibited any remarkable specificity with respect to cluster
+assignment (**supplementary figure S3**). To look at other risk factors such as
+maintenance of DNA methylation levels and telomere length we projected somatic
+mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
+observed similar lack of segregation by cluster (**supplementary figure S3**).
 
-#### Independence of Stage, mutation count and mutation profile.
+#### Independence of stage, mutation count and mutation profile.
 
-Tumor staging is based on physico-pathological criteria, including diameter, which can vary greatly in importance between different tissues.
-Stage is used clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common pathway disruption profiles, and the UMAP which features a series of lobe-like structures on a common backbone of tumor samples (**figure 1c**) could in principle also reflect progression through a series of stages.
-In support of this hypothesis, the backbone culminates in a group of tumors (class 8) which has nearly every pathway disrupted.
-However, outside of class 8 we don't observe an obvious trend in the mutation count across the backbone of the UMAP (**figure 1g**).
-Nonetheless, to test the hypothesis that the molecular-pathway disruption classes represent advancement through stages, we projected these data onto the UMAP.
-Similar to tissue of origin and other categories of tumor, we did not observe any bias amongst the stages into specific pathway disruption classes (**figure 1k**), suggesting that stage is not an determining factor in pathway disruption classes.
+Tumor staging is based on physico-pathological criteria, including diameter,
+which can vary greatly in importance between different tissues. Stage is used
+clinically as a proxy for advancement toward a more deadly state and metastasis.
+Given these assumptions, it is possible that more advanced tumors have common
+pathway disruption profiles, and the UMAP which features a series of lobe-like
+structures on a common backbone of tumor samples (**figure 1c**) could in
+principle also reflect progression through a series of stages. In support of
+this hypothesis, the backbone culminates in a group of tumors (class 8) which
+has nearly every pathway disrupted. However, outside of class 8 we don't observe
+an obvious trend in the mutation count across the backbone of the UMAP (**figure
+1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
+classes represent advancement through stages, we projected these data onto the
+UMAP. Similar to tissue of origin and other categories of tumor, we did not
+observe any bias amongst the stages into specific pathway disruption classes
+(**figure 1k**), suggesting that stage is not an determining factor in pathway
+disruption classes.
 
-Finally, as a measure of tumor advancement, metastasis is the condition in which certain phenotypic criteria are met.
-These phenotypes include (among others) loss of differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune system evasion and tissue invasiveness.
-To determine whether any of our classes correspond to an especially advanced stage of cancer accross tissue types, we projected the metastasis data onto our UMAP, and surprisingly, again, we observed an even distribution across classes (**figure 1l**).
-This final observation suggests that our pathway-disruption classification is dependent on particular combinations of gene mutations that hit different pathway profiles that can each give rise to advanced stages of disease and metastasis, regardless and independent of overall mutational burden.
+Finally, as a measure of tumor advancement, metastasis is the condition in which
+certain phenotypic criteria are met. These phenotypes include (among others)
+loss of differentiation, cell-cell contacts, epithelial to mesenchyme
+transition, immune system evasion and tissue invasiveness. To determine whether
+any of our classes correspond to an especially advanced stage of cancer accross
+tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
+again, we observed an even distribution across classes (**figure 1l**). This
+final observation suggests that our pathway-disruption classification is
+dependent on particular combinations of gene mutations that hit different
+pathway profiles that can each give rise to advanced stages of disease and
+metastasis, regardless and independent of overall mutational burden.
 
@@
 
-In order to visualize pathway disruption enrichment across all cancers (pan-cancer), we created oncoprint graphs analagous to the commonly use method, replacing genes with pathways (**figure 2**).
-A complete list of pathway disruptions with percent mutated samples and top genes is provided in **supplementary table ST1**).
-These reveal the broad importance of many well known pathways that are disrupted in cancer, including <mark>PI3K/Akt activation, DNA repair etc fix this with top 5 or so quoted pathway labels</mark>.
-The top third of pathways drop off gradually in overall mutation frequency, possibly owing to relative diversity among all tumors.
+In order to visualize pathway disruption enrichment across all cancers
+(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
+replacing genes with pathways (**figure 2**). A complete list of pathway
+disruptions with percent mutated samples and top genes is provided in
+**supplementary table ST1**). These reveal the broad importance of many well
+known pathways that are disrupted in cancer, including <mark>PI3K/Akt
+activation, DNA repair etc fix this with top 5 or so quoted pathway
+labels</mark>. The top third of pathways drop off gradually in overall mutation
+frequency, possibly owing to relative diversity among all tumors.
 
@@
 
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2: Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in which each column represents a tumor and each row represents a pathway.
-The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively.
-All 377 pathways are shown.
-Red signifies pathway disrupted pathways, grey signifies not disrupted.</p>
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
+Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
+which each column represents a tumor and each row represents a pathway. The
+pathways are ranked such that the pathway with the most disruptions across all
+7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
+top pathway disruptions successively. All 377 pathways are shown. Red signifies
+pathway disrupted pathways, grey signifies not disrupted.</p>
 
-To discover what pathways drive the UMAP clustering, we calculated enrichment for each pathway _in-class_ relative to all other classes combined (see methods) and ranked pathways for each class from highest to lowest enrichment.
-We visualized the enrichment as a heatmap (**figure 3**).
-Using this approach, we identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7, 0 pathways in class 8 (for which most pathways are putatively disrupted), 25 pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29 pathways in class 12.
-Some pairs of classes, in particular classes 6 & 7, classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways had enrichment.
-To explore the specific pathways marking each class, we projected disruptions for each of the 377 pathways onto the UMAP (**supplementary figure S4**).
-<mark>Class x</mark> was distinguished by metabolic pathways including RNA and protein biosynthesis (**supplementary figure S4**).
-Most class 4 enriched pathways were not unique to this class but shared in class 10 & 12, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution (as distinct from clinical ``stage'').
-Classes 9-12 have in common mutations in extracellular signaling, intracellular signaling, and immune-related pathways (see **figures 3** and **supplementary figure S4**).
-Cluster group 6-7 had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, $\beta$-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others.
-As cluster groups 9-12 and 6-7 were the most distinct of the 12, we chose to investigate these two groups further.
+To discover what pathways drive the UMAP clustering, we calculated enrichment
+for each pathway _in-class_ relative to all other classes combined (see methods)
+and ranked pathways for each class from highest to lowest enrichment. We
+visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
+identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
+class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
+class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
+0 pathways in class 8 (for which most pathways are putatively disrupted), 25
+pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
+pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
+classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
+had enrichment. To explore the specific pathways marking each class, we
+projected disruptions for each of the 377 pathways onto the UMAP
+(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
+metabolic pathways including RNA and protein biosynthesis (**supplementary
+figure S4**). Most class 4 enriched pathways were not unique to this class but
+shared in class 10 & 12, suggesting that this grouping reflects either greater
+heterogeneity or possibly a lower stage of tumor evolution (as distinct from
+clinical ``stage''). Classes 9-12 have in common mutations in extracellular
+signaling, intracellular signaling, and immune-related pathways (see **figures
+3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
+enrichment levels of the three, having mutations in hedgehog signaling,
+$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
+cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
+distinct of the 12, we chose to investigate these two groups further.
 
@@
 
-Since pathways like these share many common genes, we investigated whether tumors in this class  have multiple mutations across signaling pathways, or whether a small number of driver genes account for enrichment.
-The split between clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **supplementary figure S4**).
-Furthermore, clusters 9 and 11 are distinguished from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the _TP53_ panel of **Supplementary Figure S4**).
+Since pathways like these share many common genes, we investigated whether
+tumors in this class  have multiple mutations across signaling pathways, or
+whether a small number of driver genes account for enrichment. The split between
+clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
+homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
+**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
+from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
+_TP53_ panel of **Supplementary Figure S4**).
 
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited the highest pathway enrichment scores of the three major categories, though there is a significant number of tumors spatially distributed between cluster groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics of both.
-The top ten distinguishing pathways of class 6/7 tumors were "cellular response to hypoxia," "regulation of mitotic cell cycle," "assembly of the pre-replicative complex," "metabolism of polyamines," "degradation of $\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2 non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated with aspects of hedgehog signaling.
-To determine once again whether the genes driving pathway enrichment were commonly shared among these pathways we plotted and performed unsupervised clustering on tumors with an individual mutation matrix using every gene in each of the ten pathways, grouped by shared pathway membership (**figure 4**).
-Surprisingly, this class of tumors was not predominated by 1 or more highly mutated genes across all cancers.
-Instead, there is a whole set of genes in the proteosomal pathway whose activity contributes to all the aforementioned pathways, and the majority of tumors in these classes have a single mutation in one of these genes.
-Nonetheless, a significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have proteasomal mutations but instead harbor mutations in genes representing a majority of the same pathways, suggesting that most if not all of the pathways are individually important and that proteasomal mutations may be sufficient to drive the net effect of the other mutations in this cluster.
-We annotated the tumors in this class according to tissue of origin to determine whether proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and indeed we could observe some clustering.
-These data could explain how tissue-specific gene mutation signatures might result in general cancer phenotypes by converging at the pathway level.
+In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
+the highest pathway enrichment scores of the three major categories, though
+there is a significant number of tumors spatially distributed between cluster
+groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
+of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
+response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
+pre-replicative complex," "metabolism of polyamines," "degradation of
+$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
+non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
+destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
+with aspects of hedgehog signaling. To determine once again whether the genes
+driving pathway enrichment were commonly shared among these pathways we plotted
+and performed unsupervised clustering on tumors with an individual mutation
+matrix using every gene in each of the ten pathways, grouped by shared pathway
+membership (**figure 4**). Surprisingly, this class of tumors was not
+predominated by 1 or more highly mutated genes across all cancers. Instead,
+there is a whole set of genes in the proteosomal pathway whose activity
+contributes to all the aforementioned pathways, and the majority of tumors in
+these classes have a single mutation in one of these genes. Nonetheless, a
+significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
+proteasomal mutations but instead harbor mutations in genes representing a
+majority of the same pathways, suggesting that most if not all of the pathways
+are individually important and that proteasomal mutations may be sufficient to
+drive the net effect of the other mutations in this cluster. We annotated the
+tumors in this class according to tissue of origin to determine whether
+proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
+indeed we could observe some clustering. These data could explain how
+tissue-specific gene mutation signatures might result in general cancer
+phenotypes by converging at the pathway level.
 
@@
 
-<mark>As discussed previously, prior efforts to extract common signatures from pan-cancer datasets met with difficulty in distinguishing tumor samples from their tissue-specific -omics data signatures.
-Given our pathway-disruption based classification, this raises the question, are tumor phenotypes driven entirely by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance thresholds), or a combination of both?
-To answer this question, we ranked the top cluster-specific pathways for each cluster, and then tabulated the most frequently mutated genes for each pathway as a function of cancer type (**Table 1** and **Supplementary tables ST2-ST13**).
-We find that in most clusters the top pathways are determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations, wherein etc etc</mark>
+<mark>As discussed previously, prior efforts to extract common signatures from
+pan-cancer datasets met with difficulty in distinguishing tumor samples from
+their tissue-specific -omics data signatures. Given our pathway-disruption based
+classification, this raises the question, are tumor phenotypes driven entirely
+by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
+few samples to detect above statistical significance thresholds), or a
+combination of both? To answer this question, we ranked the top cluster-specific
+pathways for each cluster, and then tabulated the most frequently mutated genes
+for each pathway as a function of cancer type (**Table 1** and **Supplementary
+tables ST2-ST13**). We find that in most clusters the top pathways are
+determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
+However in many cancers we find divergence outside these top driver mutations,
+wherein etc etc</mark>
 
@@
 
-Following the same logic we used to investigate cluster-specific enrichment of pathways (**figure 3**) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters ( _e.g._ **figure 1l**).
-Using all non-metastatic tumors as background, we found very low levels of enrichment, less than $10\%$, in a handful of pathways.
-We reasoned that the individual clusters might be too different to detect global metastasis enrichment signal given the small sample size (n = 215 metastatic tumor samples).
+Following the same logic we used to investigate cluster-specific enrichment of
+pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
+tumors, as we find that metastatic tumors are distributed across all 12 clusters
+( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
+very low levels of enrichment, less than $10\%$, in a handful of pathways. We
+reasoned that the individual clusters might be too different to detect global
+metastasis enrichment signal given the small sample size (n = 215 metastatic
+tumor samples).
 
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and found 99 enriched pathways across all clusters (**table 2**).
-A number of enrichments were found in multiple clusters represented by pathways that were already shown to be enriched in a neighboring cluster.
-For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and 10.
-This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, **table 2**).
-This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters.
-These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).
-Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors.
-Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.
+Therefore, we calculated cluster-specific enrichment in metastatic tumors and
+found 99 enriched pathways across all clusters (**table 2**). A number of
+enrichments were found in multiple clusters represented by pathways that were
+already shown to be enriched in a neighboring cluster. For example,
+“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
+of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
+10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
+union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
+activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
+S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
+Cluster 1, one of two clusters for which we could not identify any
+distinguishing pathways (cluster 8 being the other), had enrichment in two
+pathways that were otherwise specific to non-metastatic tumors of other
+clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
+7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
+samples were enriched for two key neutrophil pathways, “Neutrophil
+Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
+cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
+metastases tend toward complementary mutations that are otherwise specific to
+non-metastatic tumors of other clusters.
 
-There were also a number of pathways that were enriched with lower specificity (among 4 or more clusters) which were found to be enriched in metastases.
-We identified 18 such pathways of which 12 had significantly higher mutation rates in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”, “Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF dependent signaling in response to WNT”, “Class I MHC mediated antigen processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
+There were also a number of pathways that were enriched with lower specificity
+(among 4 or more clusters) which were found to be enriched in metastases. We
+identified 18 such pathways of which 12 had significantly higher mutation rates
+in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
+“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
+dependent signaling in response to WNT”, “Class I MHC mediated antigen
+processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
+Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
+Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
 
@@
 
-If our molecular pathway disruption classes represent distinct biological states apart from tissue of origin, they may have different prognoses within cancer types or across all cancers.
-Kaplan-Meier plots reveal large differences between cancer types as expected, and some modest differences between our classes (not shown).
-These analyses are profoundly limited by confounding factors of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to name the obvious ones.
-To explore these ideas, we tested models of survival using public longevity data from the CDC, accounting for age, gender and ethnicity.
-To build on this relatively accurate model of baseline risk (see methods), we assume that individual cancer type captures some variation not accounted for in the longevity data ( _e.g._ all tissue-specific factors, including stage [see methods]), thus allowing us to avoid introducing uncertainty by modeling each of those factors explicitly.
-Therefore, our model estimates the effect of cancer type and class-specific cancer effects independently.
-We found that cancer types, as expected, have a range of prognoses relative to the general population, as shown in **figure 5a**.
-The least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a risk multiplier of between 15 and 22 relative to the background risk of death in the population.
-Among the deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers.
-Apart from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted death, the remaining cancer types ranged from about 1 to 5 in magnitude (**figure 5a**).
+If our molecular pathway disruption classes represent distinct biological states
+apart from tissue of origin, they may have different prognoses within cancer
+types or across all cancers. Kaplan-Meier plots reveal large differences between
+cancer types as expected, and some modest differences between our classes (not
+shown). These analyses are profoundly limited by confounding factors of age and
+stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
+name the obvious ones. To explore these ideas, we tested models of survival
+using public longevity data from the CDC, accounting for age, gender and
+ethnicity. To build on this relatively accurate model of baseline risk (see
+methods), we assume that individual cancer type captures some variation not
+accounted for in the longevity data ( _e.g._ all tissue-specific factors,
+including stage [see methods]), thus allowing us to avoid introducing
+uncertainty by modeling each of those factors explicitly. Therefore, our model
+estimates the effect of cancer type and class-specific cancer effects
+independently. We found that cancer types, as expected, have a range of
+prognoses relative to the general population, as shown in **figure 5a**. The
+least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
+(GBM) which has a risk multiplier of between 15 and 22 relative to the
+background risk of death in the population. Among the deadliest cancers outside
+of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
+from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
+death, the remaining cancer types ranged from about 1 to 5 in magnitude
+(**figure 5a**).
 
@@
 
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3, reflecting that some classes are either less deadly or more deadly than other classes within each cancer types, where the null hypothesis corresponds to an effect size=1 (**figure 5b**).
-For several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely indestinguishable from the prior, reflecting that either there were too few mortalities in the data to make an estimate (as expected for PRAD and THCA) or two few samples period.
-We did not observe class-specific trends that held true across cancer types, which could result from different cancers having different standards of care for example.
-In support of this interpretation, we also tested a factored model which considered both cancer type and each class independent of cancer type (**supplementary figure S7??**).
-Though we were able to successfully fit these models, they exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the factored model, here?</mark>)
+Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
+reflecting that some classes are either less deadly or more deadly than other
+classes within each cancer types, where the null hypothesis corresponds to an
+effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
+chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
+the posterior estimates are largely indestinguishable from the prior, reflecting
+that either there were too few mortalities in the data to make an estimate (as
+expected for PRAD and THCA) or two few samples period. We did not observe
+class-specific trends that held true across cancer types, which could result
+from different cancers having different standards of care for example. In
+support of this interpretation, we also tested a factored model which considered
+both cancer type and each class independent of cancer type (**supplementary
+figure S7??**). Though we were able to successfully fit these models, they
+exhibited much greater uncertainty, suggesting an inferior specification.
+(<mark>Dan are there additional arguments to be made for the inferiority of the
+factored model, here?</mark>)
 
@@
 
-All code for producing the analyses and figures herein are included in this fully reproducible manuscript.
-In addition, the R markdown files and all other models are available from repositories in the following locations on the distributed version software hub GitHub.
+All code for producing the analyses and figures herein are included in this
+fully reproducible manuscript. In addition, the R markdown files and all other
+models are available from repositories in the following locations on the
+distributed version software hub GitHub.
 
@@
 
-We calculated the enrichment of pathways in one set of tumors as the relative fraction (and probability of observing) of tumors with a mutated gene in the pathway to all tumors, inclusive of the category of interest.
-To do this operation, we use the beta distribution to permute a posterior distribution on the fraction of tumors with a pathway mutated for each category based on the observed set, and compared this to the posterior obtained from the full set of tumors (all tumors) as the distribution of differences between all permuted samples.
-We considered a pathway enriched if the $95\%$ range of credible differences thus obtained excludes 0 (as a null hypothesis) and the mean credible difference was greater than or equal to $20\%$ enrichment, which excludes a large number of small differences that are not likely to be biologically relevant.
+We calculated the enrichment of pathways in one set of tumors as the relative
+fraction (and probability of observing) of tumors with a mutated gene in the
+pathway to all tumors, inclusive of the category of interest. To do this
+operation, we use the beta distribution to permute a posterior distribution on
+the fraction of tumors with a pathway mutated for each category based on the
+observed set, and compared this to the posterior obtained from the full set of
+tumors (all tumors) as the distribution of differences between all permuted
+samples. We considered a pathway enriched if the $95\%$ range of credible
+differences thus obtained excludes 0 (as a null hypothesis) and the mean
+credible difference was greater than or equal to $20\%$ enrichment, which
+excludes a large number of small differences that are not likely to be
+biologically relevant.
 
@@
 
-We modeled baseline longevity in the U.S.
-with these different factors.
-<mark>Key assumptions?
-(Room here for further descriptions and equations)</mark> Using this base survival model, we constructed a likelihood function with parameters describing the overall lethality of each cancer type as a modifier $k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of $k$ and optimized model fit using stan[@stanref].
-The $k$ in our model can be taken as, roughly, the death rate relative to the baseline rates as depicted in **supplementary figure S5**.
-Thus, a value of $k=2$ means that a given cancer is twice as deadly.
+We modeled baseline longevity in the U.S. with these different factors.
+<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
+Using this base survival model, we constructed a likelihood function with
+parameters describing the overall lethality of each cancer type as a modifier
+$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
+$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
+taken as, roughly, the death rate relative to the baseline rates as depicted in
+**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
+twice as deadly.
 
-With an accurate model for baseline longevity in place, we sought to model cancer as an event based modifier of long-term survival.
-Thus any factors or events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for female vs.
-male caucasians, blacks and asians (hispanics are not clearly marked in the tumor data; see **supplementary figure S5** for comparisons of raw CDC data and baseline risk models).
-Thus we are able to accurately model baseline risk of death taking these factors into account.
-The data themselves encompass all other causes of death for these populations including cancer, thus freeing our model from concerns about other confounding factors ( _e.g._ Diabetes).
-For demonstration purposes we also illustrate some examples of the expected survival for a population that receives a diagnosis with doubled risk of death at ages 40, 50, 60, 70, and 80 (**supplementary figure S6**).
+With an accurate model for baseline longevity in place, we sought to model
+cancer as an event based modifier of long-term survival. Thus any factors or
+events such as cancer would modify the baseline risk function as a multiplier.
+The resulting coefficients can thus be interpreted as increasing baseline risk
+by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
+female vs. male caucasians, blacks and asians (hispanics are not clearly marked
+in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
+data and baseline risk models). Thus we are able to accurately model baseline
+risk of death taking these factors into account. The data themselves encompass
+all other causes of death for these populations including cancer, thus freeing
+our model from concerns about other confounding factors ( _e.g._ Diabetes). For
+demonstration purposes we also illustrate some examples of the expected survival
+for a population that receives a diagnosis with doubled risk of death at ages
+40, 50, 60, 70, and 80 (**supplementary figure S6**).
 
-As an aside, we did not explicitly account for stage at diagnosis in our model for the following reasons.
-In the tumor sample data, stage at diagnosis is confounded with cancer type because some cancers are screened agressively ( _e.g._ colorectal and prostate cancers) while others are diagnosed typically after they become problematic for the patient's lifestyle ( _e.g._ ovarian and pancreatic cancers).
-Secondly, such a model specification would suffer from added noise because the staging data are not well standardized across cancer types, have different criteria, and because it is unclear what the relationship between stage and advancement of disease is (for example some sub-stage 4 tumors are metastatic).
-To compound this latter issue, our tumor set is vastly under-powered given the uneven representation of stage across cancer types.
+As an aside, we did not explicitly account for stage at diagnosis in our model
+for the following reasons. In the tumor sample data, stage at diagnosis is
+confounded with cancer type because some cancers are screened agressively (
+_e.g._ colorectal and prostate cancers) while others are diagnosed typically
+after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
+pancreatic cancers). Secondly, such a model specification would suffer from
+added noise because the staging data are not well standardized across cancer
+types, have different criteria, and because it is unclear what the relationship
+between stage and advancement of disease is (for example some sub-stage 4 tumors
+are metastatic). To compound this latter issue, our tumor set is vastly
+under-powered given the uneven representation of stage across cancer types.
 
@@
 
-The basic problem in tumor research of this kind, namely that the search space, ~20,000 protein-coding genes is greater than the number of tumor samples [@hornNetSigNetworkbasedDiscovery2018].
-And that is not accounting for the fact that ~30k non-coding genes and hundreds of thousands of non-coding regulatory elements exist with oncogenic or tumor suppressor potential.
-Hence, we sought to investigate cancer as a disease of basic cellular and biochemical pathways, which we accomplished by translating gene-level mutations into pathway level disruptions.
-Importantly, we selectively removed pathways that lacked generality as cellular processes or that could be specifically assigned to differentiated cell types, while excluding others.
+The basic problem in tumor research of this kind, namely that the search space,
+~20,000 protein-coding genes is greater than the number of tumor samples
+[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
+that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
+elements exist with oncogenic or tumor suppressor potential. Hence, we sought to
+investigate cancer as a disease of basic cellular and biochemical pathways,
+which we accomplished by translating gene-level mutations into pathway level
+disruptions. Importantly, we selectively removed pathways that lacked generality
+as cellular processes or that could be specifically assigned to differentiated
+cell types, while excluding others.
 
-We hypothesize that cancer is the result of dysfunction from a limited number of basic cellular processes that are common to all living human cells, and thus all tumors belong on a spectrum of dysfunction on this axis regardless of tissue of origin.
-An alternative hypothesis is that every tumor belongs to one of a large number of syndromes which are unique to each tissue of origin, that share some mechanisms and superficial similarity in treatment strategies.
-The latter idea has gained momentum in recent years since the publication of Encode, Roadmap and TCGA consortium papers, perhaps owing to the intractable complexity of genomic data in all its forms.
-The inescapable conclusion thus far is that all these data point to tissue-of-origin as the most important driver of tumor behavior.
-Our observations contrast with this view, and instead support an interpretation in which all tumors manifest one of a limited number of metastable phenotypes represented by our classes (**figure 1a-k**), with increased fitness regardless of tissue of origin, which then converge on end stage cancer by accumulating complementary mutations to the other classes (**figure 1l** and **Table 2**).
+We hypothesize that cancer is the result of dysfunction from a limited number of
+basic cellular processes that are common to all living human cells, and thus all
+tumors belong on a spectrum of dysfunction on this axis regardless of tissue of
+origin. An alternative hypothesis is that every tumor belongs to one of a large
+number of syndromes which are unique to each tissue of origin, that share some
+mechanisms and superficial similarity in treatment strategies. The latter idea
+has gained momentum in recent years since the publication of Encode, Roadmap and
+TCGA consortium papers, perhaps owing to the intractable complexity of genomic
+data in all its forms. The inescapable conclusion thus far is that all these
+data point to tissue-of-origin as the most important driver of tumor behavior.
+Our observations contrast with this view, and instead support an interpretation
+in which all tumors manifest one of a limited number of metastable phenotypes
+represented by our classes (**figure 1a-k**), with increased fitness regardless
+of tissue of origin, which then converge on end stage cancer by accumulating
+complementary mutations to the other classes (**figure 1l** and **Table 2**).
 
-We find that most tumors segregate into one of 12 biologically distinct clusters, regardless of the cancer type (**figure 1h** and **supplementary figure S1**).
-Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS arm clusters 9 and 10, and breast cancers being excluded from the same cluster, see **supplementary figure S1**) are unevenly distributed toward a specific class, we could not identify cancer type that was exclusive to a single tumor class.
-Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental Figure S2**).
-Indeed, it would be surprising if we did not observe bias for some cancers, as it has long been known that certain treatment regimens are likely to work for a majority of patients of a given cancer type.
-Nonetheless, the basis for some tumors being treatment-refractory in spite of clinical indications remains elusive.
-Our analysis suggests the possibility of more general mechanisms underlying tumor evolution regardless of tissue, histological or cellular sub-type and could form a basis for future separation of patients into different precision treatment regimens.
+We find that most tumors segregate into one of 12 biologically distinct
+clusters, regardless of the cancer type (**figure 1h** and **supplementary
+figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
+arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
+see **supplementary figure S1**) are unevenly distributed toward a specific
+class, we could not identify cancer type that was exclusive to a single tumor
+class. Somewhat surprisingly to us, this finding extends to histological
+subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
+and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
+observe bias for some cancers, as it has long been known that certain treatment
+regimens are likely to work for a majority of patients of a given cancer type.
+Nonetheless, the basis for some tumors being treatment-refractory in spite of
+clinical indications remains elusive. Our analysis suggests the possibility of
+more general mechanisms underlying tumor evolution regardless of tissue,
+histological or cellular subtype and could form a basis for future separation of
+patients into different precision treatment regimens.
 
-It could be that the classes we identified are consistent with a signature of overall disease progression, such that successive, adjacent classes on the UMAP projection exhibit increasing specificity.
-We identified no such trend (see **figure 1k**) and each class instead was associated with unique combinations of pathways (as in **figure 3**).
-In our view, these observations suggest that cancers have available to them a limited number of stable configurations with respect to pathway disruption which are consistent with increased fitness and therefore positive selection.
+It could be that the classes we identified are consistent with a signature of
+overall disease progression, such that successive, adjacent classes on the UMAP
+projection exhibit increasing specificity. We identified no such trend (see
+**figure 1k**) and each class instead was associated with unique combinations of
+pathways (as in **figure 3**). In our view, these observations suggest that
+cancers have available to them a limited number of stable configurations with
+respect to pathway disruption which are consistent with increased fitness and
+therefore positive selection.
 
-Finally, we must remark on two of the classes, 1 and 8, which for which we did not find distinctive associations with pathways.
-Class 1 has a relatively low level of mutated pathways overall, although it is broadly enriched in many of the same tumor-promoting pathways common to the other groups.
-This likely does not reflect a true difference in stages of tumor evolution, as our data show clearly that this class is as likely to contain stage IV as it is to contain stage I tumors.
-We speculate this class represents a group of tumors with large aberrations in methylation, copy number or other genome rearrangements which we excluded from the present analysis.
-In the future, we hope to incorporate these other data into a pathway-centric analysis as we have done here for point mutations and indels.
-Class 8 represents a group of hyper-mutated tumors that harbor so many mutations that virtually no pathway is unaffected.
-This group highlights one of the limitations of the approach that we’ve taken, in that it is likely that a significant fraction of the samples flagged as “mutant” for each pathway are actually unaffected, consistent with those samples have low information content.
-Similar to class 1 we speculate that a future analysis of other data types, including improved filtering of mutations, may resolve these tumors into more meaningful categories.
+Finally, we must remark on two of the classes, 1 and 8, which for which we did
+not find distinctive associations with pathways. Class 1 has a relatively low
+level of mutated pathways overall, although it is broadly enriched in many of
+the same tumor-promoting pathways common to the other groups. This likely does
+not reflect a true difference in stages of tumor evolution, as our data show
+clearly that this class is as likely to contain stage IV as it is to contain
+stage I tumors. We speculate this class represents a group of tumors with large
+aberrations in methylation, copy number or other genome rearrangements which we
+excluded from the present analysis. In the future, we hope to incorporate these
+other data into a pathway-centric analysis as we have done here for point
+mutations and indels. Class 8 represents a group of hyper-mutated tumors that
+harbor so many mutations that virtually no pathway is unaffected. This group
+highlights one of the limitations of the approach that we’ve taken, in that it
+is likely that a significant fraction of the samples flagged as “mutant” for
+each pathway are actually unaffected, consistent with those samples have low
+information content. Similar to class 1 we speculate that a future analysis of
+other data types, including improved filtering of mutations, may resolve these
+tumors into more meaningful categories.
 
@@
 
-The fact that we are likely excluding many defects (under chromosomal abberations, amplifications, demethylation and so on) also raises the possibility that our classes are incomplete or may change at a future time when we can account for them.
-For now at least, we think that the point mutation data are representative of the most important pathways.
+The fact that we are likely excluding many defects (under chromosomal
+abberations, amplifications, demethylation and so on) also raises the
+possibility that our classes are incomplete or may change at a future time when
+we can account for them. For now at least, we think that the point mutation data
+are representative of the most important pathways.
 
@@
 
-By modeling CDC longevity data as a baseline risk function we showed that each of our pathway disruption classes exhibit cancer-type specific effects on expected survival (**figure 5**).
-Contrary to what one might expect at the outset, we found that models assuming different survival based on class _and_ cancer type fit better than models in which these variables are independent factors.
-However, considering that within each cancer type there are different clinical standards of care, and even within classes of drugs the preferred treatment can very between cancers, it makes sense in hindsight that they should all be different.
-Contrast the situation with ovarian vs breast cancer, which are both hormonally driven cancers, for example.
-Ovarian cancer has but one main treatment, platinum, whereas breast cancer patients have a variety of treatment regimens based on molecular subtype and other factors.
-It is easy to see how standard of care for one cancer and class could produce different outcomes than the same class of a different cancer type under the circumstances.
-Unfortunately, given the diversity drug classes and treatments (including non-chemical treatments such as surgery and palliative care), we lack sufficient power in the current study to explore these variables in the GDC data.
-It is our hope that more focused future studies will help to distinguish between treatment-specific effects or survival given different pathway disruption classes.
+By modeling CDC longevity data as a baseline risk function we showed that each
+of our pathway disruption classes exhibit cancer-type specific effects on
+expected survival (**figure 5**). Contrary to what one might expect at the
+outset, we found that models assuming different survival based on class _and_
+cancer type fit better than models in which these variables are independent
+factors. However, considering that within each cancer type there are different
+clinical standards of care, and even within classes of drugs the preferred
+treatment can very between cancers, it makes sense in hindsight that they should
+all be different. Contrast the situation with ovarian vs breast cancer, which
+are both hormonally driven cancers, for example. Ovarian cancer has but one main
+treatment, platinum, whereas breast cancer patients have a variety of treatment
+regimens based on molecular subtype and other factors. It is easy to see how
+standard of care for one cancer and class could produce different outcomes than
+the same class of a different cancer type under the circumstances.
+Unfortunately, given the diversity drug classes and treatments (including
+non-chemical treatments such as surgery and palliative care), we lack sufficient
+power in the current study to explore these variables in the GDC data. It is our
+hope that more focused future studies will help to distinguish between
+treatment-specific effects or survival given different pathway disruption
+classes.
 
@@
 
-Our findings imply that there are separate processes in the etiology of cancer that can be broadly thought of as general cancer promoting, class-specific mutations and metastasis.
-General cancer promoting processes must include factors that relate to genome stability either directly or indirectly.
-Such pathways are disrupted in all or most of the defined classes (see **figures 2** and **supplementary figure S1**) and often are the result of abberations involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
-Class-specific evolution must involve the acquisition of combinatorial disruptions to pathways that may individually be harmful but together produce more specialized cancer phenotypes and increased fitness (as in **figure 4** and **tables 1** and **ST2**).
-Importantly, our work does not distinguish between the order in which these mutations should accumulate.
-However, since many of the genes in the non-specific pathways involve genome stability, it is reasonable to hypothesize that such mutations promote the acquisition of class specific defects, and natural selection produces the classes we observed.
+Our findings imply that there are separate processes in the etiology of cancer
+that can be broadly thought of as general cancer promoting, class-specific
+mutations and metastasis. General cancer promoting processes must include
+factors that relate to genome stability either directly or indirectly. Such
+pathways are disrupted in all or most of the defined classes (see **figures 2**
+and **supplementary figure S1**) and often are the result of abberations
+involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic
+checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and
+**supplemental file 1**). Class-specific evolution must involve the acquisition
+of combinatorial disruptions to pathways that may individually be harmful but
+together produce more specialized cancer phenotypes and increased fitness (as in
+**figure 4** and **tables 1** and **ST2**). Importantly, our work does not
+distinguish between the order in which these mutations should accumulate.
+However, since many of the genes in the non-specific pathways involve genome
+stability, it is reasonable to hypothesize that such mutations promote the
+acquisition of class specific defects, and natural selection produces the
+classes we observed.
 
@@
 
-The class-specific metastatic pathway enrichment we observe (in **table 2**) may be a result of the pathways directly promoting metastatic phenotypes, or alternatively that such mutations are more likely to turn metastatic because they acquire a stronger or more complete oncogenic phenotype, since (again) we don’t know the order in which they were acquired.
-The fact that most metastatic enrichment is cluster-specific and has a tendency to overlap with cluster-specific pathways from non-metastatic tumors of neighboring clusters favors the latter explanation, and suggests a mechanism of complementarity in which newly acquired mutations converge on a single deadly phenotype that captures most or all of the critical features of end-stage cancer.
-If this is the case, we speculate that once we discover the vulnerabilities of each cluster we might also reasonably predict which pathways are susceptible to further selection and therefore target them  therapeutically at the same time to prevent further adaptation.
+The class-specific metastatic pathway enrichment we observe (in **table 2**) may
+be a result of the pathways directly promoting metastatic phenotypes, or
+alternatively that such mutations are more likely to turn metastatic because
+they acquire a stronger or more complete oncogenic phenotype, since (again) we
+don’t know the order in which they were acquired. The fact that most metastatic
+enrichment is cluster-specific and has a tendency to overlap with
+cluster-specific pathways from non-metastatic tumors of neighboring clusters
+favors the latter explanation, and suggests a mechanism of complementarity in
+which newly acquired mutations converge on a single deadly phenotype that
+captures most or all of the critical features of end-stage cancer. If this is
+the case, we speculate that once we discover the vulnerabilities of each cluster
+we might also reasonably predict which pathways are susceptible to further
+selection and therefore target them  therapeutically at the same time to prevent
+further adaptation.
 

### 2020-07-07 text/manuscript.Rmd

@@
 library(ggthemes)
+library(data.table)
+library(biomaRt)
+library(dplyr)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(reactome.db)
+library(plyr)
+library(cluster)
+library(pheatmap)
+library(clusterProfiler)
+library(org.Hs.eg.db)
+library(umap)
+library(dbscan)
+library(class)
+library(TCGAbiolinks)
+library(patchwork)
+library(ggplotify)
+library(ComplexHeatmap)
+library(knitr)
+library(kableExtra)
 ```
@@
 carrying mutations (this data is visualized in **figure 1a**).
-
+<div style="text-align: right"> Generation of Somatic Mutation Data </div>
 ```{r data-set-creation, eval=FALSE}
-
+## Data acquired from GDC
+tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
+## Somatic mutations focused in analysis
+selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
+## Filtering tcga_ssm for our selected mutations
+tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
+## Selecting column that we are interested in
+tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
+                   start = tcga_ssm$chromosome_start,
+                   end = tcga_ssm$chromosome_end,
+                   ref_allele = tcga_ssm$reference_genome_allele,
+                   alt_allele = tcga_ssm$mutated_to_allele,
+                   type = tcga_ssm$consequence_type,
+                   mut_type = tcga_ssm$mutation_type,
+                   ensembl_id = tcga_ssm$gene_affected,
+                   donor_id = tcga_ssm$icgc_donor_id,
+                   sample_id = tcga_ssm$submitted_sample_id,
+                   stringsAsFactors = F)
+## Removing duplicate rows
+tcga_ssm <- tcga_ssm %>% distinct()
+## Recategorizing the mutation type
+tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
+tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
+                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
+                                                 type == "splice_acceptor_variant" ~ "splice_variant",
+                                                 type == "splice_donor_variant" ~ "splice_variant",
+                                                 type == "start_lost" ~ "nonsense_variant",
+                                                 type == "stop_gained" ~ "nonsense_variant",
+                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
+                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
+                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
+                                                 type == "missense_variant" ~ "missense_variant",
+                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
+  dplyr::select(-mut_type) ## rm mut_type col
+
+## Preparing sample metadata ##
+
+complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
+
+tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
+tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
+  .[.$submitted_sample_id %in% complete_sample,] %>%
+  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
+  filter(type == "01") ## filtering 01 samples
+tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
+  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
+tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
+  rbind(., tcga_dup_id) ## rbind the fixed dups
+
+tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
+tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
+
+sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
+
+## Preparing gene metadata ##
+
+# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
+# ensembl_2_entrez_full_list <- bitr(genelist, 
+#                                    fromType = "ENSEMBL", 
+#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
+load("../data/ensembl_2_entrez_full_list.Rda")
+tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
+
+## Now we only have 19067 genes
+complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
+names(complete_ensembl) <- "ensembl_id"
+
+selected_mut <- unique(as.character(tcga_ssm$type))
+## For every mutation type.. select them.. and create a tsv file 
+for(selected_mut in selected_mut) {
+  print(selected_mut)
+  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
+    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
+    mutate(mutation = 1) %>% ## create a new column with 1's
+    distinct() %>% ## remove duplicated rows
+    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
+    remove_rownames() %>% ## move ensembl_id column to rownames
+    column_to_rownames(var = "ensembl_id") %>% 
+    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
+    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
+  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
+  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
+  names(missing_gene) <- names(df) ## insert sample_id names into col names
+  df <- rbind(df, missing_gene) ## rbind df and missing_gene
+  rm(missing_gene) ## rm data
+  gc()
+  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
+  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
+  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
+  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
+  df <- cbind(df, missing_sample) ## rbind df and missing_gene
+  rm(missing_sample) ## rm data
+  gc()
+  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
+  assign(paste0(selected_mut), value = df) ## create df in r environment
+  rm(df)
+  gc()
+  }
+
+# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
+ # genelist <- rownames(missense_variant)
+ # gene_data <- getBM(
+ #   attributes = c(
+ #     'ensembl_gene_id',
+ #     'external_gene_name',
+ #     'gene_biotype',
+ #     'chromosome_name', 
+ #     'start_position',
+ #     'end_position', 
+ #     'strand'),
+ #   filters = 'ensembl_gene_id',
+ #   values = genelist,
+ #   mart = ensembl)
+ # save(gene_data, file = "gene_data.Rda")
+load("../data/gene_data.Rda")
+######################
+
+missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
+UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
+UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
+frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
+frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
+splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
+nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
+nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
+
+## Matching df to the ensembl_id orders in gene_data
+match_ensembl <- function(x) {
+  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
+}
+
+missense_variant <- match_ensembl(missense_variant)
+UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
+UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
+frameshdel_variant <- match_ensembl(frameshdel_variant)
+frameshins_variant <- match_ensembl(frameshins_variant)
+splice_variant <- match_ensembl(splice_variant)
+nonstop_variant <- match_ensembl(nonstop_variant)
+nonsense_variant <- match_ensembl(nonsense_variant)
+
+genes <- GRanges(
+  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
+  ranges = IRanges(
+    start = gene_data$start_position,
+    end = gene_data$end_position),
+  strand = gene_data$strand,
+  mcols = list(
+    ensembl_gene_id = gene_data$ensembl_gene_name,
+    hgnc_symbol = gene_data$hgnc_symbol))
+
+gdc_tumor_coding_complete <- SummarizedExperiment(
+  assays = list(
+    missense   = missense_variant,
+    splice     = splice_variant,
+    frameshdel = frameshdel_variant,
+    frameshins = frameshins_variant,
+    utr3       = UTR_3_prime_variant,
+    utr5       = UTR_5_prime_variant,
+    nonsense   = nonsense_variant,
+    nonstop    = nonstop_variant),
+  rowRanges = genes,
+  colData = sample_sheet)
+save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
+rm(list = ls(all.names = T))
+gc()
 ```
-
-```{r kNNclusters, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE, fig.cap="**Classification of tumors by pathway disruptions.** a) Each of 377 selected Reactome pathways (columns) is classified as disrupted if one or more genes is mutated in the tumor sample (rows).
-Tumors are grouped by tissue of origin using standardized abrreviations from the TCGA project.
-b) MCA analysis was performed restricted to the 36 informative components.
-c) The resulting MCA was projected down to 3 dimensions for the purpose of density based clustering (only the first 2 dimensions are shown).
-Each dot represents a tumor d) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize the fraction of unassigned tumors.
-e) HDBSCAN clustered tumors were visualized on the UMAP projection; tumors are colored by cluster membership (see legend).
-This results in some tumors remaining unclassified (grey points).
-f) kNN was used to assign membership to the remaining tumors."}
-plot(1:10,1:10)
+<div style="text-align: right"> Reactome Pathway Creation </div>
+```{r reactome-pathway-creation, eval=FALSE}
+load("../data/gdc_tumor_coding_complete.Rdata")
+load("../data/gene_data.Rda")
+## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
+main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
+
+main2first <- main2first %>% .[,2:3] %>%
+  'colnames<-' (c("first_level", "main_level")) ## rename colnames
+
+## Filtering out master pathways from first2second_level
+first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
+first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
+  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
+  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
+  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
+
+ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
+ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
+  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
+  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
+  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
+
+## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
+rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
+                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
+                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
+                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
+                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
+                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
+                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
+                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
+
+## Ensembl to reactome pathway lookup table
+gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
+  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
+  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
+  distinct() %>% ## keep unique rows
+  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
+  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
+  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
+  replace(is.na(.), values = 0) ## replace NA's with 0's
+#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
+## Pathway names to level 2 lookup table
+pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
+  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
+  distinct(.) ## keep unique rows
+#save(pathname2level2, file = "pathname2level2.Rda")
+rm(rm_pathways, ensembl2reactome, main2first, first2second)
+gc()
 ```
-
-![](../figures/FIG1_TEMP.png)
+<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
+```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
+gtf_file <- "../data/gencode.v30.annotation.gtf"
+gencode_gtf <- rtracklayer::import(gtf_file)
+gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
+## We only want transcript only 
+gene_info <- gene_info[gene_info$type == "transcript",]
+## We only want protein_coding only
+gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
+gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
+## Filtering for transcript_id, gene_id, gene_name columns
+g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
+## Condense df to only having unique data. 
+good.g2t <- as_tibble(unique(g2t))
+
+## df with transcript_id and entrez id                      
+gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
+head(gencode_entrez[1:10,])
+colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
+## Joining data that are joined together by transcript_id
+entrez_g2t <- left_join(good.g2t, gencode_entrez)
+head(entrez_g2t[1:10,])
+## Remove transcript_id column
+entrez_g2t$transcript_id <- NULL
+## Remove . and the numbers following from the ensembl id.
+entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
+## Removing data that does not have entrez id. 83091 out of 83647 had NA values
+table(is.na(entrez_g2t$entrez_id))
+entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
+entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
+## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
+table(duplicated(entrez_g2t$gene_id))
+## These are pseudogenes... We will remove them all.
+dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
+## We're removing the rows that are associated with this pseudogene
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
+## It's gone
+which(entrez_g2t$gene_id == "ENSG00000285607")
+## Are there any duplicate gene names? Yes! 14 of them
+table(duplicated(entrez_g2t$gene_name))
+dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
+entrez_g2t[entrez_g2t$gene_name == "TBCE",]
+## ENSG00000285053  is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
+entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
+## ENSG00000284741 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
+entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
+## ENSG00000285258 is not in the database anywhere
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
+entrez_g2t[entrez_g2t$gene_name == "MATR3",]
+## ENSG00000280987 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
+entrez_g2t[entrez_g2t$gene_name == "SOD2",]
+## ENSG00000285441 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
+entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
+## ENSG00000285292 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
+entrez_g2t[entrez_g2t$gene_name == "PINX1",]
+## ENSG00000258724 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
+entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
+## ENSG00000284024 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
+entrez_g2t[entrez_g2t$gene_name == "IGF2",]
+## ENSG00000284779 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
+entrez_g2t[entrez_g2t$gene_name == "ATF7",]
+## ENSG00000267281 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
+entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
+## ENSG00000284934 is a homolog
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
+entrez_g2t[entrez_g2t$gene_name == "GGT1",]
+## ENSG00000286070 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
+entrez_g2t[entrez_g2t$gene_name == "SCO2",]
+## ENSG00000130489 no longer exist
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
+entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
+## ENSG00000269226 is a readthrough
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
+
+## Are there any dupped entrez id? 
+dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
+
+entrez_g2t[entrez_g2t$entrez_id == "55486",]
+## ENSG00000283765 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
+entrez_g2t[entrez_g2t$entrez_id == "55300",]
+## ENSG00000281028 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
+entrez_g2t[entrez_g2t$entrez_id == "653082",]
+## ENSG00000286094 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
+entrez_g2t[entrez_g2t$entrez_id == "8622",]
+## ENSG00000284762 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
+entrez_g2t[entrez_g2t$entrez_id == "221468",]
+## ENSG00000286105 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
+entrez_g2t[entrez_g2t$entrez_id == "1124",]
+## ENSG00000285162 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
+entrez_g2t[entrez_g2t$entrez_id == "102723849",]
+## ENSG00000283765 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
+entrez_g2t[entrez_g2t$entrez_id == "883",]
+## ENSG00000286112 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
+entrez_g2t[entrez_g2t$entrez_id == "22891",]
+## ENSG00000261130 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
+entrez_g2t[entrez_g2t$entrez_id == "54816",]
+## ENSG00000285253 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
+entrez_g2t[entrez_g2t$entrez_id == "23370",]
+## ENSG00000268861 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
+entrez_g2t[entrez_g2t$entrez_id == "284391",]
+## ENSG00000286098 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
+entrez_g2t[entrez_g2t$entrez_id == "440519",]
+## ENSG00000283201 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
+entrez_g2t[entrez_g2t$entrez_id == "149951",]
+## ENSG00000285382 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
+entrez_g2t[entrez_g2t$entrez_id == "388820",]
+## ENSG00000268861 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
+entrez_g2t[entrez_g2t$entrez_id == "8471",]
+## ENSG00000286098 not found on ncbi
+entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
+
+## manual gene removal complete
+
+load("../data/TCGA.RNA.Rda")
+
+## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
+tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
+  as.data.frame(.) %>%
+  dplyr::select(project_code, submitted_specimen_id) %>%
+  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
+
+TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
+  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
+  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
+  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
+  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
+
+g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
+
+express_no_express_lookup <- list()
+for(i in unique(tumor_samples$project_code)) {
+  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
+  df <- tumor_samples[df,] ## select only those rows
+  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
+  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
+    mean(x)
+  })) ## take the mean for each gene across the tumor samples
+  names(mean_gdc) <- i
+  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
+  rm(df, gdc, mean_gdc)
+  gc
+}
+express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
+express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
+
+## 
+gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
+  rownames(.)
+g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
+
+express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
+express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
+express_no_express_lookup[express_no_express_lookup >10] <- 1
+
+## convert entrez id to ensembl_id
+express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
+  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
+
+## Booleanize mutation count for each tumor sample
+missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
+splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+
+gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
+gc()
+gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
+gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
+
+gdc_gene_exp_adj <- list()
+for(i in unique(tumor_samples$project_code)) {
+  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
+  df <- tumor_samples[df,] ## select only those rows
+  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
+  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
+  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
+  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
+  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
+    exp_data * x
+  })) ## Function to multiply lookup to the data for the specific cancer type
+  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
+  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
+  rm(df, gdc, exp_data, gdc_adj)
+  gc()
+}
+## Combining list into a dataframe.
+gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
+names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
+rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
+gc()
+save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
+```
+<div style="text-align: right"> Genes to pathway transformation </div>
+```{r genes-to-pathways-transformation, eval=FALSE}
+load("../data/gdc_gene_exp_adj.Rda")
+load("../data/gene2_level2_lookup.Rda")
+sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
+complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
+
+df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
+df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
+df_gene <- as.data.frame(df_gene) ## turn matrix back to df
+
+df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
+  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
+  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
+  dplyr::select(-mutation) ## Remove column named mutation
+
+path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
+path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
+path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
+
+path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
+  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
+  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
+  dplyr::select(-path_member)
+
+df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
+  na.omit(.) ## Omit NA's that were generated due to no matches
+
+## getting the genes involved in our analysis
+genes_in_pathanalysis <- unique(df_lookup$gene)
+write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
+
+df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
+
+df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
+  remove_rownames() %>%
+  column_to_rownames(var = "Var2") ## path_id to rownames
+
+## Insert missing tumor samples due to no mutations
+df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
+mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
+names(mx) <- df_missing ## Missing tumor sample names in col names
+rownames(mx) <- rownames(df_lookup) ## Path names in rownames
+df_lookup <- cbind(df_lookup, mx) ## cbind data
+
+## Insert missing pathway due to no mutations
+path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
+mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
+names(mx) <- names(df_lookup) ## Insert tumor sample names
+rownames(mx) <- path_missing ## Insert missing path names
+df_lookup <- rbind(df_lookup, mx) ## cbind data
+
+pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
+gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
+  'rownames<-' (pathway_order) %>%
+  rownames_to_column(var = "PATH_ID") %>%
+  left_join(., pathname2level2, by = "PATH_ID") %>%
+  dplyr::select(-PATH_ID) %>% 
+  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
+rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
+gc()
+save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
+```
+<div style="text-align: right"> Figure 1 data </div>
+```{r tumor-clusters, eval = FALSE}
+####################
+## Figure 1A data ##
+####################
+load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/gdc_tumor_coding_complete.Rdata")
+
+path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
+path_daisy <- as.dist(as.matrix(path_daisy))
+
+path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
+path_daisy_row <- as.dist(as.matrix(path_daisy_row))
+
+pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
+         show_rownames = F,
+         show_colnames = F,
+         cellheight = 1,
+         cellwidth = 0.5,
+         clustering_distance_cols = path_daisy,
+         clustering_distance_rows = path_daisy_row,
+         cluster_cols = T,
+         cluster_rows = T,
+         cutree_cols = 19,
+         annotation_legend = F,
+         legend_breaks = c(0, 0.25, 0.75, 1),
+         legend_labels = c("","Not Mut", "Mut", ""),
+         color = c("grey", "firebrick3"),
+         clustering_method = "ward.D2",
+         legend = T)
+dev.off()
+pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
+df <- rownames(gdc_reactome_path_analysis)
+pathway_dendro <- df[pathway_dendro]
+
+tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
+  as.data.frame(.) %>%
+  dplyr::select(project_code, submitted_specimen_id) %>%  
+  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
+
+cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
+
+cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
+
+for(n in cancer_type) {
+  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
+    filter(project_code == n) %>% ## filter for specific cancer type
+    .[, "sample_id"] ## retain only sample_id
+  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
+  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
+                          show_rownames = F,
+                          show_colnames = F,
+                          cellheight = 1,
+                          cellwidth = 0.5,
+                          cluster_cols = T,
+                          cluster_rows = F,
+                          annotation_legend = F,
+                          color = c("grey", "firebrick3"),
+                          clustering_method = "single", ## ordering is single linkage
+                          legend = T)
+  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
+  cancer_path <- names(cancer_path) ## getting the sample_id original order 
+  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
+  rm(cancer_info, cancer_path)
+  gc()
+}
+dev.off()
+
+for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
+  print(n)
+  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
+  names(cancer_order) <- "sample_id"
+  cancer_order$type <- n
+  for(p in cancer_type[-1]) {
+    print(p)
+    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
+    names(df1) <- "sample_id"
+    df1$type <- p
+    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
+    rm(df1)
+    gc()
+  }
+}
+
+cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
+
+cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
+  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
+cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
+
+## set levels based on clustering algorithm above
+cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
+cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
+
+cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
+save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
+
+####################
+## Figure 1C data ##
+####################
+gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
+  rownames_to_column(var = "sample") %>% ## move path_names to column
+  .[match(rownames(tumor_samples), .$sample),]
+
+## umap analysis
+set.seed(3296)
+gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
+gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
+
+####################
+## Figure 1E data ##
+####################
+## To identify clusters using hdbscan, use umap in the third dimension
+set.seed(3296)
+gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
+gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
+rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
+hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
+umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
+
+####################
+## Figure 1F data ##
+####################
+## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
+umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
+umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
+                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
+                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
+                                                                  k=15)
+
+####################
+## Figure 1G data ##
+####################
+missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
+splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+
+mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
+gc()
+mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
+  colSums() %>% as.data.frame() %>% ## sum the mutation count
+  'colnames<-' (c("mut_count")) %>% ## rename colname
+  rownames_to_column(var = "V1") ## move rownames to column
+
+umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
+
+####################
+## Figure 1H data ##
+####################
+tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
+  as.data.frame(.) %>%
+  dplyr::select(project_code) %>% 
+  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
+  rownames_to_column(var = "V1")
+
+umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
+
+####################
+## Figure 1I data ##
+####################
+## getting subtype information on our cancer type
+subtypes <- PanCancerAtlas_subtypes() 
+tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
+subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
+  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
+  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
+  distinct() %>% ## only unique columns
+  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
+  filter(Subtype_Selected != "Normal") %>%
+  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
+
+#save(subtypes, file = "subtypes.Rda")
+load("../data/subtypes.Rda")
+
+umap_2d_3d_coors <- umap_2d_3d_coors %>%
+  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
+  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
+  dplyr::select(-pan.samplesID)
+
+##########################
+## Figure 1K and L data ##
+##########################
+## getting tumor_metastasis, staging
+tcga_clinical_tumor_staging <- list()
+for(n in cancer_type) {
+  print(n)
+  cancername <- paste0("TCGA-",n)
+  query <- GDCquery(project = cancername,
+                  data.category = "Clinical",
+                  data.type = "Clinical Supplement",
+                  data.format = "BCR Biotab",
+                  file.type = "patient")
+  GDCdownload(query)
+  clinical <- GDCprepare(query)[[1]]
+  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
+    cat("Yep, ajcc info in there!\n")
+    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
+  }
+  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
+    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
+    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
+    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
+  }
+  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
+    cat("Nope, ajcc info not there!\n")
+    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
+    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
+    }
+  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
+  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
+  tcga_clinical_tumor_staging[[n]] <- clinical
+}
+
+tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
+tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
+missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
+
+tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
+  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
+  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
+  rbind(tcga_clinical_tumor_staging, .) ## rbind
+
+save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
+
+umap_2d_3d_coors <- umap_2d_3d_coors %>%
+  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
+  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
+  dplyr::select(-bcr_patient_barcode)
+
+M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
+M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
+NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
+
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
+
+stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
+stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
+stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
+stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
+unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
+
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
+
+write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
+```
+
+```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
+load("../data/cancer_disrupt_heatmap.Rda")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
+fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
+  geom_tile(size = 0.6) +
+  scale_fill_manual(values = c("grey", "firebrick3")) +
+  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
+scale_y_discrete(position = "left") +
+  labs(tag = "A") +
+  ylab("") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        strip.text.y = element_text(size = 12, hjust = 0),
+        strip.text.y.left = element_text(angle = 0),
+        strip.background = element_blank(),
+        legend.position = "none")
+
+## fixing the text cutoff in heatmap
+ fig1Ag <- ggplotGrob(fig1A)
+
+ for(i in which(grepl("strip-l", fig1Ag$layout$name))){
+   fig1Ag$grobs[[i]]$layout$clip <- "off"
+ }
+
+ fig1A <- as.ggplot(fig1Ag)
+ rm(fig1Ag)
+
+## umap monochrome 2d
+fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
+  geom_point(colour = "grey") +
+  xlab("umap_1") +
+  ylab("umap_2") +
+  theme_classic() +
+  labs(tag = "C") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+## umap hdb_scan
+fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
+  guides(color = F) +
+  theme_classic() +
+  xlab("umap_1") +
+  ylab("umap_2") +
+  theme_classic() +
+  labs(tag = "E") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+
+## umap knn_clust 
+fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
+  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
+  guides(color = F) +
+  theme_classic() +
+  xlab("umap_1") +
+  ylab("umap_2") +
+  theme_classic() +
+  labs(tag = "F") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+
+umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
+
+## violin mutation count
+fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
+    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
+    geom_boxplot(aes(fill = knn_clust), width=0.2) +
+    theme_classic() +
+    scale_y_log10() +
+    xlab("") +
+    ylab("Log(Mutation Count)") +
+    labs(color = "", tag = "G") +
+    theme(axis.text.x = element_text(size = 12),
+          axis.text.y = element_text(size = 12),
+          axis.title.x = element_text(size = 12),
+          axis.title.y = element_text(size = 12),
+          legend.position = "none") +
+  scale_fill_manual(values = kelly.colours[-c(1:2)])
+
+## umap by cancer type
+fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "H", color = "") +
+  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
+                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap of brca subtype
+fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() +
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "I", color = "") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap brca somatic mutations
+load("../data/gdc_gene_exp_adj.Rda")
+load("../data/gene_data.Rda")
+brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
+brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
+  rownames_to_column(var = "ensembl_gene_id") %>%
+  left_join(., brca, by = "ensembl_gene_id") %>% 
+  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
+  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
+  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
+  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
+                           BRCA1 == "0" ~ "NA"),
+         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
+                           BRCA2 == "0" ~ "NA")) %>%
+  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
+  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
+  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
+  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
+
+fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
+  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(color = "", tag = "J") +
+  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
+                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap stage
+fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "K", color = "") +
+  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
+                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap metastasis
+fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "L", color = "") +
+  scale_color_manual(labels = c("M0", "M1"), 
+                     values = c('#a6cee3','#1f78b4')) +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+layout <- '
+AAAABBCCDD
+AAAAEEEFFF
+AAAAGGGG##
+AAAAGGGG##
+AAAAHHIIJJ
+AAAAKKLL##
+'
+
+fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
+
+```
+
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
@@
 Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of tumor stage onto the UMAP,
-regardless of cancer type. k) Projection of metastatic status onto the UMAP.
+cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
+UMAP. k) projection of tumor stage onto the UMAP,
+regardless of cancer type. l) Projection of metastatic status onto the UMAP.
 Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
@@
 
-### <mark>Selection of pathways <- Peter to write</mark>
+### Selection of pathways
+
+Pathway analysis provides additional information that genes and gene ontologies
+alone lack: 1) it aggregates molecular events across multiple genes in the same
+pathway and 2) the results, interpretable as genetic alterations, are related 
+to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
+response to hypoxia”. Using the Reactome Pathway Database, we selected the 
+second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
+signaling’ in ‘Signal Transduction’) which we believed was a suitable 
+representation of the basic cellular processes and biochemical pathways. 
+We also excluded pathways categorized under cancer, disease, gene defect, 
+disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
+Signaling in Cancer’). While some of the excluded pathways have been shown
+to play an important role in the cancer, they focus on a specific point 
+mutations. In addition, for most of the excluded pathways, there is a global
+pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
+the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
+This operation produced a lookup table that consisted of 377 pathways mapped 
+to 9,910 genes.
 

### 2020-07-08 text/manuscript.Rmd

@@
 bibliography: references.bib
+editor_options: 
+  chunk_output_type: console
 ---
@@
 exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1g**
+cancer types had representative tumors in every class (for example **figure 1h**
 and full tumor profiles in **supplementary figure S1**, see also interactive
@@
 frequency, possibly owing to relative diversity among all tumors.
-
-![](../figures/figure2_temp.png)
-
+<div style="text-align: right"> Figure 2 Data </div>
+```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+load("../data/gdc_reactome_path_analysis.Rda")
+
+######################
+## oncoprint top100 ##
+######################
+onco_color <- c("0" = "grey", "1" = "firebrick3")
+
+top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
+
+top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
+
+top100path[top100path == "1"] <- "MUT" ## categorize boolean values
+top100path[top100path == "0"] <- " "
+  col <- c("MUT" = "firebrick3")
+heatmap_legend <- list(title = "", at = c("MUT"),
+                       labels = c("Mutation"))
+alter_fun = list(
+    background = function(x, y, w, h) {
+        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
+            gp = gpar(fill = "#CCCCCC", col = NA))
+    },
+    # bug red
+    MUT = function(x, y, w, h) {
+        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
+            gp = gpar(fill = col["MUT"], col = NA))
+    })
+  column_title <- "Top 100 Pathway PanCancer"
+  onco_res <- oncoPrint(top100path,
+    alter_fun = alter_fun, col = col,
+    column_title = column_title, heatmap_legend_param = heatmap_legend,
+    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
+    top_annotation = NULL
+    )
+
+clust_col_order <- names(top100path[,onco_res@column_order])
+path_row_order <- rownames(top100path[onco_res@row_order,])
+
+onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
+  rownames_to_column(var = "path_name") %>% ## path_name to col
+  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
+  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
+         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
+save(onco_ggplot, file = "onco_ggplot.Rda")
+
+#####################
+## oncoprint panel ##
+#####################
+clust_col_order <- list()
+path_row_order <- list()
+
+for(n in 1:12) {
+  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
+  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
+  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
+  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
+  top10path[top10path == "0"] <- " "
+  col <- c("MUT" = "firebrick3")
+heatmap_legend <- list(title = "", at = c("MUT"),
+                       labels = c("Mutation"))
+alter_fun = list(
+    background = function(x, y, w, h) {
+        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
+            gp = gpar(fill = "#CCCCCC", col = NA))
+    },
+    # bug red
+    MUT = function(x, y, w, h) {
+        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
+            gp = gpar(fill = col["MUT"], col = NA))
+    })
+  column_title <- "test"
+  onco_res <- oncoPrint(top10path,
+    alter_fun = alter_fun, col = col,
+    column_title = column_title, heatmap_legend_param = heatmap_legend,
+    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
+    top_annotation = NULL
+    )
+  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
+  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
+}
+
+save(clust_col_order, path_row_order, file = "onco_clust.Rda")
+```
+
+```{r figure-2-plot, fig.width=72, fig.height=48}
+## top100 pathways pancancer oncoprint
+load("../data/onco_ggplot.Rda") ## load data
+onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
+
+fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
+  geom_tile(aes(fill = as.character(mutation_status))) +
+  scale_fill_manual(values = onco_color) +
+  xlab(label = "sample_id") +
+  labs(tag = "A") +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_text(size = 12),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line = element_blank(),
+        legend.position = "none")
+
+## top10 pathways for each cluster
+load("../data/onco_clust.Rda") ## load in data
+load("../data/gdc_reactome_path_analysis.Rda")
+
+plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
+onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
+figure_labs <- LETTERS[seq(from = 2, to = 13)]
+for(n in 1:12) {
+  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
+  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
+  rownames_to_column(var = "path_name") %>% ## path_name to col
+  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
+  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
+         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
+  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
+  geom_tile(aes(fill = as.character(mutation_status))) +
+  scale_fill_manual(values = onco_color) +
+  xlab(label = paste0("Class ", n)) +
+  labs(tag = figure_labs[n]) +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line = element_blank(),
+        legend.position = "none") 
+}
+
+## setting patchwork panel layout
+layout <- '
+AABCD
+AABCD
+AABCD
+AAEFG
+AAEFG
+AAEFG
+AAHIJ
+AAHIJ
+AAHIJ
+AAKLM
+AAKLM
+AAKLM
+'
+
+fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
+```
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
@@
 distinct of the 12, we chose to investigate these two groups further.
-
-![](../figures/fig2B_short.png)
+<div style="text-align: right"> Figure 3 data </div>
+```{r group-pathway-enrichment-data-generation, eval=FALSE}
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
+
+## Getting the total number of tumor samples in each cluster
+n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
+  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
+  mutate(clust = as.character(clust)) ## convert factor to character
+
+## Organizing rBeta data
+load("../data/gdc_reactome_path_analysis.Rda")
+clust_group <- n_tumor_in_clust$clust
+cluster_path_count <- list()
+for(n in clust_group) {
+  print(n)
+  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
+    .[,"V1"] ## get sample_ids
+  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
+    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
+    'colnames<-' (n) ## change colname to cluster number
+  cluster_path_count[[n]] <- path_mutation
+  rm(path_mutation, sample_id, n)
+  gc()
+}
+cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
+
+global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
+  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
+  'colnames<-' (clust_group) %>% ## rename column with tumor class
+  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
+  gather(., key = "clust", value = "global_mut", -path)
+
+rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
+  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
+  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
+  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
+  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
+
+rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
+gc()
+
+enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
+                                            clust = rbeta_path_analysis$clust,
+                                            fg_mut = rbeta_path_analysis$clust_mut,
+                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
+                                            bg_mut = rbeta_path_analysis$global_mut,
+                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
+                                            )
+fg_mut_rate <- list()
+bg_mut_rate <- list()
+for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
+  print(n)
+  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
+  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
+  fg_numerator <- sum(df$fg_mut)
+  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
+  bg_numerator <- sum(df$bg_mut)
+  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
+  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
+  rm(df)
+}
+
+fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
+  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
+  rownames_to_column(var = "clust") ## move rownames to col
+bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
+  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
+  rownames_to_column(var = "clust") ## move rownames to col
+
+enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
+  left_join(., bg_mut_rate, by = "clust") %>%
+  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
+  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
+  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
+  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
+rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
+gc()
+
+moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
+})
+
+moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
+})
+
+## ModerateEnrichment probability 
+prob <- moderate_fg > moderate_bg ## boolean data
+mean_prob <- apply(prob, MARGIN = 2, function(x) {
+  mean(x)
+})
+rm(prob)
+gc()
+## Moderate Difference
+CI <- 0.025 ## setting a 95% confidence interval
+
+diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
+rm(moderate_fg, moderate_bg, prob)
+gc()
+
+sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
+}))
+quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  quantile(x, c(CI, 0.5, 1-CI))
+}))
+
+moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
+                                   clust = enrich_prob_uncertainity_in_bg$clust,
+                                   probability = mean_prob,
+                                   signif = sig_fg_bg[,1],
+                                   difference = quant_diff[,2],
+                                   lower = quant_diff[,1],
+                                   upper = quant_diff[,3])
+rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
+gc()
+#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
+load("../data/moderate_enrich_prob.Rda")
+
+diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
+diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
+diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
+
+effect_size <- pheatmap(diff_heatmap,
+         show_rownames = F,
+         show_colnames = T,
+         cluster_cols = T,
+         cluster_rows = T,
+         cellwidth = 10,
+         cellheight = 0.76,
+         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
+         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
+         clustering_method = "ward.D2")
+
+effect_row_order <- effect_size$tree_row$order
+effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
+
+effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
+  gather(., key = "cluster", value = "effect_size", -path_name)
+
+effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
+effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
+effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
+
+effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
+  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
+  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
+  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
+  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
+  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
+  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
+  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
+  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
+  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
+  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
+  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
+))
+effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
+
+save(effect_heatmap, file = "effect_heatmap.Rda")
+```
+
+```{r group-pathway-enrichment heatmap}
+load("../data/effect_heatmap.Rda") ## load data
+
+effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
+
+fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
+  geom_tile(size = 0.6, aes(fill = discrete_color)) +
+  scale_fill_manual(values = effect_colors) +
+  theme(axis.text.x = element_text(size = 12),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "none")
+
+fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
+  geom_bar(stat = "identity", width = 1) +
+  theme_classic() +
+  scale_fill_manual(values = effect_colors) + 
+  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
+  theme(legend.position = "none",
+        axis.text.y = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        axis.line = element_blank())
+
+## setting patchwork panel layout
+layout <- '
+AAAAA
+AAAAA
+AAAAA
+AAAAA
+AAAAA
+AAAAA
+AAAAA
+AAAAA
+#BBB#
+'
+
+fig3 + fig3_legend + plot_layout(design = layout)
+```
 
@@
 ## Supplementary Information
-
+```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
+
+umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+
+ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
+  geom_point(size = 1) +
+  theme_classic() +
+  coord_fixed() +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  scale_colour_manual(values = rep("firebrick3", 23)) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
+```
 ![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._](../figures/figS2.png)
-
-![**Supplementary Figure S2.
-Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_](../figures/figS3.png)
-
-![**Supplementary Figure S3.
-Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_](../figures/figS4.png)
+Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
+
+```{r supplementary-figure-s2-data-generation}
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
+umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+
+ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
+  geom_point(size = 0.5) +
+  theme_classic() +
+  coord_fixed() +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
+```
+
+![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
+<div style="text-align: right"> Supplementary Figure S3 Data </div>
+```{r supplementary-figure-s3-data-generation, eval = FALSE}
+load("../data/gdc_tumor_coding_complete.Rdata")
+load("../data/gene_data.Rda")
+
+## Booleanize mutation count for each tumor sample
+missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
+splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+
+gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
+gc()
+gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
+
+##oncogene
+oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
+##tumor_supressor
+tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
+
+selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
+
+selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
+  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
+  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
+  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
+
+selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
+  left_join(., selected_mut, by = "V1") %>%
+  melt(data = ., id.vars = c("V1", "x_2d", "y_2d"), meausure.vars = tumor_suppressor$external_gene_name) %>%
+  filter(value != 0)
+
+save(selected_mut, file = "selected_mut.Rda")
+```
+
+```{r supplementary-figure-s3}
+load("../data/selected_mut.Rda")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
+umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+
+ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
+  geom_point(size = 0.5) +
+  theme_classic() +
+  coord_fixed() +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  scale_colour_manual(values = rep("firebrick3", 20)) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
+                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
+                      ncol = 5) ## facet by gene
+```
+
+![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
 ![**Supplemental figure S6.

### 2020-07-08 text/manuscript.Rmd
add Hanahan and Weinberg reference

@@
 expression and genomic signatures, with the express goal of understanding the
-common themes of all cancer independent of tissue of origin. The first of these
-(the “Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
+common themes of all cancer independent of tissue of origin. Such a view of
+cancer as a malady affecting common cellular processes has long been a goal of
+the field and is perhaps best characterized in an essay by @weinberg2000 (see also
+Discussion). The first of these studies, (the
+“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
 using integrative analysis with co-equal weighting of gene expression,
@@
 [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
-could be further related or subdivided into organ specific groups, including
+could be further subdivided into organ specific groups, including
 pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/

### 2020-07-08 text/manuscript.Rmd
add fig s4, table st1, incomplete fig 4 - still needs work

--- 2020-07-08 text/manuscript.Rmd
@@
 library(kableExtra)
+library(UpSetR)
 ```
@@
 
+```{r figure-4-data-generation, eval = FALSE}
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+load("../data/gene2_level2_lookup.Rda")
+load("../data/pathname2level2.Rda")
+load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/gene_data.Rda")
+
+## Parameters ##
+filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
+n_path <- 10 ## looking at the top10 path
+
+## Data Generation ##
+class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
+
+toppath <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top10 mutated pathways
+
+genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
+  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
+  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
+  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
+  .[rowSums(.[, -1])>0,]
+
+df <- upset(genes_in_path, 
+      sets = names(genes_in_path),
+      mb.ratio = c(0.55, 0.45), 
+      order.by = "freq", 
+      point.size = 3.5,
+      line.size = 2.0,
+      mainbar.y.label = "Gene Intersection",
+      sets.x.label = "Total Gene Per Pathway",
+      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
+      number.angles = 0)
+
+make_combinations <- function(x) {
+  l <- length(x)
+  mylist <- lapply(1:l, function(y) {
+    combn(x, y, simplify = FALSE)
+  })
+  mylist
+}
+
+path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
+
+top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
+  'colnames<-' ("sum_top10") %>% ## rename col
+  rownames_to_column(var = "gene_name") ## make rownames a col
+
+path_match_in_set <- list() ## making a list for matchings in sets
+
+for(n in 1:10) {
+  message(n)
+  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
+  path_match_within_set <- list() ## making a list for matching within sets
+  for(p in 1:length(path_combo_set)) {
+    message(p)
+    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
+    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
+    if(length(spec_combo) > 1) {
+      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
+        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
+    }
+    if(length(spec_combo) == 1) {
+      names(spec_combo) <- "combo"
+    }
+    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
+    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
+    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
+    if(!"TRUE" %in% any_matches) { ## if no matches then skip
+      next
+      }
+    if("TRUE" %in% any_matches) { ## if are matches then do this
+      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
+        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
+    }
+  }    
+  if(length(path_match_within_set) > 0) {
+    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
+  }
+}
+path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
+
+save(path_match_in_set, file = "path_match_in_set.Rda")
+
+### getting clustering order from pheatmap
+most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
+  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
+  rownames_to_column(var = "order")
+
+gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
+  arrange(order) %>% .$gene_name ## getting the gene name order
+
+## now need to get the gene mutation data and tumor samples 
+load("../data/gdc_tumor_coding_complete.Rdata")
+
+## Booleanize mutation count for each tumor sample
+missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
+splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+
+gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
+gc()
+gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
+
+filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
+  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
+  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
+  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
+rm(gdc_mut_count_patient)
+gc()
+
+save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
+pheatmap(filtered_mut_data[gene_order,],
+         show_rownames = T,
+         show_colnames = F,
+         treeheight_row = 0,
+         treeheight_col = 0,
+         cellheight = 5,
+         cellwidth = 1,
+         cluster_cols = T,
+         cluster_rows = F,
+         #gaps_row = file_dims,
+         #cutree_cols = 6,
+         annotation_legend = F,
+         legend_breaks = c(0, 0.25, 0.75, 1),
+         legend_labels = c("","Not Mut", "Mut", ""),
+         color = c("grey", "firebrick3"),
+         clustering_method = "ward.D2",
+         fontsize = 10,
+         legend = T,
+         filename = "fig4.png")
+
+## stopped here ## ignore bottom using as reference
+path_match_in_set
+cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
+                          show_rownames = F,
+                          show_colnames = F,
+                          cellheight = 1,
+                          cellwidth = 0.5,
+                          cluster_cols = T,
+                          cluster_rows = F,
+                          annotation_legend = F,
+                          color = c("grey", "firebrick3"),
+                          clustering_method = "single", ## ordering is single linkage
+                          legend = T)
+  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
+  cancer_path <- names(cancer_path) ## getting the sample_id original order 
+  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
+  rm(cancer_info, cancer_path)
+  gc()
+}
+dev.off()
+
+for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
+  print(n)
+  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
+  names(cancer_order) <- "sample_id"
+  cancer_order$type <- n
+  for(p in cancer_type[-1]) {
+    print(p)
+    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
+    names(df1) <- "sample_id"
+    df1$type <- p
+    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
+    rm(df1)
+    gc()
+  }
+}
+
+cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
+
+cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
+  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
+cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
+
+## set levels based on clustering algorithm above
+cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
+cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
+
+cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
+save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
+```
+
+```{r figure-4}
+
+```
+
 ![](../figures/figure4_temp.png)
@@
   left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d"), meausure.vars = tumor_suppressor$external_gene_name) %>%
+  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
   filter(value != 0)
@@
 ![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
+
+<div style="text-align: right"> Supplementary Figure S4 Data </div>
+```{r supplementary-figure-s4-data-generation, eval = FALSE}
+load("../data/gdc_reactome_path_analysis.Rda") ## load data
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
+n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
+
+toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
+
+selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
+
+selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
+  left_join(., selected_path, by = "V1") %>%
+  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
+  filter(value != 0)
+
+save(selected_path, file = "selected_path.Rda")
+```
+
+```{r supplementary-figure-s4, out.width = '90%'}
+load("../data/selected_path.Rda") ## load data
+load("../data/gdc_reactome_path_analysis.Rda")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+
+n_path <- 377 ## number of pathways we want to see
+
+toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
+needed for facet
+
+ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
+  geom_point(size = 0.5) +
+  theme_classic() +
+  coord_fixed() +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  scale_colour_manual(values = rep("firebrick3", 20)) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ factor(variable, levels = toppath),
+                      ncol = 5) ## facet by pathways
+
+```
+
+![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
+
+<div style="text-align: right"> Supplementary Figure ST1 Data </div>
+```{r supplemental-table-st1-data-generation, eval = FALSE}
+load("../data/gdc_tumor_coding_complete.Rdata")
+load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/gene2_level2_lookup.Rda")
+load("../data/pathname2level2.Rda")
+load("../data/gene_data.Rda")
+load("../data/gdc_gene_exp_adj.Rda")
+
+## Parameters ##
+n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
+mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
+show_perc <- TRUE ## show perc for mutated gene
+
+toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
+
+id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
+  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
+  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
+  dplyr::select(-ensembl_gene_id) %>%
+  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
+  left_join(., pathname2level2, by = "PATH_ID") %>%
+  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
+  dplyr::select(-PATH_ID) %>%
+  t() %>% as.data.frame()
+
+gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
+  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
+  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
+  dplyr::select(-ensembl_gene_id)
+
+table_st1 <- list() ## list for forloop
+
+for(n in toppath) {
+  message(n)
+  ## finding top mutated gene in pathway
+  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
+    'rownames<-' (rownames(id_to_names)) %>% 
+    filter(. == 1) %>% ## filter for genes that are in that pathway
+    rownames(.)
+  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
+      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
+      mutate(n_sample = length(gene_mutation),
+             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
+      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
+  top_gene_in_path <- list()
+  for(p in mutated_gene_position) {
+    message(p)
+    if(show_perc == TRUE) {
+    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
+    }
+    if(show_perc == FALSE) {
+    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
+    }
+  }
+  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
+   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
+  ## calculating the percent mutated samples
+  spec_path <- data.frame(Pathway = n,
+                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
+  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
+}
+table_st1 <- as.data.frame(do.call(rbind, table_st1))
+save(table_st1, file = "table_st1.Rda")  
+rm(list = ls(all.names = T))  
+gc()
+```
+
+```{r supplemental-table-st1}
+load("../data/table_st1.Rda")
+
+table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
+  column_spec(1, bold = T, color = "black")
+```
+
+![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
+
 ![**Supplemental figure S6.

### 2020-07-09 text/manuscript.Rmd
add second hallmarks reference

@@
 cancer as a malady affecting common cellular processes has long been a goal of
-the field and is perhaps best characterized in an essay by @weinberg2000 (see also
+the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
 Discussion). The first of these studies, (the

### 2020-07-10 text/manuscript.Rmd
added more code for fig4 - will be complete tomorrow, table 1, and table 2

--- 2020-07-09 text/manuscript.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
+#library(rstan)
 library(foreach)
@@
 save(path_match_in_set, file = "path_match_in_set.Rda")
+load("../data/path_match_in_set.Rda")
 
@@
 save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-pheatmap(filtered_mut_data[gene_order,],
+
+fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
          show_rownames = T,
@@
          cluster_cols = T,
-         cluster_rows = F,
-         #gaps_row = file_dims,
-         #cutree_cols = 6,
+         cluster_rows = F, ## turn off row clustering
          annotation_legend = F,
@@
          fontsize = 10,
-         legend = T,
-         filename = "fig4.png")
+         legend = T)
+
+## getting sample orders
+pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
+sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
+sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
 
 ## stopped here ## ignore bottom using as reference
-path_match_in_set
-cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
+
+class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
+  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
+class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
+
+## set levels based on clustering algorithm above
+class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
+class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
+
+class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
+save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
+rm(list = ls(all.names = T))
 gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
 ```
 
-```{r figure-4}
+```{r figure-4, eval = FALSE}
+load("../data/class67_gene_2_path_heatmap.Rda") ## load data
+load("../data/path_match_in_set.Rda")
+
+### getting facet ordering
+most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
+  'colnames<-' (c("path_combo", "Freq")) %>% 
+  rownames_to_column(var = "order")
+
+fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
+  geom_tile(size = 0.6) +
+  scale_fill_manual(values = c("grey", "firebrick3")) +
+  #facet_grid(factor(path_combo, levels = most_genes$path_combo) ~ ., scales = "free", space = "free") +
+  #scale_y_discrete(position = "left") +
+  labs(tag = "A") +
+  ylab("") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_text(size = 12),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        strip.text.y = element_blank(),
+        strip.background = element_blank(),
+        legend.position = "none") 
+fig4
+ggsave("fig4_ggplot_heatmap.png", height = 100, width = 17, limitsize = FALSE)
+
+## fixing the text cutoff in heatmap
+ fig1Ag <- ggplotGrob(fig1A)
+
+ for(i in which(grepl("strip-l", fig1Ag$layout$name))){
+   fig1Ag$grobs[[i]]$layout$clip <- "off"
+ }
+
+ fig1A <- as.ggplot(fig1Ag)
+ rm(fig1Ag)
 
@@
 wherein etc etc</mark>
-
-![](../figures/table1.png)
+<div style="text-align: right"> Table 1 Data </div>
+```{r table-1-data-generation, eval = FALSE}
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+load("../data/moderate_enrich_prob.Rda") ## load data
+load("../data/gene2_level2_lookup.Rda")
+load("../data/gene_data.Rda")
+load("../data/gdc_gene_exp_adj.Rda")
+load("../data/pathname2level2.Rda")
+
+id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
+  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
+  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
+  dplyr::select(-ensembl_gene_id) %>%
+  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
+  left_join(., pathname2level2, by = "PATH_ID") %>%
+  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
+  dplyr::select(-PATH_ID) %>%
+  t() %>% as.data.frame()
+
+gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
+  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
+  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
+  dplyr::select(-ensembl_gene_id)
+
+## parameters ##
+enrich_cutoff <- 0.30
+howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
+show_perc <- TRUE
+show_n <- TRUE
+mutated_gene_position <- 1
+################
+
+clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
+
+clust_spec_enrich_top5 <- list()
+for(n in unique(clust_spec_enrich$clust)) {
+  message(n)
+  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
+    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
+    dplyr::select(path, clust) %>% .[1:5,]
+}
+clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
+rm(clust_spec_enrich)
+
+## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
+type <- unique(tumor_clusters$project_code) ## type of cancer type in data
+cluster_spec_enrich_table <- list() ## list for forloop
+
+for(d in unique(clust_spec_enrich_top5$clust)) {
+  message(d)
+  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
+  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
+  top_genes_in_path_cancer_type <- list()
+  for(n in clust_toppath) { ## for each of the top paths do this
+    message(n)
+    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
+    'rownames<-' (rownames(id_to_names)) %>% 
+    filter(. == 1) %>% ## filter for genes that are in that pathway
+    rownames(.)
+    top_gene_in_path <- list()
+    for(t in type) { ## looping through cancer type within class
+      message(t)
+      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
+      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
+        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
+      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
+        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
+          mutate(n_sample = length(class_type_spec),
+               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
+        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
+      }
+      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
+        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
+        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
+        mutate(n_sample = length(class_type_spec),
+               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
+        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
+      }
+      if(show_perc == TRUE & show_n == FALSE) {
+         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
+      }
+      if(show_perc == TRUE & show_n == TRUE) {
+         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
+      }
+      if(show_perc == FALSE) {
+         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
+      }
+      }
+      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
+        top_gene_in_path[[t]] <- "NA" 
+      }
+    }
+    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
+    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
+    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
+  }
+  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
+  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
+  }
+
+cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
+save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
+rm(list = ls(all.names = T))
+gc()
+```
+
+```{r table-1}
+load("../data/cluster_spec_enrich_table1.Rda")
+cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
+  column_spec(1, bold = T)
+```
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
@@
 non-metastatic tumors of other clusters.
+<div style="text-align: right"> Table 2 Data </div>
+```{r table-2-data-generation, eval = FALSE}
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+load("../data/gdc_reactome_path_analysis.Rda")
+
+tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
+
+m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
+m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
+
+prepare_rbeta_data_p1 <- function(data, path_analysis) {
+  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
+  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
+  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
+  table(names(path_analysis) == data$V1) ## good they match now
+  ## Now we are going to build group specific cluster data
+  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
+  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
+  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
+  cluster_path_count <- list() ## Creating a list for the forloop
+  for(n in clust_group) {
+    print(n)
+    df <- subset(data, knn_clust == n) ## subset cluster number
+    df <- df$V1 ## getting tumor sample id
+    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
+    df1 <- as.data.frame(rowSums(df1))
+    names(df1) <- n
+    cluster_path_count[[n]] <- df1
+  }
+  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
+  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
+  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
+  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
+  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
+  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
+  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
+  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
+  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
+  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
+  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
+  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
+  return(rbeta_path_df)
+}
+
+m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
+## Combining cluster 6/7 cluster 910 cluster 1112
+combo_group <- c("6", "9", "11")
+compo_group_df <- list() ## list for loop
+for(n in combo_group) {
+  print(n) ## first cluster in set
+  m <- as.numeric(n) + 1 ## second cluster in set
+  m <- as.character(m)
+  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
+  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
+  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
+  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
+  p <- paste0(n,m)
+  df2 <- m0_rbeta[1:377,]
+  df2$clust <- p ## replace with new clust name
+  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
+  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
+  compo_group_df[[p]] <- df2 ## store in list and continue loop
+}
+compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
+m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
+save(m0_rbeta, file = "m0_rbeta.Rda")
+
+## repeat same function above with metastasis sample
+m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
+compo_group_df <- list() ## list for loop
+for(n in combo_group) {
+  print(n) ## first cluster in set
+  m <- as.numeric(n) + 1 ## second cluster in set
+  m <- as.character(m)
+  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
+  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
+  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
+  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
+  p <- paste0(n,m)
+  df2 <- m0_rbeta[1:377,]
+  df2$clust <- p ## replace with new clust name
+  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
+  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
+  compo_group_df[[p]] <- df2 ## store in list and continue loop
+}
+compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
+m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
+save(m1_rbeta, file = "m1_rbeta.Rda")
+
+## function to compiling data for rbeta analysis
+prepare_rbeta_data_p2 <- function(data) {
+  enrich_prob <- data.frame(path = data$path,
+                               clust = data$clust,
+                               fg_mut = data$clust_mut,
+                               fg_not_mut = data$tot_n_clust - data$clust_mut,
+                               bg_mut = data$global_mut,
+                               bg_not_mut = data$global_sample - data$global_mut)
+fg_mut_rate <- list()
+bg_mut_rate <- list()
+for(n in unique(as.character(enrich_prob$clust))) {
+  print(n)
+  df <- subset(enrich_prob, clust == n)
+  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
+  fg_numerator <- sum(df$fg_mut)
+  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
+  bg_numerator <- sum(df$bg_mut)
+  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
+  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
+}
+
+fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
+bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
+enrich_prob_df <- enrich_prob
+enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
+enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
+enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
+enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
+enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
+enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
+return(enrich_prob_df)
+}
+
+m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
+m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
+
+save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
+rm(list = ls(all.names = T)) ## cleaning up data
+gc()
+load("../data/m0_m1_rbeta.Rda")
+## Metastatic (foreground) - Nonmetastatic (background) Rbeta
+moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
+})
+moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
+})
+
+metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
+metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
+
+## ModerateEnrichment probability 
+prob <- moderate_fg > moderate_bg ## boolean data
+mean_prob <- apply(prob, MARGIN = 2, function(x) {
+  mean(x)
+})
+## Moderate Difference
+CI <- 0.025
+
+diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
+rm(moderate_fg, moderate_bg, prob)
+gc()
+quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
+}))
+
+sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
+}))
+
+m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
+                                   clust = m1_rbeta$clust,
+                                   probability = mean_prob,
+                                   signif = sig_fg_bg[,1],
+                                   difference = quant_diff[,2],
+                                   lower = quant_diff[,1],
+                                   upper = quant_diff[,3])
+save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
+rm(list = ls(all.names = T))
+gc()
+```
+
+```{r table-2}
+load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
+
+table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
+  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
+
+table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
+table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
+table2$Signif <- NULL ## we don't need this anymore
+
+table2 %>% knitr::kable(format = "html") %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
+  column_spec(1, bold = T, color = "black")
+```
 

### 2020-07-10 text/manuscript.Rmd
split text into smaller files for ease of editing

--- 2020-07-09 text/manuscript.Rmd
@@
 ## Abstract
-Cancer is a disease of cellular metabolism and other basic biochemical processes
-common to all cells. However, molecular signatures remain largely
-indistinguishable from tissue-of-origin, and general, unifying principles remain
-obscure. To address this challenge, we reduced mutation data to disruptions in a
-select set of pathways pointing to basic cell biology, from DNA replication to
-cellular communication. Using dimension reduction techniques, we assigned tumor
-samples into twelve classes distinct from their tissues-of-origin and largely
-free of bias from mutational burden or clinical stage. We showed that these
-tumor classes vary in prognosis by modeling relative risk of death in different
-classes and cancer types. We identified class-specific mutations in different
-tissues, demonstrating how tissue-specific signatures contribute to common
-cellular phenotypes. Moreover, germline risk mutations involved in genome
-stability, such as BRCA1 and Rb1, are equally distributed among our classes,
-somewhat contrary to the expectation that such genes are hallmarks of molecular
-subtypes. We also investigated pathways that distinguish between metastatic and
-non-metastatic tumors, and found that most meaningful differences are
-class-specific. Most metastatic enriched pathways from one class in metastatic
-tumors are class-specific pathways from another class, suggesting phenotypic
-convergence, and potentially highlighting key pathways to focus on
-therapeutically for prevention of class-specific relapse. In summary, our
-observations suggest that some genes act as drivers of genome instability in all
-classes, while different sets of class-specific genes confer increased fitness
-in a combinatorial fashion, eventually converging on one or more end-stage
-disease phenotypes. Thus, we present a coherent picture of global tumor biology,
-and explain how common cellular dysfunction might arise from tissue-specific
-mutations.
+
+```{r child='abstract.Rmd'}
+```
 
@@
 
-Recent advances in technology have made the study of cancer possible at
-unprecedented levels of detail and scale. Array and next-generation sequencing
-technologies have made it possible to assay germline and somatic mutations,
-expression profiles, DNA methylation, and copy number variations. To take
-advantage of these tools several large consortia, including the cancer genome
-atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
-numbers of tumors and collected data from multiple assays that could be analyzed
-together, with the goal of integration and increased understanding of the
-mechanisms of cancer. These efforts have focused on recognizing and quantifying
-the importance of genes in cancer etiology as the result of selection processes,
-or classification of cancers relative to common patterns of aberration in
-multiple assays.
-
-Considerable progress has been made analyzing these data. Statistical analyses
-identify hundreds of global and tissue-specific cancer driver genes
-[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
-@lawrenceDiscoverySaturationAnalysis2014;
-@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
-@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
-when genes are mutated at a greater rate than expected due to chance. Based in
-part on studies like these, it has been estimated that on the order of three
-mutations in key oncogenes and/or tumor suppressors would be sufficient to
-transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
-
-Other studies aim at providing a comprehensive view of mutations, gene
-expression and genomic signatures, with the express goal of understanding the
-common themes of all cancer independent of tissue of origin. Such a view of
-cancer as a malady affecting common cellular processes has long been a goal of
-the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
-Discussion). The first of these studies, (the
-“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
-using integrative analysis with co-equal weighting of gene expression,
-methylation, copy number and proteomics data
-[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
-tissue-of-origin is the predominant driving factor, though a significant
-proportion of tumors ($10\%$) could be reclassified independent of
-tissue-of-origin. In a second study involving 33 cancer types and a much greater
-number of tumors using a newer algorithm (iCluster)
-[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
-could be further subdivided into organ specific groups, including
-pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
-squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
-
-More recently, attempts have been made to consider pan-cancer mutations and
-other data outside the context of enrichment at the gene level. Horn _et al._
-[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
-sparseness of mutations, allowing the potential importance of a gene to be
-influenced by the aggregate of positive selection in its nearest network
-neighbors. This method is powerful especially when applied to a small number (on
-the order of hundreds) of tumors. However, the larger problem of cancer etiology
-remains: what mutations are required, in what order, and are there specific
-combinations of delinquent cellular processes sufficient to cause cancer? If
-there is a common logic to cancer, how is it to be detected underneath the
-overwhelming cell-of-origin signal?
-
-In this study, we attempt to address these questions by focusing on discrete
-molecular pathways as the unit of disruption in cancer. Our approach strips all
-cell-type-specific information from the mutation data and generalizing
-gene-level information to non-tissue specific cell-biological pathway
-phenotypes. We use this rubric to evaluate all cancers and show that,
-surprisingly, tumors may have distinct phenotypes completely independent of
-their tissues of origin.
+```{r child='introduction.Rmd'}
+```
 
@@
 
-### Classification of tumors independent of tissue-of-origin
-
-The basic problem in tumor research of this kind, namely that the search space,
-~20,000 protein-coding genes is greater than the number of tumor samples
-[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
-that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
-elements exist with oncogenic or tumor suppressor potential. Hence, we sought to
-investigate cancer as a disease of basic cellular and biochemical pathways,
-which we accomplished by translating gene-level mutations into pathway level
-disruptions. Importantly, we selectively removed pathways that lacked generality
-as cellular processes or that could be specifically assigned to differentiated
-cell types, while excluding others.
-
-We hypothesize that cancer is the result of dysfunction from a limited number of
-basic cellular processes that are common to all living human cells, and thus all
-tumors belong on a spectrum of dysfunction on this axis regardless of tissue of
-origin. An alternative hypothesis is that every tumor belongs to one of a large
-number of syndromes which are unique to each tissue of origin, that share some
-mechanisms and superficial similarity in treatment strategies. The latter idea
-has gained momentum in recent years since the publication of Encode, Roadmap and
-TCGA consortium papers, perhaps owing to the intractable complexity of genomic
-data in all its forms. The inescapable conclusion thus far is that all these
-data point to tissue-of-origin as the most important driver of tumor behavior.
-Our observations contrast with this view, and instead support an interpretation
-in which all tumors manifest one of a limited number of metastable phenotypes
-represented by our classes (**figure 1a-k**), with increased fitness regardless
-of tissue of origin, which then converge on end stage cancer by accumulating
-complementary mutations to the other classes (**figure 1l** and **Table 2**).
-
-We find that most tumors segregate into one of 12 biologically distinct
-clusters, regardless of the cancer type (**figure 1h** and **supplementary
-figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
-arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
-see **supplementary figure S1**) are unevenly distributed toward a specific
-class, we could not identify cancer type that was exclusive to a single tumor
-class. Somewhat surprisingly to us, this finding extends to histological
-subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
-and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
-observe bias for some cancers, as it has long been known that certain treatment
-regimens are likely to work for a majority of patients of a given cancer type.
-Nonetheless, the basis for some tumors being treatment-refractory in spite of
-clinical indications remains elusive. Our analysis suggests the possibility of
-more general mechanisms underlying tumor evolution regardless of tissue,
-histological or cellular subtype and could form a basis for future separation of
-patients into different precision treatment regimens.
-
-It could be that the classes we identified are consistent with a signature of
-overall disease progression, such that successive, adjacent classes on the UMAP
-projection exhibit increasing specificity. We identified no such trend (see
-**figure 1k**) and each class instead was associated with unique combinations of
-pathways (as in **figure 3**). In our view, these observations suggest that
-cancers have available to them a limited number of stable configurations with
-respect to pathway disruption which are consistent with increased fitness and
-therefore positive selection.
-
-Finally, we must remark on two of the classes, 1 and 8, which for which we did
-not find distinctive associations with pathways. Class 1 has a relatively low
-level of mutated pathways overall, although it is broadly enriched in many of
-the same tumor-promoting pathways common to the other groups. This likely does
-not reflect a true difference in stages of tumor evolution, as our data show
-clearly that this class is as likely to contain stage IV as it is to contain
-stage I tumors. We speculate this class represents a group of tumors with large
-aberrations in methylation, copy number or other genome rearrangements which we
-excluded from the present analysis. In the future, we hope to incorporate these
-other data into a pathway-centric analysis as we have done here for point
-mutations and indels. Class 8 represents a group of hyper-mutated tumors that
-harbor so many mutations that virtually no pathway is unaffected. This group
-highlights one of the limitations of the approach that we’ve taken, in that it
-is likely that a significant fraction of the samples flagged as “mutant” for
-each pathway are actually unaffected, consistent with those samples have low
-information content. Similar to class 1 we speculate that a future analysis of
-other data types, including improved filtering of mutations, may resolve these
-tumors into more meaningful categories.
-
-### Incompleteness of the pathway disruption data
-
-The fact that we are likely excluding many defects (under chromosomal
-abberations, amplifications, demethylation and so on) also raises the
-possibility that our classes are incomplete or may change at a future time when
-we can account for them. For now at least, we think that the point mutation data
-are representative of the most important pathways.
-
-### Estimates of survival reveal pathway-dependent differences
-
-By modeling CDC longevity data as a baseline risk function we showed that each
-of our pathway disruption classes exhibit cancer-type specific effects on
-expected survival (**figure 5**). Contrary to what one might expect at the
-outset, we found that models assuming different survival based on class _and_
-cancer type fit better than models in which these variables are independent
-factors. However, considering that within each cancer type there are different
-clinical standards of care, and even within classes of drugs the preferred
-treatment can very between cancers, it makes sense in hindsight that they should
-all be different. Contrast the situation with ovarian vs breast cancer, which
-are both hormonally driven cancers, for example. Ovarian cancer has but one main
-treatment, platinum, whereas breast cancer patients have a variety of treatment
-regimens based on molecular subtype and other factors. It is easy to see how
-standard of care for one cancer and class could produce different outcomes than
-the same class of a different cancer type under the circumstances.
-Unfortunately, given the diversity drug classes and treatments (including
-non-chemical treatments such as surgery and palliative care), we lack sufficient
-power in the current study to explore these variables in the GDC data. It is our
-hope that more focused future studies will help to distinguish between
-treatment-specific effects or survival given different pathway disruption
-classes.
-
-### Implications for the evolution of cancer
-
-Our findings imply that there are separate processes in the etiology of cancer
-that can be broadly thought of as general cancer promoting, class-specific
-mutations and metastasis. General cancer promoting processes must include
-factors that relate to genome stability either directly or indirectly. Such
-pathways are disrupted in all or most of the defined classes (see **figures 2**
-and **supplementary figure S1**) and often are the result of abberations
-involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic
-checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and
-**supplemental file 1**). Class-specific evolution must involve the acquisition
-of combinatorial disruptions to pathways that may individually be harmful but
-together produce more specialized cancer phenotypes and increased fitness (as in
-**figure 4** and **tables 1** and **ST2**). Importantly, our work does not
-distinguish between the order in which these mutations should accumulate.
-However, since many of the genes in the non-specific pathways involve genome
-stability, it is reasonable to hypothesize that such mutations promote the
-acquisition of class specific defects, and natural selection produces the
-classes we observed.
-
-### On metastasis as a convergence of phenotypes
-
-The class-specific metastatic pathway enrichment we observe (in **table 2**) may
-be a result of the pathways directly promoting metastatic phenotypes, or
-alternatively that such mutations are more likely to turn metastatic because
-they acquire a stronger or more complete oncogenic phenotype, since (again) we
-don’t know the order in which they were acquired. The fact that most metastatic
-enrichment is cluster-specific and has a tendency to overlap with
-cluster-specific pathways from non-metastatic tumors of neighboring clusters
-favors the latter explanation, and suggests a mechanism of complementarity in
-which newly acquired mutations converge on a single deadly phenotype that
-captures most or all of the critical features of end-stage cancer. If this is
-the case, we speculate that once we discover the vulnerabilities of each cluster
-we might also reasonably predict which pathways are susceptible to further
-selection and therefore target them  therapeutically at the same time to prevent
-further adaptation.
+```{r child='discussion.Rmd'}
+```
 

### 2020-07-10 text/abstract.Rmd

Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways pointing to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into twelve classes distinct from their tissues-of-origin and largely
free of bias from mutational burden or clinical stage. We showed that these
tumor classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating how tissue-specific signatures contribute to common
cellular phenotypes. Moreover, germline risk mutations involved in genome
stability, such as BRCA1 and Rb1, are equally distributed among our classes,
somewhat contrary to the expectation that such genes are hallmarks of molecular
subtypes. We also investigated pathways that distinguish between metastatic and
non-metastatic tumors, and found that most meaningful differences are
class-specific. Most metastatic enriched pathways from one class in metastatic
tumors are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on
therapeutically for prevention of class-specific relapse. In summary, our
observations suggest that some genes act as drivers of genome instability in all
classes, while different sets of class-specific genes confer increased fitness
in a combinatorial fashion, eventually converging on one or more end-stage
disease phenotypes. Thus, we present a coherent picture of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

### 2020-07-10 text/manuscript.Rmd

--- 2020-07-10 text/manuscript.Rmd
@@
 
-Recent advances in technology have made the study of cancer possible at
-unprecedented levels of detail and scale. Array and next-generation sequencing
-technologies have made it possible to assay germline and somatic mutations,
-expression profiles, DNA methylation, and copy number variations. To take
-advantage of these tools several large consortia, including the cancer genome
-atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
-numbers of tumors and collected data from multiple assays that could be analyzed
-together, with the goal of integration and increased understanding of the
-mechanisms of cancer. These efforts have focused on recognizing and quantifying
-the importance of genes in cancer etiology as the result of selection processes,
-or classification of cancers relative to common patterns of aberration in
-multiple assays.
-
-Considerable progress has been made analyzing these data. Statistical analyses
-identify hundreds of global and tissue-specific cancer driver genes
-[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
-@lawrenceDiscoverySaturationAnalysis2014;
-@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
-@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
-when genes are mutated at a greater rate than expected due to chance. Based in
-part on studies like these, it has been estimated that on the order of three
-mutations in key oncogenes and/or tumor suppressors would be sufficient to
-transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
-
-Other studies aim at providing a comprehensive view of mutations, gene
-expression and genomic signatures, with the express goal of understanding the
-common themes of all cancer independent of tissue of origin. Such a view of
-cancer as a malady affecting common cellular processes has long been a goal of
-the field and is perhaps best characterized in an essay by @weinberg2000 (see also
-Discussion). The first of these studies, (the
-“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
-using integrative analysis with co-equal weighting of gene expression,
-methylation, copy number and proteomics data
-[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
-tissue-of-origin is the predominant driving factor, though a significant
-proportion of tumors ($10\%$) could be reclassified independent of
-tissue-of-origin. In a second study involving 33 cancer types and a much greater
-number of tumors using a newer algorithm (iCluster)
-[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
-could be further subdivided into organ specific groups, including
-pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
-squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
-
-More recently, attempts have been made to consider pan-cancer mutations and
-other data outside the context of enrichment at the gene level. Horn _et al._
-[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
-sparseness of mutations, allowing the potential importance of a gene to be
-influenced by the aggregate of positive selection in its nearest network
-neighbors. This method is powerful especially when applied to a small number (on
-the order of hundreds) of tumors. However, the larger problem of cancer etiology
-remains: what mutations are required, in what order, and are there specific
-combinations of delinquent cellular processes sufficient to cause cancer? If
-there is a common logic to cancer, how is it to be detected underneath the
-overwhelming cell-of-origin signal?
-
-In this study, we attempt to address these questions by focusing on discrete
-molecular pathways as the unit of disruption in cancer. Our approach strips all
-cell-type-specific information from the mutation data and generalizing
-gene-level information to non-tissue specific cell-biological pathway
-phenotypes. We use this rubric to evaluate all cancers and show that,
-surprisingly, tumors may have distinct phenotypes completely independent of
-their tissues of origin.
+```{r child='introduction.Rmd'}
+```
 

### 2020-07-10 text/manuscript.Rmd
updated fig4 annotation

--- 2020-07-10 text/manuscript.Rmd
@@
 library(UpSetR)
+library(cowplot)
 ```
@@
 cancer as a malady affecting common cellular processes has long been a goal of
-the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
+the field and is perhaps best characterized in an essay by @weinberg2000 (see also
 Discussion). The first of these studies, (the
@@
 load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/moderate_enrich_prob.Rda")
 load("../data/gene_data.Rda")
@@
 n_path <- 10 ## looking at the top10 path
+enrich_cutoff <- 0.30
+################
 
@@
 
-toppath <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top10 mutated pathways
+toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
+  filter(clust %in% filtered_class) %>% ## filter the cluster
+  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
+  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
 
@@
 
-```{r figure-4, eval = FALSE}
+```{r figure-4, message = FALSE}
 load("../data/class67_gene_2_path_heatmap.Rda") ## load data
@@
   scale_fill_manual(values = c("grey", "firebrick3")) +
-  #facet_grid(factor(path_combo, levels = most_genes$path_combo) ~ ., scales = "free", space = "free") +
-  #scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
   theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
+        axis.text.y = element_blank(),
         axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
+        axis.title.y = element_blank(),
         axis.ticks = element_blank(),
@@
         legend.position = "none")
-fig4
-ggsave("fig4_ggplot_heatmap.png", height = 100, width = 17, limitsize = FALSE)
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
+
+load("../data/path_match_in_set.Rda")
+
+gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
+  arrange(order) %>% .$gene_name ## getting the gene name order
+
+## right annotation data generation
+proteasome_r_anno <- data.frame(gene = gene_order, 
+                                psm_gene = 0)
+
+proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
+
+r_anno_color <- c("0" = "white", "1" = "black")
+r_anno <- ggplot(proteasome_r_anno) +
+  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
+           stat = "identity", 
+           width = 1) +
+  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
+  theme_void() 
+legend <- plot_grid(get_legend(r_anno), ncol = 1)
+r_anno <- r_anno + theme(legend.position = "none")
+
+## left annotatoin data generation
+l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
+
+## need to get top path names again
+load("../data/moderate_enrich_prob.Rda")
+enrich_cutoff <- 0.30
+filtered_class <- c(6,7)
+n_path <- 10
+toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
+  filter(clust %in% filtered_class) %>% ## filter the cluster
+  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
+  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
+
+for(p in toppath){
+  message(p)
+  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
+  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
+  ## for any path_combo that includes the pathway name change the 0 into a 1
+  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
+    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
+  }
+  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
+  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
+  }
+}
+l_anno_ggplot <- list() ## list for storing ggplots
+path_legend <- list()
+for(n in 1:length(toppath)){
+  message(n)
+  p <- toppath[n]
+  message(p)
+  spec_path <- l_anno[,c("gene_name", p)]
+  names(spec_path)[names(spec_path) == p] <- "p"
+  path_ggplot <- ggplot(spec_path) +
+  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
+           stat = "identity", 
+           width = 1) +
+  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
+  theme_void() 
+  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
+  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
+}
+
+plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
+plot
 
@@
 ## Supplementary Information
+
+### Supplementary Figure S1
+
 ```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
@@
 
+### Supplementary Figure S2
+
 ```{r supplementary-figure-s2-data-generation}
@@
 ![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
+
+### Supplementary Figure S3
+
 <div style="text-align: right"> Supplementary Figure S3 Data </div>
@@
 
+### Supplementary Figure S4
+
 <div style="text-align: right"> Supplementary Figure S4 Data </div>
@@
 ![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
+
+### Supplementary Table ST1
 

### 2020-07-10 text/discussion.Rmd

--- 2020-07-10 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
+### Classification of tumors independent of tissue-of-origin
 
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
+The basic problem in tumor research of this kind, namely that the search space,
+~20,000 protein-coding genes is greater than the number of tumor samples
+[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
+that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
+elements exist with oncogenic or tumor suppressor potential. Hence, we sought to
+investigate cancer as a disease of basic cellular and biochemical pathways,
+which we accomplished by translating gene-level mutations into pathway level
+disruptions. Importantly, we selectively removed pathways that lacked generality
+as cellular processes or that could be specifically assigned to differentiated
+cell types, while excluding others.
 
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
+We hypothesize that cancer is the result of dysfunction from a limited number of
+basic cellular processes that are common to all living human cells, and thus all
+tumors belong on a spectrum of dysfunction on this axis regardless of tissue of
+origin. An alternative hypothesis is that every tumor belongs to one of a large
+number of syndromes which are unique to each tissue of origin, that share some
+mechanisms and superficial similarity in treatment strategies. The latter idea
+has gained momentum in recent years since the publication of Encode, Roadmap and
+TCGA consortium papers, perhaps owing to the intractable complexity of genomic
+data in all its forms. The inescapable conclusion thus far is that all these
+data point to tissue-of-origin as the most important driver of tumor behavior.
+Our observations contrast with this view, and instead support an interpretation
+in which all tumors manifest one of a limited number of metastable phenotypes
+represented by our classes (**figure 1a-k**), with increased fitness regardless
+of tissue of origin, which then converge on end stage cancer by accumulating
+complementary mutations to the other classes (**figure 1l** and **Table 2**).
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-```
+We find that most tumors segregate into one of 12 biologically distinct
+clusters, regardless of the cancer type (**figure 1h** and **supplementary
+figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
+arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
+see **supplementary figure S1**) are unevenly distributed toward a specific
+class, we could not identify cancer type that was exclusive to a single tumor
+class. Somewhat surprisingly to us, this finding extends to histological
+subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
+and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
+observe bias for some cancers, as it has long been known that certain treatment
+regimens are likely to work for a majority of patients of a given cancer type.
+Nonetheless, the basis for some tumors being treatment-refractory in spite of
+clinical indications remains elusive. Our analysis suggests the possibility of
+more general mechanisms underlying tumor evolution regardless of tissue,
+histological or cellular subtype and could form a basis for future separation of
+patients into different precision treatment regimens.
 
-## Abstract
+It could be that the classes we identified are consistent with a signature of
+overall disease progression, such that successive, adjacent classes on the UMAP
+projection exhibit increasing specificity. We identified no such trend (see
+**figure 1k**) and each class instead was associated with unique combinations of
+pathways (as in **figure 3**). In our view, these observations suggest that
+cancers have available to them a limited number of stable configurations with
+respect to pathway disruption which are consistent with increased fitness and
+therefore positive selection.
 
-```{r child='abstract.Rmd'}
-```
+Finally, we must remark on two of the classes, 1 and 8, which for which we did
+not find distinctive associations with pathways. Class 1 has a relatively low
+level of mutated pathways overall, although it is broadly enriched in many of
+the same tumor-promoting pathways common to the other groups. This likely does
+not reflect a true difference in stages of tumor evolution, as our data show
+clearly that this class is as likely to contain stage IV as it is to contain
+stage I tumors. We speculate this class represents a group of tumors with large
+aberrations in methylation, copy number or other genome rearrangements which we
+excluded from the present analysis. In the future, we hope to incorporate these
+other data into a pathway-centric analysis as we have done here for point
+mutations and indels. Class 8 represents a group of hyper-mutated tumors that
+harbor so many mutations that virtually no pathway is unaffected. This group
+highlights one of the limitations of the approach that we’ve taken, in that it
+is likely that a significant fraction of the samples flagged as “mutant” for
+each pathway are actually unaffected, consistent with those samples have low
+information content. Similar to class 1 we speculate that a future analysis of
+other data types, including improved filtering of mutations, may resolve these
+tumors into more meaningful categories.
 
-## Introduction
+### Incompleteness of the pathway disruption data
 
-```{r child='introduction.Rmd'}
-```
+The fact that we are likely excluding many defects (under chromosomal
+abberations, amplifications, demethylation and so on) also raises the
+possibility that our classes are incomplete or may change at a future time when
+we can account for them. For now at least, we think that the point mutation data
+are representative of the most important pathways.
 
-## Results
+### Estimates of survival reveal pathway-dependent differences
 
-### A taxonomy of tumors based on disrupted molecular pathways
+By modeling CDC longevity data as a baseline risk function we showed that each
+of our pathway disruption classes exhibit cancer-type specific effects on
+expected survival (**figure 5**). Contrary to what one might expect at the
+outset, we found that models assuming different survival based on class _and_
+cancer type fit better than models in which these variables are independent
+factors. However, considering that within each cancer type there are different
+clinical standards of care, and even within classes of drugs the preferred
+treatment can very between cancers, it makes sense in hindsight that they should
+all be different. Contrast the situation with ovarian vs breast cancer, which
+are both hormonally driven cancers, for example. Ovarian cancer has but one main
+treatment, platinum, whereas breast cancer patients have a variety of treatment
+regimens based on molecular subtype and other factors. It is easy to see how
+standard of care for one cancer and class could produce different outcomes than
+the same class of a different cancer type under the circumstances.
+Unfortunately, given the diversity drug classes and treatments (including
+non-chemical treatments such as surgery and palliative care), we lack sufficient
+power in the current study to explore these variables in the GDC data. It is our
+hope that more focused future studies will help to distinguish between
+treatment-specific effects or survival given different pathway disruption
+classes.
 
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
+### Implications for the evolution of cancer
 
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
+Our findings imply that there are separate processes in the etiology of cancer
+that can be broadly thought of as general cancer promoting, class-specific
+mutations and metastasis. General cancer promoting processes must include
+factors that relate to genome stability either directly or indirectly. Such
+pathways are disrupted in all or most of the defined classes (see **figures 2**
+and **supplementary figure S1**) and often are the result of abberations
+involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic
+checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and
+**supplemental file 1**). Class-specific evolution must involve the acquisition
+of combinatorial disruptions to pathways that may individually be harmful but
+together produce more specialized cancer phenotypes and increased fitness (as in
+**figure 4** and **tables 1** and **ST2**). Importantly, our work does not
+distinguish between the order in which these mutations should accumulate.
+However, since many of the genes in the non-specific pathways involve genome
+stability, it is reasonable to hypothesize that such mutations promote the
+acquisition of class specific defects, and natural selection produces the
+classes we observed.
 
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
+### On metastasis as a convergence of phenotypes
 
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-there is a whole set of genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways, and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top10 mutated pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F,
-         #gaps_row = file_dims,
-         #cutree_cols = 6,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T,
-         filename = "fig4.png")
-
-## stopped here ## ignore bottom using as reference
-path_match_in_set
-cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
-```
-
-```{r figure-4}
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-
-![](../figures/table1.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References
+The class-specific metastatic pathway enrichment we observe (in **table 2**) may
+be a result of the pathways directly promoting metastatic phenotypes, or
+alternatively that such mutations are more likely to turn metastatic because
+they acquire a stronger or more complete oncogenic phenotype, since (again) we
+don’t know the order in which they were acquired. The fact that most metastatic
+enrichment is cluster-specific and has a tendency to overlap with
+cluster-specific pathways from non-metastatic tumors of neighboring clusters
+favors the latter explanation, and suggests a mechanism of complementarity in
+which newly acquired mutations converge on a single deadly phenotype that
+captures most or all of the critical features of end-stage cancer. If this is
+the case, we speculate that once we discover the vulnerabilities of each cluster
+we might also reasonably predict which pathways are susceptible to further
+selection and therefore target them  therapeutically at the same time to prevent
+further adaptation.

### 2020-07-10 text/introduction.Rmd

--- 2020-07-10 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
+Recent advances in technology have made the study of cancer possible at
+unprecedented levels of detail and scale. Array and next-generation sequencing
+technologies have made it possible to assay germline and somatic mutations,
+expression profiles, DNA methylation, and copy number variations. To take
+advantage of these tools several large consortia, including the cancer genome
+atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
+numbers of tumors and collected data from multiple assays that could be analyzed
+together, with the goal of integration and increased understanding of the
+mechanisms of cancer. These efforts have focused on recognizing and quantifying
+the importance of genes in cancer etiology as the result of selection processes,
+or classification of cancers relative to common patterns of aberration in
+multiple assays.
 
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
+Considerable progress has been made analyzing these data. Statistical analyses
+identify hundreds of global and tissue-specific cancer driver genes
+[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
+@lawrenceDiscoverySaturationAnalysis2014;
+@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
+@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
+when genes are mutated at a greater rate than expected due to chance. Based in
+part on studies like these, it has been estimated that on the order of three
+mutations in key oncogenes and/or tumor suppressors would be sufficient to
+transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
 
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
+Other studies aim at providing a comprehensive view of mutations, gene
+expression and genomic signatures, with the express goal of understanding the
+common themes of all cancer independent of tissue of origin. Such a view of
+cancer as a malady affecting common cellular processes has long been a goal of
+the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
+Discussion). The first of these studies, (the
+“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
+using integrative analysis with co-equal weighting of gene expression,
+methylation, copy number and proteomics data
+[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
+tissue-of-origin is the predominant driving factor, though a significant
+proportion of tumors ($10\%$) could be reclassified independent of
+tissue-of-origin. In a second study involving 33 cancer types and a much greater
+number of tumors using a newer algorithm (iCluster)
+[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
+could be further subdivided into organ specific groups, including
+pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
+squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-```
+More recently, attempts have been made to consider pan-cancer mutations and
+other data outside the context of enrichment at the gene level. Horn _et al._
+[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
+sparseness of mutations, allowing the potential importance of a gene to be
+influenced by the aggregate of positive selection in its nearest network
+neighbors. This method is powerful especially when applied to a small number (on
+the order of hundreds) of tumors. However, the larger problem of cancer etiology
+remains: what mutations are required, in what order, and are there specific
+combinations of delinquent cellular processes sufficient to cause cancer? If
+there is a common logic to cancer, how is it to be detected underneath the
+overwhelming cell-of-origin signal?
 
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-there is a whole set of genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways, and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top10 mutated pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F,
-         #gaps_row = file_dims,
-         #cutree_cols = 6,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T,
-         filename = "fig4.png")
-
-## stopped here ## ignore bottom using as reference
-path_match_in_set
-cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
-```
-
-```{r figure-4}
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-
-![](../figures/table1.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References
+In this study, we attempt to address these questions by focusing on discrete
+molecular pathways as the unit of disruption in cancer. Our approach strips all
+cell-type-specific information from the mutation data and generalizing
+gene-level information to non-tissue specific cell-biological pathway
+phenotypes. We use this rubric to evaluate all cancers and show that,
+surprisingly, tumors may have distinct phenotypes completely independent of
+their tissues of origin.

### 2020-07-14 text/manuscript.Rmd

@@
 knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
+library(rstan)
 library(foreach)
@@
 
+```{r mutation-burden-plot, message = FALSE}
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
+mut_order <- list() ## store data in list 
+for(n in 1:max(umap_cancertype$knn_clust)) { ## loop through all the clusters
+  message(n)
+  mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
+    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
+    arrange(mut_count) ## arrange mutation in increasing order
+}
+mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
+
+mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
+mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
+
+## Need the breaks positions for the x-axis
+previous_class <- 0 ## starting position 1
+position_of_previous_class <- 0 ## starting position 2
+xaxis_breaks <- list() ## list for for loop
+for(n in 1) { ## loop through 1 to 12 clusters
+  message(n)
+  class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
+  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+  previous_class <- class_spec ## storing class halfway mark
+  for(n in 2:max(as.numeric(as.character(mut_order$knn_clust)))) {
+    message(n)
+    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
+    class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
+    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+    previous_class <- class_spec ## storing class halfway mark
+  }
+}
+xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
+
+total_n <- as.data.frame(table(mut_order$knn_clust))
+
+## for making dup discrete x-axis
+guide_axis_label_trans <- function(label_trans = identity, ...) {
+  axis_guide <- guide_axis(...)
+  axis_guide$label_trans <- rlang::as_function(label_trans)
+  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
+  axis_guide
+}
+
+guide_train.guide_axis_trans <- function(x, ...) {
+  trained <- NextMethod()
+  trained$key$.label <- x$label_trans(trained$key$.label)
+  trained
+}
+
+ggplot(mut_order, aes(x = V1, y = log10(mut_count + 1), color = knn_clust)) + 
+  geom_point() + 
+  ylab("Log10(Mutation Count)") + 
+  xlab("") +
+  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("Class =" ,1:12)) +
+  guides(x.sec = guide_axis_label_trans(~paste(c("n = ", total_n$Freq)))) + ## create second x-axis
+  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
+  #scale_y_log10() +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12),
+        axis.text.y = element_text(size = 12),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        legend.position = "none") +
+  scale_color_manual(values = kelly.colours[-c(1:2)])
+```
+
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
@@
 
+gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
+  arrange(order) %>% .$gene_name ## getting the gene name order
+
+class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
+
 fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
@@
 }
+
 l_anno_ggplot <- list() ## list for storing ggplots
 path_legend <- list()
+
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
 for(n in 1:length(toppath)){
@@
   names(spec_path)[names(spec_path) == p] <- "p"
+  anno_color <- c("0" = "white", "1" = kelly.colours[2+n])
   path_ggplot <- ggplot(spec_path) +
@@
            width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
+  scale_fill_manual(values = anno_color, breaks = "1", name = "", labels = p) +
   theme_void() 
@@
 plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
+plot_legend <- plot_grid(path_legend[[1]], path_legend[[2]], path_legend[[3]], path_legend[[4]], path_legend[[5]], path_legend[[6]], path_legend[[7]], path_legend[[8]], path_legend[[9]], path_legend[[10]], ncol = 5)
+
+plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5))
 

### 2020-07-14 text/manuscript.Rmd
extend proteasome text (related to figure 4)

--- 2020-07-11 text/manuscript.Rmd
@@
 predominated by 1 or more highly mutated genes across all cancers. Instead,
-there is a whole set of genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways, and the majority of tumors in
+genes in the proteosomal pathway whose activity
+contributes to all the aforementioned pathways and the majority of tumors in
 these classes have a single mutation in one of these genes. Nonetheless, a
@@
 The highlighted genes at right are all proteasomal subunits.</p>
+
+The proteasome plays key roles in the transduction of signals from the p53, WNT
+and NFkB pathways and it and proteasomal inhibitors have been approved for
+clinical use in some patient populations [@proteasome_review_Jang]. To determine
+whether this class of proteins is broadly sensitive to cancer-promoting
+mutations, we tabulated each gene as prognostic or not in the human protein
+expression atlas [@protein_atlas] and found 49 positive and 10 negative
+associations ( _i.e._ higher expression associated with unfavorable outcomes is
+"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
+Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
+majority (x out of y) are missense _vs._ various other deletarious classes (see
+**figure 4d**), suggesting that, similar to other oncogenic driver classes such
+as RAS and PI3K, proteasomal subunit mutations may result in specific
+dysregulation of the proteasomal complex activities that target cancer-promoting
+pathways. Taken together, our data suggest that mutations in proteasomal
+subunits are more widespread in cancer than previously appreciated and may
+represent a targetable category of tumors.
 

### 2020-07-14 text/data_set_generation.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
 <div style="text-align: right"> Generation of Somatic Mutation Data </div>
@@
 ```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_1.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
 <div style="text-align: right"> Figure 1 data </div>
@@
 
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
+```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5, message = FALSE}
 load("../data/cancer_disrupt_heatmap.Rda")
@@
 
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-```{r mutation-burden-plot, message = FALSE}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
+## mutation burden plot
 mut_order <- list() ## store data in list 
-for(n in 1:max(umap_cancertype$knn_clust)) { ## loop through all the clusters
+for(n in 1:max(as.numeric(as.character(umap_cancertype$knn_clust)))) { ## loop through all the clusters
   message(n)
@@
 
-total_n <- as.data.frame(table(mut_order$knn_clust))
-
-## for making dup discrete x-axis
-guide_axis_label_trans <- function(label_trans = identity, ...) {
-  axis_guide <- guide_axis(...)
-  axis_guide$label_trans <- rlang::as_function(label_trans)
-  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
-  axis_guide
-}
-
-guide_train.guide_axis_trans <- function(x, ...) {
-  trained <- NextMethod()
-  trained$key$.label <- x$label_trans(trained$key$.label)
-  trained
-}
-
-ggplot(mut_order, aes(x = V1, y = log10(mut_count + 1), color = knn_clust)) + 
+total_n <- as.data.frame(table(mut_order$knn_clust)) ## getting total n in cluster
+
+mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count
+
+mut_median <- mut_order %>% 
+  group_by(knn_clust) %>% 
+  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
+
+fig1G <- ggplot(mut_order, aes(x = V1, y = log10mut, color = knn_clust)) + 
   geom_point() + 
@@
   xlab("") +
-  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("Class =" ,1:12)) +
-  guides(x.sec = guide_axis_label_trans(~paste(c("n = ", total_n$Freq)))) + ## create second x-axis
+  theme_classic() +
+  facet_wrap(~ factor(knn_clust, levels = c(1:12)), ncol = 12, scales = "free_x") +
+  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
   scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  #scale_y_log10() +
-  theme_classic() +
+  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
   theme(axis.text.x = element_text(size = 12),
@@
         axis.title.y = element_text(size = 12),
+        axis.ticks.x = element_blank(),
+        axis.line.x = element_line(colour = "black", size = 0.5),
+        axis.line.x.top = element_line(colour = "black", size = 0.5),
+        #strip.background.x = element_blank(),
+        strip.placement = "outside",
+        strip.text.x = element_text(size = 12),
         legend.position = "none") +
   scale_color_manual(values = kelly.colours[-c(1:2)])
+
+## code to keep only the bottom border line of the facet
+fig1G <- ggplotGrob(fig1G)
+lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
+                   gp=gpar(col="black", lwd=4))
+
+for (k in grep("strip-t",fig1G$layout$name)) {
+  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
+}
+
+fig1G <- as.ggplot(fig1G)
+
+## umap by cancer type
+fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "H", color = "") +
+  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
+                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap of brca subtype
+fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() +
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "I", color = "") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap brca somatic mutations
+load("../data/gdc_gene_exp_adj.Rda")
+load("../data/gene_data.Rda")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
+brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
+  rownames_to_column(var = "ensembl_gene_id") %>%
+  left_join(., brca, by = "ensembl_gene_id") %>% 
+  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
+  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
+  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
+  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
+  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
+                           BRCA1 == "0" ~ "NA"),
+         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
+                           BRCA2 == "0" ~ "NA")) %>%
+  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
+  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
+  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
+
+fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
+  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(color = "", tag = "J") +
+  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
+                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap stage
+fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "K", color = "") +
+  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
+                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap metastasis
+fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "L", color = "") +
+  scale_color_manual(labels = c("M0", "M1"), 
+                     values = c('#a6cee3','#1f78b4')) +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+layout <- '
+AAAABBCCDD
+AAAAEEEFFF
+AAAAGGGGGG
+AAAAGGGGGG
+AAAAHHIIJJ
+AAAAKKLL##
+'
+
+fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
+
 ```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-there is a whole set of genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways, and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  anno_color <- c("0" = "white", "1" = kelly.colours[2+n])
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot_legend <- plot_grid(path_legend[[1]], path_legend[[2]], path_legend[[3]], path_legend[[4]], path_legend[[5]], path_legend[[6]], path_legend[[7]], path_legend[[8]], path_legend[[9]], path_legend[[10]], ncol = 5)
-
-plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5))
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_2.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
 <div style="text-align: right"> Figure 2 Data </div>
@@
 ```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_3.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
 <div style="text-align: right"> Figure 3 data </div>
@@
 ```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_4.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-```{r mutation-burden-plot, message = FALSE}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-mut_order <- list() ## store data in list 
-for(n in 1:max(umap_cancertype$knn_clust)) { ## loop through all the clusters
-  message(n)
-  mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
-    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
-    arrange(mut_count) ## arrange mutation in increasing order
-}
-mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
-
-mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
-mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
-
-## Need the breaks positions for the x-axis
-previous_class <- 0 ## starting position 1
-position_of_previous_class <- 0 ## starting position 2
-xaxis_breaks <- list() ## list for for loop
-for(n in 1) { ## loop through 1 to 12 clusters
-  message(n)
-  class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
-  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-  previous_class <- class_spec ## storing class halfway mark
-  for(n in 2:max(as.numeric(as.character(mut_order$knn_clust)))) {
-    message(n)
-    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
-    class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
-    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-    previous_class <- class_spec ## storing class halfway mark
-  }
-}
-xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
-
-total_n <- as.data.frame(table(mut_order$knn_clust))
-
-## for making dup discrete x-axis
-guide_axis_label_trans <- function(label_trans = identity, ...) {
-  axis_guide <- guide_axis(...)
-  axis_guide$label_trans <- rlang::as_function(label_trans)
-  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
-  axis_guide
-}
-
-guide_train.guide_axis_trans <- function(x, ...) {
-  trained <- NextMethod()
-  trained$key$.label <- x$label_trans(trained$key$.label)
-  trained
-}
-
-ggplot(mut_order, aes(x = V1, y = log10(mut_count + 1), color = knn_clust)) + 
-  geom_point() + 
-  ylab("Log10(Mutation Count)") + 
-  xlab("") +
-  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("Class =" ,1:12)) +
-  guides(x.sec = guide_axis_label_trans(~paste(c("n = ", total_n$Freq)))) + ## create second x-axis
-  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  #scale_y_log10() +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        legend.position = "none") +
-  scale_color_manual(values = kelly.colours[-c(1:2)])
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-there is a whole set of genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways, and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
 ```{r figure-4-data-generation, eval = FALSE}
@@
 ```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_4_function.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r figure-4-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/moderate_enrich_prob.Rda")
load("../data/gene_data.Rda")
# load("../data/gdc_tumor_coding_complete.Rdata")
# 
# missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
# splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
# frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
# frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
# utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
# utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
# nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
# nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
# 
# gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
# rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
# gc()
# gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
# save(gdc_mut_count_patient, file = "../data/gdc_mut_count_patient_booleanized.Rda")
load("../data/gdc_mut_count_patient_booleanized.Rda")
filtered_class <- 12
n_path <- 10
enrich_cutoff <- 0.30
oncoprint_style <- function(filtered_class, n_path, enrich_cutoff = 0.30) {
  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
  toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
  filter(clust %in% filtered_class) %>% ## filter the cluster
  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] %>% na.omit() ## getting the top10 enriched pathways
  if(length(toppath) == 0) { ## if no enriched pathways. stop and print error message
    stop("No enriched pathways... select a lower enrich_cutoff")
  }
  if(length(toppath) == 1) { ## if there is only 1 enriched path. stop and print error message
    stop("1 enriched pathway found. Use oncoprint to do your analysis")
  }
  genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
  .[rowSums(.[, -1])>0,]

  make_combinations <- function(x) { ## function to make all the possible combination of genes in pathway
  l <- length(x)
  mylist <- lapply(1:l, function(y) {
    combn(x, y, simplify = FALSE)
  })
  mylist
  }
  path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10

  top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
    'colnames<-' ("sum_top10") %>% ## rename col
    rownames_to_column(var = "gene_name") ## make rownames a col

  path_match_in_set <- list() ## making a list for matchings in sets

  for(n in 1:length(path_combo)) {
    message(n)
    path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
    path_match_within_set <- list() ## making a list for matching within sets
    for(p in 1:length(path_combo_set)) {
      message(p)
      path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
      spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
      if(length(spec_combo) > 1) {
        spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
      }
      if(length(spec_combo) == 1) {
      names(spec_combo) <- "combo"
      }
      spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
      spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
      any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
      if(!"TRUE" %in% any_matches) { ## if no matches then skip
        next
        }
      if("TRUE" %in% any_matches) { ## if are matches then do this
        path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
      }
   }    
    if(length(path_match_within_set) > 0) {
      path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
    }
  }
  path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
  ### getting clustering order from pheatmap
  most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
    'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
    rownames_to_column(var = "order")

  gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
  arrange(order) %>% .$gene_name ## getting the gene name order

  ## gene mutation data and tumor samples - Booleanize mutation count for each tumor sample
  message("working on mutation data")
  filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
    left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
    column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
    .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
  message("Running pheatmap to get coordinates")
  pheatmap_res <- pheatmap(filtered_mut_data[gene_order,],
         show_rownames = T,
         show_colnames = F,
         treeheight_row = 0,
         treeheight_col = 0,
         cellheight = 5,
         cellwidth = 1,
         cluster_cols = T,
         cluster_rows = F, ## turn off row clustering
         annotation_legend = F,
         legend_breaks = c(0, 0.25, 0.75, 1),
         legend_labels = c("","Not Mut", "Mut", ""),
         color = c("grey", "firebrick3"),
         clustering_method = "ward.D2",
         fontsize = 10,
         legend = T)

  pheatmap_order <- pheatmap_res$tree_col$order ## getting sample clustering order
  sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
  sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
  message("Preparing data for plot")
  gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
  gene_2_path_heatmap$membership <- as.character(gene_2_path_heatmap$membership) ## turn boolean values into categories
  ## set levels based on clustering algorithm above
  gene_2_path_heatmap$gene_name <- factor(gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
  gene_2_path_heatmap$sample_id <- factor(gene_2_path_heatmap$sample_id, levels = c(sample_order))

  gene_2_path_heatmap <- left_join(gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
  heatmap_ggplot <- ggplot(data = gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none") 
  ## left annotatoin data generation
  l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap

  for(p in toppath){
    message(p)
    l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
    names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
    ## for any path_combo that includes the pathway name change the 0 into a 1
    if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
    }
    if(p == "Signaling by Type 1 Insulin-like Growth Factor 1 Receptor (IGF1R)") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "Signaling by Type 1 Insulin", x = l_anno$path_combo), p] <- 1
    }
    if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA" & p != "Signaling by Type 1 Insulin-like Growth Factor 1 Receptor (IGF1R)") { ## if it is not AUF1 do it normally
    l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
    }
  }

  l_anno_ggplot <- list() ## list for storing ggplots
  path_legend <- list()

  kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
                     "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
                     "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
                     "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")

  for(n in 1:length(toppath)){
    message(n)
    p <- toppath[n]
    message(p)
    spec_path <- l_anno[,c("gene_name", p)]
    names(spec_path)[names(spec_path) == p] <- "p"
    anno_color <- c("0" = "white", "1" = kelly.colours[2+n])
    path_ggplot <- ggplot(spec_path) +
    geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
            stat = "identity", 
            width = 1) +
    scale_fill_manual(values = anno_color, breaks = "1", name = "", labels = p) +
    theme_void() 
    path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
    l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
  }

  plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], heatmap_ggplot, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, 10), 15))
plot_legend <- plot_grid(path_legend[[1]], path_legend[[2]], path_legend[[3]], path_legend[[4]], path_legend[[5]], path_legend[[6]], path_legend[[7]], path_legend[[8]], path_legend[[9]], path_legend[[10]], ncol = 5)

return(plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5)))
}

test1 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
test12 <- oncoprint_style(filtered_class = 12, n_path = 10)
test11 <- oncoprint_style(filtered_class = 11, n_path = 10)

```

### 2020-07-14 text/figure_s1.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data

umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
  geom_point(size = 1) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```

### 2020-07-14 text/figure_s2.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
 ```{r supplementary-figure-s2-data-generation}
@@
 ```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_s3.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
 <div style="text-align: right"> Supplementary Figure S3 Data </div>
@@
 ```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/figure_s4.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
----
-title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-
-```{r echo=FALSE, eval=FALSE}
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-#library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(umap)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-```
-
-## Abstract
-
-```{r child='abstract.Rmd'}
-```
-
-## Introduction
-
-```{r child='introduction.Rmd'}
-```
-
-## Results
-
-### A taxonomy of tumors based on disrupted molecular pathways
-
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
-
-It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
-
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
-
-#### Independence from tissue-of-origin.
-
-Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
-patients hosting these tumors have potentially different underlying molecular
-pathologies.
-
-#### Independence from molecular and histological subtype.
-
-Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
-
-#### Independence from drivers of genome instability.
-
-There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
-
-#### Independence of stage, mutation count and mutation profile.
-
-Tumor staging is based on physico-pathological criteria, including diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
-disruption classes.
-
-Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
-
-### Each class is defined by a specific set of pathway disruptions
-
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
-
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
-
-### Gene-Pathway redundancy contributes to cluster identity
-
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
-
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
-
-<mark>As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
-
-### Enrichment of pathways in metastasis is class-specific
-
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
-
-Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
-
-### Pathway disruption classes vary in short-term prognosis of survival
-
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
-```
-
-## Materials & Methods
-
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
-
-### Selection of pathways
-
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
-
-### <mark>Filtering genes <- Peter to write</mark>
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
-
-## Discussion
-
-```{r child='discussion.Rmd'}
-```
-
-## Supplementary Information
-
-### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-
 <div style="text-align: right"> Supplementary Figure S4 Data </div>
@@
 ```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
-## References

### 2020-07-14 text/manuscript.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

--- 2020-07-14 text/manuscript.Rmd
@@
 carrying mutations (this data is visualized in **figure 1a**).
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-<div style="text-align: right"> Reactome Pathway Creation </div>
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
-gc()
-```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-<div style="text-align: right"> Genes to pathway transformation </div>
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-<div style="text-align: right"> Figure 1 data </div>
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1C data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
-
-####################
-## Figure 1E data ##
-####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
-
-####################
-## Figure 1H data ##
-####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  labs(tag = "A") +
-  ylab("") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
-    theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
-    theme(axis.text.x = element_text(size = 12),
-          axis.text.y = element_text(size = 12),
-          axis.title.x = element_text(size = 12),
-          axis.title.y = element_text(size = 12),
-          legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
-  theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
-AAAAHHIIJJ
-AAAAKKLL##
-'
-
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
-
-```
-
+```{r child='data_set_generation.Rmd'}
+```
+
+```{r child='figure_1.Rmd'}
+```
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
@@
 frequency, possibly owing to relative diversity among all tumors.
-<div style="text-align: right"> Figure 2 Data </div>
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
-
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
-
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
-
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
-
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
-```
-
-```{r figure-2-plot, fig.width=72, fig.height=48}
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
-layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
-'
-
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
+```{r child='figure_2.Rmd'}
 ```
@@
 distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
-
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
-
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
-
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
-
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "effect_heatmap.Rda")
-```
-
-```{r group-pathway-enrichment heatmap}
-load("../data/effect_heatmap.Rda") ## load data
-
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
-
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
-
-fig3 + fig3_legend + plot_layout(design = layout)
-```
-
+```{r child='figure_3.Rmd'}
+```
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
@@
 predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
+there is a whole set of genes in the proteosomal pathway whose activity
+contributes to all the aforementioned pathways, and the majority of tumors in
 these classes have a single mutation in one of these genes. Nonetheless, a
@@
 phenotypes by converging at the pathway level.
-
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
-
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
-
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
-
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
-
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
-
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
-}
-
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r figure-4, message = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot
-
-```
-
-![](../figures/figure4_temp.png)
+```{r child='figure_4.Rmd'}
+```
 <p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
@@
 The highlighted genes at right are all proteasomal subunits.</p>
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
 
@@
 wherein etc etc</mark>
-<div style="text-align: right"> Table 1 Data </div>
-```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
-}
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
-  }
-
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
+```{r child='table_1.Rmd'}
 ```
@@
 non-metastatic tumors of other clusters.
-<div style="text-align: right"> Table 2 Data </div>
-```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
-
-m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
-m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
-
-prepare_rbeta_data_p1 <- function(data, path_analysis) {
-  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
-  ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
-  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
-  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
-  cluster_path_count <- list() ## Creating a list for the forloop
-  for(n in clust_group) {
-    print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
-    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
-    df1 <- as.data.frame(rowSums(df1))
-    names(df1) <- n
-    cluster_path_count[[n]] <- df1
-  }
-  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
-  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
-  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
-  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
-  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
-  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
-  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
-  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
-  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
-  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
-  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
-  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
-  return(rbeta_path_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
-
-## repeat same function above with metastasis sample
-m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
-
-## function to compiling data for rbeta analysis
-prepare_rbeta_data_p2 <- function(data) {
-  enrich_prob <- data.frame(path = data$path,
-                               clust = data$clust,
-                               fg_mut = data$clust_mut,
-                               fg_not_mut = data$tot_n_clust - data$clust_mut,
-                               bg_mut = data$global_mut,
-                               bg_not_mut = data$global_sample - data$global_mut)
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in unique(as.character(enrich_prob$clust))) {
-  print(n)
-  df <- subset(enrich_prob, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-}
-
-fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
-bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
-enrich_prob_df <- enrich_prob
-enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
-enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
-enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
-enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
-enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
-return(enrich_prob_df)
-}
-
-m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
-m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
-
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
-rm(list = ls(all.names = T)) ## cleaning up data
-gc()
-load("../data/m0_m1_rbeta.Rda")
-## Metastatic (foreground) - Nonmetastatic (background) Rbeta
-moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
-metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-## Moderate Difference
-CI <- 0.025
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
-gc()
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
-}))
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-
-m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
-                                   clust = m1_rbeta$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r table-2}
-load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
-
-table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
-table2$Signif <- NULL ## we don't need this anymore
-
-table2 %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
+```{r child='table_2.Rmd'}
 ```
@@
 ### Supplementary Figure S1
-
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
+```{r child = 'figure_s1.Rmd'}
 ```
@@
 ### Supplementary Figure S2
-
-```{r supplementary-figure-s2-data-generation}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
+```{r child = 'figure_s2.Rmd'}
 ```
@@
 ### Supplementary Figure S3
-
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
-```{r supplementary-figure-s3-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene_data.Rda")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
-
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
-selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
-
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_mut, file = "selected_mut.Rda")
-```
-
-```{r supplementary-figure-s3}
-load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
-```
-
+```{r child = 'figure_s3.Rmd'}
+```
 ![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
@@
 ### Supplementary Figure S4
-
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "selected_path.Rda")
-```
-
-```{r supplementary-figure-s4, out.width = '90%'}
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-
-n_path <- 377 ## number of pathways we want to see
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
-                      ncol = 5) ## facet by pathways
-
+```{r child = 'figure_s4.Rmd'}
 ```
@@
 ### Supplementary Table ST1
-
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
-
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
-
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
-
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
-```
-
-```{r supplemental-table-st1}
-load("../data/table_st1.Rda")
-
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
+```{r child = 'table_s1.Rmd'}
 ```

### 2020-07-14 text/table_1.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

<div style="text-align: right"> Table 1 Data </div>
```{r table-1-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/moderate_enrich_prob.Rda") ## load data
load("../data/gene2_level2_lookup.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")
load("../data/pathname2level2.Rda")

id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id) %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
  left_join(., pathname2level2, by = "PATH_ID") %>%
  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
  dplyr::select(-PATH_ID) %>%
  t() %>% as.data.frame()

gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id)

## parameters ##
enrich_cutoff <- 0.30
howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
show_perc <- TRUE
show_n <- TRUE
mutated_gene_position <- 1
################

clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff

clust_spec_enrich_top5 <- list()
for(n in unique(clust_spec_enrich$clust)) {
  message(n)
  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
    dplyr::select(path, clust) %>% .[1:5,]
}
clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
rm(clust_spec_enrich)

## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
type <- unique(tumor_clusters$project_code) ## type of cancer type in data
cluster_spec_enrich_table <- list() ## list for forloop

for(d in unique(clust_spec_enrich_top5$clust)) {
  message(d)
  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
  top_genes_in_path_cancer_type <- list()
  for(n in clust_toppath) { ## for each of the top paths do this
    message(n)
    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
    'rownames<-' (rownames(id_to_names)) %>% 
    filter(. == 1) %>% ## filter for genes that are in that pathway
    rownames(.)
    top_gene_in_path <- list()
    for(t in type) { ## looping through cancer type within class
      message(t)
      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>% 
          mutate(n_sample = length(class_type_spec),
               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
      }
      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
        mutate(n_sample = length(class_type_spec),
               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
      }
      if(show_perc == TRUE & show_n == FALSE) {
         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
      }
      if(show_perc == TRUE & show_n == TRUE) {
         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
      }
      if(show_perc == FALSE) {
         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
      }
      }
      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
        top_gene_in_path[[t]] <- "NA" 
      }
    }
    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
  }
  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
  }

cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
rm(list = ls(all.names = T))
gc()
```

```{r table-1}
load("../data/cluster_spec_enrich_table1.Rda")
cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```

### 2020-07-14 text/table_2.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

<div style="text-align: right"> Table 2 Data </div>
```{r table-2-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gdc_reactome_path_analysis.Rda")

tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)

m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples

prepare_rbeta_data_p1 <- function(data, path_analysis) {
  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
  table(names(path_analysis) == data$V1) ## good they match now
  ## Now we are going to build group specific cluster data
  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
  cluster_path_count <- list() ## Creating a list for the forloop
  for(n in clust_group) {
    print(n)
    df <- subset(data, knn_clust == n) ## subset cluster number
    df <- df$V1 ## getting tumor sample id
    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
    df1 <- as.data.frame(rowSums(df1))
    names(df1) <- n
    cluster_path_count[[n]] <- df1
  }
  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
  return(rbeta_path_df)
}

m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
## Combining cluster 6/7 cluster 910 cluster 1112
combo_group <- c("6", "9", "11")
compo_group_df <- list() ## list for loop
for(n in combo_group) {
  print(n) ## first cluster in set
  m <- as.numeric(n) + 1 ## second cluster in set
  m <- as.character(m)
  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
  p <- paste0(n,m)
  df2 <- m0_rbeta[1:377,]
  df2$clust <- p ## replace with new clust name
  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
  compo_group_df[[p]] <- df2 ## store in list and continue loop
}
compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
save(m0_rbeta, file = "m0_rbeta.Rda")

## repeat same function above with metastasis sample
m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
compo_group_df <- list() ## list for loop
for(n in combo_group) {
  print(n) ## first cluster in set
  m <- as.numeric(n) + 1 ## second cluster in set
  m <- as.character(m)
  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
  p <- paste0(n,m)
  df2 <- m0_rbeta[1:377,]
  df2$clust <- p ## replace with new clust name
  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
  compo_group_df[[p]] <- df2 ## store in list and continue loop
}
compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
save(m1_rbeta, file = "m1_rbeta.Rda")

## function to compiling data for rbeta analysis
prepare_rbeta_data_p2 <- function(data) {
  enrich_prob <- data.frame(path = data$path,
                               clust = data$clust,
                               fg_mut = data$clust_mut,
                               fg_not_mut = data$tot_n_clust - data$clust_mut,
                               bg_mut = data$global_mut,
                               bg_not_mut = data$global_sample - data$global_mut)
fg_mut_rate <- list()
bg_mut_rate <- list()
for(n in unique(as.character(enrich_prob$clust))) {
  print(n)
  df <- subset(enrich_prob, clust == n)
  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
  fg_numerator <- sum(df$fg_mut)
  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
  bg_numerator <- sum(df$bg_mut)
  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
}

fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
enrich_prob_df <- enrich_prob
enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
return(enrich_prob_df)
}

m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)

save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
rm(list = ls(all.names = T)) ## cleaning up data
gc()
load("../data/m0_m1_rbeta.Rda")
## Metastatic (foreground) - Nonmetastatic (background) Rbeta
moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
})
moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
})

metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)

## ModerateEnrichment probability 
prob <- moderate_fg > moderate_bg ## boolean data
mean_prob <- apply(prob, MARGIN = 2, function(x) {
  mean(x)
})
## Moderate Difference
CI <- 0.025

diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
rm(moderate_fg, moderate_bg, prob)
gc()
quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
}))

sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
}))

m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
                                   clust = m1_rbeta$clust,
                                   probability = mean_prob,
                                   signif = sig_fg_bg[,1],
                                   difference = quant_diff[,2],
                                   lower = quant_diff[,1],
                                   upper = quant_diff[,3])
save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
rm(list = ls(all.names = T))
gc()
```

```{r table-2}
load("../data/m1_minus_m0_per_cluster_rbeta.Rda")

table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col

table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
table2$Signif <- NULL ## we don't need this anymore

table2 %>% knitr::kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T, color = "black")
```

### 2020-07-14 text/table_s1.Rmd
made child docs, fixed mutation burden plot, worked on fig4 function - still needs some tweaking

<div style="text-align: right"> Supplementary Figure ST1 Data </div>
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

## Parameters ##
n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
show_perc <- TRUE ## show perc for mutated gene

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways

id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id) %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
  left_join(., pathname2level2, by = "PATH_ID") %>%
  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
  dplyr::select(-PATH_ID) %>%
  t() %>% as.data.frame()

gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id)

table_st1 <- list() ## list for forloop

for(n in toppath) {
  message(n)
  ## finding top mutated gene in pathway
  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
    'rownames<-' (rownames(id_to_names)) %>% 
    filter(. == 1) %>% ## filter for genes that are in that pathway
    rownames(.)
  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
      mutate(n_sample = length(gene_mutation),
             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
  top_gene_in_path <- list()
  for(p in mutated_gene_position) {
    message(p)
    if(show_perc == TRUE) {
    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
    }
    if(show_perc == FALSE) {
    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
    }
  }
  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
  ## calculating the percent mutated samples
  spec_path <- data.frame(Pathway = n,
                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
}
table_st1 <- as.data.frame(do.call(rbind, table_st1))
save(table_st1, file = "table_st1.Rda")  
rm(list = ls(all.names = T))  
gc()
```

```{r supplemental-table-st1}
load("../data/table_st1.Rda")

table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T, color = "black")
```

### 2020-07-15 code/table1.Rmd
move all code blocks to child Rmd files

--- 2020-07-14 text/table_1.Rmd
@@
 ```
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>

### 2020-07-15 text/manuscript.Rmd
move all code blocks to child Rmd files

@@
 carrying mutations (this data is visualized in **figure 1a**).
-```{r child='data_set_generation.Rmd'}
+
+```{r child='../code/figure1.Rmd'}
 ```
-
-```{r child='figure_1.Rmd'}
-```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
 
@@
 frequency, possibly owing to relative diversity among all tumors.
-```{r child='figure_2.Rmd'}
+
+```{r child='../code/figure2.Rmd'}
 ```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
-Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
-which each column represents a tumor and each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. All 377 pathways are shown. Red signifies
-pathway disrupted pathways, grey signifies not disrupted.</p>
 
@@
 distinct of the 12, we chose to investigate these two groups further.
-```{r child='figure_3.Rmd'}
+<div style="text-align: right"> Figure 3 data </div>
+
+```{r child='../code/figure3.Rmd'}
 ```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this).</p>
 
@@
 phenotypes by converging at the pathway level.
-```{r child='figure_4.Rmd'}
+
+```{r child='../code/figure4.Rmd'}
 ```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.</p>
 
@@
 "positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
+Second we looked at survival in mutations results results results (**figure 4c**). 
+Third, we checked the mutation classes and found the overwhelming
 majority (x out of y) are missense _vs._ various other deletarious classes (see
@@
 pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary
-tables ST2-ST13**). We find that in most clusters the top pathways are
+for each pathway as a function of cancer type (**Table 1** and **Supplementary tables ST2-ST13**). 
+We find that in most clusters the top pathways are
 determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
@@
 wherein etc etc</mark>
-```{r child='table_1.Rmd'}
+
+```{r child='../code/table1.Rmd'}
 ```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>
 
@@
 non-metastatic tumors of other clusters.
-```{r child='table_2.Rmd'}
+
+```{r child='table2.Rmd'}
 ```
@@
 
-### Supplementary Figure S1
-```{r child = 'figure_s1.Rmd'}
-```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
-### Supplementary Figure S2
-```{r child = 'figure_s2.Rmd'}
+```{r child='../code/supplements.Rmd'}
 ```
 
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
-```{r child = 'figure_s3.Rmd'}
-```
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S4
-```{r child = 'figure_s4.Rmd'}
-```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
-### Supplementary Table ST1
-```{r child = 'table_s1.Rmd'}
-```
-
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
-
 ## References

### 2020-07-15 code/figure1.Rmd

--- 2020-07-14 text/data_set_generation.Rmd
@@
 ```
+<div style="text-align: right"> Figure 1 data </div>
+```{r tumor-clusters, eval = FALSE}
+####################
+## Figure 1A data ##
+####################
+load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/gdc_tumor_coding_complete.Rdata")
+
+path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
+path_daisy <- as.dist(as.matrix(path_daisy))
+
+path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
+path_daisy_row <- as.dist(as.matrix(path_daisy_row))
+
+pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
+         show_rownames = F,
+         show_colnames = F,
+         cellheight = 1,
+         cellwidth = 0.5,
+         clustering_distance_cols = path_daisy,
+         clustering_distance_rows = path_daisy_row,
+         cluster_cols = T,
+         cluster_rows = T,
+         cutree_cols = 19,
+         annotation_legend = F,
+         legend_breaks = c(0, 0.25, 0.75, 1),
+         legend_labels = c("","Not Mut", "Mut", ""),
+         color = c("grey", "firebrick3"),
+         clustering_method = "ward.D2",
+         legend = T)
+dev.off()
+pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
+df <- rownames(gdc_reactome_path_analysis)
+pathway_dendro <- df[pathway_dendro]
+
+tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
+  as.data.frame(.) %>%
+  dplyr::select(project_code, submitted_specimen_id) %>%  
+  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
+
+cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
+
+cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
+
+for(n in cancer_type) {
+  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
+    filter(project_code == n) %>% ## filter for specific cancer type
+    .[, "sample_id"] ## retain only sample_id
+  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
+  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
+                          show_rownames = F,
+                          show_colnames = F,
+                          cellheight = 1,
+                          cellwidth = 0.5,
+                          cluster_cols = T,
+                          cluster_rows = F,
+                          annotation_legend = F,
+                          color = c("grey", "firebrick3"),
+                          clustering_method = "single", ## ordering is single linkage
+                          legend = T)
+  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
+  cancer_path <- names(cancer_path) ## getting the sample_id original order 
+  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
+  rm(cancer_info, cancer_path)
+  gc()
+}
+dev.off()
+
+for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
+  print(n)
+  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
+  names(cancer_order) <- "sample_id"
+  cancer_order$type <- n
+  for(p in cancer_type[-1]) {
+    print(p)
+    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
+    names(df1) <- "sample_id"
+    df1$type <- p
+    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
+    rm(df1)
+    gc()
+  }
+}
+
+cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
+
+cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
+  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
+cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
+
+## set levels based on clustering algorithm above
+cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
+cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
+
+cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
+save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
+
+####################
+## Figure 1C data ##
+####################
+gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
+  rownames_to_column(var = "sample") %>% ## move path_names to column
+  .[match(rownames(tumor_samples), .$sample),]
+
+## umap analysis
+set.seed(3296)
+gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
+gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
+
+####################
+## Figure 1E data ##
+####################
+## To identify clusters using hdbscan, use umap in the third dimension
+set.seed(3296)
+gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
+gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
+rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
+hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
+umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
+
+####################
+## Figure 1F data ##
+####################
+## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
+umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
+umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
+                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
+                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
+                                                                  k=15)
+
+####################
+## Figure 1G data ##
+####################
+missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
+splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+
+mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
+gc()
+mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
+  colSums() %>% as.data.frame() %>% ## sum the mutation count
+  'colnames<-' (c("mut_count")) %>% ## rename colname
+  rownames_to_column(var = "V1") ## move rownames to column
+
+umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
+
+####################
+## Figure 1H data ##
+####################
+tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
+  as.data.frame(.) %>%
+  dplyr::select(project_code) %>% 
+  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
+  rownames_to_column(var = "V1")
+
+umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
+
+####################
+## Figure 1I data ##
+####################
+## getting subtype information on our cancer type
+subtypes <- PanCancerAtlas_subtypes() 
+tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
+subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
+  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
+  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
+  distinct() %>% ## only unique columns
+  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
+  filter(Subtype_Selected != "Normal") %>%
+  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
+
+#save(subtypes, file = "subtypes.Rda")
+load("../data/subtypes.Rda")
+
+umap_2d_3d_coors <- umap_2d_3d_coors %>%
+  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
+  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
+  dplyr::select(-pan.samplesID)
+
+##########################
+## Figure 1K and L data ##
+##########################
+## getting tumor_metastasis, staging
+tcga_clinical_tumor_staging <- list()
+for(n in cancer_type) {
+  print(n)
+  cancername <- paste0("TCGA-",n)
+  query <- GDCquery(project = cancername,
+                  data.category = "Clinical",
+                  data.type = "Clinical Supplement",
+                  data.format = "BCR Biotab",
+                  file.type = "patient")
+  GDCdownload(query)
+  clinical <- GDCprepare(query)[[1]]
+  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
+    cat("Yep, ajcc info in there!\n")
+    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
+  }
+  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
+    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
+    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
+    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
+  }
+  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
+    cat("Nope, ajcc info not there!\n")
+    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
+    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
+    }
+  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
+  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
+  tcga_clinical_tumor_staging[[n]] <- clinical
+}
+
+tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
+tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
+missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
+
+tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
+  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
+  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
+  rbind(tcga_clinical_tumor_staging, .) ## rbind
+
+save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
+
+umap_2d_3d_coors <- umap_2d_3d_coors %>%
+  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
+  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
+  dplyr::select(-bcr_patient_barcode)
+
+M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
+M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
+NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
+
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
+
+stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
+stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
+stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
+stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
+unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
+
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
+
+write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
+```
+
+```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5}
+load("../data/cancer_disrupt_heatmap.Rda")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
+fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
+  geom_tile(size = 0.6) +
+  scale_fill_manual(values = c("grey", "firebrick3")) +
+  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
+scale_y_discrete(position = "left") +
+  labs(tag = "A") +
+  ylab("") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        strip.text.y = element_text(size = 12, hjust = 0),
+        strip.text.y.left = element_text(angle = 0),
+        strip.background = element_blank(),
+        legend.position = "none")
+
+## fixing the text cutoff in heatmap
+ fig1Ag <- ggplotGrob(fig1A)
+
+ for(i in which(grepl("strip-l", fig1Ag$layout$name))){
+   fig1Ag$grobs[[i]]$layout$clip <- "off"
+ }
+
+ fig1A <- as.ggplot(fig1Ag)
+ rm(fig1Ag)
+
+## umap monochrome 2d
+fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
+  geom_point(colour = "grey") +
+  xlab("umap_1") +
+  ylab("umap_2") +
+  theme_classic() +
+  labs(tag = "C") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+## umap hdb_scan
+fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
+  guides(color = F) +
+  theme_classic() +
+  xlab("umap_1") +
+  ylab("umap_2") +
+  theme_classic() +
+  labs(tag = "E") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+
+## umap knn_clust 
+fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
+  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
+  guides(color = F) +
+  theme_classic() +
+  xlab("umap_1") +
+  ylab("umap_2") +
+  theme_classic() +
+  labs(tag = "F") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+
+umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
+
+## violin mutation count
+fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
+    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
+    geom_boxplot(aes(fill = knn_clust), width=0.2) +
+    theme_classic() +
+    scale_y_log10() +
+    xlab("") +
+    ylab("Log(Mutation Count)") +
+    labs(color = "", tag = "G") +
+    theme(axis.text.x = element_text(size = 12),
+          axis.text.y = element_text(size = 12),
+          axis.title.x = element_text(size = 12),
+          axis.title.y = element_text(size = 12),
+          legend.position = "none") +
+  scale_fill_manual(values = kelly.colours[-c(1:2)])
+
+## umap by cancer type
+fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "H", color = "") +
+  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
+                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap of brca subtype
+fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() +
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "I", color = "") +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap brca somatic mutations
+load("../data/gdc_gene_exp_adj.Rda")
+load("../data/gene_data.Rda")
+brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
+brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
+  rownames_to_column(var = "ensembl_gene_id") %>%
+  left_join(., brca, by = "ensembl_gene_id") %>% 
+  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
+  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
+  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
+  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
+                           BRCA1 == "0" ~ "NA"),
+         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
+                           BRCA2 == "0" ~ "NA")) %>%
+  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
+  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
+  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
+  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
+
+fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
+  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(color = "", tag = "J") +
+  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
+                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap stage
+fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point() + 
+  theme_classic() +
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "K", color = "") +
+  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
+                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+## umap metastasis
+fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  ylab("umap_2") +
+  xlab("umap_1") +
+  labs(tag = "L", color = "") +
+  scale_color_manual(labels = c("M0", "M1"), 
+                     values = c('#a6cee3','#1f78b4')) +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(),
+        legend.position = "bottom")
+
+layout <- '
+AAAABBCCDD
+AAAAEEEFFF
+AAAAGGGG##
+AAAAGGGG##
+AAAAHHIIJJ
+AAAAKKLL##
+'
+
+fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
+
+```
+
+```{r mutation-burden-plot, message = FALSE}
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
+mut_order <- list() ## store data in list 
+for(n in 1:max(umap_cancertype$knn_clust)) { ## loop through all the clusters
+  message(n)
+  mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
+    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
+    arrange(mut_count) ## arrange mutation in increasing order
+}
+mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
+
+mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
+mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
+
+## Need the breaks positions for the x-axis
+previous_class <- 0 ## starting position 1
+position_of_previous_class <- 0 ## starting position 2
+xaxis_breaks <- list() ## list for for loop
+for(n in 1) { ## loop through 1 to 12 clusters
+  message(n)
+  class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
+  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+  previous_class <- class_spec ## storing class halfway mark
+  for(n in 2:max(as.numeric(as.character(mut_order$knn_clust)))) {
+    message(n)
+    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
+    class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
+    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+    previous_class <- class_spec ## storing class halfway mark
+  }
+}
+xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
+
+total_n <- as.data.frame(table(mut_order$knn_clust))
+
+## for making dup discrete x-axis
+guide_axis_label_trans <- function(label_trans = identity, ...) {
+  axis_guide <- guide_axis(...)
+  axis_guide$label_trans <- rlang::as_function(label_trans)
+  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
+  axis_guide
+}
+
+guide_train.guide_axis_trans <- function(x, ...) {
+  trained <- NextMethod()
+  trained$key$.label <- x$label_trans(trained$key$.label)
+  trained
+}
+
+ggplot(mut_order, aes(x = V1, y = log10(mut_count + 1), color = knn_clust)) + 
+  geom_point() + 
+  ylab("Log10(Mutation Count)") + 
+  xlab("") +
+  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("Class =" ,1:12)) +
+  guides(x.sec = guide_axis_label_trans(~paste(c("n = ", total_n$Freq)))) + ## create second x-axis
+  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
+  #scale_y_log10() +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 12),
+        axis.text.y = element_text(size = 12),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        legend.position = "none") +
+  scale_color_manual(values = kelly.colours[-c(1:2)])
+```
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
+Classification of tumors by pathway disruptions.** a) Each of 377 selected
+Reactome pathways (columns) is classified as disrupted if one or more genes is
+mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
+standardized abrreviations from the TCGA project. b) MCA analysis was performed
+restricted to the 36 informative components. c) The resulting MCA was projected
+down to 3 dimensions for the purpose of density based clustering (only the first
+2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
+perform hierarchical density based clustering tuned to minimize the fraction of
+unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
+projection; tumors are colored by cluster membership (see legend). This results
+in some tumors remaining unclassified (grey points). f) kNN was used to assign
+membership to the remaining tumors. g) Violin plots of sample mutation counts
+for tumors in each class. h) Projection of cancer type identity onto UMAP.
+Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
+Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
+cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
+UMAP. k) projection of tumor stage onto the UMAP,
+regardless of cancer type. l) Projection of metastatic status onto the UMAP.
+Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>

### 2020-07-15 code/figure2.Rmd

--- 2020-07-14 text/figure_2.Rmd
@@
 
-```{r figure-2-plot, fig.width=72, fig.height=48}
+```{r figure-2-plot, fig.width=72, fig.height=48, fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions.
+Waterfall plot; columns represent tumors.
+Each row represents a pathway.
+The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively.
+All 377 pathways are shown.
+Red signifies pathway disrupted pathways, grey signifies not disrupted."}
+
 ## top100 pathways pancancer oncoprint

### 2020-07-15 code/figure3.Rmd

--- 2020-07-14 text/figure_3.Rmd
@@
-<div style="text-align: right"> Figure 3 data </div>
 ```{r group-pathway-enrichment-data-generation, eval=FALSE}
@@
 
-```{r group-pathway-enrichment heatmap}
+```{r group-pathway-enrichment heatmap, fig.cap="**Figure 3.
+Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
+The annotation at left indicates the Reactome pathway group at the second level (check this)."}
 load("../data/effect_heatmap.Rda") ## load data

### 2020-07-15 code/figure4.Rmd

--- 2020-07-14 text/figure_4.Rmd
@@
 
-```{r figure-4, message = FALSE}
+```{r figure-4, message = FALSE, fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
+The genes are organized according to which of the top distinguishing pathways for this class that they effect.
+Some genes affect multiple pathways, as indicated by the bars at left.
+The highlighted genes at right are all proteasomal subunits.", eval=FALSE}
+
 load("../data/class67_gene_2_path_heatmap.Rda") ## load data

### 2020-07-15 code/supplements.Rmd

### Supplementary Figure S1

```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data

umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
  geom_point(size = 1) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```
![**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]

### Supplementary Figure S2

```{r supplementary-figure-s2-data-generation}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
  geom_point(size = 0.5) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S3

<div style="text-align: right"> Supplementary Figure S3 Data </div>
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

## Booleanize mutation count for each tumor sample
missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)

gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
gc()
gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1

##oncogene
oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
##tumor_supressor
tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor

selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data

selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join

selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
  left_join(., selected_mut, by = "V1") %>%
  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
  filter(value != 0)

save(selected_mut, file = "selected_mut.Rda")
```

```{r supplementary-figure-s3}
load("../data/selected_mut.Rda")
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
  geom_point(size = 0.5) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
                                           "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
                      ncol = 5) ## facet by gene
```

![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S4

<div style="text-align: right"> Supplementary Figure S4 Data </div>
```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway

selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join

selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
  left_join(., selected_path, by = "V1") %>%
  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
  filter(value != 0)

save(selected_path, file = "selected_path.Rda")
```

```{r supplementary-figure-s4, out.width = '90%'}
load("../data/selected_path.Rda") ## load data
load("../data/gdc_reactome_path_analysis.Rda")
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

n_path <- 377 ## number of pathways we want to see

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway.. needed for facet

ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
  geom_point(size = 0.5) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = toppath),
                      ncol = 5) ## facet by pathways

```

![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]

### Supplementary Table ST1

<div style="text-align: right"> Supplementary Figure ST1 Data </div>
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

## Parameters ##
n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
show_perc <- TRUE ## show perc for mutated gene

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways

id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id) %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
  left_join(., pathname2level2, by = "PATH_ID") %>%
  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
  dplyr::select(-PATH_ID) %>%
  t() %>% as.data.frame()

gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id)

table_st1 <- list() ## list for forloop

for(n in toppath) {
  message(n)
  ## finding top mutated gene in pathway
  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
    'rownames<-' (rownames(id_to_names)) %>% 
    filter(. == 1) %>% ## filter for genes that are in that pathway
    rownames(.)
  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
      mutate(n_sample = length(gene_mutation),
             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
  top_gene_in_path <- list()
  for(p in mutated_gene_position) {
    message(p)
    if(show_perc == TRUE) {
    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
    }
    if(show_perc == FALSE) {
    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
    }
  }
  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
  ## calculating the percent mutated samples
  spec_path <- data.frame(Pathway = n,
                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
}
table_st1 <- as.data.frame(do.call(rbind, table_st1))
save(table_st1, file = "table_st1.Rda")  
rm(list = ls(all.names = T))  
gc()
```

```{r supplemental-table-st1}
load("../data/table_st1.Rda")

table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T, color = "black")
```

![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]

![**Supplemental figure S6. Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively. The orange and green curves show the model fits._](../figures/fig6A.png)

### 2020-07-15 code/figure2.Rmd
merge master to Peter branch. update figure codes. delete old figure code in text folder

@@
 Red signifies pathway disrupted pathways, grey signifies not disrupted."}
-
 ## top100 pathways pancancer oncoprint

### 2020-07-15 code/figure1.Rmd

@@
 
-## violin mutation count
-fig1G <- ggplot(umap_cancertype, aes(x = knn_clust, y = mut_count)) +
-    geom_violin(aes(fill = knn_clust)) + #draw_quantiles = c(0.25, 0.5, 0.75)) 
-    geom_boxplot(aes(fill = knn_clust), width=0.2) +
+## mutation burden plot
+mut_order <- list() ## store data in list 
+for(n in 1:max(as.numeric(as.character(umap_cancertype$knn_clust)))) { ## loop through all the clusters
+  message(n)
+  mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
+    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
+    arrange(mut_count) ## arrange mutation in increasing order
+}
+mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
+
+mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
+mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
+
+## Need the breaks positions for the x-axis
+previous_class <- 0 ## starting position 1
+position_of_previous_class <- 0 ## starting position 2
+xaxis_breaks <- list() ## list for for loop
+for(n in 1) { ## loop through 1 to 12 clusters
+  message(n)
+  class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
+  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+  previous_class <- class_spec ## storing class halfway mark
+  for(n in 2:max(as.numeric(as.character(mut_order$knn_clust)))) {
+    message(n)
+    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
+    class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
+    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+    previous_class <- class_spec ## storing class halfway mark
+  }
+}
+xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
+
+total_n <- as.data.frame(table(mut_order$knn_clust)) ## getting total n in cluster
+
+mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count
+
+mut_median <- mut_order %>% 
+  group_by(knn_clust) %>% 
+  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
+
+fig1G <- ggplot(mut_order, aes(x = V1, y = log10mut, color = knn_clust)) + 
+  geom_point() + 
+  ylab("Log10(Mutation Count)") + 
+  xlab("") +
   theme_classic() +
-    scale_y_log10() +
-    xlab("") +
-    ylab("Log(Mutation Count)") +
-    labs(color = "", tag = "G") +
+  facet_wrap(~ factor(knn_clust, levels = c(1:12)), ncol = 12, scales = "free_x") +
+  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
+  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
+  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
   theme(axis.text.x = element_text(size = 12),
@@
         axis.title.y = element_text(size = 12),
+        axis.ticks.x = element_blank(),
+        axis.line.x = element_line(colour = "black", size = 0.5),
+        axis.line.x.top = element_line(colour = "black", size = 0.5),
+        #strip.background.x = element_blank(),
+        strip.placement = "outside",
+        strip.text.x = element_text(size = 12),
         legend.position = "none") +
-  scale_fill_manual(values = kelly.colours[-c(1:2)])
+  scale_color_manual(values = kelly.colours[-c(1:2)])
+
+## code to keep only the bottom border line of the facet
+fig1G <- ggplotGrob(fig1G)
+lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
+                   gp=gpar(col="black", lwd=4))
+
+for (k in grep("strip-t",fig1G$layout$name)) {
+  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
+}
+
+fig1G <- as.ggplot(fig1G)
 
@@
 load("../data/gene_data.Rda")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
 brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
@@
   t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
+  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
   mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
@@
   dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
   left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
@@
 AAAAEEEFFF
-AAAAGGGG##
-AAAAGGGG##
+AAAAGGGGGG
+AAAAGGGGGG
 AAAAHHIIJJ
@@
 
-```
-
-```{r mutation-burden-plot, message = FALSE}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-mut_order <- list() ## store data in list 
-for(n in 1:max(umap_cancertype$knn_clust)) { ## loop through all the clusters
-  message(n)
-  mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
-    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
-    arrange(mut_count) ## arrange mutation in increasing order
-}
-mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
-
-mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
-mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
-
-## Need the breaks positions for the x-axis
-previous_class <- 0 ## starting position 1
-position_of_previous_class <- 0 ## starting position 2
-xaxis_breaks <- list() ## list for for loop
-for(n in 1) { ## loop through 1 to 12 clusters
-  message(n)
-  class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
-  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-  previous_class <- class_spec ## storing class halfway mark
-  for(n in 2:max(as.numeric(as.character(mut_order$knn_clust)))) {
-    message(n)
-    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
-    class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
-    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-    previous_class <- class_spec ## storing class halfway mark
-  }
-}
-xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
-
-total_n <- as.data.frame(table(mut_order$knn_clust))
-
-## for making dup discrete x-axis
-guide_axis_label_trans <- function(label_trans = identity, ...) {
-  axis_guide <- guide_axis(...)
-  axis_guide$label_trans <- rlang::as_function(label_trans)
-  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
-  axis_guide
-}
-
-guide_train.guide_axis_trans <- function(x, ...) {
-  trained <- NextMethod()
-  trained$key$.label <- x$label_trans(trained$key$.label)
-  trained
-}
-
-ggplot(mut_order, aes(x = V1, y = log10(mut_count + 1), color = knn_clust)) + 
-  geom_point() + 
-  ylab("Log10(Mutation Count)") + 
-  xlab("") +
-  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("Class =" ,1:12)) +
-  guides(x.sec = guide_axis_label_trans(~paste(c("n = ", total_n$Freq)))) + ## create second x-axis
-  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  #scale_y_log10() +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        legend.position = "none") +
-  scale_color_manual(values = kelly.colours[-c(1:2)])
 ```

### 2020-07-15 code/figure4.Rmd

@@
 The highlighted genes at right are all proteasomal subunits.", eval=FALSE}
-
 load("../data/class67_gene_2_path_heatmap.Rda") ## load data

### 2020-07-16 code/figure1.Rmd
edits to figure 1-4. added function for fig4. adjusted table1 to show top largest n

@@
   xlab("") +
+  labs(tag = "G") +
   theme_classic() +

### 2020-07-16 code/figure3.Rmd
edits to figure 1-4. added function for fig4. adjusted table1 to show top largest n

@@
 
-```{r group-pathway-enrichment heatmap, fig.cap="**Figure 3.
+```{r group-pathway-enrichment heatmap, fig.width= 5, fig.height= 10, fig.align= 'center', fig.cap="**Figure 3.
 Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
@@
 load("../data/effect_heatmap.Rda") ## load data
-
+effect_heatmap$test <- 1
+class(effect_heatmap$cluster)
 effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
@@
   scale_fill_manual(values = effect_colors) +
-  theme(axis.text.x = element_text(size = 12),
+  ylab("377 Reactome Pathways") +
+  xlab("Cluster") +
+  theme(axis.text.x = element_text(size = 10),
         axis.text.y = element_blank(),
@@
         legend.position = "none")
-
 fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
@@
 
-## setting patchwork panel layout
-layout <- '
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-AAAAA
-#BBB#
-'
+# ## setting patchwork panel layout
+# layout <- '
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# AAAAA
+# #BBB#
+# '
+# 
+# fig3 + fig3_legend + plot_layout(design = layout)
 
-fig3 + fig3_legend + plot_layout(design = layout)
+fig3_legend <- plot_grid(NULL, fig3_legend, NULL, ncol = 3)
+
+plot_grid(fig3, fig3_legend, ncol = 1, rel_heights = c(15, 1))
 ```

### 2020-07-16 code/figure4_function.Rmd
edits to figure 1-4. added function for fig4. adjusted table1 to show top largest n

```{r figure-4-data-generation_function}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/moderate_enrich_prob.Rda")
load("../data/gene_data.Rda")
# load("../data/gdc_tumor_coding_complete.Rdata")
# 
# missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
# splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
# frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
# frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
# utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
# utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
# nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
# nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
# 
# gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
# rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
# gc()
# gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
# save(gdc_mut_count_patient, file = "../data/gdc_mut_count_patient_booleanized.Rda")
load("../data/gdc_mut_count_patient_booleanized.Rda")
filtered_class <- 3
n_path <- 10
enrich_cutoff <- 0.30
oncoprint_style <- function(filtered_class, n_path, enrich_cutoff = 0.30) {
  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
  toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
  filter(clust %in% filtered_class) %>% ## filter the cluster
  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] %>% na.omit() ## getting the top10 enriched pathways
  if(length(toppath) == 0) { ## if no enriched pathways. stop and print error message
    stop("No enriched pathways... select a lower enrich_cutoff")
  }
  if(length(toppath) == 1) { ## if there is only 1 enriched path. stop and print error message
    stop("1 enriched pathway found. Use oncoprint to do your analysis")
  }
  genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
  .[rowSums(.[, -1])>0,]

  make_combinations <- function(x) { ## function to make all the possible combination of genes in pathway
  l <- length(x)
  mylist <- lapply(1:l, function(y) {
    combn(x, y, simplify = FALSE)
  })
  mylist
  }
  path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10

  top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
    'colnames<-' ("sum_top10") %>% ## rename col
    rownames_to_column(var = "gene_name") ## make rownames a col

  path_match_in_set <- list() ## making a list for matchings in sets

  for(n in 1:length(path_combo)) {
    message(n)
    path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
    path_match_within_set <- list() ## making a list for matching within sets
    for(p in 1:length(path_combo_set)) {
      message(p)
      path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
      spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
      if(length(spec_combo) > 1) {
        spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
      }
      if(length(spec_combo) == 1) {
      names(spec_combo) <- "combo"
      }
      spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
      spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
      any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
      if(!"TRUE" %in% any_matches) { ## if no matches then skip
        next
        }
      if("TRUE" %in% any_matches) { ## if are matches then do this
        path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
      }
   }    
    if(length(path_match_within_set) > 0) {
      path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
    }
  }
  path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
  ### getting clustering order from pheatmap
  most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
    'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
    rownames_to_column(var = "order")

  gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
  arrange(as.numeric(order)) %>% .$gene_name ## getting the gene name order
  ## gene mutation data and tumor samples - Booleanize mutation count for each tumor sample
  message("working on mutation data")
  filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
    left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
    column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
    .[rownames(.) %in% gene_order, names(.) %in% class_spec] %>% ## filter by genes in top10 path of filtered class and tumor samples associated with it
    .[match(gene_order, rownames(.)),]
  message("Running pheatmap to get coordinates")
  pheatmap_res <- pheatmap(filtered_mut_data,
         show_rownames = T,
         show_colnames = F,
         treeheight_row = 0,
         treeheight_col = 0,
         cellheight = 5,
         cellwidth = 1,
         cluster_cols = T,
         cluster_rows = F, ## turn off row clustering
         annotation_legend = F,
         legend_breaks = c(0, 0.25, 0.75, 1),
         legend_labels = c("","Not Mut", "Mut", ""),
         color = c("grey", "firebrick3"),
         clustering_method = "ward.D2",
         fontsize = 10,
         legend = T)
  dev.off()
  pheatmap_order <- pheatmap_res$tree_col$order ## getting sample clustering order
  sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
  sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
  message("Preparing data for plot")
  gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
  gene_2_path_heatmap$membership <- as.character(gene_2_path_heatmap$membership) ## turn boolean values into categories
  ## set levels based on clustering algorithm above
  gene_2_path_heatmap <- left_join(gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
  gene_2_path_heatmap$gene_name <- factor(gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
  gene_2_path_heatmap$sample_id <- factor(gene_2_path_heatmap$sample_id, levels = c(sample_order))
  heatmap_ggplot <- ggplot(data = gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none") +
    ggtitle(paste0("Class ", filtered_class))
  ## left annotatoin data generation
  l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap

  for(p in toppath){
    message(p)
    l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
    names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
    ## for any path_combo that includes the pathway name change the 0 into a 1
    if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
    }
    if(p == "Signaling by Type 1 Insulin-like Growth Factor 1 Receptor (IGF1R)") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "Signaling by Type 1 Insulin", x = l_anno$path_combo), p] <- 1
    }
    if(p == "Signaling by the B Cell Receptor (BCR)") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "Signaling by the B Cell Receptor", x = l_anno$path_combo), p] <- 1
    }
    if(p == "Erythropoietin activates Phosphoinositide-3-kinase (PI3K)") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "Erythropoietin activates Phosphoinositide-3-kinase", x = l_anno$path_combo), p] <- 1
    }
    if(p == "Nonsense Mediated Decay (NMD) independent of the Exon Junction Complex (EJC)") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "independent of the Exon Junction Complex", x = l_anno$path_combo), p] <- 1
    }
    if(p == "Nonsense Mediated Decay (NMD) enhanced by the Exon Junction Complex (EJC)") { ## if it has this pathway grep a different way
     l_anno[grep(pattern = "enhanced by the Exon Junction Complex", x = l_anno$path_combo), p] <- 1
    }
    if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA" & p != "Signaling by Type 1 Insulin-like Growth Factor 1 Receptor (IGF1R)" & p != "Signaling by the B Cell Receptor (BCR)" & p != "Erythropoietin activates Phosphoinositide-3-kinase (PI3K)" & p != "Nonsense Mediated Decay (NMD) independent of the Exon Junction Complex (EJC)" & p != "Nonsense Mediated Decay (NMD) enhanced by the Exon Junction Complex (EJC)") { ## if it is not AUF1 do it normally
    l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
    }
  }

  l_anno_ggplot <- list() ## list for storing ggplots
  path_legend <- list()

  kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
                     "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
                     "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
                     "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")

  l_anno$gene_name <- factor(l_anno$gene_name, levels = c(rev(gene_order))) ##factor gene_name
  for(n in 1:length(toppath)){
    message(n)
    p <- toppath[n]
    message(p)
    spec_path <- l_anno[,c("gene_name", p)]
    names(spec_path)[names(spec_path) == p] <- "p"
    anno_color <- c("0" = "white", "1" = kelly.colours[2+n])
    path_ggplot <- ggplot(spec_path) + ## making plot for left annotation
    geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
            stat = "identity", 
            width = 1) +
    scale_fill_manual(values = anno_color, breaks = "1", name = "", labels = p) +
    theme_void() 
    path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
    l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
  }

  plotlist <- c(l_anno_ggplot, list(heatmap_ggplot)) ## make a list of plots for cowplot

  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 15))
plot_legend <- plot_grid(plotlist = path_legend, ncol = 5)

return(plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5)))
}

# class2 <- oncoprint_style(filtered_class = 2, n_path = 10)
# class2
# class3 <- oncoprint_style(filtered_class = 3, n_path = 10)
# class3
# class4 <- oncoprint_style(filtered_class = 4, n_path = 10)
# class4
# class5 <- oncoprint_style(filtered_class = 5, n_path = 10)
# class5
# class6 <- oncoprint_style(filtered_class = 6, n_path = 10)
# class6
# class7 <- oncoprint_style(filtered_class = 7, n_path = 10)
# class7
# class9 <- oncoprint_style(filtered_class = 9, n_path = 10)
# class9
# class10 <- oncoprint_style(filtered_class = 10, n_path = 10)
# class10
# class11 <- oncoprint_style(filtered_class = 11, n_path = 10)
# class11
# class12 <- oncoprint_style(filtered_class = 12, n_path = 10)
# class12

```

### 2020-07-16 code/figure4.Rmd

--- 2020-07-15 code/figure4.Rmd
@@
 
+```{r child = '../code/figure4_function.Rmd'}
+
+```
+
+```{r figure-4-oncoprint, message = FALSE, fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
+The genes are organized according to which of the top distinguishing pathways for this class that they effect.
+Some genes affect multiple pathways, as indicated by the bars at left.
+The highlighted genes at right are all proteasomal subunits."}
+
+class67 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
+class67
+```
+
 ```{r figure-4, message = FALSE, fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
@@
 Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.", eval=FALSE}
-
+The highlighted genes at right are all proteasomal subunits.", eval = FALSE}
 load("../data/class67_gene_2_path_heatmap.Rda") ## load data

### 2020-07-16 code/table1.Rmd

--- 2020-07-14 text/table_1.Rmd
@@
 load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+
+## Selecting the top 10 cancers with the largest n for table 1
+n_tumor <- as.data.frame(table(tumor_clusters$project_code))
+n_tumor <- n_tumor %>% arrange(desc(Freq)) %>% .[1:10, "Var1"] %>% as.character() %>% .[order(.)]
+
+cluster_spec_enrich_table1 %>% .[, c("Pathway", "Class", n_tumor)] %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%

### 2020-07-16 text/discussion.Rmd
add discussion of Hahahan and Weinberg essays to discussion, update references

@@
 
-The basic problem in tumor research of this kind, namely that the search space,
-~20,000 protein-coding genes is greater than the number of tumor samples
+One of the biggest technical hurdles in tumor research is that the search space
+comprising ~20,000 protein-coding genes is comparable with the number of tumor
+samples, even with multiple mutations per sample
 [@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
 that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
-elements exist with oncogenic or tumor suppressor potential. Hence, we sought to
+elements exist with oncogenic or tumor suppressor potential in spite of many
+efforts detect non-random or selected mutation above the noise[@deesMuSiC2012;
+@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
+@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
+@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
 investigate cancer as a disease of basic cellular and biochemical pathways,
 which we accomplished by translating gene-level mutations into pathway level
-disruptions. Importantly, we selectively removed pathways that lacked generality
-as cellular processes or that could be specifically assigned to differentiated
-cell types, while excluding others.
+disruptions.
 
-We hypothesize that cancer is the result of dysfunction from a limited number of
-basic cellular processes that are common to all living human cells, and thus all
-tumors belong on a spectrum of dysfunction on this axis regardless of tissue of
-origin. An alternative hypothesis is that every tumor belongs to one of a large
-number of syndromes which are unique to each tissue of origin, that share some
-mechanisms and superficial similarity in treatment strategies. The latter idea
-has gained momentum in recent years since the publication of Encode, Roadmap and
-TCGA consortium papers, perhaps owing to the intractable complexity of genomic
-data in all its forms. The inescapable conclusion thus far is that all these
-data point to tissue-of-origin as the most important driver of tumor behavior.
-Our observations contrast with this view, and instead support an interpretation
-in which all tumors manifest one of a limited number of metastable phenotypes
-represented by our classes (**figure 1a-k**), with increased fitness regardless
-of tissue of origin, which then converge on end stage cancer by accumulating
-complementary mutations to the other classes (**figure 1l** and **Table 2**).
+The hypothesis that cancer is the result of dysfunction from a limited number of
+basic cellular processes common to all living human cells was first
+introduced and later expanded on in a pair of classic essays by Hanahan and
+Weinberg [@weinberg200; @hallmarks_II]. In this view all tumors, regardless of
+tissue of origin, acquire 6 essential phenotypes; sustaining proliferative
+signaling, evading growth suppressors, replicative immortality, evading
+apoptosis, supportive angiogenesis, and tissue invasion and metastasis
+[@weinberg2000]. Advancements in the field over the ensuing decade led to the
+inclusion of new tumor promoting processes, namely deregulation of celullar
+energy cycle, tumor promoting inflammation, evasion of the immune system, genome
+instability, and on top of these acquired traits a new appreciation for the
+concept of the tumor microenvironment and its contribution to the evolution and
+promotion of metastatic tumorigenesis [@hallmarks_II].
+
+An alternative hypothesis is that every tumor belongs to one of a large number
+of syndromes which are unique to each tissue-of-origin, that share some
+mechanisms and treatment strategies. Recent publication of Encode, Roadmap and TCGA
+consortium papers present a view largely consistent with the latter
+hypothesis. Perhaps owing to the intractable complexity of genomic data in all
+its forms, the inescapable conclusion thus far is that tissue-of-origin remains
+the most important driver of tumor characteristics at every scale and by every
+technique. Our observations contrast with this view, and instead support an
+interpretation of publicly available  data in which all tumors manifest one of a
+limited number of metastable phenotypes resulting from disruptions in basic
+pathways represented by our classes (**figure 1a-k**), with increased fitness
+regardless of tissue of origin, which then converge on end stage cancer by
+accumulating complementary mutations to the other classes (**figure 1l** and
+**Table 2**).
 
@@
 arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
-see **supplementary figure S1**) are unevenly distributed toward a specific
-class, we could not identify cancer type that was exclusive to a single tumor
-class. Somewhat surprisingly to us, this finding extends to histological
-subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
-and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
-observe bias for some cancers, as it has long been known that certain treatment
-regimens are likely to work for a majority of patients of a given cancer type.
-Nonetheless, the basis for some tumors being treatment-refractory in spite of
-clinical indications remains elusive. Our analysis suggests the possibility of
-more general mechanisms underlying tumor evolution regardless of tissue,
-histological or cellular subtype and could form a basis for future separation of
-patients into different precision treatment regimens.
+see **supplementary figure S1**) are unevenly distributed, though we could not
+identify cancer type that was exclusive to a single tumor class. Somewhat
+surprisingly to us, this finding extends to histological subtypes of breast,
+head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental
+Figure S2**). Indeed, it would be surprising if we did not observe bias for some
+cancers, as it has long been known that certain treatment regimens are likely to
+work for a significant fraction of patients of a given cancer type. Nonetheless,
+the basis for some tumors being treatment-refractory in spite of clinical
+indications remains elusive. Our analysis suggests the possibility of
+identifying more informative molecular, histological or cellular subtypes and
+could form a basis for future separation of patients into different precision
+treatment regimens.
 
@@
 
-Finally, we must remark on two of the classes, 1 and 8, which for which we did
-not find distinctive associations with pathways. Class 1 has a relatively low
-level of mutated pathways overall, although it is broadly enriched in many of
-the same tumor-promoting pathways common to the other groups. This likely does
-not reflect a true difference in stages of tumor evolution, as our data show
-clearly that this class is as likely to contain stage IV as it is to contain
-stage I tumors. We speculate this class represents a group of tumors with large
-aberrations in methylation, copy number or other genome rearrangements which we
-excluded from the present analysis. In the future, we hope to incorporate these
-other data into a pathway-centric analysis as we have done here for point
-mutations and indels. Class 8 represents a group of hyper-mutated tumors that
-harbor so many mutations that virtually no pathway is unaffected. This group
-highlights one of the limitations of the approach that we’ve taken, in that it
-is likely that a significant fraction of the samples flagged as “mutant” for
-each pathway are actually unaffected, consistent with those samples have low
-information content. Similar to class 1 we speculate that a future analysis of
-other data types, including improved filtering of mutations, may resolve these
-tumors into more meaningful categories.
-
 ### Incompleteness of the pathway disruption data
 
-The fact that we are likely excluding many defects (under chromosomal
-abberations, amplifications, demethylation and so on) also raises the
-possibility that our classes are incomplete or may change at a future time when
-we can account for them. For now at least, we think that the point mutation data
-are representative of the most important pathways.
+Finally, we must remark on the limitations of our work exemplified in two of the
+classes, 1 and 8, which for which we did not find distinctive associations with
+pathways. Class 1 has a relatively low level of mutated pathways overall,
+although it is broadly enriched in many of the same tumor-promoting pathways
+common to the other groups. This likely does not reflect a true difference in
+stages of tumor evolution, as our data show clearly that this class is as likely
+to contain stage IV metastatic tumors as it is to contain those of stage I. We
+speculate this class represents a group of tumors with large aberrations in
+methylation, copy number or other genome rearrangements which we excluded from
+the present analysis. In the future, we hope to incorporate these other data
+into a pathway-centric analysis as we have done here for point mutations and
+indels. Now that we have established this as a feasible and potentially fruitful
+approach to understanding the cancer genome, it may be worth expanded effort in
+this area. In addition, Class 8 represents a group of hyper-mutated tumors that
+harbor so many mutations that virtually no pathway is unaffected. It seems
+likely that a significant fraction of the samples flagged as “mutant” for each
+pathway are actually unaffected, consistent with those samples have low
+information content. Similar to class 1 we speculate that a future analysis of
+other data types, including improved filtering of mutations across all classes,
+may resolve these tumors into improved classifications.
 
@@
 mutations and metastasis. General cancer promoting processes must include
-factors that relate to genome stability either directly or indirectly. Such
-pathways are disrupted in all or most of the defined classes (see **figures 2**
-and **supplementary figure S1**) and often are the result of abberations
-involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic
-checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and
-**supplemental file 1**). Class-specific evolution must involve the acquisition
-of combinatorial disruptions to pathways that may individually be harmful but
-together produce more specialized cancer phenotypes and increased fitness (as in
-**figure 4** and **tables 1** and **ST2**). Importantly, our work does not
-distinguish between the order in which these mutations should accumulate.
-However, since many of the genes in the non-specific pathways involve genome
-stability, it is reasonable to hypothesize that such mutations promote the
-acquisition of class specific defects, and natural selection produces the
-classes we observed.
+factors that relate to genome stability and immortality, as "enabling
+characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
+disrupted in all or most of the defined classes (see **figures 2** and
+**supplementary figure S1**) and often are the result of abberations involving
+common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
+complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
+Class-specific evolution must involve the acquisition of combinatorial
+disruptions to pathways that may individually be harmful but together produce
+more specialized cancer phenotypes and increased fitness (as in **figure 4** and
+**tables 1** and **ST2**). Importantly, our work does not distinguish between
+the order in which these mutations should accumulate. However, since many of the
+genes in the non-specific pathways involve genome stability, it is reasonable to
+hypothesize that such mutations promote the acquisition of class specific
+defects, and natural selection produces the classes we observed.
+
+One of the drawbacks of bulk tumor whole genome sequencing data is the problem
+of tumor heterogeneity, both in terms of evolving subclones with different or
+passively diverging mutation profiles which then serve as a substrate for futher
+selection. Moreover consortium samples are likely to contain contamination from
+support tissue and inflammatory and immune cells of the innate and adaptive
+immune systems, all potentially harboring cancer supporting mutations
+[@bhowmick_review]. We think it will be instructive to explore these ideas in
+the context of single cell experiments.
 

### 2020-07-16 text/discussion.Rmd
update discussion of metastasis, add knott asparagine reference

@@
 
-The class-specific metastatic pathway enrichment we observe (in **table 2**) may
-be a result of the pathways directly promoting metastatic phenotypes, or
-alternatively that such mutations are more likely to turn metastatic because
-they acquire a stronger or more complete oncogenic phenotype, since (again) we
-don’t know the order in which they were acquired. The fact that most metastatic
-enrichment is cluster-specific and has a tendency to overlap with
-cluster-specific pathways from non-metastatic tumors of neighboring clusters
-favors the latter explanation, and suggests a mechanism of complementarity in
-which newly acquired mutations converge on a single deadly phenotype that
-captures most or all of the critical features of end-stage cancer. If this is
-the case, we speculate that once we discover the vulnerabilities of each cluster
-we might also reasonably predict which pathways are susceptible to further
-selection and therefore target them  therapeutically at the same time to prevent
-further adaptation.
+We report that pathway enrichment across all classes (pooled) yielded generally
+lower effect sizes and larger p-values than the cluster specific analysis,
+suggesting that signal is diluted when the samples are analyzed this way, and
+supporting the view that metastasis has class-specific requirements. It should
+be noted, however, that in spite of this lack of global enrichment, some of the
+cluster specific pathways reflect known metastatic associations, _e.g._ pathways
+associated with asparagine and aspartate metabolism in class 10
+[@knott_asparagus]. Since the number of metastatic tumors is relatively low this
+part of our analysis is likely underpowered and subject to expanded analysis
+with larger cohorts. Nonetheless, the class-specific metastatic pathway
+enrichment we observe (in **table 2**) could be a result of the pathways
+directly promoting metastatic phenotypes in these classes, or because they
+acquire a stronger or more complete oncogenic phenotype, since (again) we don’t
+know the order in which they were acquired in individual tumor samples. The fact
+that most metastatic enrichment is cluster-specific and has a tendency to
+overlap with cluster-specific pathways from non-metastatic tumors of neighboring
+clusters favors the latter interpretation, and suggests a mechanism of
+complementarity, in which newly acquired mutations result in similar classes
+converging on a single deadly phenotype that captures most or all of the
+critical features of end-stage cancer. If this is the case, once we discover the
+requirements of each cluster with respect to metastasis we can target them
+therapeutically and prevent further adaptation.

### 2020-07-20 code/figure1.Rmd
minor updates on fig 3, 4, and supplemental - changes to plot size, labels, and plot scaling on html

--- 2020-07-15 code/figure1.Rmd
@@
 fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
-  geom_point(colour = "grey") +
+  geom_point(colour = "grey", size = 0.25) +
   xlab("umap_1") +
@@
 fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
   guides(color = F) +
@@
 fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
-  geom_point() + scale_color_manual(values = kelly.colours[-c(1:2)]) +
+  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
   guides(color = F) +
@@
 fig1G <- ggplot(mut_order, aes(x = V1, y = log10mut, color = knn_clust)) + 
-  geom_point() + 
+  geom_point(size = 0.25) + 
   ylab("Log10(Mutation Count)") + 
   xlab("") +
+  labs(tag = "G") +
   theme_classic() +
@@
   scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
-  theme(axis.text.x = element_text(size = 12),
-        axis.text.y = element_text(size = 12),
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
         axis.title.x = element_text(size = 12),
@@
 fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) + 
   theme_classic() +
@@
 fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() +
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
   theme_classic() +
@@
 fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
+  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) + 
   theme_classic() +
@@
 fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point() + 
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) + 
   theme_classic() +
@@
 fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey") +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm)) + 
+  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
   ylab("umap_2") +

### 2020-07-20 code/figure4.Rmd
small update to figure4. trying to get ggplot to show up
minor updates on fig 3, 4, and supplemental - changes to plot size, labels, and plot scaling on html

@@
 
-```{r child = '../code/figure4_function.Rmd'}
-
-```
-
 ```{r figure-4-oncoprint, message = FALSE, fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
@@
 
-class67 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
-class67
+tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+load("../data/gene2_level2_lookup.Rda")
+load("../data/pathname2level2.Rda")
+load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/moderate_enrich_prob.Rda")
+load("../data/gene_data.Rda")
+# load("../data/gdc_tumor_coding_complete.Rdata")
+# 
+# missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
+# splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+# frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+# frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+# utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+# utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+# nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+# nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+# 
+# gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+# rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
+# gc()
+# gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
+# save(gdc_mut_count_patient, file = "../data/gdc_mut_count_patient_booleanized.Rda")
+load("../data/gdc_mut_count_patient_booleanized.Rda")
+filtered_class <- 3
+n_path <- 10
+enrich_cutoff <- 0.30
+oncoprint_style <- function(filtered_class, n_path, enrich_cutoff = 0.30) {
+  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
+  toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
+  filter(clust %in% filtered_class) %>% ## filter the cluster
+  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
+  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] %>% na.omit() ## getting the top10 enriched pathways
+  if(length(toppath) == 0) { ## if no enriched pathways. stop and print error message
+    stop("No enriched pathways... select a lower enrich_cutoff")
+  }
+  if(length(toppath) == 1) { ## if there is only 1 enriched path. stop and print error message
+    stop("1 enriched pathway found. Use oncoprint to do your analysis")
+  }
+  genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
+  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
+  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
+  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
+  .[rowSums(.[, -1])>0,]
+
+  make_combinations <- function(x) { ## function to make all the possible combination of genes in pathway
+  l <- length(x)
+  mylist <- lapply(1:l, function(y) {
+    combn(x, y, simplify = FALSE)
+  })
+  mylist
+  }
+  path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
+
+  top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
+    'colnames<-' ("sum_top10") %>% ## rename col
+    rownames_to_column(var = "gene_name") ## make rownames a col
+
+  path_match_in_set <- list() ## making a list for matchings in sets
+
+  for(n in 1:length(path_combo)) {
+    message(n)
+    path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
+    path_match_within_set <- list() ## making a list for matching within sets
+    for(p in 1:length(path_combo_set)) {
+      message(p)
+      path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
+      spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
+      if(length(spec_combo) > 1) {
+        spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
+        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
+      }
+      if(length(spec_combo) == 1) {
+      names(spec_combo) <- "combo"
+      }
+      spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
+      spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
+      any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
+      if(!"TRUE" %in% any_matches) { ## if no matches then skip
+        next
+        }
+      if("TRUE" %in% any_matches) { ## if are matches then do this
+        path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
+        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
+      }
+   }    
+    if(length(path_match_within_set) > 0) {
+      path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
+    }
+  }
+  path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
+  ### getting clustering order from pheatmap
+  most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
+    'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
+    rownames_to_column(var = "order")
+
+  gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
+  arrange(as.numeric(order)) %>% .$gene_name ## getting the gene name order
+  ## gene mutation data and tumor samples - Booleanize mutation count for each tumor sample
+  message("working on mutation data")
+  filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
+    left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
+    column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
+    .[rownames(.) %in% gene_order, names(.) %in% class_spec] %>% ## filter by genes in top10 path of filtered class and tumor samples associated with it
+    .[match(gene_order, rownames(.)),]
+  message("Running pheatmap to get coordinates")
+  pheatmap_res <- pheatmap(filtered_mut_data,
+         show_rownames = T,
+         show_colnames = F,
+         treeheight_row = 0,
+         treeheight_col = 0,
+         cellheight = 5,
+         cellwidth = 1,
+         cluster_cols = T,
+         cluster_rows = F, ## turn off row clustering
+         annotation_legend = F,
+         legend_breaks = c(0, 0.25, 0.75, 1),
+         legend_labels = c("","Not Mut", "Mut", ""),
+         color = c("grey", "firebrick3"),
+         clustering_method = "ward.D2",
+         fontsize = 10,
+         legend = T)
+  dev.off()
+  pheatmap_order <- pheatmap_res$tree_col$order ## getting sample clustering order
+  sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
+  sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
+  message("Preparing data for plot")
+  gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
+  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
+  gene_2_path_heatmap$membership <- as.character(gene_2_path_heatmap$membership) ## turn boolean values into categories
+  ## set levels based on clustering algorithm above
+  gene_2_path_heatmap <- left_join(gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
+  gene_2_path_heatmap$gene_name <- factor(gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
+  gene_2_path_heatmap$sample_id <- factor(gene_2_path_heatmap$sample_id, levels = c(sample_order))
+  heatmap_ggplot <- ggplot(data = gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
+  geom_tile(size = 0.6) +
+  scale_fill_manual(values = c("grey", "firebrick3")) +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_blank(),
+        axis.ticks = element_blank(),
+        strip.text.y = element_blank(),
+        strip.background = element_blank(),
+        legend.position = "none") +
+    ggtitle(paste0("Class ", filtered_class))
+  ## left annotatoin data generation
+  l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
+
+  for(p in toppath){
+    message(p)
+    l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
+    names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
+    ## for any path_combo that includes the pathway name change the 0 into a 1
+    if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
+     l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
+    }
+    if(p == "Signaling by Type 1 Insulin-like Growth Factor 1 Receptor (IGF1R)") { ## if it has this pathway grep a different way
+     l_anno[grep(pattern = "Signaling by Type 1 Insulin", x = l_anno$path_combo), p] <- 1
+    }
+    if(p == "Signaling by the B Cell Receptor (BCR)") { ## if it has this pathway grep a different way
+     l_anno[grep(pattern = "Signaling by the B Cell Receptor", x = l_anno$path_combo), p] <- 1
+    }
+    if(p == "Erythropoietin activates Phosphoinositide-3-kinase (PI3K)") { ## if it has this pathway grep a different way
+     l_anno[grep(pattern = "Erythropoietin activates Phosphoinositide-3-kinase", x = l_anno$path_combo), p] <- 1
+    }
+    if(p == "Nonsense Mediated Decay (NMD) independent of the Exon Junction Complex (EJC)") { ## if it has this pathway grep a different way
+     l_anno[grep(pattern = "independent of the Exon Junction Complex", x = l_anno$path_combo), p] <- 1
+    }
+    if(p == "Nonsense Mediated Decay (NMD) enhanced by the Exon Junction Complex (EJC)") { ## if it has this pathway grep a different way
+     l_anno[grep(pattern = "enhanced by the Exon Junction Complex", x = l_anno$path_combo), p] <- 1
+    }
+    if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA" & p != "Signaling by Type 1 Insulin-like Growth Factor 1 Receptor (IGF1R)" & p != "Signaling by the B Cell Receptor (BCR)" & p != "Erythropoietin activates Phosphoinositide-3-kinase (PI3K)" & p != "Nonsense Mediated Decay (NMD) independent of the Exon Junction Complex (EJC)" & p != "Nonsense Mediated Decay (NMD) enhanced by the Exon Junction Complex (EJC)") { ## if it is not AUF1 do it normally
+    l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
+    }
+  }
+
+  l_anno_ggplot <- list() ## list for storing ggplots
+  path_legend <- list()
+
+  kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                     "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                     "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                     "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
+  l_anno$gene_name <- factor(l_anno$gene_name, levels = c(rev(gene_order))) ##factor gene_name
+  for(n in 1:length(toppath)){
+    message(n)
+    p <- toppath[n]
+    message(p)
+    spec_path <- l_anno[,c("gene_name", p)]
+    names(spec_path)[names(spec_path) == p] <- "p"
+    anno_color <- c("0" = "white", "1" = kelly.colours[2+n])
+    path_ggplot <- ggplot(spec_path) + ## making plot for left annotation
+    geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
+            stat = "identity", 
+            width = 1) +
+    scale_fill_manual(values = anno_color, breaks = "1", name = "", labels = p) +
+    theme_void() 
+    path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
+    l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
+  }
+
+  plotlist <- c(l_anno_ggplot, list(heatmap_ggplot)) ## make a list of plots for cowplot
+
+  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 15))
+plot_legend <- plot_grid(plotlist = path_legend, ncol = 5)
+
+return(plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5)))
+}
+
+fig4 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
+fig4
+
 ```

### 2020-07-20 code/supplements.Rmd
minor updates on fig 3, 4, and supplemental - changes to plot size, labels, and plot scaling on html

@@
 ### Supplementary Figure S1
+```{r}
+library(data.table)
+library(ggplot2)
+library(kableExtra)
+library(tidyverse)
+```
 
-```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
+```{r supplementary-figure-s1, out.width = '90%', fig.align='center'}
 umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
@@
 ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 1) +
-  geom_point(size = 1) +
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
   theme_classic() +
@@
 
-```{r supplementary-figure-s2-data-generation}
+```{r supplementary-figure-s2-data-generation, out.width = '90%', fig.align = "center"}
 umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
@@
 ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
   theme_classic() +
@@
 
-```{r supplementary-figure-s3}
+```{r supplementary-figure-s3, out.width='77%', fig.align = "center"}
 load("../data/selected_mut.Rda")
@@
 ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
+  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
   theme_classic() +
@@
 
-save(selected_path, file = "selected_path.Rda")
-```
+save(selected_path, file = "../data/selected_path.Rda")
 
-```{r supplementary-figure-s4, out.width = '90%'}
 load("../data/selected_path.Rda") ## load data
@@
 
-ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-  geom_point(size = 0.5) +
+## creating layout to plot all 377 pathways
+figs4_layout <- data.frame(start = seq(from = 1, to = length(toppath), by = 25),
+                          end = c(seq(from = 25, to = length(toppath), by = 25), length(toppath)))
+
+for(n in 1:nrow(figs4_layout)) {
+  message(n)
+  pathway_layout <- figs4_layout %>% .[n,] ## getting the layout for selecting pathways
+  pathway_layout <- toppath[pathway_layout$start:pathway_layout$end] ## filtering those pathways from toppath
+  filtered_path <- selected_path %>% filter(variable %in% pathway_layout) %>% ## filtering our data to only include those pathways
+    mutate(variable = factor(variable, levels = c(pathway_layout)))
+  ggplot_umap <- ggplot(filtered_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
+    geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.1) +
+    geom_point(size = 0.1) +
     theme_classic() +
@@
     xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
+    scale_colour_manual(values = rep("firebrick3", 25)) +
     theme(axis.text.x = element_blank(),
@@
       axis.line = element_blank(),
+      strip.text = element_text(size = 5),
       legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = toppath),
+    facet_wrap(~ variable,
                ncol = 5) ## facet by pathways
+  new_dfname <- paste0("ggplot_umap_", n)
+  assign(paste0(new_dfname), value = ggplot_umap)
+}
+
+save(ggplot_umap_1,ggplot_umap_2,ggplot_umap_3,ggplot_umap_4,ggplot_umap_5,ggplot_umap_6,ggplot_umap_7,ggplot_umap_8,ggplot_umap_9,ggplot_umap_10,ggplot_umap_11,ggplot_umap_12,ggplot_umap_13,ggplot_umap_14,ggplot_umap_15,ggplot_umap_16, file = "../data/fig_s4_ggplot.Rda")
+```
+
+```{r supplementary-figure-s4, message = FALSE, echo = FALSE, out.width = '90%', fig.align= 'center'}
+load("../data/fig_s4_ggplot.Rda")
+
+ggplot_umap_1
+ggplot_umap_2
+ggplot_umap_3
+ggplot_umap_4
+ggplot_umap_5
+ggplot_umap_6
+ggplot_umap_7
+ggplot_umap_8
+ggplot_umap_9
+ggplot_umap_10
+ggplot_umap_11
+ggplot_umap_12
+ggplot_umap_13
+ggplot_umap_14
+ggplot_umap_15
+ggplot_umap_16
+
+# ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
+#   geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
+#   geom_point(size = 0.5) +
+#   theme_classic() +
+#   coord_fixed() +
+#   xlab("UMAP 1") + ylab("UMAP 2") +
+#   scale_colour_manual(values = rep("firebrick3", 20)) +
+#   theme(axis.text.x = element_blank(),
+#     axis.text.y = element_blank(),
+#     axis.title.x = element_blank(),
+#     axis.title.y = element_blank(),
+#     axis.ticks = element_blank(),
+#     axis.line = element_blank(),
+#     legend.position = "none") +
+#   facet_wrap(~ factor(variable, levels = toppath),
+#                       ncol = 5) ## facet by pathways
 

### 2020-07-20 code/figure3.Rmd

@@
 effect_heatmap$test <- 1
-class(effect_heatmap$cluster)
+
 effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme

### 2020-07-20 code/figure4_function.Rmd

@@
-```{r figure-4-data-generation_function}
+```{r figure-4-data-generation_function, message = FALSE, eval = FALSE}
 tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
@@
 load("../data/gdc_mut_count_patient_booleanized.Rda")
-filtered_class <- 3
+filtered_class <- c(6,7)
 n_path <- 10
@@
   scale_fill_manual(values = c("grey", "firebrick3")) +
+  xlab("Tumor Samples") +
   theme(axis.text.x = element_blank(),
@@
     theme_void() 
-    path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
+    path_legend[[n]] <- get_legend(path_ggplot) ## extract legend from plot
     l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
@@
 
-  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 15))
-plot_legend <- plot_grid(plotlist = path_legend, ncol = 5)
+  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 5))
+  plot_legend <- plot_grid(plotlist = path_legend, ncol = 1, align = "v")
 
-return(plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5)))
+return(plot_grid(plot, plot_grid(NULL, plot_legend1, NULL, ncol = 1, rel_heights = c(1, 1, 1)), ncol = 2, rel_widths = c(2.5, 2), axis = "tbl"))
 }
 
-# class2 <- oncoprint_style(filtered_class = 2, n_path = 10)
-# class2
-# class3 <- oncoprint_style(filtered_class = 3, n_path = 10)
-# class3
-# class4 <- oncoprint_style(filtered_class = 4, n_path = 10)
-# class4
-# class5 <- oncoprint_style(filtered_class = 5, n_path = 10)
-# class5
-# class6 <- oncoprint_style(filtered_class = 6, n_path = 10)
-# class6
-# class7 <- oncoprint_style(filtered_class = 7, n_path = 10)
-# class7
-# class9 <- oncoprint_style(filtered_class = 9, n_path = 10)
-# class9
-# class10 <- oncoprint_style(filtered_class = 10, n_path = 10)
-# class10
-# class11 <- oncoprint_style(filtered_class = 11, n_path = 10)
-# class11
-# class12 <- oncoprint_style(filtered_class = 12, n_path = 10)
-# class12
+class67 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
+save(class67, file = "class67.Rda")
+```
 
+```{r figure-4-plot, out.width= '90%', fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
+The genes are organized according to which of the top distinguishing pathways for this class that they effect.
+Some genes affect multiple pathways, as indicated by the bars at left.
+The highlighted genes at right are all proteasomal subunits."}
+load("../data/class67.Rda")
+class67
 ```

### 2020-07-20 code/figure1.Rmd
Set up code for figure1b data  section and plotting section

@@
 ####################
+## Figure 1B data ##
+####################
+gdc_path <- gdc_reactome_path_analysis %>%
+  t() %>%
+  as.data.frame() %>%
+  rownames_to_column(var = "sample") %>% ## move path_names to column
+  .[match(rownames(tumor_samples), .$sample), ]
+
+gdc_path_mca <- left_join(gdc_path, tumor_samples[, c(1, 3)],
+  by = c("sample" = "submitted_sample_id")) %>%
+  column_to_rownames(var = "sample")
+
+gdc_path_sample_names <- rownames(gdc_path_mca)
+gdc_path_tumor_types <- gdc_path_mca$project_code.x
+
+gdc_path_mca$project_code.x <- NULL
+
+## create a categorical variable from the boolean variable
+pathway_list <- make_clean_names(colnames(gdc_path_mca))
+pathway_count <- length(pathway_list)
+gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
+  factor(x, levels = c(0, 1), labels = c("no", "yes"))
+}))
+colnames(gdc_path_mca) <- pathway_list
+gdc_path_mca$project_code <- gdc_path_tumor_types
+rownames(gdc_path_mca) <- gdc_path_sample_names
+
+## perform mca
+mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
+
+expl_var <- mca_res$eig[, 2]
+expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
+
+set.seed(4242)
+for (k in seq(100)) {
+  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
+  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
+  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
+}
+
+randomize <- tibble(
+  eigenvalue = 1:length(expl_var),
+  expl_var = expl_var,
+  perm = colMeans(expl_var_perm),
+  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
+)
+
+optimal_mca_dimensions <- randomize %>%
+  filter(pvalue < 0.05) %>%
+  filter(eigenvalue == max(eigenvalue)) %>%
+  pull(eigenvalue)
+
+mca_res <- MCA(
+  X = gdc_path_mca[, pathway_list],
+  ncp = optimal_dimensions, graph = FALSE
+)
+
+save(randomize, mca_res, file = "figure1B_data.rda")
+
+####################
 ## Figure 1C data ##
@@
  rm(fig1Ag)
+
+optimal_mca_dimensions <- randomize %>%
+  filter(pvalue < 0.05) %>%
+  filter(eigenvalue == max(eigenvalue)) %>%
+  pull(eigenvalue)
+
+permut_p <- randomize %>%
+  filter(eigenvalue < 50) %>%
+  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
+  geom_line(aes(colour = "#998ec3")) +
+  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
+  scale_color_identity(guide = guide_legend(title = ""), 
+                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
+  scale_y_continuous(labels = scales::percent) +
+  xlab("MCA Eigenvalues") +
+  ylab("explained variance") +
+  theme_classic()
+
+pvalue_p <- randomize %>%
+  filter(eigenvalue < 50) %>%
+  ggplot(aes(x = eigenvalue, y = pvalue)) +
+  geom_line() +
+  geom_point() +
+  geom_vline(xintercept = optimal_dimensions, color = "red") +
+  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
+  xlab("MCA Eigenvalues") +
+  ylab("P-value") +
+  theme_classic()
+
+fig1B <- permut_p /
+  pvalue_p + plot_annotation(title = "Significance of MCA Eigenvalues")
 

### 2020-07-20 code/figure1.Rmd
Update Figure 1C data to use uwot umap on MCA data

@@
 ####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
+rng <- RNGseq(max_n_pt * n_iter, 42)
+r <- rng[[3649]]
+
+rngtools::setRNG(r)
+gdc.umap <- uwot::umap(mca_res$ind$coord, 
+                       n_components = 3,
+                       n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
+                       n_epochs = 1000,
+                       n_threads = 1)
+gdc_umap_coors <- as.data.frame(plotting_umap) %>%
+  rename(plot_x = V1, plot_y = V2, plot_z = V3)
 

### 2020-07-20 code/figure1.Rmd
Add Figure 1D data for generating optimal minPts

@@
 ####################
+## Figure 1D data ##
+####################
+n_iter <- 40
+max_n_pt <- 200
+score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
+
+n_tumors <- nrow(mca_res$ind$coord)
+
+cl <- makeCluster(8)
+registerDoParallel(cl)
+
+for (i in 3:max_n_pt) {
+  print(paste0("MIN POINTS = ", i))
+  score_iter <- foreach(1:n_iter,
+    .combine = c,
+    .export = c("mca_res"),
+    .packages = c("dbscan", "uwot"),
+    r = rng[(i - 1) * n_iter + 1:n_iter]
+  ) %dopar% {
+    rngtools::setRNG(r)
+    umap_init_mca <- umap(mca_res$ind$coord, 
+                          min_dist = 1e-6, 
+                          n_components = 3, 
+                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
+                          n_epochs = 1000, 
+                          n_threads = 1)
+    res <- hdbscan(umap_init_mca, minPts = i)
+    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
+    return(score_iter_temp)
+  }
+  score_mca[i - 2, ] <- score_iter
+}
+
+score <- rowMeans(score_mca)
+score_df <- tibble(minPts = 3:max_n_pt, score = score)
+
+sample_df <- score_mca %>%
+  as.data.frame() %>%
+  as_tibble() %>%
+  add_column(minPts = 3:max_n_pt, .before = 1) %>%
+  pivot_longer(-minPts) %>%
+  group_by(minPts)
+
+optimal_minPts <- score_df %>%
+  filter(score == min(score)) %>%
+  pull(minPts)
+
+save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
+
+####################
 ## Figure 1E data ##

### 2020-07-20 code/figure1.Rmd
typo in Figure 1C data umap output

@@
                        n_threads = 1)
-gdc_umap_coors <- as.data.frame(plotting_umap) %>%
+gdc_umap_coors <- as.data.frame(gdc.umap) %>%
   rename(plot_x = V1, plot_y = V2, plot_z = V3)

### 2020-07-20 code/figure1.Rmd
Figure 1E data update - only umap3d coords objects change contents

@@
   X = gdc_path_mca[, pathway_list],
-  ncp = optimal_dimensions, graph = FALSE
+  ncp = optimal_mca_dimensions, graph = FALSE
 )
@@
   rename(plot_x = V1, plot_y = V2, plot_z = V3)
+rownames(gdc_umap_coors) <- rownames(gdc_path_mca)
 
@@
 ####################
-## To identify clusters using hdbscan, use umap in the third dimension
-set.seed(3296)
-gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
-gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
-rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
-hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
-umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster
+rngtools::setRNG(r)
+
+cluster_umap <- uwot::umap(mca_res$ind$coord,
+  min_dist = 1e-6,
+  n_components = 3,
+  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
+  n_epochs = 1000,
+  n_threads = 1
+)
+colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
+hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
+umap_3d_coors <- gdc_umap_coors %>%
+  bind_cols(as.data.frame(cluster_umap)) %>%
+  mutate(
+    clust = as.factor(hdb_cluster_assignment$cluster),
+    member_prob = hdb_cluster_assignment$membership_prob,
+    row.names = rownames(gdc_umap_coors)
+  )
+## clust 0 is unclassified cluster
 

### 2020-07-20 code/figure1.Rmd
Update Figure 1F data with new umap_3d_coors object

@@
 ## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
-umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
-                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
-                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
-                                                                  k=15)
+umap_3d_coors$clust_knn <- umap_3d_coors$clust
+umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
+  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
+  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
+  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
+)
 

### 2020-07-20 code/figure1.Rmd
adjust plot so legend does not bleed to adjacent plot. fixed E and F not scaled to rest of umap plots. added 50 samples with NA values in the front and back to prevent points touching y-axis

--- 2020-07-20 code/figure1.Rmd
@@
+---
+output: html_document
+editor_options: 
+  chunk_output_type: console
+---
+```{r}
+library(data.table)
+library(ggplot2)
+library(ggplotify)
+library(tidyverse)
+library(grid)
+library(patchwork)
+```
+
 <div style="text-align: right"> Generation of Somatic Mutation Data </div>
@@
 ####################
-## Figure 1B data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>%
-  t() %>%
-  as.data.frame() %>%
+## Figure 1C data ##
+####################
+gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
   rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample), ]
-
-gdc_path_mca <- left_join(gdc_path, tumor_samples[, c(1, 3)],
-  by = c("sample" = "submitted_sample_id")) %>%
-  column_to_rownames(var = "sample")
-
-gdc_path_sample_names <- rownames(gdc_path_mca)
-gdc_path_tumor_types <- gdc_path_mca$project_code.x
-
-gdc_path_mca$project_code.x <- NULL
-
-## create a categorical variable from the boolean variable
-pathway_list <- make_clean_names(colnames(gdc_path_mca))
-pathway_count <- length(pathway_list)
-gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
-  factor(x, levels = c(0, 1), labels = c("no", "yes"))
-}))
-colnames(gdc_path_mca) <- pathway_list
-gdc_path_mca$project_code <- gdc_path_tumor_types
-rownames(gdc_path_mca) <- gdc_path_sample_names
-
-## perform mca
-mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
-
-expl_var <- mca_res$eig[, 2]
-expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
-
-set.seed(4242)
-for (k in seq(100)) {
-  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
-  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
-  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
-}
-
-randomize <- tibble(
-  eigenvalue = 1:length(expl_var),
-  expl_var = expl_var,
-  perm = colMeans(expl_var_perm),
-  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
-)
-
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-mca_res <- MCA(
-  X = gdc_path_mca[, pathway_list],
-  ncp = optimal_dimensions, graph = FALSE
-)
-
-save(randomize, mca_res, file = "figure1B_data.rda")
-
-####################
-## Figure 1C data ##
-####################
-rng <- RNGseq(max_n_pt * n_iter, 42)
-r <- rng[[3649]]
-
-rngtools::setRNG(r)
-gdc.umap <- uwot::umap(mca_res$ind$coord, 
-                       n_components = 3,
-                       n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
-                       n_epochs = 1000,
-                       n_threads = 1)
-gdc_umap_coors <- as.data.frame(gdc.umap) %>%
-  rename(plot_x = V1, plot_y = V2, plot_z = V3)
-
-####################
-## Figure 1D data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
-
-n_tumors <- nrow(mca_res$ind$coord)
-
-cl <- makeCluster(8)
-registerDoParallel(cl)
-
-for (i in 3:max_n_pt) {
-  print(paste0("MIN POINTS = ", i))
-  score_iter <- foreach(1:n_iter,
-    .combine = c,
-    .export = c("mca_res"),
-    .packages = c("dbscan", "uwot"),
-    r = rng[(i - 1) * n_iter + 1:n_iter]
-  ) %dopar% {
-    rngtools::setRNG(r)
-    umap_init_mca <- umap(mca_res$ind$coord, 
-                          min_dist = 1e-6, 
-                          n_components = 3, 
-                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
-                          n_epochs = 1000, 
-                          n_threads = 1)
-    res <- hdbscan(umap_init_mca, minPts = i)
-    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
-    return(score_iter_temp)
-  }
-  score_mca[i - 2, ] <- score_iter
-}
-
-score <- rowMeans(score_mca)
-score_df <- tibble(minPts = 3:max_n_pt, score = score)
-
-sample_df <- score_mca %>%
-  as.data.frame() %>%
-  as_tibble() %>%
-  add_column(minPts = 3:max_n_pt, .before = 1) %>%
-  pivot_longer(-minPts) %>%
-  group_by(minPts)
-
-optimal_minPts <- score_df %>%
-  filter(score == min(score)) %>%
-  pull(minPts)
-
-save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
+  .[match(rownames(tumor_samples), .$sample),]
+
+## umap analysis
+set.seed(3296)
+gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
+gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
 
@@
 
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5, message = FALSE}
+```{r fig-1a-1l, warning=FALSE, fig.height=15, fig.width=16, message = FALSE}
 load("../data/cancer_disrupt_heatmap.Rda")
@@
 scale_y_discrete(position = "left") +
-  labs(tag = "A") +
   ylab("") +
+  xlab("377 Reactome Pathways") +
   theme(axis.text.x = element_blank(),
@@
 
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-permut_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
-  geom_line(aes(colour = "#998ec3")) +
-  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
-  scale_color_identity(guide = guide_legend(title = ""), 
-                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
-  scale_y_continuous(labels = scales::percent) +
-  xlab("MCA Eigenvalues") +
-  ylab("explained variance") +
-  theme_classic()
-
-pvalue_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = pvalue)) +
-  geom_line() +
-  geom_point() +
-  geom_vline(xintercept = optimal_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
-  xlab("MCA Eigenvalues") +
-  ylab("P-value") +
-  theme_classic()
-
-fig1B <- permut_p /
-  pvalue_p + plot_annotation(title = "Significance of MCA Eigenvalues")
-
 ## umap monochrome 2d
@@
   theme_classic() +
-  labs(tag = "C") +
   theme(axis.text.x = element_blank(),
@@
   theme_classic() +
-  labs(tag = "E") +
   theme(axis.text.x = element_blank(),
@@
   theme_classic() +
-  labs(tag = "F") +
   theme(axis.text.x = element_blank(),
@@
 
+## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
+mut_order <- mut_order[,c("V1", "knn_clust", "log10mut")] ## selecting relevant column
+
+begin_NA <- data.frame(V1 = paste0("sample", 1:50), 
+                       knn_clust = min(as.numeric(mut_order$knn_clust)), ## put lowest cluster to prevent a new facet
+                       log10mut = NA) ## put NA values so they don't get plotted
+end_NA <- data.frame(V1 = paste0("sample", 51:100), ## repeat the step for the ends
+                       knn_clust = max(as.numeric(mut_order$knn_clust)),
+                       log10mut = NA)
+mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
+
+mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1) ## factor samples
+mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12)) ## factor clusters
+#### test ####
+
 fig1G <- ggplot(mut_order, aes(x = V1, y = log10mut, color = knn_clust)) + 
@@
   xlab("") +
-  labs(tag = "G") +
   theme_classic() +
@@
   scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
+  scale_x_discrete(breaks = c(as.character(mut_order$V1[1]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
   theme(axis.text.x = element_text(size = 10),
@@
         axis.line.x.top = element_line(colour = "black", size = 0.5),
-        #strip.background.x = element_blank(),
+        panel.spacing = unit(1, "lines"),
+        strip.background = element_rect(size = 0.5),
         strip.placement = "outside",
-        strip.text.x = element_text(size = 12),
+        strip.text.x = element_text(size = 10),
         legend.position = "none") +
@@
   xlab("umap_1") +
-  labs(tag = "H", color = "") +
+  labs(color = "") +
   scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
@@
   xlab("umap_1") +
-  labs(tag = "I", color = "") +
+  labs(color = "") +
   theme(axis.text.x = element_blank(),
@@
   xlab("umap_1") +
-  labs(color = "", tag = "J") +
+  labs(color = "") +
   scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
@@
   xlab("umap_1") +
-  labs(tag = "K", color = "") +
+  labs(color = "") +
   scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
@@
   xlab("umap_1") +
-  labs(tag = "L", color = "") +
+  labs(color = "") +
   scale_color_manual(labels = c("M0", "M1"), 
@@
 
+empty_plot <- ggplot() + theme_void()
+
 layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGGGG
-AAAAGGGGGG
-AAAAHHIIJJ
-AAAAKKLL##
+AAAABBBCCCDDD
+AAAAEEEFFF###
+AAAAGGGGGGGGG
+AAAAGGGGGGGGG
+AAAAHHHIIIJJJ
+AAAAKKKLLL###
 '
 
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
+fig1A + empty_plot + fig1C + empty_plot + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')
 

### 2020-07-20 code/figure1.Rmd

@@
 ####################
+## Figure 1B data ##
+####################
+gdc_path <- gdc_reactome_path_analysis %>%
+  t() %>%
+  as.data.frame() %>%
+  rownames_to_column(var = "sample") %>% ## move path_names to column
+  .[match(rownames(tumor_samples), .$sample), ]
+
+gdc_path_mca <- left_join(gdc_path, tumor_samples[, c(1, 3)],
+  by = c("sample" = "submitted_sample_id")) %>%
+  column_to_rownames(var = "sample")
+
+gdc_path_sample_names <- rownames(gdc_path_mca)
+gdc_path_tumor_types <- gdc_path_mca$project_code.x
+
+gdc_path_mca$project_code.x <- NULL
+
+## create a categorical variable from the boolean variable
+pathway_list <- make_clean_names(colnames(gdc_path_mca))
+pathway_count <- length(pathway_list)
+gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
+  factor(x, levels = c(0, 1), labels = c("no", "yes"))
+}))
+colnames(gdc_path_mca) <- pathway_list
+gdc_path_mca$project_code <- gdc_path_tumor_types
+rownames(gdc_path_mca) <- gdc_path_sample_names
+
+## perform mca
+mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
+
+expl_var <- mca_res$eig[, 2]
+expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
+
+set.seed(4242)
+for (k in seq(100)) {
+  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
+  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
+  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
+}
+
+randomize <- tibble(
+  eigenvalue = 1:length(expl_var),
+  expl_var = expl_var,
+  perm = colMeans(expl_var_perm),
+  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
+)
+
+optimal_mca_dimensions <- randomize %>%
+  filter(pvalue < 0.05) %>%
+  filter(eigenvalue == max(eigenvalue)) %>%
+  pull(eigenvalue)
+
+mca_res <- MCA(
+  X = gdc_path_mca[, pathway_list],
+  ncp = optimal_mca_dimensions, graph = FALSE
+)
+
+save(randomize, mca_res, file = "figure1B_data.rda")
+
+####################
 ## Figure 1C data ##
 ####################
-gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample),]
-
-## umap analysis
-set.seed(3296)
-gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
-gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])
+rng <- RNGseq(max_n_pt * n_iter, 42)
+r <- rng[[3649]]
+
+rngtools::setRNG(r)
+gdc.umap <- uwot::umap(mca_res$ind$coord, 
+                       n_components = 3,
+                       n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
+                       n_epochs = 1000,
+                       n_threads = 1)
+gdc_umap_coors <- as.data.frame(gdc.umap) %>%
+  rename(plot_x = V1, plot_y = V2, plot_z = V3)
+rownames(gdc_umap_coors) <- rownames(gdc_path_mca)
+
+####################
+## Figure 1D data ##
+####################
+n_iter <- 40
+max_n_pt <- 200
+score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
+
+n_tumors <- nrow(mca_res$ind$coord)
+
+cl <- makeCluster(8)
+registerDoParallel(cl)
+
+for (i in 3:max_n_pt) {
+  print(paste0("MIN POINTS = ", i))
+  score_iter <- foreach(1:n_iter,
+    .combine = c,
+    .export = c("mca_res"),
+    .packages = c("dbscan", "uwot"),
+    r = rng[(i - 1) * n_iter + 1:n_iter]
+  ) %dopar% {
+    rngtools::setRNG(r)
+    umap_init_mca <- umap(mca_res$ind$coord, 
+                          min_dist = 1e-6, 
+                          n_components = 3, 
+                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
+                          n_epochs = 1000, 
+                          n_threads = 1)
+    res <- hdbscan(umap_init_mca, minPts = i)
+    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
+    return(score_iter_temp)
+  }
+  score_mca[i - 2, ] <- score_iter
+}
+
+score <- rowMeans(score_mca)
+score_df <- tibble(minPts = 3:max_n_pt, score = score)
+
+sample_df <- score_mca %>%
+  as.data.frame() %>%
+  as_tibble() %>%
+  add_column(minPts = 3:max_n_pt, .before = 1) %>%
+  pivot_longer(-minPts) %>%
+  group_by(minPts)
+
+optimal_minPts <- score_df %>%
+  filter(score == min(score)) %>%
+  pull(minPts)
+
+save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
 

### 2020-07-20 code/figure1.Rmd
change coordinate object to have  a column explicitly called sample_id and not row.names

--- 2020-07-20 code/figure1.Rmd
@@
     member_prob = hdb_cluster_assignment$membership_prob,
-    row.names = rownames(gdc_umap_coors)
+    sample_id = rownames(gdc_umap_coors)
   )

### 2020-07-20 code/figure1.Rmd
Add packages for mca, umap, random numbers, and data cleaning

@@
 rngtools::setRNG(r)
-gdc.umap <- uwot::umap(mca_res$ind$coord, 
+gdc.umap <- umap(mca_res$ind$coord, 
                  n_components = 3,
@@
 
-cluster_umap <- uwot::umap(mca_res$ind$coord,
+cluster_umap <- umap(mca_res$ind$coord,
   min_dist = 1e-6,

### 2020-07-20 code/figure1.Rmd
Change all references to umap_2d_3d_coors to umap_3d_coors ; also alter all references to V1 to read sample_id

@@
   'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "V1") ## move rownames to column
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
+  rownames_to_column(var = "sample_id") ## move rownames to column
+
+umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id")
 
@@
   mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "V1")
-
-umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
+  rownames_to_column(var = "sample_id")
+
+umap_3d_coors <- left_join(umap_3d_coors, tumor_samples, by = "sample_id")
 
@@
 
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
+umap_3d_coors <- umap_3d_coors %>%
+  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
   left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
@@
 
-umap_2d_3d_coors <- umap_2d_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
+umap_3d_coors <- umap_3d_coors %>%
+  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
   left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
@@
 
-M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
+M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
+M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
+NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
+
+umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
+umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
+umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
+
+stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
+stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
+stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
+stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
+unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
+
+umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
+umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
+umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
+umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
+umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
+
+write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
 ```
@@
 load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+umap_cancertype <- fread("../data/umap_3d_coors.tsv")
 
@@
   mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
-    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
+    dplyr::select(sample_id, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
     arrange(mut_count) ## arrange mutation in increasing order
@@
 
-mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
+mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
 mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
@@
 xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$sample_id,] ## find the tumor samples in these positions
 
@@
 
-fig1G <- ggplot(mut_order, aes(x = V1, y = log10mut, color = knn_clust)) + 
+fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = knn_clust)) + 
   geom_point(size = 0.25) + 
@@
   scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$V1[300]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
+  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[300]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
   theme(axis.text.x = element_text(size = 10),
@@
 load("../data/gene_data.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+umap_cancertype <- fread("../data/umap_3d_coors.tsv")
 brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
@@
   t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
+  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
   mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
@@
   dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
+  left_join(., umap_cancertype[,c("sample_id", "x_2d", "y_2d", "project_code")], by = "sample_id") ## leftjoin by sample id
 

### 2020-07-20 code/figure1.Rmd
alter all references from x_2d and y_2d to plot_x and plot_y

@@
 ## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=x_2d, y=y_2d)) +
+fig1C <- ggplot(umap_cancertype, aes(x=plot_x, y=plot_y)) +
   geom_point(colour = "grey", size = 0.25) +
@@
 ## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = x_2d, y = y_2d, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = plot_x, y = plot_y, color = as.factor(clust))) +
+  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
@@
 ## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = x_2d, y = y_2d, color = as.factor(knn_clust))) +
+fig1F <- ggplot(subset(umap_cancertype), aes(x = plot_x, y = plot_y, color = as.factor(knn_clust))) +
   geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
@@
   mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
-    dplyr::select(sample_id, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
+    dplyr::select(sample_id, plot_x, plot_y, knn_clust, mut_count) %>% ## select relevant columns
     arrange(mut_count) ## arrange mutation in increasing order
@@
 ## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = x_2d, y = y_2d, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = plot_x, y = plot_y, colour = project_code)) +
+  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + 
@@
 ## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = plot_x, y = plot_y, colour = Subtype_Selected)) +
+  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) +
@@
   dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("sample_id", "x_2d", "y_2d", "project_code")], by = "sample_id") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = x_2d, y = y_2d, colour = mut_type)) +
-  geom_point(data = brca, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "project_code")], by = "sample_id") ## leftjoin by sample id
+
+fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = plot_x, y = plot_y, colour = mut_type)) +
+  geom_point(data = brca, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + 
@@
 ## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = x_2d, y = y_2d, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = plot_x, y = plot_y, colour = ajcc_pathologic_tumor_stage)) +
+  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + 
@@
 ## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = x_2d, y = y_2d)) +
-  geom_point(data = umap_cancertype, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = x_2d, y = y_2d, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
+fig1L <- ggplot(umap_cancertype, aes(x = plot_x, y = plot_y)) +
+  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
   ylab("umap_2") +

### 2020-07-20 code/figure1.Rmd
modify plotting to coord_fixed for umaps; added plots 1b and 1d, rename umap coordinates UMAP 1 and UMAP 2

@@
   mutate(
+    n_pathways_mut = rowSums(gdc_path[, -1]),
     clust = as.factor(hdb_cluster_assignment$cluster),
@@
 
+load("../data/figure1B_data.rda")
 optimal_mca_dimensions <- randomize %>%
@@
   ylab("explained variance") +
-  theme_classic()
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12))
 
@@
   ylab("P-value") +
-  theme_classic()
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12))
 
@@
 
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=plot_x, y=plot_y)) +
-  geom_point(colour = "grey", size = 0.25) +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
+fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
+                aes(x=plot_x, y=plot_y)) +
+  geom_point(size = 0.25) +
+  xlab("UMAP 1") +
+  ylab("UMAP 2") +
+  coord_fixed() +
+  theme_classic() +
+  scale_color_viridis_c()
   labs(tag = "C") +
@@
         axis.ticks = element_blank())
+
+load("../data/optimal_minpts_score.rda")
+fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
+  geom_point(data = sample_df, color = "grey", aes(y = value)) +
+  geom_point() + 
+  geom_line() +
+  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
+  scale_y_log10() +
+  xlab("Minimum Cluster Size") +
+  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank())
+
 ## umap hdb_scan
@@
   theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
+  xlab("UMAP 1") +
+  ylab("UMAP 2") +
+  coord_fixed() +
   theme_classic() +
@@
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
   labs(tag = "H", color = "") +
@@
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
   labs(tag = "I", color = "") +
@@
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
   labs(color = "", tag = "J") +
@@
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
   labs(tag = "K", color = "") +
@@
   geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
   labs(tag = "L", color = "") +
@@
 AAAABBCCDD
-AAAAEEEFFF
+AAAABBEEFF
 AAAAGGGGGG
@@
 
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
+fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
 

### 2020-07-20 code/figure1.Rmd
minor updates to figure1.Rmd...reviewing simons code

--- 2020-07-20 code/figure1.Rmd
@@
 rngtools::setRNG(r)
-gdc.umap <- umap(mca_res$ind$coord, 
+gdc.umap <- uwot::umap(mca_res$ind$coord, 
                        n_components = 3,
@@
 
-cluster_umap <- umap(mca_res$ind$coord,
+cluster_umap <- uwot::umap(mca_res$ind$coord,
   min_dist = 1e-6,
@@
     member_prob = hdb_cluster_assignment$membership_prob,
-    sample_id = rownames(gdc_umap_coors)
+    row.names = rownames(gdc_umap_coors)
   )
@@
   'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "sample_id") ## move rownames to column
-
-umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id")
+  rownames_to_column(var = "V1") ## move rownames to column
+
+umap_2d_3d_coors <- left_join(umap_2d_3d_coors, mutation_count, by = "V1")
 
@@
   mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "sample_id")
-
-umap_3d_coors <- left_join(umap_3d_coors, tumor_samples, by = "sample_id")
+  rownames_to_column(var = "V1")
+
+umap_2d_3d_coors <- left_join(umap_2d_3d_coors, tumor_samples, by = "V1")
 
@@
 
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
+umap_2d_3d_coors <- umap_2d_3d_coors %>%
+  mutate(pan.samplesID = substring(text = .$V1, first = 1, last = 12)) %>%
   left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
@@
 
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
+umap_2d_3d_coors <- umap_2d_3d_coors %>%
+  mutate(bcr_patient_barcode = substring(text = .$V1, first = 1, last = 12)) %>%
   left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
@@
 
-M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
+M0 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
+M1 <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
+NA.value <- umap_2d_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
+
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
+umap_2d_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
+
+stage1 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
+stage2 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
+stage3 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
+stage4 <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
+unknown <- umap_2d_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
+
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
+umap_2d_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
+
+write.table(x = umap_2d_3d_coors, file = "umap_2d_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
 ```
 
-```{r fig-1a-1l, warning=FALSE, fig.height=11, fig.width=8.5, message = FALSE}
+```{r fig-1a-1l, warning=FALSE, fig.height=15, fig.width=16, message = FALSE}
 load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
 
@@
 
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-permut_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
-  geom_line(aes(colour = "#998ec3")) +
-  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
-  scale_color_identity(guide = guide_legend(title = ""), 
-                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
-  scale_y_continuous(labels = scales::percent) +
-  xlab("MCA Eigenvalues") +
-  ylab("explained variance") +
-  theme_classic()
-
-pvalue_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = pvalue)) +
-  geom_line() +
-  geom_point() +
-  geom_vline(xintercept = optimal_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
-  xlab("MCA Eigenvalues") +
-  ylab("P-value") +
-  theme_classic()
-
-fig1B <- permut_p /
-  pvalue_p + plot_annotation(title = "Significance of MCA Eigenvalues")
+# optimal_mca_dimensions <- randomize %>%
+#   filter(pvalue < 0.05) %>%
+#   filter(eigenvalue == max(eigenvalue)) %>%
+#   pull(eigenvalue)
+# 
+# permut_p <- randomize %>%
+#   filter(eigenvalue < 50) %>%
+#   ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
+#   geom_line(aes(colour = "#998ec3")) +
+#   geom_line(aes(y = perm / 100, colour = "#f1a340")) +
+#   scale_color_identity(guide = guide_legend(title = ""), 
+#                        labels = c("Eigenvalues", "Permuted Eigenvalues")) +
+#   scale_y_continuous(labels = scales::percent) +
+#   xlab("MCA Eigenvalues") +
+#   ylab("explained variance") +
+#   theme_classic()
+# 
+# pvalue_p <- randomize %>%
+#   filter(eigenvalue < 50) %>%
+#   ggplot(aes(x = eigenvalue, y = pvalue)) +
+#   geom_line() +
+#   geom_point() +
+#   geom_vline(xintercept = optimal_dimensions, color = "red") +
+#   ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
+#   xlab("MCA Eigenvalues") +
+#   ylab("P-value") +
+#   theme_classic()
+# 
+# fig1B <- permut_p /
+#   pvalue_p + plot_annotation(title = "Significance of MCA Eigenvalues")
 
@@
   mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
-    dplyr::select(sample_id, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
+    dplyr::select(V1, x_2d, y_2d, knn_clust, mut_count) %>% ## select relevant columns
     arrange(mut_count) ## arrange mutation in increasing order
@@
 
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
+mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1)
 mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
@@
 xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$sample_id,] ## find the tumor samples in these positions
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
 
@@
 
-fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = knn_clust)) + 
+## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
+mut_order <- mut_order[,c("V1", "knn_clust", "log10mut")] ## selecting relevant column
+
+begin_NA <- data.frame(V1 = paste0("sample", 1:50), 
+                       knn_clust = min(as.numeric(mut_order$knn_clust)), ## put lowest cluster to prevent a new facet
+                       log10mut = NA) ## put NA values so they don't get plotted
+end_NA <- data.frame(V1 = paste0("sample", 51:100), ## repeat the step for the ends
+                       knn_clust = max(as.numeric(mut_order$knn_clust)),
+                       log10mut = NA)
+mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
+
+mut_order$V1 <- factor(mut_order$V1, levels = mut_order$V1) ## factor samples
+mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12)) ## factor clusters
+#### test ####
+
+fig1G <- ggplot(mut_order, aes(x = V1, y = log10mut, color = knn_clust)) + 
   geom_point(size = 0.25) + 
@@
   scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[300]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
+  scale_x_discrete(breaks = c(as.character(mut_order$V1[1]), as.character(xaxis_breaks$V1)), labels = c("n =" ,total_n$Freq)) +
   theme(axis.text.x = element_text(size = 10),
@@
   xlab("umap_1") +
-  labs(tag = "H", color = "") +
+  labs(color = "") +
+  guides(fill = guide_legend(override.aes = list(size=8))) +
   scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
@@
 load("../data/gene_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
+umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
 brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
@@
   t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
+  rownames_to_column(var = "V1") %>% ## getting data ready for leftjoin 
   mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
@@
   dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("sample_id", "x_2d", "y_2d", "project_code")], by = "sample_id") ## leftjoin by sample id
+  left_join(., umap_cancertype[,c("V1", "x_2d", "y_2d", "project_code")], by = "V1") ## leftjoin by sample id
 

### 2020-07-20 text/manuscript.Rmd

@@
 library(org.Hs.eg.db)
-library(umap)
 library(dbscan)
@@
 library(cowplot)
+library(doRNG)
+library(uwot)
+library(janitor)
+library(FactoMineR)
 ```
@@
 ```
-
 In order to classify tumors using this dataset, we used multiple correspondence
@@
 
-```{r child='../code/figure4.Rmd'}
+```{r child='../code/figure4_function.Rmd'}
 ```
@@
 
-```{r child='table2.Rmd'}
+```{r child='../code/table2.Rmd'}
 ```

### 2020-07-21 text/discussion.Rmd
add pcawg annotation-aware annotation reference to discussion

@@
 
-We find that most tumors segregate into one of 12 biologically distinct
+We find that most tumors segregate into one of twelve biologically distinct
 clusters, regardless of the cancer type (**figure 1h** and **supplementary

### 2020-07-21 code/figure2.Rmd
updated code to work with new cluster assignments by simon

@@
 ```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
 load("../data/gdc_reactome_path_analysis.Rda")
@@
 
-for(n in 1:12) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("V1", "knn_clust")] %>% filter(knn_clust == n) %>% .[,.$V1] ## get class spec sample ids
+for(n in 1:max(tumor_clusters$clust_knn)) {
+  class_spec_sample_id <- tumor_clusters %>% .[,c("sample_id", "clust_knn")] %>% filter(clust_knn == n) %>% .[,.$sample_id] ## get class spec sample ids
   top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
@@
 onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-figure_labs <- LETTERS[seq(from = 2, to = 13)]
-for(n in 1:12) {
+for(n in 1:length(clust_col_order)) {
   onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
@@
   xlab(label = paste0("Class ", n)) +
-  labs(tag = figure_labs[n]) +
   theme_classic() +
@@
 
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plots_onco[[11]] + plots_onco[[12]] + plot_layout(design = layout)
+fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
 ```

### 2020-07-21 code/figure4.Rmd
updated code to work with new cluster assignments by simon

@@
 ```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
 load("../data/gene2_level2_lookup.Rda")
@@
 ## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
+filtered_class <- c(2) ## filtering for only class 2 samples
 n_path <- 10 ## looking at the top10 path

### 2020-07-21 code/table1.Rmd
updated code to work with new cluster assignments by simon

--- 2020-07-14 text/table_1.Rmd
@@
 ```{r table-1-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
 load("../data/moderate_enrich_prob.Rda") ## load data
@@
   clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust %in% d) ## get the smaple ids and cancer type for spec class
+  class_spec <- tumor_clusters %>% dplyr::select(sample_id, clust_knn, project_code) %>% filter(clust_knn %in% d) ## get the smaple ids and cancer type for spec class
   top_genes_in_path_cancer_type <- list()
@@
       message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust, project_code) %>% filter(knn_clust == d) ## sample_id for class
+      class_spec <- tumor_clusters %>% dplyr::select(sample_id, clust_knn, project_code) %>% filter(clust_knn == d) ## sample_id for class
       if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "V1") %>% rownames() ## get cancer type specific tumor classes
+        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "sample_id") %>% rownames() ## get cancer type specific tumor classes
       if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
         top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
+          rownames_to_column(var = "gene") %>%
           mutate(n_sample = length(class_type_spec),
                perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
+         arrange(desc(obs_count))
       }
@@
         rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
+        rownames_to_column(var = "gene") %>%
         mutate(n_sample = length(class_type_spec),
                perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-        rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
+         arrange(desc(obs_count))
       }
@@
 
-cluster_spec_enrich_table1 <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table1, file = "cluster_spec_enrich_table1.Rda")
+cluster_spec_enrich_table <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
+save(cluster_spec_enrich_table, file = "../data/cluster_spec_enrich_table.Rda")
 rm(list = ls(all.names = T))
@@
 ```{r table-1}
-load("../data/cluster_spec_enrich_table1.Rda")
-cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
+load("../data/cluster_spec_enrich_table.Rda")
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
+
+## Selecting the top 10 cancers with the largest n for table 1
+n_tumor <- as.data.frame(table(tumor_clusters$project_code))
+n_tumor <- n_tumor %>% arrange(desc(Freq)) %>% .[1:10, "Var1"] %>% as.character() %>% .[order(.)]
+
+cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%

### 2020-07-21 code/table2.Rmd
updated code to work with new cluster assignments by simon

--- 2020-07-14 text/table_2.Rmd
@@
 ```{r table-2-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
 load("../data/gdc_reactome_path_analysis.Rda")
 
-tumor_clusters <- tumor_clusters %>% dplyr::select(V1, ajcc_metastasis_pathologic_pm, knn_clust)
+tumor_clusters <- tumor_clusters %>% dplyr::select(sample_id, ajcc_metastasis_pathologic_pm, clust_knn)
 
@@
   n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
-  path_analysis <- path_analysis[,names(path_analysis) %in% data$V1] ## Filter only group specific cancers
-  data <- data[match(names(path_analysis), data$V1),] ## Match tumor sample ids
-  table(names(path_analysis) == data$V1) ## good they match now
+  path_analysis <- path_analysis[,names(path_analysis) %in% data$sample_id] ## Filter only group specific cancers
+  data <- data[match(names(path_analysis), data$sample_id),] ## Match tumor sample ids
+  table(names(path_analysis) == data$sample_id) ## good they match now
   ## Now we are going to build group specific cluster data
-  n_tumor_in_clust <- as.data.frame(table(data$knn_clust)) ## getting total n in each cluster
+  n_tumor_in_clust <- as.data.frame(table(data$clust_knn)) ## getting total n in each cluster
   names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
@@
     print(n)
-    df <- subset(data, knn_clust == n) ## subset cluster number
-    df <- df$V1 ## getting tumor sample id
+    df <- subset(data, clust_knn == n) ## subset cluster number
+    df <- df$sample_id ## getting tumor sample id
     df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
@@
 m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
-## Combining cluster 6/7 cluster 910 cluster 1112
-combo_group <- c("6", "9", "11")
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m0_rbeta <- rbind(m0_rbeta, compo_group_df) ## rbind m0_rbeta and the compo group together
-save(m0_rbeta, file = "m0_rbeta.Rda")
+save(m0_rbeta, file = "../data/m0_rbeta.Rda")
 
@@
 m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
-compo_group_df <- list() ## list for loop
-for(n in combo_group) {
-  print(n) ## first cluster in set
-  m <- as.numeric(n) + 1 ## second cluster in set
-  m <- as.character(m)
-  df <- m0_rbeta[m0_rbeta$clust == n,] ## get rbeta info for first cluster in set
-  df1 <- m0_rbeta[m0_rbeta$clust == m,] ## get rbeta info for second cluster in set
-  tot_clust_mut <- df$clust_mut + df1$clust_mut ## add both clust muts
-  tot_n_clust <- df$tot_n_clust + df1$tot_n_clust ## add both clust n's
-  p <- paste0(n,m)
-  df2 <- m0_rbeta[1:377,]
-  df2$clust <- p ## replace with new clust name
-  df2$clust_mut <- tot_clust_mut ## replace with new clust mut count
-  df2$tot_n_clust <- tot_n_clust ## replace with new clust size
-  compo_group_df[[p]] <- df2 ## store in list and continue loop
-}
-compo_group_df <- do.call(rbind, compo_group_df) ## rbind list
-m1_rbeta <- rbind(m1_rbeta, compo_group_df) ## Combine the new groups
-save(m1_rbeta, file = "m1_rbeta.Rda")
+save(m1_rbeta, file = "../data/m1_rbeta.Rda")
 
@@
 
-save(m0_rbeta, m1_rbeta, file = "m0_m1_rbeta.Rda")
+save(m0_rbeta, m1_rbeta, file = "../data/m0_m1_rbeta.Rda")
 rm(list = ls(all.names = T)) ## cleaning up data
@@
                                    upper = quant_diff[,3])
-save(m1_minus_m0_per_cluster_rbeta, file = "m1_minus_m0_per_cluster_rbeta.Rda")
+save(m1_minus_m0_per_cluster_rbeta, file = "../data/m1_minus_m0_per_cluster_rbeta.Rda")
 rm(list = ls(all.names = T))

### 2020-07-21 text/manuscript.Rmd

@@
 [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation]. Data selected in this manner are likely to be
-noisy for several reasons. First, point mutations can be deleterious
-(attenuating or hypomorphic) or activating (neomorphic) in genes that act
-functionally as either activators or repressors. Within the scope of this study
-we must accept this uncertainty and assume mutations are generically disruptive
-to the cell regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, as previously mentioned, we know that
-low-expressed genes and non-expressed genes are known to accumulate mutations at
-a higher rate due to transcription coupled repair. To address this issue, we
-identified genes with low expression in each type of cancer and eliminated them
-for that cancer type only (see methods for details). <strike>Highly expressed
-genes also have high mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. We felt that this mechanism results in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose to
-include these genes in our analysis.</strike>
-
+@Pervasive_lesion_segregation; @expression-aware-annotation]. Data selected in
+this manner are likely to be noisy for several reasons. First, point mutations
+can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in
+genes that act functionally as either activators or repressors. Within the scope
+of this study we must accept this uncertainty and assume mutations are
+generically disruptive to the cell regardless of whether they are activating or
+deactivating or whether they affect a gene that we might classify as one thing
+or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, as previously
+noted, we know that low-expressed genes and non-expressed genes are known to
+accumulate mutations at a higher rate due to transcription coupled repair. To
+address this issue, we identified genes with low expression in each type of
+cancer and eliminated them for that cancer type only (see methods for details).
+<strike>Highly expressed genes also may have high mutation rates owing to
+transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
+that this mechanism results in cell-type-specific biases that might be
+biologically meaningful for predisposition to different classes of cancer
+however, and therefore chose to include these genes in our analysis.</strike>
 After selecting our pathways and genes, we then compiled a matrix of the

### 2020-07-21 code/figure1.Rmd

--- 2020-07-20 code/figure1.Rmd
@@
 load("../data/gdc_tumor_coding_complete.Rdata")
+load("../data/gene2_level2_lookup.Rda")
 
@@
   as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%  
+  dplyr::select(project_code, submitted_sample_id, submitted_specimen_id) %>%  
   mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
@@
   rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(rownames(tumor_samples), .$sample), ]
-
-gdc_path_mca <- left_join(gdc_path, tumor_samples[, c(1, 3)],
+  .[match(tumor_samples$submitted_sample_id, .$sample), ]
+
+gdc_path_mca <- left_join(gdc_path, tumor_samples[, c("project_code", "submitted_sample_id")],
   by = c("sample" = "submitted_sample_id")) %>%
@@
 
-gdc_path_sample_names <- rownames(gdc_path_mca)
-gdc_path_tumor_types <- gdc_path_mca$project_code.x
-
-gdc_path_mca$project_code.x <- NULL
+gdc_path_sample_names <- rownames(gdc_path_mca) ## Getting sample id order
+gdc_path_tumor_types <- gdc_path_mca$project_code ## Getting cancer type order
+
+gdc_path_mca$project_code <- NULL ## Remove project_code column from df
 
@@
 ####################
+n_iter <- 40
+max_n_pt <- 200
 rng <- RNGseq(max_n_pt * n_iter, 42)
@@
 umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id") 
+rm(mutation_count)
 
@@
 ####################
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code) %>% 
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) %>%
-  rownames_to_column(var = "sample_id")
-
-umap_3d_coors <- left_join(umap_3d_coors, tumor_samples, by = "sample_id")
+umap_3d_coors <- left_join(umap_3d_coors, tumor_samples[,c("project_code", "submitted_sample_id")], by = c("sample_id" = "submitted_sample_id"))
 
@@
 
+##figure 1b
+load("../data/figure1B_data.rda")
 optimal_mca_dimensions <- randomize %>%
@@
   ylab("explained variance") +
-  theme_classic()
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        legend.position = c(0.8, 0.8))
 
@@
   geom_point() +
-  geom_vline(xintercept = optimal_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
+  geom_vline(xintercept = optimal_mca_dimensions, color = "red") +
+  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_mca_dimensions)) +
   xlab("MCA Eigenvalues") +
   ylab("P-value") +
-  theme_classic()
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12))
 
 fig1B <- permut_p /
-  pvalue_p + plot_annotation(title = "Significance of MCA Eigenvalues")
-
-## umap monochrome 2d
-fig1C <- ggplot(umap_cancertype, aes(x=plot_x, y=plot_y)) +
-  geom_point(colour = "grey", size = 0.25) +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
+  pvalue_p
+
+fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
+                aes(y=plot_x, x=plot_y)) +
+  geom_point(size = 0.25) +
+  xlab("UMAP 1") +
+  ylab("UMAP 2") +
+  coord_fixed() +
+  theme_classic() +
+  scale_color_viridis_c() +
   labs(tag = "C") +
@@
         axis.ticks = element_blank())
+
+load("../data/optimal_minpts_score.rda")
+fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
+  geom_point(data = sample_df, color = "grey", aes(y = value), size = 0.25) +
+  geom_point(size = 0.25) + 
+  geom_line() +
+  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
+  scale_y_log10() +
+  xlab("Minimum Cluster Size") +
+  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
+  theme_classic() +
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12))
+
 ## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = plot_x, y = plot_y, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
+fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(y = plot_x, x = plot_y, color = as.factor(clust))) +
+  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
@@
   theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "E") +
+  xlab("UMAP 1") +
+  ylab("UMAP 2") +
+  coord_fixed() +
+  theme_classic() +
   theme(axis.text.x = element_blank(),
@@
 ## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(x = plot_x, y = plot_y, color = as.factor(knn_clust))) +
+fig1F <- ggplot(subset(umap_cancertype), aes(y = plot_x, x = plot_y, color = as.factor(clust_knn))) +
   geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
@@
   theme_classic() +
-  xlab("umap_1") +
-  ylab("umap_2") +
-  theme_classic() +
-  labs(tag = "F") +
+  xlab("UMAP 1") +
+  ylab("UMAP 2") +
+  coord_fixed() +
+  theme_classic() +
   theme(axis.text.x = element_blank(),
@@
 
-umap_cancertype$knn_clust <- factor(umap_cancertype$knn_clust, levels = c(1:12))
-
 ## mutation burden plot
 mut_order <- list() ## store data in list 
-for(n in 1:max(as.numeric(as.character(umap_cancertype$knn_clust)))) { ## loop through all the clusters
+for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
   message(n)
-  mut_order[[n]] <- umap_cancertype %>% filter(knn_clust == n) %>%  ## filter for specific class
-    dplyr::select(sample_id, plot_x, plot_y, knn_clust, mut_count) %>% ## select relevant columns
+  mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
+    dplyr::select(sample_id, plot_x, plot_y, clust_knn, mut_count) %>% ## select relevant columns
     arrange(mut_count) ## arrange mutation in increasing order
@@
 mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
-mut_order$knn_clust <- factor(mut_order$knn_clust, levels = c(1:12))
+mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10))
 
@@
   message(n)
-  class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
+  class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
   xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
   previous_class <- class_spec ## storing class halfway mark
-  for(n in 2:max(as.numeric(as.character(mut_order$knn_clust)))) {
+  for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
     message(n)
     position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
-    class_spec <- round(mut_order %>% filter(knn_clust == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
+    class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
     xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
@@
 xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$sample_id,] ## find the tumor samples in these positions
-
-total_n <- as.data.frame(table(mut_order$knn_clust)) ## getting total n in cluster
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
+
+total_n <- as.data.frame(table(mut_order$clust_knn)) ## getting total n in cluster
 
@@
 mut_median <- mut_order %>% 
-  group_by(knn_clust) %>% 
+  group_by(clust_knn) %>% 
   dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
 
-fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = knn_clust)) + 
+## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
+mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column
+
+begin_NA <- data.frame(sample_id = paste0("sample", 1:50), 
+                       clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
+                       log10mut = NA) ## put NA values so they don't get plotted
+end_NA <- data.frame(sample_id = paste0("sample", 51:100), ## repeat the step for the ends
+                       clust_knn = max(as.numeric(mut_order$clust_knn)),
+                       log10mut = NA)
+mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
+
+mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id) ## factor samples
+mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10)) ## factor clusters
+
+fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = clust_knn)) + 
   geom_point(size = 0.25) + 
@@
   xlab("") +
-  labs(tag = "G") +
-  theme_classic() +
-  facet_wrap(~ factor(knn_clust, levels = c(1:12)), ncol = 12, scales = "free_x") +
+  theme_classic() +
+  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") +
   geom_hline(data= mut_median, aes(yintercept=mut_med)) +
   scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[300]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
+  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[1]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
   theme(axis.text.x = element_text(size = 10),
@@
 ## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = plot_x, y = plot_y, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
+fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(y = plot_x, x = plot_y, colour = project_code)) +
+  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + 
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "H", color = "") +
+  guides(colour = guide_legend(override.aes = list(size=3))) +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
+  labs(color = "") +
   scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
@@
 ## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = plot_x, y = plot_y, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
+fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(y = plot_x, x = plot_y, colour = Subtype_Selected)) +
+  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) +
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "I", color = "") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
+  labs(color = "") +
+  guides(colour = guide_legend(override.aes = list(size=3))) +
   theme(axis.text.x = element_blank(),
@@
 
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = plot_x, y = plot_y, colour = mut_type)) +
-  geom_point(data = brca, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
+fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(y = plot_x, x = plot_y, colour = mut_type)) +
+  geom_point(data = brca, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + 
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(color = "", tag = "J") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
+  labs(color = "") +
+  guides(colour = guide_legend(override.aes = list(size=3))) +
   scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
@@
 ## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = plot_x, y = plot_y, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
+fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(y = plot_x, x = plot_y, colour = ajcc_pathologic_tumor_stage)) +
+  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) + 
   theme_classic() +
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "K", color = "") +
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
+  labs(color = "") +
+  guides(colour = guide_legend(override.aes = list(size=3))) +
   scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
@@
 ## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(x = plot_x, y = plot_y)) +
-  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  ylab("umap_2") +
-  xlab("umap_1") +
-  labs(tag = "L", color = "") +
+fig1L <- ggplot(umap_cancertype, aes(y = plot_x, x = plot_y)) +
+  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
+  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
+  ylab("UMAP 2") +
+  xlab("UMAP 1") +
+  coord_fixed() +
+  labs(color = "") +
+  guides(colour = guide_legend(override.aes = list(size=3))) +
   scale_color_manual(labels = c("M0", "M1"), 
@@
 
+empty_plot <- ggplot() + theme_void()
+
 layout <- '
-AAAABBCCDD
-AAAAEEEFFF
-AAAAGGGGGG
-AAAAGGGGGG
-AAAAHHIIJJ
-AAAAKKLL##
+AAAABBBCCCDDD
+AAAABBBCCCDDD
+AAAABBBEEEFFF
+AAAABBBEEEFFF
+AAAAGGGGGGGGG
+AAAAGGGGGGGGG
+AAAAGGGGGGGGG
+AAAAHHHIIIJJJ
+AAAAHHHIIIJJJ
+AAAAKKKLLL###
+AAAAKKKLLL###
 '
 
-fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
+fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')
 

### 2020-07-21 code/figure3.Rmd

@@
 ```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(V1, knn_clust) ## select the relevant columns
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
+tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(sample_id, clust_knn) ## select the relevant columns
 
 ## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$knn_clust)) %>% ## table the number of tumors in each cluster
+n_tumor_in_clust <- as.data.frame(table(tumor_clusters$clust_knn)) %>% ## table the number of tumors in each cluster
   'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
@@
   print(n)
-  sample_id <- tumor_clusters %>% filter(knn_clust == n) %>% ## filter for tumor class
-    .[,"V1"] ## get sample_ids
+  sample_id <- tumor_clusters %>% filter(clust_knn == n) %>% ## filter for tumor class
+    .[,"sample_id"] ## get sample_ids
   path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
@@
 global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = 12)] %>% ## replicate the column 12 times for the 12 tumor classes
+  .[rep(1:ncol(.), each = max(tumor_clusters$clust_knn))] %>% ## replicate the column 0 times for the 10 tumor classes
   'colnames<-' (clust_group) %>% ## rename column with tumor class
@@
 diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg, prob)
+rm(moderate_fg, moderate_bg)
 gc()
@@
 gc()
-#save(moderate_enrich_prob, file = "moderate_enrich_prob.Rda")
+save(moderate_enrich_prob, file = "../data/moderate_enrich_prob.Rda")
 load("../data/moderate_enrich_prob.Rda")
@@
 effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:12))
+effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:max(as.numeric(effect_heatmap$cluster))))
 effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
@@
 
-save(effect_heatmap, file = "effect_heatmap.Rda")
+save(effect_heatmap, file = "../data/effect_heatmap.Rda")
+rm(list = ls(all.names = T))
+gc()
 ```
@@
 load("../data/effect_heatmap.Rda") ## load data
-effect_heatmap$test <- 1
 
@@
 
-# ## setting patchwork panel layout
-# layout <- '
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# AAAAA
-# #BBB#
-# '
-# 
-# fig3 + fig3_legend + plot_layout(design = layout)
-
 fig3_legend <- plot_grid(NULL, fig3_legend, NULL, ncol = 3)

### 2020-07-21 code/figure4_function.Rmd

@@
 ```{r figure-4-data-generation_function, message = FALSE, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
 load("../data/gene2_level2_lookup.Rda")
@@
 load("../data/gdc_mut_count_patient_booleanized.Rda")
-filtered_class <- c(6,7)
+filtered_class <- 2
 n_path <- 10
@@
 oncoprint_style <- function(filtered_class, n_path, enrich_cutoff = 0.30) {
-  class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
+  class_spec <- tumor_clusters %>% dplyr::select(sample_id, clust_knn) %>% filter(clust_knn %in% filtered_class) %>% .[,.$sample_id] ## sample_id for class specific
   toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
@@
 
-return(plot_grid(plot, plot_grid(NULL, plot_legend1, NULL, ncol = 1, rel_heights = c(1, 1, 1)), ncol = 2, rel_widths = c(2.5, 2), axis = "tbl"))
+return(plot_grid(plot, plot_grid(NULL, plot_legend, NULL, ncol = 1, rel_heights = c(1, 1, 1)), ncol = 2, rel_widths = c(2.5, 2), axis = "tbl"))
 }
 
-class67 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
-save(class67, file = "class67.Rda")
+class_spec_fig4 <- oncoprint_style(filtered_class = 2, n_path = 10)
+save(class_spec_fig4, file = "../data/class_spec_fig4.Rda")
+rm(list = ls(all.names = T))
+gc()
 ```
@@
 The highlighted genes at right are all proteasomal subunits."}
-load("../data/class67.Rda")
-class67
+load("../data/class_spec_fig4.Rda")
+class_spec_fig4
 ```

### 2020-07-22 text/discussion.Rmd
add PCAWG main paper and evolution refs to disc

@@
 disrupted in all or most of the defined classes (see **figures 2** and
-**supplementary figure S1**) and often are the result of abberations involving
+**supplementary figure S1**) and often are the result of aberrations involving
 common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
@@
 passively diverging mutation profiles which then serve as a substrate for futher
-selection. Moreover consortium samples are likely to contain contamination from
-support tissue and inflammatory and immune cells of the innate and adaptive
-immune systems, all potentially harboring cancer supporting mutations
-[@bhowmick_review]. We think it will be instructive to explore these ideas in
-the context of single cell experiments.
+selection [@PCAWG-MAIN; @PCAWG-EVO]. Moreover consortium samples are likely to
+contain contamination from support tissue and inflammatory and immune cells of
+the innate and adaptive immune systems, potentially harboring cancer supporting
+mutations [@bhowmick_review]. We think it will be instructive to explore these
+ideas in the context of single cell experiments.
 
@@
 converging on a single deadly phenotype that captures most or all of the
-critical features of end-stage cancer. If this is the case, once we discover the
-requirements of each cluster with respect to metastasis we can target them
-therapeutically and prevent further adaptation.
+critical features of end-stage cancer. This is consistent with the observation
+that tumors increase diversity of driver mutations in later stages [@PCAWG-EVO].
+When we discover the requirements of each cluster with respect to metastasis we
+can target them therapeutically and prevent further adaptation.

### 2020-07-22 code/figure2.Rmd
fixed scaling and text of figure 2. remove y-axis labels

@@
 <div style="text-align: right"> Figure 2 Data </div>
+```{r}
+library(ggplot2)
+library(patchwork)
+library(tidyverse)
+```
+
 ```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
@@
 
-```{r figure-2-plot, fig.width=72, fig.height=48, fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions.
+```{r figure-2-plot, fig.width=72, fig.height=60, fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions.
 Waterfall plot; columns represent tumors.
@@
   scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
   theme_classic() +
   theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
+        axis.text.y = element_blank(),
+        axis.title.x = element_blank(),
         axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5, size = 60),
+        plot.tag = element_text(size = 48, hjust = 0, vjust = 0),
         axis.ticks = element_blank(),
         axis.line = element_blank(),
-        legend.position = "none")
+        legend.position = "none") +
+  ggtitle("Pan-Cancer")
 
@@
   scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
   theme_classic() +
@@
         axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
+        axis.title.x = element_blank(),
         axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5, size = 60),
         axis.ticks = element_blank(),
         axis.line = element_blank(),
-        legend.position = "none") 
+        plot.tag = element_text(size = 48, hjust = 0, vjust = 0),
+        legend.position = "none") +
+  ggtitle(paste0("Class ", n))
 }
@@
 ```
+## oncoprint style of union set pathways
+```{r, eval = FALSE}
+tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
+load("../data/gdc_reactome_path_analysis.Rda")
+
+## getting the top 5 pathways from each cluster
+n_toppaths <- 5
+
+toppaths <- list() ## list to store data
+for(n in 1:max(tumor_clusters$clust_knn)) {
+  spec_class <- tumor_clusters %>% filter(clust_knn == n) %>% .$sample_id
+  toppaths[[n]] <- gdc_reactome_path_analysis %>% .[, names(.) %in% spec_class] %>% 
+    rowSums() %>% as.data.frame() %>%
+    'colnames<-' ("Freq") %>%
+    rownames_to_column(var = "pathname") %>%
+    arrange(desc(Freq)) %>%
+    mutate(clust = n) %>% .[1:n_toppaths,]
+}
+
+toppaths <- as.data.frame(do.call(rbind, toppaths))
+## union set of the top pathways from each cluster
+union_toppaths <- unique(toppaths$pathname)
+
+clust_col_order1 <- list()
+path_row_order1 <- list()
+
+for(n in 1:max(tumor_clusters$clust_knn)) {
+  class_spec_sample_id <- tumor_clusters %>% .[,c("sample_id", "clust_knn")] %>% filter(clust_knn == n) %>% .[,.$sample_id] ## get class spec sample ids
+  top10path <- gdc_reactome_path_analysis[union_toppaths, class_spec_sample_id]
+  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
+  top10path[top10path == "0"] <- " "
+  col <- c("MUT" = "firebrick3")
+heatmap_legend <- list(title = "", at = c("MUT"),
+                       labels = c("Mutation"))
+alter_fun = list(
+    background = function(x, y, w, h) {
+        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
+            gp = gpar(fill = "#CCCCCC", col = NA))
+    },
+    # bug red
+    MUT = function(x, y, w, h) {
+        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
+            gp = gpar(fill = col["MUT"], col = NA))
+    })
+  column_title <- "test"
+  onco_res <- oncoPrint(top10path,
+    alter_fun = alter_fun, col = col,
+    column_title = column_title, heatmap_legend_param = heatmap_legend,
+    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
+    top_annotation = NULL
+    )
+  clust_col_order1[[n]] <- names(top10path[,onco_res@column_order])
+  path_row_order1[[n]] <- rownames(top10path[onco_res@row_order,])
+}
+
+save(clust_col_order1, path_row_order1, file = "../data/onco_clust1.Rda")
+
+## getting tumor sample orders 
+for(n in 1) {
+df <- as.data.frame(clust_col_order1[[n]])
+names(df) <- "sample_id"
+df$cluster <- n
+for(p in 2:length(clust_col_order1)) {
+  df1 <- as.data.frame(clust_col_order1[[p]])
+  names(df1) <- "sample_id"
+  df1$cluster <- p
+  df <- rbind(df, df1)
+  rm(df1)
+}
+}
+
+## setting up for ggplot geom_tile
+union_pathways_analysis <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order1[[1]],] %>% 
+  t() %>% as.data.frame() %>% 
+  rownames_to_column(var = "sample_id") %>% 
+  gather(., key = path_name, value = "mutation_status", -sample_id) %>%
+  left_join(., tumor_clusters[,c("sample_id", "clust_knn")]) %>%
+  mutate(path_name = factor(.$path_name, levels = c(path_row_order1[[1]])), ## factor to force oncoprint order
+         sample_id = factor(.$sample_id, levels = c(rev(df$sample_id))))
+
+onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
+
+fig2_test <- ggplot(data = union_pathways_analysis, aes(y=sample_id, x=path_name)) + ## ggplot code
+  geom_tile(aes(fill = as.character(mutation_status))) +
+  scale_fill_manual(values = onco_color) +
+  theme_classic() +
+  theme(axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line = element_blank(),
+        legend.position = "none") +
+  facet_grid(clust_knn ~ ., scales = "free", space = "free", switch = "y")
+ggsave("../data/fig2_test.png")
+
+```

### 2020-07-23 code/figure2.Rmd

@@
 save(onco_ggplot, file = "onco_ggplot.Rda")
+
+###################
+## barplot panel ##
+###################
+
+load("../data/kelly.colours.rda")
+load("../data/moderate_enrich_prob.Rda")
+
+enrich_pathways <- moderate_enrich_prob %>%
+  group_by(clust) %>%
+  filter(difference > 0.3 & signif <= 2500) %>%
+  mutate(pathway = make_clean_names(path)) %>%
+  rename(clust_knn = "clust") %>%
+  select(clust_knn, pathway) %>%
+  mutate(clust_knn = as.numeric(clust_knn)) %>%
+  mutate(specific = TRUE)
+
+pathway_counts <- tumor_clusters %>%
+  left_join(t(gdc_reactome_path_analysis) %>% 
+              as.data.frame() %>% 
+              clean_names() %>% 
+              rownames_to_column("sample_id")) %>% 
+  group_by(clust_knn) %>%
+  summarise(total = n(), across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, sum)) %>%
+  mutate(across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, ~ .x / total)) %>%
+  select(-total) %>%
+  pivot_longer(-clust_knn, names_to = "pathway", values_to = "proportion") %>%
+  group_by(clust_knn) %>%
+  filter(pathway %in% enrich_pathways$pathway) %>%
+  left_join(enrich_pathways %>% select(clust_knn, pathway, specific), by = c("clust_knn", "pathway"))
+
+pathway_counts[is.na(pathway_counts$specific), ]$specific <- FALSE
+
+pathway_mat <- pathway_counts %>%
+  select(-specific) %>%
+  pivot_wider(names_from = pathway, values_from = proportion) %>%
+  ungroup() %>%
+  select(-clust_knn) %>%
+  as.matrix()
+
+pathway_dist <- dist(t(pathway_mat))
+pathway_clust <- as.hclust(rev(as.dendrogram(seriate(pathway_dist, method = "OLO_ward")[[1]])))
+pathway_counts$pathway <- factor(pathway_counts$pathway, levels = colnames(pathway_mat)[pathway_clust$order])
+bp.cluster.colors <- c(kelly.colours[3:12], "gray15")
+names(bp.cluster.colors) <- 1:11
+
+pathway_counts <- pathway_counts %>%
+  mutate(color = case_when(specific ~ clust_knn, TRUE ~ 11)) %>%
+  mutate(color = factor(color))
+save(pathway_counts, bp.cluster.colors, file = "barplot_pathways_fig2b.rda")
 
@@
   ggtitle("Pan-Cancer")
+
+load("../barplot_pathways_fig2b.rda")
+fig2B <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
+  geom_col() +
+  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
+  ylab("Proportion of samples mutated in pathway") +
+  xlab("Pathways enriched with mutations within clusters") +
+  theme_classic() +
+  scale_fill_manual(values = bp.cluster.colors) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    axis.ticks = element_blank(),
+    strip.text.y = element_text(size = 12, hjust = .5),
+    strip.text.y.right = element_text(angle = 0),
+    strip.background = element_blank(),
+    legend.position = "none"
+  )
+
+fig2B_rotate <- pathway_counts %>% ggplot(aes(y = pathway, x = proportion, fill = color)) +
+  geom_col() +
+  facet_wrap(~clust_knn, nrow = 1, strip.position = "top") +
+  xlab("Proportion of samples mutated in pathway") +
+  ylab("Pathways enriched with mutations within clusters") +
+  theme_classic() +
+  scale_fill_manual(values = bp.cluster.colors) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    axis.ticks = element_blank(),
+    strip.text.x = element_text(size = 12, hjust = .5),
+    strip.background = element_blank(),
+    legend.position = "none"
+  )
 

### 2020-07-23 text/manuscript.Rmd

@@
 library(FactoMineR)
+library(seriation)
 ```

### 2020-07-23 text/manuscript.Rmd
fix study source in first sent of results

--- 2020-07-21 text/manuscript.Rmd
@@
 
-To study cancer pathways we obtained a set of 7,607 tumor samples from GDC. We
-chose to start with somatic mutations in exome sequencing data because the
-affected gene is known almost unambiguously. Therefore, we focused on all
-missense, nonsense, frameshift, and splicing mutations. In order to minimize
-bias from well-studied diseases and processes, we selected 377 Reactome pathways
+To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
+Cancer Genome Atlas (TCGA) through the [Genomic Data Commons (GDC)](gdc.cancer.gov)
+portal. TCGA data because were
+most appropriate for our study given the relative completeness of the patient
+metadata, particularly for survival and staging. We chose to start with somatic
+mutations in exome sequencing data because the affected target gene is known 
+unambiguously. Therefore, we focused on all missense, nonsense, frameshift, and
+splicing mutations. In order to minimize bias from well-studied diseases and
+processes, we selected 377 Reactome pathways
 [reactome.org](https://reactome.org/) of interest corresponding to basic

### 2020-07-23 code/figure2.Rmd

--- 2020-07-21 code/figure2.Rmd
@@
 
+load("../barplot_pathways_fig2b.rda")
+fig2B <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
+  geom_col() +
+  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
+  ylab("Proportion of samples mutated in pathway") +
+  xlab("Pathways enriched with mutations within clusters") +
+  theme_classic() +
+  scale_fill_manual(values = bp.cluster.colors) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    axis.ticks = element_blank(),
+    strip.text.y = element_text(size = 12, hjust = .5),
+    strip.text.y.right = element_text(angle = 0),
+    strip.background = element_blank(),
+    legend.position = "none"
+  )
+
+fig2B_rotate <- pathway_counts %>% ggplot(aes(y = pathway, x = proportion, fill = color)) +
+  geom_col() +
+  facet_wrap(~clust_knn, nrow = 1, strip.position = "top") +
+  xlab("Proportion of samples mutated in pathway") +
+  ylab("Pathways enriched with mutations within clusters") +
+  theme_classic() +
+  scale_fill_manual(values = bp.cluster.colors) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    axis.ticks = element_blank(),
+    strip.text.x = element_text(size = 12, hjust = .5),
+    strip.background = element_blank(),
+    legend.position = "none"
+  )
+
 ## top10 pathways for each cluster

### 2020-07-23 code/supplements.Rmd

@@
 ### Supplementary Figure S1
-```{r}
-library(data.table)
-library(ggplot2)
-library(kableExtra)
-library(tidyverse)
-```
-
 ```{r supplementary-figure-s1, out.width = '90%', fig.align='center'}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
 
@@
 
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = project_code)) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) +
@@
 ```{r supplementary-figure-s2-data-generation, out.width = '90%', fig.align = "center"}
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
 umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
 
-ggplot(umap_cancertype, aes(x = x_2d, y = y_2d, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) +
@@
   column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "V1")  ## prep for left_join
+  t() %>% as.data.frame() %>% rownames_to_column(var = "sample_id")  ## prep for left_join
 
-selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_mut, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
+selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
+  left_join(., selected_mut, by = "sample_id") %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
   filter(value != 0)
 
-save(selected_mut, file = "selected_mut.Rda")
+save(selected_mut, file = "../data/selected_mut.Rda")
 ```
@@
 load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
 umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
 
-ggplot(selected_mut, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.25) +
+ggplot(selected_mut, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
   geom_point(size = 0.25) +
@@
 load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
 n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
@@
 
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join
+selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "sample_id") ## prepare pathway for left_join
 
-selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
-  left_join(., selected_path, by = "V1") %>%
-  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
+selected_path <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
+  left_join(., selected_path, by = "sample_id") %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
   filter(value != 0)
@@
 load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
+umap_cancertype <- fread("../data/umap_3d_coors.tsv")
 umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
@@
     mutate(variable = factor(variable, levels = c(pathway_layout)))
-  ggplot_umap <- ggplot(filtered_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-    geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.1) +
-    geom_point(size = 0.1) +
+  ggplot_umap <- ggplot(filtered_path, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+    geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+    geom_point(size = 0.25) +
     theme_classic() +
@@
       rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
+    rownames_to_column(var = "gene") %>%
       mutate(n_sample = length(gene_mutation),
              perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
+       arrange(desc(obs_count))
   top_gene_in_path <- list()
@@
 table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "table_st1.Rda")  
+save(table_st1, file = "../data/table_st1.Rda")  
 rm(list = ls(all.names = T))  

### 2020-07-23 code/figure2.Rmd
updated figure2 to include figure 3. removed figure 3 code from manuscript

@@
 <div style="text-align: right"> Figure 2 Data </div>
+```{r}
+library(ggplot2)
+library(patchwork)
+library(tidyverse)
+library(cowplot)
+```
+
 ```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
@@
 
-```{r figure-2-plot, fig.width=72, fig.height=48, fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions.
+```{r figure-2-plot, out.height='90%', out.width='90%',fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions.
 Waterfall plot; columns represent tumors.
@@
 Red signifies pathway disrupted pathways, grey signifies not disrupted."}
+
 ## top100 pathways pancancer oncoprint
@@
   scale_fill_manual(values = onco_color) +
-  xlab(label = "sample_id") +
-  labs(tag = "A") +
   theme_classic() +
+  xlab("Pan-Cancer (n=7,607 tumor samples)") +
+  ylab("377 pathways") +
   theme(axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 12),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 10),
+        axis.title.y = element_text(size = 10),
+        #plot.tag = element_text(size = 8, hjust = 0, vjust = 0),
         axis.ticks = element_blank(),
@@
 
-load("../barplot_pathways_fig2b.rda")
-fig2B <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
+load("../data/effect_heatmap.Rda") ## load data
+effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
+
+guide_axis_label_trans <- function(label_trans = identity, ...) {
+  axis_guide <- guide_axis(...)
+  axis_guide$label_trans <- rlang::as_function(label_trans)
+  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
+  axis_guide
+}
+
+guide_train.guide_axis_trans <- function(x, ...) {
+  trained <- NextMethod()
+  trained$key$.label <- x$label_trans(trained$key$.label)
+  trained
+}
+
+fig2B <- ggplot(data = effect_heatmap, aes(y = cluster, x = path_name)) +
+  geom_tile(size = 1, aes(fill = effect_size)) +
+  scale_fill_fermenter(
+    type = "div",
+    palette = "PuOr",
+    n.breaks = 12,
+    limits = c(-1, 1),
+    name = "Effect Size",
+    labels = c(-1, "", 0.5, "", "", 0, "", "", 0.5, "", 1)
+  ) +
+  facet_wrap(~cluster, ncol = 1, strip.position = "right", scales = "free_y") +
+  guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
+  xlab("377 pathways, ward.D2 clustered") +
+  ylab("Clusters") +
+  scale_x_discrete(position = "top") +
+  guides(y.sec = guide_axis_label_trans(~ paste(.x))) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    axis.ticks = element_blank(),
+    strip.text.y = element_text(size = 10, hjust = .5),
+    strip.text.y.right = element_text(angle = 0),
+    strip.background = element_blank(),
+    panel.background = element_blank(),
+    panel.spacing = unit(0, "lines"),
+    legend.position = "bottom",
+    legend.key.height = unit(0.3, "lines"),
+    legend.text = element_text(size = 5),
+    legend.title = element_text(size = 5),
+    legend.margin=margin(0,0,0,0),
+    legend.box.margin=margin(-10,-10,0,-10)
+  )
+
+load("../data/barplot_pathways_fig2b.rda")
+fig2C <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
   geom_col() +
   facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
-  ylab("Proportion of samples mutated in pathway") +
-  xlab("Pathways enriched with mutations within clusters") +
+  ylab("Proportion of mutated samples") +
+  xlab("Significant pathways") +
   theme_classic() +
@@
     axis.text.y = element_blank(),
-    axis.title.x = element_text(size = 12),
-    axis.title.y = element_text(size = 12),
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
     axis.ticks = element_blank(),
-    strip.text.y = element_text(size = 12, hjust = .5),
+    axis.line = element_blank(),
+    #strip.text.y = element_text(size = 10, hjust = .5),
+    #plot.tag = element_text(size = 12, hjust = 0, vjust = 0),
     strip.text.y.right = element_text(angle = 0),
@@
     legend.position = "none"
-  )
+  ) #+ 
+  ggtitle("Cluster-specific enrichment score")
 
-fig2B_rotate <- pathway_counts %>% ggplot(aes(y = pathway, x = proportion, fill = color)) +
-  geom_col() +
-  facet_wrap(~clust_knn, nrow = 1, strip.position = "top") +
-  xlab("Proportion of samples mutated in pathway") +
-  ylab("Pathways enriched with mutations within clusters") +
-  theme_classic() +
-  scale_fill_manual(values = bp.cluster.colors) +
-  theme(
-    axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_text(size = 12),
-    axis.title.y = element_text(size = 12),
-    axis.ticks = element_blank(),
-    strip.text.x = element_text(size = 12, hjust = .5),
-    strip.background = element_blank(),
-    legend.position = "none"
-  )
-
-## top10 pathways for each cluster
-load("../data/onco_clust.Rda") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
-
-plots_onco <- list() ## creating a list to store all the ggplots in the for loop below
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-for(n in 1:length(clust_col_order)) {
-  onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order[[n]],] %>% ## only include the top 10 paths
-  .[,names(.) %in% clust_col_order[[n]]] %>% ## only include specific tumor class
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order[[n]]))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order[[n]])))
-  plots_onco[[n]] <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + ## ggplot code
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  xlab(label = paste0("Class ", n)) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none") 
-}
-
-## setting patchwork panel layout
 layout <- '
-AABCD
-AABCD
-AABCD
-AAEFG
-AAEFG
-AAEFG
-AAHIJ
-AAHIJ
-AAHIJ
-AAKLM
-AAKLM
-AAKLM
+AAABBBB
+AAABBBB
+AAABBBB
+AAACCCC
+AAACCCC
+AAACCCC
+AAACCCC
 '
 
-fig2A + plots_onco[[1]] + plots_onco[[2]] + plots_onco[[3]] + plots_onco[[4]] + plots_onco[[5]] + plots_onco[[6]] + plots_onco[[7]] + plots_onco[[8]] + plots_onco[[9]] + plots_onco[[10]] + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
+fig2A + fig2B + fig2C + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
 ```

### 2020-07-23 text/discussion.Rmd

@@
 
-One of the biggest technical hurdles in tumor research is that the search space
+One of the biggest hurdles in cancer research is that the search space
 comprising ~20,000 protein-coding genes is comparable with the number of tumor
-samples, even with multiple mutations per sample
-[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
-that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
-elements exist with oncogenic or tumor suppressor potential in spite of many
-efforts detect non-random or selected mutation above the noise[@deesMuSiC2012;
+samples, even with multiple mutations per sample. And that is not accounting for
+the fact that ~30k non-coding genes and hundreds of thousands of non-coding
+regulatory elements exist with oncogenic or tumor suppressor potential
+[@deesMuSiC2012;
 @tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@@
 which we accomplished by translating gene-level mutations into pathway level
-disruptions.
+disruptions. Our approach differs from previously proposed methods [@PATHNET] in
+that we chose to focus on pathways as the unit of disruption instead of the
+gene, using the conceit that individual mutations may be sufficient to disrupt
+entire pathways (as is often the case with drivers). Rather than evaluating the
+burden of mutations on each pathway within or across tumors, which can be biased
+by the number of genes or the coverage of pathway-linked genes in the genome,
+and which we find to be relatively uniformly distributed, we scored each pathway
+in binary fashion, allowing one mutation or multiple mutations to count equally
+in a sample. To our knowledge, this approach has not previously been attempted
+in the literature in spite of its relative simplicity.
+
+We limited our analysis to mutations with likely deleterious effects, including
+missense, nonsense, frame-shift stop-gain and splice site mutations in genes
+that are actively expressed in each cancer type, thus avoiding bias from
+transcription coupled repair. Our method of filtration differs from the
+"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
+account for the same biases. We also filtered genes on the basis of pathway
+membership in biochemical pathways, excluding curated genesets related to
+diseases, syndromes, or classes of proteins ( _e.g._ transcription factors). We
+chose this approach to limit redundancy, which is already high in the remaining
+pathways for some driver genes, and to exclude biologically unrelated
+collections of genes. We do not account for non-coding mutations, amplification
+events or structural variants, though we hope to include them in subsequent
+studies (see below). Our pathway-centric analysis also allows us to circumvent
+the problem that some cancer-promoting mutations likely occur at low enough
+frequencies as to be indistinguishable from background. Attempts to capture
+these associations better using Bayesian networks have yielded important new
+insights into this problem and new genes that would not have been possible with
+traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the case
+that pathway-centric approaches like ours have the potential to increase
+statistical power.
 
@@
 introduced and later expanded on in a pair of classic essays by Hanahan and
-Weinberg [@weinberg200; @hallmarks_II]. In this view all tumors, regardless of
-tissue of origin, acquire 6 essential phenotypes; sustaining proliferative
+Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors regardless of
+tissue of origin acquire 6 essential phenotypes; sustaining proliferative
 signaling, evading growth suppressors, replicative immortality, evading
-apoptosis, supportive angiogenesis, and tissue invasion and metastasis
+apoptosis, supportive angiogenesis, tissue invasion and metastasis
 [@weinberg2000]. Advancements in the field over the ensuing decade led to the
-inclusion of new tumor promoting processes, namely deregulation of celullar
+inclusion of new tumor promoting processes, namely deregulation of cellular
 energy cycle, tumor promoting inflammation, evasion of the immune system, genome
@@
 of syndromes which are unique to each tissue-of-origin, that share some
-mechanisms and treatment strategies. Recent publication of Encode, Roadmap and TCGA
-consortium papers present a view largely consistent with the latter
-hypothesis. Perhaps owing to the intractable complexity of genomic data in all
-its forms, the inescapable conclusion thus far is that tissue-of-origin remains
-the most important driver of tumor characteristics at every scale and by every
-technique. Our observations contrast with this view, and instead support an
-interpretation of publicly available  data in which all tumors manifest one of a
-limited number of metastable phenotypes resulting from disruptions in basic
-pathways represented by our classes (**figure 1a-k**), with increased fitness
-regardless of tissue of origin, which then converge on end stage cancer by
-accumulating complementary mutations to the other classes (**figure 1l** and
-**Table 2**).
+mechanisms and treatment strategies. Recent publication of TCGA consortium
+papers present a view largely, and surprisingly, consistent with this hypothesis
+[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
+Perhaps owing to the intractable complexity of genomic, proteomic data and
+patient metadata in all its forms, the inescapable conclusion thus far is that
+tissue-of-origin remains the most important driver of tumor characteristics at
+every scale and by every measure. Our observations contrast with this view, and
+instead support an interpretation of publicly available data in which all tumors
+manifest one of a limited number of metastable phenotypes resulting from
+disruptions in basic pathways represented by our classes (**figure 1a-k**). Each
+of these classes may represent unique combinations of pathway disruptions with
+increased fitness regardless of tissue of origin, which then converge on end
+stage cancer by accumulating complementary mutations to the other classes
+(**figure 1l** and **Table 2**).
 

### 2020-07-24 text/discussion.Rmd
update discussion of histological subtypes, metastasis, add machine learning references

@@
 regulatory elements exist with oncogenic or tumor suppressor potential
-[@deesMuSiC2012;
-@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
+[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
+@lawrenceDiscoverySaturationAnalysis2014;
 @kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@@
 manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways represented by our classes (**figure 1a-k**). Each
+disruptions in basic pathways represented by our classes (**figure 1**). Each
 of these classes may represent unique combinations of pathway disruptions with
@@
 stage cancer by accumulating complementary mutations to the other classes
-(**figure 1l** and **Table 2**).
+(**figures 1L and M** and **Table 2**).
 
 We find that most tumors segregate into one of twelve biologically distinct
-clusters, regardless of the cancer type (**figure 1h** and **supplementary
+clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
 figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
-arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
-see **supplementary figure S1**) are unevenly distributed, though we could not
-identify cancer type that was exclusive to a single tumor class. Somewhat
+arm clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
+figure S1**) are unevenly distributed, though we could not identify cancer type
+that was exclusive to a single tumor class. Only cancer types with the fewest
+samples were found to be absent from one or more classes at all. Somewhat
 surprisingly to us, this finding extends to histological subtypes of breast,
-head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental
-Figure S2**). Indeed, it would be surprising if we did not observe bias for some
-cancers, as it has long been known that certain treatment regimens are likely to
-work for a significant fraction of patients of a given cancer type. Nonetheless,
-the basis for some tumors being treatment-refractory in spite of clinical
-indications remains elusive. Our analysis suggests the possibility of
-identifying more informative molecular, histological or cellular subtypes and
-could form a basis for future separation of patients into different precision
-treatment regimens.
+head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
+Figure S2**). This result implies that many, if not most, histological subtypes
+primarily reflect differences in cell-of-origin, rather than fundamental
+differences in proximal cause. It has been hypothesized that differences in the
+molecular regulators of development in the array of cell types present in the
+breast epithelium drive the various histological phenotypes [@breast-subtypes].
+Consistent with this view we found that breast tumor samples of the Luminal A
+subtype were heavily biased toward membership in clusters 1 and 7, and basal
+tumor samples were biased toward cluster 6 (**figure 1J**). However, both
+subtypes contained samples in every other class without exception, and Luminal B
+and Her2 positive samples are distributed across clusters. We interpret these
+data to mean that cell-of-origin molecular machinery can predispose certain cell
+types within the breast epithelium to forming tumors of one class or another but
+is not determinative; this view is compatible with the previously stated
+hypothesis but opens the way for a more granular view of individual tumors.
+Indeed, it would be surprising if we did not observe bias for some cancers and
+subtypes amongst our classes, since some treatment regimens have greater
+efficacy for a significant fraction of patients of a given cancer/histological
+subtype. Nonetheless, the basis for some tumors being treatment-refractory in
+spite of receiving the standard of clinical care for diagnostic markers remains
+elusive. Doubtless some of this is due to bad luck, as tumors can metastasis and
+remain dormant years before they are detected at distal sites [@hallmarks_II], or
+resistant clones may have already arisen at undetectable levels, but our
+analysis suggests the possibility of identifying more informative molecular,
+histological or cellular subtypes that could form a basis for future separation
+of patients into different precision treatment regimens.
 
@@
 projection exhibit increasing specificity. We identified no such trend (see
-**figure 1k**) and each class instead was associated with unique combinations of
-pathways (as in **figure 3**). In our view, these observations suggest that
+**figure 1H**) and each class instead was associated with unique combinations of
+pathways (as in **figures 2B, C**). In our view, these observations suggest that
 cancers have available to them a limited number of stable configurations with
@@
 Finally, we must remark on the limitations of our work exemplified in two of the
-classes, 1 and 8, which for which we did not find distinctive associations with
-pathways. Class 1 has a relatively low level of mutated pathways overall,
-although it is broadly enriched in many of the same tumor-promoting pathways
-common to the other groups. This likely does not reflect a true difference in
-stages of tumor evolution, as our data show clearly that this class is as likely
-to contain stage IV metastatic tumors as it is to contain those of stage I. We
-speculate this class represents a group of tumors with large aberrations in
-methylation, copy number or other genome rearrangements which we excluded from
-the present analysis. In the future, we hope to incorporate these other data
-into a pathway-centric analysis as we have done here for point mutations and
-indels. Now that we have established this as a feasible and potentially fruitful
-approach to understanding the cancer genome, it may be worth expanded effort in
-this area. In addition, Class 8 represents a group of hyper-mutated tumors that
-harbor so many mutations that virtually no pathway is unaffected. It seems
-likely that a significant fraction of the samples flagged as “mutant” for each
-pathway are actually unaffected, consistent with those samples have low
-information content. Similar to class 1 we speculate that a future analysis of
-other data types, including improved filtering of mutations across all classes,
-may resolve these tumors into improved classifications.
+classes, 1 and 10, which for which we did not find distinctive associations with
+pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
+pathways across the board, although it is broadly enriched in many of the same
+tumor-promoting pathways common to the other groups. This likely does not
+reflect a true difference in stages of tumor evolution, as our data show clearly
+that this class is as likely to contain stage IV metastatic tumors as it is to
+contain those of stage I (**figure 1L, M**). This class likely represents a
+group of tumors with large aberrations in methylation, copy number alterations,
+structural variants, or combinations thereof which we excluded from our
+analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
+proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
+skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
+represents a group of hyper-mutated tumors that harbor so many mutations that
+virtually no pathway is unaffected. It seems likely that a significant fraction
+of the samples flagged as “mutant” for each pathway are actually unaffected,
+consistent with those samples have low information content. This could be
+addressed with application of ever-increasingly sophisticated bioinformatics
+algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
+@deep-learning-clinical-variants]. In the future, we hope to incorporate these
+other data into a pathway-centric analysis as we have done here for point
+mutations and indels.
 
@@
 of our pathway disruption classes exhibit cancer-type specific effects on
-expected survival (**figure 5**). Contrary to what one might expect at the
+expected survival (**figure 4**). Contrary to what one might expect at the
 outset, we found that models assuming different survival based on class _and_
@@
 non-chemical treatments such as surgery and palliative care), we lack sufficient
-power in the current study to explore these variables in the GDC data. It is our
+power in the current study to explore these variables in the TCGA data. It is our
 hope that more focused future studies will help to distinguish between
@@
 characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
-disrupted in all or most of the defined classes (see **figures 2** and
-**supplementary figure S1**) and often are the result of aberrations involving
-common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
-complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
-Class-specific evolution must involve the acquisition of combinatorial
-disruptions to pathways that may individually be harmful but together produce
-more specialized cancer phenotypes and increased fitness (as in **figure 4** and
-**tables 1** and **ST2**). Importantly, our work does not distinguish between
-the order in which these mutations should accumulate. However, since many of the
-genes in the non-specific pathways involve genome stability, it is reasonable to
-hypothesize that such mutations promote the acquisition of class specific
-defects, and natural selection produces the classes we observed.
+disrupted in all or most of the defined classes (**supplementary figure S1**)
+and often are the result of aberrations involving common driver genes such as
+BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
+(**supplementary figure S3** and **supplemental file 1**). Class-specific
+evolution must involve the acquisition of combinatorial disruptions to pathways
+that may individually be harmful but together produce more specialized cancer
+phenotypes and increased fitness (as in **figure 4** and **tables 1** and
+**ST2**). Importantly, our work does not distinguish between the order in which
+these mutations should accumulate. This could be addressed in a future study by
+integrating our analysis with evolutionary analysis of clonality, drawing
+inferences from variant allele frequencies as in @PCAWG-EVO. However, since many
+of the genes in the non-cluster-specific pathways involve genome stability, it
+is reasonable to hypothesize that such mutations promote the acquisition of
+class-specific defects via random mutation and natural selection, thus producing
+the classes we observed. Indeed, a subset of known oncogenic driver mutations
+are highly enriched in early arising clones, whereas later arising clones have
+much greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes
+that are known to be responsible for discrete mutation signatures such as
+APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the varying
+selective pressures in different tissues [@PCAWG-MAIN].
 
 One of the drawbacks of bulk tumor whole genome sequencing data is the problem
-of tumor heterogeneity, both in terms of evolving subclones with different or
-passively diverging mutation profiles which then serve as a substrate for futher
-selection [@PCAWG-MAIN; @PCAWG-EVO]. Moreover consortium samples are likely to
+of tumor heterogeneity, thus consortium samples are likely to
 contain contamination from support tissue and inflammatory and immune cells of
@@
 
-We report that pathway enrichment across all classes (pooled) yielded generally
-lower effect sizes and larger p-values than the cluster specific analysis,
-suggesting that signal is diluted when the samples are analyzed this way, and
-supporting the view that metastasis has class-specific requirements. It should
-be noted, however, that in spite of this lack of global enrichment, some of the
-cluster specific pathways reflect known metastatic associations, _e.g._ pathways
-associated with asparagine and aspartate metabolism in class 10
-[@knott_asparagus]. Since the number of metastatic tumors is relatively low this
-part of our analysis is likely underpowered and subject to expanded analysis
-with larger cohorts. Nonetheless, the class-specific metastatic pathway
+We report that pathway enrichment in metastatic tumors across all classes
+(pooled) yielded generally lower effect sizes and larger p-values than the
+cluster specific analysis, suggesting that signal is diluted when the samples
+are analyzed this way, and supporting the view that metastasis has
+class-specific requirements. Since the number of metastatic tumors is relatively
+low this part of our analysis is likely underpowered and subject to expanded
+analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
 enrichment we observe (in **table 2**) could be a result of the pathways
 directly promoting metastatic phenotypes in these classes, or because they
-acquire a stronger or more complete oncogenic phenotype, since (again) we don’t
-know the order in which they were acquired in individual tumor samples. The fact
-that most metastatic enrichment is cluster-specific and has a tendency to
-overlap with cluster-specific pathways from non-metastatic tumors of neighboring
+acquire a stronger or more complete oncogenic phenotype, since we don’t know the
+order in which they were acquired in individual tumor samples. The fact that
+most metastatic enrichment is cluster-specific and has a tendency to overlap
+with cluster-specific pathways from non-metastatic tumors of neighboring
 clusters favors the latter interpretation, and suggests a mechanism of
@@
 converging on a single deadly phenotype that captures most or all of the
-critical features of end-stage cancer. This is consistent with the observation
-that tumors increase diversity of driver mutations in later stages [@PCAWG-EVO].
-When we discover the requirements of each cluster with respect to metastasis we
-can target them therapeutically and prevent further adaptation.
+critical features of end-stage cancer. Some clusters were characterized by
+cluster-specific pathways that seem superficially unrelated; for example cluster
+3 has mutations impacting protein translation and turnover, whereas cluster 5
+has defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
+traditional extra-cellular signaling pathways (as exemplified in **table 1**).
+It may be that the primary effects of mutations in polymer metabolism in cluster
+3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
+Clusters 3 and 5 are also marked by a decidedly different set of pathways in
+metastasis. Clusters 3 and 5 notably have (in addition to EGFR signaling)
+mutations in glucagon and glycogen metabolism, respectively, in their metastases
+(**table 2**), consistent with energy metabolism as one of the new hallmarks
+[@hallmarks_II], but also distinguishing these classes from the others in the
+order or nature of hallmark acquisition. We submit that when we discover the
+requirements of each cluster with respect to metastasis we may be able to target
+them therapeutically and prevent further adaptation.

### 2020-07-24 text/discussion.Rmd
edit discussion of metastasis

@@
 
-One of the biggest hurdles in cancer research is that the search space
-comprising ~20,000 protein-coding genes is comparable with the number of tumor
-samples, even with multiple mutations per sample. And that is not accounting for
-the fact that ~30k non-coding genes and hundreds of thousands of non-coding
-regulatory elements exist with oncogenic or tumor suppressor potential
-[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
+One of the biggest hurdles in cancer research is the size of the search space;
+~20,000 protein-coding genes is comparable with the number of tumor samples,
+even with multiple mutations per sample. And that is not accounting for the fact
+that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
+elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
+@tamboreroOncodriveCLUST2013;
 @lawrenceDiscoverySaturationAnalysis2014;
@@
 @zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
-investigate cancer as a disease of basic cellular and biochemical pathways,
-which we accomplished by translating gene-level mutations into pathway level
-disruptions. Our approach differs from previously proposed methods [@PATHNET] in
-that we chose to focus on pathways as the unit of disruption instead of the
-gene, using the conceit that individual mutations may be sufficient to disrupt
-entire pathways (as is often the case with drivers). Rather than evaluating the
-burden of mutations on each pathway within or across tumors, which can be biased
-by the number of genes or the coverage of pathway-linked genes in the genome,
-and which we find to be relatively uniformly distributed, we scored each pathway
-in binary fashion, allowing one mutation or multiple mutations to count equally
-in a sample. To our knowledge, this approach has not previously been attempted
-in the literature in spite of its relative simplicity.
-
-We limited our analysis to mutations with likely deleterious effects, including
-missense, nonsense, frame-shift stop-gain and splice site mutations in genes
-that are actively expressed in each cancer type, thus avoiding bias from
-transcription coupled repair. Our method of filtration differs from the
-"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
-account for the same biases. We also filtered genes on the basis of pathway
-membership in biochemical pathways, excluding curated genesets related to
-diseases, syndromes, or classes of proteins ( _e.g._ transcription factors). We
-chose this approach to limit redundancy, which is already high in the remaining
-pathways for some driver genes, and to exclude biologically unrelated
-collections of genes. We do not account for non-coding mutations, amplification
-events or structural variants, though we hope to include them in subsequent
-studies (see below). Our pathway-centric analysis also allows us to circumvent
-the problem that some cancer-promoting mutations likely occur at low enough
-frequencies as to be indistinguishable from background. Attempts to capture
-these associations better using Bayesian networks have yielded important new
-insights into this problem and new genes that would not have been possible with
-traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the case
-that pathway-centric approaches like ours have the potential to increase
+simplify the problem by investigating cancer as a disease of basic cellular and
+biochemical pathways, which we accomplished by translating gene-level mutations
+into pathway level disruptions. Our approach differs from previously proposed
+methods [@PATHNET] in that we chose to focus on the pathway as the unit of
+disruption instead of the gene, employing the conceit that individual mutations
+may be sufficient to disrupt entire pathways (as is often the case with
+drivers). Rather than evaluating the burden of mutations on each pathway within
+or across tumors, which can be biased by the number of genes or the coverage of
+pathway-linked genes in the genome, and which we find to be relatively uniformly
+distributed, we scored each pathway in binary fashion, allowing one mutation or
+multiple mutations to count equally for each sample. To our knowledge, this
+approach has not previously been attempted in the literature in spite of its
+relative simplicity. We limited our analysis to mutations with likely
+deleterious effects, including missense, nonsense, frame-shift, stop-gain and
+splice site mutations in genes that are actively expressed in each cancer type,
+thus avoiding bias from transcription coupled repair. Our method of filtration
+differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
+reasonable attempt to account for the same biases. We also filtered genes on the
+basis of pathway membership in biochemical pathways, excluding curated genesets
+related to diseases, syndromes, or classes of proteins ( _e.g._ transcription
+factors). We chose this approach to limit redundancy, which is already high in
+the remaining pathways for some driver genes, and to exclude biologically
+unrelated collections of genes. We do not account for non-coding mutations,
+amplification events or structural variants, though we hope to include them in
+subsequent studies (see below). Our pathway-centric analysis also allows us to
+circumvent the problem that some cancer-promoting mutations likely occur at low
+enough frequencies as to be indistinguishable from background. Attempts to
+capture these associations better using Bayesian networks have yielded important
+new insights into this problem and new genes that would not have been possible
+with traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the
+case that pathway-centric approaches like ours have the potential to increase
 statistical power.
 
-The hypothesis that cancer is the result of dysfunction from a limited number of
-basic cellular processes common to all living human cells was first
-introduced and later expanded on in a pair of classic essays by Hanahan and
-Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors regardless of
-tissue of origin acquire 6 essential phenotypes; sustaining proliferative
-signaling, evading growth suppressors, replicative immortality, evading
-apoptosis, supportive angiogenesis, tissue invasion and metastasis
+The hypothesis that cancer is the result of dysfunction from
+a limited number of basic cellular processes common to all living human cells
+was first introduced and later expanded on in a pair of classic essays by
+Hanahan and Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors
+regardless of tissue of origin acquire 6 essential phenotypes; sustaining
+proliferative signaling, evading growth suppressors, replicative immortality,
+evading apoptosis, supportive angiogenesis, tissue invasion and metastasis
 [@weinberg2000]. Advancements in the field over the ensuing decade led to the
@@
 manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways represented by our classes (**figure 1**). Each
-of these classes may represent unique combinations of pathway disruptions with
+disruptions in basic pathways represented by our classes (**figure 1**). Each of
+these classes may represent unique combinations of pathway disruptions with
 increased fitness regardless of tissue of origin, which then converge on end
@@
 spite of receiving the standard of clinical care for diagnostic markers remains
-elusive. Doubtless some of this is due to bad luck, as tumors can metastasis and
-remain dormant years before they are detected at distal sites [@hallmarks_II], or
-resistant clones may have already arisen at undetectable levels, but our
+elusive. Doubtless some of this is due to bad luck, as tumors can metastasize
+and remain dormant years before they are detected at distal sites, or resistant
+clones may have already arisen at undetectable levels [@hallmarks_II], but our
 analysis suggests the possibility of identifying more informative molecular,
@@
 
-By modeling CDC longevity data as a baseline risk function we showed that each
-of our pathway disruption classes exhibit cancer-type specific effects on
-expected survival (**figure 4**). Contrary to what one might expect at the
+By modeling CDC longevity data as a baseline risk function we showed
+that each of our pathway disruption classes exhibit cancer-type specific effects
+on expected survival (**figure 4**). Contrary to what one might expect at the
 outset, we found that models assuming different survival based on class _and_
@@
 non-chemical treatments such as surgery and palliative care), we lack sufficient
-power in the current study to explore these variables in the TCGA data. It is our
-hope that more focused future studies will help to distinguish between
+power in the current study to explore these variables in the TCGA data. It is
+our hope that more focused future studies will help to distinguish between
 treatment-specific effects or survival given different pathway disruption
@@
 
-One of the drawbacks of bulk tumor whole genome sequencing data is the problem
-of tumor heterogeneity, thus consortium samples are likely to
+One of the drawbacks of bulk tumor whole genome sequencing data
+is the problem of tumor heterogeneity, thus consortium samples are likely to
 contain contamination from support tissue and inflammatory and immune cells of
@@
 
-We report that pathway enrichment in metastatic tumors across all classes
-(pooled) yielded generally lower effect sizes and larger p-values than the
-cluster specific analysis, suggesting that signal is diluted when the samples
-are analyzed this way, and supporting the view that metastasis has
-class-specific requirements. Since the number of metastatic tumors is relatively
-low this part of our analysis is likely underpowered and subject to expanded
-analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
-enrichment we observe (in **table 2**) could be a result of the pathways
-directly promoting metastatic phenotypes in these classes, or because they
-acquire a stronger or more complete oncogenic phenotype, since we don’t know the
-order in which they were acquired in individual tumor samples. The fact that
-most metastatic enrichment is cluster-specific and has a tendency to overlap
-with cluster-specific pathways from non-metastatic tumors of neighboring
-clusters favors the latter interpretation, and suggests a mechanism of
-complementarity, in which newly acquired mutations result in similar classes
+We report that pathway enrichment in metastatic tumors
+across all classes (pooled) yielded generally lower effect sizes and larger
+p-values than the cluster specific analysis, suggesting that signal is diluted
+when the samples are analyzed this way, and supporting the view that metastasis
+has class-specific requirements. Since the number of metastatic tumors is
+relatively low this part of our analysis is likely underpowered and subject to
+expanded analysis with larger cohorts. Nonetheless, the class-specific
+metastatic pathway enrichment we observe (in **table 2**) could be a result of
+the pathways directly promoting metastatic phenotypes in these classes, or
+because they acquire a stronger or more complete oncogenic phenotype, since we
+don’t know the order in which they were acquired in individual tumor samples.
+The fact that most metastatic enrichment is cluster-specific and has a tendency
+to overlap with cluster-specific pathways from non-metastatic tumors of
+neighboring clusters favors the latter interpretation, and suggests a mechanism
+of complementarity, in which newly acquired mutations result in similar classes
 converging on a single deadly phenotype that captures most or all of the
@@
 traditional extra-cellular signaling pathways (as exemplified in **table 1**).
-It may be that the primary effects of mutations in polymer metabolism in cluster
-3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
-Clusters 3 and 5 are also marked by a decidedly different set of pathways in
-metastasis. Clusters 3 and 5 notably have (in addition to EGFR signaling)
-mutations in glucagon and glycogen metabolism, respectively, in their metastases
-(**table 2**), consistent with energy metabolism as one of the new hallmarks
-[@hallmarks_II], but also distinguishing these classes from the others in the
-order or nature of hallmark acquisition. We submit that when we discover the
-requirements of each cluster with respect to metastasis we may be able to target
-them therapeutically and prevent further adaptation.
+It may be that the primary effects of mutations in protein and RNA metabolism in
+clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
+pathways. Clusters 3 and 5 are also marked by a decidedly different set of
+pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
+pathways) mutations in glucagon and glycogen metabolism, respectively, in their
+metastases (**table 2**), consistent with energy metabolism as one of the new
+hallmarks [@hallmarks_II], but also distinguishing these classes from the others
+in the order or nature of hallmark acquisition. Thus, even at relatively low
+power, our analysis of metastasis further supports pathway-centric analysis of
+tumors as indicative of real differences in underlying biology. As a major
+caveat to this, we acknowledge that many therapies do target specific signaling
+pathways ( _e.g._ EGFR or HER2), but understanding the broader context of the
+genetic background and pathway vulnerability of tumors containing such markers
+may aid in creating smarter combination therapies. We submit that when we
+discover the requirements of each cluster with respect to pathway disruptions
+and metastasis we may be able to target them therapeutically and prevent further
+adaptation.

### 2020-07-25 code/supplements.Rmd
update figure references to reflect new kNN clusters, update figure 3 results, add APOBEC to supplement

@@
 ##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
+tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL", "APOBEC3B") ## list of selected tumor suppressor
 

### 2020-07-25 text/manuscript.Rmd
updated figure2 to include figure 3. removed figure 3 code from manuscript
update figure references to reflect new kNN clusters, update figure 3 results, add APOBEC to supplement

@@
 To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
-Cancer Genome Atlas (TCGA) through the [Genomic Data Commons (GDC)](gdc.cancer.gov)
-portal. TCGA data because were
-most appropriate for our study given the relative completeness of the patient
-metadata, particularly for survival and staging. We chose to start with somatic
-mutations in exome sequencing data because the affected target gene is known 
-unambiguously. Therefore, we focused on all missense, nonsense, frameshift, and
-splicing mutations. In order to minimize bias from well-studied diseases and
-processes, we selected 377 Reactome pathways
-[reactome.org](https://reactome.org/) of interest corresponding to basic
-cellular processes and biochemical pathways, excluding gene sets that correspond
-to miscellaneous categories ( _e.g._ “transcription factors”) or disease
-associations ( _e.g._ “mutated in colon cancer”) and filtered our gene list on
-membership in these pathways (<mark>total of x genes;</mark> see methods for
-details).
+Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
+(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
+study given the relative completeness of the patient metadata, particularly for
+survival and staging. We chose to start with somatic mutations in exome
+sequencing data because the affected target gene is known unambiguously.
+Therefore, we focused on all missense, nonsense, frameshift, and splicing
+mutations. In order to minimize bias from well-studied diseases and processes,
+we selected 377 Reactome pathways [reactome.org](https://reactome.org/) of
+interest corresponding to basic cellular processes and biochemical pathways,
+excluding gene sets that correspond to miscellaneous categories ( _e.g._
+“transcription factors”) or disease associations ( _e.g._ “mutated in colon
+cancer”) and filtered our gene list on membership in these pathways (<mark>total
+of x genes;</mark> see methods for details).
 
 It does not matter for our purposes whether a single tumor is “enriched” for
-multiple pathway mutations, chiefly because such metrics favor larger pathways (
-_i.e._ pathways with more member genes) over smaller ones, and non-expressed
-genes over expressed ones due to transcription coupled repair
+mutations of individual pathways, chiefly because such metrics favor larger
+pathways ( _i.e._ pathways with more member genes) over smaller ones. Data
+binarized by pathway are likely to be noisy for several reasons. First, point
+mutations can be deleterious (attenuating or hypomorphic) or activating
+(neomorphic) in genes that act functionally as either activators or repressors.
+Within the scope of this study we accept this uncertainty and assume mutations
+are generically disruptive to the pathway regardless of whether they are
+activating or deactivating or whether they affect a gene that we might classify
+as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, we
+know that low-expressed genes and non-expressed genes are known to accumulate
+mutations at a higher rate due to transcription coupled repair
 [@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
-@Pervasive_lesion_segregation; @expression-aware-annotation]. Data selected in
-this manner are likely to be noisy for several reasons. First, point mutations
-can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in
-genes that act functionally as either activators or repressors. Within the scope
-of this study we must accept this uncertainty and assume mutations are
-generically disruptive to the cell regardless of whether they are activating or
-deactivating or whether they affect a gene that we might classify as one thing
-or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, as previously
-noted, we know that low-expressed genes and non-expressed genes are known to
-accumulate mutations at a higher rate due to transcription coupled repair. To
-address this issue, we identified genes with low expression in each type of
-cancer and eliminated them for that cancer type only (see methods for details).
-<strike>Highly expressed genes also may have high mutation rates owing to
+@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
+issue, we identified genes with low expression in each type of cancer and
+eliminated them for that cancer type only (see methods for details).
+<strike>Highly expressed genes could also have high mutation rates owing to
 transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
@@
 biologically meaningful for predisposition to different classes of cancer
-however, and therefore chose to include these genes in our analysis.</strike>
-After selecting our pathways and genes, we then compiled a matrix of the
-pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
-member gene carries a mutation) to the pathway regardless of the number of genes
-carrying mutations (this data is visualized in **figure 1a**).
+however, and therefore chose not to exclude these genes from our analysis. After
+selecting our pathways and genes, we then compiled a matrix of the pathways,
+assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene
+carries a mutation) to the pathway regardless of the number of genes carrying
+mutations (**figure 1A**).
 
@@
 analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figure 1b**). Then we performed a UMAP
+containing useful information (**figures 1B, C**). Then we performed a UMAP
 analysis, both in order to summarize the MCA graphically, and as a preprocessing
@@
 reproducible regardless of random seed setting. A representative version of this
-UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
-mapping we attempted to define groupings of similar tumors within the spatial
-map using HDBSCAN, which performs hierarchical clustering and provides metrics
-of cluster stability and probabilities of cluster membership for each node.
-However, HDBSCAN is sensitive to several parameters; key for our analysis are
-the minimum number of tumor samples in a cluster that capture the maximum number
-of tumor samples, measured by probability of membership of $\geq 5\%$ in at
-least one class. Thus, we created a score metric as the fraction of classified
-tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
-to minimize the score function (**figure 1d**). HDBSCAN with these settings
-resulted in 12 distinct high-density clusters which we then projected onto the
-UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
-still resulted in a significant fraction of unclassified tumor samples. Since we
-ultimately wish to be able to classify any tumor using this scheme, we performed
-k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
-tumor in the set and then lets the $k$ most similar tumors "vote" as to the
-identity of the query tumor sample based on their classification labels. We set
-$k$ to be the square root of the number of tumor samples (87). Using this
-method, we assigned cluster membership to the remaining tumors (**figure 1f**).
+3D UMAP is shown in **figure 1D** as a 2D projection, rotated to enhance the
+visibility of the major features. Following this spatial mapping we attempted to
+define groupings of similar tumors within the spatial map using HDBSCAN, which
+performs hierarchical clustering and provides metrics of cluster stability and
+probabilities of cluster membership for each node. However, HDBSCAN is sensitive
+to several parameters; key for our analysis are the minimum number of tumor
+samples in a cluster that capture the maximum number of tumor samples, measured
+by probability of membership of $\geq 5\%$ in at least one class. Thus, we
+created a score metric as the fraction of classified tumors with max probability
+$< 5\%$ in one class and chose a cluster size of 95 to minimize the score
+function (**figure 1E**). HDBSCAN with these settings resulted in 12 distinct
+high-density clusters which we then projected onto the UMAP (**figure 1F**).
+This classified <mark>x out of 7,607 tumors</mark> but still resulted in a
+significant fraction of unclassified tumor samples. Since we ultimately wish to
+be able to classify any tumor using this scheme, we performed k Nearest
+Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
+the set and then lets the $k$ most similar tumors "vote" as to the identity of
+the query tumor sample based on their classification labels. We set $k$ to be
+the square root of the number of tumor samples (87). Using this method, we
+assigned cluster membership to the remaining tumors (**figure 1G**).
 
@@
 exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1h**
+cancer types had representative tumors in every class (for example **figure 1I**
 and full tumor profiles in **supplementary figure S1**, see also interactive
@@
 example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
-distinguished  by KRAS mutations among others. But even PAAD comprises tumors
+pancreatic adenocarcinoma (PAAD) was biased in class 8 & 9, which is
+distinguished  by KRAS mutations among others (**supplementary figure S3**). But even PAAD comprises tumors
 from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
@@
 prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
+pathway for these subtype annotations (**Figure 1J** and **supplemental
 file 1**). We also projected histological subtype data for the remaining cancers
@@
 break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
+breast, prostate and ovarian cancers. The mechanism of risk is thought to
 involve loss of heterozygosity [@brca-loh], so we projected the somatic
 mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1j**). We made
+segregation of these mutations into specific clusters (**figure 1K**). We made
 similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
@@
 pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1c**) could in
+structures on a common backbone of tumor samples (**figure 1D**) could in
 principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone culminates in a group of tumors (class 8) which
-has nearly every pathway disrupted. However, outside of class 8 we don't observe
-an obvious trend in the mutation count across the backbone of the UMAP (**figure
-1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected these data onto the
+this hypothesis, the backbone starts with a cluster of tumors (class 1) that has the fewest point mutations and culminates in a cluster (class 10) which
+has nearly every pathway disrupted (**figure 1H**). However, outside of class 10 we don't observe
+an obvious trend in the overall mutation burden across the backbone of the UMAP (**figure
+1H**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
+classes represent advancement through stages, we projected staging data onto the
 UMAP. Similar to tissue of origin and other categories of tumor, we did not
 observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1k**), suggesting that stage is not an determining factor in pathway
+(**figure 1L**), suggesting that stage is not an determining factor in pathway
 disruption classes.
@@
 loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness. To determine whether
+transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To determine whether
 any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
-again, we observed an even distribution across classes (**figure 1l**). This
+tissue types, we projected the metastasis data onto our UMAP, and surprisingly we observed an even distribution across classes (**figure 1M**). This
 final observation suggests that our pathway-disruption classification is
@@
 pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden.
+metastasis, regardless and independent of overall mutational burden (see Discussion for further analysis).
 
@@
 In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
-replacing genes with pathways (**figure 2**). A complete list of pathway
-disruptions with percent mutated samples and top genes is provided in
-**supplementary table ST1**). These reveal the broad importance of many well
-known pathways that are disrupted in cancer, including <mark>PI3K/Akt
-activation, DNA repair etc fix this with top 5 or so quoted pathway
-labels</mark>. The top third of pathways drop off gradually in overall mutation
-frequency, possibly owing to relative diversity among all tumors.
+(pan-cancer), we created waterfall graphs analagous to the oncoprint
+visualizations on cBioportal [@cbioportal], replacing genes with pathways
+(**figure 2A**). A complete list of pathway disruptions with percent mutated
+samples and top genes is provided in **supplementary table ST1**). As expected,
+these reveal the broad importance of many well known pathways that are disrupted
+in cancer, including "PIP3 activates Akt signaling" (77% of samples),
+"MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
+samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
+samples), _etc._ . The top third of pathways drop off gradually in overall
+mutation frequency (**figure 1A** and **supplementary table ST1**), possibly
+owing to relative diversity among all tumors.
 
@@
 and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
-identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
-class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
-class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
-0 pathways in class 8 (for which most pathways are putatively disrupted), 25
-pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
-pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
-classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
-had enrichment. To explore the specific pathways marking each class, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
-metabolic pathways including RNA and protein biosynthesis (**supplementary
-figure S4**). Most class 4 enriched pathways were not unique to this class but
-shared in class 10 & 12, suggesting that this grouping reflects either greater
-heterogeneity or possibly a lower stage of tumor evolution (as distinct from
-clinical ``stage''). Classes 9-12 have in common mutations in extracellular
-signaling, intracellular signaling, and immune-related pathways (see **figures
-3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
+visualized the enrichment as a heatmap (**figure 2B**). Using this approach, we
+identified 25 pathways highly enriched (enrichment score > 0.3, see methods) in
+class 2, 9 pathways enriched in class 3, 24 pathways in class 4, 10 pathways in
+class 5, 10 pathways in class 6, 24 pathways in class 7, 24 pathways in class 8
+(for which most pathways are putatively disrupted), 10 pathways in class 9, and
+0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
+terms of which pathways had enrichment. To explore the specific pathways marking
+each class, we projected disruptions for each of the 377 pathways onto the UMAP
+(**supplementary figure S4**). Classes 3 and 5 were distinguished by metabolic
+pathways including RNA and protein biosynthesis (**supplementary figure S4**).
+Similarly, class 4 was distinguished by mutations affecting regulation of DNA
+and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
+and "Nucleosome assembly"). Classes 7-9 have in common mutations in
+extracellular, intracellular, and immune-related signaling pathways (see
+**figure3** and **supplementary figure S4**). Cluster 2 had the highest pathway
 enrichment levels of the three, having mutations in hedgehog signaling,
-$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
-cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
-distinct of the 12, we chose to investigate these two groups further.
-<div style="text-align: right"> Figure 3 data </div>
-
-```{r child='../code/figure3.Rmd'}
-```
+"$\beta$-catenin degradation", "cellular response to hypoxia", "regulation of
+cell cycle" and "apoptosis" among others.
 
@@
 
-<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>
-
-Since pathways like these share many common genes, we investigated whether
-tumors in this class  have multiple mutations across signaling pathways, or
+Since pathways share some common driver genes, we investigated whether
+tumors in each class  have multiple mutations across signaling pathways, or
 whether a small number of driver genes account for enrichment. The split between
-clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
+clusters 7 and 8 is largely driven by mutations in _PI3K_ and its
 homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
-from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
+**supplementary figure S4**). Furthermore, cluster 1 is distinguished
+from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
 _TP53_ panel of **Supplementary Figure S4**).
 
-In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
-the highest pathway enrichment scores of the three major categories, though
-there is a significant number of tumors spatially distributed between cluster
-groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
-of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
-response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
-pre-replicative complex," "metabolism of polyamines," "degradation of
-$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
-non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
-destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
-with aspects of hedgehog signaling. To determine once again whether the genes
-driving pathway enrichment were commonly shared among these pathways we plotted
-and performed unsupervised clustering on tumors with an individual mutation
-matrix using every gene in each of the ten pathways, grouped by shared pathway
-membership (**figure 4**). Surprisingly, this class of tumors was not
+Cluster 2 exhibited the highest pathway enrichment scores of the ten major
+classes, although there is a significant number of tumors spatially distributed
+between cluster groupings 7-9 and 2 on the UMAP projection that have pathway
+characteristics of both. The top ten distinguishing pathways of class 2 tumors
+were "cellular response to hypoxia," "regulation of mitotic cell cycle,"
+"assembly of the pre-replicative complex," "metabolism of polyamines,"
+"degradation of $\beta$-catenin by the destruction complex", "tumor necrosis
+factor receptor 2 non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0)
+binds and destabilizes RNA", "regulation of apoptosis" and 3 other pathways
+associated with aspects of hedgehog signaling. To determine once again whether
+the genes driving pathway enrichment were commonly shared among these pathways
+we plotted and performed unsupervised clustering on tumors with an individual
+mutation matrix using every gene in each of the ten pathways, grouped by shared
+pathway membership (**figure 3A**). Surprisingly, this class of tumors was not
 predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity
-contributes to all the aforementioned pathways and the majority of tumors in
-these classes have a single mutation in one of these genes. Nonetheless, a
-significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
-proteasomal mutations but instead harbor mutations in genes representing a
-majority of the same pathways, suggesting that most if not all of the pathways
-are individually important and that proteasomal mutations may be sufficient to
-drive the net effect of the other mutations in this cluster. We annotated the
-tumors in this class according to tissue of origin to determine whether
-proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
-indeed we could observe some clustering. These data could explain how
-tissue-specific gene mutation signatures might result in general cancer
-phenotypes by converging at the pathway level.
+genes in the proteosomal pathway whose activity contributes to all the
+aforementioned pathways and the majority of tumors in these classes have a
+single mutation in one of these genes. We annotated the tumors in this class
+according to tissue of origin to determine whether proteasome mutations have a
+tissue-of-origin (or cancer-specific) bias, and indeed we could observe some
+clustering. These data could explain how tissue-specific gene mutation
+signatures might result in general cancer phenotypes by converging at the
+pathway level.
 
@@
 associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
-Second we looked at survival in mutations results results results (**figure 4c**). 
-Third, we checked the mutation classes and found the overwhelming
-majority (x out of y) are missense _vs._ various other deletarious classes (see
-**figure 4d**), suggesting that, similar to other oncogenic driver classes such
-as RAS and PI3K, proteasomal subunit mutations may result in specific
+"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 3B**).
+Second we looked at survival in mutations (**figure 3C**).
+Only a handful of proteasome mutations were significantly associated with favorable or unfavorable outcomes, probably owing to lack of sufficient power.
+
+Third, we checked the mutation classes using MAFtools [@maftools] and found the overwhelming
+majority (x out of y) are missense _vs._ various other deleterious classes (see
+**figure 3D**), suggesting that, similar to other oncogenic driver classes such
+as KRAS and PIK3C, proteasomal subunit mutations may result in specific
 dysregulation of the proteasomal complex activities that target cancer-promoting

### 2020-07-26 text/discussion.Rmd
update Breast histological subtypes discussion, add review references

@@
 Figure S2**). This result implies that many, if not most, histological subtypes
-primarily reflect differences in cell-of-origin, rather than fundamental
-differences in proximal cause. It has been hypothesized that differences in the
-molecular regulators of development in the array of cell types present in the
-breast epithelium drive the various histological phenotypes [@breast-subtypes].
-Consistent with this view we found that breast tumor samples of the Luminal A
-subtype were heavily biased toward membership in clusters 1 and 7, and basal
-tumor samples were biased toward cluster 6 (**figure 1J**). However, both
-subtypes contained samples in every other class without exception, and Luminal B
-and Her2 positive samples are distributed across clusters. We interpret these
-data to mean that cell-of-origin molecular machinery can predispose certain cell
-types within the breast epithelium to forming tumors of one class or another but
-is not determinative; this view is compatible with the previously stated
-hypothesis but opens the way for a more granular view of individual tumors.
-Indeed, it would be surprising if we did not observe bias for some cancers and
-subtypes amongst our classes, since some treatment regimens have greater
-efficacy for a significant fraction of patients of a given cancer/histological
-subtype. Nonetheless, the basis for some tumors being treatment-refractory in
-spite of receiving the standard of clinical care for diagnostic markers remains
-elusive. Doubtless some of this is due to bad luck, as tumors can metastasize
-and remain dormant years before they are detected at distal sites, or resistant
-clones may have already arisen at undetectable levels [@hallmarks_II], but our
-analysis suggests the possibility of identifying more informative molecular,
-histological or cellular subtypes that could form a basis for future separation
-of patients into different precision treatment regimens.
+could reflect differences in cell-of-origin, rather than fundamental differences
+in proximal cause. The five major subtypes of breast cancer have histological
+and molecular correlates that define them and how they respond to experimental
+stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
+molecular regulators of development in the precursor cell types present in the
+breast epithelium drive the various histological phenotypes [@breast-subtypes;
+@breast-subtypes3]. Consistent with this view we found that breast tumor samples
+of the Luminal A subtype were heavily biased toward membership in clusters 1 and
+7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
+However, both subtypes also contained samples in every other class (without
+exception), and Luminal B and Her2 positive samples are distributed across
+clusters. We interpret these data to mean that inherited cell-of-origin
+molecular profiles may predispose certain precursor cell types within the breast
+epithelium to forming tumors of one class or another but is not determinative;
+this view is compatible with the previously stated hypothesis but opens the way
+for a more granular view of individual tumors. It would be surprising if we did
+not observe bias for some cancers and subtypes amongst our classes, since some
+treatment regimens have greater efficacy for a significant fraction of patients
+of a given cancer/histological subtype [@breast-subtypes2]. Nonetheless, the
+basis for some tumors being treatment-refractory in spite of receiving the
+standard of clinical care for diagnostic markers remains elusive. Doubtless some
+of this is due to bad luck, as tumors can metastasize and remain dormant years
+before they are detected at distal sites, or resistant clones may have already
+arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
+possibility of identifying more informative molecular, histological or cellular
+subtypes that could form a basis for future separation of patients into
+different precision treatment regimens.
 

### 2020-07-27 text/manuscript.Rmd
Pull survival text out into separate survival.Rmd file, add a Makefile to build the document from the command line

@@
 
-If our molecular pathway disruption classes represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. To build on this relatively accurate model of baseline risk (see
-methods), we assume that individual cancer type captures some variation not
-accounted for in the longevity data ( _e.g._ all tissue-specific factors,
-including stage [see methods]), thus allowing us to avoid introducing
-uncertainty by modeling each of those factors explicitly. Therefore, our model
-estimates the effect of cancer type and class-specific cancer effects
-independently. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. Among the deadliest cancers outside
-of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD) cancers. Apart
-from these top 4 cancers which ranged from 6-22 in relative risk of age-adjusted
-death, the remaining cancer types ranged from about 1 to 5 in magnitude
-(**figure 5a**).
-
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types, where the null hypothesis corresponds to an
-effect size=1 (**figure 5b**). For several cancers ( _e.g._ PRAD, kidney
-chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer (THCA))
-the posterior estimates are largely indestinguishable from the prior, reflecting
-that either there were too few mortalities in the data to make an estimate (as
-expected for PRAD and THCA) or two few samples period. We did not observe
-class-specific trends that held true across cancer types, which could result
-from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit these models, they
-exhibited much greater uncertainty, suggesting an inferior specification.
-(<mark>Dan are there additional arguments to be made for the inferiority of the
-factored model, here?</mark>)
-
-```{r survival-model, eval=FALSE}
-
+```{r child='survival.Rmd'}
 ```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
 
@@
 fully reproducible manuscript. In addition, the R markdown files and all other
-models are available from repositories in the following locations on the
-distributed version software hub GitHub.
+models are available from
+[our repositories](https://github.com/dennishazelett/TTmanu) in the following
+locations on the distributed version control site GitHub
 

### 2020-07-27 text/manuscript.Rmd
pull out survival methods into seperate .Rmd

@@
 
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
-
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
-
-As an aside, we did not explicitly account for stage at diagnosis in our model
-for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
-pancreatic cancers). Secondly, such a model specification would suffer from
-added noise because the staging data are not well standardized across cancer
-types, have different criteria, and because it is unclear what the relationship
-between stage and advancement of disease is (for example some sub-stage 4 tumors
-are metastatic). To compound this latter issue, our tumor set is vastly
-under-powered given the uneven representation of stage across cancer types.
+```{r child='survival-methods.Rmd'}
+```
 

### 2020-07-27 text/survival-methods.Rmd
pull out survival methods into seperate .Rmd

We modeled baseline longevity in the U.S. with these different factors.
<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
Using this base survival model, we constructed a likelihood function with
parameters describing the overall lethality of each cancer type as a modifier
$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
taken as, roughly, the death rate relative to the baseline rates as depicted in
**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
twice as deadly.

With an accurate model for baseline longevity in place, we sought to model
cancer as an event based modifier of long-term survival. Thus any factors or
events such as cancer would modify the baseline risk function as a multiplier.
The resulting coefficients can thus be interpreted as increasing baseline risk
by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
female vs. male caucasians, blacks and asians (hispanics are not clearly marked
in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
data and baseline risk models). Thus we are able to accurately model baseline
risk of death taking these factors into account. The data themselves encompass
all other causes of death for these populations including cancer, thus freeing
our model from concerns about other confounding factors ( _e.g._ Diabetes). For
demonstration purposes we also illustrate some examples of the expected survival
for a population that receives a diagnosis with doubled risk of death at ages
40, 50, 60, 70, and 80 (**supplementary figure S6**).

As an aside, we did not explicitly account for stage at diagnosis in our model
for the following reasons. In the tumor sample data, stage at diagnosis is
confounded with cancer type because some cancers are screened agressively (
_e.g._ colorectal and prostate cancers) while others are diagnosed typically
after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
pancreatic cancers). Secondly, such a model specification would suffer from
added noise because the staging data are not well standardized across cancer
types, have different criteria, and because it is unclear what the relationship
between stage and advancement of disease is (for example some sub-stage 4 tumors
are metastatic). To compound this latter issue, our tumor set is vastly
under-powered given the uneven representation of stage across cancer types.

### 2020-07-27 text/discussion.Rmd
fix bug in references, edit discussion

@@
+---
+title: "Discussion"
+output: 
+  bookdown::html_document2:
+    fig_captions: TRUE
+    number_sections: FALSE
+    toc: FALSE
+    code_folding: hide
+    theme: flatly
+    smart: true
+bibliography: references.bib
+editor_options: 
+  chunk_output_type: console
+---
+
 ### Classification of tumors independent of tissue-of-origin
@@
 circumvent the problem that some cancer-promoting mutations likely occur at low
-enough frequencies as to be indistinguishable from background. Attempts to
-capture these associations better using Bayesian networks have yielded important
-new insights into this problem and new genes that would not have been possible
-with traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the
-case that pathway-centric approaches like ours have the potential to increase
-statistical power.
+enough frequencies as to be indistinguishable from background. Previous attempts
+to capture these associations better using Bayesian networks have yielded
+important new insights into this problem and new genes
+[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
+traditional analysis, supporting the case that pathway-centric approaches like
+ours have the potential to increase statistical power.
 
@@
 Hanahan and Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors
-regardless of tissue of origin acquire 6 essential phenotypes; sustaining
+regardless of tissue of origin acquire six essential phenotypes; sustaining
 proliferative signaling, evading growth suppressors, replicative immortality,
@@
 [@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
-Perhaps owing to the intractable complexity of genomic, proteomic data and
+Perhaps owing to the intractable complexity of genomics, proteomics, and
 patient metadata in all its forms, the inescapable conclusion thus far is that
@@
 manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways represented by our classes (**figure 1**). Each of
-these classes may represent unique combinations of pathway disruptions with
-increased fitness regardless of tissue of origin, which then converge on end
-stage cancer by accumulating complementary mutations to the other classes
-(**figures 1L and M** and **Table 2**).
+disruptions in basic pathways represented by our classes (**figure 1**).
+
+We attempted to account for our clusters in terms of more trivial explanations.
+For example, it could be that the classes we identified are consistent with a
+signature of overall disease progression, such that successive, adjacent classes
+on the UMAP projection exhibit increasing specificity. We identified no such
+trend in the number of mutations (see **figure 1H**), the relative staging
+(**figure 1L**) or metastasis (**figure 1M**); and each class instead was
+associated with unique combinations of pathways (as in **figures 2B, C**).
 
@@
 figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
-arm clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
-figure S1**) are unevenly distributed, though we could not identify cancer type
-that was exclusive to a single tumor class. Only cancer types with the fewest
+mutant clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
+figure S1**) are unevenly distributed, though we could not identify any cancers
+that were exclusive to a single tumor class. Only cancer types with the fewest
 samples were found to be absent from one or more classes at all. Somewhat
@@
 could reflect differences in cell-of-origin, rather than fundamental differences
-in proximal cause. The five major subtypes of breast cancer have histological
-and molecular correlates that define them and how they respond to experimental
+in proximal cause. The four major subtypes of breast cancer correspond to histological
+and molecular expression profiles that define them and how they respond to experimental
 stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
@@
 not observe bias for some cancers and subtypes amongst our classes, since some
-treatment regimens have greater efficacy for a significant fraction of patients
-of a given cancer/histological subtype [@breast-subtypes2]. Nonetheless, the
+treatment regimens have greater efficacy for patients
+of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
 basis for some tumors being treatment-refractory in spite of receiving the
@@
 different precision treatment regimens.
-
-It could be that the classes we identified are consistent with a signature of
-overall disease progression, such that successive, adjacent classes on the UMAP
-projection exhibit increasing specificity. We identified no such trend (see
-**figure 1H**) and each class instead was associated with unique combinations of
-pathways (as in **figures 2B, C**). In our view, these observations suggest that
-cancers have available to them a limited number of stable configurations with
-respect to pathway disruption which are consistent with increased fitness and
-therefore positive selection.
 
@@
 contain those of stage I (**figure 1L, M**). This class likely represents a
-group of tumors with large aberrations in methylation, copy number alterations,
+group of tumors with aberrations in methylation, copy number alterations,
 structural variants, or combinations thereof which we excluded from our
@@
 @deep-learning-clinical-variants]. In the future, we hope to incorporate these
-other data into a pathway-centric analysis as we have done here for point
-mutations and indels.
+other data into a comprehensive pathway-centric analysis as we have done here
+for point mutations and indels.
 
@@
 
-By modeling CDC longevity data as a baseline risk function we showed
-that each of our pathway disruption classes exhibit cancer-type specific effects
-on expected survival (**figure 4**). Contrary to what one might expect at the
+By modeling CDC longevity data as a baseline risk function we showed that each
+of our pathway disruption classes exhibit cancer-type specific effects on
+expected survival (**figure 4**). Contrary to what one might expect at the
 outset, we found that models assuming different survival based on class _and_
@@
 treatment can very between cancers, it makes sense in hindsight that they should
-all be different. Contrast the situation with ovarian vs breast cancer, which
+all be different. Contrast the situation with ovarian _vs._ breast cancer, which
 are both hormonally driven cancers, for example. Ovarian cancer has but one main
-treatment, platinum, whereas breast cancer patients have a variety of treatment
-regimens based on molecular subtype and other factors. It is easy to see how
-standard of care for one cancer and class could produce different outcomes than
-the same class of a different cancer type under the circumstances.
-Unfortunately, given the diversity drug classes and treatments (including
-non-chemical treatments such as surgery and palliative care), we lack sufficient
-power in the current study to explore these variables in the TCGA data. It is
-our hope that more focused future studies will help to distinguish between
-treatment-specific effects or survival given different pathway disruption
-classes.
+treatment axis, platinum, whereas breast cancer patients have a variety of
+treatment regimens based on molecular subtype and other factors. It is easy to
+see how standard of care for one cancer and class could produce different
+outcomes than the same class of a different cancer type. Unfortunately, given
+the diversity drug classes and treatments (including non-chemical treatments
+such as surgery and palliative care), we lack sufficient power in the current
+study to explore these variables in the TCGA data. It is our hope that more
+focused future studies will help to distinguish between treatment-specific
+effects or survival given different pathway disruption classes.
 
@@
 (**supplementary figure S3** and **supplemental file 1**). Class-specific
-evolution must involve the acquisition of combinatorial disruptions to pathways
-that may individually be harmful but together produce more specialized cancer
-phenotypes and increased fitness (as in **figure 4** and **tables 1** and
-**ST2**). Importantly, our work does not distinguish between the order in which
-these mutations should accumulate. This could be addressed in a future study by
-integrating our analysis with evolutionary analysis of clonality, drawing
-inferences from variant allele frequencies as in @PCAWG-EVO. However, since many
-of the genes in the non-cluster-specific pathways involve genome stability, it
-is reasonable to hypothesize that such mutations promote the acquisition of
+evolution must involve the acquisition of disruptions to pathways that may
+individually be harmful ( _e.g._ highly proliferative cells are more likely to
+senesce) but together produce more specialized cancer phenotype and increased
+fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our work
+does not distinguish between the order in which these mutations should
+accumulate. This could be addressed in a future study by integrating our
+analysis with evolutionary analysis of clonality, drawing inferences from
+variant allele frequencies as in @PCAWG-EVO. However, since many of the genes in
+the non-cluster-specific pathways involve genome stability, it is reasonable to
+hypothesize that these mutations promote or enable the acquisition of
 class-specific defects via random mutation and natural selection, thus producing
-the classes we observed. Indeed, a subset of known oncogenic driver mutations
-are highly enriched in early arising clones, whereas later arising clones have
-much greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes
-that are known to be responsible for discrete mutation signatures such as
-APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the varying
-selective pressures in different tissues [@PCAWG-MAIN].
+the classes we observed. In support of this, the pan-cancer analysis of whole
+genomes consortium (PCAWG) found that a subset of known oncogenic driver
+mutations are highly enriched in early arising clones, whereas later arising
+clones have much greater diversity in driver mutations [@PCAWG-EVO]. Moreover,
+driver genes that are known to be responsible for discrete mutation signatures
+such as APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the
+varying selective pressures in different tissues [@PCAWG-MAIN].
 
-One of the drawbacks of bulk tumor whole genome sequencing data
-is the problem of tumor heterogeneity, thus consortium samples are likely to
-contain contamination from support tissue and inflammatory and immune cells of
-the innate and adaptive immune systems, potentially harboring cancer supporting
-mutations [@bhowmick_review]. We think it will be instructive to explore these
-ideas in the context of single cell experiments.
+One of the drawbacks of bulk tumor whole genome sequencing data is the problem
+of tumor heterogeneity. Consortium samples are likely to contain contamination
+from support tissue and inflammatory and immune cells of the innate and adaptive
+immune systems, potentially harboring cancer supporting mutations
+[@bhowmick_review]. We think it will be instructive to explore these ideas in
+the context of single cell experiments.
 
@@
 
-We report that pathway enrichment in metastatic tumors
-across all classes (pooled) yielded generally lower effect sizes and larger
-p-values than the cluster specific analysis, suggesting that signal is diluted
-when the samples are analyzed this way, and supporting the view that metastasis
-has class-specific requirements. Since the number of metastatic tumors is
-relatively low this part of our analysis is likely underpowered and subject to
-expanded analysis with larger cohorts. Nonetheless, the class-specific
-metastatic pathway enrichment we observe (in **table 2**) could be a result of
-the pathways directly promoting metastatic phenotypes in these classes, or
-because they acquire a stronger or more complete oncogenic phenotype, since we
-don’t know the order in which they were acquired in individual tumor samples.
-The fact that most metastatic enrichment is cluster-specific and has a tendency
-to overlap with cluster-specific pathways from non-metastatic tumors of
-neighboring clusters favors the latter interpretation, and suggests a mechanism
-of complementarity, in which newly acquired mutations result in similar classes
+We report that pathway enrichment in metastatic tumors across all classes
+(pooled) yielded generally lower effect sizes and larger p-values than the
+cluster specific analysis, suggesting that signal is diluted when the samples
+are analyzed this way, and supporting the view that metastasis has
+class-specific requirements. Since the number of metastatic tumors is relatively
+low this part of our analysis is likely underpowered and subject to expanded
+analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
+enrichment we observe (in **table 2**) could be a result of the pathways
+directly promoting metastatic phenotypes in these classes, or because they
+acquire a stronger or more complete oncogenic phenotype, as we don’t know the
+order in which they were acquired in individual tumor samples. The fact that
+most metastatic enrichment is cluster-specific and has a tendency to overlap
+with cluster-specific pathways from non-metastatic tumors of neighboring
+clusters favors the latter interpretation, and suggests a mechanism of
+complementarity, in which newly acquired mutations result in similar classes
 converging on a single deadly phenotype that captures most or all of the
-critical features of end-stage cancer. Some clusters were characterized by
-cluster-specific pathways that seem superficially unrelated; for example cluster
-3 has mutations impacting protein translation and turnover, whereas cluster 5
-has defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
-traditional extra-cellular signaling pathways (as exemplified in **table 1**).
-It may be that the primary effects of mutations in protein and RNA metabolism in
-clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
-pathways. Clusters 3 and 5 are also marked by a decidedly different set of
-pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
-pathways) mutations in glucagon and glycogen metabolism, respectively, in their
-metastases (**table 2**), consistent with energy metabolism as one of the new
-hallmarks [@hallmarks_II], but also distinguishing these classes from the others
-in the order or nature of hallmark acquisition. Thus, even at relatively low
-power, our analysis of metastasis further supports pathway-centric analysis of
-tumors as indicative of real differences in underlying biology. As a major
-caveat to this, we acknowledge that many therapies do target specific signaling
-pathways ( _e.g._ EGFR or HER2), but understanding the broader context of the
-genetic background and pathway vulnerability of tumors containing such markers
-may aid in creating smarter combination therapies. We submit that when we
-discover the requirements of each cluster with respect to pathway disruptions
-and metastasis we may be able to target them therapeutically and prevent further
-adaptation.
+critical features of end-stage cancer. One explanation for this could be that
+tumors must accumulate mutations in a certain order or face elimination due to
+any of the selective bottlenecks they face (senescence, the immune system,
+_etc._) [@hallmarks_II]. Some clusters were characterized by cluster-specific
+pathways that seem superficially unrelated; for example cluster 3 has mutations
+impacting protein translation and turnover, whereas cluster 5 has defects in RNA
+metabolism and clusters 7, 8 and 9 have defects in the more traditional
+extra-cellular signaling pathways (as exemplified in **table 1**). It may be
+that the primary effects of mutations in protein and RNA metabolism in clusters
+3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
+Clusters 3 and 5 are also marked by a decidedly different set of pathways in
+metastasis. Clusters 3 and 5 notably have (in addition to signaling pathways)
+mutations in glucagon and glycogen metabolism, respectively, in their metastases
+(**table 2**), consistent with energy metabolism as one of the new hallmarks
+[@hallmarks_II], but also distinguishing these classes from the others in the
+order or nature of hallmark acquisition. Thus, even at relatively low power, our
+analysis of metastasis further supports pathway-centric analysis of tumors as
+indicative of real differences in underlying biology. As a major caveat to this,
+we acknowledge that many therapies _do_ target specific signaling pathways (
+_e.g._ EGFR or HER2), but understanding the broader context of the genetic
+background and pathway vulnerability of tumors containing such markers may aid
+in creating smarter combination therapies. We submit that when we discover the
+requirements of each cluster with respect to pathway disruptions and metastasis
+we may be able to target them therapeutically and prevent further adaptation.

### 2020-07-27 text/discussion.Rmd

@@
 distributed, we scored each pathway in binary fashion, allowing one mutation or
-multiple mutations to count equally for each sample. To our knowledge, this
-approach has not previously been attempted in the literature in spite of its
-relative simplicity. We limited our analysis to mutations with likely
-deleterious effects, including missense, nonsense, frame-shift, stop-gain and
-splice site mutations in genes that are actively expressed in each cancer type,
-thus avoiding bias from transcription coupled repair. Our method of filtration
-differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
-reasonable attempt to account for the same biases. We also filtered genes on the
-basis of pathway membership in biochemical pathways, excluding curated genesets
-related to diseases, syndromes, or classes of proteins ( _e.g._ transcription
-factors). We chose this approach to limit redundancy, which is already high in
-the remaining pathways for some driver genes, and to exclude biologically
-unrelated collections of genes. We do not account for non-coding mutations,
-amplification events or structural variants, though we hope to include them in
-subsequent studies (see below). Our pathway-centric analysis also allows us to
-circumvent the problem that some cancer-promoting mutations likely occur at low
-enough frequencies as to be indistinguishable from background. Previous attempts
-to capture these associations better using Bayesian networks have yielded
-important new insights into this problem and new genes
-[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
-traditional analysis, supporting the case that pathway-centric approaches like
-ours have the potential to increase statistical power.
+multiple mutations to count equally for each sample.
+
+To our knowledge, this approach has not previously been attempted in the
+literature in spite of its relative simplicity. We limited our analysis to
+mutations with likely deleterious effects, including missense, nonsense,
+frame-shift, stop-gain and splice site mutations in genes that are actively
+expressed in each cancer type, thus avoiding bias from transcription coupled
+repair. Our method of filtration differs from the "rank-and-cut" method of
+[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
+biases. We also filtered genes on the basis of pathway membership in biochemical
+pathways, excluding curated genesets related to diseases, syndromes, or classes
+of proteins ( _e.g._ transcription factors). We chose this approach to limit
+redundancy, which is already high in the remaining pathways for some driver
+genes, and to exclude biologically unrelated collections of genes. We do not
+account for non-coding mutations, amplification events or structural variants,
+though we hope to include them in subsequent studies (see below). Our
+pathway-centric analysis also allows us to circumvent the problem that some
+cancer-promoting mutations likely occur at low enough frequencies as to be
+indistinguishable from background. Previous attempts to capture these
+associations better using Bayesian networks have yielded important new insights
+into this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that
+would not have been possible with traditional analysis, supporting the case that
+pathway-centric approaches like ours have the potential to increase statistical
+power.
 
@@
 was first introduced and later expanded on in a pair of classic essays by
-Hanahan and Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors
+Hanahan and Weinberg [@weinberg2000; @hallmarks_II]. In this view, all tumors
 regardless of tissue of origin acquire six essential phenotypes; sustaining
@@
 concept of the tumor microenvironment and its contribution to the evolution and
-promotion of metastatic tumorigenesis [@hallmarks_II].
+promotion of metastatic disease [@hallmarks_II].
 
@@
 molecular profiles may predispose certain precursor cell types within the breast
-epithelium to forming tumors of one class or another but is not determinative;
+epithelium to forming tumors of one class or another but are not determinative;
 this view is compatible with the previously stated hypothesis but opens the way
@@
 Finally, we must remark on the limitations of our work exemplified in two of the
-classes, 1 and 10, which for which we did not find distinctive associations with
+classes, 1 and 10, for which we did not find distinctive associations with
 pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
@@
 senesce) but together produce more specialized cancer phenotype and increased
-fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our work
-does not distinguish between the order in which these mutations should
-accumulate. This could be addressed in a future study by integrating our
-analysis with evolutionary analysis of clonality, drawing inferences from
-variant allele frequencies as in @PCAWG-EVO. However, since many of the genes in
-the non-cluster-specific pathways involve genome stability, it is reasonable to
+fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our
+observations do not imply the order in which these mutations should accumulate.
+This could be addressed in a future study by integrating our analysis with
+evolutionary analysis of clonality, drawing inferences from variant allele
+frequencies as in @PCAWG-EVO. However, since many of the genes in the
+non-cluster-specific pathways involve genome stability, it is reasonable to
 hypothesize that these mutations promote or enable the acquisition of
@@
 of tumor heterogeneity. Consortium samples are likely to contain contamination
-from support tissue and inflammatory and immune cells of the innate and adaptive
-immune systems, potentially harboring cancer supporting mutations
+from support tissue, stroma, inflammatory, and immune cells of the innate and
+adaptive immune systems, all potentially harboring cancer supporting mutations
 [@bhowmick_review]. We think it will be instructive to explore these ideas in
@@
 cluster specific analysis, suggesting that signal is diluted when the samples
-are analyzed this way, and supporting the view that metastasis has
-class-specific requirements. Since the number of metastatic tumors is relatively
-low this part of our analysis is likely underpowered and subject to expanded
-analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
-enrichment we observe (in **table 2**) could be a result of the pathways
-directly promoting metastatic phenotypes in these classes, or because they
-acquire a stronger or more complete oncogenic phenotype, as we don’t know the
-order in which they were acquired in individual tumor samples. The fact that
-most metastatic enrichment is cluster-specific and has a tendency to overlap
-with cluster-specific pathways from non-metastatic tumors of neighboring
-clusters favors the latter interpretation, and suggests a mechanism of
-complementarity, in which newly acquired mutations result in similar classes
-converging on a single deadly phenotype that captures most or all of the
-critical features of end-stage cancer. One explanation for this could be that
-tumors must accumulate mutations in a certain order or face elimination due to
-any of the selective bottlenecks they face (senescence, the immune system,
-_etc._) [@hallmarks_II]. Some clusters were characterized by cluster-specific
-pathways that seem superficially unrelated; for example cluster 3 has mutations
-impacting protein translation and turnover, whereas cluster 5 has defects in RNA
-metabolism and clusters 7, 8 and 9 have defects in the more traditional
-extra-cellular signaling pathways (as exemplified in **table 1**). It may be
-that the primary effects of mutations in protein and RNA metabolism in clusters
-3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
-Clusters 3 and 5 are also marked by a decidedly different set of pathways in
-metastasis. Clusters 3 and 5 notably have (in addition to signaling pathways)
-mutations in glucagon and glycogen metabolism, respectively, in their metastases
-(**table 2**), consistent with energy metabolism as one of the new hallmarks
-[@hallmarks_II], but also distinguishing these classes from the others in the
-order or nature of hallmark acquisition. Thus, even at relatively low power, our
-analysis of metastasis further supports pathway-centric analysis of tumors as
-indicative of real differences in underlying biology. As a major caveat to this,
-we acknowledge that many therapies _do_ target specific signaling pathways (
-_e.g._ EGFR or HER2), but understanding the broader context of the genetic
-background and pathway vulnerability of tumors containing such markers may aid
-in creating smarter combination therapies. We submit that when we discover the
-requirements of each cluster with respect to pathway disruptions and metastasis
-we may be able to target them therapeutically and prevent further adaptation.
+are analyzed together, and supporting the view that metastasis has
+class-specific requirements. Since the number of metastatic samples is
+relatively low this part of our analysis is likely underpowered and subject to
+expanded analysis with larger cohorts. Nonetheless, the class-specific
+metastatic pathway enrichment we observe (in **table 2**) could be a result of
+the pathways directly promoting metastatic phenotypes in these classes, or
+because they acquire a stronger or more complete oncogenic phenotype, as we
+don’t know the order in which they were acquired. The fact that most metastatic
+enrichment is cluster-specific and has a tendency to overlap with
+cluster-specific pathways from non-metastatic tumors of neighboring clusters
+favors the latter interpretation, and suggests a mechanism of complementarity,
+in which newly acquired mutations result in similar classes converging on a
+single deadly phenotype that captures most of the critical features of end-stage
+cancer. One explanation for this could be that tumors must accumulate mutations
+in a certain order or face elimination due to any of the selective bottlenecks
+they face (senescence, the immune system, _etc._) [@hallmarks_II]. Some clusters
+were characterized by cluster-specific pathways that seem superficially
+unrelated; for example cluster 2 has mutations in the proteasome subunits,
+cluster 3 has mutations impacting protein translation, whereas cluster 5 has
+defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
+traditional extra-cellular signaling pathways (as exemplified in **table 1**).
+It may be that the primary effects of mutations in protein and RNA metabolism in
+clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
+pathways. Clusters 3 and 5 are also marked by a decidedly different set of
+pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
+pathways) mutations in glucagon and glycogen metabolism, respectively, in their
+metastases (**table 2**), consistent with energy metabolism as one of the new
+hallmarks [@hallmarks_II], but also distinguishing these classes from the others
+in the order or nature of hallmark acquisition. Thus, even at relatively low
+power, our analysis of metastasis further supports the pathway-centric approach
+as a fruitful way to uncover real differences in underlying biology. As a major
+caveat to this, we acknowledge that many therapies _do_ target highly-specific
+driver genes, markers, and signaling pathways ( _e.g._ TP53, EGFR or HER2), but
+understanding the broader context of the genetic background and pathway
+vulnerability of tumors containing such markers may aid in creating smarter
+combination therapies. We submit that when we discover the requirements of each
+cluster with respect to pathway disruptions and metastasis we may be able to
+target them therapeutically and prevent further adaptation.

### 2020-07-27 text/survival-methods.Rmd

@@
-We modeled baseline longevity in the U.S. with these different factors.
-<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
-Using this base survival model, we constructed a likelihood function with
-parameters describing the overall lethality of each cancer type as a modifier
-$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
-$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
-taken as, roughly, the death rate relative to the baseline rates as depicted in
-**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
-twice as deadly.
+In order to understand how the diagnosis of cancer changes the expectation for
+longevity of patients, we first began by modeling the longevity of the overall
+population of the US based on CDC life tables. Although childhood cancers exist
+and are tragic, most cancer patients have diagnosis later in life. We modeled
+baseline longevity using a risk rate function:
 
-With an accurate model for baseline longevity in place, we sought to model
-cancer as an event based modifier of long-term survival. Thus any factors or
-events such as cancer would modify the baseline risk function as a multiplier.
-The resulting coefficients can thus be interpreted as increasing baseline risk
-by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
-female vs. male caucasians, blacks and asians (hispanics are not clearly marked
-in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
-data and baseline risk models). Thus we are able to accurately model baseline
-risk of death taking these factors into account. The data themselves encompass
-all other causes of death for these populations including cancer, thus freeing
-our model from concerns about other confounding factors ( _e.g._ Diabetes). For
-demonstration purposes we also illustrate some examples of the expected survival
-for a population that receives a diagnosis with doubled risk of death at ages
-40, 50, 60, 70, and 80 (**supplementary figure S6**).
+$$R(t) = R_{20}\exp(a(A-20) + D(t) k a (\frac{t}{365}))$$
+
+Time is split into two components. $A$ going from 0 up to the age of the patient
+at diagnosis and then remaining fixed, whereas $t$ starts at 0 upon diagnosis
+and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
+is an indicator for whether diagnosis has occurred or not (before diagnosis D(t)
+= 0, after it equals 1). The model essentially assumes that the risk of dying
+per year at age 20 years is $R_{20}$ and that it increases exponentially with a
+constant rate $a$ up to diagnosis, and then with a new rate $ka$
+thereafter. Though the model does extend back to earlier ages, relatively few
+patients were under age 20.
+
+Given this risk per unit time, the probability to die at time $T+dT$ is the
+probability to survive to time $T$ which is $(1-P(T))$ and then die in the
+remaining interval which is $R(T)dT$
+
+$$dP = (1-P(T)) R(T) dT$$
+
+leading to the differential equation for the cumulative probability of death
+$P(T)$
+
+$$\frac{dP}{dT} = R(T)(1-P(T))$$
+
+The solution of this ODE is:
+
+$$P(T) = 1 - \exp(-\int_0^T R(t) dt)$$
+
+And the density of deaths per unit time is the derivative with respect to T:
+
+$$p(T) = R(T) \exp(-\int_0^T R(t)dt)$$
+
+The $R(T)$ function has two important parameters, the $a$ value which represents
+the background risk rates, and the $k$ value, which is a function of both cancer
+tissue type, and the cancer genomic class.
+
+$$k = k_{tis} k_{cl}$$
+
+The product $k_{tis}k_{cl}$ is obviously symmetric between the two factors. To
+disambiguate the meaning of the two, the prior distribution of $k_{cl}$ has peak
+probability at 1 and a relatively narrow width, due to the fact that it is a
+tissue specific multiplicative modifying factor, this allows the overall
+magnitude of k to be primarily determined by the $k_{tis}$ value whose prior is
+significantly less constrained so that there is a wider range of dangers across
+tissue types.
+
+The $k$ in our model can be taken as, roughly, the aging rate relative to the
+baseline rates as depicted in **supplementary figure S5**. Thus, a value of
+$k=2$ means that a given cancer causes you to experience the same risk in time
+$dt$ as a non-cancer patient would experience at the same age, in time $2dt$.
+
+Using this model, we split the data by Male/Female sex, and by White, vs Black
+race as these were the main well-identified categories available to us in the
+dataset. Sex is known to be a risk factor for death, with males dying at
+slightly higher rates for all ages of interest in our dataset. Black vs White
+race is an important category due to differences in access to care and other
+socioeconomic and education related factors.
+
+Using the model described above we fit the data using Stan via the rstan package
+in R. Posterior inferences are shown in **FIXME WHICH FIGURE?**.
+
+**supplementary figure S5** gives a comparison on log-scale between the post-fit
+survival curves and the CDC data it is based on. For demonstration purposes we
+also illustrate some examples of the expected survival for a population that
+receives a diagnosis with doubled risk of death at ages 40, 50, 60, 70, and 80
+(**supplementary figure S6**).
 
@@
 for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened agressively (
+confounded with cancer type because some cancers are screened aggressively (
 _e.g._ colorectal and prostate cancers) while others are diagnosed typically

### 2020-07-27 text/survival.Rmd

If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. Kaplan-Meier plots reveal large differences between
cancer types as expected, and some modest differences between our classes (not
shown). These analyses are profoundly limited by confounding factors of age and
stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
name the obvious ones. To explore these ideas, we tested models of survival
using public longevity data from the CDC, accounting for age, gender and
ethnicity. Our model assumes that at birth each person has a small initial
probability per unit time to die, and that this probability per unit time of
death conditional on age, grows exponentially in time at some baseline rate that
matches the observed CDC data. The rate of growth in risk is assumed to jump to
a higher level at diagnosis, which we take as a proxy for onset. This represents
an accelerated aging type model. To fit the model, we constrain the baseline
risks using the available CDC life tables. We then constrain the cancer specific
multiplier using the longevity data within the cancer tumor dataset. Our model
estimates the effect of cancer type and class-specific cancer effects
independently, resulting in a cancer and class specific estimate of the aging
rate multiplier. We found that cancer types, as expected, have a range of
prognoses relative to the general population, as shown in **figure 5a**. The
least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
(GBM) which has a risk multiplier of between 15 and 22 relative to the
background risk of death in the population. This undoubtedly is due to the fact
that relatively young patients are affected by GBM. Among the deadliest cancers
outside of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD)
cancers. Apart from these top 4 cancers which ranged from 6-22 in relative risk
of age-adjusted death, the remaining cancer types ranged from about 1 to 5 in
magnitude (**figure 5a**). A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

![](../figures/fig6B.png)
![](../figures/fig6C.png)
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters. a) Posterior distributions of estimated relative age-dependent survival by cancer type. b) Posterior distributions of estimated class-dependent survival for each cancer type. See methods for model details.</p>

Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
reflecting that some classes are either less deadly or more deadly than other
classes within each cancer types. A class specific rate of 1 represents typical
rate for this tissue type (**figure 5b**). For several cancers ( _e.g._ PRAD,
kidney chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer
(THCA)) the posterior estimates are largely indestinguishable from the prior,
reflecting that either there were too few mortalities in the data to make an
estimate (as expected for PRAD and THCA) or two few samples period. We did not
observe class-specific trends that held true across cancer types, which could
result from different cancers having different standards of care for example. In
support of this interpretation, we also tested a factored model which considered
both cancer type and each class independent of cancer type (**supplementary
figure S7??**). Though we were able to successfully fit this model, it is a
special case of the more general model where class and tissue can independently
affect longevity, and there is no reason to believe that class specific effects
would necessarily remain constant across tissue types given how widely the
patients vary across tissue types as well as the different drug and surgical
treatments typical used for each tissue.

```{r survival-model, eval=FALSE}

```

![**Figure 6B**](../figures/fig6B.png)

![**Figure 6C**](../figures/fig6C.png)

### 2020-07-27 code/figure1.Rmd

@@
 ```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
 load("../data/gene_data.Rda")
@@
   .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] ## include only second level pathways
-
-## Removing pathways that are not molecular functions such as pathways that related to disease, defect, bacteria, disorder, myosin and etc
-rm_pathways <- c("R-HSA-1222499", "R-HSA-1226099", "R-HSA-1300642", "R-HSA-1300645", "R-HSA-162906",  "R-HSA-1643713", "R-HSA-168254",
-                 "R-HSA-212436", "R-HSA-2160456", "R-HSA-2219528", "R-HSA-2474795", "R-HSA-2534343", "R-HSA-2644603", "R-HSA-2978092",
-                 "R-HSA-3296482", "R-HSA-3304351", "R-HSA-3560782", "R-HSA-3781860", "R-HSA-388396",  "R-HSA-3906995", "R-HSA-4791275",
-                 "R-HSA-5339562", "R-HSA-5387390", "R-HSA-5545483", "R-HSA-5576886", "R-HSA-5576890", "R-HSA-5576892", "R-HSA-5576893",
-                 "R-HSA-5576894", "R-HSA-5578768", "R-HSA-5578775", "R-HSA-5579029", "R-HSA-5602358", "R-HSA-5609975", "R-HSA-5619084",
-                 "R-HSA-5619102", "R-HSA-5632927", "R-HSA-5632928", "R-HSA-5632968", "R-HSA-5632987", "R-HSA-5663084", "R-HSA-5687613",
-                 "R-HSA-6802957", "R-HSA-8862803", "R-HSA-8876384", "R-HSA-9005891", "R-HSA-9605310", "R-HSA-9616333", "R-HSA-9616334",
-                 "R-HSA-9629232", "R-HSA-9630750", "R-HSA-9632693", "R-HSA-9645722", "R-HSA-9646303", "R-HSA-9646304")
-
-## Ensembl to reactome pathway lookup table
-gene2level2_lookup <- ensembl2reactome %>% dplyr::select(-PATH_NAME) %>% ## remove PATH_NAME col
+  .[.$PATH_ID %in% first2second$second_level,] %>% ## include only second level pathways
+  left_join(., main2first, by = c("PATH_ID" = "first_level")) %>% distinct() %>% 
+  left_join(., ensembl2reactome[,c(2,4)], by = c("main_level" = "V2")) %>% distinct()
+
+## Filtering pathways not relevant to our analysis by keywords
+filtered_ensembl2reactome <- ensembl2reactome %>% filter(!V4 %in% c("Disease", "Muscle contraction", "Reproduction")) ## Pathways in disease, muscle contraction and reproduction parent pathway were removed
+
+toMatch <- c("disease", "defect", "cancer", "oncogenic", "disorder", "infection", "mutant", "bacteria", "listeria", "loss") ## pathways with the following key terms were removed.
+
+filtered_ensembl2reactome <- filtered_ensembl2reactome %>% mutate(l_pathname = tolower(.$PATH_NAME)) %>% 
+  .[!grep(paste(toMatch,collapse="|"), .$l_pathname),]
+
+filtered_ensembl2reactome <- filtered_ensembl2reactome %>% .[which(!.$PATH_ID %in% c("R-HSA-212436", "R-HSA-388396")),] ## Removing Generic Transcription Pathway and GPCR downstream signalling as they are broad pathways
+
+gene2level2_lookup <- filtered_ensembl2reactome %>% dplyr::select(ENSEMBL, PATH_ID) %>% ## remove PATH_NAME col
   .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
@@
   column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  .[,!names(.) %in% rm_pathways] %>% ## filter rm_pathways from lookup table
   replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "gene2_level2_lookup.Rda")
+#save(gene2level2_lookup, file = "../data/gene2_level2_lookup.Rda")
 ## Pathway names to level 2 lookup table
-pathname2level2 <- ensembl2reactome %>% dplyr::select(-ENSEMBL) %>% ## remove ENSEMBL col
+pathname2level2 <- filtered_ensembl2reactome %>% dplyr::select(PATH_ID, PATH_NAME) %>% ## remove ENSEMBL col
   .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
   distinct(.) ## keep unique rows
-#save(pathname2level2, file = "pathname2level2.Rda")
-rm(rm_pathways, ensembl2reactome, main2first, first2second)
+#save(pathname2level2, file = "../data/pathname2level2.Rda")
+rm(list = ls(all.names = T))
 gc()

### 2020-07-28 text/manuscript.Rmd
add Khatri et al (Atul Butte) Pathway review reference to discussion

@@
 knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-library(doRNG)
-library(uwot)
-library(janitor)
-library(FactoMineR)
-library(seriation)
+suppressPackageStartupMessages(
+  library(rstan),
+  library(foreach),
+  library(sqldf),
+  library(tidyverse),
+  library(ggthemes),
+  library(data.table),
+  library(biomaRt),
+  library(dplyr),
+  library(GenomicRanges),
+  library(SummarizedExperiment),
+  library(reactome.db),
+  library(plyr),
+  library(cluster),
+  library(pheatmap),
+  library(clusterProfiler),
+  library(org.Hs.eg.db),
+  library(dbscan),
+  library(class),
+  library(TCGAbiolinks),
+  library(patchwork),
+  library(ggplotify),
+  library(ComplexHeatmap),
+  library(knitr),
+  library(kableExtra),
+  library(UpSetR),
+  library(cowplot),
+  library(doRNG),
+  library(uwot),
+  library(janitor),
+  library(FactoMineR),
+  library(seriation),
+)
 ```

### 2020-07-30 code/figure2.Rmd
update text and figures

@@
 
-```{r figure-2-plot, out.height='90%', out.width='90%',fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions.
-Waterfall plot; columns represent tumors.
-Each row represents a pathway.
-The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively.
-All 377 pathways are shown.
-Red signifies pathway disrupted pathways, grey signifies not disrupted."}
+```{r figure-2-plot, out.height='90%', out.width='90%'}
 
@@
 ```
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 2: Pan-cancer enrichment of pathway disruptions.** a) Waterfall plot; columns represent tumors.
+Each row represents a pathway.
+The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively.
+The top 100 most mutated pathways are shown.
+Red signifies pathway disrupted pathways, grey signifies not disrupted.
+b) Heatmap shows the relative enrichment of each pathway (columns) within each numbered cluster (rows).
+Effect size is displayed as colors (orange, enrichment; purple, depletion).
+c) Proportion of mutated samples in the each significant pathway (columns; union set of pathways with effect size >= 0.30 in each cluster) within each numbered cluster (rows).</p>

### 2020-07-30 text/manuscript.Rmd
update text and figures

--- 2020-07-27 text/manuscript.Rmd
@@
 library(seriation)
+library(rtracklayer)
+library(poibin)
 ```
@@
 sequencing data because the affected target gene is known unambiguously.
-Therefore, we focused on all missense, nonsense, frameshift, and splicing
-mutations. In order to minimize bias from well-studied diseases and processes,
-we selected 377 Reactome pathways [reactome.org](https://reactome.org/) of
-interest corresponding to basic cellular processes and biochemical pathways,
+Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
+region, and splicing mutations. In order to minimize bias from well-studied diseases
+and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
+of interest corresponding to basic cellular processes and biochemical pathways,
 excluding gene sets that correspond to miscellaneous categories ( _e.g._
 “transcription factors”) or disease associations ( _e.g._ “mutated in colon
-cancer”) and filtered our gene list on membership in these pathways (<mark>total
-of x genes;</mark> see methods for details).
+cancer”) and filtered our gene list on membership in these pathways (total
+of 9040 genes; see methods for details).
 
@@
 mutations can be deleterious (attenuating or hypomorphic) or activating
-(neomorphic) in genes that act functionally as either activators or repressors.
+(neomorphic or hypermorphic) in genes that act functionally as either activators or repressors.
 Within the scope of this study we accept this uncertainty and assume mutations
@@
 eliminated them for that cancer type only (see methods for details). 
-<strike>Highly expressed genes could also have high mutation rates owing to
+<strike> Highly expressed genes could also have high mutation rates owing to
 transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
@@
 carries a mutation) to the pathway regardless of the number of genes carrying
-mutations (**figure 1A**).
+mutations (**figure 1A**). </strike>
 
@@
 In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
-containing useful information (**figures 1B, C**). Then we performed a UMAP
+analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
+containing useful information. (**figures 1B, C**). Then we performed a UMAP
 analysis, both in order to summarize the MCA graphically, and as a preprocessing
@@
 created a score metric as the fraction of classified tumors with max probability
-$< 5\%$ in one class and chose a cluster size of 95 to minimize the score
-function (**figure 1E**). HDBSCAN with these settings resulted in 12 distinct
+$< 5\%$ in one class and chose a cluster size of 92 to minimize the score
+function (**figure 1E**). HDBSCAN with these settings resulted in ten distinct
 high-density clusters which we then projected onto the UMAP (**figure 1F**).
-This classified <mark>x out of 7,607 tumors</mark> but still resulted in a
+This classified 6,038 out of 7,607 tumors but still resulted in a
 significant fraction of unclassified tumor samples. Since we ultimately wish to
@@
 example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 8 & 9, which is
-distinguished  by KRAS mutations among others (**supplementary figure S3**). But even PAAD comprises tumors
-from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
+pancreatic adenocarcinoma (PAAD) was biased in class 8, which is
+distinguished  by KRAS mutations among others (**supplementary figure S3**). 
+But even PAAD comprises tumors from the nine remaining classes, meaning that 
 patients hosting these tumors have potentially different underlying molecular
@@
 cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
+normal-like, and luminal A and B subtypes onto the UMAP. These are among the
 most heavily studied molecular subtypes in cancer, which each have different
@@
 segregation of these mutations into specific clusters (**figure 1K**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
+similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_,
+_MLH3_, _PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
 except for _APC_ exhibited any remarkable specificity with respect to cluster
 assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
+maintenance of DNA methylation levels and telomere length, we projected somatic
+mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
 observed similar lack of segregation by cluster (**supplementary figure S3**).

### 2020-07-30 text/discussion.Rmd

@@
 @kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
-@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
-simplify the problem by investigating cancer as a disease of basic cellular and
+@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
+simplify the problem by employing a "knowledge-base driven analysis"
+[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
 biochemical pathways, which we accomplished by translating gene-level mutations
-into pathway level disruptions. Our approach differs from previously proposed
+into pathway level disruptions. Our approach differs from previously described
 methods [@PATHNET] in that we chose to focus on the pathway as the unit of
 disruption instead of the gene, employing the conceit that individual mutations
-may be sufficient to disrupt entire pathways (as is often the case with
-drivers). Rather than evaluating the burden of mutations on each pathway within
-or across tumors, which can be biased by the number of genes or the coverage of
-pathway-linked genes in the genome, and which we find to be relatively uniformly
-distributed, we scored each pathway in binary fashion, allowing one mutation or
-multiple mutations to count equally for each sample.
+may be sufficient to disrupt or misdirect pathway activity (as is often the case
+with driver mutations). Rather than evaluating the burden of mutations on each
+pathway within or across tumors, which can be biased by the number of genes or
+the coverage of pathway-linked genes in the genome, we scored each pathway in
+binary fashion, allowing one mutation or multiple mutations to count equally for
+each sample.
 
@@
 pathways, excluding curated genesets related to diseases, syndromes, or classes
-of proteins ( _e.g._ transcription factors). We chose this approach to limit
-redundancy, which is already high in the remaining pathways for some driver
-genes, and to exclude biologically unrelated collections of genes. We do not
-account for non-coding mutations, amplification events or structural variants,
-though we hope to include them in subsequent studies (see below). Our
-pathway-centric analysis also allows us to circumvent the problem that some
-cancer-promoting mutations likely occur at low enough frequencies as to be
-indistinguishable from background. Previous attempts to capture these
-associations better using Bayesian networks have yielded important new insights
-into this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that
-would not have been possible with traditional analysis, supporting the case that
-pathway-centric approaches like ours have the potential to increase statistical
-power.
+of proteins ( _e.g._ transcription factors) which are potentially problematic
+[@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
+already high in the remaining pathways for some driver genes, and to exclude
+biologically unrelated collections of genes. We do not account for non-coding
+mutations, amplification events or structural variants, though we hope to
+include them in subsequent studies (see below). Our pathway-centric analysis
+also allows us to circumvent the problem that some cancer-promoting mutations
+likely occur at low enough frequencies as to be indistinguishable from
+background. Previous attempts to capture these associations better using
+Bayesian networks have yielded important new insights into this problem and new
+genes [@hornNetSigNetworkbasedDiscovery2018] that would not have been possible
+with traditional analysis, supporting the case that pathway-centric approaches
+like ours have the potential to increase statistical power.
 

### 2020-07-30 text/introduction.Rmd
add temporary header to discussion and intro to compile references, fix typos, update discussion edits

@@
+---
+title: "Introduction"
+output: 
+  bookdown::html_document2:
+    fig_captions: TRUE
+    number_sections: FALSE
+    toc: FALSE
+    theme: flatly
+bibliography: references.bib
+---
+
 Recent advances in technology have made the study of cancer possible at
@@
 sparseness of mutations, allowing the potential importance of a gene to be
-influenced by the aggregate of positive selection in its nearest network
+influenced by the aggregate of mutations in its nearest network
 neighbors. This method is powerful especially when applied to a small number (on
@@
 gene-level information to non-tissue specific cell-biological pathway
-phenotypes. We use this rubric to evaluate all cancers and show that,
+phenotypes. We use this statistic to evaluate all cancers and show that,
 surprisingly, tumors may have distinct phenotypes completely independent of

### 2020-07-30 code/supplements.Rmd

--- 2020-07-23 code/supplements.Rmd
@@
+```{r}
+library(data.table)
+library(ggplot2)
+library(tidyverse)
+library(kableExtra)
+```
+
 ### Supplementary Figure S1
@@
 
-##oncogene
-oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-##tumor_supressor
-tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor
+## MMR genes
+genes <- c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC",
+           "TET2", "TET3", "TERT", "TEP1", "DKC1")
 
-selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
+selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% genes) ## getting ensembl_id for the mmr genes for filtering gdc data
+
+# ##oncogene
+# oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
+# ##tumor_supressor
+# tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL", "APOBEC3B") ## list of selected tumor suppressor
+# 
+# selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
 
@@
     legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6", 
-                                  "APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL")),
-                      ncol = 5) ## facet by gene
+  facet_wrap(~ factor(variable, levels = c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")), ncol = 5) ## facet by gene
 ```

### 2020-07-30 text/manuscript.Rmd

@@
 knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-library(doRNG)
-library(uwot)
-library(janitor)
-library(FactoMineR)
-library(seriation)
-library(rtracklayer)
+suppressPackageStartupMessages(
+  library(rstan),
+  library(foreach),
+  library(sqldf),
+  library(tidyverse),
+  library(ggthemes),
+  library(data.table),
+  library(biomaRt),
+  library(dplyr),
+  library(GenomicRanges),
+  library(SummarizedExperiment),
+  library(reactome.db),
+  library(plyr),
+  library(cluster),
+  library(pheatmap),
+  library(clusterProfiler),
+  library(org.Hs.eg.db),
+  library(dbscan),
+  library(class),
+  library(TCGAbiolinks),
+  library(patchwork),
+  library(ggplotify),
+  library(ComplexHeatmap),
+  library(knitr),
+  library(kableExtra),
+  library(UpSetR),
+  library(cowplot),
+  library(doRNG),
+  library(uwot),
+  library(janitor),
+  library(FactoMineR),
+  library(seriation),
+  library(rtracklayer),
   library(poibin)
+)
 ```

### 2020-07-31 code/figure1.Rmd

@@
 
-begin_NA <- data.frame(sample_id = paste0("sample", 1:50), 
+begin_NA <- data.frame(sample_id = paste0("sample", 1:100), 
                        clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
                        log10mut = NA) ## put NA values so they don't get plotted
-end_NA <- data.frame(sample_id = paste0("sample", 51:100), ## repeat the step for the ends
+end_NA <- data.frame(sample_id = paste0("sample", 101:200), ## repeat the step for the ends
                        clust_knn = max(as.numeric(mut_order$clust_knn)),
@@
 mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abrreviations from the TCGA project. b) MCA analysis was performed
-restricted to the 36 informative components. c) The resulting MCA was projected
-down to 3 dimensions for the purpose of density based clustering (only the first
-2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
-perform hierarchical density based clustering tuned to minimize the fraction of
-unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
+standardized abbreviations from the TCGA project. b) MCA analysis was performed to 
+determine the eigenvalue with the most explanatory power from the categorical data.
+Average of 100 permutations of the data was used as a baseline. c) The p-value of
+each eigenvalue was plotted and the max eigenvalue with a pvalue < 0.05 was selected. 
+d) The resulting MCA was projected down to 3 dimensions for the purpose of density 
+based clustering (only the first 2 dimensions are shown). Each dot represents a tumor 
+e) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize
+the fraction of unassigned tumors. f) HDBSCAN clustered tumors were visualized on the UMAP
 projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). f) kNN was used to assign
-membership to the remaining tumors. g) Violin plots of sample mutation counts
-for tumors in each class. h) Projection of cancer type identity onto UMAP.
-Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
-Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
-cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
-UMAP. k) projection of tumor stage onto the UMAP,
-regardless of cancer type. l) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>
+in some tumors remaining unclassified (grey points). g) kNN was used to assign
+membership to the remaining tumors. h) Mutational burden plot where each dot corresponds
+to the sample mutation counts for tumors in each class. i) Projection of cancer 
+type identity onto UMAP. Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma,
+HNSC = Head and neck squamous cell, PAAD = Pancreatic adenocarcinoma. j) Projection of breast
+cancer subtypes onto the UMAP. k) Projection of BRCA1/2 somatic mutation onto the
+UMAP. l) Projection of tumor stage onto the UMAP, regardless of cancer type. m) 
+Projection of metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, 
+M1 = metastatic tumors.</p>

### 2020-07-31 text/discussion.Rmd

@@
 
+### Tissue specific manifestation of pathway-centric disruptions
+
+Our results illustrate how unique combinations of mutations in pan-cancer driver genes with tissue-specific pathway disruptions result in common categories when viewed at the level of the pathway knowledge-base.
+For example, in table 1 we find that `snRNP assembly`, one of the unique driver pathways of cluster 5, the top gene in breast cancer is represented by TPR (mutated in 26% of samples).
+By contrast the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of samples, respectively).
+For some other clusters, _e.g._ 7, 8 & 9, the top genes are known drivers such as TP53, PIK3CA and KRAS.
+None of these genes is 100% mutated in any cluster however, and therefore it is the combinations with secondary mutations, many of which are tissue-specific, that round out the complement of pathways in all clusters.
+These top genes are found in most of the clusters (supplemental figure S5), and in spite of the fact that they contribute to many cluster-specific pathways.
+This can only be explained as a result of other less common driver genes complementing the pathway disruption of the unique combination of driver genes that are disrupted in each tumor, and we speculate that many of these are relatively tissue-specific in terms of their sensitivity to mutation.
+Consistent with our observations, @moonlight showed how tissue- and cancer-specific driver genes and druggable targets of pathway moduli can be discovered by machine learning on integrated multi-omics datasets.
+Marrying these approaches in future work will greatly enhance our understanding of tumor biology.
+
 ### Incompleteness of the pathway disruption data
@@
 frequencies as in @PCAWG-EVO. However, since many of the genes in the
-non-cluster-specific pathways involve genome stability, it is reasonable to
-hypothesize that these mutations promote or enable the acquisition of
+non-cluster-specific pathways involve the known driver genes, it is reasonable to
+hypothesize that these mutations promote or enable the subsequent acquisition of
 class-specific defects via random mutation and natural selection, thus producing
 the classes we observed. In support of this, the pan-cancer analysis of whole
-genomes consortium (PCAWG) found that a subset of known oncogenic driver
+genomes consortium (PCAWG) found that oncogenic driver
 mutations are highly enriched in early arising clones, whereas later arising

### 2020-08-03 text/discussion.Rmd
double check results, updated text in results and discussion section. added ubiq-proteasome gene annotation on right side. added figure s5

--- 2020-07-30 text/discussion.Rmd
@@
 mutations with likely deleterious effects, including missense, nonsense,
-frame-shift, stop-gain and splice site mutations in genes that are actively
+frame-shift, stop-gain, untranslated region and splice site mutations in genes that are actively
 expressed in each cancer type, thus avoiding bias from transcription coupled
@@
 
-We find that most tumors segregate into one of twelve biologically distinct
+We find that most tumors segregate into one of ten biologically distinct
 clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
 figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
-mutant clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
+mutant clusters 8, and Kidney Chromophobe in cluster 1, see **supplementary
 figure S1**) are unevenly distributed, though we could not identify any cancers
@@
 analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
-proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
-skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
+proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and are heavily
+skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 10
 represents a group of hyper-mutated tumors that harbor so many mutations that

### 2020-08-03 text/manuscript.Rmd
double check results, updated text in results and discussion section. added ubiq-proteasome gene annotation on right side. added figure s5

--- 2020-07-30 text/manuscript.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(rstan)
-library(foreach)
-library(sqldf)
-library(tidyverse)
-library(ggthemes)
-library(data.table)
-library(biomaRt)
-library(dplyr)
-library(GenomicRanges)
-library(SummarizedExperiment)
-library(reactome.db)
-library(plyr)
-library(cluster)
-library(pheatmap)
-library(clusterProfiler)
-library(org.Hs.eg.db)
-library(dbscan)
-library(class)
-library(TCGAbiolinks)
-library(patchwork)
-library(ggplotify)
-library(ComplexHeatmap)
-library(knitr)
-library(kableExtra)
-library(UpSetR)
-library(cowplot)
-library(doRNG)
-library(uwot)
-library(janitor)
-library(FactoMineR)
-library(seriation)
-library(rtracklayer)
+suppressPackageStartupMessages(c(
+  library(rstan),
+  library(foreach),
+  library(sqldf),
+  library(tidyverse),
+  library(ggthemes),
+  library(data.table),
+  library(biomaRt),
+  library(dplyr),
+  library(GenomicRanges),
+  library(SummarizedExperiment),
+  library(reactome.db),
+  library(plyr),
+  library(cluster),
+  library(pheatmap),
+  library(clusterProfiler),
+  library(org.Hs.eg.db),
+  library(dbscan),
+  library(class),
+  library(TCGAbiolinks),
+  library(patchwork),
+  library(ggplotify),
+  library(ComplexHeatmap),
+  library(knitr),
+  library(kableExtra),
+  library(UpSetR),
+  library(cowplot),
+  library(doRNG),
+  library(uwot),
+  library(janitor),
+  library(FactoMineR),
+  library(seriation),
+  library(rtracklayer),
   library(poibin)
+))
 ```
@@
 cancer”) and filtered our gene list on membership in these pathways (total
-of 9040 genes; see methods for details).
+of 8940 genes; see methods for details).
 
@@
 homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S4**). Furthermore, cluster 1 is distinguished
+**supplementary figure S5**). Furthermore, cluster 1 is distinguished
 from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S4**).
+_TP53_ panel of **Supplementary Figure S5**).
 
@@
 Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all 12 clusters
-( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
+pathways (**figure 2**) we applied this method to metastatic vs non-metastatic
+tumors, as we find that metastatic tumors are distributed across all ten clusters
+( _e.g._ **figure 1m**). Using all non-metastatic tumors as background, we found
 very low levels of enrichment, less than $10\%$, in a handful of pathways. We
@@
 Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found 99 enriched pathways across all clusters (**table 2**). A number of
+found a total of 27 enriched pathways across all clusters (**table 2**). A number of
 enrichments were found in multiple clusters represented by pathways that were
 already shown to be enriched in a neighboring cluster. For example,
-“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
-of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
-10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
-union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
-activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
-Cluster 1, one of two clusters for which we could not identify any
-distinguishing pathways (cluster 8 being the other), had enrichment in two
-pathways that were otherwise specific to non-metastatic tumors of other
+“Signaling by PTK6” is enriched in non-metastatic samples of cluster 8 (see 
+**supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
+metastatic tumors of clusters 7 and 9 (p < 10-3, **table 2**). This is also true of 
+“Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
+S4**) and also in metastatic tumors of clusters 7 and 9. <strike>Cluster 1, one of two clusters 
+for which we could not identify any distinguishing pathways (cluster 8 being the other), 
+had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other
 clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
-samples were enriched for two key neutrophil pathways, “Neutrophil
-Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
-cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
+7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).</strike> Cluster 4 metastatic
+samples was enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathways, 
+which are specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
+specific to metastases tend toward complementary mutations that are otherwise specific to
 non-metastatic tumors of other clusters.
@@
 ```
-
-There were also a number of pathways that were enriched with lower specificity
+<mark>Statement does not apply anymore. The highest # of enriched path is 3. Not many are in 
+metastasis. Only 2. “Fc epsilon receptor (FCERI) signaling” and "Signaling by SCF-KIT" </mark>
+<strike>There were also a number of pathways that were enriched with lower specificity
 (among 4 or more clusters) which were found to be enriched in metastases. We
@@
 Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).
+Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**). </strike>
 

### 2020-08-03 code/figure4_function.Rmd

@@
         strip.background = element_blank(),
+        plot.margin = unit(c(0,0,0,0), "cm"),
         legend.position = "none") +
@@
 
-  plotlist <- c(l_anno_ggplot, list(heatmap_ggplot)) ## make a list of plots for cowplot
+  proteasome_r_anno <- data.frame(gene = gene_order, 
+                                psm_gene = 0)
 
-  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 5))
+  proteasome_r_anno[grep(pattern = "PSM|SEM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
+
+  r_anno_color <- c("0" = "white", "1" = "black")
+  proteasome_r_anno$gene <- factor(proteasome_r_anno$gene, levels = c(rev(gene_order)))
+  r_anno <- ggplot(proteasome_r_anno) +
+    geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
+             stat = "identity", 
+            width = 1) +
+    scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
+    theme_void() 
+  legend <- plot_grid(get_legend(r_anno), ncol = 1)
+  r_anno <- r_anno + theme(legend.position = "none")
+
+  plotlist <- c(l_anno_ggplot, list(heatmap_ggplot), list(r_anno)) ## make a list of plots for cowplot
+
+  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 5, 0.1))
   plot_legend <- plot_grid(plotlist = path_legend, ncol = 1, align = "v")
@@
 
-```{r figure-4-plot, out.width= '90%', fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits."}
+```{r figure-4-plot, out.width= '90%'}
 load("../data/class_spec_fig4.Rda")
@@
 ```
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3: Mutations in proteasomes found in Class 2 affect multiple enriched pathways.** a) Each column is a tumor in cluster 2, each row represents a gene with mutation (red) or not mutated (grey).
+The genes are organized according to which of the top distinguishing pathways for this class that they effect.
+Some genes affect multiple pathways, as indicated by the bars at left.
+The highlighted genes at right are all proteasomal subunits.</p>

### 2020-08-03 code/supplements.Rmd

@@
 
+```{r, supplemental-s5-data-generation, eval = FALSE}
+load("../data/gdc_tumor_coding_complete.Rdata")
+load("../data/gene_data.Rda")
+
+## Booleanize mutation count for each tumor sample
+missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
+splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
+frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
+frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
+utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
+utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
+nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
+nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
+
+gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
+rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
+gc()
+gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
+
+## class 7 and 8 distinguishing genes
+genes <- c("PIK3CA", "KRAS", "TP53")
+
+selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% genes) ## getting ensembl_id for the genes for filtering gdc data
+
+selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
+  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., selected_mut, by = "ensembl_gene_id") %>% ## replace ensembl ids with gene name 
+  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>%
+  t() %>% as.data.frame() %>% rownames_to_column(var = "sample_id")  ## prep for left_join
+
+s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
+  left_join(., selected_mut, by = "sample_id") %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
+  filter(value != 0)
+
+save(s5_data, file = "../data/s5_data.Rda")
+```
+
+```{r supplemental-s5-plot}
+load("../data/s5_data.Rda")
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
+umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+
+clust_178 <- umap_cancertype %>% filter(clust_knn %in% c(1, 7, 8))
+
+umap_clust178 <- ggplot(clust_178, aes(x = plot_y, y = plot_x, colour = as.character(clust_knn))) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
+  theme_classic() +
+  coord_fixed() +
+  xlab("UMAP 1") + ylab("UMAP 2") + labs(colour = "Cluster") +
+  guides(colour = guide_legend(override.aes = list(size=3))) +
+  scale_colour_manual(values = c("gold2", "gray51", "springgreen4")) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    axis.ticks = element_blank(),
+    legend.position = c(0.95, 0.8))
+
+load("../data/s5_data.Rda")
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
+class7_8 <- umap_cancertype %>% filter(clust_knn %in% c(7, 8))
+umap_bg_7_8 <- class7_8[,-"project_code"] ## get background data for facet
+
+# pik3ca_kras <- ggplot(s5_data %>% filter(sample_id %in% umap_bg_7_8$sample_id) %>% filter(variable != "TP53") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+#   geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+#   geom_point(size = 0.25) +
+#   theme_classic() +
+#   coord_fixed() +
+#   #ylim(-0.50, 1.5) +
+#   xlab("UMAP 1") + ylab("UMAP 2") +
+#   scale_colour_manual(values = rep("firebrick3", 20)) +
+#   theme(axis.text.x = element_blank(),
+#     axis.text.y = element_blank(),
+#     axis.title.x = element_blank(),
+#     axis.title.y = element_blank(),
+#     axis.ticks = element_blank(),
+#     axis.line = element_blank(),
+#     legend.position = "none") +
+#   facet_wrap(~ variable, ncol = 2)
+
+tp53 <- ggplot(s5_data %>% filter(variable == "TP53") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
+  theme_classic() +
+  coord_fixed() +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  scale_colour_manual(values = rep("firebrick3", 20)) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ variable, ncol = 1)
+
+pik3ca <- ggplot(s5_data %>%  filter(variable == "PIK3CA") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
+  theme_classic() +
+  coord_fixed() +
+  #ylim(-0.50, 1.5) +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  scale_colour_manual(values = rep("firebrick3", 20)) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ variable, ncol = 1)
+
+kras <- ggplot(s5_data %>%  filter(variable == "KRAS") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+  geom_point(size = 0.25) +
+  theme_classic() +
+  coord_fixed() +
+  #ylim(-0.50, 1.5) +
+  xlab("UMAP 1") + ylab("UMAP 2") +
+  scale_colour_manual(values = rep("firebrick3", 20)) +
+  theme(axis.text.x = element_blank(),
+    axis.text.y = element_blank(),
+    axis.title.x = element_blank(),
+    axis.title.y = element_blank(),
+    axis.ticks = element_blank(),
+    axis.line = element_blank(),
+    legend.position = "none") +
+  facet_wrap(~ variable, ncol = 1)
+
+layout <- '
+AABB
+AACC
+AADD
+'
+
+umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
+```
+
 ### Supplementary Table ST1

### 2020-08-03 code/table1.Rmd

@@
 ```
-
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1: Most frequently mutated gene in class-specific enriched pathways (effect size >= 0.30) for specific cancer**</p>
 ```{r table-1}
@@
 
-cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
+cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(format = "html", row.names = FALSE) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
@@
 ```
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>

### 2020-08-03 text/discussion.Rmd

@@
 figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
-mutant clusters 8, and Kidney Chromophobe in cluster 1, see **supplementary
+mutant cluster 8, and Kidney Chromophobe in cluster 1, see **supplementary
 figure S1**) are unevenly distributed, though we could not identify any cancers

### 2020-08-05 code/survival/kmcurve.R
Add all the survival code, including kmplots for simulated data, and modify some of the methods section on survival

@@ -1,0 +1,121 @@
+library(survival)
+library(ggfortify)
+library(gridExtra)
+library(readr)
+
+
+install.packages("devtools")
+devtools::install_github("sachsmc/ggkm")
+library(ggkm)
+
+## read in estimates from the stan runs, and then generate pseudo K-M curves
+
+
+## ethnicity based model output for mean r20 values, as printed in the
+## Rout file:
+
+## these are scaled in the stan file, so we need to scale here as well
+r20scale =1e-5
+ageratescale = .05
+    
+
+r20wf = 0.365209 * r20scale
+r20wm = 0.367732 * r20scale
+r20bf = 0.093299 * r20scale
+r20bm = 0.098039 * r20scale
+
+ageratewf = 1.197896 * ageratescale
+ageratewm = 1.291932 * ageratescale
+ageratebf = 1.649803 * ageratescale
+ageratebm = 1.734934 * ageratescale
+
+
+lccdf = function(r20,agerate,age,k,t){
+    return(-r20*((365*exp((agerate*k*t)/365+age*agerate-20*agerate))/(agerate*k)
+        -(365*exp(age*agerate-20*agerate))/(agerate*k)));
+}
+
+curve(lccdf(r20wf,ageratewf,30,1,x)-log(runif(1)),0,365*100)
+
+simdeath = function(r20,agerate,age,k){
+    r = runif(1)
+    lr = log(r)
+    tdays = uniroot(function(t) return(lccdf(r20,agerate,age,k,t)-lr),c(0,365*400))
+    return(tdays$root) ## years since diagnosis at "age"
+}
+
+## create example data for a fixed 30 year old without vs with a k=2
+## cancer
+
+
+patientsurvivalsbkd = replicate(100,simdeath(r20wf,ageratewf,30,1))/365
+patientsurvivalsk2 = replicate(100,simdeath(r20wf,ageratewf,30,2))/365
+
+
+fit1 <- survfit(Surv(patientsurvivalsbkd) ~ 1) 
+fit2 <- survfit(Surv(patientsurvivalsk2) ~ 1) 
+p1 = autoplot(fit1) + coord_cartesian(xlim=c(0,80)) + labs(x="Survival Time (years past 30th birthday)",y="Probability",title="Survival For 30 year old White Females (baseline)")
[…]

### 2020-08-05 code/survival/stananalyze.R

@@ -1,0 +1,408 @@
+library(sqldf)
+library(readr)
+library(tidyverse)
+library(rstan)
+library(foreach)
+library(bayesplot)
+library(gridExtra)
+library(openxlsx)
+library(ggridges)
+
+cancerdat = read_csv("../../data/survival/clinical_plus_cluster.csv");
+treatdays = read_csv("../../data/survival/tcga_clinical_data.csv")
+knnclust = read_tsv("../../data/survival/peter_clusters_knn.tsv")
+
+## mortality data from CDC:
+namesmort = c("agerng","pdead","nsurv","ndie","pyrs","pyrsabove","expectedlife")
+menmort = read_csv("../../data/survival/lifetables/Table02.csv",skip=2)
+## drop the final row, it's just a text info statement
+menmort = menmort[1:NROW(menmort)-1,] 
+wommort = read_csv("../../data/survival/lifetables/Table03.csv",skip=2)
+wommort = wommort[1:NROW(wommort)-1,]
+
+names(menmort) = namesmort
+names(wommort) = namesmort
+yrs=seq(0,100,by=1);
+menmort$yrnum = yrs
+wommort$yrnum = yrs
+
+ggplot(menmort) + geom_line(aes(yrnum,nsurv/100000),col="blue") +geom_line(aes(yrnum,nsurv/100000),col="red",data=wommort)
+
+
+sqldf("select count(*) from cancerdat where age_at_initial_pathologic_diagnosis < 20")
+sqldf("select count(*) from cancerdat where age_at_initial_pathologic_diagnosis < 10")
+
+sqldf("select type,count(*) from cancerdat group by type");
+sqldf("select type,avg(age_at_initial_pathologic_diagnosis) as meanage from cancerdat group by type");
+#sqldf("select type,count( as meanage from cancerdat group by type");
+
+
+
+## Create a dataset merging various components
+
+
+
[…]

### 2020-08-05 text/manuscript.Rmd

@@
 
-Pathway analysis provides additional information that genes and gene ontologies
-alone lack: 1) it aggregates molecular events across multiple genes in the same
-pathway and 2) the results, interpretable as genetic alterations, are related 
-to molecular functions such as “intrinsic pathway for apoptosis” or “cellular 
-response to hypoxia”. Using the Reactome Pathway Database, we selected the 
-second sub-pathways for each pathway (e.g. ‘Beta-catenin independent WNT 
-signaling’ in ‘Signal Transduction’) which we believed was a suitable 
-representation of the basic cellular processes and biochemical pathways. 
-We also excluded pathways categorized under cancer, disease, gene defect, 
-disorder, infection, reproduction, and muscle contraction (e.g. ‘PI3K/AKT 
-Signaling in Cancer’). While some of the excluded pathways have been shown
-to play an important role in the cancer, they focus on a specific point 
-mutations. In addition, for most of the excluded pathways, there is a global
-pathway (e.g. ‘PIP3 activates Akt signaling’) that exist. Finally, we mapped
-the 18,577 Ensembl IDs from the GDC dataset to the highly selected pathways.
-This operation produced a lookup table that consisted of 377 pathways mapped 
-to 9,910 genes.
+To understand the molecular mechanism of cancer at a pathway level, our pathway
+analysis approach focused on Reactome (https://reactome.org/), a pathway knowledge-based 
+database. The mapping files of ENSEMBL to pathways, pathway hierarchy relationship,
+and complexes to top pathways were downloaded from the download page 
+(https://reactome.org/download-data). Using the data, we imposed pathway criteria to
+meet the basic cellular processes and biochemical pathways: (1) human-derived pathways
+(_“HSA”_) (2) limited to grandchild node for each parent pathway (_e.g._ ‘Beta-catenin 
+independent WNT signaling’ in ‘Signal Transduction’) (3) exclusion of pathways in the
+parent pathway: Disease, Muscle contraction, and Reproduction or pathway names that 
+include any of the following keywords: disease, defect, cancer, oncogenic, disorder, 
+mutant, loss, infection, bacteria, or listeria. While some of the excluded pathways 
+have been shown to play an important role in cancer, they are highly specialized 
+(_e.g._ ‘PI3K/AKT Signaling in Cancer’). Additionally, for most of the excluded pathways,
+a neutral version pathway of the pathway exist (_e.g._ ‘PIP3 activates Akt signaling’). 
+Finally, we mapped the 18,577 Ensembl IDs from the TCGA dataset to the highly selected
+Reactome pathways. This operation produced a lookup table that consisted of 377 pathways
+mapped to 8,940 genes.
 

### 2020-08-05 code/survival/kmcurve.R
added a set.seed at the top of the kmcurve.R file for reproducibility

@@ -8,6 +8,9 @@ install.packages("devtools")
 devtools::install_github("sachsmc/ggkm")
 library(ggkm)
 
+
+set.seed(20200805) #a date based seed for reproducibility
+
 ## read in estimates from the stan runs, and then generate pseudo K-M curves
 
 

### 2020-08-06 text/manuscript.Rmd
Add filtering gene section to methods

@@
 
-### <mark>Filtering genes <- Peter to write</mark>
+### Filtering genes
+
+One of the limitations of knowledge-based pathway analysis is that it heavily relies on 
+experimental results for the curation and validation of its pathways. The resulting 
+pathways are therefore not specific to a particular tissue type but provides information
+for all tissue type. Before utilizing the gene to pathway lookup table, we considered the
+impact of deleterious passenger mutations for a given cancer (non-specific genes that are
+not expressed) as they may introduce biases in our pathway analysis. We addressed this by
+identifying genes that are lowly/not expressed in a tissue type. Using the TCGA RNA 
+sequencing data adjusted for batch effect dataset (https://pancanatlas.xenahubs.net), we 
+filtered the genes and tumor samples that were not included in our analysis, grouped the 
+tumor samples by tissue type and computed the mean expression value for each gene. A minimum
+threshold of 10 transcripts per million was set for expressed genes based on an inflection
+point observed when plotting the mean expression values of genes ranked by increasing 
+expression values for each tissue type. Genes that did not meet this threshold were 
+considered not expressed. This operation produced a lookup table for gene expression
+status in each tumor sample for 18,127 genes.
 

### 2020-08-07 text/manuscript.Rmd
add ms word style reference and update yaml header

--- 2020-08-05 text/manuscript.Rmd
@@
 title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
+author: Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and
+  Dennis J. Hazelett$^{1,3}$
+bibliography: references.bib
 output:
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
+  word_document:
+    reference_docx: word-style-reference-ttmanu.docx
+always_allow_html: TRUE
 editor_options:
@@
     keep_tex: yes
+
+output:
+  word_document:
+    reference_docx: word-styles-reference-01.docx
+  number_sections: FALSE
+  toc: FALSE
+
+output: 
+bookdown::html_document2:
+  fig_captions: TRUE
+  number_sections: FALSE
+  toc: TRUE
+  toc_float: TRUE
+  toc_depth: 3
+  code_folding: hide
+  theme: flatly
+  smart: true
+
 ```

### 2020-08-08 code/figure2.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

@@
-<div style="text-align: right"> Figure 2 Data </div>
-```{r}
-library(ggplot2)
-library(patchwork)
-library(tidyverse)
-library(cowplot)
-```
-
 ```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
@@
 
-```{r figure-2-plot, out.height='90%', out.width='90%'}
-
+```{r figure-2-plot, fig.align= 'center', fig.height= 7, fig.width = 7}
+#out.height='90%', out.width='90%'
 ## top100 pathways pancancer oncoprint
@@
     legend.position = "none"
-  ) #+ 
-  ggtitle("Cluster-specific enrichment score")
+  )
 

### 2020-08-08 code/figure4.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

--- 2020-07-16 code/figure4.Rmd
@@
-```{r figure-4-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/moderate_enrich_prob.Rda")
-load("../data/gene_data.Rda")
+```{r forest-heatmap-data-generation, eval = FALSE}
+# rm(list = ls())
+# gc()
+# setwd("/data/GDC_TCGA/TTmanu/wd/")
 
-## Parameters ##
-filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
-n_path <- 10 ## looking at the top10 path
-enrich_cutoff <- 0.30
-################
+library(maftools)
+library(TCGAbiolinks)
+library(tidyverse)
+library(survival)
 
-## Data Generation ##
-class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific
+kelly.colours <- c(
+  "gray95", "gray13", "gold2", "plum4", "darkorange1",
+  "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+  "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+  "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19"
+)
+cluster.colors <- kelly.colours[-c(1)]
+names(cluster.colors) <- 0:20
+cluster_colors <- cluster.colors[2:11]
+names(cluster_colors) <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
 
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
+### zcat simple_somatic_mutation.open.tsv.gz | grep -Fw -f mutation_classes.txt > simple_somatic_mutation.open.selmuts.tsv
+### cat <(zcat simple_somatic_mutation.open.tsv.gz | head -1) simple_somatic_mutation.open.selmuts.tsv | gzip -c - >simple_somatic_mutation.open.selmuts.tsv.gz
+all_muts <- icgcSimpleMutationToMAF("../data/simple_somatic_mutation.open.selmuts.tsv.gz", removeDuplicatedVariants = FALSE, addHugoSymbol = TRUE)
+all_muts <- distinct(.data = all_muts,
+                     Chromosome, Start_Position, End_Position, 
+                     Reference_Allele, Tumor_Seq_Allele2, consequence_type, 
+                     mutation_type, ens_id, icgc_donor_id, 
+                     submitted_sample_id, .keep_all = TRUE) %>% 
+  rename(icgc_sample_id = Tumor_Sample_Barcode) %>% 
+  mutate(Tumor_Sample_Barcode = stringr::str_trunc(submitted_sample_id, 16, ellipsis = ""))
 
-genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
-  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
-  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
-  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
-  .[rowSums(.[, -1])>0,]
+# vep.export <- all_muts %>% select(Chromosome, Start_Position, End_Position, Reference_Allele, Tumor_Seq_Allele2, Strand, icgc_mutation_id)
+# vep.export <- vep.export %>% arrange(Chromosome, Start_Position)
+# vep.export <- distinct(vep.export)
+# vep.export <- with(vep.export,
+#                    paste(Chromosome, Start_Position, End_Position,
+#                          paste(Reference_Allele, Tumor_Seq_Allele2, sep = "/"),
+#                          Strand, icgc_mutation_id, sep = "\t"))
+# writeLines(vep.export, con = "../data/vep.export.tsv")
+# vep settings availible in "../data/VEP_settings.png"
 
-df <- upset(genes_in_path, 
-      sets = names(genes_in_path),
-      mb.ratio = c(0.55, 0.45), 
-      order.by = "freq", 
-      point.size = 3.5,
-      line.size = 2.0,
-      mainbar.y.label = "Gene Intersection",
-      sets.x.label = "Total Gene Per Pathway",
-      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
-      number.angles = 0)
+clinical_plus_cluster <- readr::read_csv("../data/clinical_plus_cluster.csv")
+patient_ids <- clinical_plus_cluster$patient_id
+solid_tumor_studies <- unique(clinical_plus_cluster$type)
+tumor_clusters <- readr::read_tsv("../data/umap_3d_coors.tsv", guess_max = 10000)
 
-make_combinations <- function(x) {
-  l <- length(x)
-  mylist <- lapply(1:l, function(y) {
-    combn(x, y, simplify = FALSE)
-  })
-  mylist
+clinical_plus_cluster <- clinical_plus_cluster %>% rename(Tumor_Sample_Barcode = sample_id)
+clinical_plus_cluster <- left_join(clinical_plus_cluster,
+                                   tumor_clusters %>% 
+                                     select(starts_with(c("plot", "cluster", "clust_knn", "sample_id"))),
+                                   by = c("dna_mutation_id" = "sample_id"))
+all_muts <- all_muts %>% filter(Tumor_Sample_Barcode %in% clinical_plus_cluster$Tumor_Sample_Barcode)
+
+vep_output <- readr::read_tsv("../data/vep_output.basic.txt", na = c("", "NA", "-"), guess_max = 10000)
+
+selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", 
+                  "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant",
+                  "stop_lost", "start_lost", "stop_gained")
+selected_mut_alt <- c("3'UTR", "5'UTR", 
+                      "Frame_Shift_Del", "Frame_Shift_Ins",
+                      "Missense_Mutation", 
+                      "Splice_Site", "Translation_Start_Site", 
+                      "Nonsense_Mutation", "Nonstop_Mutation")
+
+vep_output <- vep_output %>% 
+  rename(icgc_mutation_id = "#Uploaded_variation")
+
+all_muts <- all_muts %>% 
+  left_join(vep_output, 
+            by = c("consequence_type" = "Consequence", 
+                   "icgc_mutation_id" = "icgc_mutation_id",
+                   "ens_id" = "Gene"))
+
+clinical_plus_cluster$clust_knn_char <- factor(clinical_plus_cluster$clust_knn, 
+                                               levels = c(1:10), 
+                                               labels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10"))
+
+all_muts <- left_join(all_muts, clinical_plus_cluster %>% select(Tumor_Sample_Barcode, clust_knn, clust_knn_char, type) %>% distinct())
+
+for( cluster_c in levels(clinical_plus_cluster$clust_knn_char) ) {
+  c_muts <- read.maf(maf = all_muts %>% filter(clust_knn_char == cluster_c),
+                      clinicalData = clinical_plus_cluster %>% filter(clust_knn_char == cluster_c),
+                      vc_nonSyn = selected_mut_alt, 
+                      removeDuplicatedVariants = FALSE)
+  other_muts <- read.maf(maf = all_muts %>% filter(clust_knn_char != cluster_c),
+                         clinicalData = clinical_plus_cluster %>% filter(clust_knn_char != cluster_c),
+                         vc_nonSyn = selected_mut_alt, 
+                         removeDuplicatedVariants = FALSE)
+
+  c_compare <- mafCompare( c_muts,
+                           other_muts )
+  c_compare <- c_compare$results %>% 
+    as_tibble() %>% 
+    filter(adjPval < 0.01) %>%
+    arrange(or) %>% 
+    mutate(Hugo_Symbol = factor(Hugo_Symbol, levels = Hugo_Symbol)) %>%
+    mutate(ci.up = log(ci.up), ci.low = log(ci.low), or = log(or)) %>%
+    mutate(cluster = cluster_c)
+
+  if (cluster_c == "01") {
+    all_compare <- c_compare
+  } else {
+    all_compare <- bind_rows(all_compare, c_compare)
+  }
 }
 
-path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10
-
-top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
-  'colnames<-' ("sum_top10") %>% ## rename col
-  rownames_to_column(var = "gene_name") ## make rownames a col
-
-path_match_in_set <- list() ## making a list for matchings in sets
-
-for(n in 1:10) {
-  message(n)
-  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
-  path_match_within_set <- list() ## making a list for matching within sets
-  for(p in 1:length(path_combo_set)) {
-    message(p)
-    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
-    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
-    if(length(spec_combo) > 1) {
-      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
-        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
-    }
-    if(length(spec_combo) == 1) {
-      names(spec_combo) <- "combo"
-    }
-    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
-    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
-    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
-    if(!"TRUE" %in% any_matches) { ## if no matches then skip
-      next
-      }
-    if("TRUE" %in% any_matches) { ## if are matches then do this
-      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
-        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
-    }
-  }    
-  if(length(path_match_within_set) > 0) {
-    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
-  }
-}
-path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together
-
-save(path_match_in_set, file = "path_match_in_set.Rda")
-load("../data/path_match_in_set.Rda")
-
-### getting clustering order from pheatmap
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## now need to get the gene mutation data and tumor samples 
-load("../data/gdc_tumor_coding_complete.Rdata")
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
-gc()
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
-  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
-  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
-rm(gdc_mut_count_patient)
-gc()
-
-save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
-
-fig4_pheatmap <- pheatmap(filtered_mut_data[gene_order,],
-         show_rownames = T,
-         show_colnames = F,
-         treeheight_row = 0,
-         treeheight_col = 0,
-         cellheight = 5,
-         cellwidth = 1,
-         cluster_cols = T,
-         cluster_rows = F, ## turn off row clustering
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         fontsize = 10,
-         legend = T)
-
-## getting sample orders
-pheatmap_order <- fig4_pheatmap$tree_col$order ## getting sample clustering order
-sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
-sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
-
-## stopped here ## ignore bottom using as reference
-
-class67_gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
-class67_gene_2_path_heatmap$membership <- as.character(class67_gene_2_path_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(rownames(filtered_mut_data))))
-class67_gene_2_path_heatmap$sample_id <- factor(class67_gene_2_path_heatmap$sample_id, levels = c(sample_order))
-
-class67_gene_2_path_heatmap <- left_join(class67_gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
-save(class67_gene_2_path_heatmap, file = "class67_gene_2_path_heatmap.Rda") 
-rm(list = ls(all.names = T))
-gc()
+save(all_compare, all_muts, clinical_plus_cluster, cluster_colors, file = "../data/forest.heatmap.data.rda")
 ```
 
-```{r child = '../code/figure4_function.Rmd'}
+```{r forest-heatmap-data-plot, fig.align='center', fig.width= 9, fig.height= 8, message=FALSE}
+#fig.width= 16, fig.height= 20,
+load("../data/forest.heatmap.data.rda")
+
+down_compare_plot <- all_compare %>%
+  group_by(cluster) %>%
+  filter(or < 0) %>%
+  filter(adjPval < 0.01) %>%
+  arrange(-or) %>%
+  # filter(M1 > 5) %>%
+  filter(is.finite(or)) %>%
+  slice_max(order_by = or, n = 10) %>%
+  arrange(cluster, or, .by.group = TRUE)
+down_compare_plot$Hugo_Symbol <- with(down_compare_plot, 
+                                      factor(as.character(Hugo_Symbol), 
+                                             levels = unique(as.character(Hugo_Symbol))))
+
+up_compare_plot <- all_compare %>%
+  group_by(cluster) %>%
+  filter(or > 0) %>%
+  arrange(-or) %>%
+  # filter(M2 > 5) %>%
+  filter(is.finite(or)) %>%
+  slice_max(order_by = or, n = 10) %>%
+  arrange(cluster, or, .by.group = TRUE)
+up_compare_plot$Hugo_Symbol <- with(up_compare_plot,
+                                    factor(as.character(Hugo_Symbol),
+                                           levels = unique(as.character(Hugo_Symbol))))
+
+up_p <- ggplot(up_compare_plot, aes(x = or, 
+                                    y = Hugo_Symbol, 
+                                    color = cluster, 
+                                    group = cluster)) +
+  geom_vline(xintercept = 0, linetype = 2) + 
+  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
+  scale_color_manual(values = cluster_colors) +
+  theme_classic() +
+  xlab("log odds ratio") +
+  xlim(c(0, 9)) +
+  ylab("Gene") +
+  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
+  theme(panel.grid.major.y = element_line(colour = "#f0f0f0"),
+        axis.title.y = element_blank()) +
+  ggtitle("Enriched")
+
+dn_p <- ggplot(down_compare_plot, aes(x = or, y = Hugo_Symbol, color = cluster, group = cluster)) +
+  geom_vline(xintercept = 0, linetype = 2) + 
+  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
+  scale_color_manual(values = cluster_colors, guide = guide_none()) +
+  theme_classic() +
+  xlab("log odds ratio") +
+  xlim(c(-9, 0)) + 
+  ylab("Gene") +
+  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
+  theme(panel.grid.major.y = element_line(colour = "#f0f0f0")) +
+  ggtitle("Depleted")
+
+tdat_up <- all_muts %>%
+  filter(Hugo_Symbol %in% up_compare_plot$Hugo_Symbol) %>%
+  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
+  select(type, Hugo_Symbol) %>%
+  table()
+
+tdat_up <- left_join(as.data.frame(tdat_up), 
+                  clinical_plus_cluster %>%
+                    select(Tumor_Sample_Barcode, type, clust_knn_char) %>% 
+                    distinct() %>% 
+                    group_by(type, clust_knn_char) %>% 
+                    summarise(ntumors = n())) %>% 
+  mutate(prop = Freq / ntumors) %>%
+  select(type, Hugo_Symbol, prop) %>%
+  left_join(up_compare_plot %>% select(Hugo_Symbol, cluster))
+tdat_up$Hugo_Symbol <- factor(tdat_up$Hugo_Symbol, levels = levels(up_compare_plot$Hugo_Symbol))
+
+tdat_up_p <- tdat_up %>% ggplot(aes(type, Hugo_Symbol, fill = prop)) + 
+  geom_tile() + 
+  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within \nCluster", limits = c(0,1)) + 
+  xlab("Tumor Type") +
+  ylab("Gene Symbol") +
+  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
+  theme_classic() +
+  theme(strip.text = element_blank(), 
+        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+        axis.text.y = element_blank(),
+        axis.title.y = element_blank(),
+        panel.grid.major.y = element_line(colour = "#f0f0f0"))
+
+tdat_dn <- all_muts %>%
+  filter(Hugo_Symbol %in% down_compare_plot$Hugo_Symbol) %>%
+  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
+  select(type, Hugo_Symbol) %>%
+  table()
+tdat_dn <- left_join(as.data.frame(tdat_dn), 
+                     clinical_plus_cluster %>%
+                       select(Tumor_Sample_Barcode, type, clust_knn_char) %>% 
+                       distinct() %>% 
+                       group_by(type, clust_knn_char) %>% 
+                       summarise(ntumors = n())) %>% 
+  mutate(prop = Freq / ntumors) %>%
+  select(type, Hugo_Symbol, prop) %>%
+  left_join(down_compare_plot %>% select(Hugo_Symbol, cluster))
+tdat_dn$Hugo_Symbol <- factor(tdat_dn$Hugo_Symbol, levels = levels(down_compare_plot$Hugo_Symbol))
+
+tdat_dn_p <- tdat_dn %>% ggplot(aes(type, Hugo_Symbol, fill = prop)) + 
+  geom_tile() + 
+  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within \nCluster", limits = c(0,1)) + 
+  xlab("Tumor Type") +
+  ylab("Gene Symbol") +
+  guides(fill = FALSE) +
+  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
+  theme_classic() +
+  theme(strip.text = element_blank(), 
+        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+        axis.text.y = element_blank(),
+        axis.title.y = element_blank(),
+        panel.grid.major.y = element_line(colour = "#f0f0f0"))
+
+layout <- "ABBCDD"
+dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
 
@@
 
-```{r figure-4-oncoprint, message = FALSE, fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits."}
-
-class67 <- oncoprint_style(filtered_class = c(6,7), n_path = 10)
-class67
-```
-
-```{r figure-4, message = FALSE, fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey).
-The genes are organized according to which of the top distinguishing pathways for this class that they effect.
-Some genes affect multiple pathways, as indicated by the bars at left.
-The highlighted genes at right are all proteasomal subunits.", eval = FALSE}
-load("../data/class67_gene_2_path_heatmap.Rda") ## load data
-load("../data/path_match_in_set.Rda")
-
-### getting facet ordering
-most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
-  'colnames<-' (c("path_combo", "Freq")) %>% 
-  rownames_to_column(var = "order")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-class67_gene_2_path_heatmap$gene_name <- factor(class67_gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
-
-fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_blank(),
-        axis.ticks = element_blank(),
-        strip.text.y = element_blank(),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-load("../data/path_match_in_set.Rda")
-
-gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
-  arrange(order) %>% .$gene_name ## getting the gene name order
-
-## right annotation data generation
-proteasome_r_anno <- data.frame(gene = gene_order, 
-                                psm_gene = 0)
-
-proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1
-
-r_anno_color <- c("0" = "white", "1" = "black")
-r_anno <- ggplot(proteasome_r_anno) +
-  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
-  theme_void() 
-legend <- plot_grid(get_legend(r_anno), ncol = 1)
-r_anno <- r_anno + theme(legend.position = "none")
-
-## left annotatoin data generation
-l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap
-
-## need to get top path names again
-load("../data/moderate_enrich_prob.Rda")
-enrich_cutoff <- 0.30
-filtered_class <- c(6,7)
-n_path <- 10
-toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
-  filter(clust %in% filtered_class) %>% ## filter the cluster
-  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
-  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways
-
-for(p in toppath){
-  message(p)
-  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
-  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
-  ## for any path_combo that includes the pathway name change the 0 into a 1
-  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
-    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
-  }
-  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
-  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
-  }
-}
-
-l_anno_ggplot <- list() ## list for storing ggplots
-path_legend <- list()
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-for(n in 1:length(toppath)){
-  message(n)
-  p <- toppath[n]
-  message(p)
-  spec_path <- l_anno[,c("gene_name", p)]
-  names(spec_path)[names(spec_path) == p] <- "p"
-  anno_color <- c("0" = "white", "1" = kelly.colours[2+n])
-  path_ggplot <- ggplot(spec_path) +
-  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
-           stat = "identity", 
-           width = 1) +
-  scale_fill_manual(values = anno_color, breaks = "1", name = "", labels = p) +
-  theme_void() 
-  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
-  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
-}
-
-plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
-plot_legend <- plot_grid(path_legend[[1]], path_legend[[2]], path_legend[[3]], path_legend[[4]], path_legend[[5]], path_legend[[6]], path_legend[[7]], path_legend[[8]], path_legend[[9]], path_legend[[10]], ncol = 5)
-
-plot_grid(plot, plot_legend, ncol = 1, rel_heights = c(10, 1.5))
-
-```
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Gene level analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence interval.
+Row represents the significant genes from each cluster.
+Each cluster was compared against the background (all other clusters) to find deferentially mutated genes.
+Significant genes (pvalue < 0.01) were selected and limited to the top ten results for each clusters.
+Heatmap; Column represents the cancer type.
+Row represents the significant genes as described above.
+The heatmap shows the proportion of samples per cluster, per cancer type that is mutated for each of the significant genes.
+Depleted significant genes (left).
+Enriched significant genes (right).</p>

### 2020-08-08 code/supplements.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

@@
-```{r}
-library(data.table)
-library(ggplot2)
-library(tidyverse)
-library(kableExtra)
-```
+### Supplementary Figure S1
 
-### Supplementary Figure S1
-```{r supplementary-figure-s1, out.width = '90%', fig.align='center'}
+```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
+#out.width = '90%',
 umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
@@
 
-```{r supplementary-figure-s2-data-generation, out.width = '90%', fig.align = "center"}
+```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
 umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
@@
 
-<div style="text-align: right"> Supplementary Figure S3 Data </div>
 ```{r supplementary-figure-s3-data-generation, eval = FALSE}
@@
 
-```{r supplementary-figure-s3, out.width='77%', fig.align = "center"}
+```{r supplementary-figure-s3, fig.width = 7, fig.height= 5, fig.align = "center"}
+# out.width = '77%'
 load("../data/selected_mut.Rda")
@@
 
-<div style="text-align: right"> Supplementary Figure S4 Data </div>
 ```{r supplementary-figure-s4-data-generation, eval = FALSE}
@@
 
-```{r supplementary-figure-s4, message = FALSE, echo = FALSE, out.width = '90%', fig.align= 'center'}
+```{r supplementary-figure-s4, message = FALSE, echo = FALSE, out.width = '90%', fig.align= 'center', eval = FALSE}
 load("../data/fig_s4_ggplot.Rda")
@@
 
-<div style="text-align: right"> Supplementary Figure ST1 Data </div>
 ```{r supplemental-table-st1-data-generation, eval = FALSE}
@@
 
-```{r supplemental-table-st1}
+```{r supplemental-table-st1, eval = FALSE}
 load("../data/table_st1.Rda")
+## for html only
+# table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable() %>%
+#   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
+#   column_spec(1, bold = T, color = "black")
 
-table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T, color = "black")
+## for word doc
+# table_st1 %>% flextable() %>% 
+#   set_caption(caption = "") %>%
+#   align(part = "all") %>% 
+#   width(., width = c(4,0.5,0.5,0.5,0.5)) %>% 
+#   fontsize(size = 8, part = "body")
+
+write.xlsx(table_st1, file = "../data/table_st1.xlsx", row.names = FALSE)
 ```
 
-![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]
+```{r enrich-pathway-in-clust}
+## docx friendly
+load("../data/moderate_enrich_prob.Rda")
+
+st1 <- moderate_enrich_prob %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
+  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
+
+st1$log10pvalue <- round(-log10((st1$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
+st1$Enrichment_Score <- round(st1$Enrichment_Score, digits = 2)
+st1$Signif <- NULL ## we don't need this anymore
+
+st1 %>% flextable() %>% 
+  set_caption(caption = "") %>%
+  align(part = "all") %>% 
+  width(., width = c(4,0.75,1,1)) %>% 
+  fontsize(size = 10, part = "body")
+```
 

### 2020-08-08 text/manuscript.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

--- 2020-08-06 text/manuscript.Rmd
@@
 output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
+    bookdown::word_document2:
     number_sections: FALSE
-    toc: TRUE
-    toc_float: TRUE
-    toc_depth: 3
-    code_folding: hide
-    theme: flatly
-    smart: true
+    toc: FALSE
 bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
 ---
@@
     keep_tex: yes
+
+  bookdown::html_document2:
+    fig_captions: TRUE
+    number_sections: FALSE
+    toc: TRUE
+    toc_float: TRUE
+    toc_depth: 3
+    code_folding: hide
+    theme: flatly
+    smart: true
+
+  bookdown::word_document2:
+    number_sections: FALSE
+    toc: FALSE
 ```
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, dpi=600)
 suppressPackageStartupMessages(c(
@@
   library(rtracklayer),
-  library(poibin)
+  library(poibin),
+  library(flextable)
 ))
@@
 ```{r child='../code/figure2.Rmd'}
+
 ```
@@
 
+<strike>
 ### Gene-Pathway redundancy contributes to cluster identity
@@
 
-```{r child='../code/figure4_function.Rmd'}
-```
-
 The proteasome plays key roles in the transduction of signals from the p53, WNT
@@
 represent a targetable category of tumors.
+</strike>
+
+```{r child = '../code/figure4.Rmd'}
+
+```
 
@@
 
-<mark>As discussed previously, prior efforts to extract common signatures from
+As discussed previously, prior efforts to extract common signatures from
 pan-cancer datasets met with difficulty in distinguishing tumor samples from
@@
 few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we ranked the top cluster-specific
-pathways for each cluster, and then tabulated the most frequently mutated genes
-for each pathway as a function of cancer type (**Table 1** and **Supplementary tables ST2-ST13**). 
-We find that in most clusters the top pathways are
-determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
-However in many cancers we find divergence outside these top driver mutations,
-wherein etc etc</mark>
-
-```{r child='../code/table1.Rmd'}
-```
+combination of both? To answer this question, we compared each cluster relative
+to the background to find deferentially mutated genes. We ranked the odd ratios 
+and selected the top ten enriched and depleted genes (pvalue < 0.01) for each 
+cluster (**Figure 4**; odds ratios plot). We find that the clusters possess 
+unique gene sets. Cluster 2 is enriched for genes responsible for protein
+homeostasis. Cluster 4 is enriched for genes responsible for nucleosome 
+structure. Clusters 7 and 8, which shared multiple enrichment in signaling
+pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
+genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
+Interestingly, cluster 9 is enriched for both _PIK3CA_ and _KRAS_ which may
+suggest tumors in this cluster derived from either clusters 7 or 8. Next, we 
+investigated the proportion of samples per cancer type for the significant genes 
+within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors 
+were not predominated by one or more highly mutated genes across all cancers. 
+Instead, some cancer-specific tumors in these classes had mutation profiles that
+differ from one another (_e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC
+was enriched for _UBC_, and PAAD was enriched for _PSMA1_).
 
@@
 Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found a total of 27 enriched pathways across all clusters (**table 2**). A number of
+found a total of 27 enriched pathways across all clusters (**table 1**). A number of
 enrichments were found in multiple clusters represented by pathways that were
@@
 **supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
-metastatic tumors of clusters 7 and 9 (p < 10-3, **table 2**). This is also true of 
+metastatic tumors of clusters 7 and 9 (p < 10-3, **table 1**). This is also true of 
 “Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
@@
 Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**). </strike>
+Response”, and “Pre-NOTCH Expression and Processing” (**Table 1**). </strike>
 
@@
 ```{r child='survival.Rmd'}
-```
-
-```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
-set.seed(21)
-plot(x=rnorm(1000), y=rnorm(1000))
 ```

### 2020-08-09 text/manuscript.Rmd
Move figures to end of paper. Move dicussion in front of method. Order is now: intro, result, discussion, methods, main figs, supplemental figs

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, dpi=600)
+knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
 suppressPackageStartupMessages(c(
@@
 
-```{r child='../code/figure1.Rmd'}
-```
 In order to classify tumors using this dataset, we used multiple correspondence
@@
 
-```{r child='../code/figure2.Rmd'}
-
-```
-
 To discover what pathways drive the UMAP clustering, we calculated enrichment
@@
 
-```{r child = '../code/figure4.Rmd'}
-
-```
-
 ### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
@@
 
-```{r child='../code/table2.Rmd'}
-```
 <mark>Statement does not apply anymore. The highest # of enriched path is 3. Not many are in 
@@
 ```{r child='survival.Rmd'}
+```
+
+## Discussion
+
+```{r child='discussion.Rmd'}
 ```
@@
 
-## Discussion
+```{r child='../code/figure1.Rmd'}
+```
 
-```{r child='discussion.Rmd'}
+```{r child='../code/figure2.Rmd'}
 ```
+
+```{r child = '../code/figure4.Rmd'}
+```
+
+```{r child='../code/table2.Rmd'}
+```
+
+![](../figures/fig6B.png)
+![](../figures/fig6C.png)
 

### 2020-08-09 text/manuscript.Rmd
Update filtering of genes method

@@
 
-One of the limitations of knowledge-based pathway analysis is that it heavily relies on 
-experimental results for the curation and validation of its pathways. The resulting 
-pathways are therefore not specific to a particular tissue type but provides information
-for all tissue type. Before utilizing the gene to pathway lookup table, we considered the
-impact of deleterious passenger mutations for a given cancer (non-specific genes that are
-not expressed) as they may introduce biases in our pathway analysis. We addressed this by
-identifying genes that are lowly/not expressed in a tissue type. Using the TCGA RNA 
-sequencing data adjusted for batch effect dataset (https://pancanatlas.xenahubs.net), we 
-filtered the genes and tumor samples that were not included in our analysis, grouped the 
-tumor samples by tissue type and computed the mean expression value for each gene. A minimum
-threshold of 10 transcripts per million was set for expressed genes based on an inflection
-point observed when plotting the mean expression values of genes ranked by increasing 
-expression values for each tissue type. Genes that did not meet this threshold were 
-considered not expressed. This operation produced a lookup table for gene expression
+We filtered likely erroneous mutations due to transcription coupled repair. Our approach
+was to determine the status of each gene (expressed or not expressed) per tissue type in
+order to exclude gene mutations that we have evidence for no expression. To do this, we 
+obtained the TCGA RNA sequencing data adjusted for batch effect dataset (https://pancanatlas.xenahubs.net).
+Using the data, we removed the genes and tumor samples that were not included in our analysis, 
+grouped the tumor samples by tissue type and computed the mean expression value for each 
+gene. A minimum threshold of 10 transcripts per million was set for expressed genes based 
+on an inflection point observed when plotting the mean expression values of genes ranked 
+by increasing expression values for each tissue type. Genes that did not meet this threshold 
+were considered not expressed. This operation produced a lookup table for gene expression
 status in each tumor sample for 18,127 genes.

### 2020-08-10 code/figure1.Rmd

@@
----
-output: html_document
-editor_options: 
-  chunk_output_type: console
----
-```{r}
-library(data.table)
-library(ggplot2)
-library(ggplotify)
-library(tidyverse)
-library(grid)
-library(patchwork)
-```
-
-<div style="text-align: right"> Generation of Somatic Mutation Data </div>
 ```{r data-set-creation, eval=FALSE}
@@
 ```
-<div style="text-align: right"> Reactome Pathway Creation </div>
+
 ```{r reactome-pathway-creation, eval=FALSE}
@@
 ```
-<div style="text-align: right"> Filter tissue-specific low expressed genes </div>
+
 ```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
@@
 ```
-<div style="text-align: right"> Genes to pathway transformation </div>
+
 ```{r genes-to-pathways-transformation, eval=FALSE}
@@
 ```
-<div style="text-align: right"> Figure 1 data </div>
+
 ```{r tumor-clusters, eval = FALSE}
@@
 
-```{r fig-1a-1l, warning=FALSE, fig.height=15, fig.width=16, message = FALSE}
+```{r fig-1a-1l, warning=FALSE, fig.height=12, fig.width=11, message = FALSE}
+#fig.height=15, fig.width=16,
 load("../data/cancer_disrupt_heatmap.Rda")

### 2020-08-10 code/table1.Rmd

@@
 
-cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(format = "html", row.names = FALSE) %>%
+cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(row.names = FALSE) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%

### 2020-08-10 code/table2.Rmd

@@
-<div style="text-align: right"> Table 2 Data </div>
 ```{r table-2-data-generation, eval = FALSE}
@@
 
-```{r table-2}
+```{r table-2, eval = FALSE}
+## this is for html only
+library(kableExtra)
+library(tidyverse)
+knitr::opts_chunk$set(echo = FALSE, dpi=600)
 load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
@@
 
-table2 %>% knitr::kable(format = "html") %>%
+table2 %>% knitr::kable() %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
@@
 ```
+
+```{r, echo = FALSE}
+## docx friendly
+load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
+
+table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
+  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
+
+table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
+table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
+table2$Signif <- NULL ## we don't need this anymore
+
+table2 %>% flextable() %>% 
+  set_caption(caption = "") %>%
+  align(part = "all") %>% 
+  width(., width = c(4,0.75,1,1)) %>% 
+  fontsize(size = 10, part = "body")
+
+```

### 2020-08-10 text/introduction.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

@@
----
-title: "Introduction"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: FALSE
-    theme: flatly
-bibliography: references.bib
----
-
 Recent advances in technology have made the study of cancer possible at
@@
 sparseness of mutations, allowing the potential importance of a gene to be
-influenced by the aggregate of mutations in its nearest network
-neighbors. This method is powerful especially when applied to a small number (on
-the order of hundreds) of tumors. However, the larger problem of cancer etiology
-remains: what mutations are required, in what order, and are there specific
-combinations of delinquent cellular processes sufficient to cause cancer? If
-there is a common logic to cancer, how is it to be detected underneath the
-overwhelming cell-of-origin signal?
+influenced by the aggregate of mutations in its nearest network neighbors. This
+method is powerful especially when applied to a small number (on the order of
+hundreds) of tumors. One of the challenges in the field remains that beyond a
+couple hundred common driver genes and mutations, there are likely thousands of
+moderate effect genes that occur at such low frequency given tissue diversity
+and low sample numbers that they are impossible to detect using positive
+selection theory. Some groups have attempted to address this challenge by
+applying machine learning to cancer data sets to discover groups of functionally
+related genes as they interact with larger pathways and networks
+[@kimPancancerAnalysisSomatic2018, @mourikis-ml-pathways, @moonlight]. These
+studies have proven very effective at highlighting some potentially fundamental
+disease phenotypes at the pathway level that cross cancers with different
+origins at the cellular and tissue level.
 
@@
 molecular pathways as the unit of disruption in cancer. Our approach strips all
-cell-type-specific information from the mutation data and generalizing
-gene-level information to non-tissue specific cell-biological pathway
-phenotypes. We use this statistic to evaluate all cancers and show that,
-surprisingly, tumors may have distinct phenotypes completely independent of
-their tissues of origin.
+cell-type-specific information from the mutation data and generalizes gene-level
+information to cell-biological pathway phenotypes. We use this heuristic to
+evaluate all cancers and show that, surprisingly, tumors fall into categories of
+common pathway disruptions with distinctive origins arising from tissue-specific
+gene alterations and having unique prognosis in each cancer type.

### 2020-08-10 text/manuscript.Rmd

@@
 title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
+author: Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and
+  Dennis J. Hazelett$^{1,3}$
+bibliography: references.bib
 output:
-    bookdown::word_document2:
-    number_sections: FALSE
-    toc: FALSE
-bibliography: references.bib
+  word_document:
+    reference_docx: word-style-reference-ttmanu.docx
+always_allow_html: TRUE
+editor_options:
+  chunk_output_type: console
 ---
@@
 
+output:
+  word_document:
+    reference_docx: word-styles-reference-01.docx
+  number_sections: FALSE
+  toc: FALSE
+
+output: 
 bookdown::html_document2:
@@
   smart: true
-
-  bookdown::word_document2:
-    number_sections: FALSE
-    toc: FALSE
 ```

### 2020-08-10 text/survival.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx
merge with upstream, and make progress on doing a large multi-component survival figure

@@
 
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
+```{r survfig}
+
+library(ggridges)
+library(patchwork)
+
+load("../code/survival/survmodel-ethbkd.stansave")
+
+cancerdat = read_csv("../data/survival/clinical_plus_cluster.csv");
+cancerdat$type = as.factor(cancerdat$type)
+
+dfsamps = as.data.frame(sampseth)
+
+kcancplot = function(dfsamps,tisnames){
+	ktisnames = grep("ktis",names(dfsamps),value=TRUE)
+	dfktis = dfsamps[,ktisnames]
+	dfktis = stack(dfktis) ## convert to two column "values" and "ind" with "ind" being the name of the variable
+	names(dfktis) = c("value","parname")
+
+	dim(tisnames) = c(length(tisnames),1)
+
+	dfktis$cancname = sapply(dfktis$parname,function(x) tisnames[x])
+	ggplot(dfktis) + geom_density_ridges(aes_(~value,~cancname))+ theme_minimal() + scale_fill_brewer(palette="Paired") + xlim(0,22) + labs(x="ktis multiplier",y="Cancer Type",title="Tissue Specific Effects")
+}
+
+kclassplot = function(dfsamps,tisnames){
+	kclnames = grep("k\\[",names(dfsamps), value=TRUE)
+	dfclas = dfsamps[,kclnames]
+	dfclas = stack(dfclas)
+	names(dfclas) = c("value","coefname")
+	dfclas$class = as.numeric(gsub("k\\[(.*),(.*)\\]","\\2",dfclas$coefname))
+	dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
+	ggplot(dfclas) + geom_density_ridges(aes_(~value,~class)) +
+		facet_wrap(. ~ tiss,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
+		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients")
+}
+
+p1 = kcancplot(dfsamps,levels(cancerdat$type))
+p2 = kclassplot(dfsamps,levels(cancerdat$type))
+
+p1 + p2
+
+```
 
@@
 treatments typical used for each tissue.
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)

### 2020-08-10 code/survival/kmcurve.R

@@ -2,10 +2,11 @@ library(survival)
 library(ggfortify)
 library(gridExtra)
 library(readr)
+library(sqldf)
 
-
-install.packages("devtools")
-devtools::install_github("sachsmc/ggkm")
+# run the following if you don't have ggkm
+# install.packages("devtools")
+# devtools::install_github("sachsmc/ggkm")
 library(ggkm)
 
 
@@ -75,10 +76,28 @@ asampseth = as.array(sampseth)
 
 cancerdat = read_csv("../../data/survival/clinical_plus_cluster.csv");
 cancerdat$type = as.factor(cancerdat$type) ## the tissue
+endpoints = sqldf("select case when death_days_to is NULL then last_contact_days_to else death_days_to end as finaltime, case when death_days_to is NULL then 0 else 1 end as finalstatus from cancerdat;")
+
+cancerdat = cbind(cancerdat,endpoints);
+sum(is.na(cancerdat$finaltime))
+cancerdat = cancerdat[!is.na(cancerdat$finaltime),]
+
+
+
+plotkmcomp = function(sim,ref,act,actstatus){
+    ggplot(data.frame(stime=sim,rtime=ref,acttime=act,status=rep(1,length(ref)),actstatus=actstatus)) +
+        geom_km(aes(time=stime,status=status),color="red") +
+        geom_km(aes(time=rtime,status=status),color="orange") +
+        geom_km(aes(time=acttime,status=actstatus),color="blue")
+}
+    
+
 
 plotcancertype = function(cancerdat,i){
-    agesw = cancerdat[cancerdat$type == i & cancerdat$gender=="FEMALE","age_at_initial_pathologic_diagnosis"]$age_at_initial_pathologic_diagnosis
-    agesm = cancerdat[cancerdat$type == i & cancerdat$gender=="MALE","age_at_initial_pathologic_diagnosis"]$age_at_initial_pathologic_diagnosis
+    women = cancerdat[cancerdat$type == i & cancerdat$gender=="FEMALE",]
+    men = cancerdat[cancerdat$type == i & cancerdat$gender=="MALE",]
+    agesw = women$age_at_initial_pathologic_diagnosis
+    agesm = men$age_at_initial_pathologic_diagnosis
     kmean = mean(asampseth[,,sprintf("ktis[%d]",j)])
     print(sprintf("Cancer %s kmean = %f\n",i,kmean))
     try(hist(agesw),silent=TRUE)
@@ -86,29 +105,31 @@ plotcancertype = function(cancerdat,i){
 
[…]

### 2020-08-11 code/figure1.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

@@
 ```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abbreviations from the TCGA project. b) MCA analysis was performed to 
-determine the eigenvalue with the most explanatory power from the categorical data.
-Average of 100 permutations of the data was used as a baseline. c) The p-value of
-each eigenvalue was plotted and the max eigenvalue with a pvalue < 0.05 was selected. 
-d) The resulting MCA was projected down to 3 dimensions for the purpose of density 
-based clustering (only the first 2 dimensions are shown). Each dot represents a tumor 
-e) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize
-the fraction of unassigned tumors. f) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). g) kNN was used to assign
-membership to the remaining tumors. h) Mutational burden plot where each dot corresponds
-to the sample mutation counts for tumors in each class. i) Projection of cancer 
-type identity onto UMAP. Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma,
-HNSC = Head and neck squamous cell, PAAD = Pancreatic adenocarcinoma. j) Projection of breast
-cancer subtypes onto the UMAP. k) Projection of BRCA1/2 somatic mutation onto the
-UMAP. l) Projection of tumor stage onto the UMAP, regardless of cancer type. m) 
-Projection of metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, 
-M1 = metastatic tumors.</p>

### 2020-08-11 code/figure2.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

@@
 ```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 2: Pan-cancer enrichment of pathway disruptions.** a) Waterfall plot; columns represent tumors.
-Each row represents a pathway.
-The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively.
-The top 100 most mutated pathways are shown.
-Red signifies pathway disrupted pathways, grey signifies not disrupted.
-b) Heatmap shows the relative enrichment of each pathway (columns) within each numbered cluster (rows).
-Effect size is displayed as colors (orange, enrichment; purple, depletion).
-c) Proportion of mutated samples in the each significant pathway (columns; union set of pathways with effect size >= 0.30 in each cluster) within each numbered cluster (rows).</p>

### 2020-08-11 code/figure4.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

@@
 dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
-
 ```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Gene level analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence interval.
-Row represents the significant genes from each cluster.
-Each cluster was compared against the background (all other clusters) to find deferentially mutated genes.
-Significant genes (pvalue < 0.01) were selected and limited to the top ten results for each clusters.
-Heatmap; Column represents the cancer type.
-Row represents the significant genes as described above.
-The heatmap shows the proportion of samples per cluster, per cancer type that is mutated for each of the significant genes.
-Depleted significant genes (left).
-Enriched significant genes (right).</p>

### 2020-08-11 code/supplements.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

@@
 ### Supplementary Figure S1
-
 ```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
@@
 ```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
-
+\newpage
 ### Supplementary Figure S2
-
 ```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
@@
 ```
-
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
+\newpage
 ### Supplementary Figure S3
-
 ```{r supplementary-figure-s3-data-generation, eval = FALSE}
@@
 ```
-
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
-
+\newpage
 ### Supplementary Figure S4
-
 ```{r supplementary-figure-s4-data-generation, eval = FALSE}
@@
 ```
-
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
-
+\newpage
+### Supplementary Figure S5
 ```{r, supplemental-s5-data-generation, eval = FALSE}
@@
 ```
-
-### Supplementary Table ST1
-
+\newpage
+### Supplementary Table 1
 ```{r supplemental-table-st1-data-generation, eval = FALSE}

### 2020-08-11 code/table2.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

@@
 ## this is for html only
-library(kableExtra)
-library(tidyverse)
-knitr::opts_chunk$set(echo = FALSE, dpi=600)
 load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
@@
 
-```{r, echo = FALSE}
+```{r table2-docx-friendly}
 ## docx friendly
@@
 
-table2 %>% flextable() %>% 
+table2 <-table2 %>% flextable() %>% 
    set_caption(caption = "") %>%
@@
 
+knitr::knit_print(table2)
 ```

### 2020-08-11 text/manuscript.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

@@
   word_document:
-    reference_docx: word-style-reference-ttmanu.docx
+    reference_docx: "../text/word-style-reference-ttmanu.docx"
 always_allow_html: TRUE
@@
 ### A taxonomy of tumors based on disrupted molecular pathways
-
 To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
@@
 
-As discussed previously, prior efforts to extract common signatures from
-pan-cancer datasets met with difficulty in distinguishing tumor samples from
-their tissue-specific -omics data signatures. Given our pathway-disruption based
-classification, this raises the question, are tumor phenotypes driven entirely
-by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
-few samples to detect above statistical significance thresholds), or a
-combination of both? To answer this question, we compared each cluster relative
-to the background to find deferentially mutated genes. We ranked the odd ratios 
-and selected the top ten enriched and depleted genes (pvalue < 0.01) for each 
-cluster (**Figure 4**; odds ratios plot). We find that the clusters possess 
-unique gene sets. Cluster 2 is enriched for genes responsible for protein
-homeostasis. Cluster 4 is enriched for genes responsible for nucleosome 
-structure. Clusters 7 and 8, which shared multiple enrichment in signaling
-pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
+As discussed previously, prior efforts to extract common signatures from pan-cancer 
+datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
+-omics data signatures. Given our pathway-disruption based classification, this raises 
+the question, are tumor phenotypes driven entirely by common driver genes, or by “silent”
+tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
+thresholds), or a combination of both? To answer this question, we compared each cluster 
+relative to the background to find deferentially mutated genes. We ranked the odd ratios
+and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
+(**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
+signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
 genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
-Interestingly, cluster 9 is enriched for both _PIK3CA_ and _KRAS_ which may
-suggest tumors in this cluster derived from either clusters 7 or 8. Next, we 
-investigated the proportion of samples per cancer type for the significant genes 
-within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors 
-were not predominated by one or more highly mutated genes across all cancers. 
-Instead, some cancer-specific tumors in these classes had mutation profiles that
-differ from one another (_e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC
-was enriched for _UBC_, and PAAD was enriched for _PSMA1_).
+Interestingly, cluster 9, which also shared multiple enrichment in signaling pathways 
+with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
+that tumors in this cluster could be a derivative either cluster 7 or 8. Clusters 3 and 
+5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
+nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible for 
+nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
+Next, we investigated the proportion of samples per cancer type for the significant genes
+within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors were not 
+predominated by one or more highly mutated genes across all cancers. Instead, several 
+cancer-specific tumors in these classes had mutation profiles that differ from one another 
+( _e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC was enriched for _UBC_, and 
+PAAD was enriched for _PSMA1_). Taken together, our data identifies a framework cancer 
+type specific mutation associated with a specific cluster.
 
@@
 ```
+\newpage
+```{r child='../text/fig_legend.Rmd'}
+```
 
+\newpage
 ```{r child='../code/figure1.Rmd'}
 ```
-
+\newpage
 ```{r child='../code/figure2.Rmd'}
 ```
-
+\newpage
 ```{r child = '../code/figure4.Rmd'}
 ```
-
+\newpage
 ```{r child='../code/table2.Rmd'}
 ```
+\newpage
 
 ![](../figures/fig6B.png)
+\pagebreak
 ![](../figures/fig6C.png)

### 2020-08-11 code/figure4.Rmd
fixes error in plotting of heatmap in figure4 that used to incorrectly compute the number of mutations independent of cluster.

--- 2020-08-08 code/figure4.Rmd
@@
   distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
-  select(type, Hugo_Symbol) %>%
+  select(type, clust_knn_char, Hugo_Symbol) %>%
   table()
@@
   mutate(prop = Freq / ntumors) %>%
+  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
   select(type, Hugo_Symbol, prop) %>%

### 2020-08-11 code/figure4.Rmd
fixes the previous fix, this time correctly calculating the within cluster / tumor type proportion

--- 2020-08-08 code/figure4.Rmd
@@
   arrange(-or) %>%
-  # filter(M1 > 5) %>%
   filter(is.finite(or)) %>%
@@
   arrange(-or) %>%
-  # filter(M2 > 5) %>%
   filter(is.finite(or)) %>%
@@
 
-up_p <- ggplot(up_compare_plot, aes(x = or, 
+up_p <- ggplot(up_compare_plot, 
+               aes(x = or,
                    y = Hugo_Symbol,
@@
   facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  theme(panel.grid.major.y = element_line(colour = "#f0f0f0"),
-        axis.title.y = element_blank()) +
   ggtitle("Enriched")
+  theme(
+    panel.grid.major.y = element_line(colour = "#f0f0f0"),
+    axis.title.y = element_blank()
+  )
 
-dn_p <- ggplot(down_compare_plot, aes(x = or, y = Hugo_Symbol, color = cluster, group = cluster)) +
+dn_p <- ggplot(down_compare_plot, 
+               aes(x = or, 
+                   y = Hugo_Symbol, 
+                   color = cluster, 
+                   group = cluster)) +
   geom_vline(xintercept = 0, linetype = 2) +
@@
   facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  theme(panel.grid.major.y = element_line(colour = "#f0f0f0")) +
-  ggtitle("Depleted")
+  ggtitle("Depleted") +
+  theme(panel.grid.major.y = element_line(colour = "#f0f0f0"))
 
@@
   distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
-  select(type, Hugo_Symbol) %>%
+  select(type, clust_knn_char, Hugo_Symbol) %>%
   table()
-
-tdat_up <- left_join(as.data.frame(tdat_up), 
+tdat_up <- left_join(
+  as.data.frame(tdat_up),
   clinical_plus_cluster %>%
@@
   mutate(prop = Freq / ntumors) %>%
-  select(type, Hugo_Symbol, prop) %>%
-  left_join(up_compare_plot %>% select(Hugo_Symbol, cluster))
+  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
+  select(type, clust_knn_char, Hugo_Symbol, prop) %>%
+  rename(cluster = clust_knn_char)
 tdat_up$Hugo_Symbol <- factor(tdat_up$Hugo_Symbol, levels = levels(up_compare_plot$Hugo_Symbol))
+tdat_up <- up_compare_plot %>% left_join(tdat_up)
 
@@
   geom_tile() +
-  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within \nCluster", limits = c(0,1)) + 
+  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
   xlab("Tumor Type") +
@@
   theme_classic() +
-  theme(strip.text = element_blank(), 
+  theme(
+    strip.text = element_blank(),
     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
@@
     axis.title.y = element_blank(),
-        panel.grid.major.y = element_line(colour = "#f0f0f0"))
+    panel.grid.major.y = element_line(colour = "#f0f0f0")
+  )
 
@@
   distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
-  select(type, Hugo_Symbol) %>%
+  select(type, clust_knn_char, Hugo_Symbol) %>%
   table()
-tdat_dn <- left_join(as.data.frame(tdat_dn), 
+tdat_dn <- left_join(
+  as.data.frame(tdat_dn),
   clinical_plus_cluster %>%
@@
   mutate(prop = Freq / ntumors) %>%
-  select(type, Hugo_Symbol, prop) %>%
-  left_join(down_compare_plot %>% select(Hugo_Symbol, cluster))
+  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
+  select(type, clust_knn_char, Hugo_Symbol, prop) %>%
+  rename(cluster = clust_knn_char)
 tdat_dn$Hugo_Symbol <- factor(tdat_dn$Hugo_Symbol, levels = levels(down_compare_plot$Hugo_Symbol))
+tdat_dn <- down_compare_plot %>% left_join(tdat_dn)
 
@@
   geom_tile() +
-  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within \nCluster", limits = c(0,1)) + 
+  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
   xlab("Tumor Type") +
   ylab("Gene Symbol") +
-  guides(fill = FALSE) +
   facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
   theme_classic() +
-  theme(strip.text = element_blank(), 
+  theme(
+    strip.text = element_blank(),
     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
@@
     axis.title.y = element_blank(),
-        panel.grid.major.y = element_line(colour = "#f0f0f0"))
+    panel.grid.major.y = element_line(colour = "#f0f0f0")
+  )
 
 layout <- "ABBCDD"
-dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
-
+dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = "collect", design = layout)
 ```

### 2020-08-11 code/figure4.Rmd

@@
 layout <- "ABBCDD"
-dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = "collect", design = layout)
+
+dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
 ```
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Gene level analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence interval.
-Row represents the significant genes from each cluster.
-Each cluster was compared against the background (all other clusters) to find deferentially mutated genes.
-Significant genes (pvalue < 0.01) were selected and limited to the top ten results for each clusters.
-Heatmap; Column represents the cancer type.
-Row represents the significant genes as described above.
-The heatmap shows the proportion of samples per cluster, per cancer type that is mutated for each of the significant genes.
-Depleted significant genes (left).
-Enriched significant genes (right).</p>

### 2020-08-13 code/figure1v2.Rmd
Split figure 1 into 4 figures, update figure4.rmd with simon's fix, add page breaks

--- 2020-08-11 code/figure1.Rmd
@@
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] %>% ## include only second level pathways
-  left_join(., main2first, by = c("PATH_ID" = "first_level")) %>% distinct() %>% 
-  left_join(., ensembl2reactome[,c(2,4)], by = c("main_level" = "V2")) %>% distinct()
-
-## Filtering pathways not relevant to our analysis by keywords
-filtered_ensembl2reactome <- ensembl2reactome %>% filter(!V4 %in% c("Disease", "Muscle contraction", "Reproduction")) ## Pathways in disease, muscle contraction and reproduction parent pathway were removed
-
-toMatch <- c("disease", "defect", "cancer", "oncogenic", "disorder", "infection", "mutant", "bacteria", "listeria", "loss") ## pathways with the following key terms were removed.
-
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% mutate(l_pathname = tolower(.$PATH_NAME)) %>% 
-  .[!grep(paste(toMatch,collapse="|"), .$l_pathname),]
-
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% .[which(!.$PATH_ID %in% c("R-HSA-212436", "R-HSA-388396")),] ## Removing Generic Transcription Pathway and GPCR downstream signalling as they are broad pathways
-
-gene2level2_lookup <- filtered_ensembl2reactome %>% dplyr::select(ENSEMBL, PATH_ID) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "../data/gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- filtered_ensembl2reactome %>% dplyr::select(PATH_ID, PATH_NAME) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "../data/pathname2level2.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
+```{r figure-1-data-generation, eval = FALSE}
+###################
+## Figure 1 data ##
+###################
 load("../data/gdc_reactome_path_analysis.Rda")
@@
 save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
-
-####################
-## Figure 1B data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>%
-  t() %>%
-  as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(tumor_samples$submitted_sample_id, .$sample), ]
-
-gdc_path_mca <- left_join(gdc_path, tumor_samples[, c("project_code", "submitted_sample_id")],
-  by = c("sample" = "submitted_sample_id")) %>%
-  column_to_rownames(var = "sample")
-
-gdc_path_sample_names <- rownames(gdc_path_mca) ## Getting sample id order
-gdc_path_tumor_types <- gdc_path_mca$project_code ## Getting cancer type order
-
-gdc_path_mca$project_code <- NULL ## Remove project_code column from df
-
-## create a categorical variable from the boolean variable
-pathway_list <- make_clean_names(colnames(gdc_path_mca))
-pathway_count <- length(pathway_list)
-gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
-  factor(x, levels = c(0, 1), labels = c("no", "yes"))
-}))
-colnames(gdc_path_mca) <- pathway_list
-gdc_path_mca$project_code <- gdc_path_tumor_types
-rownames(gdc_path_mca) <- gdc_path_sample_names
-
-## perform mca
-mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
-
-expl_var <- mca_res$eig[, 2]
-expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
-
-set.seed(4242)
-for (k in seq(100)) {
-  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
-  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
-  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
-}
-
-randomize <- tibble(
-  eigenvalue = 1:length(expl_var),
-  expl_var = expl_var,
-  perm = colMeans(expl_var_perm),
-  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
-)
-
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-mca_res <- MCA(
-  X = gdc_path_mca[, pathway_list],
-  ncp = optimal_mca_dimensions, graph = FALSE
-)
-
-save(randomize, mca_res, file = "figure1B_data.rda")
-
-####################
-## Figure 1C data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-rng <- RNGseq(max_n_pt * n_iter, 42)
-r <- rng[[3649]]
-
-rngtools::setRNG(r)
-gdc.umap <- umap(mca_res$ind$coord, 
-                 n_components = 3,
-                 n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
-                 n_epochs = 1000,
-                 n_threads = 1)
-gdc_umap_coors <- as.data.frame(gdc.umap) %>%
-  rename(plot_x = V1, plot_y = V2, plot_z = V3)
-rownames(gdc_umap_coors) <- rownames(gdc_path_mca)
-
-####################
-## Figure 1D data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
-
-n_tumors <- nrow(mca_res$ind$coord)
-
-cl <- makeCluster(8)
-registerDoParallel(cl)
-
-for (i in 3:max_n_pt) {
-  print(paste0("MIN POINTS = ", i))
-  score_iter <- foreach(1:n_iter,
-    .combine = c,
-    .export = c("mca_res"),
-    .packages = c("dbscan", "uwot"),
-    r = rng[(i - 1) * n_iter + 1:n_iter]
-  ) %dopar% {
-    rngtools::setRNG(r)
-    umap_init_mca <- umap(mca_res$ind$coord, 
-                          min_dist = 1e-6, 
-                          n_components = 3, 
-                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
-                          n_epochs = 1000, 
-                          n_threads = 1)
-    res <- hdbscan(umap_init_mca, minPts = i)
-    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
-    return(score_iter_temp)
-  }
-  score_mca[i - 2, ] <- score_iter
-}
-
-score <- rowMeans(score_mca)
-score_df <- tibble(minPts = 3:max_n_pt, score = score)
-
-sample_df <- score_mca %>%
-  as.data.frame() %>%
-  as_tibble() %>%
-  add_column(minPts = 3:max_n_pt, .before = 1) %>%
-  pivot_longer(-minPts) %>%
-  group_by(minPts)
-
-optimal_minPts <- score_df %>%
-  filter(score == min(score)) %>%
-  pull(minPts)
-
-save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
-
-####################
-## Figure 1E data ##
-####################
-rngtools::setRNG(r)
-
-cluster_umap <- umap(mca_res$ind$coord,
-  min_dist = 1e-6,
-  n_components = 3,
-  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
-  n_epochs = 1000,
-  n_threads = 1
-)
-colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
-hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
-umap_3d_coors <- gdc_umap_coors %>%
-  bind_cols(as.data.frame(cluster_umap)) %>%
-  mutate(
-    n_pathways_mut = rowSums(gdc_path[, -1]),
-    clust = as.factor(hdb_cluster_assignment$cluster),
-    member_prob = hdb_cluster_assignment$membership_prob,
-    sample_id = rownames(gdc_umap_coors)
-  )
-## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_3d_coors$clust_knn <- umap_3d_coors$clust
-umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
-  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
-  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
-  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
-)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "sample_id") ## move rownames to column
-
-umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id") 
-rm(mutation_count)
-
-####################
-## Figure 1H data ##
-####################
-umap_3d_coors <- left_join(umap_3d_coors, tumor_samples[,c("project_code", "submitted_sample_id")], by = c("sample_id" = "submitted_sample_id"))
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
 ```
 
-```{r fig-1a-1l, warning=FALSE, fig.height=12, fig.width=11, message = FALSE}
-#fig.height=15, fig.width=16,
+```{r, fig.width = 9, fig.height= 4}
 load("../data/cancer_disrupt_heatmap.Rda")
@@
 
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
+fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(x=sample_id, y=path_name, fill = membership)) + 
   geom_tile(size = 0.6) +
   scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
+  facet_grid(.~type, scales = "free", space = "free", switch = "x") +
 scale_y_discrete(position = "left") +
-  ylab("") +
-  xlab("377 Reactome Pathways") +
+  xlab("") +
+  ylab("377 Reactome Pathways") +
   theme(axis.text.x = element_blank(),
@@
         strip.text.y.left = element_text(angle = 0),
+        strip.text.x = element_text(angle = 90),
         strip.background = element_blank(),
@@
 
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
+ for(i in which(grepl("strip-b", fig1Ag$layout$name))){
    fig1Ag$grobs[[i]]$layout$clip <- "off"
@@
 
-##figure 1b
-load("../data/figure1B_data.rda")
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
+opar <- par(mfrow = c(3,3), mai = c(0.1, 0.1, 0.1, 0.1))
 
-permut_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
-  geom_line(aes(colour = "#998ec3")) +
-  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
-  scale_color_identity(guide = guide_legend(title = ""), 
-                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
-  scale_y_continuous(labels = scales::percent) +
-  xlab("MCA Eigenvalues") +
-  ylab("explained variance") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        legend.position = c(0.8, 0.8))
-
-pvalue_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = pvalue)) +
-  geom_line() +
-  geom_point() +
-  geom_vline(xintercept = optimal_mca_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_mca_dimensions)) +
-  xlab("MCA Eigenvalues") +
-  ylab("P-value") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-fig1B <- permut_p /
-  pvalue_p
-
-fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
-                aes(y=plot_x, x=plot_y)) +
-  geom_point(size = 0.25) +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  scale_color_viridis_c() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-load("../data/optimal_minpts_score.rda")
-fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
-  geom_point(data = sample_df, color = "grey", aes(y = value), size = 0.25) +
-  geom_point(size = 0.25) + 
-  geom_line() +
-  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
-  scale_y_log10() +
-  xlab("Minimum Cluster Size") +
-  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(y = plot_x, x = plot_y, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(y = plot_x, x = plot_y, color = as.factor(clust_knn))) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## mutation burden plot
-mut_order <- list() ## store data in list 
-for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
-  message(n)
-  mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
-    dplyr::select(sample_id, plot_x, plot_y, clust_knn, mut_count) %>% ## select relevant columns
-    arrange(mut_count) ## arrange mutation in increasing order
-}
-mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
-
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
-mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10))
-
-## Need the breaks positions for the x-axis
-previous_class <- 0 ## starting position 1
-position_of_previous_class <- 0 ## starting position 2
-xaxis_breaks <- list() ## list for for loop
-for(n in 1) { ## loop through 1 to 12 clusters
-  message(n)
-  class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
-  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-  previous_class <- class_spec ## storing class halfway mark
-  for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
-    message(n)
-    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
-    class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
-    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-    previous_class <- class_spec ## storing class halfway mark
-  }
-}
-xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
-
-total_n <- as.data.frame(table(mut_order$clust_knn)) ## getting total n in cluster
-
-mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count
-
-mut_median <- mut_order %>% 
-  group_by(clust_knn) %>% 
-  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
-
-## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
-mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column
-
-begin_NA <- data.frame(sample_id = paste0("sample", 1:100), 
-                       clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
-                       log10mut = NA) ## put NA values so they don't get plotted
-end_NA <- data.frame(sample_id = paste0("sample", 101:200), ## repeat the step for the ends
-                       clust_knn = max(as.numeric(mut_order$clust_knn)),
-                       log10mut = NA)
-mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
-
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id) ## factor samples
-mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10)) ## factor clusters
-
-fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = clust_knn)) + 
-  geom_point(size = 0.25) + 
-  ylab("Log10(Mutation Count)") + 
-  xlab("") +
-  theme_classic() +
-  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") +
-  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
-  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[1]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks.x = element_blank(),
-        axis.line.x = element_line(colour = "black", size = 0.5),
-        axis.line.x.top = element_line(colour = "black", size = 0.5),
-        panel.spacing = unit(1, "lines"),
-        strip.background = element_rect(size = 0.5),
-        strip.placement = "outside",
-        strip.text.x = element_text(size = 10),
-        legend.position = "none") +
-  scale_color_manual(values = kelly.colours[-c(1:2)])
-
-## code to keep only the bottom border line of the facet
-fig1G <- ggplotGrob(fig1G)
-lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
-                   gp=gpar(col="black", lwd=4))
-
-for (k in grep("strip-t",fig1G$layout$name)) {
-  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
-}
-
-fig1G <- as.ggplot(fig1G)
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(y = plot_x, x = plot_y, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(y = plot_x, x = plot_y, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "project_code")], by = "sample_id") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(y = plot_x, x = plot_y, colour = mut_type)) +
-  geom_point(data = brca, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(y = plot_x, x = plot_y, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(y = plot_x, x = plot_y)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-empty_plot <- ggplot() + theme_void()
-
-layout <- '
-AAAABBBCCCDDD
-AAAABBBCCCDDD
-AAAABBBEEEFFF
-AAAABBBEEEFFF
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAHHHIIIJJJ
-AAAAHHHIIIJJJ
-AAAAKKKLLL###
-AAAAKKKLLL###
-'
-
-fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')
-
+fig1A
 ```

### 2020-08-13 code/figure4.Rmd
Split figure 1 into 4 figures, update figure4.rmd with simon's fix, add page breaks

@@
   facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  ggtitle("Enriched")
+  ggtitle("Enriched") +
   theme(

### 2020-08-13 code/figure4v2.Rmd
Split figure 1 into 4 figures, update figure4.rmd with simon's fix, add page breaks

```{r figure-4-plot, fig.width = 9}
umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
ditto_color <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#666666",
  "#AD7700", "#1C91D4", "#007756", "#D5C711",
  "#005685", "#A04700", "#B14380", "#4D4D4D",
  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
  "#00446B", "#803800", "#8D3666", "#3D3D3D")
cluster.colours <- c("grey", kelly.colours[c(3:12)])
names(cluster.colours) <- c(0:10)

### I - Cancer Type
par(opar)
par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
I_colors <- c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')
names(I_colors) <- c("BRCA", "COAD", "HNSC", "PAAD")
with(umap_cancertype %>% 
       mutate(I_color = case_when(project_code %in% c("COAD", "BRCA", "HNSC", "PAAD") ~ project_code,
                                  TRUE ~ "Other")) %>%
       filter(I_color != "Other"),
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = I_colors[I_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
legend("top", inset = c(0), legend=rev(names(I_colors)), col=rev(I_colors), pch = 20, bty = "n", horiz = T, cex = 1.2)
title(xlab="UMAP 2", line=0)
### J - BRCA subtype
J_colors <- c(ditto_color[c(2,3,4,5)])#, 'grey')
names(J_colors) <- c("Basal", "Her2", "LumA", "LumB")#, "Other")
with(umap_cancertype %>% 
       mutate(J_color = case_when(project_code == "BRCA" ~ Subtype_Selected,
                                  TRUE ~ "Other")) %>%
       filter(J_color != "Other"),
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
legend("top", inset = c(0), legend=names(J_colors), col=J_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
title(xlab="UMAP 2", line=0)
### K - BRCA Somatic Mutations
load("../data/gdc_gene_exp_adj.Rda")
load("../data/gene_data.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv")
brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
  rownames_to_column(var = "ensembl_gene_id") %>%
  left_join(., brca, by = "ensembl_gene_id") %>% 
  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
                           BRCA1 == "0" ~ "NA"),
         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
                           BRCA2 == "0" ~ "NA")) %>%
  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "plot_z", "project_code")], by = "sample_id") ## leftjoin by sample id

K_colors <- c('#33a02c', '#e31a1c', '#ff7f00')#, "grey")
names(K_colors) <- c("BRCA1", "BRCA1\nBRCA2", "BRCA2")#, "Other")
with(brca %>% filter(mut_type != "NA"),
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = K_colors[mut_type], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
legend("top", inset = c(0), legend=names(K_colors), col=K_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
title(xlab="UMAP 2", line=0)
### L - Stage
L_colors <- c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')
names(L_colors) <- c("Stage I", "Stage II", "Stage III", "Stage IV")
with(umap_cancertype %>% filter(ajcc_pathologic_tumor_stage != "Not Available"),
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = L_colors[ajcc_pathologic_tumor_stage], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
legend("top", inset = c(0), legend=names(L_colors), col=L_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
title(xlab="UMAP 2", line=0)
### M - Metastasis
M_colors <- c('#a6cee3','#1f78b4')
names(M_colors) <- c("M0", "M1")
with(umap_cancertype %>% filter(ajcc_metastasis_pathologic_pm %in% c("M0", "M1")),
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = M_colors[ajcc_metastasis_pathologic_pm], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
legend("top", inset = c(0), legend=names(M_colors), col=M_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
title(xlab="UMAP 2", line=0)

plot.new()
```

### 2020-08-13 code/figure2.Rmd

@@
 
-```{r figure-2-plot, fig.align= 'center', fig.height= 7, fig.width = 7}
+```{r figure-5-plot, fig.align= 'center', fig.height= 7, fig.width = 7}
 #out.height='90%', out.width='90%'

### 2020-08-13 code/figure2v2.Rmd

--- 2020-08-11 code/figure1.Rmd
@@
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] %>% ## include only second level pathways
-  left_join(., main2first, by = c("PATH_ID" = "first_level")) %>% distinct() %>% 
-  left_join(., ensembl2reactome[,c(2,4)], by = c("main_level" = "V2")) %>% distinct()
-
-## Filtering pathways not relevant to our analysis by keywords
-filtered_ensembl2reactome <- ensembl2reactome %>% filter(!V4 %in% c("Disease", "Muscle contraction", "Reproduction")) ## Pathways in disease, muscle contraction and reproduction parent pathway were removed
-
-toMatch <- c("disease", "defect", "cancer", "oncogenic", "disorder", "infection", "mutant", "bacteria", "listeria", "loss") ## pathways with the following key terms were removed.
-
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% mutate(l_pathname = tolower(.$PATH_NAME)) %>% 
-  .[!grep(paste(toMatch,collapse="|"), .$l_pathname),]
-
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% .[which(!.$PATH_ID %in% c("R-HSA-212436", "R-HSA-388396")),] ## Removing Generic Transcription Pathway and GPCR downstream signalling as they are broad pathways
-
-gene2level2_lookup <- filtered_ensembl2reactome %>% dplyr::select(ENSEMBL, PATH_ID) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "../data/gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- filtered_ensembl2reactome %>% dplyr::select(PATH_ID, PATH_NAME) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "../data/pathname2level2.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene2_level2_lookup.Rda")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_sample_id, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
+```{r figure-2-data-generation, eval = FALSE}
 ####################
@@
 )
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "sample_id") ## move rownames to column
-
-umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id") 
-rm(mutation_count)
-
-####################
-## Figure 1H data ##
-####################
-umap_3d_coors <- left_join(umap_3d_coors, tumor_samples[,c("project_code", "submitted_sample_id")], by = c("sample_id" = "submitted_sample_id"))
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
 ```
 
-```{r fig-1a-1l, warning=FALSE, fig.height=12, fig.width=11, message = FALSE}
-#fig.height=15, fig.width=16,
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  ylab("") +
-  xlab("377 Reactome Pathways") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-##figure 1b
+```{r figure-2-plot, fig.width = 9}
 load("../data/figure1B_data.rda")
@@
   pull(eigenvalue)
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+ditto_color <- c(
+  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
+  "#0072B2", "#D55E00", "#CC79A7", "#666666",
+  "#AD7700", "#1C91D4", "#007756", "#D5C711",
+  "#005685", "#A04700", "#B14380", "#4D4D4D",
+  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
+  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
+  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
+  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
+  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
+  "#00446B", "#803800", "#8D3666", "#3D3D3D")
+cluster.colours <- c("grey", kelly.colours[c(3:12)])
+names(cluster.colours) <- c(0:10)
+load("../data/optimal_minpts_score.rda")
 
-permut_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
-  geom_line(aes(colour = "#998ec3")) +
-  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
-  scale_color_identity(guide = guide_legend(title = ""), 
-                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
-  scale_y_continuous(labels = scales::percent) +
-  xlab("MCA Eigenvalues") +
-  ylab("explained variance") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        legend.position = c(0.8, 0.8))
+opar <- par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(4,6,2,0) + 0.1)
+with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = expl_var / 100, 
+                                                 type = "l", yaxt = "n", bty = "l",
+                                                 col = "#998ec3", 
+                                                 xlab = "", ylab = ""))
+with(randomize %>% filter(eigenvalue < 50), lines(x = eigenvalue, y = perm / 100,
+                                                  col = "#f1a340"))
+yticks_val <- scales::breaks_pretty(n=5)(randomize %>% filter(eigenvalue < 50) %>% pull(expl_var) / 100)
+axis(2, at=yticks_val, lab=scales::percent(yticks_val), las = 1)
+title(ylab = "explained variance", line = 3.5)
+title(xlab = "MCA Eigenvalues", line = 2)
+legend("topright", legend=c("Eigenvalues", "Permuted Eigenvalues"),
+       col=c("#998ec3", "#f1a340"), lty=1, cex=0.8)
+### D
+par(mar = c(2,2,0,0), bg = NA)
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "white", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+### E
+par(mar = c(4,6,2,0) + 0.1)
+with(sample_df, plot(x = minPts, y = value, 
+                     col = "grey", pch = 20, yaxt = "n",
+                     xlab = "", ylab = "", bty = "l",
+                     las = 1, cex = 0.3))
+with(score_df, lines(x = minPts, y = score,
+                     col = "black", las = 1))
+abline(v = optimal_minPts, col = "red")
+title(ylab = "score", line = 3.5)
+title(xlab = "Minimum Cluster Size", line = 2)
+axis(2, at=c(0.25, 0.35, 0.45, 0.55), las = 1)
 
-pvalue_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = pvalue)) +
-  geom_line() +
-  geom_point() +
-  geom_vline(xintercept = optimal_mca_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_mca_dimensions)) +
-  xlab("MCA Eigenvalues") +
-  ylab("P-value") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
+### C
+with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = pvalue, 
+                                                 type = "l", bty = "l",
+                                                 col = "black", las = 1,
+                                                 xlab = "", ylab = ""))
+with(randomize %>% filter(eigenvalue < 50), points(x = eigenvalue, y = pvalue, pch = 20,
+                                                   col = "black"))
+abline(v = optimal_mca_dimensions, col = "red")
+title(main = paste0("Significant dimensions = ", optimal_mca_dimensions), adj = 0, font = 3, cex.main = .8)
+title(ylab = "P-value", line = 3.5)
+title(xlab = "MCA Eigenvalues", line = 2)
+### F
+par(mar = c(2,2,0,0), bg = NA)
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = cluster.colours[as.character(clust)], pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+### G
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
 
-fig1B <- permut_p /
-  pvalue_p
-
-fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
-                aes(y=plot_x, x=plot_y)) +
-  geom_point(size = 0.25) +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  scale_color_viridis_c() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-load("../data/optimal_minpts_score.rda")
-fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
-  geom_point(data = sample_df, color = "grey", aes(y = value), size = 0.25) +
-  geom_point(size = 0.25) + 
-  geom_line() +
-  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
-  scale_y_log10() +
-  xlab("Minimum Cluster Size") +
-  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(y = plot_x, x = plot_y, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(y = plot_x, x = plot_y, color = as.factor(clust_knn))) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## mutation burden plot
-mut_order <- list() ## store data in list 
-for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
-  message(n)
-  mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
-    dplyr::select(sample_id, plot_x, plot_y, clust_knn, mut_count) %>% ## select relevant columns
-    arrange(mut_count) ## arrange mutation in increasing order
-}
-mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
-
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
-mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10))
-
-## Need the breaks positions for the x-axis
-previous_class <- 0 ## starting position 1
-position_of_previous_class <- 0 ## starting position 2
-xaxis_breaks <- list() ## list for for loop
-for(n in 1) { ## loop through 1 to 12 clusters
-  message(n)
-  class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
-  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-  previous_class <- class_spec ## storing class halfway mark
-  for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
-    message(n)
-    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
-    class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
-    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-    previous_class <- class_spec ## storing class halfway mark
-  }
-}
-xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
-
-total_n <- as.data.frame(table(mut_order$clust_knn)) ## getting total n in cluster
-
-mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count
-
-mut_median <- mut_order %>% 
-  group_by(clust_knn) %>% 
-  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
-
-## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
-mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column
-
-begin_NA <- data.frame(sample_id = paste0("sample", 1:100), 
-                       clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
-                       log10mut = NA) ## put NA values so they don't get plotted
-end_NA <- data.frame(sample_id = paste0("sample", 101:200), ## repeat the step for the ends
-                       clust_knn = max(as.numeric(mut_order$clust_knn)),
-                       log10mut = NA)
-mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
-
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id) ## factor samples
-mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10)) ## factor clusters
-
-fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = clust_knn)) + 
-  geom_point(size = 0.25) + 
-  ylab("Log10(Mutation Count)") + 
-  xlab("") +
-  theme_classic() +
-  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") +
-  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
-  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[1]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks.x = element_blank(),
-        axis.line.x = element_line(colour = "black", size = 0.5),
-        axis.line.x.top = element_line(colour = "black", size = 0.5),
-        panel.spacing = unit(1, "lines"),
-        strip.background = element_rect(size = 0.5),
-        strip.placement = "outside",
-        strip.text.x = element_text(size = 10),
-        legend.position = "none") +
-  scale_color_manual(values = kelly.colours[-c(1:2)])
-
-## code to keep only the bottom border line of the facet
-fig1G <- ggplotGrob(fig1G)
-lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
-                   gp=gpar(col="black", lwd=4))
-
-for (k in grep("strip-t",fig1G$layout$name)) {
-  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
-}
-
-fig1G <- as.ggplot(fig1G)
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(y = plot_x, x = plot_y, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(y = plot_x, x = plot_y, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "project_code")], by = "sample_id") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(y = plot_x, x = plot_y, colour = mut_type)) +
-  geom_point(data = brca, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(y = plot_x, x = plot_y, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(y = plot_x, x = plot_y)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-empty_plot <- ggplot() + theme_void()
-
-layout <- '
-AAAABBBCCCDDD
-AAAABBBCCCDDD
-AAAABBBEEEFFF
-AAAABBBEEEFFF
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAHHHIIIJJJ
-AAAAHHHIIIJJJ
-AAAAKKKLLL###
-AAAAKKKLLL###
-'
-
-fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')
-
+plot.new()
 ```

### 2020-08-13 code/figure3v2.Rmd

--- 2020-08-11 code/figure1.Rmd
@@
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
-
-## Preparing sample metadata ##
-
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
-
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
-  }
-
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
-
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
-
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
-
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] %>% ## include only second level pathways
-  left_join(., main2first, by = c("PATH_ID" = "first_level")) %>% distinct() %>% 
-  left_join(., ensembl2reactome[,c(2,4)], by = c("main_level" = "V2")) %>% distinct()
-
-## Filtering pathways not relevant to our analysis by keywords
-filtered_ensembl2reactome <- ensembl2reactome %>% filter(!V4 %in% c("Disease", "Muscle contraction", "Reproduction")) ## Pathways in disease, muscle contraction and reproduction parent pathway were removed
-
-toMatch <- c("disease", "defect", "cancer", "oncogenic", "disorder", "infection", "mutant", "bacteria", "listeria", "loss") ## pathways with the following key terms were removed.
-
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% mutate(l_pathname = tolower(.$PATH_NAME)) %>% 
-  .[!grep(paste(toMatch,collapse="|"), .$l_pathname),]
-
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% .[which(!.$PATH_ID %in% c("R-HSA-212436", "R-HSA-388396")),] ## Removing Generic Transcription Pathway and GPCR downstream signalling as they are broad pathways
-
-gene2level2_lookup <- filtered_ensembl2reactome %>% dplyr::select(ENSEMBL, PATH_ID) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "../data/gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- filtered_ensembl2reactome %>% dplyr::select(PATH_ID, PATH_NAME) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "../data/pathname2level2.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene2_level2_lookup.Rda")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_sample_id, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1B data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>%
-  t() %>%
-  as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(tumor_samples$submitted_sample_id, .$sample), ]
-
-gdc_path_mca <- left_join(gdc_path, tumor_samples[, c("project_code", "submitted_sample_id")],
-  by = c("sample" = "submitted_sample_id")) %>%
-  column_to_rownames(var = "sample")
-
-gdc_path_sample_names <- rownames(gdc_path_mca) ## Getting sample id order
-gdc_path_tumor_types <- gdc_path_mca$project_code ## Getting cancer type order
-
-gdc_path_mca$project_code <- NULL ## Remove project_code column from df
-
-## create a categorical variable from the boolean variable
-pathway_list <- make_clean_names(colnames(gdc_path_mca))
-pathway_count <- length(pathway_list)
-gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
-  factor(x, levels = c(0, 1), labels = c("no", "yes"))
-}))
-colnames(gdc_path_mca) <- pathway_list
-gdc_path_mca$project_code <- gdc_path_tumor_types
-rownames(gdc_path_mca) <- gdc_path_sample_names
-
-## perform mca
-mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
-
-expl_var <- mca_res$eig[, 2]
-expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
-
-set.seed(4242)
-for (k in seq(100)) {
-  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
-  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
-  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
-}
-
-randomize <- tibble(
-  eigenvalue = 1:length(expl_var),
-  expl_var = expl_var,
-  perm = colMeans(expl_var_perm),
-  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
-)
-
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-mca_res <- MCA(
-  X = gdc_path_mca[, pathway_list],
-  ncp = optimal_mca_dimensions, graph = FALSE
-)
-
-save(randomize, mca_res, file = "figure1B_data.rda")
-
-####################
-## Figure 1C data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-rng <- RNGseq(max_n_pt * n_iter, 42)
-r <- rng[[3649]]
-
-rngtools::setRNG(r)
-gdc.umap <- umap(mca_res$ind$coord, 
-                 n_components = 3,
-                 n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
-                 n_epochs = 1000,
-                 n_threads = 1)
-gdc_umap_coors <- as.data.frame(gdc.umap) %>%
-  rename(plot_x = V1, plot_y = V2, plot_z = V3)
-rownames(gdc_umap_coors) <- rownames(gdc_path_mca)
-
-####################
-## Figure 1D data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
-
-n_tumors <- nrow(mca_res$ind$coord)
-
-cl <- makeCluster(8)
-registerDoParallel(cl)
-
-for (i in 3:max_n_pt) {
-  print(paste0("MIN POINTS = ", i))
-  score_iter <- foreach(1:n_iter,
-    .combine = c,
-    .export = c("mca_res"),
-    .packages = c("dbscan", "uwot"),
-    r = rng[(i - 1) * n_iter + 1:n_iter]
-  ) %dopar% {
-    rngtools::setRNG(r)
-    umap_init_mca <- umap(mca_res$ind$coord, 
-                          min_dist = 1e-6, 
-                          n_components = 3, 
-                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
-                          n_epochs = 1000, 
-                          n_threads = 1)
-    res <- hdbscan(umap_init_mca, minPts = i)
-    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
-    return(score_iter_temp)
-  }
-  score_mca[i - 2, ] <- score_iter
-}
-
-score <- rowMeans(score_mca)
-score_df <- tibble(minPts = 3:max_n_pt, score = score)
-
-sample_df <- score_mca %>%
-  as.data.frame() %>%
-  as_tibble() %>%
-  add_column(minPts = 3:max_n_pt, .before = 1) %>%
-  pivot_longer(-minPts) %>%
-  group_by(minPts)
-
-optimal_minPts <- score_df %>%
-  filter(score == min(score)) %>%
-  pull(minPts)
-
-save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
-
-####################
-## Figure 1E data ##
-####################
-rngtools::setRNG(r)
-
-cluster_umap <- umap(mca_res$ind$coord,
-  min_dist = 1e-6,
-  n_components = 3,
-  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
-  n_epochs = 1000,
-  n_threads = 1
-)
-colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
-hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
-umap_3d_coors <- gdc_umap_coors %>%
-  bind_cols(as.data.frame(cluster_umap)) %>%
-  mutate(
-    n_pathways_mut = rowSums(gdc_path[, -1]),
-    clust = as.factor(hdb_cluster_assignment$cluster),
-    member_prob = hdb_cluster_assignment$membership_prob,
-    sample_id = rownames(gdc_umap_coors)
-  )
-## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_3d_coors$clust_knn <- umap_3d_coors$clust
-umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
-  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
-  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
-  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
-)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "sample_id") ## move rownames to column
-
-umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id") 
-rm(mutation_count)
-
-####################
-## Figure 1H data ##
-####################
-umap_3d_coors <- left_join(umap_3d_coors, tumor_samples[,c("project_code", "submitted_sample_id")], by = c("sample_id" = "submitted_sample_id"))
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=12, fig.width=11, message = FALSE}
-#fig.height=15, fig.width=16,
-load("../data/cancer_disrupt_heatmap.Rda")
+```{r figure-3-plot, message=FALSE, warning=FALSE, fig.width= 9, fig.height= 3}
 umap_cancertype <- fread("../data/umap_3d_coors.tsv")
@@
                    "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  ylab("") +
-  xlab("377 Reactome Pathways") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-##figure 1b
-load("../data/figure1B_data.rda")
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-permut_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
-  geom_line(aes(colour = "#998ec3")) +
-  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
-  scale_color_identity(guide = guide_legend(title = ""), 
-                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
-  scale_y_continuous(labels = scales::percent) +
-  xlab("MCA Eigenvalues") +
-  ylab("explained variance") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        legend.position = c(0.8, 0.8))
-
-pvalue_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = pvalue)) +
-  geom_line() +
-  geom_point() +
-  geom_vline(xintercept = optimal_mca_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_mca_dimensions)) +
-  xlab("MCA Eigenvalues") +
-  ylab("P-value") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-fig1B <- permut_p /
-  pvalue_p
-
-fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
-                aes(y=plot_x, x=plot_y)) +
-  geom_point(size = 0.25) +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  scale_color_viridis_c() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-load("../data/optimal_minpts_score.rda")
-fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
-  geom_point(data = sample_df, color = "grey", aes(y = value), size = 0.25) +
-  geom_point(size = 0.25) + 
-  geom_line() +
-  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
-  scale_y_log10() +
-  xlab("Minimum Cluster Size") +
-  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(y = plot_x, x = plot_y, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(y = plot_x, x = plot_y, color = as.factor(clust_knn))) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
 ## mutation burden plot
@@
 fig1G <- as.ggplot(fig1G)
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(y = plot_x, x = plot_y, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(y = plot_x, x = plot_y, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "project_code")], by = "sample_id") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(y = plot_x, x = plot_y, colour = mut_type)) +
-  geom_point(data = brca, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(y = plot_x, x = plot_y, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(y = plot_x, x = plot_y)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-empty_plot <- ggplot() + theme_void()
-
-layout <- '
-AAAABBBCCCDDD
-AAAABBBCCCDDD
-AAAABBBEEEFFF
-AAAABBBEEEFFF
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAHHHIIIJJJ
-AAAAHHHIIIJJJ
-AAAAKKKLLL###
-AAAAKKKLLL###
-'
-
-fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')
-
+fig1G
 ```

### 2020-08-13 code/supplements.Rmd

@@
+\newpage
 ### Supplementary Figure S1
@@
 ```
-\newpage
+\pagebreak
 ### Supplementary Figure S2
@@
 ```
-\newpage
+\pagebreak
 ### Supplementary Figure S3
@@
 ```
-\newpage
+\pagebreak
 ### Supplementary Figure S4
@@
 ```
-\newpage
+\pagebreak
 ### Supplementary Figure S5
@@
 ```
-\newpage
+\pagebreak
 ### Supplementary Table 1

### 2020-08-13 text/fig_legend.Rmd

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
Classification of tumors by pathway disruptions.** a) Each of 377 selected
Reactome pathways (columns) is classified as disrupted if one or more genes is
mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
standardized abbreviations from the TCGA project. b) MCA analysis was performed to 
determine the eigenvalue with the most explanatory power from the categorical data.
Average of 100 permutations of the data was used as a baseline. c) The p-value of
each eigenvalue was plotted and the max eigenvalue with a pvalue < 0.05 was selected. 
d) The resulting MCA was projected down to 3 dimensions for the purpose of density 
based clustering (only the first 2 dimensions are shown). Each dot represents a tumor 
e) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize
the fraction of unassigned tumors. f) HDBSCAN clustered tumors were visualized on the UMAP
projection; tumors are colored by cluster membership (see legend). This results
in some tumors remaining unclassified (grey points). g) kNN was used to assign
membership to the remaining tumors. h) Mutational burden plot where each dot corresponds
to the sample mutation counts for tumors in each class. i) Projection of cancer 
type identity onto UMAP. Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma,
HNSC = Head and neck squamous cell, PAAD = Pancreatic adenocarcinoma. j) Projection of breast
cancer subtypes onto the UMAP. k) Projection of BRCA1/2 somatic mutation onto the
UMAP. l) Projection of tumor stage onto the UMAP, regardless of cancer type. m) 
Projection of metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, 
M1 = metastatic tumors.</p>

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 2: Pan-cancer 
enrichment of pathway disruptions.** a) Waterfall plot; columns represent tumors. 
Each row represents a pathway. The pathways are ranked such that the pathway with 
the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are
ordered to prioritize the top pathway disruptions successively. The top 100 most 
mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies
not disrupted. b) Heatmap shows the relative enrichment of each pathway (columns) 
within each numbered cluster (rows). Effect size is displayed as colors (orange, 
enrichment; purple, depletion). c) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size >= 0.30 in each cluster)
within each numbered cluster (rows).</p>

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3: Gene level 
analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses 
a logarithmic axis to represent odds ratio with a 95% confidence interval. Row 
represents the significant genes from each cluster. Each cluster was compared 
against the background (all other clusters) to find deferentially mutated genes. 
Significant genes (pvalue < 0.01) were selected and limited to the top ten results 
for each clusters. Heatmap; Column represents the cancer type. Row represents the 
significant genes as described above. The heatmap shows the proportion of samples 
per cluster, per cancer type that is mutated for each of the significant genes. 
Depleted significant genes (left). Enriched significant genes (right).</p>

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1: Cluster-specific 
enriched pathways (effect size >= 0.30) in metastatsis.**

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary figure S1. 
Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
in **Figure 1D of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S2. 
Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
cancers projected onto the tumors of the UMAP from **Figure 1D of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S3.
Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15  
genes involved in genome stability projected on the UMAP from **Figure 1D of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S4. 
Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
pathways projected on the UMAP from **Figure 1D of the main text** ordered by the 
hierarchical pathway order (column) from **Figure 2B of the main text**.

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S5.
Distribution of oncogenes on UMAP clusters.** The left panel shows the identity of 
clusters 1, 7 and 8. The multipanel on the right shows one of the three oncogenes 
projected onto the tumors of the UMAP from **Figure 1D of the main text**.

### 2020-08-13 text/manuscript.Rmd

@@
   library(poibin),
-  library(flextable)
+  library(flextable),
+  library(plot3D)
 ))
@@
 
-<strike>
-### Gene-Pathway redundancy contributes to cluster identity
-
-Since pathways share some common driver genes, we investigated whether
-tumors in each class  have multiple mutations across signaling pathways, or
-whether a small number of driver genes account for enrichment. The split between
-clusters 7 and 8 is largely driven by mutations in _PI3K_ and its
-homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
-**supplementary figure S5**). Furthermore, cluster 1 is distinguished
-from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
-_TP53_ panel of **Supplementary Figure S5**).
-
-Cluster 2 exhibited the highest pathway enrichment scores of the ten major
-classes, although there is a significant number of tumors spatially distributed
-between cluster groupings 7-9 and 2 on the UMAP projection that have pathway
-characteristics of both. The top ten distinguishing pathways of class 2 tumors
-were "cellular response to hypoxia," "regulation of mitotic cell cycle,"
-"assembly of the pre-replicative complex," "metabolism of polyamines,"
-"degradation of $\beta$-catenin by the destruction complex", "tumor necrosis
-factor receptor 2 non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0)
-binds and destabilizes RNA", "regulation of apoptosis" and 3 other pathways
-associated with aspects of hedgehog signaling. To determine once again whether
-the genes driving pathway enrichment were commonly shared among these pathways
-we plotted and performed unsupervised clustering on tumors with an individual
-mutation matrix using every gene in each of the ten pathways, grouped by shared
-pathway membership (**figure 3A**). Surprisingly, this class of tumors was not
-predominated by 1 or more highly mutated genes across all cancers. Instead,
-genes in the proteosomal pathway whose activity contributes to all the
-aforementioned pathways and the majority of tumors in these classes have a
-single mutation in one of these genes. We annotated the tumors in this class
-according to tissue of origin to determine whether proteasome mutations have a
-tissue-of-origin (or cancer-specific) bias, and indeed we could observe some
-clustering. These data could explain how tissue-specific gene mutation
-signatures might result in general cancer phenotypes by converging at the
-pathway level.
-
-The proteasome plays key roles in the transduction of signals from the p53, WNT
-and NFkB pathways and it and proteasomal inhibitors have been approved for
-clinical use in some patient populations [@proteasome_review_Jang]. To determine
-whether this class of proteins is broadly sensitive to cancer-promoting
-mutations, we tabulated each gene as prognostic or not in the human protein
-expression atlas [@protein_atlas] and found 49 positive and 10 negative
-associations ( _i.e._ higher expression associated with unfavorable outcomes is
-"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 3B**).
-Second we looked at survival in mutations (**figure 3C**).
-Only a handful of proteasome mutations were significantly associated with favorable or unfavorable outcomes, probably owing to lack of sufficient power.
-
-Third, we checked the mutation classes using MAFtools [@maftools] and found the overwhelming
-majority (x out of y) are missense _vs._ various other deleterious classes (see
-**figure 3D**), suggesting that, similar to other oncogenic driver classes such
-as KRAS and PIK3C, proteasomal subunit mutations may result in specific
-dysregulation of the proteasomal complex activities that target cancer-promoting
-pathways. Taken together, our data suggest that mutations in proteasomal
-subunits are more widespread in cancer than previously appreciated and may
-represent a targetable category of tumors.
-</strike>
-
 ### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
@@
 nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
-Next, we investigated the proportion of samples per cancer type for the significant genes
-within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors were not 
-predominated by one or more highly mutated genes across all cancers. Instead, several 
-cancer-specific tumors in these classes had mutation profiles that differ from one another 
-( _e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC was enriched for _UBC_, and 
-PAAD was enriched for _PSMA1_). Taken together, our data identifies a framework cancer 
-type specific mutation associated with a specific cluster.
+We also observed genes that were enriched for a cluster are depleted from others ( _e.g._ 
+_TP53_ is enriched in cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster
+7, but depleted in clusters 3 and 8). Next, we investigated the proportion of samples per 
+cancer type for the significant genes within a cluster (**Figure 4**; heatmap). Surprisingly, 
+cluster-specific tumors were not predominated by one or more highly mutated genes across all 
+cancers. Instead, when observing the mutation rate for these genes within samples that belong
+to a cluster, the mutation rate is heterogeneous across tumors by tissue origin ( _e.g._ In 
+cluster 4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched
+for _HIST1H2AC_). Even among the top most mutationally enriched genes within clusters there is
+no global pattern, indicating that our classes are not driven by individual genes, but rather 
+networks as a whole. Taken together, our data identifies a framework of cancer type-specific 
+mutation associated with a specific cluster.
 
@@
 ```
-\newpage
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
 ```{r child='../text/fig_legend.Rmd'}
@@
 
-\newpage
-```{r child='../code/figure1.Rmd'}
+```{r, results='asis'}
+cat("\\newpage")
 ```
-\newpage
-```{r child='../code/figure2.Rmd'}
+#### Figure 1
+```{r child='../code/figure1v2.Rmd'}
 ```
-\newpage
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 2
+```{r child='../code/figure2v2.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 3
+```{r child = '../code/figure3v2.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 4
+```{r child = '../code/figure4v2.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 5
+```{r child = '../code/figure2.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 6
 ```{r child = '../code/figure4.Rmd'}
 ```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Table 1
+```{r child = '../code/table2.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+![](../figures/fig6B.png)
 \newpage
-```{r child='../code/table2.Rmd'}
-```
-\newpage
-
-![](../figures/fig6B.png)
-\pagebreak
 ![](../figures/fig6C.png)
@@
 
-```{r child='../code/supplements.Rmd'}
+```{r }
+#child='../code/supplements.Rmd'
 ```

### 2020-08-13 text/survival-methods.Rmd
Add all the survival code, including kmplots for simulated data, and modify some of the methods section on survival
Create a single large panel figure for the survival stuff, rewrite the text to deal with that appropriately. Also adjust the Makefile to build the script properly now that we're using .docx
Reorganize some of the survival code.

@@
 
-$$\frac{dP}{dT} = R(T)(1-P(T))$$
+$$\frac{dP}{dT} = (1-P(T))R(T)$$
 
@@
 
-The $k$ in our model can be taken as, roughly, the aging rate relative to the
-baseline rates as depicted in **supplementary figure S5**. Thus, a value of
-$k=2$ means that a given cancer causes you to experience the same risk in time
-$dt$ as a non-cancer patient would experience at the same age, in time $2dt$.
+The $k$ in our model can be taken as, roughly, the aging rate relative
+to the baseline rates. Thus, a value of $k=2$ means that a given
+cancer causes you to experience the same risk in time $dt$ as a
+non-cancer patient would experience at the same age, in time $2dt$.
 
@@
 race is an important category due to differences in access to care and other
-socioeconomic and education related factors.
+socioeconomic and education related factors. However, due to the low number of
+Black patient tumors, we use ethnicity only to identify the baseline risk
+function, while the tissue and class specific multipliers are not ethnicity
+specific.
 
 Using the model described above we fit the data using Stan via the rstan package
-in R. Posterior inferences are shown in **FIXME WHICH FIGURE?**.
+in R. Posterior inferences are shown in **Figure 7**.
 
-**supplementary figure S5** gives a comparison on log-scale between the post-fit
-survival curves and the CDC data it is based on. For demonstration purposes we
-also illustrate some examples of the expected survival for a population that
-receives a diagnosis with doubled risk of death at ages 40, 50, 60, 70, and 80
-(**supplementary figure S6**).
+**supplementary figure S5** gives a comparison on log-scale between
+the post-fit survival curves and the CDC data it is based on.
 

### 2020-08-13 text/survival.Rmd
Create a single large panel figure for the survival stuff, rewrite the text to deal with that appropriately. Also adjust the Makefile to build the script properly now that we're using .docx
Reorganize some of the survival code.

--- 2020-07-27 text/survival.Rmd
@@
 independently, resulting in a cancer and class specific estimate of the aging
-rate multiplier. We found that cancer types, as expected, have a range of
-prognoses relative to the general population, as shown in **figure 5a**. The
-least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
-(GBM) which has a risk multiplier of between 15 and 22 relative to the
-background risk of death in the population. This undoubtedly is due to the fact
-that relatively young patients are affected by GBM. Among the deadliest cancers
-outside of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD)
-cancers. Apart from these top 4 cancers which ranged from 6-22 in relative risk
-of age-adjusted death, the remaining cancer types ranged from about 1 to 5 in
-magnitude (**figure 5a**). A rate of 1 represents patients who die relatively
+rate multiplier.
+
+To understand the model, see **Figure 7A** where we show simulated survival time
+curves for four related groups. We start with a cohort of women aged 30 and show
+their expected survival (blue). By setting their cancer rate multiplier to k=2
+we can simulate the effect of a moderately deadly cancer (red). Compare this to
+a randomly selected group of women ranging in age from 30 to 70 years old (dark
+blue). Immediately the survival curve changes due to the mixture of ages,
+without any malignancy. Adding a k=2 malignancy further reduces expected
+lifespan (dark red). We can see that for different tissues simply due to the
+change in distribution of age-at-diagnosis we should expect equally deadly
+cancers to have *different* survival curves. The model takes this into account
+and the aggressiveness of the cancer can be estimated without the confounding
+aspect of age.
+
+We found that cancer types, as expected, have a range of prognoses relative to
+the general population. For example in **Figure 7B-E** we see three particularly
+deadly cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one
+cancer where diagnosis apparently decreases risk relative to background
+(Thyroid: THCA). Cancers with posterior probability for relative risk of less
+than 1 such as THCA may be those where diagnosis is associated with patients who
+have relatively healthy behaviors, are seen regularly by doctors, and comply
+with health recommendations.
+
+The degree to which a given cancer accelerates aging can be determined by the
+ktis multiplier (**Figure 7F**) multiplied by the class specific modifier
+(**Figure 7G**). Looking at the tissue specific multiplier, the least deadly was
+prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
+risk multiplier of between 15 and 22 relative to the background risk of death in
+the population. This undoubtedly is due to the fact that relatively young
+patients are affected by GBM and that it is extremely deadly. Among the
+deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
+pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
+in relative risk of age-adjusted death, the remaining cancer types ranged from
+about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
 close to the age predicted purely from their demographics independent of the
@@
 
-![](../figures/fig6B.png)
-![](../figures/fig6C.png)
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters.
-a) Posterior distributions of estimated relative age-dependent survival by cancer type.
-b) Posterior distributions of estimated class-dependent survival for each cancer type.
-See methods for model details.</p>
-
-Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
-reflecting that some classes are either less deadly or more deadly than other
-classes within each cancer types. A class specific rate of 1 represents typical
-rate for this tissue type (**figure 5b**). For several cancers ( _e.g._ PRAD,
-kidney chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer
-(THCA)) the posterior estimates are largely indestinguishable from the prior,
-reflecting that either there were too few mortalities in the data to make an
-estimate (as expected for PRAD and THCA) or two few samples period. We did not
-observe class-specific trends that held true across cancer types, which could
-result from different cancers having different standards of care for example. In
-support of this interpretation, we also tested a factored model which considered
-both cancer type and each class independent of cancer type (**supplementary
-figure S7??**). Though we were able to successfully fit this model, it is a
-special case of the more general model where class and tissue can independently
-affect longevity, and there is no reason to believe that class specific effects
-would necessarily remain constant across tissue types given how widely the
-patients vary across tissue types as well as the different drug and surgical
-treatments typical used for each tissue.
-
-```{r survival-model, eval=FALSE}
-
-```
-
-![**Figure 6B**](../figures/fig6B.png)
-
-![**Figure 6C**](../figures/fig6C.png)
+Our estimates of tissue-specific class effects, in contrast, ranged from less
+than 1 up to about 3 or 4, reflecting that some classes are either less deadly
+or more deadly than other classes within each cancer types (**Figure 7G**). A
+class specific rate of 1 represents typical rate for this tissue type. For
+several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
+lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
+indestinguishable from the prior, reflecting that either there were too few
+mortalities in the data to make an estimate (as expected for PRAD and THCA) or
+two few samples period. We did not observe class-specific trends that held true
+across cancer types, which could result from different cancers having different
+standards of care for example. In support of this interpretation, we also tested
+a factored model which considered both cancer type and each class independent of
+cancer type (not shown). Though we were able to successfully fit this model, it
+is a special case of the more general model where class and tissue can
+independently affect longevity, and there is no reason to believe that class
+specific effects would necessarily remain constant across tissue types given how
+widely the patients vary across tissue types as well as the different drug and
+surgical treatments typical used for each tissue.

### 2020-08-13 code/survival/kmcurve.R

@@ -39,7 +39,6 @@ lccdf = function(r20,agerate,age,k,t){
         -(365*exp(age*agerate-20*agerate))/(agerate*k)));
 }
 
-curve(lccdf(r20wf,ageratewf,30,1,x)-log(runif(1)),0,365*100)
 
 simdeath = function(r20,agerate,age,k){
     r = runif(1)
@@ -48,19 +47,6 @@ simdeath = function(r20,agerate,age,k){
     return(tdays$root) ## years since diagnosis at "age"
 }
 
-## create example data for a fixed 30 year old without vs with a k=2
-## cancer
-
-
-patientsurvivalsbkd = replicate(100,simdeath(r20wf,ageratewf,30,1))/365
-patientsurvivalsk2 = replicate(100,simdeath(r20wf,ageratewf,30,2))/365
-
-
-fit1 <- survfit(Surv(patientsurvivalsbkd) ~ 1) 
-fit2 <- survfit(Surv(patientsurvivalsk2) ~ 1) 
-p1 = autoplot(fit1) + coord_cartesian(xlim=c(0,80)) + labs(x="Survival Time (years past 30th birthday)",y="Probability",title="Survival For 30 year old White Females (baseline)")
-p2 = autoplot(fit2) + coord_cartesian(xlim=c(0,80)) + labs(x="Survival Time (years past 30th birthday)",y="Probability",title="Survival For 30 year old White Females with k = 2 cancer") 
-grid.arrange(p1,p2)
 
 
 ## do a survival plot for the age distribution of each cancer using
@@ -87,22 +73,31 @@ cancerdat = cancerdat[!is.na(cancerdat$finaltime),]
 plotkmcomp = function(sim,ref,act,actstatus){
     ggplot(data.frame(stime=sim,rtime=ref,acttime=act,status=rep(1,length(ref)),actstatus=actstatus)) +
         geom_km(aes(time=stime,status=status),color="red") +
-        geom_km(aes(time=rtime,status=status),color="orange") +
-        geom_km(aes(time=acttime,status=actstatus),color="blue")
+        geom_km(aes(time=rtime,status=status),color="blue") +
+        geom_km(aes(time=acttime,status=actstatus),color="darkgreen")
 }
     
 
 
 plotcancertype = function(cancerdat,i){
+
+    lev = which(levels(cancerdat$type) == i)[[1]]
+    
+    
[…]

### 2020-08-14 code/supplemental_v2.Rmd
Update figures with 3d umap, fig ref in text, fig legend text

--- 2020-08-08 code/supplements.Rmd
@@
-### Supplementary Figure S1
+---
+output:
+  bookdown::pdf_book:
+    number_sections: FALSE
+    toc: FALSE
+    keep_tex: yes
+always_allow_html: TRUE
+editor_options:
+  chunk_output_type: console
+---
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE, message = FALSE)
+suppressPackageStartupMessages(c(
+  library(tidyverse),
+  library(data.table),
+  library(plot3D),
+  library(flextable)
+))
+```
 
-```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
-#out.width = '90%',
+```{r supplementary-figure-s1, fig.width = 9}
 umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
 
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+cancer_type <- unique(umap_cancertype$project_code) ## Get cancer type in data set
+cancer_type <- cancer_type[order(cancer_type)] ## Order alphabetically
 
-ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = project_code)) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 23)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in cancer_type[1:6]) {
+with(umap_cancertype %>% 
+       filter(project_code == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
+par(mar = c(2,2,0,0))
+for(n in cancer_type[7:12]) {
+with(umap_cancertype %>%
+       filter(project_code == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in cancer_type[13:15]) {
+with(umap_cancertype %>%
+       filter(project_code == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in cancer_type[19:23]) {
+with(umap_cancertype %>%
+       filter(project_code == n),
+     scatter3D(plot_y, -plot_z, plot_x,
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+
 ```
-![**Supplementary figure S1.
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]
+**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
+individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
+in **Figure 2C of the main text**.
 
-### Supplementary Figure S2
+\newpage
 
-```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
-umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+```{r supplementary-figure-s2-data-generation, fig.width = 9}
+ditto_color <- c(
+  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
+  "#0072B2", "#D55E00", "#CC79A7", "#666666",
+  "#AD7700", "#1C91D4", "#007756", "#D5C711",
+  "#005685", "#A04700", "#B14380", "#4D4D4D",
+  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
+  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
+  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
+  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
+  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
+  "#00446B", "#803800", "#8D3666", "#3D3D3D")
 
-ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
+cancer_w_subtypes <- list()
+for(n in cancer_type) {
+  spec_cancer <- umap_cancertype %>% 
+    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
+                               TRUE ~ "Other")) %>%
+    filter(J_color != "Other")
+  if(nrow(spec_cancer) == 0) {next
+  }
+  cancer_w_subtypes[[n]] <- n
+}
+cancer_w_subtypes <- names(cancer_w_subtypes)
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+
+for(n in cancer_w_subtypes[1:6]) {
+  spec_cancer <- umap_cancertype %>% 
+    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
+                               TRUE ~ "Other")) %>%
+    filter(J_color != "Other")
+  if(nrow(spec_cancer) == 0) {next
+    }
+  subtypes <- unique(spec_cancer$J_color)
+  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
+  names(J_colors) <- subtypes
+
+  with(spec_cancer,
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+  legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  title(xlab="UMAP 2", line=0)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+
+for(n in cancer_w_subtypes[7:12]) {
+  spec_cancer <- umap_cancertype %>% 
+    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
+                               TRUE ~ "Other")) %>%
+    filter(J_color != "Other")
+  if(nrow(spec_cancer) == 0) {next
+    }
+  subtypes <- unique(spec_cancer$J_color)
+  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
+  names(J_colors) <- subtypes
+
+  with(spec_cancer,
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+  legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  title(xlab="UMAP 2", line=0)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+
+for(n in cancer_w_subtypes[13:15]) {
+  spec_cancer <- umap_cancertype %>% 
+    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
+                               TRUE ~ "Other")) %>%
+    filter(J_color != "Other")
+  if(nrow(spec_cancer) == 0) {next
+    }
+  subtypes <- unique(spec_cancer$J_color)
+  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
+  names(J_colors) <- subtypes
+
+  with(spec_cancer,
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+  legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  title(xlab="UMAP 2", line=0)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+
+for(n in cancer_w_subtypes[16:19]) {
+  spec_cancer <- umap_cancertype %>% 
+    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
+                               TRUE ~ "Other")) %>%
+    filter(J_color != "Other")
+  if(nrow(spec_cancer) == 0) {next
+    }
+  subtypes <- unique(spec_cancer$J_color)
+  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
+  names(J_colors) <- subtypes
+
+  with(spec_cancer,
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+  legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+#legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  title(xlab="UMAP 2", line=0)
+}
+
+# ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
+#   geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+#   geom_point(size = 0.25) +
+#   theme_classic() +
+#   coord_fixed() +
+#   xlab("UMAP 1") + ylab("UMAP 2") +
+#   theme(axis.text.x = element_blank(),
+#     axis.text.y = element_blank(),
+#     axis.title.x = element_blank(),
+#     axis.title.y = element_blank(),
+#     axis.ticks = element_blank(),
+#     axis.line = element_blank(),
+#     legend.position = "none") +
+#   facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
 ```
+**Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
+cancers projected onto the tumors of the UMAP from **Figure 2C of the main text**.
 
-![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
-
-### Supplementary Figure S3
+\newpage
 
@@
 
-selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
+selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
   left_join(., selected_mut, by = "sample_id") %>%
-  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
   filter(value != 0)
@@
 
-```{r supplementary-figure-s3, fig.width = 7, fig.height= 5, fig.align = "center"}
+```{r supplementary-figure-s3, fig.width = 9}
 # out.width = '77%'
 load("../data/selected_mut.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
 
-ggplot(selected_mut, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ factor(variable, levels = c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")), ncol = 5) ## facet by gene
+genes <- c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in genes[1:6]) {
+with(selected_mut %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
+par(mar = c(2,2,0,0))
+for(n in genes[7:12]) {
+with(selected_mut %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in genes[13:15]) {
+with(selected_mut %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
 ```
+**Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
+genes involved in genome stability projected on the UMAP from **Figure 2C of the main text**.
 
-![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
+\newpage
 
 ### Supplementary Figure S4
+This is the 377 pathway UMAP. I will need to develop a loop that does this.
 
+**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
+pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
+(column) from **Figure 2C of the main text**.
 ```{r supplementary-figure-s4-data-generation, eval = FALSE}
@@
 
-![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]
+\newpage
 
@@
 load("../data/gene_data.Rda")
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
 
@@
 
-s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
+s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
   left_join(., selected_mut, by = "sample_id") %>%
-  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
   filter(value != 0)
@@
 
-```{r supplemental-s5-plot}
-load("../data/s5_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+```{r supplemental-s5-plot, fig.width=9}
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
 
-clust_178 <- umap_cancertype %>% filter(clust_knn %in% c(1, 7, 8))
+umap_cancertype$clust_knn <- as.character(umap_cancertype$clust_knn)
 
-umap_clust178 <- ggplot(clust_178, aes(x = plot_y, y = plot_x, colour = as.character(clust_knn))) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") + labs(colour = "Cluster") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_colour_manual(values = c("gold2", "gray51", "springgreen4")) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_text(size = 12),
-    axis.title.y = element_text(size = 12),
-    axis.ticks = element_blank(),
-    legend.position = c(0.95, 0.8))
+clusters178 <- umap_cancertype %>% 
+       mutate(color = case_when(clust_knn %in% c("1", "7", "8") ~ clust_knn,
+                                  TRUE ~ "Other"))
+
+I_outlines <- c("black", "black", "black", "#00000000")
+names(I_outlines) <- c("1", "7", "8", "Other")
+
+I_colors <- c("gold2", "gray51", "springgreen4", "#00000011")
+names(I_colors) <- c("1", "7", "8", "Other")
 
 load("../data/s5_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
-class7_8 <- umap_cancertype %>% filter(clust_knn %in% c(7, 8))
-umap_bg_7_8 <- class7_8[,-"project_code"] ## get background data for facet
+genes <- as.character(unique(s5_data$variable))
 
-# pik3ca_kras <- ggplot(s5_data %>% filter(sample_id %in% umap_bg_7_8$sample_id) %>% filter(variable != "TP53") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-#   geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-#   geom_point(size = 0.25) +
-#   theme_classic() +
-#   coord_fixed() +
-#   #ylim(-0.50, 1.5) +
-#   xlab("UMAP 1") + ylab("UMAP 2") +
-#   scale_colour_manual(values = rep("firebrick3", 20)) +
-#   theme(axis.text.x = element_blank(),
-#     axis.text.y = element_blank(),
-#     axis.title.x = element_blank(),
-#     axis.title.y = element_blank(),
-#     axis.ticks = element_blank(),
-#     axis.line = element_blank(),
-#     legend.position = "none") +
-#   facet_wrap(~ variable, ncol = 2)
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+plot.new()
+with(clusters178, 
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = I_colors[color], col = I_outlines[color],
+                                pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+legend("top", inset = c(0.05), legend=names(I_colors[c(1:3)]), col=I_colors[c(1:3)], pch = 20, bty = "n", horiz = T, cex = 1.2)
+title(xlab="UMAP 2", line=0)
+plot.new()
 
-tp53 <- ggplot(s5_data %>% filter(variable == "TP53") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ variable, ncol = 1)
+for(n in genes) {
+with(s5_data %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+```
+**Supplementary Figure S5. Distribution of oncogenes on UMAP clusters.** The left panel shows the identity of 
+clusters 1, 7 and 8. The multipanel on the right shows one of the three oncogenes projected onto the tumors of 
+the UMAP from **Figure 2C of the main text**.
 
-pik3ca <- ggplot(s5_data %>%  filter(variable == "PIK3CA") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  #ylim(-0.50, 1.5) +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ variable, ncol = 1)
-
-kras <- ggplot(s5_data %>%  filter(variable == "KRAS") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  coord_fixed() +
-  #ylim(-0.50, 1.5) +
-  xlab("UMAP 1") + ylab("UMAP 2") +
-  scale_colour_manual(values = rep("firebrick3", 20)) +
-  theme(axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_blank(),
-    axis.title.y = element_blank(),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    legend.position = "none") +
-  facet_wrap(~ variable, ncol = 1)
-
-layout <- '
-AABB
-AACC
-AADD
-'
-
-umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
-```
-
-### Supplementary Table ST1
-
+\newpage 
 ```{r supplemental-table-st1-data-generation, eval = FALSE}
@@
 
-```{r enrich-pathway-in-clust}
+```{r enrich-pathway-in-clust, eval = FALSE}
 ## docx friendly
@@
 Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._](../figures/fig6A.png)
+The orange and green curves show the model fits._]

### 2020-08-14 text/fig_legend.Rmd
Update figures with 3d umap, fig ref in text, fig legend text

@@
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
-Classification of tumors by pathway disruptions.** a) Each of 377 selected
-Reactome pathways (columns) is classified as disrupted if one or more genes is
-mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
-standardized abbreviations from the TCGA project. b) MCA analysis was performed to 
-determine the eigenvalue with the most explanatory power from the categorical data.
-Average of 100 permutations of the data was used as a baseline. c) The p-value of
-each eigenvalue was plotted and the max eigenvalue with a pvalue < 0.05 was selected. 
-d) The resulting MCA was projected down to 3 dimensions for the purpose of density 
-based clustering (only the first 2 dimensions are shown). Each dot represents a tumor 
-e) HDBSCAN was used to perform hierarchical density based clustering tuned to minimize
-the fraction of unassigned tumors. f) HDBSCAN clustered tumors were visualized on the UMAP
-projection; tumors are colored by cluster membership (see legend). This results
-in some tumors remaining unclassified (grey points). g) kNN was used to assign
-membership to the remaining tumors. h) Mutational burden plot where each dot corresponds
-to the sample mutation counts for tumors in each class. i) Projection of cancer 
-type identity onto UMAP. Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma,
-HNSC = Head and neck squamous cell, PAAD = Pancreatic adenocarcinoma. j) Projection of breast
-cancer subtypes onto the UMAP. k) Projection of BRCA1/2 somatic mutation onto the
-UMAP. l) Projection of tumor stage onto the UMAP, regardless of cancer type. m) 
-Projection of metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, 
-M1 = metastatic tumors.</p>
+**Figure 1: Molecular pathway profiles of tumor samples with one or more
+mutations.** Each of 377 selected Reactome pathways (columns) is classified
+as disrupted if one or more genes is mutated in the tumor sample (rows) where
+red and grey represents the pathway is mutated and not, respectively. Tumors
+are grouped by tissue of origin using standardized abbreviations 
+from the TCGA project.
 
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 2: Pan-cancer 
-enrichment of pathway disruptions.** a) Waterfall plot; columns represent tumors. 
-Each row represents a pathway. The pathways are ranked such that the pathway with 
-the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are
-ordered to prioritize the top pathway disruptions successively. The top 100 most 
-mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies
+**Figure 2: Method for tumor classification using pathway aberration data** 
+a) MCA analysis was performed to determine the eigenvalue with the most 
+explanatory power from the categorical data. Average of 100 permutations of 
+the data was used as a baseline. b) The p-value of each eigenvalue was plotted
+and the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA 
+was projected down to 3 dimensions for the purpose of density based clustering 
+(only the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN
+was used to perform hierarchical density based clustering tuned to minimize
+the fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized 
+on the UMAP projection; tumors are colored by cluster membership (see legend). 
+This results in some tumors remaining unclassified (grey points). f) kNN was 
+used to assign membership to the remaining tumors.
+
+**Figure 3: Pathway-based classification is independent of tissue-of-origin features.**
+a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer, 
+COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD = 
+Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the UMAP.
+c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of tumor 
+stage onto the UMAP, regardless of cancer type. e) Projection of metastatic status
+onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.
+
+**Figure 4: Somatic mutation frequencies for tumors in each class** Column uses 
+a logarithmic axis to represent mutation count. Row represents the tumor samples. 
+Each dot corresponds to the sample mutation count for tumors in each class, 
+ordered by mutation counts. The median mutation count in each cluster is indicated 
+by the horizontal line.
+
+**Figure 5: Pan-cancer enrichment of pathway disruptions.** a) Waterfall plot; columns
+represent tumors. Each row represents a pathway. The pathways are ranked such that the 
+pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and 
+tumors are ordered to prioritize the top pathway disruptions successively. The top 100 
+most mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies
 not disrupted. b) Heatmap shows the relative enrichment of each pathway (columns) 
@@
 pathway (columns; union set of pathways with effect size >= 0.30 in each cluster)
-within each numbered cluster (rows).</p>
+within each numbered cluster (rows).
 
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3: Gene level 
-analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses 
-a logarithmic axis to represent odds ratio with a 95% confidence interval. Row 
-represents the significant genes from each cluster. Each cluster was compared 
-against the background (all other clusters) to find deferentially mutated genes. 
-Significant genes (pvalue < 0.01) were selected and limited to the top ten results 
-for each clusters. Heatmap; Column represents the cancer type. Row represents the 
+**Figure 6: Gene level analysis reveals tissue-specific class signatures.** Odds ratio 
+plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence 
+interval. Row represents the significant genes from each cluster. Each cluster was 
+compared against the background (all other clusters) to find deferentially mutated 
+genes. Significant genes (pvalue < 0.01) were selected and limited to the top ten 
+results for each clusters. Heatmap; Column represents the cancer type. Row represents the 
 significant genes as described above. The heatmap shows the proportion of samples 
 per cluster, per cancer type that is mutated for each of the significant genes. 
-Depleted significant genes (left). Enriched significant genes (right).</p>
+Depleted significant genes (left). Enriched significant genes (right).
 
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1: Cluster-specific 
-enriched pathways (effect size >= 0.30) in metastatsis.**
+**Figure 7: **
 
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary figure S1. 
-Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
-individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Figure 1D of the main text**.
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S2. 
-Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Figure 1D of the main text**.
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S3.
-Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15  
-genes involved in genome stability projected on the UMAP from **Figure 1D of the main text**.
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S4. 
-Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
-pathways projected on the UMAP from **Figure 1D of the main text** ordered by the 
-hierarchical pathway order (column) from **Figure 2B of the main text**.
-
-<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S5.
-Distribution of oncogenes on UMAP clusters.** The left panel shows the identity of 
-clusters 1, 7 and 8. The multipanel on the right shows one of the three oncogenes 
-projected onto the tumors of the UMAP from **Figure 1D of the main text**.
+**Table 1: Cluster-specific enriched pathways (effect size >= 0.30) in metastatsis.**

### 2020-08-14 code/survival/survfig.Rmd

```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=9, fig.width=6.5}

library(ggridges)
library(patchwork)

source("./kmcurve.R")

## create example data for a fixed 30 year old without vs with a k=2
## cancer

patientsurvivalsbkd = replicate(100,simdeath(r20wf,ageratewf,30,1))/365
patientsurvivalsk2 = replicate(100,simdeath(r20wf,ageratewf,30,2))/365

mixedages = runif(100,30,70)
mixedsurvwfk1 = sapply(mixedages, function(x) simdeath(r20wf,ageratewf,x,1)/365)
mixedsurvwfk2 = sapply(mixedages, function(x) simdeath(r20wf,ageratewf,x,2)/365)

p1 = ggplot(data.frame(r=patientsurvivalsbkd,c=patientsurvivalsk2,rm = mixedsurvwfk1, cm=mixedsurvwfk2, status=rep(1,length(patientsurvivalsbkd)))) +
geom_km(aes(time=r,status=status),color="blue") +
geom_km(aes(time=c,status=status),color="red") +
geom_km(aes(time=rm,status=status),color="darkblue") +
geom_km(aes(time=cm,status=status),color="darkred") +
annotate("text",rep(90,4),c(.9,.7,.5,.3),color=c("blue","red","darkblue","darkred"),label=c("Random 30 year old women","Random 30 year old women cancer k=2","Random uniform(30,70) year old women","Random uniform(30,70) year old women k=2"),hjust="right")+
labs(title="Simulated patients",x="Time (years)",y="Fraction Surviving")

pstad = plotcancertype(cancerdat,"STAD")
pgbm = plotcancertype(cancerdat,"GBM")
ppaad = plotcancertype(cancerdat,"PAAD")
thca = plotcancertype(cancerdat,"THCA")

load("./survmodel-ethbkd.stansave")

dfsamps = as.data.frame(sampseth)

kcancplot = function(dfsamps,tisnames){
	ktisnames = grep("ktis",names(dfsamps),value=TRUE)
	dfktis = dfsamps[,ktisnames]
	dfktis = stack(dfktis) ## convert to two column "values" and "ind" with "ind" being the name of the variable
	names(dfktis) = c("value","parname")

	dim(tisnames) = c(length(tisnames),1)

	dfktis$cancname = as.factor(sapply(dfktis$parname,function(x) tisnames[x]))
	ggplot(dfktis,aes_(quote(value),quote(cancname))) + geom_density_ridges()+ theme_minimal() + scale_fill_brewer(palette="Paired") + xlim(0,22) + labs(x="ktis multiplier",y="Cancer Type",title="Tissue Specific Effects")
}

kclassplot = function(dfsamps,tisnames){
	kclnames = grep("k\\[",names(dfsamps), value=TRUE)
	dfclas = dfsamps[,kclnames]
	dfclas = stack(dfclas)
	names(dfclas) = c("value","coefname")
	dfclas$class = as.factor(as.numeric(gsub("k\\[(.*),(.*)\\]","\\2",dfclas$coefname)))
	dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
	ggplot(dfclas,aes_(quote(value),quote(class))) + geom_density_ridges() +
		facet_wrap(. ~ tiss,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients")
}

pcanc = kcancplot(dfsamps,levels(cancerdat$type))
pclass = kclassplot(dfsamps,levels(cancerdat$type))

mytheme = theme(
#    axis.text.x = element_blank(),
#    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
#    axis.ticks = element_blank(),
#    axis.line = element_blank(),
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_blank(),
    legend.position = "none",
    title = element_text(size=10)
  )

mylayout = '
AAAAAAAA
BBBBBBBB
BBBBBBBB
CCCCCCCC
CCCCCCCC
CCCCCCCC
'

print(
((p1+mytheme)/((pstad[[1]]+mytheme)+(pgbm[[1]]+mytheme) +
(ppaad[[1]]+mytheme) + (thca[[1]] + mytheme)))/((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
)

```

### 2020-08-14 code/figure2v2.Rmd

@@
        col=c("#998ec3", "#f1a340"), lty=1, cex=0.8)
-### D
+mtext("A", side=3, line=0, cex=1.2, adj=-0.35)
+### C
 par(mar = c(2,2,0,0), bg = NA)
@@
 title(xlab="UMAP 2", line=0)
-### E
+mtext("C", side=3, line=-2.05, cex=1.2, adj=0)
+### D
 par(mar = c(4,6,2,0) + 0.1)
@@
 axis(2, at=c(0.25, 0.35, 0.45, 0.55), las = 1)
-
-### C
+mtext("D", side=3, line=0, cex=1.2, adj=-0.27)
+### B
 with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = pvalue, 
@@
 title(xlab = "MCA Eigenvalues", line = 2)
-### F
+mtext("B", side=3, line=0, cex=1.2, adj=-0.35)
+### E
 par(mar = c(2,2,0,0), bg = NA)
@@
 title(xlab="UMAP 2", line=0)
-### G
+mtext("E", side=3, line=-2.05, cex=1.2, adj=0)
+### F
 with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
@@
 title(xlab="UMAP 2", line=0)
+mtext("F", side=3, line=-2.05, cex=1.2, adj=0)
 

### 2020-08-14 code/figure4v2.Rmd

@@
+---
+output: html_document
+editor_options: 
+  chunk_output_type: console
+---
 ```{r figure-4-plot, fig.width = 9}
@@
 
-### I - Cancer Type
+### A - Cancer Type
 par(opar)
@@
                colvar = NULL))
-legend("top", inset = c(0), legend=rev(names(I_colors)), col=rev(I_colors), pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0), legend=rev(names(I_colors[1:2])), col=rev(I_colors[1:2]), pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=rev(names(I_colors[3:4])), col=rev(I_colors[3:4]), pch = 20, bty = "n", horiz = T, cex = 1.2)
 title(xlab="UMAP 2", line=0)
-### J - BRCA subtype
+mtext("A", side=3, line=-2.05, cex=1.2, adj=0)
+### B - BRCA subtype
 J_colors <- c(ditto_color[c(2,3,4,5)])#, 'grey')
@@
                colvar = NULL))
-legend("top", inset = c(0), legend=names(J_colors), col=J_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
 title(xlab="UMAP 2", line=0)
-### K - BRCA Somatic Mutations
+mtext("B", side=3, line=-2.05, cex=1.2, adj=0)
+### C - BRCA Somatic Mutations
 load("../data/gdc_gene_exp_adj.Rda")
@@
                colvar = NULL))
-legend("top", inset = c(0), legend=names(K_colors), col=K_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0), legend=names(K_colors[c(1,3)]), col=K_colors[c(1,3)], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend="BRCA1,BRCA2", col=K_colors[2], pch = 20, bty = "n", horiz = T, cex = 1.2)
 title(xlab="UMAP 2", line=0)
-### L - Stage
+mtext("C", side=3, line=-2.05, cex=1.2, adj=0)
+### D - Stage
 L_colors <- c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')
@@
                colvar = NULL))
-legend("top", inset = c(0), legend=names(L_colors), col=L_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0), legend=names(L_colors[1:2]), col=L_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=names(L_colors[3:4]), col=L_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
 title(xlab="UMAP 2", line=0)
-### M - Metastasis
+mtext("D", side=3, line=-2.05, cex=1.2, adj=0)
+### E - Metastasis
 M_colors <- c('#a6cee3','#1f78b4')
@@
 title(xlab="UMAP 2", line=0)
+mtext("E", side=3, line=-2.05, cex=1.2, adj=0)
 

### 2020-08-14 code/supplements.Rmd

@@
 The orange and green curves show the model fits._](../figures/fig6A.png)
+
+#### Supplemental figure text
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary figure S1. 
+Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
+individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
+in **Figure 1D of the main text**.
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S2. 
+Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
+cancers projected onto the tumors of the UMAP from **Figure 1D of the main text**.
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S3.
+Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15  
+genes involved in genome stability projected on the UMAP from **Figure 1D of the main text**.
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S4. 
+Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
+pathways projected on the UMAP from **Figure 1D of the main text** ordered by the 
+hierarchical pathway order (column) from **Figure 2B of the main text**.
+
+<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Supplementary Figure S5.
+Distribution of oncogenes on UMAP clusters.** The left panel shows the identity of 
+clusters 1, 7 and 8. The multipanel on the right shows one of the three oncogenes 
+projected onto the tumors of the UMAP from **Figure 1D of the main text**.

### 2020-08-14 text/discussion.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

@@
----
-title: "Discussion"
-output: 
-  bookdown::html_document2:
-    fig_captions: TRUE
-    number_sections: FALSE
-    toc: FALSE
-    code_folding: hide
-    theme: flatly
-    smart: true
-bibliography: references.bib
-editor_options: 
-  chunk_output_type: console
----
-
 ### Classification of tumors independent of tissue-of-origin
@@
 on the UMAP projection exhibit increasing specificity. We identified no such
-trend in the number of mutations (see **figure 1H**), the relative staging
-(**figure 1L**) or metastasis (**figure 1M**); and each class instead was
-associated with unique combinations of pathways (as in **figures 2B, C**).
+trend in the number of mutations (see **figure 4**), the relative staging
+(**figure 2D**) or metastasis (**figure 2E**); and each class instead was
+associated with unique combinations of pathways (as in **figures 5B, C**).
 
 We find that most tumors segregate into one of ten biologically distinct
-clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
+clusters, regardless of the cancer type (**figure 2F, 3A** and **supplementary
 figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
@@
 surprisingly to us, this finding extends to histological subtypes of breast,
-head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
+head and neck cancers, leukemias, _etc._ (see **Figure 3B** and **Supplemental
 Figure S2**). This result implies that many, if not most, histological subtypes
@@
 of the Luminal A subtype were heavily biased toward membership in clusters 1 and
-7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
+7, and basal tumor samples were biased toward cluster 6 (**figure 3B**).
 However, both subtypes also contained samples in every other class (without
@@
 classes, 1 and 10, for which we did not find distinctive associations with
-pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
+pathways (**figure 5B, C**). Class 1 has a relatively low proportion of mutated
 pathways across the board, although it is broadly enriched in many of the same
@@
 that this class is as likely to contain stage IV metastatic tumors as it is to
-contain those of stage I (**figure 1L, M**). This class likely represents a
+contain those of stage I (**figure 3D, E**). This class likely represents a
 group of tumors with aberrations in methylation, copy number alterations,
@@
 of our pathway disruption classes exhibit cancer-type specific effects on
-expected survival (**figure 4**). Contrary to what one might expect at the
+expected survival (**figure 7**). Contrary to what one might expect at the
 outset, we found that models assuming different survival based on class _and_
@@
 senesce) but together produce more specialized cancer phenotype and increased
-fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our
+fitness (as in **figure 6** and **tables 1** and **ST2**). Importantly, our
 observations do not imply the order in which these mutations should accumulate.
@@
 expanded analysis with larger cohorts. Nonetheless, the class-specific
-metastatic pathway enrichment we observe (in **table 2**) could be a result of
+metastatic pathway enrichment we observe (in **table 1**) could be a result of
 the pathways directly promoting metastatic phenotypes in these classes, or
@@
 defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
-traditional extra-cellular signaling pathways (as exemplified in **table 1**).
+traditional extra-cellular signaling pathways (as exemplified in **table S1**).
 It may be that the primary effects of mutations in protein and RNA metabolism in
@@
 pathways) mutations in glucagon and glycogen metabolism, respectively, in their
-metastases (**table 2**), consistent with energy metabolism as one of the new
+metastases (**table 1**), consistent with energy metabolism as one of the new
 hallmarks [@hallmarks_II], but also distinguishing these classes from the others

### 2020-08-14 text/manuscript.Rmd
Create a single large panel figure for the survival stuff, rewrite the text to deal with that appropriately. Also adjust the Makefile to build the script properly now that we're using .docx
Reorganize some of the survival code.

@@
 carries a mutation) to the pathway regardless of the number of genes carrying
-mutations (**figure 1A**). </strike>
+mutations (**figure 1**). </strike>
 
@@
 analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
-containing useful information. (**figures 1B, C**). Then we performed a UMAP
+containing useful information. (**figures 2A, B**). Then we performed a UMAP
 analysis, both in order to summarize the MCA graphically, and as a preprocessing
@@
 reproducible regardless of random seed setting. A representative version of this
-3D UMAP is shown in **figure 1D** as a 2D projection, rotated to enhance the
+3D UMAP is shown in **figure 2C** as a 2D projection, rotated to enhance the
 visibility of the major features. Following this spatial mapping we attempted to
@@
 $< 5\%$ in one class and chose a cluster size of 92 to minimize the score
-function (**figure 1E**). HDBSCAN with these settings resulted in ten distinct
-high-density clusters which we then projected onto the UMAP (**figure 1F**).
+function (**figure 2D**). HDBSCAN with these settings resulted in ten distinct
+high-density clusters which we then projected onto the UMAP (**figure 2E**).
 This classified 6,038 out of 7,607 tumors but still resulted in a
@@
 the square root of the number of tumor samples (87). Using this method, we
-assigned cluster membership to the remaining tumors (**figure 1G**).
+assigned cluster membership to the remaining tumors (**figure 2F**).
 
@@
 exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 1I**
+cancer types had representative tumors in every class (for example **figure 3A**
 and full tumor profiles in **supplementary figure S1**, see also interactive
@@
 prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations (**Figure 1J** and **supplemental
+pathway for these subtype annotations (**Figure 3B** and **supplemental
 file 1**). We also projected histological subtype data for the remaining cancers
@@
 mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 1K**). We made
+segregation of these mutations into specific clusters (**figure 3C**). We made
 similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_,
@@
 pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 1D**) could in
+structures on a common backbone of tumor samples (**figure 2C**) could in
 principle also reflect progression through a series of stages. In support of
 this hypothesis, the backbone starts with a cluster of tumors (class 1) that has the fewest point mutations and culminates in a cluster (class 10) which
-has nearly every pathway disrupted (**figure 1H**). However, outside of class 10 we don't observe
+has nearly every pathway disrupted (**figure 4**). However, outside of class 10 we don't observe
 an obvious trend in the overall mutation burden across the backbone of the UMAP (**figure
@@
 observe any bias amongst the stages into specific pathway disruption classes
-(**figure 1L**), suggesting that stage is not an determining factor in pathway
+(**figure 3D**), suggesting that stage is not an determining factor in pathway
 disruption classes.
@@
 any of our classes correspond to an especially advanced stage of cancer accross
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly we observed an even distribution across classes (**figure 1M**). This
+tissue types, we projected the metastasis data onto our UMAP, and surprisingly we 
+observed an even distribution across classes (**figure 3E**). This
 final observation suggests that our pathway-disruption classification is
@@
 visualizations on cBioportal [@cbioportal], replacing genes with pathways
-(**figure 2A**). A complete list of pathway disruptions with percent mutated
+(**figure 5A**). A complete list of pathway disruptions with percent mutated
 samples and top genes is provided in **supplementary table ST1**). As expected,
@@
 samples), _etc._ . The top third of pathways drop off gradually in overall
-mutation frequency (**figure 1A** and **supplementary table ST1**), possibly
+mutation frequency (**figure 5A** and **supplementary table ST1**), possibly
 owing to relative diversity among all tumors.
@@
 and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 2B**). Using this approach, we
+visualized the enrichment as a heatmap (**figure 5B**). Using this approach, we
 identified 25 pathways highly enriched (enrichment score > 0.3, see methods) in
@@
 extracellular, intracellular, and immune-related signaling pathways (see
-**figure3** and **supplementary figure S4**). Cluster 2 had the highest pathway
+**figure5C** and **supplementary figure S4**). Cluster 2 had the highest pathway
 enrichment levels of the three, having mutations in hedgehog signaling,
@@
 and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
-(**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
+(**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
 signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
@@
 7, but depleted in clusters 3 and 8). Next, we investigated the proportion of samples per 
-cancer type for the significant genes within a cluster (**Figure 4**; heatmap). Surprisingly, 
+cancer type for the significant genes within a cluster (**Figure 6**; heatmap). Surprisingly, 
 cluster-specific tumors were not predominated by one or more highly mutated genes across all 
@@
 Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 2**) we applied this method to metastatic vs non-metastatic
+pathways (**figure 5B**) we applied this method to metastatic vs non-metastatic
 tumors, as we find that metastatic tumors are distributed across all ten clusters
-( _e.g._ **figure 1m**). Using all non-metastatic tumors as background, we found
+( _e.g._ **figure 3E**). Using all non-metastatic tumors as background, we found
 very low levels of enrichment, less than $10\%$, in a handful of pathways. We
@@
 ```
+#### Figure 1
+```{r child='../code/figure1v2.Rmd'}
+#
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 2
+```{r child='../code/figure2v2.Rmd'}
+#
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 3
+```{r child = '../code/figure4v2.Rmd'}
+#
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 4
+```{r child = '../code/figure3v2.Rmd'}
+#
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 5
+```{r child = '../code/figure2.Rmd'}
+#
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 6
+```{r child = '../code/figure4.Rmd'}
+#
+```
+#### Figure 7
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+```{r child='../code/survival/survfig.Rmd'}
+
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Table 1
+```{r child = '../code/table2.Rmd'}
+#
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 
@@
 ```
-#### Figure 1
-```{r child='../code/figure1v2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 2
-```{r child='../code/figure2v2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 3
-```{r child = '../code/figure3v2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 4
-```{r child = '../code/figure4v2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 5
-```{r child = '../code/figure2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 6
-```{r child = '../code/figure4.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Table 1
-```{r child = '../code/table2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-![](../figures/fig6B.png)
-\newpage
-![](../figures/fig6C.png)
-
-## Supplementary Information
-
-```{r }
-#child='../code/supplements.Rmd'
-```
 

### 2020-08-14 text/abstract.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

@@
 obscure. To address this challenge, we reduced mutation data to disruptions in a
-select set of pathways pointing to basic cell biology, from DNA replication to
+select set of pathways relevant to basic cell biology, from DNA replication to
 cellular communication. Using dimension reduction techniques, we assigned tumor
-samples into twelve classes distinct from their tissues-of-origin and largely
+samples into ten classes distinct from their tissues-of-origin and largely
 free of bias from mutational burden or clinical stage. We showed that these
@@
 classes and cancer types. We identified class-specific mutations in different
-tissues, demonstrating how tissue-specific signatures contribute to common
-cellular phenotypes. Moreover, germline risk mutations involved in genome
+tissues, demonstrating that tissue-specific signatures contribute to common
+cellular phenotypes. Moreover, driver gene mutations involved in replication fidelity and genome
 stability, such as BRCA1 and Rb1, are equally distributed among our classes,
-somewhat contrary to the expectation that such genes are hallmarks of molecular
-subtypes. We also investigated pathways that distinguish between metastatic and
+contrary to the expectation that such genes are avatars of molecular
+subtype. We also investigated pathways that distinguish between metastatic and
 non-metastatic tumors, and found that most meaningful differences are
-class-specific. Most metastatic enriched pathways from one class in metastatic
+class-specific. Many metastastasis enriched pathways from one class of
 tumors are class-specific pathways from another class, suggesting phenotypic
 convergence, and potentially highlighting key pathways to focus on
-therapeutically for prevention of class-specific relapse. In summary, our
+for prevention of class-specific relapse. In summary, our
 observations suggest that some genes act as drivers of genome instability in all
@@
 in a combinatorial fashion, eventually converging on one or more end-stage
-disease phenotypes. Thus, we present a coherent picture of global tumor biology,
+disease phenotypes. Thus, we present a coherent view of global tumor biology,
 and explain how common cellular dysfunction might arise from tissue-specific

### 2020-08-14 text/introduction.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

@@
 together, with the goal of integration and increased understanding of the
-mechanisms of cancer. These efforts have focused on recognizing and quantifying
-the importance of genes in cancer etiology as the result of selection processes,
-or classification of cancers relative to common patterns of aberration in
-multiple assays.
+mechanisms of cancer.
 
@@
 @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
-when genes are mutated at a greater rate than expected due to chance. Based in
-part on studies like these, it has been estimated that on the order of three
+when genes are mutated at a greater rate than expected due to chance. It has been estimated that on the order of three
 mutations in key oncogenes and/or tumor suppressors would be sufficient to
@@
 expression and genomic signatures, with the express goal of understanding the
-common themes of all cancer independent of tissue of origin. Such a view of
-cancer as a malady affecting common cellular processes has long been a goal of
+common themes of all cancer independent of tissue of origin. Understanding cancer
+as a disease of the cell has long been a goal of
 the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
-Discussion). The first of these studies, (the
-“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
+Discussion). The first of these studies in the enomic era identified 11 subtypes from 12 cancer types,
 using integrative analysis with co-equal weighting of gene expression,
@@
 tissue-of-origin is the predominant driving factor, though a significant
-proportion of tumors ($10\%$) could be reclassified independent of
+proportion of tumors (~10%$) could be reclassified independent of
 tissue-of-origin. In a second study involving 33 cancer types and a much greater
-number of tumors using a newer algorithm (iCluster)
-[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
+number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
 could be further subdivided into organ specific groups, including
@@
 More recently, attempts have been made to consider pan-cancer mutations and
-other data outside the context of enrichment at the gene level. Horn _et al._
+other data outside the context of enrichment at the gene level. @PCAWG-PATH carried out near-exhaustive analysis of pathway enrichment using whole genome data. Horn _et al._
 [@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
@@
 couple hundred common driver genes and mutations, there are likely thousands of
-moderate effect genes that occur at such low frequency given tissue diversity
-and low sample numbers that they are impossible to detect using positive
+moderate effect genes that occur at such low frequency--given tissue diversity--and low sample numbers that they are impossible to detect using positive
 selection theory. Some groups have attempted to address this challenge by
@@
 molecular pathways as the unit of disruption in cancer. Our approach strips all
-cell-type-specific information from the mutation data and generalizes gene-level
-information to cell-biological pathway phenotypes. We use this heuristic to
+cell-type-specific information from the mutation data and equates gene-level
+mutations to cell-biological pathway disruptions. We use this heuristic to
 evaluate all cancers and show that, surprisingly, tumors fall into categories of

### 2020-08-14 text/manuscript.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

@@
 Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
-(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
+(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our
 study given the relative completeness of the patient metadata, particularly for
-survival and staging. We chose to start with somatic mutations in exome
+survival and staging. We chose to analyze somatic mutations in exome
 sequencing data because the affected target gene is known unambiguously.
-Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
+Therefore, we selected all missense, nonsense, frameshift, stop-gain, untranslated
 region, and splicing mutations. In order to minimize bias from well-studied diseases
@@
 
-It does not matter for our purposes whether a single tumor is “enriched” for
-mutations of individual pathways, chiefly because such metrics favor larger
-pathways ( _i.e._ pathways with more member genes) over smaller ones. Data
+To avoid bias toward larger
+pathways ( _i.e._ pathways with more member genes), we counted pathways as disrupted if one or more member genes were mutated.
+We do not attempt to calculate enrichment for mutations within a pathway, for example.
+Data
 binarized by pathway are likely to be noisy for several reasons. First, point
 mutations can be deleterious (attenuating or hypomorphic) or activating
-(neomorphic or hypermorphic) in genes that act functionally as either activators or repressors.
+(neomorphic or hypermorphic) in genes, and these can in turn act functionally as either activators or repressors.
 Within the scope of this study we accept this uncertainty and assume mutations
-are generically disruptive to the pathway regardless of whether they are
+are generically disruptive to pathways regardless of whether they are
 activating or deactivating or whether they affect a gene that we might classify
@@
 eliminated them for that cancer type only (see methods for details). 
-<strike> Highly expressed genes could also have high mutation rates owing to
+Highly expressed genes could also have high mutation rates owing to
 transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
-that this mechanism results in cell-type-specific biases that might be
+that such mechanisms result in cell-type-specific biases that might be
 biologically meaningful for predisposition to different classes of cancer
@@
 carries a mutation) to the pathway regardless of the number of genes carrying
-mutations (**figure 1**). </strike>
+mutations (**figure 1**).
 
@@
 step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several large, robust projections
-reproducible regardless of random seed setting. A representative version of this
-3D UMAP is shown in **figure 2C** as a 2D projection, rotated to enhance the
+notable for its lobed structure, with several reproducible projections
+regardless of random seed setting. A representative version of this
+3D UMAP is shown in **figure 2C**, rotated to enhance the
 visibility of the major features. Following this spatial mapping we attempted to
@@
 samples in a cluster that capture the maximum number of tumor samples, measured
-by probability of membership of $\geq 5\%$ in at least one class. Thus, we
+by probability of membership of ≥ 5% in at least one class. Thus, we
 created a score metric as the fraction of classified tumors with max probability
-$< 5\%$ in one class and chose a cluster size of 92 to minimize the score
+< 5% in one class and chose a cluster size of 92 to minimize the score
 function (**figure 2D**). HDBSCAN with these settings resulted in ten distinct
@@
 Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
-the set and then lets the $k$ most similar tumors "vote" as to the identity of
-the query tumor sample based on their classification labels. We set $k$ to be
+the set and then lets the _k_ most similar tumors "vote" as to the identity of
+the query tumor sample based on their classification labels. We set _k_ to be
 the square root of the number of tumor samples (87). Using this method, we
@@
 exceptions) were not heavily biased in one class, and all well-represented
-cancer types had representative tumors in every class (for example **figure 3A**
+cancer types had tumors in every class (see **figure 3A**
 and full tumor profiles in **supplementary figure S1**, see also interactive
@@
 least, our novel pathway-disruption classification method identifies molecular
-pathologies largely independent of tissue-of-origin considerations. As an
+pathology largely independent of tissue-of-origin. As an
 example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was biased in class 8, which is
+pancreatic adenocarcinoma (PAAD) was predominantly found in class 8, which is
 distinguished  by KRAS mutations among others (**supplementary figure S3**). 
 But even PAAD comprises tumors from the nine remaining classes, meaning that 
-patients hosting these tumors have potentially different underlying molecular
+patients suffering these tumors have potentially different underlying molecular
 pathologies.
@@
 result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. For example we projected annotations for each of the breast
+pathway disruptions. To illustrate this, we projected annotations for each of the breast
 cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
@@
 the subtypes, though often biased towards one or more classes, are almost never
-exclusive to one of the classes. We interpret these data in aggregate to mean
+exclusive. We interpret these data in aggregate to mean
 that our pathway disruption classes do not correspond to molecular subtypes
@@
 treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability, proof-reading and DNA damage repair, and
+factors such as genome stability generally, proof-reading and DNA damage repair, and
 telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
@@
 
-Tumor staging is based on physico-pathological criteria, including diameter,
+Tumor staging is based on physico-pathological criteria, including tumor diameter,
 which can vary greatly in importance between different tissues. Stage is used
@@
 UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias amongst the stages into specific pathway disruption classes
+observe any bias among the stages to specific pathway disruption classes
 (**figure 3D**), suggesting that stage is not an determining factor in pathway
@@
 transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To determine whether
-any of our classes correspond to an especially advanced stage of cancer accross
+any of our classes correspond to an especially advanced stage of cancer across
 tissue types, we projected the metastasis data onto our UMAP, and surprisingly we 
-observed an even distribution across classes (**figure 3E**). This
+observed an even distribution of metastases across classes (**figure 3E**). This
 final observation suggests that our pathway-disruption classification is
@@
 (**figure 5A**). A complete list of pathway disruptions with percent mutated
-samples and top genes is provided in **supplementary table ST1**). As expected,
-these reveal the broad importance of many well known pathways that are disrupted
+samples and top genes is provided in **supplementary table ST1**. As expected,
+these analyses reveal the broad importance of many well known pathways that are disrupted
 in cancer, including "PIP3 activates Akt signaling" (77% of samples),
@@
 mutation frequency (**figure 5A** and **supplementary table ST1**), possibly
-owing to relative diversity among all tumors.
+owing to the relative diversity of tumors.
 
@@
 0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
-terms of which pathways had enrichment. To explore the specific pathways marking
+terms of pathway enrichment. To explore the specific pathways marking
 each class, we projected disruptions for each of the 377 pathways onto the UMAP
@@
 enrichment levels of the three, having mutations in hedgehog signaling,
-"$\beta$-catenin degradation", "cellular response to hypoxia", "regulation of
+"β-catenin degradation", "cellular response to hypoxia", "regulation of
 cell cycle" and "apoptosis" among others.
 
-### A combination of common driver mutations and tissue-specific pathway genes create tissue-specific class signatures
+### A combination of common driver mutations and tissue-specific genes contribute to common class signatures
 
-As discussed previously, prior efforts to extract common signatures from pan-cancer 
+Prior efforts to extract common signatures from pan-cancer 
 datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
@@
 tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
-thresholds), or a combination of both? To answer this question, we compared each cluster 
-relative to the background to find deferentially mutated genes. We ranked the odd ratios
+thresholds), or a combination of both? To answer this question, we compared top pathway genes for each cluster 
+relative to the TCGA background to find differentially mutated genes. We ranked odds ratios
 and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
@@
 with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
-that tumors in this cluster could be a derivative either cluster 7 or 8. Clusters 3 and 
+that tumors in this cluster could be derivative of either cluster 7 or 8. Clusters 3 and 
 5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
@@
 nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
-We also observed genes that were enriched for a cluster are depleted from others ( _e.g._ 
+We also observed that genes that were enriched for one cluster are depleted from others ( _e.g._ 
 _TP53_ is enriched in cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster
@@
 networks as a whole. Taken together, our data identifies a framework of cancer type-specific 
-mutation associated with a specific cluster.
+mutations associated with specific clusters.
 
@@
 ( _e.g._ **figure 3E**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment, less than $10\%$, in a handful of pathways. We
+very low levels of enrichment (≤ 10%) in a handful of pathways. We
 reasoned that the individual clusters might be too different to detect global
@@
 Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found a total of 27 enriched pathways across all clusters (**table 1**). A number of
+found a total of 31 enriched pathways across all clusters (**table 1**). A number of
 enrichments were found in multiple clusters represented by pathways that were
@@
 **supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
-metastatic tumors of clusters 7 and 9 (p < 10-3, **table 1**). This is also true of 
+metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$, **table 1**). This is also true of 
 “Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 7 and 9. <strike>Cluster 1, one of two clusters 
-for which we could not identify any distinguishing pathways (cluster 8 being the other), 
-had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other
-clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
-7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).</strike> Cluster 4 metastatic
-samples was enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathways, 
-which are specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
+S4**) and also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic
+samples were enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathway, 
+which is also specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
 specific to metastases tend toward complementary mutations that are otherwise specific to
 non-metastatic tumors of other clusters.
-
-<mark>Statement does not apply anymore. The highest # of enriched path is 3. Not many are in 
-metastasis. Only 2. “Fc epsilon receptor (FCERI) signaling” and "Signaling by SCF-KIT" </mark>
-<strike>There were also a number of pathways that were enriched with lower specificity
-(among 4 or more clusters) which were found to be enriched in metastases. We
-identified 18 such pathways of which 12 had significantly higher mutation rates
-in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
-“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
-dependent signaling in response to WNT”, “Class I MHC mediated antigen
-processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
-Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
-Response”, and “Pre-NOTCH Expression and Processing” (**Table 1**). </strike>
 
@@
 All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript. In addition, the R markdown files and all other
+fully reproducible manuscript in R markdown format. R markdown files and all other
 models are available from
 [our repositories](https://github.com/dennishazelett/TTmanu) in the following
-locations on the distributed version control site GitHub
+locations on the distributed version control site GitHub.
 
@@
 
-To understand the molecular mechanism of cancer at a pathway level, our pathway
-analysis approach focused on Reactome (https://reactome.org/), a pathway knowledge-based 
-database. The mapping files of ENSEMBL to pathways, pathway hierarchy relationship,
-and complexes to top pathways were downloaded from the download page 
-(https://reactome.org/download-data). Using the data, we imposed pathway criteria to
-meet the basic cellular processes and biochemical pathways: (1) human-derived pathways
-(_“HSA”_) (2) limited to grandchild node for each parent pathway (_e.g._ ‘Beta-catenin 
+To understand the molecular mechanism of cancer at a pathway level, we used
+Reactome (https://reactome.org/), a knowledge-based pathway
+database. The mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships,
+and complexes to top pathways were downloaded from https://reactome.org/download-data. Using these data, we imposed pathway criteria to
+define basic cellular processes and biochemical pathways: (1) human-derived pathways
+( _“HSA”_) (2) limited to grandchild node for each parent pathway ( _e.g._ ‘Beta-catenin 
 independent WNT signaling’ in ‘Signal Transduction’) (3) exclusion of pathways in the
-parent pathway: Disease, Muscle contraction, and Reproduction or pathway names that 
-include any of the following keywords: disease, defect, cancer, oncogenic, disorder, 
-mutant, loss, infection, bacteria, or listeria. While some of the excluded pathways 
+parent pathway: "Disease", "Muscle contraction", and "Reproduction" or pathway names that 
+include any of the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder", 
+"mutant", "loss", "infection", "bacteria", or "listeria". While some of the excluded pathways 
 have been shown to play an important role in cancer, they are highly specialized 
-(_e.g._ ‘PI3K/AKT Signaling in Cancer’). Additionally, for most of the excluded pathways,
-a neutral version pathway of the pathway exist (_e.g._ ‘PIP3 activates Akt signaling’). 
+( _e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for most of the excluded pathways,
+a neutral version pathway of the pathway exists ( _e.g._ "PIP3 activates Akt signaling"). 
 Finally, we mapped the 18,577 Ensembl IDs from the TCGA dataset to the highly selected
@@
 
-We filtered likely erroneous mutations due to transcription coupled repair. Our approach
-was to determine the status of each gene (expressed or not expressed) per tissue type in
-order to exclude gene mutations that we have evidence for no expression. To do this, we 
-obtained the TCGA RNA sequencing data adjusted for batch effect dataset (https://pancanatlas.xenahubs.net).
-Using the data, we removed the genes and tumor samples that were not included in our analysis, 
-grouped the tumor samples by tissue type and computed the mean expression value for each 
-gene. A minimum threshold of 10 transcripts per million was set for expressed genes based 
-on an inflection point observed when plotting the mean expression values of genes ranked 
-by increasing expression values for each tissue type. Genes that did not meet this threshold 
-were considered not expressed. This operation produced a lookup table for gene expression
-status in each tumor sample for 18,127 genes.
+We filtered likely erroneous mutations due to transcription coupled repair. Our
+approach was to determine the status of each gene ( _i.e._ expressed or not
+expressed) in each tissue type in order to exclude gene mutations low expressed
+enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
+effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
+the genes and tumor samples that were not included in our analysis, grouped the
+tumor samples by tissue type and computed the mean expression value for each
+gene. A minimum threshold of 10 transcripts per million was set for expressed
+genes based on an inflection point observed when plotting the mean expression
+values of genes ranked by expression in each tissue type. Genes that did not
+meet this threshold were considered not expressed. This operation produced a
+lookup table for gene expression status in each tumor sample for 18,127 genes.
 
@@
 We calculated the enrichment of pathways in one set of tumors as the relative
-fraction (and probability of observing) of tumors with a mutated gene in the
+fraction of tumors (with estimated uncertainty) with a mutated gene in the
 pathway to all tumors, inclusive of the category of interest. To do this
@@
 tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the $95\%$ range of credible
-differences thus obtained excludes 0 (as a null hypothesis) and the mean
-credible difference was greater than or equal to $20\%$ enrichment, which
+samples. We considered a pathway enriched if the 95% range of credible
+differences thus obtained excludes 0 and the mean
+credible difference was greater than or equal to 30% enrichment, which
 excludes a large number of small differences that are not likely to be
@@
 ```
+
+## Acknowledgements
+
+## Declarations
+
+The authors have nothing to declare. This work was supported
 

### 2020-08-14 text/survival-methods.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

@@
 In order to understand how the diagnosis of cancer changes the expectation for
-longevity of patients, we first began by modeling the longevity of the overall
-population of the US based on CDC life tables. Although childhood cancers exist
-and are tragic, most cancer patients have diagnosis later in life. We modeled
+longevity of patients, we first modeled the longevity of the overall
+population of the U.S. based on CDC life tables. Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes accuracy in the right tail of the distribution. We modeled
 baseline longevity using a risk rate function:
@@
 and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
-is an indicator for whether diagnosis has occurred or not (before diagnosis D(t)
-= 0, after it equals 1). The model essentially assumes that the risk of dying
-per year at age 20 years is $R_{20}$ and that it increases exponentially with a
-constant rate $a$ up to diagnosis, and then with a new rate $ka$
-thereafter. Though the model does extend back to earlier ages, relatively few
+is an indicator for whether diagnosis has occurred or not (before diagnosis $D(t) = 0$, 
+after $D(t) = 1$). The model assumes that the risk of death
+is $R_{20}$ at age 20 years and increases exponentially with a
+constant rate $a$ up to the age at diagnosis. Risk of death assumes a new rate $ka$
+thereafter. Though the model extends to earlier ages, relatively few
 patients were under age 20.
 
-Given this risk per unit time, the probability to die at time $T+dT$ is the
+Given this risk per unit time, the probability of death at time $T+dT$ is the
 probability to survive to time $T$ which is $(1-P(T))$ and then die in the
@@
 probability at 1 and a relatively narrow width, due to the fact that it is a
-tissue specific multiplicative modifying factor, this allows the overall
-magnitude of k to be primarily determined by the $k_{tis}$ value whose prior is
-significantly less constrained so that there is a wider range of dangers across
+tissue specific multiplicative modifying factor. This allows the overall
+magnitude of $k$ to be primarily determined by the $k_{tis}$ value whose prior is
+significantly less constrained so that there is a wider range of risk across
 tissue types.
@@
 
-Using this model, we split the data by Male/Female sex, and by White, vs Black
+Using this model, we split the data by Male/Female sex, and by White, _vs._ Black
 race as these were the main well-identified categories available to us in the
 dataset. Sex is known to be a risk factor for death, with males dying at
-slightly higher rates for all ages of interest in our dataset. Black vs White
+slightly higher rates for all ages of interest in our dataset. Black _vs._ White
 race is an important category due to differences in access to care and other

### 2020-08-14 text/abstract.Rmd
edit abstract

@@
 cellular communication. Using dimension reduction techniques, we assigned tumor
-samples into ten classes distinct from their tissues-of-origin and largely
-free of bias from mutational burden or clinical stage. We showed that these
-tumor classes vary in prognosis by modeling relative risk of death in different
+samples into ten classes distinct from their tissues-of-origin largely free of
+bias from mutational burden or clinical stage. We showed that these tumor
+classes vary in prognosis by modeling relative risk of death in different
 classes and cancer types. We identified class-specific mutations in different
 tissues, demonstrating that tissue-specific signatures contribute to common
-cellular phenotypes. Moreover, driver gene mutations involved in replication fidelity and genome
-stability, such as BRCA1 and Rb1, are equally distributed among our classes,
-contrary to the expectation that such genes are avatars of molecular
-subtype. We also investigated pathways that distinguish between metastatic and
-non-metastatic tumors, and found that most meaningful differences are
-class-specific. Many metastastasis enriched pathways from one class of
-tumors are class-specific pathways from another class, suggesting phenotypic
-convergence, and potentially highlighting key pathways to focus on
-for prevention of class-specific relapse. In summary, our
-observations suggest that some genes act as drivers of genome instability in all
-classes, while different sets of class-specific genes confer increased fitness
-in a combinatorial fashion, eventually converging on one or more end-stage
-disease phenotypes. Thus, we present a coherent view of global tumor biology,
-and explain how common cellular dysfunction might arise from tissue-specific
-mutations.
+cellular phenotypes. Moreover, driver gene mutations involved in replication
+fidelity and genome stability, such as BRCA1 and Rb1, are equally distributed
+among our classes, contrary to the expectation that such genes are avatars of
+molecular subtype. We also investigated pathways that distinguish between
+metastatic and non-metastatic tumors, and found that most meaningful differences
+are class-specific. Many metastasis enriched pathways from one class of tumors
+are class-specific pathways from another class, suggesting phenotypic
+convergence, and potentially highlighting key pathways to focus on for
+prevention of class-specific relapse. Taken as a whole, our observations suggest
+that some genes act as drivers of the cancer phenotype in all classes, while
+specific pathway disruptions confer increased fitness in a combinatorial
+fashion, producing a limited number of distinct molecular class phenotypes.
+Eventually these classes converge on one or more end-stage disease phenotypes.
+Thus, we present a coherent view of global tumor biology, and explain how common
+cellular dysfunction might arise from tissue-specific mutations.

### 2020-08-14 text/manuscript.Rmd
minor updates to get dan's rmd working

--- 2020-08-14 text/manuscript.Rmd
@@
   library(flextable),
-  library(plot3D)
+  library(plot3D),
+  library(ggfortify)
 ))

### 2020-08-14 text/discussion.Rmd

@@
 
-One of the biggest hurdles in cancer research is the size of the search space;
+One of the biggest hurdles in cancer research is the sparsity of data; 
 ~20,000 protein-coding genes is comparable with the number of tumor samples,
 even with multiple mutations per sample. And that is not accounting for the fact
-that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
+that ~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
 elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@@
 biases. We also filtered genes on the basis of pathway membership in biochemical
-pathways, excluding curated genesets related to diseases, syndromes, or classes
-of proteins ( _e.g._ transcription factors) which are potentially problematic
+pathways, excluding curated gene sets related to diseases, syndromes, or classes
+of proteins with shared catalytic activity or conserved domains ( _e.g._ transcription factors) which are potentially problematic
 [@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
@@
 likely occur at low enough frequencies as to be indistinguishable from
-background. Previous attempts to capture these associations better using
+background. Previous attempts to capture these associations using
 Bayesian networks have yielded important new insights into this problem and new
@@
 with traditional analysis, supporting the case that pathway-centric approaches
-like ours have the potential to increase statistical power.
+have the potential to increase statistical power.
 
@@
 clusters, regardless of the cancer type (**figure 2F, 3A** and **supplementary
-figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
-mutant cluster 8, and Kidney Chromophobe in cluster 1, see **supplementary
+figure S1**). Some cancer types ( _e.g._ Pancreatic:PAAD in the RAS
+mutant cluster 8, and Kidney Chromophobe:KICH in cluster 1, see **supplementary
 figure S1**) are unevenly distributed, though we could not identify any cancers
@@
 Our results illustrate how unique combinations of mutations in pan-cancer driver genes with tissue-specific pathway disruptions result in common categories when viewed at the level of the pathway knowledge-base.
-For example, in table 1 we find that `snRNP assembly`, one of the unique driver pathways of cluster 5, the top gene in breast cancer is represented by TPR (mutated in 26% of samples).
+For example, we find that for "snRNP assembly", one of the unique driver pathways of cluster 5, the top gene in breast cancer is represented by TPR (mutated in 26% of samples).
 By contrast the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of samples, respectively).
@@
 None of these genes is 100% mutated in any cluster however, and therefore it is the combinations with secondary mutations, many of which are tissue-specific, that round out the complement of pathways in all clusters.
-These top genes are found in most of the clusters (supplemental figure S5), and in spite of the fact that they contribute to many cluster-specific pathways.
+Some top genes are found in most of the clusters (supplemental figure S5), in spite of the fact that they contribute to many cluster-specific pathways.
 This can only be explained as a result of other less common driver genes complementing the pathway disruption of the unique combination of driver genes that are disrupted in each tumor, and we speculate that many of these are relatively tissue-specific in terms of their sensitivity to mutation.
@@
 Finally, we must remark on the limitations of our work exemplified in two of the
-classes, 1 and 10, for which we did not find distinctive associations with
-pathways (**figure 5B, C**). Class 1 has a relatively low proportion of mutated
+classes, 1 and 10, for which we did not find many distinctive associations with
+pathways (**figure 5B, C**). Class 1 had a relatively low proportion of mutated
 pathways across the board, although it is broadly enriched in many of the same
@@
 virtually no pathway is unaffected. It seems likely that a significant fraction
-of the samples flagged as “mutant” for each pathway are actually unaffected,
-consistent with those samples have low information content. This could be
-addressed with application of ever-increasingly sophisticated bioinformatics
-algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
+of the “mutant” samples for each pathway are burdened with excess passenger mutations. This could be
+addressed with application of ever-increasingly sophisticated filtering likely passenger mutations [ _e.g._ @spliceAI;
 @deep-learning-clinical-variants]. In the future, we hope to incorporate these
@@
 expected survival (**figure 7**). Contrary to what one might expect at the
-outset, we found that models assuming different survival based on class _and_
-cancer type fit better than models in which these variables are independent
+outset, we found that models assuming different survival based on class and
+cancer type fit better than models in which they are independent
 factors. However, considering that within each cancer type there are different
 clinical standards of care, and even within classes of drugs the preferred
-treatment can very between cancers, it makes sense in hindsight that they should
-all be different. Contrast the situation with ovarian _vs._ breast cancer, which
+treatment can very between cancers, it makes sense that we observe tissue-specific class effects. Contrast the situation with ovarian _vs._ breast cancer, which
 are both hormonally driven cancers, for example. Ovarian cancer has but one main
@@
 the diversity drug classes and treatments (including non-chemical treatments
-such as surgery and palliative care), we lack sufficient power in the current
-study to explore these variables in the TCGA data. It is our hope that more
-focused future studies will help to distinguish between treatment-specific
-effects or survival given different pathway disruption classes.
+such as surgery and palliative care), we lack sufficient power 
+to explore these variables in the TCGA data. It is our hope that 
+future studies will help to distinguish between treatment-specific
+effects on survival given different pathway disruption classes.
 
@@
 disrupted in all or most of the defined classes (**supplementary figure S1**)
-and often are the result of aberrations involving common driver genes such as
+and are frequently the result of aberrations involving common driver genes such as
 BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
@@
 of tumor heterogeneity. Consortium samples are likely to contain contamination
-from support tissue, stroma, inflammatory, and immune cells of the innate and
+from support tissue, stroma, inflammatory cells, immune cells of the innate and
 adaptive immune systems, all potentially harboring cancer supporting mutations
@@
 cluster specific analysis, suggesting that signal is diluted when the samples
-are analyzed together, and supporting the view that metastasis has
+are analyzed together, and supporting the view that, within the parameters of our analysis, metastasis has
 class-specific requirements. Since the number of metastatic samples is
@@
 favors the latter interpretation, and suggests a mechanism of complementarity,
-in which newly acquired mutations result in similar classes converging on a
-single deadly phenotype that captures most of the critical features of end-stage
+in which newly acquired mutations result in similar classes converging on
+one or more deadly phenotypes that capture most of the critical features of end-stage
 cancer. One explanation for this could be that tumors must accumulate mutations
@@
 It may be that the primary effects of mutations in protein and RNA metabolism in
-clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
+clusters 3 and 5 point to currently underappreciated mechanisms impacting hallmark
 pathways. Clusters 3 and 5 are also marked by a decidedly different set of

### 2020-08-14 text/survival.Rmd

@@
 ethnicity. Our model assumes that at birth each person has a small initial
-probability per unit time to die, and that this probability per unit time of
-death conditional on age, grows exponentially in time at some baseline rate that
+probability per unit time to die, and that this probability 
+grows exponentially in time at some baseline rate that
 matches the observed CDC data. The rate of growth in risk is assumed to jump to
@@
 
-To understand the model, see **Figure 7A** where we show simulated survival time
+To understand the model, we show in **Figure 7A** simulated survival time
 curves for four related groups. We start with a cohort of women aged 30 and show
-their expected survival (blue). By setting their cancer rate multiplier to k=2
+their expected survival (blue). By setting their cancer rate multiplier to $k=2$
 we can simulate the effect of a moderately deadly cancer (red). Compare this to
@@
 blue). Immediately the survival curve changes due to the mixture of ages,
-without any malignancy. Adding a k=2 malignancy further reduces expected
-lifespan (dark red). We can see that for different tissues simply due to the
-change in distribution of age-at-diagnosis we should expect equally deadly
+without any malignancy. Adding a $k=2$ malignancy further reduces expected
+lifespan (dark red). We can see that for different tissues, simply due to the
+change in distribution of age-at-diagnosis, we should expect equally deadly
 cancers to have *different* survival curves. The model takes this into account
@@
 The degree to which a given cancer accelerates aging can be determined by the
-ktis multiplier (**Figure 7F**) multiplied by the class specific modifier
+$k_{tis}$ multiplier (**Figure 7F**) multiplied by the class specific modifier
 (**Figure 7G**). Looking at the tissue specific multiplier, the least deadly was
@@
 risk multiplier of between 15 and 22 relative to the background risk of death in
-the population. This undoubtedly is due to the fact that relatively young
+the population. This undoubtedly is influenced by the fact that relatively young
 patients are affected by GBM and that it is extremely deadly. Among the
@@
 lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
-indestinguishable from the prior, reflecting that either there were too few
+indistinguishable from the prior, reflecting that either there were too few
 mortalities in the data to make an estimate (as expected for PRAD and THCA) or
-two few samples period. We did not observe class-specific trends that held true
+two few samples, period. We did not observe class-specific trends that held true
 across cancer types, which could result from different cancers having different
@@
 widely the patients vary across tissue types as well as the different drug and
-surgical treatments typical used for each tissue.
+surgical treatments typically used for each tissue.

### 2020-08-14 code/survival/survfig.Rmd
Updates survival plots with new color scheme + key

@@
 
-p1 = ggplot(data.frame(r=patientsurvivalsbkd,c=patientsurvivalsk2,rm = mixedsurvwfk1, cm=mixedsurvwfk2, status=rep(1,length(patientsurvivalsbkd)))) +
-geom_km(aes(time=r,status=status),color="blue") +
-geom_km(aes(time=c,status=status),color="red") +
-geom_km(aes(time=rm,status=status),color="darkblue") +
-geom_km(aes(time=cm,status=status),color="darkred") +
-annotate("text",rep(90,4),c(.9,.7,.5,.3),color=c("blue","red","darkblue","darkred"),label=c("Random 30 year old women","Random 30 year old women cancer k=2","Random uniform(30,70) year old women","Random uniform(30,70) year old women k=2"),hjust="right")+
-labs(title="Simulated patients",x="Time (years)",y="Fraction Surviving")
+p1 = data.frame(
+    r = patientsurvivalsbkd,
+    c = patientsurvivalsk2,
+    rm = mixedsurvwfk1,
+    cm = mixedsurvwfk2,
+    status = rep(1, length(patientsurvivalsbkd))
+) %>%
+    rownames_to_column(var = "index") %>%
+    pivot_longer(-c(index, status), values_to = "time", names_to = "condition") %>%
+    ggplot(aes(time = time, color = condition, status = status)) +
+    geom_km() +
+    scale_color_manual(
+        values = c(r = "#56b4e9", c = "#e69f00", rm = "#0072b2", cm = "#d55e00"),
+        breaks = c("r", "c", "rm", "cm"),
+        labels = c("Random 30 year old women", "Random 30 year old women cancer k=2", "Random uniform(30,70) year old women", "Random uniform(30,70) year old women k=2")
+    ) +
+    theme_classic() +
+    theme(legend.position = c(.8, .75))
 
@@
     strip.background = element_blank(),
-    legend.position = "none",
+    legend.key.height = unit(0.4, "lines"),
+    legend.title = element_blank(),
+    #legend.position = "none",
     title = element_text(size=10)

### 2020-08-14 code/survival/kmcurve.R

@@ -71,10 +71,23 @@ cancerdat = cancerdat[!is.na(cancerdat$finaltime),]
 
 
 plotkmcomp = function(sim,ref,act,actstatus){
-    ggplot(data.frame(stime=sim,rtime=ref,acttime=act,status=rep(1,length(ref)),actstatus=actstatus)) +
-        geom_km(aes(time=stime,status=status),color="red") +
-        geom_km(aes(time=rtime,status=status),color="blue") +
-        geom_km(aes(time=acttime,status=actstatus),color="darkgreen")
+    data.frame(
+        stime = sim,
+        rtime = ref,
+        acttime = act,
+        status = rep(1, length(ref)),
+        actstatus = actstatus
+    ) %>%
+        pivot_longer(-c(status, actstatus),
+            values_to = "time",
+            names_to = "condition"
+        ) %>%
+            mutate(status = case_when(condition != "acttime" ~ as.integer(status),
+            condition == "acttime" ~ as.integer(actstatus))) %>%
+        ggplot(aes(time = time, color = condition, status = status)) +
+        geom_km() +
+        theme_classic() +
+        theme(legend.position = c(.8, .75))
 }
     
 
@@ -103,17 +116,31 @@ plotcancertype = function(cancerdat,i){
 
     survyrswm = sapply(agesm,function(a) tryCatch(simdeath(r20wm,ageratewm,a,kmean)/365,error= function(e) {0}))
     survyrswmref = sapply(agesm,function(a) tryCatch(simdeath(r20wm,ageratewm,a,1)/365,error= function(e) {0}))
-
-    p1 = tryCatch(plotkmcomp(survyrswf,survyrswfref,women$finaltime/365,women$finalstatus)+
-                  coord_cartesian(xlim=c(0,80)) +
-                  labs(x="Survival Time (years past diagnosis)",
-                       y="Probability",
-                       title=sprintf("Survival for %s simulated white female patients",i)) +
-                  annotate("text",x=80, y=.9,label="Simulated General Population k=1",color="blue",hjust="right") +
-                  annotate("text",x=80, y=.7,label=sprintf("Simulated Patient Population k=%.1f",kmean),color="red",hjust="right")+
-                  annotate("text",x=80, y=.5,label=sprintf("Actual Patient Population",kmean),color="darkgreen",hjust="right")                  
-                 ,
-                  error= function(e) {print(e);ggplot()})
[…]

### 2020-08-14 code/survival/survfig.Rmd
Collect legends for cancer specific km plots

@@
 
+kmlayout = "
+AB
+AB
+AB
+CD
+CD
+CD
+EE
+"
+kmlegend <- theme(legend.direction = "horizontal")
+
 print(
-((p1+mytheme)/((pstad[[1]]+mytheme)+(pgbm[[1]]+mytheme) +
-(ppaad[[1]]+mytheme) + (thca[[1]] + mytheme)))/((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
+((p1+mytheme)/((pstad[[1]]+mytheme + kmlegend)+(pgbm[[1]]+mytheme + kmlegend) +
+(ppaad[[1]]+mytheme + kmlegend) + (thca[[1]] + mytheme + kmlegend) + guide_area() + plot_layout(design = kmlayout, guides = "collect")))/((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
 )

### 2020-08-14 code/survival/kmcurve.R

@@ -121,7 +121,8 @@ plotcancertype = function(cancerdat,i){
         labs(
             x = "Survival Time (years past diagnosis)",
             y = "Probability",
-            title = sprintf("Survival for %s simulated white female patients", i)
+            title = sprintf("Survival for %s simulated white female patients", i),
+            subtitle = sprintf("Simulated Patient Population; k = %.1f", kmean)
         ) +
         scale_color_manual(
             values = c(
@@ -132,7 +133,7 @@ plotcancertype = function(cancerdat,i){
             breaks = c("rtime", "stime", "acttime"),
             labels = c(
                 "Simulated General Population; k = 1",
-                sprintf("Simulated Patient Population; k = %.1f", kmean),
+                "Simulated Patient Population",
                 "Actual Patient Population"
             )
         ),

### 2020-08-14 code/supplemental_v2.Rmd
Add ST1

@@
   library(plot3D),
-  library(flextable)
+  library(flextable),
+  library(knitr),
+  library(kableExtra)
 ))
@@
 
-```{r enrich-pathway-in-clust, eval = FALSE}
+**Supplementary Table 1. Cluster-specific enriched pathways (effect size >= 0.30).**
+```{r enrich-pathway-in-clust, fig.width= 9}
 ## docx friendly
@@
 
-st1 %>% flextable() %>% 
-  set_caption(caption = "") %>%
-  align(part = "all") %>% 
-  width(., width = c(4,0.75,1,1)) %>% 
-  fontsize(size = 10, part = "body")
+# st1 <- st1 %>% flextable() %>% 
+#   set_caption(caption = "") %>%
+#   align(part = "all") %>% 
+#   width(., width = c(4,0.75,1,1)) %>% 
+#   fontsize(size = 10, part = "body")
+# 
+# knitr::knit_print(st1)
+names(st1) <- c("Pathway", "Cluster", "Effect Size", "-log(p)")
+ st1 %>%  knitr::kable(format = "latex", longtable = TRUE, align = c("l", rep("c",3))) %>%
+   kable_styling(latex_options=c("scale_down", "repeat_header"), full_width = FALSE, font_size = 12) %>%
+   column_spec(1, color = "black", width = "4in")
 ```
+
+\newpage
 

### 2020-08-14 code/supplemental_v2.Rmd
Update supplemental s2 with key, inprogress s4 -does not run using loop

@@
 }
+
 cancer_w_subtypes <- names(cancer_w_subtypes)
@@
                colvar = NULL))
-  legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-#legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-#legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
+  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
   title(xlab="UMAP 2", line=0)
@@
 
-### Supplementary Figure S4
-This is the 377 pathway UMAP. I will need to develop a loop that does this.
-
-**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
-pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
-(column) from **Figure 2C of the main text**.
 ```{r supplementary-figure-s4-data-generation, eval = FALSE}
@@
 
-selected_path <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
+selected_path <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
   left_join(., selected_path, by = "sample_id") %>%
-  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
   filter(value != 0)
@@
 
-load("../data/selected_path.Rda") ## load data
-load("../data/gdc_reactome_path_analysis.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+# load("../data/selected_path.Rda") ## load data
+# load("../data/gdc_reactome_path_analysis.Rda")
+# umap_cancertype <- fread("../data/umap_3d_coors.tsv")
+# umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
+# 
+# n_path <- 377 ## number of pathways we want to see
+# 
+# toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
+needed for facet
+# 
+# ## creating layout to plot all 377 pathways
+# figs4_layout <- data.frame(start = seq(from = 1, to = length(toppath), by = 25),
+#                           end = c(seq(from = 25, to = length(toppath), by = 25), length(toppath)))
+# 
+# for(n in 1:nrow(figs4_layout)) {
+#   message(n)
+#   pathway_layout <- figs4_layout %>% .[n,] ## getting the layout for selecting pathways
+#   pathway_layout <- toppath[pathway_layout$start:pathway_layout$end] ## filtering those pathways from toppath
+#   filtered_path <- selected_path %>% filter(variable %in% pathway_layout) %>% ## filtering our data to only include those pathways
+#     mutate(variable = factor(variable, levels = c(pathway_layout)))
+#   ggplot_umap <- ggplot(filtered_path, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
+#     geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
+#     geom_point(size = 0.25) +
+#     theme_classic() +
+#     coord_fixed() +
+#     xlab("UMAP 1") + ylab("UMAP 2") +
+#     scale_colour_manual(values = rep("firebrick3", 25)) +
+#     theme(axis.text.x = element_blank(),
+#       axis.text.y = element_blank(),
+#       axis.title.x = element_blank(),
+#       axis.title.y = element_blank(),
+#       axis.ticks = element_blank(),
+#       axis.line = element_blank(),
+#       strip.text = element_text(size = 5),
+#       legend.position = "none") +
+#     facet_wrap(~ variable,
+#                ncol = 5) ## facet by pathways
+#   new_dfname <- paste0("ggplot_umap_", n)
+#   assign(paste0(new_dfname), value = ggplot_umap)
+# }
+# 
+# save(ggplot_umap_1,ggplot_umap_2,ggplot_umap_3,ggplot_umap_4,ggplot_umap_5,ggplot_umap_6,ggplot_umap_7,ggplot_umap_8,ggplot_umap_9,ggplot_umap_10,ggplot_umap_11,ggplot_umap_12,ggplot_umap_13,ggplot_umap_14,ggplot_umap_15,ggplot_umap_16, file = "../data/fig_s4_ggplot.Rda")
+```
 
-n_path <- 377 ## number of pathways we want to see
+```{r figure-s4-plot, fig.width= 9}
+load("../data/selected_path.Rda")
+load("../data/effect_heatmap.Rda") ## load effect heatmap data
+pathorder <- as.character(levels(effect_heatmap$path_name)) ## getting col order
 
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
+startingpoint <- seq(1, 377, by = 15)
+startingpointlist <- list()
+for(n in startingpoint) {
+  start1 <- n
+  end1 <- start1 + 5
+  start2 <- end1 + 1
+  end2 <- start2 + 5
+  start3 <- end2 + 1
+  end3 <- start3 + 2
+  startingpointlist[[n]] <- data.frame(start1 = start1,
+                                       end1 = end1,
+                                       start2 = start2,
+                                       end2 = end2,
+                                       start3 = start3,
+                                       end3 = end3)
+}
+startingpointlist <- as.data.frame(do.call(rbind, startingpointlist))
 
-## creating layout to plot all 377 pathways
-figs4_layout <- data.frame(start = seq(from = 1, to = length(toppath), by = 25),
-                          end = c(seq(from = 25, to = length(toppath), by = 25), length(toppath)))
-
-for(n in 1:nrow(figs4_layout)) {
-  message(n)
-  pathway_layout <- figs4_layout %>% .[n,] ## getting the layout for selecting pathways
-  pathway_layout <- toppath[pathway_layout$start:pathway_layout$end] ## filtering those pathways from toppath
-  filtered_path <- selected_path %>% filter(variable %in% pathway_layout) %>% ## filtering our data to only include those pathways
-    mutate(variable = factor(variable, levels = c(pathway_layout)))
-  ggplot_umap <- ggplot(filtered_path, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-    geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-    geom_point(size = 0.25) +
-    theme_classic() +
-    coord_fixed() +
-    xlab("UMAP 1") + ylab("UMAP 2") +
-    scale_colour_manual(values = rep("firebrick3", 25)) +
-    theme(axis.text.x = element_blank(),
-      axis.text.y = element_blank(),
-      axis.title.x = element_blank(),
-      axis.title.y = element_blank(),
-      axis.ticks = element_blank(),
-      axis.line = element_blank(),
-      strip.text = element_text(size = 5),
-      legend.position = "none") +
-    facet_wrap(~ variable,
-               ncol = 5) ## facet by pathways
-  new_dfname <- paste0("ggplot_umap_", n)
-  assign(paste0(new_dfname), value = ggplot_umap)
+s4plot <- function(x) {
+  a <-  x[,"start1"]
+  b <- x[,"end1"]
+  c <- x[,"start2"]
+  d <- x[,"end2"]
+  e <- x[,"start3"]
+  f <- x[,"end3"]
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in pathorder[x[,"start1"]:x[,"end1"]]) {
+  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
+  pathname <- pathname[,1] ## get the character vectors
+  with(selected_path %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
+path1 <- gsub("NA", "", path1)
+path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
+path2 <- gsub("NA", "", path2)
+legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
+legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
 }
 
-save(ggplot_umap_1,ggplot_umap_2,ggplot_umap_3,ggplot_umap_4,ggplot_umap_5,ggplot_umap_6,ggplot_umap_7,ggplot_umap_8,ggplot_umap_9,ggplot_umap_10,ggplot_umap_11,ggplot_umap_12,ggplot_umap_13,ggplot_umap_14,ggplot_umap_15,ggplot_umap_16, file = "../data/fig_s4_ggplot.Rda")
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in pathorder[x[,"start2"]:x[,"end2"]]) {
+  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
+  pathname <- pathname[,1] ## get the character vectors
+  with(selected_path %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
+path1 <- gsub("NA", "", path1)
+path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
+path2 <- gsub("NA", "", path2)
+legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
+legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
+}
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in pathorder[x[,"start3"]:x[,"end3"]]) {
+  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
+  pathname <- pathname[,1] ## get the character vectors
+  with(selected_path %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
+path1 <- gsub("NA", "", path1)
+path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
+path2 <- gsub("NA", "", path2)
+legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
+legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
+}
+}
+
+a = 1
+b = 6
+c = 7
+d = 12
+e = 13
+f = 15
+
+for(p in 1:nrow(startingpointlist)-1) {
+  data <- startingpointlist[p,]
+  s4plot(x = data)
+}
+
+# par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
+# par(mar = c(2,2,0,0))
+# for(n in genes[7:12]) {
+# with(selected_mut %>% 
+#        filter(variable == n),
+#      scatter3D(plot_y, -plot_z, plot_x, 
+#                                 bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+#                                 theta = 0, phi = 65, scale = F,
+#                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+#                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+#                                 colvar = NULL))
+# title(xlab="UMAP 2", line=0)
+# legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+# }
+# par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+# par(mar = c(2,2,0,0))
+# for(n in genes[13:15]) {
+# with(selected_mut %>% 
+#        filter(variable == n),
+#      scatter3D(plot_y, -plot_z, plot_x, 
+#                                 bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+#                                 theta = 0, phi = 65, scale = F,
+#                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+#                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+#                                 colvar = NULL))
+# title(xlab="UMAP 2", line=0)
+# legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+# }
+
 ```
+**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
+pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
+(column) from **Figure 2C of the main text**.
 

### 2020-08-14 code/survival/survfig.Rmd
Update supplemental s2 with key, inprogress s4 -does not run using loop

--- 2020-08-14 code/survival/survfig.Rmd
@@
-```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=9, fig.width=6.5}
+#### Figure 7
+```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=9, fig.width=9}
 
@@
 
+kmlayout = "
+AB
+AB
+AB
+CD
+CD
+CD
+EE
+"
+kmlegend <- theme(legend.direction = "horizontal")
+
+print(
+((p1+mytheme)/((pstad[[1]]+mytheme + kmlegend)+(pgbm[[1]]+mytheme + kmlegend) +
+(ppaad[[1]]+mytheme + kmlegend) + (thca[[1]] + mytheme + kmlegend) + guide_area() + plot_layout(design = kmlayout, guides = "collect"))) + plot_annotation(tag_levels = "A")
+)
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 8
+```{r, message = FALSE, fig.width = 9}
 mylayout = '
-AAAAAAAA
-BBBBBBBB
-BBBBBBBB
-CCCCCCCC
-CCCCCCCC
-CCCCCCCC
+ABB
+ABB
+#BB
 '
 
-print(
-((p1+mytheme)/((pstad[[1]]+mytheme)+(pgbm[[1]]+mytheme) +
-(ppaad[[1]]+mytheme) + (thca[[1]] + mytheme)))/((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
-)
-
+((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
 ```

### 2020-08-14 text/manuscript.Rmd
Update supplemental s2 with key, inprogress s4 -does not run using loop

@@
 #### Figure 1
-```{r child='../code/figure1v2.Rmd'}
-#
+```{r }
+#child='../code/figure1v2.Rmd'
 ```
@@
 #### Figure 2
-```{r child='../code/figure2v2.Rmd'}
-#
+```{r }
+#child='../code/figure2v2.Rmd'
 ```
@@
 #### Figure 3
-```{r child = '../code/figure4v2.Rmd'}
-#
+```{r }
+#child = '../code/figure4v2.Rmd'
 ```
@@
 #### Figure 4
-```{r child = '../code/figure3v2.Rmd'}
-#
+```{r }
+#child = '../code/figure3v2.Rmd'
 ```
@@
 #### Figure 5
-```{r child = '../code/figure2.Rmd'}
-#
+```{r }
+#child = '../code/figure2.Rmd'
 ```
@@
 #### Figure 6
-```{r child = '../code/figure4.Rmd'}
-#
+```{r }
+#child = '../code/figure4.Rmd'
 ```
-#### Figure 7
+
 ```{r, results='asis'}
@@
 #### Table 1
-```{r child = '../code/table2.Rmd'}
-#
+```{r }
+#child = '../code/table2.Rmd'
 ```

### 2020-08-15 code/supplemental_v2.Rmd

@@
   f <- x[,"end3"]
+  path01<- pathorder[a:b]
+  path02<- pathorder[c:d]
+  path03<- pathorder[e:f]
 par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
 par(mar = c(2,2,0,0))
-for(n in pathorder[x[,"start1"]:x[,"end1"]]) {
+for(n in path01) { #pathorder[x[,"start1"]:x[,"end1"]]
   pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
@@
 
-a = 1
-b = 6
-c = 7
-d = 12
-e = 13
-f = 15
+data <- startingpointlist[1,]
+s4plot(x = data)
+data <- startingpointlist[2,]
+s4plot(x = data)
+data <- startingpointlist[3,]
+s4plot(x = data)
+data <- startingpointlist[4,]
+s4plot(x = data)
+data <- startingpointlist[5,]
+s4plot(x = data)
+data <- startingpointlist[6,]
+s4plot(x = data)
+data <- startingpointlist[7,]
+s4plot(x = data)
+data <- startingpointlist[8,]
+s4plot(x = data)
+data <- startingpointlist[9,]
+s4plot(x = data)
+data <- startingpointlist[10,]
+s4plot(x = data)
+data <- startingpointlist[11,]
+s4plot(x = data)
+data <- startingpointlist[12,]
+s4plot(x = data)
+data <- startingpointlist[13,]
+s4plot(x = data)
+data <- startingpointlist[14,]
+s4plot(x = data)
+data <- startingpointlist[15,]
+s4plot(x = data)
+data <- startingpointlist[16,]
+s4plot(x = data)
+data <- startingpointlist[17,]
+s4plot(x = data)
+data <- startingpointlist[18,]
+s4plot(x = data)
+data <- startingpointlist[19,]
+s4plot(x = data)
+data <- startingpointlist[20,]
+s4plot(x = data)
+data <- startingpointlist[21,]
+s4plot(x = data)
+data <- startingpointlist[22,]
+s4plot(x = data)
+data <- startingpointlist[23,]
+s4plot(x = data)
+data <- startingpointlist[24,]
+s4plot(x = data)
+data <- startingpointlist[25,]
+s4plot(x = data)
 
-for(p in 1:nrow(startingpointlist)-1) {
-  data <- startingpointlist[p,]
-  s4plot(x = data)
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in pathorder[376:377]) { 
+  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
+  pathname <- pathname[,1] ## get the character vectors
+  with(selected_path %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
+path1 <- gsub("NA", "", path1)
+path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
+path2 <- gsub("NA", "", path2)
+legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
+legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
 }
-
-# par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
-# par(mar = c(2,2,0,0))
-# for(n in genes[7:12]) {
-# with(selected_mut %>% 
-#        filter(variable == n),
-#      scatter3D(plot_y, -plot_z, plot_x, 
-#                                 bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-#                                 theta = 0, phi = 65, scale = F,
-#                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-#                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-#                                 colvar = NULL))
-# title(xlab="UMAP 2", line=0)
-# legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-# }
-# par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-# par(mar = c(2,2,0,0))
-# for(n in genes[13:15]) {
-# with(selected_mut %>% 
-#        filter(variable == n),
-#      scatter3D(plot_y, -plot_z, plot_x, 
-#                                 bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-#                                 theta = 0, phi = 65, scale = F,
-#                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-#                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-#                                 colvar = NULL))
-# title(xlab="UMAP 2", line=0)
-# legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-# }
-
 ```
@@
 (column) from **Figure 2C of the main text**.
-
-```{r supplementary-figure-s4, message = FALSE, echo = FALSE, out.width = '90%', fig.align= 'center', eval = FALSE}
-load("../data/fig_s4_ggplot.Rda")
-
-ggplot_umap_1
-ggplot_umap_2
-ggplot_umap_3
-ggplot_umap_4
-ggplot_umap_5
-ggplot_umap_6
-ggplot_umap_7
-ggplot_umap_8
-ggplot_umap_9
-ggplot_umap_10
-ggplot_umap_11
-ggplot_umap_12
-ggplot_umap_13
-ggplot_umap_14
-ggplot_umap_15
-ggplot_umap_16
-
-# ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
-#   geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
-#   geom_point(size = 0.5) +
-#   theme_classic() +
-#   coord_fixed() +
-#   xlab("UMAP 1") + ylab("UMAP 2") +
-#   scale_colour_manual(values = rep("firebrick3", 20)) +
-#   theme(axis.text.x = element_blank(),
-#     axis.text.y = element_blank(),
-#     axis.title.x = element_blank(),
-#     axis.title.y = element_blank(),
-#     axis.ticks = element_blank(),
-#     axis.line = element_blank(),
-#     legend.position = "none") +
-#   facet_wrap(~ factor(variable, levels = toppath),
-#                       ncol = 5) ## facet by pathways
-
-```
 
@@
 ## class 7 and 8 distinguishing genes
-genes <- c("PIK3CA", "KRAS", "TP53")
+genes <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","PIK3CA","CDKN2C","PTEN","RB1","SMARCA4","TP53","VHL")
 
@@
 ```{r supplemental-s5-plot, fig.width=9}
-umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
-
-umap_cancertype$clust_knn <- as.character(umap_cancertype$clust_knn)
-
-clusters178 <- umap_cancertype %>% 
-       mutate(color = case_when(clust_knn %in% c("1", "7", "8") ~ clust_knn,
-                                  TRUE ~ "Other"))
-
-I_outlines <- c("black", "black", "black", "#00000000")
-names(I_outlines) <- c("1", "7", "8", "Other")
-
-I_colors <- c("gold2", "gray51", "springgreen4", "#00000011")
-names(I_colors) <- c("1", "7", "8", "Other")
+# umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
+# 
+# umap_cancertype$clust_knn <- as.character(umap_cancertype$clust_knn)
+# 
+# clusters178 <- umap_cancertype %>% 
+#        mutate(color = case_when(clust_knn %in% c("1", "7", "8") ~ clust_knn,
+#                                   TRUE ~ "Other"))
+# 
+# I_outlines <- c("black", "black", "black", "#00000000")
+# names(I_outlines) <- c("1", "7", "8", "Other")
+# 
+# I_colors <- c("gold2", "gray51", "springgreen4", "#00000011")
+# names(I_colors) <- c("1", "7", "8", "Other")
+# 
+# load("../data/s5_data.Rda")
+# genes <- as.character(unique(s5_data$variable))
+# 
+# par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+# par(mar = c(2,2,0,0))
+# plot.new()
+# with(clusters178, 
+#      scatter3D(plot_y, -plot_z, plot_x, 
+#                                 bg = I_colors[color], col = I_outlines[color],
+#                                 pch = 21, cex = 0.8, lwd = 0.2,
+#                                 theta = 0, phi = 65, scale = F,
+#                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+#                                 colvar = NULL))
+# legend("top", inset = c(0.05), legend=names(I_colors[c(1:3)]), col=I_colors[c(1:3)], pch = 20, bty = "n", horiz = T, cex = 1.2)
+# title(xlab="UMAP 2", line=0)
+# plot.new()
 
 load("../data/s5_data.Rda")
-genes <- as.character(unique(s5_data$variable))
+genes <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","PIK3CA","CDKN2C","PTEN","RB1","SMARCA4","TP53","VHL")
 
@@
 par(mar = c(2,2,0,0))
-plot.new()
-with(clusters178, 
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = I_colors[color], col = I_outlines[color],
-                                pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-legend("top", inset = c(0.05), legend=names(I_colors[c(1:3)]), col=I_colors[c(1:3)], pch = 20, bty = "n", horiz = T, cex = 1.2)
-title(xlab="UMAP 2", line=0)
-plot.new()
-
-for(n in genes) {
+for(n in genes[1:6]) {
 with(s5_data %>% 
@@
 }
+
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(2,2,0,0))
+for(n in genes[7:12]) {
+with(s5_data %>% 
+       filter(variable == n),
+     scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+}
+
 ```
-**Supplementary Figure S5. Distribution of oncogenes on UMAP clusters.** The left panel shows the identity of 
-clusters 1, 7 and 8. The multipanel on the right shows one of the three oncogenes projected onto the tumors of 
-the UMAP from **Figure 2C of the main text**.
+**Supplementary Figure S5.
+Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 2C of the main text**.
 

### 2020-08-15 code/survival/survfig.Rmd

@@
 #### Figure 8
-```{r, message = FALSE, fig.width = 9}
-mylayout = '
-ABB
-ABB
-#BB
-'
+```{r, message = FALSE, fig.width = 3}
+pcanc+mytheme
+```
 
-((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
+```{r, results='asis'}
+cat("\\newpage")
 ```
+
+#### Figure 9
+```{r, message = FALSE, fig.width = 9, fig.height= 9}
+kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
+                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
+                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
+                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
+
+cluster.colours <- c("grey", kelly.colours[c(3:12)])
+names(cluster.colours) <- c(0:10)
+
+kclassplot = function(dfsamps,tisnames){
+	kclnames = grep("k\\[",names(dfsamps), value=TRUE)
+	dfclas = dfsamps[,kclnames]
+	dfclas = stack(dfclas)
+	names(dfclas) = c("value","coefname")
+	dfclas$class = as.factor(as.numeric(gsub("k\\[(.*),(.*)\\]","\\2",dfclas$coefname)))
+	dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
+	ggplot(dfclas,aes_(quote(value),quote(class), fill = quote(class))) + geom_density_ridges() +
+		facet_wrap(. ~ tiss,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
+		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients")
+}
+
+pclass = kclassplot(dfsamps,levels(cancerdat$type))
+
+pclass+mytheme+scale_fill_manual(values = cluster.colours)
+```

### 2020-08-15 text/manuscript.Rmd
Update fig s4, split dan's plot into fig 7-9
Update fig ref in text

--- 2020-08-14 text/manuscript.Rmd
@@
 pancreatic adenocarcinoma (PAAD) was predominantly found in class 8, which is
-distinguished  by KRAS mutations among others (**supplementary figure S3**). 
+distinguished  by KRAS mutations among others (**supplementary figure S5**). 
 But even PAAD comprises tumors from the nine remaining classes, meaning that 
@@
 has nearly every pathway disrupted (**figure 4**). However, outside of class 10 we don't observe
-an obvious trend in the overall mutation burden across the backbone of the UMAP (**figure
-1H**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
+an obvious trend in the overall mutation burden across the backbone of the UMAP. 
+Nonetheless, to test the hypothesis that the molecular-pathway disruption
 classes represent advancement through stages, we projected staging data onto the
@@
 extracellular, intracellular, and immune-related signaling pathways (see
-**figure5C** and **supplementary figure S4**). Cluster 2 had the highest pathway
+**figure 5C** and **supplementary figure S4**). Cluster 2 had the highest pathway
 enrichment levels of the three, having mutations in hedgehog signaling,
@@
 ```
-#### Figure 7
+
 ```{r, results='asis'}

### 2020-08-15 text/survival.Rmd
Update fig ref in text

@@
 The degree to which a given cancer accelerates aging can be determined by the
-$k_{tis}$ multiplier (**Figure 7F**) multiplied by the class specific modifier
-(**Figure 7G**). Looking at the tissue specific multiplier, the least deadly was
+$k_{tis}$ multiplier (**Figure 8**) multiplied by the class specific modifier
+(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
 prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
@@
 than 1 up to about 3 or 4, reflecting that some classes are either less deadly
-or more deadly than other classes within each cancer types (**Figure 7G**). A
+or more deadly than other classes within each cancer types (**Figure 9**). A
 class specific rate of 1 represents typical rate for this tissue type. For

### 2020-08-15 text/fig_legend.Rmd

@@
 
-**Figure 7: **
+**Figure 7: ** Comparison of different cancers, simulated and actual
+results results. A) To understand the risk model, we plot survival of
+four simulated cohorts. Compare survival of a group consisting of
+randomly selected 30 year old women (light blue) to an equal sized
+group of 30 year old women with cancer, k = 2 (light orange). To
+understand how the mix of ages alters the survival curve, we compare a
+randomly selected group of people with ages uniformly distributed from
+30 to 70 (dark blue) and a randomly selected group with uniformly
+distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a
+randomly selected group with correct age distribution for the given
+cancer (blue) to the model predicted survival for the given average k
+(red) and the actual survival for the patients in our dataset (CHOOSE
+COLOR HERE, NOT GREEN). As can be seen, Stomach Cancer (STAD),
+Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
+deadly. The modeled results fit the general trend for the actual
+patients. Thyroid Cancer (THCA) on the other hand has very little if
+any effect on life expectancy, in fact there may be a slight benefit
+which may be associated with a bias in diagnosis towards patients who
+are more health conscious, or higher socioeconomic status, with more
+access to care compared to random population members.
+
+**Figure 8** Tissue specific effects are compared by showing the
+probability density for the posterior value of the k_tis
+multiplier. These values reflect the general deadliness of the cancer
+type across all clusters. GBM, PAAD, and SKCM stand out as having
+values in the 10 to 20 range with a wide degree of uncertainty. Other
+cancers such as Prostate (PRAD) and Thyroid (THCA) show values down
+near or even below 1 suggesting that these populations may have
+healthier lifestyles and more access to care. Most cancer types fall
+in the range 2-7 as might be expected.
+
+**Figure 9** Class specific $k_class$ values shown for each cancer
+type. The overall $k$ value is the product of $k_tis \times
+k_class$. The class specific value shows the relative aggressiveness
+of that particular class within the tissue. For example class 1, 2,
+and 6 Glioblastoma (GBM) are apparently much less deadly than other
+types of GBM. A similar situation can be seen with Pancreatic Cancer
+(PAAD) where 1, 6, and 8 are less deadly compared to classes 7,
+9, 10. For some cancer types there are few data points and little
+ability to estimate values with precision (DLBC and KICH for example).
 

### 2020-08-15 text/discussion.Rmd

@@
 trend in the number of mutations (see **figure 4**), the relative staging
-(**figure 2D**) or metastasis (**figure 2E**); and each class instead was
+(**figure 3D**) or metastasis (**figure 3E**); and each class instead was
 associated with unique combinations of pathways (as in **figures 5B, C**).
@@
 of our pathway disruption classes exhibit cancer-type specific effects on
-expected survival (**figure 7**). Contrary to what one might expect at the
+expected survival (**figure 9**). Contrary to what one might expect at the
 outset, we found that models assuming different survival based on class and

### 2020-08-16 text/discussion.Rmd
edit full document

--- 2020-08-14 text/discussion.Rmd
@@
 
-One of the biggest hurdles in cancer research is the sparsity of data; 
-~20,000 protein-coding genes is comparable with the number of tumor samples,
-even with multiple mutations per sample. And that is not accounting for the fact
-that ~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
+One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
+protein-coding genes is comparable with the number of tumor samples, even with
+multiple mutations per sample. And that is not accounting for the fact that
+~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
 elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@@
 mutations with likely deleterious effects, including missense, nonsense,
-frame-shift, stop-gain, untranslated region and splice site mutations in genes that are actively
-expressed in each cancer type, thus avoiding bias from transcription coupled
-repair. Our method of filtration differs from the "rank-and-cut" method of
-[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
-biases. We also filtered genes on the basis of pathway membership in biochemical
-pathways, excluding curated gene sets related to diseases, syndromes, or classes
-of proteins with shared catalytic activity or conserved domains ( _e.g._ transcription factors) which are potentially problematic
-[@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
-already high in the remaining pathways for some driver genes, and to exclude
-biologically unrelated collections of genes. We do not account for non-coding
-mutations, amplification events or structural variants, though we hope to
-include them in subsequent studies (see below). Our pathway-centric analysis
-also allows us to circumvent the problem that some cancer-promoting mutations
-likely occur at low enough frequencies as to be indistinguishable from
-background. Previous attempts to capture these associations using
-Bayesian networks have yielded important new insights into this problem and new
-genes [@hornNetSigNetworkbasedDiscovery2018] that would not have been possible
-with traditional analysis, supporting the case that pathway-centric approaches
-have the potential to increase statistical power.
+frame-shift, stop-gain, untranslated region and splice site mutations in genes
+that are actively expressed in each cancer type, thus avoiding bias from
+transcription coupled repair. Our method of filtration differs from the
+"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
+account for the same biases. We also filtered genes on the basis of pathway
+membership in biochemical pathways, excluding curated gene sets related to
+diseases, syndromes, or classes of proteins with shared catalytic activity or
+conserved domains ( _e.g._ transcription factors) which are potentially
+problematic [@pathways-Atul-Butte]. We chose this approach to limit redundancy,
+which is already high in the remaining pathways for some driver genes, and to
+exclude biologically unrelated collections of genes. We do not account for
+non-coding mutations, amplification events or structural variants, though we
+hope to include them in subsequent studies (see below). Our pathway-centric
+analysis also allows us to circumvent the problem that some cancer-promoting
+mutations likely occur at such low frequencies as to be indistinguishable from
+background. Previous attempts to capture these associations using Bayesian
+networks have yielded important new insights into this problem and new genes
+[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
+traditional analysis, supporting the case that pathway-centric approaches have
+the potential to increase statistical power.
 
@@
 manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways represented by our classes (**figure 1**).
+disruptions in basic pathways as represented by our classes (**figures 1 & 2**).
 
@@
 clusters, regardless of the cancer type (**figure 2F, 3A** and **supplementary
-figure S1**). Some cancer types ( _e.g._ Pancreatic:PAAD in the RAS
-mutant cluster 8, and Kidney Chromophobe:KICH in cluster 1, see **supplementary
-figure S1**) are unevenly distributed, though we could not identify any cancers
-that were exclusive to a single tumor class. Only cancer types with the fewest
+figure S1**). Some cancer types ( _e.g._ Pancreatic:PAAD in the RAS mutant
+cluster 8, and Kidney Chromophobe:KICH in cluster 1, see **supplementary figure
+S1**) are unevenly distributed, though we could not identify any cancers that
+were exclusive to a single tumor class. Only cancer types with the fewest
 samples were found to be absent from one or more classes at all. Somewhat
@@
 could reflect differences in cell-of-origin, rather than fundamental differences
-in proximal cause. The four major subtypes of breast cancer correspond to histological
-and molecular expression profiles that define them and how they respond to experimental
-stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
-molecular regulators of development in the precursor cell types present in the
-breast epithelium drive the various histological phenotypes [@breast-subtypes;
+in proximal cause. The four major subtypes of breast cancer correspond to
+histological and molecular expression profiles that define them and how they
+respond to experimental stimuli [@breast-subtypes2]. It has been hypothesized
+that differences in the molecular regulators of development in the precursor
+cell types present in the breast epithelium drive the various histological
+phenotypes [@breast-subtypes;
 @breast-subtypes3]. Consistent with this view we found that breast tumor samples
@@
 not observe bias for some cancers and subtypes amongst our classes, since some
-treatment regimens have greater efficacy for patients
-of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
-basis for some tumors being treatment-refractory in spite of receiving the
-standard of clinical care for diagnostic markers remains elusive. Doubtless some
-of this is due to bad luck, as tumors can metastasize and remain dormant years
-before they are detected at distal sites, or resistant clones may have already
-arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
-possibility of identifying more informative molecular, histological or cellular
-subtypes that could form a basis for future separation of patients into
-different precision treatment regimens.
+treatment regimens have greater efficacy for patients of a given cancer or
+histological subtype [@breast-subtypes2]. Nonetheless, the basis for some tumors
+being treatment-refractory in spite of receiving the standard of clinical care
+for diagnostic markers remains elusive. Doubtless some of this is due to bad
+luck, as tumors can metastasize and remain dormant years before they are
+detected at distal sites, or resistant clones may have already arisen at
+undetectable levels [@hallmarks_II], but our analysis suggests the possibility
+of identifying more informative molecular, histological or cellular subtypes
+that could form a basis for future separation of patients into different
+precision treatment regimens.
 
@@
 
-Our results illustrate how unique combinations of mutations in pan-cancer driver genes with tissue-specific pathway disruptions result in common categories when viewed at the level of the pathway knowledge-base.
-For example, we find that for "snRNP assembly", one of the unique driver pathways of cluster 5, the top gene in breast cancer is represented by TPR (mutated in 26% of samples).
-By contrast the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of samples, respectively).
-For some other clusters, _e.g._ 7, 8 & 9, the top genes are known drivers such as TP53, PIK3CA and KRAS.
-None of these genes is 100% mutated in any cluster however, and therefore it is the combinations with secondary mutations, many of which are tissue-specific, that round out the complement of pathways in all clusters.
-Some top genes are found in most of the clusters (supplemental figure S5), in spite of the fact that they contribute to many cluster-specific pathways.
-This can only be explained as a result of other less common driver genes complementing the pathway disruption of the unique combination of driver genes that are disrupted in each tumor, and we speculate that many of these are relatively tissue-specific in terms of their sensitivity to mutation.
-Consistent with our observations, @moonlight showed how tissue- and cancer-specific driver genes and druggable targets of pathway moduli can be discovered by machine learning on integrated multi-omics datasets.
-Marrying these approaches in future work will greatly enhance our understanding of tumor biology.
+Our results illustrate how unique combinations of mutations in pan-cancer driver
+genes with tissue-specific pathway disruptions result in common categories when
+viewed at the level of the pathway knowledge-base. For example, we find that for
+"snRNP assembly", one of the unique driver pathways of cluster 5, the top gene
+in breast cancer is represented by TPR (mutated in 26% of samples). By contrast
+the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of
+samples, respectively). For some other clusters, _e.g._ 7, 8 & 9, the top genes
+are known drivers such as TP53, PIK3CA and KRAS. None of these genes is 100%
+mutated in any cluster however, and therefore it is the combinations with
+secondary mutations, many of which are tissue-specific, that round out the
+complement of pathways in all clusters. Some top genes are found in most of the
+clusters (**supplemental figure S5**), in spite of the fact that they contribute to
+many cluster-specific pathways. This can only be explained as a result of other
+less common driver genes complementing the pathway disruption of the unique
+combination of driver genes that are disrupted in each tumor, and we speculate
+that many of these are relatively tissue-specific in terms of their sensitivity
+to mutation. Consistent with our observations, @moonlight showed how tissue- and
+cancer-specific driver genes and druggable targets of pathway moduli can be
+discovered by machine learning on integrated multi-omics datasets. Marrying
+these approaches in future work will greatly enhance our understanding of tumor
+biology.
 
@@
 analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
-proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and are heavily
-skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 10
-represents a group of hyper-mutated tumors that harbor so many mutations that
-virtually no pathway is unaffected. It seems likely that a significant fraction
-of the “mutant” samples for each pathway are burdened with excess passenger mutations. This could be
-addressed with application of ever-increasingly sophisticated filtering likely passenger mutations [ _e.g._ @spliceAI;
-@deep-learning-clinical-variants]. In the future, we hope to incorporate these
-other data into a comprehensive pathway-centric analysis as we have done here
-for point mutations and indels.
+proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and
+are heavily skewed to cluster 1 membership (**Supplementary figure S1**).
+Likewise, Class 10 represents a group of hyper-mutated tumors that harbor so
+many mutations that virtually no pathway is unaffected. It seems likely that a
+significant fraction of the “mutant” samples for each pathway are burdened with
+excess passenger mutations. This could be addressed with application of
+ever-increasingly sophisticated filtering likely passenger mutations [ _e.g._
+@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
+incorporate these other data into a comprehensive pathway-centric analysis as we
+have done here for point mutations and indels.
 
@@
 outset, we found that models assuming different survival based on class and
-cancer type fit better than models in which they are independent
-factors. However, considering that within each cancer type there are different
-clinical standards of care, and even within classes of drugs the preferred
-treatment can very between cancers, it makes sense that we observe tissue-specific class effects. Contrast the situation with ovarian _vs._ breast cancer, which
-are both hormonally driven cancers, for example. Ovarian cancer has but one main
+cancer type fit better than models in which they are independent factors.
+However, considering that within each cancer type there are different clinical
+standards of care, and even within classes of drugs the preferred treatment can
+very between cancers, it makes sense that we observe tissue-specific class
+effects. Contrast the situation with ovarian _vs._ breast cancer, which are both
+hormonally driven cancers, for example. Ovarian cancer has but one main
 treatment axis, platinum, whereas breast cancer patients have a variety of
@@
 outcomes than the same class of a different cancer type. Unfortunately, given
-the diversity drug classes and treatments (including non-chemical treatments
-such as surgery and palliative care), we lack sufficient power 
-to explore these variables in the TCGA data. It is our hope that 
-future studies will help to distinguish between treatment-specific
-effects on survival given different pathway disruption classes.
+the diversity of drug classes and treatments (including non-chemical treatments
+such as surgery and palliative care), we lack sufficient power to explore these
+variables in the TCGA data. It is our hope that future studies will help to
+distinguish between treatment-specific effects on survival given different
+pathway disruption classes.
 
@@
 disrupted in all or most of the defined classes (**supplementary figure S1**)
-and are frequently the result of aberrations involving common driver genes such as
-BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
+and are frequently the result of aberrations involving common driver genes such
+as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
 (**supplementary figure S3** and **supplemental file 1**). Class-specific
@@
 frequencies as in @PCAWG-EVO. However, since many of the genes in the
-non-cluster-specific pathways involve the known driver genes, it is reasonable to
-hypothesize that these mutations promote or enable the subsequent acquisition of
-class-specific defects via random mutation and natural selection, thus producing
-the classes we observed. In support of this, the pan-cancer analysis of whole
-genomes consortium (PCAWG) found that oncogenic driver
-mutations are highly enriched in early arising clones, whereas later arising
-clones have much greater diversity in driver mutations [@PCAWG-EVO]. Moreover,
-driver genes that are known to be responsible for discrete mutation signatures
-such as APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the
-varying selective pressures in different tissues [@PCAWG-MAIN].
+non-cluster-specific pathways involve the known driver genes, it is reasonable
+to hypothesize that these mutations promote or enable the subsequent acquisition
+of class-specific defects via random mutation and natural selection, thus
+producing the classes we observed. In support of this, the pan-cancer analysis
+of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
+highly enriched in early arising clones, whereas later arising clones have much
+greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
+are known to be responsible for discrete mutation signatures such as APOBEC,
+BRCA1 and BRCA2 produce mutational hotspots reflecting the varying selective
+pressures in different tissues [@PCAWG-MAIN].
 
@@
 cluster specific analysis, suggesting that signal is diluted when the samples
-are analyzed together, and supporting the view that, within the parameters of our analysis, metastasis has
-class-specific requirements. Since the number of metastatic samples is
-relatively low this part of our analysis is likely underpowered and subject to
-expanded analysis with larger cohorts. Nonetheless, the class-specific
-metastatic pathway enrichment we observe (in **table 1**) could be a result of
-the pathways directly promoting metastatic phenotypes in these classes, or
-because they acquire a stronger or more complete oncogenic phenotype, as we
-don’t know the order in which they were acquired. The fact that most metastatic
-enrichment is cluster-specific and has a tendency to overlap with
-cluster-specific pathways from non-metastatic tumors of neighboring clusters
-favors the latter interpretation, and suggests a mechanism of complementarity,
-in which newly acquired mutations result in similar classes converging on
-one or more deadly phenotypes that capture most of the critical features of end-stage
-cancer. One explanation for this could be that tumors must accumulate mutations
-in a certain order or face elimination due to any of the selective bottlenecks
-they face (senescence, the immune system, _etc._) [@hallmarks_II]. Some clusters
-were characterized by cluster-specific pathways that seem superficially
-unrelated; for example cluster 2 has mutations in the proteasome subunits,
-cluster 3 has mutations impacting protein translation, whereas cluster 5 has
-defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
-traditional extra-cellular signaling pathways (as exemplified in **table S1**).
-It may be that the primary effects of mutations in protein and RNA metabolism in
-clusters 3 and 5 point to currently underappreciated mechanisms impacting hallmark
-pathways. Clusters 3 and 5 are also marked by a decidedly different set of
-pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
-pathways) mutations in glucagon and glycogen metabolism, respectively, in their
-metastases (**table 1**), consistent with energy metabolism as one of the new
-hallmarks [@hallmarks_II], but also distinguishing these classes from the others
-in the order or nature of hallmark acquisition. Thus, even at relatively low
-power, our analysis of metastasis further supports the pathway-centric approach
-as a fruitful way to uncover real differences in underlying biology. As a major
-caveat to this, we acknowledge that many therapies _do_ target highly-specific
-driver genes, markers, and signaling pathways ( _e.g._ TP53, EGFR or HER2), but
-understanding the broader context of the genetic background and pathway
-vulnerability of tumors containing such markers may aid in creating smarter
-combination therapies. We submit that when we discover the requirements of each
-cluster with respect to pathway disruptions and metastasis we may be able to
-target them therapeutically and prevent further adaptation.
+are analyzed together, and supporting the view that, within the parameters of
+our analysis, metastasis has class-specific requirements. Since the number of
+metastatic samples is relatively low this part of our analysis is likely
+underpowered and subject to expanded analysis with larger cohorts. Nonetheless,
+the class-specific metastatic pathway enrichment we observe (in **table 1**)
+could be a result of the pathways directly promoting metastatic phenotypes in
+these classes, or because they acquire a stronger or more complete oncogenic
+phenotype, as we don’t know the order in which they were acquired. The fact that
+most metastatic enrichment is cluster-specific and has a tendency to overlap
+with cluster-specific pathways from non-metastatic tumors of neighboring
+clusters favors the latter interpretation, and suggests a mechanism of
+complementarity, in which newly acquired mutations result in similar classes
+converging on one or more deadly phenotypes that capture most of the critical
+features of end-stage cancer. One explanation for this could be that tumors must
+accumulate mutations in a certain order or face elimination due to any of the
+selective bottlenecks they face (senescence, the immune system, _etc._)
+[@hallmarks_II]. Some clusters were characterized by cluster-specific pathways
+that seem superficially unrelated; for example cluster 2 has mutations in the
+proteasome subunits, cluster 3 has mutations impacting protein translation,
+whereas cluster 5 has defects in RNA metabolism and clusters 7, 8 and 9 have
+defects in the more traditional extra-cellular signaling pathways (as
+exemplified in **table S1**). It may be that the primary effects of mutations in
+protein and RNA metabolism in clusters 3 and 5 point to currently
+underappreciated mechanisms impacting hallmark pathways. Clusters 3 and 5 are
+also marked by a decidedly different set of pathways in metastasis. Clusters 3
+and 5 notably have (in addition to signaling pathways) mutations in glucagon and
+glycogen metabolism, respectively, in their metastases (**table 1**), consistent
+with energy metabolism as one of the new hallmarks [@hallmarks_II], but also
+distinguishing these classes from the others in the order or nature of hallmark
+acquisition. Thus, even at relatively low power, our analysis of metastasis
+further supports the pathway-centric approach as a fruitful way to uncover real
+differences in underlying biology. As a major caveat to this, we acknowledge
+that many therapies _do_ target highly-specific driver genes, markers, and
+signaling pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader
+context of the genetic background and pathway vulnerability of tumors containing
+such markers may aid in creating smarter combination therapies. We submit that
+when we discover the requirements of each cluster with respect to pathway
+disruptions and metastasis we may be able to target them therapeutically and
+prevent further adaptation.

### 2020-08-16 text/introduction.Rmd
modify spacing before and after all section & subsection headings in word style template; edit and reflow abstract and introduction
edit full document

@@
 @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
-when genes are mutated at a greater rate than expected due to chance. It has been estimated that on the order of three
-mutations in key oncogenes and/or tumor suppressors would be sufficient to
-transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].
+when genes are mutated at a greater rate than expected due to chance. It has
+been estimated that on the order of three mutations in key oncogenes and/or
+tumor suppressors would be sufficient to transform a normal cell to a cancerous
+state [@vogelsteinPathCancerThree2015].
 
@@
 expression and genomic signatures, with the express goal of understanding the
-common themes of all cancer independent of tissue of origin. Understanding cancer
-as a disease of the cell has long been a goal of
-the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
-Discussion). The first of these studies in the enomic era identified 11 subtypes from 12 cancer types,
-using integrative analysis with co-equal weighting of gene expression,
-methylation, copy number and proteomics data
+common themes of all cancer independent of tissue of origin. Understanding
+cancer as a disease of the cell has long been a goal of the field and is perhaps
+best characterized in essays by @weinberg2000; @hallmarks_II (see also
+Discussion). The first of these studies in the genomic era identified 11 subtypes
+from 12 cancer types, using integrative analysis with co-equal weighting of gene
+expression, methylation, copy number and proteomics data
 [@hoadleyMultiplatformAnalysis2014]. The principal finding was that
 tissue-of-origin is the predominant driving factor, though a significant
-proportion of tumors (~10%$) could be reclassified independent of
+proportion of tumors (~10%) could be reclassified independent of
 tissue-of-origin. In a second study involving 33 cancer types and a much greater
-number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
-could be further subdivided into organ specific groups, including
-pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
-squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
+number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify
+28 clusters that could be further subdivided into organ specific groups,
+including pan-gastrointestinal, pan-gynecological, pan-squamous,
+pan-gynecological/ squamous and pan-kidney
+[@hoadleyCellofOriginPatternsDominate2018].
 
 More recently, attempts have been made to consider pan-cancer mutations and
-other data outside the context of enrichment at the gene level. @PCAWG-PATH carried out near-exhaustive analysis of pathway enrichment using whole genome data. Horn _et al._
-[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
-sparseness of mutations, allowing the potential importance of a gene to be
-influenced by the aggregate of mutations in its nearest network neighbors. This
-method is powerful especially when applied to a small number (on the order of
-hundreds) of tumors. One of the challenges in the field remains that beyond a
-couple hundred common driver genes and mutations, there are likely thousands of
-moderate effect genes that occur at such low frequency--given tissue diversity--and low sample numbers that they are impossible to detect using positive
-selection theory. Some groups have attempted to address this challenge by
-applying machine learning to cancer data sets to discover groups of functionally
-related genes as they interact with larger pathways and networks
-[@kimPancancerAnalysisSomatic2018, @mourikis-ml-pathways, @moonlight]. These
+other data as disruptions of normal pathway activity. @PCAWG-PATH
+carried out near-exhaustive analysis of pathway enrichment using whole genome
+data. Horn _et al._ [@hornNetSigNetworkbasedDiscovery2018] introduce approaches
+to address the sparseness of mutations, allowing the potential importance of a
+gene to be influenced by the aggregate of mutations in its nearest network
+neighbors. This method is powerful especially when applied to a small number (on
+the order of hundreds) of tumors. One of the challenges in the field remains
+that beyond a couple hundred common driver genes and mutations, there are likely
+thousands of moderate effect genes that occur at such low frequency--given
+tissue diversity--and low sample numbers that they are impossible to detect
+using positive selection theory. Some groups have attempted to address this
+challenge by applying machine learning to cancer data sets to discover groups of
+functionally related genes as they interact with larger pathways and networks
+[@kimPancancerAnalysisSomatic2018; @mourikis-ml-pathways; @moonlight]. These
 studies have proven very effective at highlighting some potentially fundamental

### 2020-08-16 text/manuscript.Rmd
edit full document

--- 2020-08-14 text/manuscript.Rmd
@@
   library(flextable),
-  library(plot3D),
-  library(ggfortify)
+  library(plot3D)
 ))
@@
 ### A taxonomy of tumors based on disrupted molecular pathways
+
 To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
 Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
-(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our
-study given the relative completeness of the patient metadata, particularly for
-survival and staging. We chose to analyze somatic mutations in exome
-sequencing data because the affected target gene is known unambiguously.
-Therefore, we selected all missense, nonsense, frameshift, stop-gain, untranslated
-region, and splicing mutations. In order to minimize bias from well-studied diseases
-and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
-of interest corresponding to basic cellular processes and biochemical pathways,
+(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
+given the relative completeness of the patient metadata, particularly for
+survival and staging. We chose to analyze somatic mutations in exome sequencing
+data because the affected target gene is known unambiguously. Therefore, we
+selected all missense, nonsense, frameshift, stop-gain, untranslated region, and
+splicing mutations. In order to minimize bias from well-studied diseases and
+processes, we selected 377 [Reactome pathways](https://reactome.org/) of
+interest corresponding to basic cellular processes and biochemical pathways,
 excluding gene sets that correspond to miscellaneous categories ( _e.g._
 “transcription factors”) or disease associations ( _e.g._ “mutated in colon
-cancer”) and filtered our gene list on membership in these pathways (total
-of 8940 genes; see methods for details).
+cancer”) and filtered our gene list on membership in these pathways (total of
+8940 genes; see methods for details).
 
-To avoid bias toward larger
-pathways ( _i.e._ pathways with more member genes), we counted pathways as disrupted if one or more member genes were mutated.
-We do not attempt to calculate enrichment for mutations within a pathway, for example.
-Data
+To avoid bias toward larger pathways ( _i.e._ pathways with more member genes),
+we counted pathways as disrupted if one or more member genes were mutated. We do
+not attempt to calculate enrichment for mutations within a pathway. Data
 binarized by pathway are likely to be noisy for several reasons. First, point
 mutations can be deleterious (attenuating or hypomorphic) or activating
-(neomorphic or hypermorphic) in genes, and these can in turn act functionally as either activators or repressors.
-Within the scope of this study we accept this uncertainty and assume mutations
-are generically disruptive to pathways regardless of whether they are
-activating or deactivating or whether they affect a gene that we might classify
-as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, we
-know that low-expressed genes and non-expressed genes are known to accumulate
-mutations at a higher rate due to transcription coupled repair
-[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
+(neomorphic or hypermorphic) in genes, and these can in turn act functionally as
+either activators or repressors. Within the scope of this study we accept this
+uncertainty and assume mutations are generically disruptive to normal pathway
+activity regardless of whether they are activating or deactivating or whether
+they affect a gene that we might classify as one thing or another ( _e.g._
+"Tumor suppressor" or "Oncogene"). Second, we know that low-expressed genes and
+non-expressed genes are known to accumulate mutations at a higher rate due to
+transcription coupled repair [@kandothMutationalLandscapeSignificance2013;
+@kimPancancerAnalysisSomatic2018;
 @Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
 issue, we identified genes with low expression in each type of cancer and
-eliminated them for that cancer type only (see methods for details). 
-Highly expressed genes could also have high mutation rates owing to
-transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
-that such mechanisms result in cell-type-specific biases that might be
-biologically meaningful for predisposition to different classes of cancer
-however, and therefore chose not to exclude these genes from our analysis. After
-selecting our pathways and genes, we then compiled a matrix of the pathways,
-assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene
-carries a mutation) to the pathway regardless of the number of genes carrying
-mutations (**figure 1**).
+eliminated them for that cancer type only (see methods for details). Highly
+expressed genes could also have high mutation rates owing to transcription
+induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that such
+mechanisms result in cell-type-specific biases that might be biologically
+meaningful for predisposition to different classes of cancer however, and
+therefore chose not to exclude these genes from our analysis. After selecting
+our pathways and genes, we then compiled a matrix of the pathways, assigning a
+Boolean value of 1 to each pathway with one or more genes mutated and 0 for all
+others (**figure 1**).
 
@@
 analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
-containing useful information. (**figures 2A, B**). Then we performed a UMAP
+containing useful information by selecting the eigenvalue with the most
+explanatory power, using the average of 100 permutations of the data as baseline
+(**figure 2A**). We then chose the maximum eigenvalue for which the p-value
+remained ≤ 0.05 (see cutoff in **figure 2B**). Then we performed a UMAP
 analysis, both in order to summarize the MCA graphically, and as a preprocessing
@@
 notable for its lobed structure, with several reproducible projections
-regardless of random seed setting. A representative version of this
-3D UMAP is shown in **figure 2C**, rotated to enhance the
-visibility of the major features. Following this spatial mapping we attempted to
-define groupings of similar tumors within the spatial map using HDBSCAN, which
-performs hierarchical clustering and provides metrics of cluster stability and
-probabilities of cluster membership for each node. However, HDBSCAN is sensitive
-to several parameters; key for our analysis are the minimum number of tumor
-samples in a cluster that capture the maximum number of tumor samples, measured
-by probability of membership of ≥ 5% in at least one class. Thus, we
-created a score metric as the fraction of classified tumors with max probability
-< 5% in one class and chose a cluster size of 92 to minimize the score
-function (**figure 2D**). HDBSCAN with these settings resulted in ten distinct
+regardless of random seed setting. A representative version of this 3D UMAP is
+shown in **figure 2C**, rotated to enhance the visibility of the major features.
+Following this spatial mapping we attempted to define groupings of similar
+tumors within the spatial map using HDBSCAN, which performs hierarchical
+clustering and provides metrics of cluster stability and probabilities of
+cluster membership for each node. However, HDBSCAN is sensitive to several
+parameters; key for our analysis are the minimum number of tumor samples in a
+cluster that capture the maximum number of tumor samples, measured by
+probability of membership of ≥ 5% in at least one class. Thus, we created a
+score metric as the fraction of classified tumors with max probability < 5% in
+one class and chose a cluster size of 92 to minimize the score function
+(**figure 2D**). HDBSCAN with these settings resulted in ten distinct
 high-density clusters which we then projected onto the UMAP (**figure 2E**).
-This classified 6,038 out of 7,607 tumors but still resulted in a
-significant fraction of unclassified tumor samples. Since we ultimately wish to
-be able to classify any tumor using this scheme, we performed k Nearest
-Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
-the set and then lets the _k_ most similar tumors "vote" as to the identity of
-the query tumor sample based on their classification labels. We set _k_ to be
-the square root of the number of tumor samples (87). Using this method, we
-assigned cluster membership to the remaining tumors (**figure 2F**).
+This classified 6,038 out of 7,607 tumors but still resulted in a significant
+fraction of unclassified tumor samples. Since we ultimately wish to be able to
+classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
+analysis, which computes a similarity metric to every tumor in the set and then
+lets the _k_ most similar tumors "vote" as to the identity of the query tumor
+sample based on their classification labels. We set _k_ to be the square root of
+the number of tumor samples (87). Using this method, we assigned cluster
+membership to the remaining tumors (**figure 2F**).
 
@@
 exceptions) were not heavily biased in one class, and all well-represented
-cancer types had tumors in every class (see **figure 3A**
-and full tumor profiles in **supplementary figure S1**, see also interactive
-media from **supplementary file 1 (html)**), suggesting that, in principle at
-least, our novel pathway-disruption classification method identifies molecular
-pathology largely independent of tissue-of-origin. As an
-example of one type of cancer that does have a biased pathway profile,
-pancreatic adenocarcinoma (PAAD) was predominantly found in class 8, which is
-distinguished  by KRAS mutations among others (**supplementary figure S3**). 
-But even PAAD comprises tumors from the nine remaining classes, meaning that 
-patients suffering these tumors have potentially different underlying molecular
-pathologies.
+cancer types had tumors in every class (see **figure 3A** and full tumor
+profiles in **supplementary figure S1**, see also interactive media from
+**supplementary file 1 (html)**), suggesting that, in principle at least, our
+novel pathway-disruption classification method identifies molecular pathology
+largely independent of tissue-of-origin. As an example of one type of cancer
+that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD) was
+predominantly found in class 8, which is distinguished  by KRAS mutations among
+others (**supplementary figure S3**). But even PAAD comprises tumors from the
+nine remaining classes, meaning that patients suffering these tumors have
+potentially different underlying molecular pathologies.
 
@@
 result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. To illustrate this, we projected annotations for each of the breast
-cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
-normal-like, and luminal A and B subtypes onto the UMAP. These are among the
-most heavily studied molecular subtypes in cancer, which each have different
-prognoses and standards of care. We did not observe any exclusive segregation by
-pathway for these subtype annotations (**Figure 3B** and **supplemental
-file 1**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** for the full set of projections); we find that
-the subtypes, though often biased towards one or more classes, are almost never
-exclusive. We interpret these data in aggregate to mean
-that our pathway disruption classes do not correspond to molecular subtypes
-within the parent cancer type. Furthermore, this result raises the possibility
-that, molecular and histo-pathological subtypes could represent different
-precursor cell origins within each cancer type. This interpretation still leaves
-open the possibility that different precursor cell types, being susceptible to
-different teratogens and selection pressures ( _e.g._ insulin dependence) could
-be biased toward one or more of the pathway disruption profiles.
+pathway disruptions. To illustrate this, we projected annotations for each of
+the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
+positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
+among the most heavily studied molecular subtypes in cancer, which each have
+different prognoses and standards of care. We did not observe any exclusive
+segregation by pathway for these subtype annotations (**Figure 3B** and
+**supplemental file 1**). We also projected histological subtype data for the
+remaining cancers (see **supplementary figure S2** for the full set of
+projections); we find that the subtypes, though often biased towards one or more
+classes, are almost never exclusive. We interpret these data in aggregate to
+mean that our pathway disruption classes do not correspond to previously
+identified molecular subtypes within the parent cancer type. Furthermore, this
+result raises the possibility that, molecular and histo-pathological subtypes
+could represent different precursor cell origins within each cancer type. This
+interpretation still leaves open the possibility that different precursor cell
+types, being susceptible to different teratogens and selection pressures (
+_e.g._ insulin dependence) could be biased toward one or more of the pathway
+disruption profiles.
 
@@
 treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability generally, proof-reading and DNA damage repair, and
-telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
+factors such as genome stability generally, proof-reading and DNA damage repair,
+and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
 break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
@@
 similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_,
-_MLH3_, _PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length, we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
+_MLH3_, _PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these
+genes except for _APC_ exhibited any remarkable specificity with respect to
+cluster assignment (**supplementary figure S3**). To look at other risk factors
+such as maintenance of DNA methylation levels and telomere length, we projected
+somatic mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and
+_DKC1_, and observed similar lack of segregation by cluster (**supplementary
+figure S3**).
 
@@
 
-Tumor staging is based on physico-pathological criteria, including tumor diameter,
-which can vary greatly in importance between different tissues. Stage is used
-clinically as a proxy for advancement toward a more deadly state and metastasis.
-Given these assumptions, it is possible that more advanced tumors have common
-pathway disruption profiles, and the UMAP which features a series of lobe-like
-structures on a common backbone of tumor samples (**figure 2C**) could in
-principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone starts with a cluster of tumors (class 1) that has the fewest point mutations and culminates in a cluster (class 10) which
-has nearly every pathway disrupted (**figure 4**). However, outside of class 10 we don't observe
-an obvious trend in the overall mutation burden across the backbone of the UMAP (**figure
-1H**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
-classes represent advancement through stages, we projected staging data onto the
-UMAP. Similar to tissue of origin and other categories of tumor, we did not
-observe any bias among the stages to specific pathway disruption classes
-(**figure 3D**), suggesting that stage is not an determining factor in pathway
-disruption classes.
+Tumor staging is based on physico-pathological criteria, including tumor
+diameter, which can vary greatly in importance between different tissues. Stage
+is used clinically as a proxy for advancement toward a more deadly state and
+metastasis. Given these assumptions, it is possible that more advanced tumors
+have common pathway disruption profiles, and the UMAP which features a series of
+lobe-like structures on a common backbone of tumor samples (**figure 2C**) could
+in principle also reflect progression through a series of stages. In support of
+this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
+the fewest point mutations and culminates in a cluster (class 10) which has
+nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
+don't observe an obvious trend in the overall mutation burden across the
+backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
+the molecular-pathway disruption classes represent advancement through stages,
+we projected staging data onto the UMAP. Similar to tissue of origin and other
+categories of tumor, we did not observe any bias among the stages to specific
+pathway disruption classes (**figure 3D**), suggesting that stage is not an
+determining factor in pathway disruption classes.
 
@@
 loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To determine whether
-any of our classes correspond to an especially advanced stage of cancer across
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly we 
-observed an even distribution of metastases across classes (**figure 3E**). This
-final observation suggests that our pathway-disruption classification is
-dependent on particular combinations of gene mutations that hit different
-pathway profiles that can each give rise to advanced stages of disease and
-metastasis, regardless and independent of overall mutational burden (see Discussion for further analysis).
+transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To
+determine whether any of our classes correspond to an especially advanced stage
+of cancer across tissue types, we projected the metastasis data onto our UMAP,
+and surprisingly we observed an even distribution of metastases across classes
+(**figure 3E**). This final observation suggests that our pathway-disruption
+classification is dependent on particular combinations of gene mutations
+affecting different pathways that can each give rise to advanced stages of
+disease and metastasis, regardless and independent of overall mutational burden
+(see Discussion for further analysis).
 
@@
 samples and top genes is provided in **supplementary table ST1**. As expected,
-these analyses reveal the broad importance of many well known pathways that are disrupted
-in cancer, including "PIP3 activates Akt signaling" (77% of samples),
+these analyses reveal the broad importance of many well known pathways that are
+disrupted in cancer, including "PIP3 activates Akt signaling" (77% of samples),
 "MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
@@
 0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
-terms of pathway enrichment. To explore the specific pathways marking
-each class, we projected disruptions for each of the 377 pathways onto the UMAP
+terms of pathway enrichment. To explore the specific pathways marking each
+class, we projected disruptions for each of the 377 pathways onto the UMAP
 (**supplementary figure S4**). Classes 3 and 5 were distinguished by metabolic
@@
 enrichment levels of the three, having mutations in hedgehog signaling,
-"β-catenin degradation", "cellular response to hypoxia", "regulation of
-cell cycle" and "apoptosis" among others.
+"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
+cycle" and "apoptosis" among others.
 
-### A combination of common driver mutations and tissue-specific genes contribute to common class signatures
+### A combination of driver mutations and tissue-specific genes contribute to common class signatures
 
-Prior efforts to extract common signatures from pan-cancer 
-datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
--omics data signatures. Given our pathway-disruption based classification, this raises 
-the question, are tumor phenotypes driven entirely by common driver genes, or by “silent”
-tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
-thresholds), or a combination of both? To answer this question, we compared top pathway genes for each cluster 
-relative to the TCGA background to find differentially mutated genes. We ranked odds ratios
-and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
-(**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
-signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
-genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
-Interestingly, cluster 9, which also shared multiple enrichment in signaling pathways 
-with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
-that tumors in this cluster could be derivative of either cluster 7 or 8. Clusters 3 and 
-5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
-nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible for 
-nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
-We also observed that genes that were enriched for one cluster are depleted from others ( _e.g._ 
-_TP53_ is enriched in cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster
-7, but depleted in clusters 3 and 8). Next, we investigated the proportion of samples per 
-cancer type for the significant genes within a cluster (**Figure 6**; heatmap). Surprisingly, 
-cluster-specific tumors were not predominated by one or more highly mutated genes across all 
-cancers. Instead, when observing the mutation rate for these genes within samples that belong
-to a cluster, the mutation rate is heterogeneous across tumors by tissue origin ( _e.g._ In 
-cluster 4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched
-for _HIST1H2AC_). Even among the top most mutationally enriched genes within clusters there is
-no global pattern, indicating that our classes are not driven by individual genes, but rather 
-networks as a whole. Taken together, our data identifies a framework of cancer type-specific 
-mutations associated with specific clusters.
+Prior efforts to extract common signatures from pan-cancer datasets met with
+difficulty in distinguishing tumor samples from their tissue-specific -omics
+data signatures. Given our pathway-disruption based classification, this raises
+the question, are tumor phenotypes driven entirely by common driver genes, or by
+“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
+statistical significance thresholds), or a combination of both? To answer this
+question, we compared top pathway genes for each cluster relative to the TCGA
+background to find differentially mutated genes. We ranked odds ratios and
+selected the top ten enriched and depleted genes (pvalue < 0.01) for each
+cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
+multiple enrichment in signaling pathways, are largely driven by mutations in
+_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
+_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
+also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
+enriched for both _PIK3CA_ and _KRAS_. This result suggests that tumors in this
+cluster could be derivative of either cluster 7 or 8. Clusters 3 and 5, defined
+by enrichment in metabolic pathways, had mutations in ribosomal proteins and
+nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible
+for nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
+involved in protein degradation. We also observed that genes that were enriched
+for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster
+6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
+clusters 3 and 8). Next, we investigated the proportion of samples per cancer
+type for the significant genes within a cluster (**Figure 6**; heatmap).
+Surprisingly, cluster-specific tumors were not predominated by one or more
+highly mutated genes across all cancers. Instead, when observing the mutation
+rate for these genes within samples that belong to a cluster, the mutation rate
+is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
+enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
+_HIST1H2AC_). Even among the top most mutationally enriched genes within
+clusters there is no global pattern, indicating that our classes are not driven
+by individual genes, but rather networks as a whole. Taken together, our data
+identifies a framework of cancer type-specific mutations associated with
+specific clusters.
 
@@
 pathways (**figure 5B**) we applied this method to metastatic vs non-metastatic
-tumors, as we find that metastatic tumors are distributed across all ten clusters
-( _e.g._ **figure 3E**). Using all non-metastatic tumors as background, we found
-very low levels of enrichment (≤ 10%) in a handful of pathways. We
+tumors, as we find that metastatic tumors are distributed across all ten
+clusters ( _e.g._ **figure 3E**). Using all non-metastatic tumors as background,
+we found very low levels of enrichment (≤ 10%) in a handful of pathways. We
 reasoned that the individual clusters might be too different to detect global
@@
 Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found a total of 31 enriched pathways across all clusters (**table 1**). A number of
-enrichments were found in multiple clusters represented by pathways that were
-already shown to be enriched in a neighboring cluster. For example,
+found a total of 31 enriched pathways across all clusters (**table 1**). A
+number of enrichments were found in multiple clusters represented by pathways
+that were already shown to be enriched in a neighboring cluster. For example,
 “Signaling by PTK6” is enriched in non-metastatic samples of cluster 8 (see
 **supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
-metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$, **table 1**). This is also true of 
-“Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic
-samples were enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathway, 
-which is also specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
-specific to metastases tend toward complementary mutations that are otherwise specific to
-non-metastatic tumors of other clusters.
+metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$, **table 1**). This is also
+true of “Erythropoietin activates RAS”, which is enriched in cluster 8
+(**supplementary figure S4**) and also in metastatic tumors of clusters 7 and 9.
+Cluster 4 metastatic samples were enriched for “Fc epsilon receptor (FCERI)
+signaling”, a key neutrophil pathway, which is also specific to clusters 2, 7
+and 8 non-metastatic tumors. Thus, pathway lesions specific to metastases tend
+toward complementary mutations that are otherwise specific to non-metastatic
+tumors of other clusters.
 
@@
 All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript in R markdown format. R markdown files and all other
-models are available from
-[our repositories](https://github.com/dennishazelett/TTmanu) in the following
+fully reproducible manuscript in R markdown format. R markdown files and all
+other models are available from [our
+repositories](https://github.com/dennishazelett/TTmanu) in the following
 locations on the distributed version control site GitHub.
@@
 To understand the molecular mechanism of cancer at a pathway level, we used
-Reactome (https://reactome.org/), a knowledge-based pathway
-database. The mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships,
-and complexes to top pathways were downloaded from https://reactome.org/download-data. Using these data, we imposed pathway criteria to
-define basic cellular processes and biochemical pathways: (1) human-derived pathways
-( _“HSA”_) (2) limited to grandchild node for each parent pathway ( _e.g._ ‘Beta-catenin 
-independent WNT signaling’ in ‘Signal Transduction’) (3) exclusion of pathways in the
-parent pathway: "Disease", "Muscle contraction", and "Reproduction" or pathway names that 
-include any of the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder", 
-"mutant", "loss", "infection", "bacteria", or "listeria". While some of the excluded pathways 
-have been shown to play an important role in cancer, they are highly specialized 
-( _e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for most of the excluded pathways,
-a neutral version pathway of the pathway exists ( _e.g._ "PIP3 activates Akt signaling"). 
-Finally, we mapped the 18,577 Ensembl IDs from the TCGA dataset to the highly selected
-Reactome pathways. This operation produced a lookup table that consisted of 377 pathways
-mapped to 8,940 genes.
+Reactome (https://reactome.org/), a knowledge-based pathway database. The
+mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
+complexes to top pathways were downloaded from
+https://reactome.org/download-data. Using these data, we imposed pathway
+criteria to define basic cellular processes and biochemical pathways: (1)
+human-derived pathways ( _“HSA”_) (2) limited to grandchild node for each parent
+pathway ( _e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
+Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
+"Muscle contraction", and "Reproduction" or pathway names that include any of
+the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
+"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
+excluded pathways have been shown to play an important role in cancer, they are
+highly specialized ( _e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
+most of the excluded pathways, a neutral version pathway of the pathway exists (
+_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577 Ensembl
+IDs from the TCGA dataset to the highly selected Reactome pathways. This
+operation produced a lookup table that consisted of 377 pathways mapped to 8,940
+genes.
 
@@
 samples. We considered a pathway enriched if the 95% range of credible
-differences thus obtained excludes 0 and the mean
-credible difference was greater than or equal to 30% enrichment, which
-excludes a large number of small differences that are not likely to be
-biologically relevant.
+differences thus obtained excludes 0 and the mean credible difference was
+greater than or equal to 30% enrichment, which excludes a large number of small
+differences that are not likely to be biologically relevant.
 
@@
 
-The authors have nothing to declare. This work was supported
+The authors have nothing to declare. This work was supported by a grant from the Cedars-Sinai Precision Health Initiative.
 

### 2020-08-16 text/survival.Rmd
edit full document

--- 2020-08-14 text/survival.Rmd
@@
 ethnicity. Our model assumes that at birth each person has a small initial
-probability per unit time to die, and that this probability 
-grows exponentially in time at some baseline rate that
-matches the observed CDC data. The rate of growth in risk is assumed to jump to
-a higher level at diagnosis, which we take as a proxy for onset. This represents
-an accelerated aging type model. To fit the model, we constrain the baseline
-risks using the available CDC life tables. We then constrain the cancer specific
-multiplier using the longevity data within the cancer tumor dataset. Our model
-estimates the effect of cancer type and class-specific cancer effects
-independently, resulting in a cancer and class specific estimate of the aging
-rate multiplier.
+probability per unit time to die, and that this probability grows exponentially
+in time at some baseline rate that matches the observed CDC data. The rate of
+growth in risk is assumed to jump to a higher level at diagnosis, which we take
+as a proxy for onset. This represents an accelerated aging type model. To fit
+the model, we constrain the baseline risks using the available CDC life tables.
+We then constrain the cancer specific multiplier using the longevity data within
+the cancer tumor dataset. Our model estimates the effect of cancer type and
+class-specific cancer effects independently, resulting in a cancer and class
+specific estimate of the aging rate multiplier.
 
-To understand the model, we show in **Figure 7A** simulated survival time
-curves for four related groups. We start with a cohort of women aged 30 and show
-their expected survival (blue). By setting their cancer rate multiplier to $k=2$
-we can simulate the effect of a moderately deadly cancer (red). Compare this to
-a randomly selected group of women ranging in age from 30 to 70 years old (dark
-blue). Immediately the survival curve changes due to the mixture of ages,
+To understand the model, we show in **Figure 7A** simulated survival time curves
+for four related groups. We start with a cohort of women aged 30 and show their
+expected survival (light blue). By setting their cancer rate multiplier to $k=2$
+we can simulate the effect of a moderately deadly cancer (yellow). Compare this
+to a randomly selected group of women ranging in age from 30 to 70 years old
+(dark blue). Immediately the survival curve changes due to the mixture of ages,
 without any malignancy. Adding a $k=2$ malignancy further reduces expected
-lifespan (dark red). We can see that for different tissues, simply due to the
-change in distribution of age-at-diagnosis, we should expect equally deadly
-cancers to have *different* survival curves. The model takes this into account
-and the aggressiveness of the cancer can be estimated without the confounding
-aspect of age.
+lifespan (red). We can see that for different tissues, simply due to the change
+in distribution of age-at-diagnosis, we should expect equally deadly cancers to
+have *different* survival curves. The model takes this into account and the
+aggressiveness of the cancer can be estimated without the confounding aspect of
+age.
 

### 2020-08-16 text/fig_legend.Rmd
Fix some bad LaTeX in fig 9 label

@@
-**Figure 1: Molecular pathway profiles of tumor samples with one or more
-mutations.** Each of 377 selected Reactome pathways (columns) is classified
-as disrupted if one or more genes is mutated in the tumor sample (rows) where
-red and grey represents the pathway is mutated and not, respectively. Tumors
-are grouped by tissue of origin using standardized abbreviations 
-from the TCGA project.
+**Figure 1: Molecular pathway profiles of tumor samples with one or more mutations.** 
+Each of 377 selected Reactome pathways (columns) is classified as disrupted if
+one or more genes is mutated in the tumor sample (rows) where red and grey
+represents the pathway is mutated and not, respectively. Tumors are grouped by
+tissue of origin using standardized abbreviations from the TCGA project.
 
@@
 a) MCA analysis was performed to determine the eigenvalue with the most
-explanatory power from the categorical data. Average of 100 permutations of 
-the data was used as a baseline. b) The p-value of each eigenvalue was plotted
-and the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA 
-was projected down to 3 dimensions for the purpose of density based clustering 
-(only the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN
-was used to perform hierarchical density based clustering tuned to minimize
-the fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized 
-on the UMAP projection; tumors are colored by cluster membership (see legend). 
-This results in some tumors remaining unclassified (grey points). f) kNN was 
-used to assign membership to the remaining tumors.
+explanatory power from the categorical data. Average of 100 permutations of the
+data was used as a baseline. b) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
+projected down to 3 dimensions for the purpose of density based clustering (only
+the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
+used to perform hierarchical density based clustering tuned to minimize the
+fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
+the UMAP projection; tumors are colored by cluster membership (see legend). This
+results in some tumors remaining unclassified (grey points). f) kNN was used to
+assign membership to the remaining tumors.
 
@@
 COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
-Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the UMAP.
-c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of tumor 
-stage onto the UMAP, regardless of cancer type. e) Projection of metastatic status
-onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.
+Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the
+UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of
+tumor stage onto the UMAP, regardless of cancer type. e) Projection of
+metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
+metastatic tumors.
 
-**Figure 4: Somatic mutation frequencies for tumors in each class** Column uses 
-a logarithmic axis to represent mutation count. Row represents the tumor samples. 
-Each dot corresponds to the sample mutation count for tumors in each class, 
-ordered by mutation counts. The median mutation count in each cluster is indicated 
-by the horizontal line.
+**Figure 4: Somatic mutation frequencies for tumors in each class** 
+Column uses a logarithmic axis to represent mutation count. Row represents the
+tumor samples. Each dot corresponds to the sample mutation count for tumors in
+each class, ordered by mutation counts. The median mutation count in each
+cluster is indicated by the horizontal line.
 
-**Figure 5: Pan-cancer enrichment of pathway disruptions.** a) Waterfall plot; columns
-represent tumors. Each row represents a pathway. The pathways are ranked such that the 
-pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and 
-tumors are ordered to prioritize the top pathway disruptions successively. The top 100 
-most mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies
-not disrupted. b) Heatmap shows the relative enrichment of each pathway (columns) 
-within each numbered cluster (rows). Effect size is displayed as colors (orange, 
-enrichment; purple, depletion). c) Proportion of mutated samples in the each significant
-pathway (columns; union set of pathways with effect size >= 0.30 in each cluster)
-within each numbered cluster (rows).
+**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
+a) Waterfall plot; columns represent tumors. Each row represents a pathway. The
+pathways are ranked such that the pathway with the most disruptions across all
+7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
+top pathway disruptions successively. The top 100 most mutated pathways are
+shown. Red signifies pathway disrupted pathways, grey signifies not disrupted.
+b) Heatmap shows the relative enrichment of each pathway (columns) within each
+numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
+purple, depletion). c) Proportion of mutated samples in the each significant
+pathway (columns; union set of pathways with effect size >= 0.30 in each
+cluster) within each numbered cluster (rows).
 
-**Figure 6: Gene level analysis reveals tissue-specific class signatures.** Odds ratio 
-plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence 
-interval. Row represents the significant genes from each cluster. Each cluster was 
-compared against the background (all other clusters) to find deferentially mutated 
-genes. Significant genes (pvalue < 0.01) were selected and limited to the top ten 
-results for each clusters. Heatmap; Column represents the cancer type. Row represents the 
-significant genes as described above. The heatmap shows the proportion of samples 
-per cluster, per cancer type that is mutated for each of the significant genes. 
-Depleted significant genes (left). Enriched significant genes (right).
+**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
+Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
+95% confidence interval. Row represents the significant genes from each cluster.
+Each cluster was compared against the background (all other clusters) to find
+deferentially mutated genes. Significant genes (pvalue < 0.01) were selected and
+limited to the top ten results for each clusters. Heatmap; Column represents the
+cancer type. Row represents the significant genes as described above. The
+heatmap shows the proportion of samples per cluster, per cancer type that is
+mutated for each of the significant genes. Depleted significant genes (left).
+Enriched significant genes (right).
 
-**Figure 7: ** Comparison of different cancers, simulated and actual
-results results. A) To understand the risk model, we plot survival of
-four simulated cohorts. Compare survival of a group consisting of
-randomly selected 30 year old women (light blue) to an equal sized
-group of 30 year old women with cancer, k = 2 (light orange). To
-understand how the mix of ages alters the survival curve, we compare a
-randomly selected group of people with ages uniformly distributed from
-30 to 70 (dark blue) and a randomly selected group with uniformly
-distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a
-randomly selected group with correct age distribution for the given
-cancer (blue) to the model predicted survival for the given average k
-(red) and the actual survival for the patients in our dataset (CHOOSE
-COLOR HERE, NOT GREEN). As can be seen, Stomach Cancer (STAD),
-Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
-deadly. The modeled results fit the general trend for the actual
-patients. Thyroid Cancer (THCA) on the other hand has very little if
-any effect on life expectancy, in fact there may be a slight benefit
-which may be associated with a bias in diagnosis towards patients who
-are more health conscious, or higher socioeconomic status, with more
-access to care compared to random population members.
+**Figure 7: ** Comparison of different cancers, simulated and actual results
+results. A) To understand the risk model, we plot survival of four simulated
+cohorts. Compare survival of a group consisting of randomly selected 30 year old
+women (light blue) to an equal sized group of 30 year old women with cancer, k =
+2 (light orange). To understand how the mix of ages alters the survival curve,
+we compare a randomly selected group of people with ages uniformly distributed
+from 30 to 70 (dark blue) and a randomly selected group with uniformly
+distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
+selected group with correct age distribution for the given cancer (blue) to the
+model predicted survival for the given average k (red) and the actual survival
+for the patients in our dataset (CHOOSE COLOR HERE, NOT GREEN). As can be seen,
+Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all
+very deadly. The modeled results fit the general trend for the actual patients.
+Thyroid Cancer (THCA) on the other hand has very little if any effect on life
+expectancy, in fact there may be a slight benefit which may be associated with a
+bias in diagnosis towards patients who are more health conscious, or higher
+socioeconomic status, with more access to care compared to random population
+members.
 
-**Figure 8** Tissue specific effects are compared by showing the
-probability density for the posterior value of the k_tis
-multiplier. These values reflect the general deadliness of the cancer
-type across all clusters. GBM, PAAD, and SKCM stand out as having
-values in the 10 to 20 range with a wide degree of uncertainty. Other
-cancers such as Prostate (PRAD) and Thyroid (THCA) show values down
-near or even below 1 suggesting that these populations may have
-healthier lifestyles and more access to care. Most cancer types fall
-in the range 2-7 as might be expected.
+**Figure 8** Tissue specific effects are compared by showing the probability
+density for the posterior value of the $k_{tis}$ multiplier. These values
+reflect the general deadliness of the cancer type across all clusters. GBM,
+PAAD, and SKCM stand out as having values in the 10 to 20 range with a wide
+degree of uncertainty. Other cancers such as Prostate (PRAD) and Thyroid (THCA)
+show values down near or even below 1 suggesting that these populations may have
+healthier lifestyles and more access to care. Most cancer types fall in the
+range 2-7 as might be expected.
 
-**Figure 9** Class specific $k_class$ values shown for each cancer
-type. The overall $k$ value is the product of $k_tis \times
-k_class$. The class specific value shows the relative aggressiveness
-of that particular class within the tissue. For example class 1, 2,
-and 6 Glioblastoma (GBM) are apparently much less deadly than other
-types of GBM. A similar situation can be seen with Pancreatic Cancer
-(PAAD) where 1, 6, and 8 are less deadly compared to classes 7,
-9, 10. For some cancer types there are few data points and little
-ability to estimate values with precision (DLBC and KICH for example).
+**Figure 9** Class specific $k_{class}$ values shown for each cancer type. The
+overall $k$ value is the product of $k_{tis} \times k_{class}$. The class
+specific value shows the relative aggressiveness of that particular class within
+the tissue. For example class 1, 2, and 6 Glioblastoma (GBM) are apparently much
+less deadly than other types of GBM. A similar situation can be seen with
+Pancreatic Cancer (PAAD) where 1, 6, and 8 are less deadly compared to classes
+7, 9, 10. For some cancer types there are few data points and little ability to
+estimate values with precision (DLBC and KICH for example).
 

### 2020-08-16 text/discussion.Rmd

@@
 excess passenger mutations. This could be addressed with application of
-ever-increasingly sophisticated filtering likely passenger mutations [ _e.g._
+ever-increasingly sophisticated filtering of likely passenger mutations [ _e.g._
 @spliceAI; @deep-learning-clinical-variants]. In the future, we hope to

### 2020-08-18 code/supplemental_file.Rmd
Add supplemental file

---
runtime: shiny
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    code_folding: hide
    theme: flatly
    smart: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
suppressPackageStartupMessages(c(
library(shiny),
library(tidyverse),
library(SummarizedExperiment),
library(plot3D),
library(data.table)
))
```

## Interactive Supplemental File {.tabset}

### Cancer type
```{r shiny-widget, echo = FALSE}
## Code to get tumor sample id and associated cancer type
# load("../data/gdc_tumor_coding_complete.Rdata")
# tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
#   as.data.frame(.) %>%
#   dplyr::select(project_code, submitted_specimen_id) %>% 
#   mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
# save(tumor_samples, file = "../data/tumor_samples.Rda")

load("../data/tumor_samples.Rda") ## load tumor samples data

cancer_type <- tumor_samples %>% .$project_code %>% unique %>% ## get cancer type info
  .[order(.)] ## Order alphabetically

selectInput("cancer_type", label = "Cancer type",
              choices = cancer_type, selected = "BRCA")
```

```{r, fig.align='center'}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")

ditto_color <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#666666",
  "#AD7700", "#1C91D4", "#007756", "#D5C711",
  "#005685", "#A04700", "#B14380", "#4D4D4D",
  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
  "#00446B", "#803800", "#8D3666", "#3D3D3D")

cluster.colours <- c(kelly.colours[c(3:12)])
names(cluster.colours) <- c(1:10)

renderPlot({
  par(mfrow = c(1,2), mai = c(0.1, 0.1, 0.1, 0.1))
  par(mar = c(2,2,0,0))

  with(umap_cancertype %>% ## using umap_cancertype data
       filter(project_code == input$cancer_type), ## Only include selected cancer type
     scatter3D(plot_y, -plot_z, plot_x, ## Plot cancer type using umap 3d coors
                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
  title(xlab="UMAP 2", line=-3, cex = 1.2)

  ## Class frequency
  class_freq <- umap_cancertype %>% dplyr::select(clust_knn, project_code) %>% ## select relevant columns
    .[.$project_code %in% input$cancer_type,] %>% ## Only include selected cancer type
    mutate(clust_knn = as.character(.$clust_knn)) ## Convert cluster membership to character

  freq_by_class <- list() ## created list to store data in the for loop below

  for(n in seq_along(1:max(umap_cancertype$clust_knn))) { ## Seq along clusters 1-10
    df <- data.frame(clust_knn = n,
                     freq = nrow(class_freq[class_freq$clust_knn %in% n, ]), ## count number of tumor in each cluster
                     stringsAsFactors = FALSE)
    freq_by_class[[n]] <- df ## store data in list
  }
  freq_by_class <- freq_by_class %>% do.call(rbind, .) %>% ## rbind list
    mutate(clust_knn = factor(x = as.character(.$clust_knn), levels = as.character(c(1:max(umap_cancertype$clust_knn))))) ## turn to factor and relevel 
  par(mar = c(4,6,2,0) + 0.1)
  barplot(height = freq_by_class$freq / nrow(class_freq), ## freq / tot_n_cancertype 
          names.arg = c(1:10), col = cluster.colours,
          ylab = "Relative Proportion",
          las = 1) ## always horizontal axis
})

renderPlot({
  par(mfrow = c(1,2), mai = c(0.1, 0.1, 0.1, 0.1))
  par(mar = c(2,2,0,0))

  spec_cancer <- umap_cancertype %>% filter(project_code == input$cancer_type) %>% ## Filter for selected cancer type
    dplyr::select(plot_x, plot_y, plot_z, Subtype_Selected) ## Select relevant column 
  spec_cancer[is.na(spec_cancer)] <- "NA" ## Convert NA to character "NA"

  s_name <- unique(spec_cancer$Subtype_Selected) ## What are the subtype names
  s_name <- s_name[order(s_name,na.last=T)] ## Order the names ABC and have NA last

  s_color <- ditto_color[c(2:(length(s_name) + 1))]
  names(s_color) <- s_name
  ## color scheme
  with(spec_cancer,
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = s_color, pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
  title(xlab="UMAP 2", line=-3)
  legend("top", inset = c(0.12), legend=names(s_color[1:2]), col=s_color[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.17), legend=names(s_color[3:4]), col=s_color[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.22), legend=names(s_color[5:6]), col=s_color[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.27), legend=names(s_color[7:8]), col=s_color[7:8], pch = 20, bty = "n", horiz = T, cex = 1.2)

  # Subtype frequency
  class_freq <- umap_cancertype %>% dplyr::select(clust_knn, project_code, Subtype_Selected) %>%
    .[.$project_code %in% input$cancer_type,] %>%
    dplyr::select(-project_code) %>%
    mutate(clust_knn = as.character(.$clust_knn))
  class_freq[is.na(class_freq)] <- "NA"

  class_freq$clust_knn <- factor(class_freq$clust_knn, levels = 1:10)
  counts <- table(class_freq$Subtype_Selected, class_freq$clust_knn)

  for(n in 1:ncol(counts)) {
    message(n)
    tot_n_clust <- sum(counts[,n])
    counts[,n] <- round(counts[,n] / tot_n_clust, digits = 3)
  }
  sum(counts[,1])

  par(mar = c(4,6,2,0) + 0.1)
  barplot(height = counts,
          names.arg = c(1:10), col = s_color,
          ylab = "Relative Proportion",
          las = 1) ## always horizontal axis
})
```

### Gene
```{r gene-data-processing, eval = FALSE}
load("../data/gdc_gene_exp_adj.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/gene_data.Rda")

## Only include genes that were used in our pathway analysis
supplemental_gene <- gdc_gene_exp_adj %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>%
  rownames_to_column(var = "ensembl_gene_id") %>%
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")]) %>%
  dplyr::select(-ensembl_gene_id) %>%
  column_to_rownames(var = "external_gene_name") %>%
  t() %>% as.data.frame()

save(supplemental_gene, file = "../data/supplemental_gene.Rda")
rm(list = ls(all.names = T))
gc()
```

```{r gene-widet, echo = FALSE}
load("../data/supplemental_gene.Rda")

gene <- colnames(supplemental_gene) ## get gene names
gene <- gene[order(gene)] ## order alphabetically

selectInput("gene", label = "Gene",
              choices = gene, selected = "TP53")
```

```{r gene-plot}

renderPlot({
  par(mfrow = c(1,2), mai = c(0.1, 0.1, 0.1, 0.1)) ## creating plot layout 1 x 2
  par(mar = c(2,2,0,0))
  with(supplemental_gene %>% dplyr::select(input$gene) %>% ## filtering for gene
    "colnames<-" ("selectedgene") %>% ## rename column
    rownames_to_column(var = "sample_id") %>% ## move sample id to column
    left_join(umap_cancertype[,c("plot_x", "plot_y", "plot_z", "sample_id")], ., by = "sample_id") %>% ## left join data by sample_id
    filter(selectedgene == 1), ## only select tumor samples that have a mutation in the selected gene
    scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
  title(xlab="UMAP 2", line=-3, cex = 1.2)
})
```

### Pathway
```{r pathway-widget}
load("../data/gdc_reactome_path_analysis.Rda")

pathway <- rownames(gdc_reactome_path_analysis) ## get pathway names
pathway <- pathway[order(pathway)] ## order alphabetically

selectInput("pathway", label = "Pathway",
              choices = pathway, selected = "Cellular response to hypoxia")
```

```{r pathway-plot}

renderPlot({
  par(mfrow = c(1,2), mai = c(0.1, 0.1, 0.1, 0.1)) ## creating plot layout 1 x 2
  par(mar = c(2,2,0,0))
  with(gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% ## transpose data for left_joining later
    dplyr::select(input$pathway) %>% ## filtering for pathway
    "colnames<-" ("selectedpath") %>% ## rename column
    rownames_to_column(var = "sample_id") %>% ## move sample id to column
    left_join(umap_cancertype[,c("plot_x", "plot_y", "plot_z", "sample_id")], ., by = "sample_id") %>% ## left join data by sample_id
    filter(selectedpath == 1), ## only select tumor samples that have a mutation in the selected gene
    scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
  title(xlab="UMAP 2", line=-3, cex = 1.2)
})
```

### 2020-08-19 code/figure2v2.Rmd
Clean and annotate code for figure 1-3, add fig legend for figure 1, remove a small text from figure 2 text

--- 2020-08-13 code/figure2v2.Rmd
@@
-```{r figure-2-data-generation, eval = FALSE}
-####################
-## Figure 1B data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>%
-  t() %>%
-  as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(tumor_samples$submitted_sample_id, .$sample), ]
+```{r figure-2-plot, fig.width = 9}
+load("../data/figure1B_data.rda") ## Load in data processed in figure1.Rmd
+load("../data/optimal_minpts_score.rda")
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
+load("../data/kelly.colours.rda")
+load("../data/ditto.colours.rda")
 
-gdc_path_mca <- left_join(gdc_path, tumor_samples[, c("project_code", "submitted_sample_id")],
-  by = c("sample" = "submitted_sample_id")) %>%
-  column_to_rownames(var = "sample")
-
-gdc_path_sample_names <- rownames(gdc_path_mca) ## Getting sample id order
-gdc_path_tumor_types <- gdc_path_mca$project_code ## Getting cancer type order
-
-gdc_path_mca$project_code <- NULL ## Remove project_code column from df
-
-## create a categorical variable from the boolean variable
-pathway_list <- make_clean_names(colnames(gdc_path_mca))
-pathway_count <- length(pathway_list)
-gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
-  factor(x, levels = c(0, 1), labels = c("no", "yes"))
-}))
-colnames(gdc_path_mca) <- pathway_list
-gdc_path_mca$project_code <- gdc_path_tumor_types
-rownames(gdc_path_mca) <- gdc_path_sample_names
-
-## perform mca
-mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
-
-expl_var <- mca_res$eig[, 2]
-expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
-
-set.seed(4242)
-for (k in seq(100)) {
-  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
-  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
-  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
-}
-
-randomize <- tibble(
-  eigenvalue = 1:length(expl_var),
-  expl_var = expl_var,
-  perm = colMeans(expl_var_perm),
-  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
-)
-
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-mca_res <- MCA(
-  X = gdc_path_mca[, pathway_list],
-  ncp = optimal_mca_dimensions, graph = FALSE
-)
-
-save(randomize, mca_res, file = "figure1B_data.rda")
-
-####################
-## Figure 1C data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-rng <- RNGseq(max_n_pt * n_iter, 42)
-r <- rng[[3649]]
-
-rngtools::setRNG(r)
-gdc.umap <- umap(mca_res$ind$coord, 
-                 n_components = 3,
-                 n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
-                 n_epochs = 1000,
-                 n_threads = 1)
-gdc_umap_coors <- as.data.frame(gdc.umap) %>%
-  rename(plot_x = V1, plot_y = V2, plot_z = V3)
-rownames(gdc_umap_coors) <- rownames(gdc_path_mca)
-
-####################
-## Figure 1D data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
-
-n_tumors <- nrow(mca_res$ind$coord)
-
-cl <- makeCluster(8)
-registerDoParallel(cl)
-
-for (i in 3:max_n_pt) {
-  print(paste0("MIN POINTS = ", i))
-  score_iter <- foreach(1:n_iter,
-    .combine = c,
-    .export = c("mca_res"),
-    .packages = c("dbscan", "uwot"),
-    r = rng[(i - 1) * n_iter + 1:n_iter]
-  ) %dopar% {
-    rngtools::setRNG(r)
-    umap_init_mca <- umap(mca_res$ind$coord, 
-                          min_dist = 1e-6, 
-                          n_components = 3, 
-                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
-                          n_epochs = 1000, 
-                          n_threads = 1)
-    res <- hdbscan(umap_init_mca, minPts = i)
-    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
-    return(score_iter_temp)
-  }
-  score_mca[i - 2, ] <- score_iter
-}
-
-score <- rowMeans(score_mca)
-score_df <- tibble(minPts = 3:max_n_pt, score = score)
-
-sample_df <- score_mca %>%
-  as.data.frame() %>%
-  as_tibble() %>%
-  add_column(minPts = 3:max_n_pt, .before = 1) %>%
-  pivot_longer(-minPts) %>%
-  group_by(minPts)
-
-optimal_minPts <- score_df %>%
-  filter(score == min(score)) %>%
-  pull(minPts)
-
-save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
-
-####################
-## Figure 1E data ##
-####################
-rngtools::setRNG(r)
-
-cluster_umap <- umap(mca_res$ind$coord,
-  min_dist = 1e-6,
-  n_components = 3,
-  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
-  n_epochs = 1000,
-  n_threads = 1
-)
-colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
-hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
-umap_3d_coors <- gdc_umap_coors %>%
-  bind_cols(as.data.frame(cluster_umap)) %>%
-  mutate(
-    n_pathways_mut = rowSums(gdc_path[, -1]),
-    clust = as.factor(hdb_cluster_assignment$cluster),
-    member_prob = hdb_cluster_assignment$membership_prob,
-    sample_id = rownames(gdc_umap_coors)
-  )
-## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_3d_coors$clust_knn <- umap_3d_coors$clust
-umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
-  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
-  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
-  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
-)
-```
-
-```{r figure-2-plot, fig.width = 9}
-load("../data/figure1B_data.rda")
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-ditto_color <- c(
-  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
-  "#0072B2", "#D55E00", "#CC79A7", "#666666",
-  "#AD7700", "#1C91D4", "#007756", "#D5C711",
-  "#005685", "#A04700", "#B14380", "#4D4D4D",
-  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
-  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
-  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
-  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
-  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
-  "#00446B", "#803800", "#8D3666", "#3D3D3D")
-cluster.colours <- c("grey", kelly.colours[c(3:12)])
-names(cluster.colours) <- c(0:10)
-load("../data/optimal_minpts_score.rda")
-
-opar <- par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(4,6,2,0) + 0.1)
-with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = expl_var / 100, 
+opar <- par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
+par(mar = c(4,6,2,0) + 0.1) ## setting plot margins for the following plots
+## Figure 2A: Explained variance of eigenvalues
+with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = expl_var / 100, ## plotting the explained variance of observed
                                                  type = "l", yaxt = "n", bty = "l",
@@
                                                  xlab = "", ylab = ""))
-with(randomize %>% filter(eigenvalue < 50), lines(x = eigenvalue, y = perm / 100,
+with(randomize %>% filter(eigenvalue < 50), lines(x = eigenvalue, y = perm / 100, ## plotting the explained variance of permuted
                                                   col = "#f1a340"))
 yticks_val <- scales::breaks_pretty(n=5)(randomize %>% filter(eigenvalue < 50) %>% pull(expl_var) / 100)
-axis(2, at=yticks_val, lab=scales::percent(yticks_val), las = 1)
+axis(2, at=yticks_val, lab=scales::percent(yticks_val), las = 1) ## Axis labels
 title(ylab = "explained variance", line = 3.5)
@@
        col=c("#998ec3", "#f1a340"), lty=1, cex=0.8)
-### D
-par(mar = c(2,2,0,0), bg = NA)
+mtext("A", side=3, line=0, cex=1.2, adj=-0.35) ## plot tag
+
+par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
+## Figure 2C: MCA projected down to 3 dimensions (UMAP)
 with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
@@
 title(xlab="UMAP 2", line=0)
-### E
-par(mar = c(4,6,2,0) + 0.1)
+mtext("C", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
+
+par(mar = c(4,6,2,0) + 0.1) ## setting plot margins for the following plots
+## Figure 2D: Optimization of HDBscan
 with(sample_df, plot(x = minPts, y = value, 
@@
                      col = "black", las = 1))
-abline(v = optimal_minPts, col = "red")
+abline(v = optimal_minPts, col = "red") ## Red vertical line to show optimal minimum point for HDBscan
 title(ylab = "score", line = 3.5)
@@
 axis(2, at=c(0.25, 0.35, 0.45, 0.55), las = 1)
+mtext("D", side=3, line=0, cex=1.2, adj=-0.27) ## plot tag
 
-### C
-with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = pvalue, 
+## Figure 2B: pvalue of eigenvalue 
+with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = pvalue,  ## line plot
                                                  type = "l", bty = "l",
@@
                                                  xlab = "", ylab = ""))
-with(randomize %>% filter(eigenvalue < 50), points(x = eigenvalue, y = pvalue, pch = 20,
+with(randomize %>% filter(eigenvalue < 50), points(x = eigenvalue, y = pvalue, pch = 20, ## add dots to line plot
                                                    col = "black"))
-abline(v = optimal_mca_dimensions, col = "red")
-title(main = paste0("Significant dimensions = ", optimal_mca_dimensions), adj = 0, font = 3, cex.main = .8)
+
+## Determining the optimal mca dimension
+optimal_mca_dimensions <- randomize %>%
+  filter(pvalue < 0.05) %>% ## filtering for pvalue < 0.05
+  filter(eigenvalue == max(eigenvalue)) %>% ## filtering for the maximum eigenvalue
+  pull(eigenvalue)
+
+abline(v = optimal_mca_dimensions, col = "red") ## Red vertical line to show optimal MCA dimension
+title(main = paste0("Significant dimensions = ", optimal_mca_dimensions), adj = 0, font = 3, cex.main = .8) ## Axis labels
 title(ylab = "P-value", line = 3.5)
 title(xlab = "MCA Eigenvalues", line = 2)
-### F
-par(mar = c(2,2,0,0), bg = NA)
+mtext("B", side=3, line=0, cex=1.2, adj=-0.35) ## plot tag
+
+par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
+## Figure 2E: HDBscan
+cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
+names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
 with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = cluster.colours[as.character(clust)], pch = 21, cex = 0.8, lwd = 0.2,
+                                bg = cluster.colours[as.character(clust)], pch = 21, cex = 0.8, lwd = 0.2, # clust is hdbscan result
                                 theta = 0, phi = 65, scale = F,
@@
 title(xlab="UMAP 2", line=0)
-### G
+mtext("E", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
+
+## FIgure 2F: kNN 
 with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
                                 theta = 0, phi = 65, scale = F,
@@
 title(xlab="UMAP 2", line=0)
+mtext("F", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
 
-plot.new()
 ```

### 2020-08-19 code/figure3v2.Rmd
Clean and annotate code for figure 1-3, add fig legend for figure 1, remove a small text from figure 2 text

@@
 umap_cancertype <- fread("../data/umap_3d_coors.tsv")
+load("../data/kelly.colours.rda")
 
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-## mutation burden plot
+## Arranging mutation freq for each cluster
 mut_order <- list() ## store data in list 
 for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
-  message(n)
   mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
@@
 
+## Convert sample_id and clust_knn into factor
 mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
@@
 
-## Need the breaks positions for the x-axis
+## Determing the breaks positions for the x-axis
 previous_class <- 0 ## starting position 1
 position_of_previous_class <- 0 ## starting position 2
-xaxis_breaks <- list() ## list for for loop
-for(n in 1) { ## loop through 1 to 12 clusters
-  message(n)
+xaxis_breaks <- list() ## list for the for loop
+for(n in 1) { ## loop through 1 to 10 clusters
   class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
@@
   for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
-    message(n)
     position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
@@
 
+## Calculate median for each class
 mut_median <- mut_order %>% 
   group_by(clust_knn) %>% 
-  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
+  dplyr::summarize(mut_med = median(log10mut))
 
-## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
+## Need to add 100 pseudo samples front and end of data so they don't touch the y-axis
 mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column
@@
   theme_classic() +
-  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") +
-  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
+  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") + ## facet by cluster
+  geom_hline(data= mut_median, aes(yintercept=mut_med)) + ## add median line
   scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
@@
 fig1G <- ggplotGrob(fig1G)
-lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
+
+lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), ## All lines off expect bottom line of facet
                    gp=gpar(col="black", lwd=4))
 
-for (k in grep("strip-t",fig1G$layout$name)) {
+for (k in grep("strip-t",fig1G$layout$name)) { ## turn off 3 lines from the facet border text
   fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg

### 2020-08-20 code/figure2.Rmd
Annotate code for fig 2-5, add interactive 3dplot to supplemental_file, delete some rmd not used

@@
 ```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
+load("../data/gdc_reactome_path_analysis.Rda") ## load data
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
+load("../data/moderate_enrich_prob.Rda") 
+load("../data/kelly.colours.rda")
 
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
+### Figure A - oncoprint style top 100 pathways
 
@@
 top100path[top100path == "0"] <- " "
-col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
+
 onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
+                      remove_empty_columns = F, remove_empty_rows = F,
+                      right_annotation = NULL, top_annotation = NULL)
 
@@
 
-###################
-## barplot panel ##
-###################
+### Figure B - Pathway enrichment by cluster
 
-load("../data/kelly.colours.rda")
-load("../data/moderate_enrich_prob.Rda")
+## Getting the total number of tumor samples in each cluster
+n_tumor_in_clust <- as.data.frame(table(umap_cancertype$clust_knn)) %>% ## table the number of tumors in each cluster
+  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
+  mutate(clust = as.character(clust)) ## convert factor to character
+
+## Organizing rBeta data
+load("../data/gdc_reactome_path_analysis.Rda")
+clust_group <- n_tumor_in_clust$clust
+cluster_path_count <- list()
+for(n in clust_group) {
+  sample_id <- umap_cancertype %>% filter(clust_knn == n) %>% ## filter for tumor class
+    .[,.$sample_id] ## get sample_ids
+  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
+    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
+    'colnames<-' (n) ## change colname to cluster number
+  cluster_path_count[[n]] <- path_mutation
+  rm(path_mutation, sample_id, n)
+  gc()
+}
+cluster_path_count <- as.data.frame(do.call(cbind, cluster_path_count)) ## Combine the list together
+
+global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
+  .[rep(1:ncol(.), each = max(umap_cancertype$clust_knn))] %>% ## replicate the column 0 times for the 10 tumor classes
+  'colnames<-' (clust_group) %>% ## rename column with tumor class
+  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
+  gather(., key = "clust", value = "global_mut", -path)
+
+rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
+  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
+  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
+  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
+  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
+
+rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
+gc()
+
+enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
+                                            clust = rbeta_path_analysis$clust,
+                                            fg_mut = rbeta_path_analysis$clust_mut,  
+                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
+                                            bg_mut = rbeta_path_analysis$global_mut,
+                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
+                                            )
+
+## Calculating the foreground and background mutation rate for each cluster
+fg_mut_rate <- list()
+bg_mut_rate <- list()
+for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
+  df <- subset(enrich_prob_uncertainity_in_bg, clust == n) ## Only include pathways in selected cluster
+  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust
+  fg_numerator <- sum(df$fg_mut) ## sum of pathway mutation in selected cluster
+  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1]) ## n_path * tot_n_clust
+  bg_numerator <- sum(df$bg_mut) ## sum of no pathway mutation in selectedcluster
+  fg_mut_rate[[n]] <- fg_numerator / fg_denominator ## calculate mutation rate and store in list
+  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
+  rm(df)
+}
+
+fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
+  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
+  rownames_to_column(var = "clust") ## move rownames to col
+bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
+  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
+  rownames_to_column(var = "clust") ## move rownames to col
+
+enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>% 
+  left_join(., bg_mut_rate, by = "clust") %>% ## add fg and bg mut rate to enrich df
+  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>% ## calculating relative odds ratio
+  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
+  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
+  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
+rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
+gc()
+
+## beta distribution for 1e5 simulations for foreground and background
+moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"]) 
+})
+
+moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
+})
+
+## ModerateEnrichment probability 
+prob <- moderate_fg > moderate_bg ## boolean data
+mean_prob <- apply(prob, MARGIN = 2, function(x) {
+  mean(x) ## calculating the mean probability
+})
+rm(prob)
+gc()
+## Moderate Difference
+CI <- 0.025 ## setting a 95% confidence interval
+
+diff_fg_bg <- moderate_fg - moderate_bg ## Calculating the difference between fg and bg for the next function 
+rm(moderate_fg, moderate_bg)
+gc()
+
+sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
+}))
+quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  quantile(x, c(CI, 0.5, 1-CI)) ## calculating the 1st quantile, median, and 3rd quantile
+}))
+
+moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
+                                   clust = enrich_prob_uncertainity_in_bg$clust,
+                                   probability = mean_prob, ## probabilty of pathway mutation
+                                   signif = sig_fg_bg[,1], ## how many times fg-bg is less than 0
+                                   difference = quant_diff[,2], ## effect size
+                                   lower = quant_diff[,1], ## 1st quantile
+                                   upper = quant_diff[,3]) ## 3rd quantile
+rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
+gc()
+save(moderate_enrich_prob, file = "../data/moderate_enrich_prob.Rda") ## save result
+
+## Processing data for effect size heatmap
+diff_heatmap <- moderate_enrich_prob %>% dplyr::select(path, clust, difference) %>% ## select relevant column
+  spread(., key = "clust", value = "difference") %>% ## long to wide format
+  remove_rownames() %>% column_to_rownames(var = "path") ## pathway names to rownames
+
+## Determining pathway order for heatmap
+effect_size <- pheatmap(diff_heatmap,
+         cluster_cols = T,
+         cluster_rows = T,
+         clustering_method = "ward.D2")
+## Extracting row order from pheatmap
+effect_row_order <- effect_size$tree_row$order
+effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames()
+
+## Revert df back to long format
+effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>%  ## pathway names to column
+  gather(., key = "cluster", value = "effect_size", -path_name)  ## wide to long
+
+## Factor column to force order when generating plot
+effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))  ## pathway name ordering
+effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:max(as.numeric(effect_heatmap$cluster)))) ## cluster ordering
+effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)) ## Creating breaks for heatmap legend (color scale)
+
+effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when( ## create discrete variable using continuous
+  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
+  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
+  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
+  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
+  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
+  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
+  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
+  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
+  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
+  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
+  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
+))
+effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11)) ## Factor order the discrete color
+
+save(effect_heatmap, file = "../data/effect_heatmap.Rda")
+
+### Figure C - barplot panel
 
@@
   group_by(clust) %>% 
-  filter(difference > 0.3 & signif <= 2500) %>%
+  filter(difference > 0.3 & signif <= 2500) %>% ## Filter pathways that have an effect size > 0.30 with a 95% CI (signif 2500/1e5)
   mutate(pathway = make_clean_names(path)) %>% 
-  rename(clust_knn = "clust") %>%
-  select(clust_knn, pathway) %>%
+  rename(clust_knn = "clust") %>%  ## rename clust column to clust_knn
+  select(clust_knn, pathway) %>% ## select relevant columns
   mutate(clust_knn = as.numeric(clust_knn)) %>%
-  mutate(specific = TRUE)
+  mutate(specific = TRUE) ## Add "specific" column with values TRUE
 
-pathway_counts <- tumor_clusters %>%
-  left_join(t(gdc_reactome_path_analysis) %>% 
+pathway_counts <- umap_cancertype %>%
+  left_join(t(gdc_reactome_path_analysis) %>% ## combine umap_cancertype and gdc_path_analysis by sample_id
               as.data.frame() %>% 
@@
   group_by(clust_knn) %>% 
-  summarise(total = n(), across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, sum)) %>%
-  mutate(across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, ~ .x / total)) %>%
-  select(-total) %>%
-  pivot_longer(-clust_knn, names_to = "pathway", values_to = "proportion") %>%
+  summarise(total = n(), across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, sum)) %>% ## Calculate the total_n_clust and total pathway mutation across cluster
+  mutate(across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, ~ .x / total)) %>% ## Divide sum mutation of cluster / total_n_clust to get percent mutation
+  select(-total) %>% ## Remove total column
+  pivot_longer(-clust_knn, names_to = "pathway", values_to = "proportion") %>% ## wide to long format
   group_by(clust_knn) %>% 
-  filter(pathway %in% enrich_pathways$pathway) %>%
+  filter(pathway %in% enrich_pathways$pathway) %>% ## only include pathways that are enriched
   left_join(enrich_pathways %>% select(clust_knn, pathway, specific), by = c("clust_knn", "pathway"))
 
-pathway_counts[is.na(pathway_counts$specific), ]$specific <- FALSE
+pathway_counts[is.na(pathway_counts$specific), ]$specific <- FALSE ## Clusters not enriched for a pathway (NA) convert to FALSE
 
+## Determine the order of pathways on the barplot
 pathway_mat <- pathway_counts %>%
   select(-specific) %>%
-  pivot_wider(names_from = pathway, values_from = proportion) %>%
+  pivot_wider(names_from = pathway, values_from = proportion) %>% ## long to wide format
   ungroup() %>%
@@
 
-pathway_dist <- dist(t(pathway_mat))
-pathway_clust <- as.hclust(rev(as.dendrogram(seriate(pathway_dist, method = "OLO_ward")[[1]])))
-pathway_counts$pathway <- factor(pathway_counts$pathway, levels = colnames(pathway_mat)[pathway_clust$order])
-bp.cluster.colors <- c(kelly.colours[3:12], "gray15")
-names(bp.cluster.colors) <- 1:11
+pathway_dist <- dist(t(pathway_mat)) ## Computing distance matrix of all values
+pathway_clust <- as.hclust(rev(as.dendrogram(seriate(pathway_dist, method = "OLO_ward")[[1]]))) ## cluster pathways based on distance matrix
+pathway_counts$pathway <- factor(pathway_counts$pathway, levels = colnames(pathway_mat)[pathway_clust$order]) ## force pathway order by factoring
 
-pathway_counts <- pathway_counts %>%
-  mutate(color = case_when(specific ~ clust_knn, TRUE ~ 11)) %>%
+bp.cluster.colors <- c(kelly.colours[3:12], "gray15") ## Color scheme for plot
+names(bp.cluster.colors) <- 1:11 ## Cluster 11 is
+
+pathway_counts1 <- pathway_counts %>%
+  mutate(color = case_when(specific ~ clust_knn, TRUE ~ 11)) %>% ## Color: specific == FALSE is 11; specific == TRUE is its cluster
   mutate(color = factor(color)) 
 save(pathway_counts, bp.cluster.colors, file = "barplot_pathways_fig2b.rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:max(tumor_clusters$clust_knn)) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("sample_id", "clust_knn")] %>% filter(clust_knn == n) %>% .[,.$sample_id] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
 ```
@@
 ```{r figure-5-plot, fig.align= 'center', fig.height= 7, fig.width = 7}
-#out.height='90%', out.width='90%'
-## top100 pathways pancancer oncoprint
+
+### Figure A - Waterfall plot of top 100 mutated pathways
 load("../data/onco_ggplot.Rda") ## load data
@@
 
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
+fig5A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
   geom_tile(aes(fill = as.character(mutation_status))) +
@@
         axis.title.y = element_text(size = 10),
-        #plot.tag = element_text(size = 8, hjust = 0, vjust = 0),
         axis.ticks = element_blank(),
@@
 
+### Figure B - Effect size heatmap
 load("../data/effect_heatmap.Rda") ## load data
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
 
+## Function to duplicate discrete axis labels
 guide_axis_label_trans <- function(label_trans = identity, ...) {
@@
 
-fig2B <- ggplot(data = effect_heatmap, aes(y = cluster, x = path_name)) +
+fig5B <- ggplot(data = effect_heatmap, aes(y = cluster, x = path_name)) +
   geom_tile(size = 1, aes(fill = effect_size)) +
-  scale_fill_fermenter(
+  scale_fill_fermenter( ## custom color scale
     type = "div",
@@
     n.breaks = 12,
-    limits = c(-1, 1),
+    limits = c(-1, 1), ## Set scale range
     name = "Effect Size",
@@
   ) +
-  facet_wrap(~cluster, ncol = 1, strip.position = "right", scales = "free_y") +
+  facet_wrap(~cluster, ncol = 1, strip.position = "right", scales = "free_y") + ## facet by cluster
   guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
@@
   ylab("Clusters") +
-  scale_x_discrete(position = "top") +
-  guides(y.sec = guide_axis_label_trans(~ paste(.x))) +
+  scale_x_discrete(position = "top") + ## move x labels to top
+  guides(y.sec = guide_axis_label_trans(~ paste(.x))) + ## Duplicate axis.text on right side (cluster)
   theme(
     axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
+    axis.text.y = element_blank(), ## turn off y-axis text
     axis.title.x = element_text(size = 10),
@@
     strip.text.y = element_text(size = 10, hjust = .5),
-    strip.text.y.right = element_text(angle = 0),
+    strip.text.y.right = element_text(angle = 0), ## Fix orientation of dupped right y-axis text
     strip.background = element_blank(),
@@
     panel.spacing = unit(0, "lines"),
+    ## legend parameters
     legend.position = "bottom",
@@
 
-load("../data/barplot_pathways_fig2b.rda")
-fig2C <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
+### Figure B - Effect size heatmap
+load("../data/barplot_pathways_fig2b.rda") ## load data
+fig5C <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
   geom_col() +
-  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
+  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") + ## facet by cluster and move facet label to the right
   ylab("Proportion of mutated samples") +
@@
     axis.line = element_blank(),
-    #strip.text.y = element_text(size = 10, hjust = .5),
-    #plot.tag = element_text(size = 12, hjust = 0, vjust = 0),
     strip.text.y.right = element_text(angle = 0),
@@
 
+## Layout for plots using patchwork
 layout <- '

### 2020-08-20 code/figure3v2.Rmd
Annotate code for fig 2-5, add interactive 3dplot to supplemental_file, delete some rmd not used

@@
 
-## Determing the breaks positions for the x-axis
+## Determining the breaks positions for the x-axis
 previous_class <- 0 ## starting position 1

### 2020-08-20 code/supplemental_file.Rmd
Annotate code for fig 2-5, add interactive 3dplot to supplemental_file, delete some rmd not used

@@
 library(plot3D),
-library(data.table)
+library(data.table),
+library(rgl),
+library(htmltools),
+library(dplyr),
+library(rgl),
+library(htmltools)
 ))
@@
 ## Interactive Supplemental File {.tabset}
+
+### UMAP cluster
+```{r}
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
+load("../data/kelly.colours.rda")
+
+cluster.colours <- c(kelly.colours[c(3:12)], "grey") ## Color scheme
+names(cluster.colours) <- c(1:11) ## Assign cluster to color
+
+ui <- (fluidPage(
+#sliderInput("sld", label = "Slider", min = 0., max = 4*pi, value = 0., step = .1),
+playwidgetOutput("player"),
+ rglwidgetOutput("plot",  width = 500, height = 350)
+))
+
+server <- (function(input, output) {
+
+  output$plot <- renderRglwidget ({
+    open3d(useNULL=TRUE)
+    plot3d(umap_cancertype[, c("plot_x", "plot_y", "plot_z"),],
+  type = "s", size = 2.5,
+  xlab = "UMAP 1", ylab = "UMAP 2", zlab = "UMAP 3",
+  aspect = c(8.3, 10.5, 7.6),
+  col = cluster.colours[as.character(umap_cancertype$clust_knn)])["data"]
+    rglwidget()
+
+    })
+
+  })   
+shinyApp(ui = ui, server = server)
+```
 
@@
   "#00446B", "#803800", "#8D3666", "#3D3D3D")
-
-cluster.colours <- c(kelly.colours[c(3:12)])
-names(cluster.colours) <- c(1:10)
 

### 2020-08-20 text/manuscript.Rmd
Annotate code for fig 2-5, add interactive 3dplot to supplemental_file, delete some rmd not used

@@
   library(plot3D),
-  library(ggfortify)
+  library(ggfortify),
+  library(maftool),
+  library(survival)
 ))

### 2020-08-20 code/table1.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

--- 2020-07-21 code/table1.Rmd
@@
-<div style="text-align: right"> Table 1 Data </div>
 ```{r table-1-data-generation, eval = FALSE}
@@
 
-cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
+cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(row.names = FALSE) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%

### 2020-08-20 text/discussion.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

@@
 into pathway level disruptions. Our approach differs from previously described
-methods [@PATHNET] in that we chose to focus on the pathway as the unit of
-disruption instead of the gene, employing the conceit that individual mutations
-may be sufficient to disrupt or misdirect pathway activity (as is often the case
-with driver mutations). Rather than evaluating the burden of mutations on each
-pathway within or across tumors, which can be biased by the number of genes or
-the coverage of pathway-linked genes in the genome, we scored each pathway in
-binary fashion, allowing one mutation or multiple mutations to count equally for
-each sample.
+methods [@PATHNET] in that we chose to focus on the pathway (defined in the
+methods) as the unit of disruption instead of the gene, employing the conceit
+that individual mutations may be sufficient to disrupt or misdirect pathway
+activity (as is often the case with driver mutations). Rather than evaluating
+the burden of mutations on each pathway within or across tumors, which can be
+biased by the number of genes or the coverage of pathway-linked genes in the
+genome, we scored each pathway in binary fashion, allowing one mutation or
+multiple mutations to count equally for each sample.
 
@@
 mutations with likely deleterious effects, including missense, nonsense,
-frame-shift, stop-gain, untranslated region and splice site mutations in genes
+frame-shift, stop-loss, untranslated region and splice site mutations in genes
 that are actively expressed in each cancer type, thus avoiding bias from
@@
 
-The hypothesis that cancer is the result of dysfunction from
-a limited number of basic cellular processes common to all living human cells
-was first introduced and later expanded on in a pair of classic essays by
-Hanahan and Weinberg [@weinberg2000; @hallmarks_II]. In this view, all tumors
-regardless of tissue of origin acquire six essential phenotypes; sustaining
-proliferative signaling, evading growth suppressors, replicative immortality,
-evading apoptosis, supportive angiogenesis, tissue invasion and metastasis
-[@weinberg2000]. Advancements in the field over the ensuing decade led to the
-inclusion of new tumor promoting processes, namely deregulation of cellular
-energy cycle, tumor promoting inflammation, evasion of the immune system, genome
-instability, and on top of these acquired traits a new appreciation for the
-concept of the tumor microenvironment and its contribution to the evolution and
-promotion of metastatic disease [@hallmarks_II].
+The hypothesis that cancer is the result of dysfunction from a limited number of
+basic cellular processes common to all living human cells was first introduced
+and later expanded on in a pair of classic essays by Hanahan and Weinberg
+[@weinberg2000; @hallmarks_II]. In this view, all tumors regardless of tissue of
+origin acquire six essential phenotypes; sustaining proliferative signaling,
+evading growth suppressors, replicative immortality, evading apoptosis,
+supportive angiogenesis, tissue invasion and metastasis [@weinberg2000].
+Advancements in the field over the ensuing decade led to the inclusion of new
+tumor promoting processes, namely deregulation of cellular energy cycle, tumor
+promoting inflammation, evasion of the immune system, genome instability, and on
+top of these acquired traits a new appreciation for the concept of the tumor
+microenvironment and its contribution to the evolution and promotion of
+metastatic disease [@hallmarks_II].
 
@@
 manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways as represented by our classes (**figures 1 & 2**).
+disruptions in basic pathways as represented by our clusters (**figures 1 & 2**).
 
 We attempted to account for our clusters in terms of more trivial explanations.
-For example, it could be that the classes we identified are consistent with a
-signature of overall disease progression, such that successive, adjacent classes
-on the UMAP projection exhibit increasing specificity. We identified no such
-trend in the number of mutations (see **figure 4**), the relative staging
-(**figure 3D**) or metastasis (**figure 3E**); and each class instead was
+For example, it could be that the clusters we identified are consistent with a
+signature of overall disease progression, such that successive, adjacent
+clusters on the UMAP projection exhibit increasing specificity. We identified no
+such trend in the number of mutations (see **figure 4**), the relative staging
+(**figure 3D**) or metastasis (**figure 3E**); and each cluster instead was
 associated with unique combinations of pathways (as in **figures 5B, C**).
@@
 S1**) are unevenly distributed, though we could not identify any cancers that
-were exclusive to a single tumor class. Only cancer types with the fewest
-samples were found to be absent from one or more classes at all. Somewhat
+were exclusive to a single tumor cluster. Only cancer types with the fewest
+samples were found to be absent from one or more clusters at all. Somewhat
 surprisingly to us, this finding extends to histological subtypes of breast,
@@
 7, and basal tumor samples were biased toward cluster 6 (**figure 3B**).
-However, both subtypes also contained samples in every other class (without
+However, both subtypes also contained samples in every other cluster (without
 exception), and Luminal B and Her2 positive samples are distributed across
@@
 molecular profiles may predispose certain precursor cell types within the breast
-epithelium to forming tumors of one class or another but are not determinative;
-this view is compatible with the previously stated hypothesis but opens the way
-for a more granular view of individual tumors. It would be surprising if we did
-not observe bias for some cancers and subtypes amongst our classes, since some
-treatment regimens have greater efficacy for patients of a given cancer or
-histological subtype [@breast-subtypes2]. Nonetheless, the basis for some tumors
-being treatment-refractory in spite of receiving the standard of clinical care
-for diagnostic markers remains elusive. Doubtless some of this is due to bad
-luck, as tumors can metastasize and remain dormant years before they are
-detected at distal sites, or resistant clones may have already arisen at
+epithelium to forming tumors of one cluster or another but are not
+determinative; this view is compatible with the previously stated hypothesis but
+opens the way for a more granular view of individual tumors. It would be
+surprising if we did not observe bias for some cancers and subtypes amongst our
+classes, since some treatment regimens have greater efficacy for patients of a
+given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
+for some tumors being treatment-refractory in spite of receiving the standard of
+clinical care for diagnostic markers remains elusive. Doubtless some of this is
+due to bad luck, as tumors can metastasize and remain dormant years before they
+are detected at distal sites, or resistant clones may have already arisen at
 undetectable levels [@hallmarks_II], but our analysis suggests the possibility
@@
 Finally, we must remark on the limitations of our work exemplified in two of the
-classes, 1 and 10, for which we did not find many distinctive associations with
-pathways (**figure 5B, C**). Class 1 had a relatively low proportion of mutated
-pathways across the board, although it is broadly enriched in many of the same
-tumor-promoting pathways common to the other groups. This likely does not
-reflect a true difference in stages of tumor evolution, as our data show clearly
-that this class is as likely to contain stage IV metastatic tumors as it is to
-contain those of stage I (**figure 3D, E**). This class likely represents a
-group of tumors with aberrations in methylation, copy number alterations,
-structural variants, or combinations thereof which we excluded from our
-analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
-proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and
-are heavily skewed to cluster 1 membership (**Supplementary figure S1**).
-Likewise, Class 10 represents a group of hyper-mutated tumors that harbor so
-many mutations that virtually no pathway is unaffected. It seems likely that a
-significant fraction of the “mutant” samples for each pathway are burdened with
-excess passenger mutations. This could be addressed with application of
-ever-increasingly sophisticated filtering of likely passenger mutations [ _e.g._
+clusters, 1 and 10, for which we did not find many distinctive associations with
+pathways (**figure 5B, C**). Cluster 1 had a relatively low proportion of
+mutated pathways across the board, although it is broadly enriched in many of
+the same tumor-promoting pathways common to the other groups. This likely does
+not reflect a true difference in stages of tumor evolution, as our data show
+clearly that this cluster is as likely to contain stage IV metastatic tumors as
+it is to contain those of stage I (**figure 3D, E**). This cluster likely
+represents a group of tumors with aberrations in methylation, copy number
+alterations, structural variants, or combinations thereof which we excluded from
+our analysis. Consistent with this, kidney chromophobe and thyroid cancers have
+high proportions of structural variations _vs._ other variant types
+[@PCAWG-MAIN] and are heavily skewed to cluster 1 membership (**Supplementary
+figure S1**). Likewise, cluster 10 represents a group of hyper-mutated tumors
+that harbor so many mutations that virtually no pathway is unaffected. It seems
+likely that a significant fraction of the “mutant” samples for each pathway are
+burdened with excess passenger mutations. This could be addressed with
+application of ever-increasingly sophisticated filtering of likely passenger
+mutations [ _e.g._
 @spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
@@
 By modeling CDC longevity data as a baseline risk function we showed that each
-of our pathway disruption classes exhibit cancer-type specific effects on
+of our pathway disruption clusters exhibit cancer-type specific effects on
 expected survival (**figure 9**). Contrary to what one might expect at the
-outset, we found that models assuming different survival based on class and
+outset, we found that models assuming different survival based on cluster and
 cancer type fit better than models in which they are independent factors.
@@
 standards of care, and even within classes of drugs the preferred treatment can
-very between cancers, it makes sense that we observe tissue-specific class
+very between cancers, it makes sense that we observe tissue-specific cluster
 effects. Contrast the situation with ovarian _vs._ breast cancer, which are both
@@
 treatment regimens based on molecular subtype and other factors. It is easy to
-see how standard of care for one cancer and class could produce different
-outcomes than the same class of a different cancer type. Unfortunately, given
+see how standard of care for one cancer and cluster could produce different
+outcomes than the same cluster of a different cancer type. Unfortunately, given
 the diversity of drug classes and treatments (including non-chemical treatments
@@
 distinguish between treatment-specific effects on survival given different
-pathway disruption classes.
+pathway disruption clusters.
 
@@
 Our findings imply that there are separate processes in the etiology of cancer
-that can be broadly thought of as general cancer promoting, class-specific
+that can be broadly thought of as general cancer promoting, cluster-specific
 mutations and metastasis. General cancer promoting processes must include
@@
 characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
-disrupted in all or most of the defined classes (**supplementary figure S1**)
+disrupted in all or most of the defined clusters (**supplementary figure S1**)
 and are frequently the result of aberrations involving common driver genes such
 as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
-(**supplementary figure S3** and **supplemental file 1**). Class-specific
+(**supplementary figure S3** and **supplemental file 1**). Cluster-specific
 evolution must involve the acquisition of disruptions to pathways that may
@@
 to hypothesize that these mutations promote or enable the subsequent acquisition
-of class-specific defects via random mutation and natural selection, thus
-producing the classes we observed. In support of this, the pan-cancer analysis
+of cluster-specific defects via random mutation and natural selection, thus
+producing the clusters we observed. In support of this, the pan-cancer analysis
 of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
@@
 
-We report that pathway enrichment in metastatic tumors across all classes
+We report that pathway enrichment in metastatic tumors across all clusters
 (pooled) yielded generally lower effect sizes and larger p-values than the
@@
 are analyzed together, and supporting the view that, within the parameters of
-our analysis, metastasis has class-specific requirements. Since the number of
+our analysis, metastasis has cluster-specific requirements. Since the number of
 metastatic samples is relatively low this part of our analysis is likely
 underpowered and subject to expanded analysis with larger cohorts. Nonetheless,
-the class-specific metastatic pathway enrichment we observe (in **table 1**)
+the cluster-specific metastatic pathway enrichment we observe (in **table 1**)
 could be a result of the pathways directly promoting metastatic phenotypes in
-these classes, or because they acquire a stronger or more complete oncogenic
+these clusters, or because they acquire a stronger or more complete oncogenic
 phenotype, as we don’t know the order in which they were acquired. The fact that
@@
 clusters favors the latter interpretation, and suggests a mechanism of
-complementarity, in which newly acquired mutations result in similar classes
+complementarity, in which newly acquired mutations result in similar clusters
 converging on one or more deadly phenotypes that capture most of the critical
@@
 with energy metabolism as one of the new hallmarks [@hallmarks_II], but also
-distinguishing these classes from the others in the order or nature of hallmark
+distinguishing these clusters from the others in the order or nature of hallmark
 acquisition. Thus, even at relatively low power, our analysis of metastasis

### 2020-08-20 text/introduction.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

@@
-Recent advances in technology have made the study of cancer possible at
-unprecedented levels of detail and scale. Array and next-generation sequencing
+Recent advances in technology have greatly expanded our view into the mechanisms
+of cancer at the systems biology level. Array and next-generation sequencing
 technologies have made it possible to assay germline and somatic mutations,
-expression profiles, DNA methylation, and copy number variations. To take
-advantage of these tools several large consortia, including the cancer genome
-atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
-numbers of tumors and collected data from multiple assays that could be analyzed
-together, with the goal of integration and increased understanding of the
-mechanisms of cancer.
+expression profiles, DNA methylation, and copy number variations from the same
+tissue samples. To take advantage of these tools several large consortia,
+including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
+genomes (PCAWG), sequenced large numbers of tumors and collected data from
+multiple assays that could be analyzed together, with the goal of integration
+and increased understanding of the mechanisms of cancer.
 
@@
 when genes are mutated at a greater rate than expected due to chance. It has
-been estimated that on the order of three mutations in key oncogenes and/or
-tumor suppressors would be sufficient to transform a normal cell to a cancerous
-state [@vogelsteinPathCancerThree2015].
+been estimated that fewer than five mutations in key oncogenes and/or tumor
+suppressors would be sufficient to transform a normal cell to a cancerous state
+[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
+mutations in cancer driver genes are commonplace in somatic tissues and
+correlate with age and environmental exposures [@mutations-healthy-tissues].
 
-Other studies aim at providing a comprehensive view of mutations, gene
-expression and genomic signatures, with the express goal of understanding the
-common themes of all cancer independent of tissue of origin. Understanding
-cancer as a disease of the cell has long been a goal of the field and is perhaps
-best characterized in essays by @weinberg2000; @hallmarks_II (see also
-Discussion). The first of these studies in the genomic era identified 11 subtypes
-from 12 cancer types, using integrative analysis with co-equal weighting of gene
-expression, methylation, copy number and proteomics data
-[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
+Other studies provide a comprehensive view of mutations, gene expression and
+genomic signatures, with the express goal of understanding the common themes of
+all cancer independent of tissue of origin. Understanding cancer as a disease of
+the cell has long been a goal of the field as characterized in essays by
+@weinberg2000; @hallmarks_II. The first of these studies in the genomic era
+identified 11 subtypes from 12 cancer types, using integrative analysis with
+co-equal weighting of gene expression, methylation, copy number and proteomics
+data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that
 tissue-of-origin is the predominant driving factor, though a significant
@@
 including pan-gastrointestinal, pan-gynecological, pan-squamous,
-pan-gynecological/ squamous and pan-kidney
-[@hoadleyCellofOriginPatternsDominate2018].
+pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
 
 More recently, attempts have been made to consider pan-cancer mutations and
-other data as disruptions of normal pathway activity. @PCAWG-PATH
-carried out near-exhaustive analysis of pathway enrichment using whole genome
-data. Horn _et al._ [@hornNetSigNetworkbasedDiscovery2018] introduce approaches
-to address the sparseness of mutations, allowing the potential importance of a
-gene to be influenced by the aggregate of mutations in its nearest network
-neighbors. This method is powerful especially when applied to a small number (on
-the order of hundreds) of tumors. One of the challenges in the field remains
-that beyond a couple hundred common driver genes and mutations, there are likely
-thousands of moderate effect genes that occur at such low frequency--given
-tissue diversity--and low sample numbers that they are impossible to detect
-using positive selection theory. Some groups have attempted to address this
-challenge by applying machine learning to cancer data sets to discover groups of
-functionally related genes as they interact with larger pathways and networks
-[@kimPancancerAnalysisSomatic2018; @mourikis-ml-pathways; @moonlight]. These
-studies have proven very effective at highlighting some potentially fundamental
-disease phenotypes at the pathway level that cross cancers with different
-origins at the cellular and tissue level.
+other data as disruptions of normal pathway activity. @PCAWG-PATH carried out
+near-exhaustive analysis of pathway enrichment using whole genome data. Network
+approaches address the sparseness of mutations, allowing genes to be influenced
+by the aggregate of mutations in the nearest network neighbors
+[@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
+powerful especially when applied to a small number (on the order of hundreds) of
+tumors. One of the challenges in the field remains that beyond a couple hundred
+common driver genes and mutations, there may be thousands of moderate effect
+genes that occur at such low frequency--given tissue diversity--and low sample
+numbers that they are impossible to detect using positive selection theory. Some
+groups have attempted to address this challenge by applying machine learning to
+cancer data sets to discover groups of functionally related genes as they
+interact with larger pathways and networks [@kimPancancerAnalysisSomatic2018;
+@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
+highlighting some potentially fundamental disease phenotypes at the pathway
+level that cross cancers with different origins at the cellular and tissue
+level.
 
@@
 mutations to cell-biological pathway disruptions. We use this heuristic to
-evaluate all cancers and show that, surprisingly, tumors fall into categories of
-common pathway disruptions with distinctive origins arising from tissue-specific
-gene alterations and having unique prognosis in each cancer type.
+evaluate all cancers and show that, surprisingly, tumors that exhibit
+tissue-specific gene mutation patterns nonetheless fall into common categories
+of pathway disruption having unique prognoses in each cancer type.

### 2020-08-20 text/manuscript.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

--- 2020-08-16 text/manuscript.Rmd
@@
 data because the affected target gene is known unambiguously. Therefore, we
-selected all missense, nonsense, frameshift, stop-gain, untranslated region, and
+selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
 splicing mutations. In order to minimize bias from well-studied diseases and
-processes, we selected 377 [Reactome pathways](https://reactome.org/) of
-interest corresponding to basic cellular processes and biochemical pathways,
-excluding gene sets that correspond to miscellaneous categories ( _e.g._
-“transcription factors”) or disease associations ( _e.g._ “mutated in colon
-cancer”) and filtered our gene list on membership in these pathways (total of
-8940 genes; see methods for details).
+processes, we selected 377 [Reactome pathways](https://reactome.org/) (see
+supplemental table ST1) of interest corresponding to basic cellular processes
+and biochemical pathways, excluding gene sets that correspond to miscellaneous
+categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._
+“mutated in colon cancer”) and filtered our gene list on membership in these
+pathways (total of 8,940 genes; see methods for details).
 
@@
 (neomorphic or hypermorphic) in genes, and these can in turn act functionally as
-either activators or repressors. Within the scope of this study we accept this
-uncertainty and assume mutations are generically disruptive to normal pathway
-activity regardless of whether they are activating or deactivating or whether
-they affect a gene that we might classify as one thing or another ( _e.g._
-"Tumor suppressor" or "Oncogene"). Second, we know that low-expressed genes and
-non-expressed genes are known to accumulate mutations at a higher rate due to
-transcription coupled repair [@kandothMutationalLandscapeSignificance2013;
+either activators or repressors. For this study we assume a significant fraction
+of these mutations are generically disruptive to normal pathway activity if they
+are activating or deactivating or affect a gene that one might classify as a
+tumor suppressor or oncogene. Second, we know that low-expressed genes and
+non-expressed genes accumulate mutations at a higher rate due to transcription
+coupled repair [@kandothMutationalLandscapeSignificance2013;
 @kimPancancerAnalysisSomatic2018;
@@
 others (**figure 1**).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+#### Figure 1
+```{r child='../code/figure1v2.Rmd'}
+```
+
+**Figure 1: Molecular pathway profiles of tumor samples with one or more
+mutations.** Each of 377 selected Reactome pathways (columns) is classified as
+disrupted if one or more genes is mutated in the tumor sample (rows) where red
+and grey represents the pathway is mutated and not, respectively. Tumors are
+grouped by tissue of origin using standardized abbreviations from the TCGA
+project.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 
@@
 
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 2
+```{r child='../code/figure2v2.Rmd'}
+```
+
+**Figure 2: Method for tumor classification using pathway aberration data** 
+a) MCA analysis was performed to determine the eigenvalue with the most
+explanatory power from the categorical data. Average of 100 permutations of the
+data was used as a baseline. b) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
+projected down to 3 dimensions for the purpose of density based clustering (only
+the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
+used to perform hierarchical density based clustering tuned to minimize the
+fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
+the UMAP projection; tumors are colored by cluster membership (see legend). This
+results in some tumors remaining unclassified (grey points). f) kNN was used to
+assign membership to the remaining tumors.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
 #### Independence from tissue-of-origin.
@@
 **supplementary file 1 (html)**), suggesting that, in principle at least, our
-novel pathway-disruption classification method identifies molecular pathology
-largely independent of tissue-of-origin. As an example of one type of cancer
-that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD) was
-predominantly found in class 8, which is distinguished  by KRAS mutations among
-others (**supplementary figure S3**). But even PAAD comprises tumors from the
-nine remaining classes, meaning that patients suffering these tumors have
-potentially different underlying molecular pathologies.
+pathway-disruptions identify clusters of molecular pathology largely independent
+of tissue-of-origin. As an example of one type of cancer that does have a biased
+pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
+class 8 (**figure 3A** and **supplementary figure S1**). But even PAAD comprises
+tumors from the nine remaining classes, meaning that patients suffering these
+tumors have potentially different underlying molecular pathologies.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 3
+```{r child = '../code/figure4v2.Rmd'}
+```
+
+**Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
+a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
+COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
+Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the
+UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of
+tumor stage onto the UMAP, regardless of cancer type. e) Projection of
+metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
+metastatic tumors.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 
@@
 mean that our pathway disruption classes do not correspond to previously
-identified molecular subtypes within the parent cancer type. Furthermore, this
-result raises the possibility that, molecular and histo-pathological subtypes
-could represent different precursor cell origins within each cancer type. This
-interpretation still leaves open the possibility that different precursor cell
-types, being susceptible to different teratogens and selection pressures (
-_e.g._ insulin dependence) could be biased toward one or more of the pathway
-disruption profiles.
+identified molecular subtypes within the parent cancer type.
 
@@
 and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
-breast, prostate and ovarian cancers. The mechanism of risk is thought to
-involve loss of heterozygosity [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
+break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
+these genes confer elevated risk for breast, prostate and ovarian cancers. The
+mechanism of risk is thought to involve loss of heterozygosity resulting in loss
+of the wildtype, functional allele [@brca-loh], so we projected the somatic
+mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP, but did not observe
 segregation of these mutations into specific clusters (**figure 3C**). We made
@@
 backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
-the molecular-pathway disruption classes represent advancement through stages,
+the molecular-pathway disruption clusters represent advancement through stages,
 we projected staging data onto the UMAP. Similar to tissue of origin and other
@@
 
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 4
+```{r child = '../code/figure3v2.Rmd'}
+```
+
+**Figure 4: Somatic mutation frequencies for tumors in each class** 
+Column uses a logarithmic axis to represent mutation count. Row represents the
+tumor samples. Each dot corresponds to the sample mutation count for tumors in
+each class, ordered by mutation counts. The median mutation count in each
+cluster is indicated by the horizontal line.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
 Finally, as a measure of tumor advancement, metastasis is the condition in which
@@
 transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To
-determine whether any of our classes correspond to an especially advanced stage
+determine whether any of our clusters correspond to an especially advanced stage
 of cancer across tissue types, we projected the metastasis data onto our UMAP,
@@
 
-### Each class is defined by a specific set of pathway disruptions
+### Tissue specific genes define cluster membership
 
 In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created waterfall graphs analagous to the oncoprint
-visualizations on cBioportal [@cbioportal], replacing genes with pathways
+(pan-cancer), we created waterfall plots, replacing genes with pathways
 (**figure 5A**). A complete list of pathway disruptions with percent mutated
@@
 samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
-samples), _etc._ . The top third of pathways drop off gradually in overall
-mutation frequency (**figure 5A** and **supplementary table ST1**), possibly
-owing to the relative diversity of tumors.
+samples), _etc._
 
-To discover what pathways drive the UMAP clustering, we calculated enrichment
-for each pathway _in-class_ relative to all other classes combined (see methods)
-and ranked pathways for each class from highest to lowest enrichment. We
-visualized the enrichment as a heatmap (**figure 5B**). Using this approach, we
-identified 25 pathways highly enriched (enrichment score > 0.3, see methods) in
-class 2, 9 pathways enriched in class 3, 24 pathways in class 4, 10 pathways in
-class 5, 10 pathways in class 6, 24 pathways in class 7, 24 pathways in class 8
-(for which most pathways are putatively disrupted), 10 pathways in class 9, and
-0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
-terms of pathway enrichment. To explore the specific pathways marking each
-class, we projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). Classes 3 and 5 were distinguished by metabolic
-pathways including RNA and protein biosynthesis (**supplementary figure S4**).
-Similarly, class 4 was distinguished by mutations affecting regulation of DNA
-and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
-and "Nucleosome assembly"). Classes 7-9 have in common mutations in
-extracellular, intracellular, and immune-related signaling pathways (see
-**figure 5C** and **supplementary figure S4**). Cluster 2 had the highest pathway
-enrichment levels of the three, having mutations in hedgehog signaling,
-"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
-cycle" and "apoptosis" among others.
+```{r, results='asis'}
+cat("\\newpage")
+```
 
-### A combination of driver mutations and tissue-specific genes contribute to common class signatures
+#### Figure 5
+```{r child = '../code/figure2.Rmd'}
+```
+
+**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
+a) Waterfall plot; columns represent tumors. Each row represents a pathway. The
+pathways are ranked such that the pathway with the most disruptions across all
+7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
+top pathway disruptions successively. The top 100 most mutated pathways are
+shown. Red signifies pathway disrupted pathways, grey signifies not disrupted.
+b) Heatmap shows the relative enrichment of each pathway (columns) within each
+numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
+purple, depletion). c) Proportion of mutated samples in the each significant
+pathway (columns; union set of pathways with effect size >= 0.30 in each
+cluster) within each numbered cluster (rows).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+To discover what pathways drive the UMAP clustering, we calculated percent
+enrichment for each pathway _within cluster_ relative to all other clusters
+combined (see methods) and ranked pathways for each cluster from highest to
+lowest enrichment. We visualized the enrichment as a heatmap (**figure 5B**).
+Using this approach, we identified 25 pathways highly enriched (enrichment score
+$> 0.3$, see methods) in cluster 2, 9 pathways enriched in cluster 3, 24
+pathways in cluster 4, 10 pathways in cluster 5, 10 pathways in cluster 6, 24
+pathways in cluster 7, 24 pathways in cluster 8, 10 pathways in cluster 9, and 0
+pathways for clusters 1 and 10. Clusters 7, 8, and 9 had several pathways in
+common. To explore the specific pathways marking each cluster, we projected
+disruptions for each of the 377 pathways onto the UMAP (**supplementary figure
+S4**). Clusters 3 and 5 were distinguished by metabolic pathways including RNA
+and protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was
+distinguished by mutations affecting regulation of DNA and histone methylation
+("DNA methylation", "PRC2 methylates histones and DNA", and "Nucleosome
+assembly"). Clusters 7-9 have in common mutations in extracellular,
+intracellular, and immune-related signaling pathways (see **figure 5C** and
+**supplementary figure S4**). Cluster 2 had the highest pathway enrichment
+levels of the three, having mutations in hedgehog signaling, "β-catenin
+degradation", "cellular response to hypoxia", "regulation of cell cycle" and
+"apoptosis" among others.
 
@@
 difficulty in distinguishing tumor samples from their tissue-specific -omics
-data signatures. Given our pathway-disruption based classification, this raises
+data signatures. Given our pathway-disruption based clustering, this raises
 the question, are tumor phenotypes driven entirely by common driver genes, or by
@@
 _HIST1H2AC_). Even among the top most mutationally enriched genes within
-clusters there is no global pattern, indicating that our classes are not driven
+clusters there is no global pattern, indicating that our clusters are not driven
 by individual genes, but rather networks as a whole. Taken together, our data
@@
 
-### Enrichment of pathways in metastasis is class-specific
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 6
+```{r child = '../code/figure4.Rmd'}
+```
+
+**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
+Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
+95% confidence interval. Row represents the significant genes from each cluster.
+Each cluster was compared against the background (all other clusters) to find
+deferentially mutated genes. Significant genes (pvalue < 0.01) were selected and
+limited to the top ten results for each clusters. Heatmap; Column represents the
+cancer type. Row represents the significant genes as described above. The
+heatmap shows the proportion of samples per cluster, per cancer type that is
+mutated for each of the significant genes. Depleted significant genes (left).
+Enriched significant genes (right).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+### Enrichment of pathways in metastasis is cluster-specific
 
 Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 5B**) we applied this method to metastatic vs non-metastatic
+pathways (**figure 5B**) we compared metastatic _vs._ non-metastatic
 tumors, as we find that metastatic tumors are distributed across all ten
@@
 that were already shown to be enriched in a neighboring cluster. For example,
-“Signaling by PTK6” is enriched in non-metastatic samples of cluster 8 (see
+“Signaling by _PTK6_” is enriched in non-metastatic samples of cluster 8 (see
 **supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
 metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$, **table 1**). This is also
-true of “Erythropoietin activates RAS”, which is enriched in cluster 8
-(**supplementary figure S4**) and also in metastatic tumors of clusters 7 and 9.
-Cluster 4 metastatic samples were enriched for “Fc epsilon receptor (FCERI)
-signaling”, a key neutrophil pathway, which is also specific to clusters 2, 7
-and 8 non-metastatic tumors. Thus, pathway lesions specific to metastases tend
-toward complementary mutations that are otherwise specific to non-metastatic
-tumors of other clusters.
+true of “Erythropoietin activates _RAS_”, which is enriched in non-metastatic
+tumors of cluster 8 (**supplementary figure S4**) and also in metastatic tumors
+of clusters 7 and 9. Cluster 4 metastatic samples were enriched for “Fc epsilon
+receptor (FCERI) signaling”, a key neutrophil pathway, which is also specific to
+clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions specific to
+metastases tend toward complementary mutations that are otherwise specific to
+non-metastatic tumors of other clusters.
 
-### Pathway disruption classes vary in short-term prognosis of survival
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Table 1
+```{r child = '../code/table2.Rmd'}
+```
+
+**Table 1: Cluster-specific enriched pathways (effect size >= 0.30) in metastatsis.**
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+### Pathway disruption clusters vary in short-term prognosis of survival
 
@@
 ```
-#### Figure 1
-```{r child='../code/figure1v2.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 2
-```{r child='../code/figure2v2.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 3
-```{r child = '../code/figure4v2.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 4
-```{r child = '../code/figure3v2.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 5
-```{r child = '../code/figure2.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 6
-```{r child = '../code/figure4.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-```{r child='../code/survival/survfig.Rmd'}
-
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Table 1
-```{r child = '../code/table2.Rmd'}
-#
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
 

### 2020-08-21 code/figure3.Rmd
Change figure file name, update child doc reference in manuscript to reflect these changes

@@
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(sample_id, clust_knn) ## select the relevant columns
+```{r sankey-plot}
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
+load("../data/kelly.colours.rda")
+load("../data/ditto.colours.rda")
 
-## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$clust_knn)) %>% ## table the number of tumors in each cluster
-  'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
-  mutate(clust = as.character(clust)) ## convert factor to character
+sankey.data <- umap_cancertype %>% dplyr::select(sample_id, project_code, clust_knn) %>% ## select relevant columns
+  gather(., key = "type", value = "value", -sample_id) ## wide to long format
 
-## Organizing rBeta data
-load("../data/gdc_reactome_path_analysis.Rda")
-clust_group <- n_tumor_in_clust$clust
-cluster_path_count <- list()
-for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(clust_knn == n) %>% ## filter for tumor class
-    .[,"sample_id"] ## get sample_ids
-  path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
-    rowSums() %>% as.data.frame() %>% ## getting the observed mutation count of tumor class
-    'colnames<-' (n) ## change colname to cluster number
-  cluster_path_count[[n]] <- path_mutation
-  rm(path_mutation, sample_id, n)
-  gc()
-}
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
+## Cancer names in alphabetical order
+cancerorder <- unique(umap_cancertype$project_code)
+cancerorder <- cancerorder[order(cancerorder)]
+## Cluster in numerical order
+clustorder <- unique(umap_cancertype$clust_knn)
+clustorder <- as.character(clustorder[order(clustorder)])
 
-global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = max(tumor_clusters$clust_knn))] %>% ## replicate the column 0 times for the 10 tumor classes
-  'colnames<-' (clust_group) %>% ## rename column with tumor class
-  rownames_to_column(var = "path") %>% ## move rownames to column (pathways) %>%
-  gather(., key = "clust", value = "global_mut", -path)
+## Factor to force ordering
+sankey.data$type <- factor(sankey.data$type, levels = c( "clust_knn", "project_code"))
+sankey.data$value <- factor(sankey.data$value, levels= c(cancerorder, clustorder))
 
-rbeta_path_analysis <- cluster_path_count %>% rownames_to_column(var="path") %>%
-  gather(key = "clust", value = "clust_mut", -path) %>% ## Wide to long format
-  left_join(., global_path_mut, by = c("path", "clust")) %>% ## add global mutation count 
-  left_join(., n_tumor_in_clust, by = "clust") %>% ## add total number of tumor samples per clust 
-  mutate(global_sample = sum(n_tumor_in_clust$tot_n_clust)) ## add total number of tumor samples
+## Color scheme for sankey plot
+sankey.colours <- c(ditto_colours[2:24],kelly.colours[3:12])
+names(sankey.colours) <- c(cancerorder, clustorder)
 
-rm(global_path_mut, cluster_path_count, n_tumor_in_clust, clust_group, tumor_clusters, gdc_reactome_path_analysis)
-gc()
-
-enrich_prob_uncertainity_in_bg <- data.frame(path = rbeta_path_analysis$path,
-                                            clust = rbeta_path_analysis$clust,
-                                            fg_mut = rbeta_path_analysis$clust_mut,
-                                            fg_not_mut = rbeta_path_analysis$tot_n_clust - rbeta_path_analysis$clust_mut,
-                                            bg_mut = rbeta_path_analysis$global_mut,
-                                            bg_not_mut = rbeta_path_analysis$global_sample - rbeta_path_analysis$global_mut
-                                            )
-fg_mut_rate <- list()
-bg_mut_rate <- list()
-for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
-  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
-  rm(df)
-}
-
-fg_mut_rate <- fg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("fg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-bg_mut_rate <- bg_mut_rate %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind the mut rates from list
-  'colnames<-' ("bg_mut_rate") %>% ## rebind col names
-  rownames_to_column(var = "clust") ## move rownames to col
-
-enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>%
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
-  mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
-  mutate(x_adj =(.$fg_mut + .$fg_not_mut) * (.$bg_mut * .$rel_odds) / .$denominator) %>%
-  mutate(y_adj = (.$fg_mut + .$fg_not_mut) - .$x_adj)
-rm(fg_mut_rate, bg_mut_rate, rbeta_path_analysis)
-gc()
-
-moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
-})
-
-moderate_bg <- apply(enrich_prob_uncertainity_in_bg[,c("x_adj", "y_adj")], MARGIN = 1, function(x) {
-  rbeta(n = 1e5, shape1=x["x_adj"], shape2=x["y_adj"])
-})
-
-## ModerateEnrichment probability 
-prob <- moderate_fg > moderate_bg ## boolean data
-mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
-})
-rm(prob)
-gc()
-## Moderate Difference
-CI <- 0.025 ## setting a 95% confidence interval
-
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
-rm(moderate_fg, moderate_bg)
-gc()
-
-sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
-}))
-quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
-}))
-
-moderate_enrich_prob <- data.frame(path = enrich_prob_uncertainity_in_bg$path,
-                                   clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
-rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
-gc()
-save(moderate_enrich_prob, file = "../data/moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
-
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
-
-effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
-         cluster_cols = T,
-         cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
-         clustering_method = "ward.D2")
-
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
-
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
-
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:max(as.numeric(effect_heatmap$cluster))))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
-  effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
-  effect_size > effect_segments[2] & effect_size <= effect_segments[3] ~ "2",
-  effect_size > effect_segments[3] & effect_size <= effect_segments[4] ~ "3",
-  effect_size > effect_segments[4] & effect_size <= effect_segments[5] ~ "4",
-  effect_size > effect_segments[5] & effect_size <= effect_segments[6] ~ "5",
-  effect_size > effect_segments[6] & effect_size <= effect_segments[7] ~ "6",
-  effect_size > effect_segments[7] & effect_size <= effect_segments[8] ~ "7",
-  effect_size > effect_segments[8] & effect_size <= effect_segments[9] ~ "8",
-  effect_size > effect_segments[9] & effect_size <= effect_segments[10] ~ "9",
-  effect_size > effect_segments[10] & effect_size <= effect_segments[11] ~ "10",
-  effect_size > effect_segments[11] & effect_size <= effect_segments[12] ~ "11"
-))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
-
-save(effect_heatmap, file = "../data/effect_heatmap.Rda")
-rm(list = ls(all.names = T))
-gc()
+## Sankey plot
+ggplot(sankey.data,
+       aes(x = type, stratum = value, alluvium = sample_id,
+           fill = value, label = value)) +
+  scale_fill_manual(values = sankey.colours) +
+  geom_flow(stat = "alluvium", lode.guidance = "frontback") +
+  geom_stratum(alpha = 0.5) +
+  geom_text(stat = "stratum", size = 3) +
+  theme_classic() +
+  xlab(label = "") +
+  theme(legend.position = "none",
+        axis.text.x = element_text(size = 10),
+        axis.text.y = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line.x = element_blank(),
+        axis.line.y = element_blank()) +
+  scale_x_discrete(labels= c("Cluster", "Cancer Type"))
 ```
 
-```{r group-pathway-enrichment heatmap, fig.width= 5, fig.height= 10, fig.align= 'center', fig.cap="**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this)."}
-load("../data/effect_heatmap.Rda") ## load data
+```{r figure-4-plot, fig.width = 9}
+### Figure A - Cancer Type
+par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
+par(mar = c(2,2,0,0)) ## setting plot margins for the following plots
 
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
+selectedcancer <- c("BRCA", "COAD", "HNSC", "PAAD") ## selecting the 4 cancer types we're interested in
+A_colors <- c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4') ## Color scheme for cancer type
+names(A_colors) <- selectedcancer ## Assigning cancertype a color
 
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  ylab("377 Reactome Pathways") +
-  xlab("Cluster") +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "none")
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
+with(umap_cancertype %>% 
+       mutate(A_color = case_when(project_code %in% selectedcancer ~ project_code, ## Create a new col named A color
+                                  TRUE ~ "Other")) %>% ## Convert all cancer type that isn't the selected cancers to "Other"
+       filter(A_color != "Other"), ## Remove "Other" type
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = A_colors[A_color], pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=rev(names(A_colors[1:2])), col=rev(A_colors[1:2]), pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=rev(names(A_colors[3:4])), col=rev(A_colors[3:4]), pch = 20, bty = "n", horiz = T, cex = 1.2)
+title(xlab="UMAP 2", line=0)
+mtext("A", side=3, line=-2.05, cex=1.2, adj=0)
 
-fig3_legend <- plot_grid(NULL, fig3_legend, NULL, ncol = 3)
+### Figure B - BRCA subtype
+brca_subtype <- umap_cancertype %>% filter(project_code == "BRCA") %>% ## Filter for selected cancer type
+  dplyr::select(plot_x, plot_y, plot_z, Subtype_Selected) ## Select relevant column 
+brca_subtype[is.na(brca_subtype)] <- "NA" ## Convert NA to character "NA"
 
-plot_grid(fig3, fig3_legend, ncol = 1, rel_heights = c(15, 1))
+## Color scheme for subtype
+B_name <- unique(brca_subtype$Subtype_Selected) ## What are the subtype names
+B_name <- B_name[order(B_name,na.last=T)] ## Order the names ABC and have NA last
+B_color <- ditto_colours[c(2:(length(B_name) + 1))] ## Selecting enough colors from ditto
+names(B_color) <- B_name ## Assigning subtypes to color
+
+with(brca_subtype,
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = B_color, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=names(B_color[1:2]), col=B_color[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=names(B_color[3:4]), col=B_color[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.20), legend=names(B_color[5:6]), col=B_color[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
+title(xlab="UMAP 2", line=0)
+mtext("B", side=3, line=-2.05, cex=1.2, adj=0)
+
+### Figure C - BRCA Somatic Mutations
+load("../data/gdc_gene_exp_adj.Rda")
+load("../data/gene_data.Rda")
+brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
+brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## Include BRCA gene
+  rownames_to_column(var = "ensembl_gene_id") %>% ## Move ensembl id to column
+  left_join(., brca, by = "ensembl_gene_id") %>% ## left join with brca df 
+  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace row with gene symbol
+  dplyr::select(-ensembl_gene_id) %>% ## get rid of ensembl_id col
+  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
+  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
+  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1", ## Convert booleans to categorical variables
+                           BRCA1 == "0" ~ "NA"),
+         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
+                           BRCA2 == "0" ~ "NA")) %>%
+  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
+  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
+  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "plot_z", "project_code")], by = "sample_id") ## leftjoin by sample id
+
+C_colors <- c('#33a02c', '#e31a1c', '#ff7f00') ## Color scheme 
+names(C_colors) <- c("BRCA1", "BRCA2", "BRCA1,BRCA2") ## Assigning mutation type to color
+with(brca %>% filter(mut_type != "NA"), ## Filter samples that do not have a mutation in BRCA1 or BRCA2
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = C_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=names(C_colors[c(1:2)]), col=C_colors[c(1:2)], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=names(C_colors[c(3)]), col=C_colors[3], pch = 20, bty = "n", horiz = T, cex = 1.2)
+title(xlab="UMAP 2", line=0)
+mtext("C", side=3, line=-2.05, cex=1.2, adj=0)
+
+### Figure D - Stage
+D_colors <- c('darkgoldenrod3','dodgerblue3','yellow3','tomato2') ## Color scheme
+names(D_colors) <- c("Stage I", "Stage II", "Stage III", "Stage IV") ## Assign stage to color
+with(umap_cancertype %>% filter(ajcc_pathologic_tumor_stage != "Not Available"), ## Filter samples that do not have stage info
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = D_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=names(D_colors[1:2]), col=D_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
+legend("top", inset = c(0.10), legend=names(D_colors[3:4]), col=D_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
+title(xlab="UMAP 2", line=0)
+mtext("D", side=3, line=-2.05, cex=1.2, adj=0)
+
+### Figure E - Metastasis
+E_colors <- c('#a6cee3','#1f78b4') ## Color scheme
+names(E_colors) <- c("M0", "M1") ## Assign metastasis to color
+with(umap_cancertype %>% filter(ajcc_metastasis_pathologic_pm %in% c("M0", "M1")), ## Only M0 and M1 tumor samples
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = E_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=names(E_colors), col=E_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
+title(xlab="UMAP 2", line=0)
+mtext("E", side=3, line=-2.05, cex=1.2, adj=0)
+
+plot.new()
 ```

### 2020-08-21 text/manuscript.Rmd
Change figure file name, update child doc reference in manuscript to reflect these changes

--- 2020-08-16 text/manuscript.Rmd
@@
   library(plot3D),
-  library(ggfortify)
+  library(ggfortify),
+  library(maftools),
+  library(survival),
+  library(ggalluvial)
 ))
@@
 #### Figure 2
-```{r child='../code/figure2v2.Rmd'}
+```{r child='../code/figure2.Rmd'}
 #
@@
 #### Figure 3
-```{r child = '../code/figure4v2.Rmd'}
+```{r child = '../code/figure3.Rmd'}
 #
@@
 #### Figure 4
-```{r child = '../code/figure3v2.Rmd'}
+```{r child = '../code/figure4.Rmd'}
 #
@@
 #### Figure 5
-```{r child = '../code/figure2.Rmd'}
+```{r child = '../code/figure5.Rmd'}
 #
@@
 #### Figure 6
-```{r child = '../code/figure4.Rmd'}
+```{r child = '../code/figure6.Rmd'}
 #
@@
 #### Table 1
-```{r child = '../code/table2.Rmd'}
+```{r child = '../code/table1.Rmd'}
 #

### 2020-08-21 code/figure9.Rmd
add figure9.Rmd

--- 2020-08-15 code/survival/survfig.Rmd
@@
-#### Figure 7
-```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=9, fig.width=9}
-
-library(ggridges)
-library(patchwork)
-
-source("./kmcurve.R")
-
-## create example data for a fixed 30 year old without vs with a k=2
-## cancer
-
-patientsurvivalsbkd = replicate(100,simdeath(r20wf,ageratewf,30,1))/365
-patientsurvivalsk2 = replicate(100,simdeath(r20wf,ageratewf,30,2))/365
-
-mixedages = runif(100,30,70)
-mixedsurvwfk1 = sapply(mixedages, function(x) simdeath(r20wf,ageratewf,x,1)/365)
-mixedsurvwfk2 = sapply(mixedages, function(x) simdeath(r20wf,ageratewf,x,2)/365)
-
-p1 = data.frame(
-    r = patientsurvivalsbkd,
-    c = patientsurvivalsk2,
-    rm = mixedsurvwfk1,
-    cm = mixedsurvwfk2,
-    status = rep(1, length(patientsurvivalsbkd))
-) %>%
-    rownames_to_column(var = "index") %>%
-    pivot_longer(-c(index, status), values_to = "time", names_to = "condition") %>%
-    ggplot(aes(time = time, color = condition, status = status)) +
-    geom_km() +
-    scale_color_manual(
-        values = c(r = "#56b4e9", c = "#e69f00", rm = "#0072b2", cm = "#d55e00"),
-        breaks = c("r", "c", "rm", "cm"),
-        labels = c("Random 30 year old women", "Random 30 year old women cancer k=2", "Random uniform(30,70) year old women", "Random uniform(30,70) year old women k=2")
-    ) +
-    theme_classic() +
-    theme(legend.position = c(.8, .75))
-
-pstad = plotcancertype(cancerdat,"STAD")
-pgbm = plotcancertype(cancerdat,"GBM")
-ppaad = plotcancertype(cancerdat,"PAAD")
-thca = plotcancertype(cancerdat,"THCA")
-
-load("./survmodel-ethbkd.stansave")
-
-dfsamps = as.data.frame(sampseth)
-
-kcancplot = function(dfsamps,tisnames){
-	ktisnames = grep("ktis",names(dfsamps),value=TRUE)
-	dfktis = dfsamps[,ktisnames]
-	dfktis = stack(dfktis) ## convert to two column "values" and "ind" with "ind" being the name of the variable
-	names(dfktis) = c("value","parname")
-
-	dim(tisnames) = c(length(tisnames),1)
-
-	dfktis$cancname = as.factor(sapply(dfktis$parname,function(x) tisnames[x]))
-	ggplot(dfktis,aes_(quote(value),quote(cancname))) + geom_density_ridges()+ theme_minimal() + scale_fill_brewer(palette="Paired") + xlim(0,22) + labs(x="ktis multiplier",y="Cancer Type",title="Tissue Specific Effects")
-}
-
-kclassplot = function(dfsamps,tisnames){
-	kclnames = grep("k\\[",names(dfsamps), value=TRUE)
-	dfclas = dfsamps[,kclnames]
-	dfclas = stack(dfclas)
-	names(dfclas) = c("value","coefname")
-	dfclas$class = as.factor(as.numeric(gsub("k\\[(.*),(.*)\\]","\\2",dfclas$coefname)))
-	dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
-	ggplot(dfclas,aes_(quote(value),quote(class))) + geom_density_ridges() +
-		facet_wrap(. ~ tiss,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
-		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients")
-}
-
-pcanc = kcancplot(dfsamps,levels(cancerdat$type))
-pclass = kclassplot(dfsamps,levels(cancerdat$type))
-
-mytheme = theme(
-#    axis.text.x = element_blank(),
-#    axis.text.y = element_blank(),
-    axis.title.x = element_text(size = 10),
-    axis.title.y = element_text(size = 10),
-#    axis.ticks = element_blank(),
-#    axis.line = element_blank(),
-    strip.text.y.right = element_text(angle = 0),
-    strip.background = element_blank(),
-    legend.key.height = unit(0.4, "lines"),
-    legend.title = element_blank(),
-    #legend.position = "none",
-    title = element_text(size=10)
-  )
-
-kmlayout = "
-AB
-AB
-AB
-CD
-CD
-CD
-EE
-"
-kmlegend <- theme(legend.direction = "horizontal")
-
-print(
-  ((p1+mytheme)/((pstad[[1]]+mytheme + kmlegend)+(pgbm[[1]]+mytheme + kmlegend) +
-                   (ppaad[[1]]+mytheme + kmlegend) + (thca[[1]] + mytheme + kmlegend) + guide_area() + plot_layout(design = kmlayout, guides = "collect"))) + plot_annotation(tag_levels = "A")
-)
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 8
-```{r, message = FALSE, fig.width = 3}
-pcanc+mytheme
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 9
 ```{r, message = FALSE, fig.width = 9, fig.height= 9}

### 2020-08-21 code/survival/survfig.Rmd

--- 2020-08-14 code/survival/survfig.Rmd
@@
-#### Figure 7
 ```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=9, fig.width=9}
@@
 ```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 8
-```{r, message = FALSE, fig.width = 9}
-mylayout = '
-ABB
-ABB
-#BB
-'
-
-((pcanc+mytheme) + (pclass+mytheme))+plot_layout(design=mylayout) +plot_annotation(tag_levels = "A")
-```

### 2020-08-21 code/table1.Rmd

@@
 tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
-load("../data/moderate_enrich_prob.Rda") ## load data
-load("../data/gene2_level2_lookup.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/pathname2level2.Rda")
+load("../data/gdc_reactome_path_analysis.Rda")
 
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
+tumor_clusters <- tumor_clusters %>% dplyr::select(sample_id, ajcc_metastasis_pathologic_pm, clust_knn)
 
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
+m0_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M0",] ## filter for nonmetastatic samples
+m1_clust <- tumor_clusters[tumor_clusters$ajcc_metastasis_pathologic_pm == "M1",] ## filter for metastatic samples
 
-## parameters ##
-enrich_cutoff <- 0.30
-howmanypaths <- 5 ## top _ pathways; how many top pathways do you want to see
-show_perc <- TRUE
-show_n <- TRUE
-mutated_gene_position <- 1
-################
-
-clust_spec_enrich <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) ## signif is 95% CI with a specified cutoff
-
-clust_spec_enrich_top5 <- list()
-for(n in unique(clust_spec_enrich$clust)) {
-  message(n)
-  clust_spec_enrich_top5[[n]] <- clust_spec_enrich %>% filter(clust == n) %>% ## filter for specific cluster
-    arrange(desc(difference)) %>% ## arrange enrichment score in decreasing order 
-    dplyr::select(path, clust) %>% .[1:5,]
+prepare_rbeta_data_p1 <- function(data, path_analysis) {
+  n_tumor_in_group <- as.data.frame(table(data$ajcc_metastasis_pathologic_pm)) ## Getting the total number of tumor samples in each group
+  path_analysis <- path_analysis[,names(path_analysis) %in% data$sample_id] ## Filter only group specific cancers
+  data <- data[match(names(path_analysis), data$sample_id),] ## Match tumor sample ids
+  table(names(path_analysis) == data$sample_id) ## good they match now
+  ## Now we are going to build group specific cluster data
+  n_tumor_in_clust <- as.data.frame(table(data$clust_knn)) ## getting total n in each cluster
+  names(n_tumor_in_clust) <- c("clust", "tot_n_clust") ## rename colnames
+  clust_group <- as.character(n_tumor_in_clust$clust) ## Getting the cluster numbers
+  cluster_path_count <- list() ## Creating a list for the forloop
+  for(n in clust_group) {
+    print(n)
+    df <- subset(data, clust_knn == n) ## subset cluster number
+    df <- df$sample_id ## getting tumor sample id
+    df1 <- path_analysis[,names(path_analysis) %in% df] ## getting the tumor samples from path analysis
+    df1 <- as.data.frame(rowSums(df1))
+    names(df1) <- n
+    cluster_path_count[[n]] <- df1
   }
-clust_spec_enrich_top5 <- clust_spec_enrich_top5 %>% do.call(rbind, .) %>% as.data.frame() ## combine list from for loop
-rm(clust_spec_enrich)
-
-## finding the top mutated gene for each cluster in each pathway for the top 5 pathways 
-type <- unique(tumor_clusters$project_code) ## type of cancer type in data
-cluster_spec_enrich_table <- list() ## list for forloop
-
-for(d in unique(clust_spec_enrich_top5$clust)) {
-  message(d)
-  clust_toppath <- clust_spec_enrich_top5 %>% filter(clust == d) %>% .$path %>% as.character() ## get the top pathway names for spec class
-  class_spec <- tumor_clusters %>% dplyr::select(sample_id, clust_knn, project_code) %>% filter(clust_knn %in% d) ## get the smaple ids and cancer type for spec class
-  top_genes_in_path_cancer_type <- list()
-  for(n in clust_toppath) { ## for each of the top paths do this
-    message(n)
-    gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-    top_gene_in_path <- list()
-    for(t in type) { ## looping through cancer type within class
-      message(t)
-      class_spec <- tumor_clusters %>% dplyr::select(sample_id, clust_knn, project_code) %>% filter(clust_knn == d) ## sample_id for class
-      if(t %in% class_spec$project_code) { ## if cancer type is in class then run this 
-        class_type_spec <- class_spec %>% filter(project_code == t) %>% column_to_rownames(var = "sample_id") %>% rownames() ## get cancer type specific tumor classes
-      if(length(class_type_spec) == 1) { ## if you only have 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% rownames_to_column(var = "gene") %>% .[.$gene %in% gene_in_path, names(.) %in% c("gene",  class_type_spec)] %>% remove_rownames() %>% column_to_rownames(var = "gene") %>% 'colnames<-' ("obs_count") %>%
-          rownames_to_column(var = "gene") %>%
-          mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-         arrange(desc(obs_count))
-      }
-      if(length(class_type_spec) > 1) { ## if you have more than 1 sample_id in the class
-        top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, names(.) %in% class_type_spec] %>% 
-        rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>%
-        rownames_to_column(var = "gene") %>%
-        mutate(n_sample = length(class_type_spec),
-               perc_mut = round((obs_count / length(class_type_spec)) * 100, digits = 1)) %>% 
-         arrange(desc(obs_count))
-      }
-      if(show_perc == TRUE & show_n == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)")  
-      }
-      if(show_perc == TRUE & show_n == TRUE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position], " (", top_mut_gene$perc_mut[mutated_gene_position],"%)", "    n = ", length(class_type_spec))  
-      }
-      if(show_perc == FALSE) {
-         top_gene_in_path[[t]] <- paste0(top_mut_gene$gene[mutated_gene_position])  
-      }
-      }
-      if(!(t %in% class_spec$project_code)){ ## if cancer type is not in class then insert NA
-        top_gene_in_path[[t]] <- "NA" 
-      }
-    }
-    top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% 'colnames<-' (n)
-    clust_row <- data.frame(Class = d, row.names = n) %>% t() ## creating a row for cluster membership
-    top_genes_in_path_cancer_type[[n]] <- rbind(clust_row, top_gene_in_path)
-  }
-  cluster_spec_enrich_table[[n]] <- top_genes_in_path_cancer_type %>% do.call(cbind, .) %>% as.data.frame() %>% 
-  t() %>% as.data.frame() %>% rownames_to_column(var = "Pathway")
+  cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
+  rbeta_path_df <- cluster_path_count %>% rownames_to_column(var="path") 
+  rbeta_path_df <- rbeta_path_df %>% gather(key = "clust", value = "mutation", -path) ## Wide to long format
+  global_path_mut <- as.data.frame(rowSums(cluster_path_count))
+  global_path_mut <- global_path_mut[rep(1:ncol(global_path_mut), each = max(as.numeric(as.character(rbeta_path_df$clust))))] ## Repeat column 12 times
+  names(global_path_mut) <- as.character(clust_group) ## Replace names with cluster numbers
+  global_path_mut <- global_path_mut %>% rownames_to_column(var="path") %>% gather( key = "clust", value = "global", -path) ## Wide to long format
+  rbeta_path_df <- left_join(rbeta_path_df, global_path_mut, by = c("path", "clust")) ## left_join rbetadf and global
+  rbeta_path_df <- left_join(rbeta_path_df, n_tumor_in_clust, by = "clust")
+  rbeta_path_df$global_sample <- as.numeric(sum(n_tumor_in_clust$tot_n_clust)) ## Getting total tumor sample number
+  names(rbeta_path_df) <- c("path", "clust", "clust_mut", "global_mut", "tot_n_clust", "global_sample")
+  rbeta_path_df$tot_n_clust <- as.numeric(as.character(rbeta_path_df$tot_n_clust)) ## convert tot_n_clust numeric
+  return(rbeta_path_df)
 }
 
-cluster_spec_enrich_table <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
-save(cluster_spec_enrich_table, file = "../data/cluster_spec_enrich_table.Rda")
+m0_rbeta <- prepare_rbeta_data_p1(data = m0_clust, path_analysis = gdc_reactome_path_analysis)
+save(m0_rbeta, file = "../data/m0_rbeta.Rda")
+
+## repeat same function above with metastasis sample
+m1_rbeta <- prepare_rbeta_data_p1(data = m1_clust, path_analysis = gdc_reactome_path_analysis)
+save(m1_rbeta, file = "../data/m1_rbeta.Rda")
+
+## function to compiling data for rbeta analysis
+prepare_rbeta_data_p2 <- function(data) {
+  enrich_prob <- data.frame(path = data$path,
+                               clust = data$clust,
+                               fg_mut = data$clust_mut,
+                               fg_not_mut = data$tot_n_clust - data$clust_mut,
+                               bg_mut = data$global_mut,
+                               bg_not_mut = data$global_sample - data$global_mut)
+fg_mut_rate <- list()
+bg_mut_rate <- list()
+for(n in unique(as.character(enrich_prob$clust))) {
+  print(n)
+  df <- subset(enrich_prob, clust == n)
+  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
+  fg_numerator <- sum(df$fg_mut)
+  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
+  bg_numerator <- sum(df$bg_mut)
+  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
+  bg_mut_rate[[n]] <- bg_numerator / bg_denominator
+}
+
+fg_mut_rate <- as.data.frame(do.call(rbind, fg_mut_rate)) ## rbind the mut rates from list
+bg_mut_rate <- as.data.frame(do.call(rbind, bg_mut_rate)) ## rbind the mut rates from list
+enrich_prob_df <- enrich_prob
+enrich_prob_df$fg_mut_rate <- rep(x = fg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
+enrich_prob_df$bg_mut_rate <- rep(x = bg_mut_rate$V1, each = length(unique(enrich_prob_df$path)))
+enrich_prob_df$rel_odds <- enrich_prob_df$fg_mut_rate / enrich_prob_df$bg_mut_rate
+enrich_prob_df$denominator <- (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) +   (enrich_prob_df$bg_not_mut / enrich_prob_df$rel_odds)
+enrich_prob_df$x_adj <- round((enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) * (enrich_prob_df$bg_mut * enrich_prob_df$rel_odds) / enrich_prob_df$denominator)
+enrich_prob_df$y_adj <- (enrich_prob_df$fg_mut + enrich_prob_df$fg_not_mut) -   enrich_prob_df$x_adj
+return(enrich_prob_df)
+}
+
+m0_rbeta <- prepare_rbeta_data_p2(m0_rbeta)                               
+m1_rbeta <- prepare_rbeta_data_p2(m1_rbeta)
+
+save(m0_rbeta, m1_rbeta, file = "../data/m0_m1_rbeta.Rda")
+rm(list = ls(all.names = T)) ## cleaning up data
+gc()
+load("../data/m0_m1_rbeta.Rda")
+## Metastatic (foreground) - Nonmetastatic (background) Rbeta
+moderate_fg <- apply(m1_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
+})
+moderate_bg <- apply(m0_rbeta[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
+  rbeta(n = 1e5, shape1=x["fg_mut"], shape2=x["fg_not_mut"])
+})
+
+metastasis_cluster_moderate <- moderate_fg - moderate_bg  ## Fg - Bg ... to see how many times the Fg is larger than the Bg
+metastasis_cluster_moderate <- as.data.frame(metastasis_cluster_moderate)
+
+## ModerateEnrichment probability 
+prob <- moderate_fg > moderate_bg ## boolean data
+mean_prob <- apply(prob, MARGIN = 2, function(x) {
+  mean(x)
+})
+## Moderate Difference
+CI <- 0.025
+
+diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
+rm(moderate_fg, moderate_bg, prob)
+gc()
+quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  quantile(x, c(CI, 0.5, 1-CI)) ## 95% CI
+}))
+
+sig_fg_bg <- as.data.frame(apply(diff_fg_bg, MARGIN = 2, function(x) {
+  sum(x < 0 ) ## out of 1e5 simulations, how many times is fg-bg less than 0
+}))
+
+m1_minus_m0_per_cluster_rbeta <- data.frame(path = m1_rbeta$path,
+                                   clust = m1_rbeta$clust,
+                                   probability = mean_prob,
+                                   signif = sig_fg_bg[,1],
+                                   difference = quant_diff[,2],
+                                   lower = quant_diff[,1],
+                                   upper = quant_diff[,3])
+save(m1_minus_m0_per_cluster_rbeta, file = "../data/m1_minus_m0_per_cluster_rbeta.Rda")
 rm(list = ls(all.names = T))
@@
 
-## Selecting the top 10 cancers with the largest n for table 1
-n_tumor <- as.data.frame(table(tumor_clusters$project_code))
-n_tumor <- n_tumor %>% arrange(desc(Freq)) %>% .[1:10, "Var1"] %>% as.character() %>% .[order(.)]
+```{r table-2, eval = FALSE}
+## this is for html only
+load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
 
-cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(row.names = FALSE) %>%
+table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.20 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
+  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
+
+table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
+table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
+table2$Signif <- NULL ## we don't need this anymore
+
+table2 %>% knitr::kable() %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-  column_spec(1, bold = T)
+  column_spec(1, bold = T, color = "black")
 ```
+
+```{r table2-docx-friendly}
+## docx friendly
+load("../data/m1_minus_m0_per_cluster_rbeta.Rda")
+
+table2 <- m1_minus_m0_per_cluster_rbeta %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
+  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
+
+table2$log10pvalue <- round(-log10((table2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
+table2$Enrichment_Score <- round(table2$Enrichment_Score, digits = 2)
+table2$Signif <- NULL ## we don't need this anymore
+
+table2 <-table2 %>% flextable() %>% 
+   set_caption(caption = "") %>%
+   align(part = "all") %>% 
+   width(., width = c(4,0.75,1,1)) %>% 
+   fontsize(size = 10, part = "body")
+
+knitr::knit_print(table2)
+```

### 2020-08-21 text/abstract.Rmd

--- 2020-08-14 text/abstract.Rmd
@@
-Cancer is a disease of cellular metabolism and other basic biochemical processes
-common to all cells. However, molecular signatures remain largely
-indistinguishable from tissue-of-origin, and general, unifying principles remain
-obscure. To address this challenge, we reduced mutation data to disruptions in a
-select set of pathways relevant to basic cell biology, from DNA replication to
-cellular communication. Using dimension reduction techniques, we assigned tumor
-samples into ten classes distinct from their tissues-of-origin and largely
-free of bias from mutational burden or clinical stage. We showed that these
-tumor classes vary in prognosis by modeling relative risk of death in different
-classes and cancer types. We identified class-specific mutations in different
+Cancer is a complex disease involving dysfunctional cellular metabolism, other
+basic biochemical processes, and the microenvironment. However, in spite of some
+generally agreed upon unifying principles [@weinberg2000; @hallmarks_II],
+molecular signatures remain largely indistinguishable from tissue-of-origin,
+thus obscuring opportunities for novel mechanistic insights and potential
+treatment vulnerabilities that hold true across different cancer types. To
+address this challenge, we reduced mutation data to disruptions in a select set
+of pathways relevant to basic cell biology, from DNA replication to cellular
+communication. Using dimensionality reduction techniques, we assigned tumor
+samples into ten clusters distinct from their tissues-of-origin and largely free
+of bias from mutational burden or clinical stage. We showed that these tumor
+classes vary in prognosis by modeling relative risk of death in different
+classes and cancer types. We identify cluster-specific mutations in different
 tissues, demonstrating that tissue-specific signatures contribute to common
-cellular phenotypes. Moreover, driver gene mutations involved in replication fidelity and genome
-stability, such as BRCA1 and Rb1, are equally distributed among our classes,
-contrary to the expectation that such genes are avatars of molecular
-subtype. We also investigated pathways that distinguish between metastatic and
-non-metastatic tumors, and found that most meaningful differences are
-class-specific. Many metastastasis enriched pathways from one class of
-tumors are class-specific pathways from another class, suggesting phenotypic
-convergence, and potentially highlighting key pathways to focus on
-for prevention of class-specific relapse. In summary, our
-observations suggest that some genes act as drivers of genome instability in all
-classes, while different sets of class-specific genes confer increased fitness
-in a combinatorial fashion, eventually converging on one or more end-stage
-disease phenotypes. Thus, we present a coherent view of global tumor biology,
-and explain how common cellular dysfunction might arise from tissue-specific
-mutations.
+cellular phenotypes. Moreover, driver gene mutations involved in replication
+fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally distributed
+among our clusters, contrary to the expectation that such genes are avatars of
+molecular subtype. We also investigated pathways that distinguish between
+metastatic and non-metastatic tumors, and found that most meaningful differences
+are cluster-specific. Some metastasis enriched pathways from one cluster are
+cluster-specific pathways from non-metastatic tumors of another cluster,
+suggesting phenotypic convergence, and potentially highlighting key pathways to
+focus on for prevention of cluster-specific relapse. Taken as a whole, our
+observations suggest that common driver genes combine with tissue-specific
+disruptions in complementary sets of tumor-promoting pathways to produce a
+limited number of distinct molecular phenotypes. Eventually these signatures
+converge on one or more end-stage disease phenotypes. Thus, we present a
+coherent view of global tumor biology, and explain how common cellular
+dysfunction might arise from tissue-specific mutations.

### 2020-08-21 text/manuscript.Rmd

--- 2020-08-20 text/manuscript.Rmd
@@
 #### Figure 2
-```{r child='../code/figure2v2.Rmd'}
+```{r child='../code/figure2.Rmd'}
 ```
@@
 #### Figure 3
-```{r child = '../code/figure4v2.Rmd'}
+```{r child = '../code/figure3.Rmd'}
 ```
@@
 #### Figure 4
-```{r child = '../code/figure3v2.Rmd'}
+```{r child = '../code/figure4.Rmd'}
 ```
@@
 #### Figure 5
-```{r child = '../code/figure2.Rmd'}
+```{r child = '../code/figure5.Rmd'}
 ```
@@
 #### Figure 6
-```{r child = '../code/figure4.Rmd'}
+```{r child = '../code/figure6.Rmd'}
 ```
@@
 #### Table 1
-```{r child = '../code/table2.Rmd'}
+```{r child = '../code/table1.Rmd'}
 ```

### 2020-08-21 text/survival.Rmd

@@
+---
+output:
+  word_document: default
+  html_document: default
+---
 If our molecular pathway disruption classes represent distinct biological states
@@
 
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 7
+```{r child='../code/survival/survfig.Rmd'}
+```
+
+**Figure 7: Comparison of different cancers, simulated and actual results
+results.** A) To understand the risk model, we plot survival of four simulated
+cohorts. Compare survival of a group consisting of randomly selected 30 year old
+women (light blue) to an equal sized group of 30 year old women with cancer, k =
+2 (light orange). To understand how the mix of ages alters the survival curve,
+we compare a randomly selected group of people with ages uniformly distributed
+from 30 to 70 (dark blue) and a randomly selected group with uniformly
+distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
+selected group with correct age distribution for the given cancer (blue) to the
+model predicted survival for the given average k (red) and the actual survival
+for the patients in our dataset (CHOOSE COLOR HERE, NOT GREEN). As can be seen,
+Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all
+very deadly. The modeled results fit the general trend for the actual patients.
+Thyroid Cancer (THCA) on the other hand has very little if any effect on life
+expectancy, in fact there may be a slight benefit which may be associated with a
+bias in diagnosis towards patients who are more health conscious, or higher
+socioeconomic status, with more access to care compared to random population
+members.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
 We found that cancer types, as expected, have a range of prognoses relative to
@@
 
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 8
+```{r fig.width=3, message=FALSE, warning=FALSE}
+pcanc+mytheme
+```
+
+**Figure 8 Tissue specific effects** Effects are compared between cancer type by
+showing the probability density for the posterior value of the $k_{tis}$
+multiplier. These values reflect the general deadliness of the cancer type
+across all clusters. GBM, PAAD, and SKCM stand out as having values in the 10 to
+20 range with a wide degree of uncertainty. Other cancers such as Prostate
+(PRAD) and Thyroid (THCA) show values down near or even below 1 suggesting that
+these populations may have healthier lifestyles and more access to care. Most
+cancer types fall in the range 2-7 as might be expected.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure 9
+
+```{r child='../code/figure9.Rmd'}
+```
+
+**Figure 9 Class specific $k_{class}$ values shown for each cancer type.** The
+overall $k$ value is the product of $k_{tis} \times k_{class}$. The class
+specific value shows the relative aggressiveness of that particular class within
+the tissue. For example class 1, 2, and 6 Glioblastoma (GBM) are apparently much
+less deadly than other types of GBM. A similar situation can be seen with
+Pancreatic Cancer (PAAD) where 1, 6, and 8 are less deadly compared to classes
+7, 9, 10. For some cancer types there are few data points and little ability to
+estimate values with precision (DLBC and KICH for example).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
 Our estimates of tissue-specific class effects, in contrast, ranged from less

### 2020-08-21 code/figure1v2.Rmd

@@
-```{r figure-1-data-generation, eval = FALSE}
-###################
-## Figure 1 data ##
-###################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene2_level2_lookup.Rda")
+```{r, fig.width = 9, fig.height= 4}
+load("../data/cancer_disrupt_heatmap.Rda") ## Load  in data processed in figure1.Rmd
 
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
+fig1 <- ggplot(data = cancer_disrupt_heatmap, aes(x=sample_id, y=path_name, fill = membership)) + 
+  geom_tile(size = 0.6) +
+  scale_fill_manual(values = c("grey", "firebrick3"),
+                    labels = c("Not Mutated", "Mutated")) +
+  facet_grid(.~type, scales = "free", space = "free", switch = "x") + ## facet by cancer type; switch: move facet labels top to bottom
+  xlab("") + ylab("377 Reactome Pathways") + labs(fill = "") +
+  guides(fill = guide_legend(override.aes = list(size = 2))) + ## make fill legend shape smaller
+  theme(axis.text.x = element_blank(), ## turn off x axis text
+        axis.text.y = element_blank(), ## turn off y axis text
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(), ## remove tick marks
+        strip.text.x = element_text(angle = 90), ## rotate x-axis
+        strip.background = element_blank(),
+        legend.position = "bottom", ## legend parameters
+        legend.key.height = unit(1, "lines"),
+        legend.text = element_text(size = 8),
+        legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(-10,-10,0,-10))
 
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
+## fixing the text cutoff in heatmap
+ fig1 <- ggplotGrob(fig1)
 
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_sample_id, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
+ for(i in which(grepl("strip-b", fig1$layout$name))){ ## grep all text strips at the bottom of plot 
+   fig1$grobs[[i]]$layout$clip <- "off" ## turn off clipping of text
  }
 
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
+fig1 <- as.ggplot(fig1) ## turn back fig into 
+fig1
 ```
-
-```{r, fig.width = 9, fig.height= 4}
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(x=sample_id, y=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(.~type, scales = "free", space = "free", switch = "x") +
-scale_y_discrete(position = "left") +
-  xlab("") +
-  ylab("377 Reactome Pathways") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.text.x = element_text(angle = 90),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-b", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-opar <- par(mfrow = c(3,3), mai = c(0.1, 0.1, 0.1, 0.1))
-
-fig1A
-```

### 2020-08-21 code/figure2.Rmd

--- 2020-08-13 code/figure2.Rmd
@@
-```{r oncoprint-top100-plus-panel-data-generation, eval = FALSE}
-tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
-load("../data/gdc_reactome_path_analysis.Rda")
+```{r figure-2-plot, fig.width = 9}
+load("../data/figure1B_data.rda") ## Load in data processed in figure1.Rmd
+load("../data/optimal_minpts_score.rda")
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
+load("../data/kelly.colours.rda")
+load("../data/ditto.colours.rda")
 
-######################
-## oncoprint top100 ##
-######################
-onco_color <- c("0" = "grey", "1" = "firebrick3")
+opar <- par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
+par(mar = c(4,6,2,0) + 0.1) ## setting plot margins for the following plots
+## Figure 2A: Explained variance of eigenvalues
+with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = expl_var / 100, ## plotting the explained variance of observed
+                                                 type = "l", yaxt = "n", bty = "l",
+                                                 col = "#998ec3", 
+                                                 xlab = "", ylab = ""))
+with(randomize %>% filter(eigenvalue < 50), lines(x = eigenvalue, y = perm / 100, ## plotting the explained variance of permuted
+                                                  col = "#f1a340"))
+yticks_val <- scales::breaks_pretty(n=5)(randomize %>% filter(eigenvalue < 50) %>% pull(expl_var) / 100)
+axis(2, at=yticks_val, lab=scales::percent(yticks_val), las = 1) ## Axis labels
+title(ylab = "explained variance", line = 3.5)
+title(xlab = "MCA Eigenvalues", line = 2)
+legend("topright", legend=c("Eigenvalues", "Permuted Eigenvalues"),
+       col=c("#998ec3", "#f1a340"), lty=1, cex=0.8)
+mtext("A", side=3, line=0, cex=1.2, adj=-0.35) ## plot tag
 
-top100path <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:100,] %>% .[, "path_name"] ## getting the top100 mutated pathways in class
+par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
+## Figure 2C: MCA projected down to 3 dimensions (UMAP)
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = "white", pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+mtext("C", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
 
-top100path <- gdc_reactome_path_analysis[top100path, ] ## selecting the top 100 paths
+par(mar = c(4,6,2,0) + 0.1) ## setting plot margins for the following plots
+## Figure 2D: Optimization of HDBscan
+with(sample_df, plot(x = minPts, y = value, 
+                     col = "grey", pch = 20, yaxt = "n",
+                     xlab = "", ylab = "", bty = "l",
+                     las = 1, cex = 0.3))
+with(score_df, lines(x = minPts, y = score,
+                     col = "black", las = 1))
+abline(v = optimal_minPts, col = "red") ## Red vertical line to show optimal minimum point for HDBscan
+title(ylab = "score", line = 3.5)
+title(xlab = "Minimum Cluster Size", line = 2)
+axis(2, at=c(0.25, 0.35, 0.45, 0.55), las = 1)
+mtext("D", side=3, line=0, cex=1.2, adj=-0.27) ## plot tag
 
-top100path[top100path == "1"] <- "MUT" ## categorize boolean values
-top100path[top100path == "0"] <- " "
-col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "Top 100 Pathway PanCancer"
-  onco_res <- oncoPrint(top100path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
+## Figure 2B: pvalue of eigenvalue 
+with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = pvalue,  ## line plot
+                                                 type = "l", bty = "l",
+                                                 col = "black", las = 1,
+                                                 xlab = "", ylab = ""))
+with(randomize %>% filter(eigenvalue < 50), points(x = eigenvalue, y = pvalue, pch = 20, ## add dots to line plot
+                                                   col = "black"))
 
-clust_col_order <- names(top100path[,onco_res@column_order])
-path_row_order <- rownames(top100path[onco_res@row_order,])
+## Determining the optimal mca dimension
+optimal_mca_dimensions <- randomize %>%
+  filter(pvalue < 0.05) %>% ## filtering for pvalue < 0.05
+  filter(eigenvalue == max(eigenvalue)) %>% ## filtering for the maximum eigenvalue
+  pull(eigenvalue)
 
-onco_ggplot <- gdc_reactome_path_analysis %>% .[rownames(.) %in% path_row_order,] %>% ## only include the top 10 paths
-  rownames_to_column(var = "path_name") %>% ## path_name to col
-  gather(., key = "sample_id", value = "mutation_status", -path_name) %>% ## wide to long format
-  mutate(path_name = factor(.$path_name, levels = c(rev(path_row_order))), ## factor to force oncoprint order
-         sample_id = factor(.$sample_id, levels = c(clust_col_order)))
-save(onco_ggplot, file = "onco_ggplot.Rda")
+abline(v = optimal_mca_dimensions, col = "red") ## Red vertical line to show optimal MCA dimension
+title(main = paste0("Significant dimensions = ", optimal_mca_dimensions), adj = 0, font = 3, cex.main = .8) ## Axis labels
+title(ylab = "P-value", line = 3.5)
+title(xlab = "MCA Eigenvalues", line = 2)
+mtext("B", side=3, line=0, cex=1.2, adj=-0.35) ## plot tag
 
-###################
-## barplot panel ##
-###################
+par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
+## Figure 2E: HDBscan
+cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
+names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = cluster.colours[as.character(clust)], pch = 21, cex = 0.8, lwd = 0.2, # clust is hdbscan result
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+mtext("E", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
 
-load("../data/kelly.colours.rda")
-load("../data/moderate_enrich_prob.Rda")
+## Figure 2F: kNN 
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                theta = 0, phi = 65, scale = F,
+                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                colvar = NULL))
+title(xlab="UMAP 2", line=0)
+mtext("F", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
 
-enrich_pathways <- moderate_enrich_prob %>%
-  group_by(clust) %>%
-  filter(difference > 0.3 & signif <= 2500) %>%
-  mutate(pathway = make_clean_names(path)) %>%
-  rename(clust_knn = "clust") %>%
-  select(clust_knn, pathway) %>%
-  mutate(clust_knn = as.numeric(clust_knn)) %>%
-  mutate(specific = TRUE)
-
-pathway_counts <- tumor_clusters %>%
-  left_join(t(gdc_reactome_path_analysis) %>% 
-              as.data.frame() %>% 
-              clean_names() %>% 
-              rownames_to_column("sample_id")) %>% 
-  group_by(clust_knn) %>%
-  summarise(total = n(), across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, sum)) %>%
-  mutate(across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, ~ .x / total)) %>%
-  select(-total) %>%
-  pivot_longer(-clust_knn, names_to = "pathway", values_to = "proportion") %>%
-  group_by(clust_knn) %>%
-  filter(pathway %in% enrich_pathways$pathway) %>%
-  left_join(enrich_pathways %>% select(clust_knn, pathway, specific), by = c("clust_knn", "pathway"))
-
-pathway_counts[is.na(pathway_counts$specific), ]$specific <- FALSE
-
-pathway_mat <- pathway_counts %>%
-  select(-specific) %>%
-  pivot_wider(names_from = pathway, values_from = proportion) %>%
-  ungroup() %>%
-  select(-clust_knn) %>%
-  as.matrix()
-
-pathway_dist <- dist(t(pathway_mat))
-pathway_clust <- as.hclust(rev(as.dendrogram(seriate(pathway_dist, method = "OLO_ward")[[1]])))
-pathway_counts$pathway <- factor(pathway_counts$pathway, levels = colnames(pathway_mat)[pathway_clust$order])
-bp.cluster.colors <- c(kelly.colours[3:12], "gray15")
-names(bp.cluster.colors) <- 1:11
-
-pathway_counts <- pathway_counts %>%
-  mutate(color = case_when(specific ~ clust_knn, TRUE ~ 11)) %>%
-  mutate(color = factor(color))
-save(pathway_counts, bp.cluster.colors, file = "barplot_pathways_fig2b.rda")
-
-#####################
-## oncoprint panel ##
-#####################
-clust_col_order <- list()
-path_row_order <- list()
-
-for(n in 1:max(tumor_clusters$clust_knn)) {
-  class_spec_sample_id <- tumor_clusters %>% .[,c("sample_id", "clust_knn")] %>% filter(clust_knn == n) %>% .[,.$sample_id] ## get class spec sample ids
-  top10path <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec_sample_id] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:10,] %>% .[, "path_name"] ## getting the top10 mutated pathways in class
-  top10path <- gdc_reactome_path_analysis[top10path, class_spec_sample_id]
-  top10path[top10path == "1"] <- "MUT" ## categorize boolean values
-  top10path[top10path == "0"] <- " "
-  col <- c("MUT" = "firebrick3")
-heatmap_legend <- list(title = "", at = c("MUT"),
-                       labels = c("Mutation"))
-alter_fun = list(
-    background = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = "#CCCCCC", col = NA))
-    },
-    # bug red
-    MUT = function(x, y, w, h) {
-        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
-            gp = gpar(fill = col["MUT"], col = NA))
-    })
-  column_title <- "test"
-  onco_res <- oncoPrint(top10path,
-    alter_fun = alter_fun, col = col,
-    column_title = column_title, heatmap_legend_param = heatmap_legend,
-    remove_empty_columns = F, remove_empty_rows = F, right_annotation = NULL,
-    top_annotation = NULL
-    )
-  clust_col_order[[n]] <- names(top10path[,onco_res@column_order])
-  path_row_order[[n]] <- rownames(top10path[onco_res@row_order,])
-}
-
-save(clust_col_order, path_row_order, file = "onco_clust.Rda")
 ```
-
-```{r figure-5-plot, fig.align= 'center', fig.height= 7, fig.width = 7}
-#out.height='90%', out.width='90%'
-## top100 pathways pancancer oncoprint
-load("../data/onco_ggplot.Rda") ## load data
-onco_color <- c("0" = "grey", "1" = "firebrick3") ## color scheme for oncoprint
-
-fig2A <- ggplot(data = onco_ggplot, aes(x=sample_id, y=path_name)) + 
-  geom_tile(aes(fill = as.character(mutation_status))) +
-  scale_fill_manual(values = onco_color) +
-  theme_classic() +
-  xlab("Pan-Cancer (n=7,607 tumor samples)") +
-  ylab("377 pathways") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 10),
-        axis.title.y = element_text(size = 10),
-        #plot.tag = element_text(size = 8, hjust = 0, vjust = 0),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        legend.position = "none")
-
-load("../data/effect_heatmap.Rda") ## load data
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
-
-guide_axis_label_trans <- function(label_trans = identity, ...) {
-  axis_guide <- guide_axis(...)
-  axis_guide$label_trans <- rlang::as_function(label_trans)
-  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
-  axis_guide
-}
-
-guide_train.guide_axis_trans <- function(x, ...) {
-  trained <- NextMethod()
-  trained$key$.label <- x$label_trans(trained$key$.label)
-  trained
-}
-
-fig2B <- ggplot(data = effect_heatmap, aes(y = cluster, x = path_name)) +
-  geom_tile(size = 1, aes(fill = effect_size)) +
-  scale_fill_fermenter(
-    type = "div",
-    palette = "PuOr",
-    n.breaks = 12,
-    limits = c(-1, 1),
-    name = "Effect Size",
-    labels = c(-1, "", 0.5, "", "", 0, "", "", 0.5, "", 1)
-  ) +
-  facet_wrap(~cluster, ncol = 1, strip.position = "right", scales = "free_y") +
-  guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
-  xlab("377 pathways, ward.D2 clustered") +
-  ylab("Clusters") +
-  scale_x_discrete(position = "top") +
-  guides(y.sec = guide_axis_label_trans(~ paste(.x))) +
-  theme(
-    axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_text(size = 10),
-    axis.title.y = element_text(size = 10),
-    axis.ticks = element_blank(),
-    strip.text.y = element_text(size = 10, hjust = .5),
-    strip.text.y.right = element_text(angle = 0),
-    strip.background = element_blank(),
-    panel.background = element_blank(),
-    panel.spacing = unit(0, "lines"),
-    legend.position = "bottom",
-    legend.key.height = unit(0.3, "lines"),
-    legend.text = element_text(size = 5),
-    legend.title = element_text(size = 5),
-    legend.margin=margin(0,0,0,0),
-    legend.box.margin=margin(-10,-10,0,-10)
-  )
-
-load("../data/barplot_pathways_fig2b.rda")
-fig2C <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
-  geom_col() +
-  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") +
-  ylab("Proportion of mutated samples") +
-  xlab("Significant pathways") +
-  theme_classic() +
-  scale_fill_manual(values = bp.cluster.colors) +
-  theme(
-    axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    axis.title.x = element_text(size = 10),
-    axis.title.y = element_text(size = 10),
-    axis.ticks = element_blank(),
-    axis.line = element_blank(),
-    #strip.text.y = element_text(size = 10, hjust = .5),
-    #plot.tag = element_text(size = 12, hjust = 0, vjust = 0),
-    strip.text.y.right = element_text(angle = 0),
-    strip.background = element_blank(),
-    legend.position = "none"
-  )
-
-layout <- '
-AAABBBB
-AAABBBB
-AAABBBB
-AAACCCC
-AAACCCC
-AAACCCC
-AAACCCC
-'
-
-fig2A + fig2B + fig2C + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
-```

### 2020-08-21 code/figure6.Rmd

--- 2020-08-13 code/figure4.Rmd
@@
 ```{r forest-heatmap-data-generation, eval = FALSE}
-# rm(list = ls())
-# gc()
-# setwd("/data/GDC_TCGA/TTmanu/wd/")
+load("../data/kelly.colours.rda")
 
-library(maftools)
-library(TCGAbiolinks)
-library(tidyverse)
-library(survival)
-
-kelly.colours <- c(
-  "gray95", "gray13", "gold2", "plum4", "darkorange1",
-  "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-  "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-  "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19"
-)
 cluster.colors <- kelly.colours[-c(1)]

### 2020-08-22 text/discussion.Rmd
reduce discussion, remove references to the results

@@
 elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
-@tamboreroOncodriveCLUST2013;
-@lawrenceDiscoverySaturationAnalysis2014;
+@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
 @kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@@
 
-To our knowledge, this approach has not previously been attempted in the
-literature in spite of its relative simplicity. We limited our analysis to
-mutations with likely deleterious effects, including missense, nonsense,
-frame-shift, stop-loss, untranslated region and splice site mutations in genes
-that are actively expressed in each cancer type, thus avoiding bias from
-transcription coupled repair. Our method of filtration differs from the
-"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
-account for the same biases. We also filtered genes on the basis of pathway
-membership in biochemical pathways, excluding curated gene sets related to
-diseases, syndromes, or classes of proteins with shared catalytic activity or
-conserved domains ( _e.g._ transcription factors) which are potentially
-problematic [@pathways-Atul-Butte]. We chose this approach to limit redundancy,
-which is already high in the remaining pathways for some driver genes, and to
-exclude biologically unrelated collections of genes. We do not account for
-non-coding mutations, amplification events or structural variants, though we
-hope to include them in subsequent studies (see below). Our pathway-centric
-analysis also allows us to circumvent the problem that some cancer-promoting
-mutations likely occur at such low frequencies as to be indistinguishable from
-background. Previous attempts to capture these associations using Bayesian
-networks have yielded important new insights into this problem and new genes
-[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
-traditional analysis, supporting the case that pathway-centric approaches have
-the potential to increase statistical power.
+To our knowledge, this approach has not previously been attempted in spite of
+its relative simplicity. We limited our analysis to mutations with likely
+deleterious effects, including missense, nonsense, frame-shift, stop-loss,
+untranslated region and splice site mutations in genes that are actively
+expressed in each cancer type, thus avoiding bias from transcription coupled
+repair. Our method of filtration differs from the "rank-and-cut" method of
+[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
+biases. We also filtered genes on the basis of pathway membership in biochemical
+pathways, excluding curated gene sets related to diseases, syndromes, or classes
+of proteins with shared catalytic activity or conserved domains which are
+potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
+redundancy and exclude biologically unrelated collections of genes. Our
+pathway-centric analysis allows us to circumvent the problem that some
+cancer-promoting mutations likely occur at such low frequencies as to be
+indistinguishable from background. Previous attempts to capture these
+associations using Bayesian networks have yielded important new insights into
+this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that would not
+have been possible with traditional analysis, supporting the case that
+pathway-centric approaches have the potential to increase statistical power.
 
 The hypothesis that cancer is the result of dysfunction from a limited number of
-basic cellular processes common to all living human cells was first introduced
-and later expanded on in a pair of classic essays by Hanahan and Weinberg
-[@weinberg2000; @hallmarks_II]. In this view, all tumors regardless of tissue of
-origin acquire six essential phenotypes; sustaining proliferative signaling,
-evading growth suppressors, replicative immortality, evading apoptosis,
-supportive angiogenesis, tissue invasion and metastasis [@weinberg2000].
-Advancements in the field over the ensuing decade led to the inclusion of new
-tumor promoting processes, namely deregulation of cellular energy cycle, tumor
-promoting inflammation, evasion of the immune system, genome instability, and on
-top of these acquired traits a new appreciation for the concept of the tumor
+basic cellular processes common to all living human cells was introduced and
+later expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
+@hallmarks_II]. In this view, all tumors regardless of tissue of origin acquire
+six essential phenotypes; sustaining proliferative signaling, evading growth
+suppressors, replicative immortality, evading apoptosis, supportive
+angiogenesis, tissue invasion and metastasis [@weinberg2000]. Advancements in
+the field over the ensuing decade led to the inclusion of new tumor promoting
+processes, namely deregulation of cellular energy cycle, tumor promoting
+inflammation, evasion of the immune system, genome instability, and on top of
+these acquired traits a new appreciation for the concept of the tumor
 microenvironment and its contribution to the evolution and promotion of
@@
 manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways as represented by our clusters (**figures 1 & 2**).
+disruptions in basic pathways as represented by our clusters.
 
@@
 clusters on the UMAP projection exhibit increasing specificity. We identified no
-such trend in the number of mutations (see **figure 4**), the relative staging
-(**figure 3D**) or metastasis (**figure 3E**); and each cluster instead was
-associated with unique combinations of pathways (as in **figures 5B, C**).
+such trend in the number of mutations, the relative staging or metastasis; and
+each cluster instead was associated with unique combinations of pathways.
 
 We find that most tumors segregate into one of ten biologically distinct
-clusters, regardless of the cancer type (**figure 2F, 3A** and **supplementary
-figure S1**). Some cancer types ( _e.g._ Pancreatic:PAAD in the RAS mutant
-cluster 8, and Kidney Chromophobe:KICH in cluster 1, see **supplementary figure
-S1**) are unevenly distributed, though we could not identify any cancers that
-were exclusive to a single tumor cluster. Only cancer types with the fewest
-samples were found to be absent from one or more clusters at all. Somewhat
-surprisingly to us, this finding extends to histological subtypes of breast,
-head and neck cancers, leukemias, _etc._ (see **Figure 3B** and **Supplemental
-Figure S2**). This result implies that many, if not most, histological subtypes
-could reflect differences in cell-of-origin, rather than fundamental differences
-in proximal cause. The four major subtypes of breast cancer correspond to
-histological and molecular expression profiles that define them and how they
-respond to experimental stimuli [@breast-subtypes2]. It has been hypothesized
-that differences in the molecular regulators of development in the precursor
-cell types present in the breast epithelium drive the various histological
-phenotypes [@breast-subtypes;
-@breast-subtypes3]. Consistent with this view we found that breast tumor samples
-of the Luminal A subtype were heavily biased toward membership in clusters 1 and
-7, and basal tumor samples were biased toward cluster 6 (**figure 3B**).
-However, both subtypes also contained samples in every other cluster (without
-exception), and Luminal B and Her2 positive samples are distributed across
-clusters. We interpret these data to mean that inherited cell-of-origin
-molecular profiles may predispose certain precursor cell types within the breast
-epithelium to forming tumors of one cluster or another but are not
-determinative; this view is compatible with the previously stated hypothesis but
-opens the way for a more granular view of individual tumors. It would be
+clusters, regardless of the cancer type. Some cancer types are unevenly
+distributed, though we could not identify any cancers that were exclusive to a
+single tumor cluster. Only cancer types with the fewest samples were found to be
+absent from one or more clusters at all. Somewhat surprisingly to us, this
+finding extends to histological subtypes of breast, head and neck cancers,
+leukemias, _etc._ This result implies that many, if not most, histological
+subtypes could reflect differences in cell-of-origin, rather than fundamental
+differences in proximal cause. The four major subtypes of breast cancer
+correspond to histological and molecular expression profiles that define them
+and how they respond to experimental stimuli [@breast-subtypes2]. It has been
+hypothesized that differences in the molecular regulators of development in the
+precursor cell types present in the breast epithelium drive the various
+histological phenotypes [@breast-subtypes; @breast-subtypes3]. Consistent with
+this view we found that breast tumor samples of the Luminal A subtype were
+heavily biased toward membership in clusters 1 and 7, and basal tumor samples
+were biased toward cluster 6. However, both subtypes also contained samples in
+every other cluster (without exception), and Luminal B and Her2 positive samples
+are distributed across clusters. We interpret these data to mean that inherited
+cell-of-origin molecular profiles may predispose certain precursor cell types
+within the breast epithelium to forming tumors of one cluster or another but are
+not determinative; this view is compatible with the previously stated hypothesis
+but opens the way for a more granular view of individual tumors. It would be
 surprising if we did not observe bias for some cancers and subtypes amongst our
@@
 genes with tissue-specific pathway disruptions result in common categories when
-viewed at the level of the pathway knowledge-base. For example, we find that for
-"snRNP assembly", one of the unique driver pathways of cluster 5, the top gene
-in breast cancer is represented by TPR (mutated in 26% of samples). By contrast
-the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of
-samples, respectively). For some other clusters, _e.g._ 7, 8 & 9, the top genes
-are known drivers such as TP53, PIK3CA and KRAS. None of these genes is 100%
-mutated in any cluster however, and therefore it is the combinations with
-secondary mutations, many of which are tissue-specific, that round out the
-complement of pathways in all clusters. Some top genes are found in most of the
-clusters (**supplemental figure S5**), in spite of the fact that they contribute to
-many cluster-specific pathways. This can only be explained as a result of other
-less common driver genes complementing the pathway disruption of the unique
-combination of driver genes that are disrupted in each tumor, and we speculate
-that many of these are relatively tissue-specific in terms of their sensitivity
-to mutation. Consistent with our observations, @moonlight showed how tissue- and
-cancer-specific driver genes and druggable targets of pathway moduli can be
-discovered by machine learning on integrated multi-omics datasets. Marrying
-these approaches in future work will greatly enhance our understanding of tumor
-biology.
+viewed at the level of the pathway knowledge-base. Top cancer driver genes (e.g.
+PIK3C and TP53) are found in most of the clusters, in spite of the fact that
+they contribute to many cluster-specific pathways. This can only be explained as
+a result of other less common driver genes complementing the pathway disruption
+of the unique combination of driver genes that are disrupted in each tumor, and
+we speculate that many of these are relatively tissue-specific in terms of their
+sensitivity to mutation. Consistent with our observations, @moonlight showed how
+tissue- and cancer-specific driver genes and druggable targets of pathway moduli
+can be discovered by machine learning on integrated multi-omics datasets.
+Marrying these approaches in future work will greatly enhance our understanding
+of tumor biology.
 
@@
 clusters, 1 and 10, for which we did not find many distinctive associations with
-pathways (**figure 5B, C**). Cluster 1 had a relatively low proportion of
-mutated pathways across the board, although it is broadly enriched in many of
-the same tumor-promoting pathways common to the other groups. This likely does
-not reflect a true difference in stages of tumor evolution, as our data show
-clearly that this cluster is as likely to contain stage IV metastatic tumors as
-it is to contain those of stage I (**figure 3D, E**). This cluster likely
-represents a group of tumors with aberrations in methylation, copy number
-alterations, structural variants, or combinations thereof which we excluded from
-our analysis. Consistent with this, kidney chromophobe and thyroid cancers have
-high proportions of structural variations _vs._ other variant types
-[@PCAWG-MAIN] and are heavily skewed to cluster 1 membership (**Supplementary
-figure S1**). Likewise, cluster 10 represents a group of hyper-mutated tumors
-that harbor so many mutations that virtually no pathway is unaffected. It seems
-likely that a significant fraction of the “mutant” samples for each pathway are
-burdened with excess passenger mutations. This could be addressed with
-application of ever-increasingly sophisticated filtering of likely passenger
-mutations [ _e.g._
-@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
-incorporate these other data into a comprehensive pathway-centric analysis as we
-have done here for point mutations and indels.
+pathways. Cluster 1 had a relatively low proportion of mutated pathways across
+the board, although it is broadly enriched in many of the same tumor-promoting
+pathways common to the other groups. This likely does not reflect a true
+difference in stages of tumor evolution, as our data show clearly that this
+cluster is as likely to contain stage IV metastatic tumors as it is to contain
+those of stage I. This cluster likely represents a group of tumors with
+aberrations in methylation, key deletions, or other structural variants.
+Consistent with this, kidney chromophobe and thyroid cancers have high
+proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and
+are heavily skewed to cluster 1 membership. Likewise, cluster 10 represents a
+group of hyper-mutated tumors that harbor so many mutations that virtually no
+pathway is unaffected. It seems likely that a significant fraction of the
+“mutant” samples for each pathway are burdened with excess passenger mutations.
+This could be addressed with application of ever-increasingly sophisticated
+filtering of likely passenger mutations [ _e.g._ @spliceAI;
+@deep-learning-clinical-variants]. In the future, we hope to incorporate these
+other data into a comprehensive pathway-centric analysis as we have done here
+for point mutations and indels.
 
@@
 of our pathway disruption clusters exhibit cancer-type specific effects on
-expected survival (**figure 9**). Contrary to what one might expect at the
-outset, we found that models assuming different survival based on cluster and
-cancer type fit better than models in which they are independent factors.
-However, considering that within each cancer type there are different clinical
-standards of care, and even within classes of drugs the preferred treatment can
-very between cancers, it makes sense that we observe tissue-specific cluster
-effects. Contrast the situation with ovarian _vs._ breast cancer, which are both
-hormonally driven cancers, for example. Ovarian cancer has but one main
-treatment axis, platinum, whereas breast cancer patients have a variety of
-treatment regimens based on molecular subtype and other factors. It is easy to
-see how standard of care for one cancer and cluster could produce different
-outcomes than the same cluster of a different cancer type. Unfortunately, given
-the diversity of drug classes and treatments (including non-chemical treatments
-such as surgery and palliative care), we lack sufficient power to explore these
-variables in the TCGA data. It is our hope that future studies will help to
-distinguish between treatment-specific effects on survival given different
-pathway disruption clusters.
+expected survival. Contrary to our initial expectation, we found that models
+assuming different survival based on cluster $\times$ cancer type fit better than
+models in which they are independent factors. However, considering that within
+each cancer type there are different clinical standards of care, and even within
+classes of drugs the preferred treatment can very between cancers, it makes
+sense that we observe tissue-specific cluster effects. Contrast the situation
+with ovarian _vs._ breast cancer, which are both hormonally driven cancers, for
+example. Ovarian cancer has but one main treatment axis, platinum, whereas
+breast cancer patients have a variety of treatment regimens based on molecular
+subtype and other factors. Standard of care for one cancer and cluster could
+produce different outcomes than the same cluster of a different cancer type.
+Unfortunately, given the diversity of drug classes and treatments (including
+non-chemical treatments such as surgery and palliative care), we lack sufficient
+power to explore these variables in the TCGA data. It is our hope that future
+studies will help to distinguish between treatment-specific effects on survival
+given different pathway disruption clusters.
 
@@
 characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
-disrupted in all or most of the defined clusters (**supplementary figure S1**)
+disrupted in all or most of the defined clusters
 and are frequently the result of aberrations involving common driver genes such
 as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
-(**supplementary figure S3** and **supplemental file 1**). Cluster-specific
+Cluster-specific
 evolution must involve the acquisition of disruptions to pathways that may
@@
 senesce) but together produce more specialized cancer phenotype and increased
-fitness (as in **figure 6** and **tables 1** and **ST2**). Importantly, our
+fitness. Importantly, our
 observations do not imply the order in which these mutations should accumulate.
@@
 cluster specific analysis, suggesting that signal is diluted when the samples
-are analyzed together, and supporting the view that, within the parameters of
-our analysis, metastasis has cluster-specific requirements. Since the number of
-metastatic samples is relatively low this part of our analysis is likely
-underpowered and subject to expanded analysis with larger cohorts. Nonetheless,
-the cluster-specific metastatic pathway enrichment we observe (in **table 1**)
-could be a result of the pathways directly promoting metastatic phenotypes in
-these clusters, or because they acquire a stronger or more complete oncogenic
-phenotype, as we don’t know the order in which they were acquired. The fact that
-most metastatic enrichment is cluster-specific and has a tendency to overlap
-with cluster-specific pathways from non-metastatic tumors of neighboring
-clusters favors the latter interpretation, and suggests a mechanism of
+are analyzed together, and supporting the view that metastasis has
+cluster-specific requirements. Since the number of metastatic samples is
+relatively low, this part of our analysis is likely underpowered and subject to
+expanded analysis with larger cohorts. The fact that most metastatic enrichment
+is cluster-specific and has a tendency to overlap with cluster-specific pathways
+from non-metastatic tumors of neighboring clusters suggests a mechanism of
 complementarity, in which newly acquired mutations result in similar clusters
 converging on one or more deadly phenotypes that capture most of the critical
-features of end-stage cancer. One explanation for this could be that tumors must
-accumulate mutations in a certain order or face elimination due to any of the
-selective bottlenecks they face (senescence, the immune system, _etc._)
-[@hallmarks_II]. Some clusters were characterized by cluster-specific pathways
-that seem superficially unrelated; for example cluster 2 has mutations in the
-proteasome subunits, cluster 3 has mutations impacting protein translation,
-whereas cluster 5 has defects in RNA metabolism and clusters 7, 8 and 9 have
-defects in the more traditional extra-cellular signaling pathways (as
-exemplified in **table S1**). It may be that the primary effects of mutations in
-protein and RNA metabolism in clusters 3 and 5 point to currently
-underappreciated mechanisms impacting hallmark pathways. Clusters 3 and 5 are
-also marked by a decidedly different set of pathways in metastasis. Clusters 3
-and 5 notably have (in addition to signaling pathways) mutations in glucagon and
-glycogen metabolism, respectively, in their metastases (**table 1**), consistent
-with energy metabolism as one of the new hallmarks [@hallmarks_II], but also
-distinguishing these clusters from the others in the order or nature of hallmark
-acquisition. Thus, even at relatively low power, our analysis of metastasis
-further supports the pathway-centric approach as a fruitful way to uncover real
-differences in underlying biology. As a major caveat to this, we acknowledge
+features of end-stage cancer. Thus, even at relatively low power, our analysis
+of metastasis further supports the pathway-centric approach as a fruitful way to
+uncover biologically meaningful differences between tumors against the noisy
+backdrop of tissue specific profiles. As a major caveat to this, we acknowledge
 that many therapies _do_ target highly-specific driver genes, markers, and

### 2020-08-24 text/manuscript.Rmd
add supplementary methods, move clustering and survival methods to supp methods doc, rename figure2.Rmd-->figureSM1.Rmd

@@
 
-In order to classify tumors using this dataset, we used multiple correspondence
-analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
-containing useful information by selecting the eigenvalue with the most
-explanatory power, using the average of 100 permutations of the data as baseline
-(**figure 2A**). We then chose the maximum eigenvalue for which the p-value
-remained ≤ 0.05 (see cutoff in **figure 2B**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several reproducible projections
-regardless of random seed setting. A representative version of this 3D UMAP is
-shown in **figure 2C**, rotated to enhance the visibility of the major features.
-Following this spatial mapping we attempted to define groupings of similar
-tumors within the spatial map using HDBSCAN, which performs hierarchical
-clustering and provides metrics of cluster stability and probabilities of
-cluster membership for each node. However, HDBSCAN is sensitive to several
-parameters; key for our analysis are the minimum number of tumor samples in a
-cluster that capture the maximum number of tumor samples, measured by
-probability of membership of ≥ 5% in at least one class. Thus, we created a
-score metric as the fraction of classified tumors with max probability < 5% in
-one class and chose a cluster size of 92 to minimize the score function
-(**figure 2D**). HDBSCAN with these settings resulted in ten distinct
-high-density clusters which we then projected onto the UMAP (**figure 2E**).
-This classified 6,038 out of 7,607 tumors but still resulted in a significant
-fraction of unclassified tumor samples. Since we ultimately wish to be able to
-classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
-analysis, which computes a similarity metric to every tumor in the set and then
-lets the _k_ most similar tumors "vote" as to the identity of the query tumor
-sample based on their classification labels. We set _k_ to be the square root of
-the number of tumor samples (87). Using this method, we assigned cluster
-membership to the remaining tumors (**figure 2F**).
+We investigated this dataset further using multiple correspondence analysis
+(MCA), and visually summarized the analysis with UMAP (**figure 2**). We used
+the resulting UMAP graph coordinates to perform density based clustering with
+HDBSCAN, which resulted in identification of 10 well-defined clusters capturing
+about 80% of the tumor samples. To capture the remaining samples into one of
+these 10 clusters we used kNN (see **Supplementary Methods** for details on all
+clustering algorithms).
 
@@
 
-#### Figure 2
-```{r child='../code/figure2.Rmd'}
+```{r, child='figure2.Rmd'}
 ```
-
-**Figure 2: Method for tumor classification using pathway aberration data** 
-a) MCA analysis was performed to determine the eigenvalue with the most
-explanatory power from the categorical data. Average of 100 permutations of the
-data was used as a baseline. b) The p-value of each eigenvalue was plotted and
-the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
-projected down to 3 dimensions for the purpose of density based clustering (only
-the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
-used to perform hierarchical density based clustering tuned to minimize the
-fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
-the UMAP projection; tumors are colored by cluster membership (see legend). This
-results in some tumors remaining unclassified (grey points). f) kNN was used to
-assign membership to the remaining tumors.
 
@@
 
-```{r child='survival-methods.Rmd'}
-```
+We used Bayesian inference to explore the relative impact of cancer diagnosis on
+survival at the time of diagnosis. Our model assumes that at birth each person
+has a small initial probability per unit time to die, and that this probability
+grows exponentially in time at some baseline rate that matches the observed CDC
+data. The rate of growth in risk is assumed to jump to a higher level at
+diagnosis, which we take as a proxy for onset. This represents an accelerated
+aging type model. To fit the model, we constrain the baseline risks using the
+available CDC life tables. We then constrain the cancer specific multiplier
+using the longevity data within the cancer tumor dataset. Our model estimates
+the effect of cancer type and class-specific cancer effects independently,
+resulting in a cancer and class specific estimate of the aging rate multiplier.
+A detailed description of the CDC longevity model and likelihood is provided in
+the supplementary methods.
 

### 2020-08-24 text/manuscript.Rmd
add acknowledgements, cut survival section a little more

@@
 
-To discover what pathways drive the UMAP clustering, we calculated percent
-enrichment for each pathway _within cluster_ relative to all other clusters
-combined (see methods) and ranked pathways for each cluster from highest to
-lowest enrichment. We visualized the enrichment as a heatmap (**figure 5B**).
-Using this approach, we identified 25 pathways highly enriched (enrichment score
-$> 0.3$, see methods) in cluster 2, 9 pathways enriched in cluster 3, 24
-pathways in cluster 4, 10 pathways in cluster 5, 10 pathways in cluster 6, 24
-pathways in cluster 7, 24 pathways in cluster 8, 10 pathways in cluster 9, and 0
-pathways for clusters 1 and 10. Clusters 7, 8, and 9 had several pathways in
-common. To explore the specific pathways marking each cluster, we projected
-disruptions for each of the 377 pathways onto the UMAP (**supplementary figure
-S4**). Clusters 3 and 5 were distinguished by metabolic pathways including RNA
-and protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was
-distinguished by mutations affecting regulation of DNA and histone methylation
-("DNA methylation", "PRC2 methylates histones and DNA", and "Nucleosome
-assembly"). Clusters 7-9 have in common mutations in extracellular,
-intracellular, and immune-related signaling pathways (see **figure 5C** and
-**supplementary figure S4**). Cluster 2 had the highest pathway enrichment
-levels of the three, having mutations in hedgehog signaling, "β-catenin
-degradation", "cellular response to hypoxia", "regulation of cell cycle" and
-"apoptosis" among others.
+To discover what pathways are most important for clustering, we calculated
+percent enrichment for each pathway _within cluster_ relative to all other
+clusters combined (see methods) and ranked pathways for each cluster from
+highest to lowest enrichment. We visualized the enrichment as a heatmap
+(**figure 5B**). Using this approach, we identified 25 pathways highly enriched
+(enrichment score $> 0.3$, see methods) in cluster 2, 9 pathways enriched in
+cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10 pathways in
+cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10 pathways in
+cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and 9 had
+several pathways in common. To explore the specific pathways marking each
+cluster, we projected disruptions for each of the 377 pathways onto the UMAP
+(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
+pathways including RNA and protein biosynthesis (**supplementary figure S4**).
+Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
+and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
+and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
+extracellular, intracellular, and immune-related signaling pathways (see
+**figure 5C** and **supplementary figure S4**). Cluster 2 had the highest
+pathway enrichment levels of the three, having mutations in hedgehog signaling,
+"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
+cycle" and "apoptosis" among others.
 
@@
 A detailed description of the CDC longevity model and likelihood is provided in
-the supplementary methods.
+the **Supplementary methods**.
 
@@
 
+Wish wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
+reading of this manuscript, and Ivetth Corona for early discussions and
+feedback.
+
 ## Declarations
 
-The authors have nothing to declare. This work was supported by a grant from the Cedars-Sinai Precision Health Initiative.
+The authors have no conflicts of interest to declare. This work was supported by
+a grant from the Cedars-Sinai Precision Health Initiative.
 

### 2020-08-24 text/manuscript.Rmd
Update fig3 with sankey plot, remove oncoprint in fig5

--- 2020-08-21 text/manuscript.Rmd
@@
   library(survival),
-  library(ggalluvial)
+  library(ggalluvial),
+  library(grid),
+  library(gridBase)
 ))

### 2020-08-24 text/supplemental_methods.Rmd

--- 2020-08-14 text/survival-methods.Rmd
@@
+---
+title: "Supplementary methods"
+author: "Peter T. Nguyen, Simon G. Coetzee, Daniel L. Lakeland, and Dennis J. Hazelett"
+date: "8/24/2020"
+bibliography: references.bib
+output:
+  word_document:
+    reference_docx: "../text/word-style-reference-ttmanu.docx"
+always_allow_html: TRUE
+editor_options:
+  chunk_output_type: console
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
+library(foreach)
+library(sqldf)
+library(tidyverse)
+library(ggthemes)
+library(data.table)
+library(biomaRt)
+library(dplyr)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(reactome.db)
+library(plyr)
+library(cluster)
+library(pheatmap)
+library(clusterProfiler)
+library(org.Hs.eg.db)
+library(dbscan)
+library(class)
+library(TCGAbiolinks)
+library(patchwork)
+library(ggplotify)
+library(ComplexHeatmap)
+library(knitr)
+library(kableExtra)
+library(UpSetR)
+library(cowplot)
+library(doRNG)
+library(uwot)
+library(janitor)
+library(FactoMineR)
+library(seriation)
+library(rtracklayer)
+library(poibin)
+library(flextable)
+library(plot3D)
+library(ggfortify)
+#library(maftools)
+library(survival)
+library(ggalluvial)
+```
+
+## Supplementary Methods
+
+### Clustering
+
+In order to classify tumors using this dataset, we used multiple correspondence
+analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
+containing useful information by selecting the eigenvalue with the most
+explanatory power, using the average of 100 permutations of the data as baseline
+(**figure SM1A**). We then chose the maximum eigenvalue for which the p-value
+remained ≤ 0.05 (see cutoff in **figure SM1B**). Then we performed a UMAP
+analysis, both in order to summarize the MCA graphically, and as a preprocessing
+step to boost the performance of density based clustering. The resulting map was
+notable for its lobed structure, with several reproducible projections
+regardless of random seed setting. A representative version of this 3D UMAP is
+shown in **figure SM1C**, rotated to enhance the visibility of the major features.
+Following this spatial mapping we attempted to define groupings of similar
+tumors within the spatial map using HDBSCAN, which performs hierarchical
+clustering and provides metrics of cluster stability and probabilities of
+cluster membership for each node. However, HDBSCAN is sensitive to several
+parameters; key for our analysis are the minimum number of tumor samples in a
+cluster that capture the maximum number of tumor samples, measured by
+probability of membership of ≥ 5% in at least one class. Thus, we created a
+score metric as the fraction of classified tumors with max probability < 5% in
+one class and chose a cluster size of 92 to minimize the score function
+(**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
+high-density clusters which we then projected onto the UMAP (**figure SM1E**).
+This classified 6,038 out of 7,607 tumors but still resulted in a significant
+fraction of unclassified tumor samples. Since we ultimately wish to be able to
+classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
+analysis, which computes a similarity metric to every tumor in the set and then
+lets the _k_ most similar tumors "vote" as to the identity of the query tumor
+sample based on their classification labels. We set _k_ to be the square root of
+the number of tumor samples (87). Using this method, we assigned cluster
+membership to the remaining tumors (**figure SM1F**).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure SM1
+```{r child='../code/figureSM1.Rmd'}
+```
+
+**Figure SM1: Method for tumor classification using pathway aberration data** 
+a) MCA analysis was performed to determine the eigenvalue with the most
+explanatory power from the categorical data. Average of 100 permutations of the
+data was used as a baseline. b) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
+projected down to 3 dimensions for the purpose of density based clustering (only
+the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
+used to perform hierarchical density based clustering tuned to minimize the
+fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
+the UMAP projection; tumors are colored by cluster membership (see legend). This
+results in some tumors remaining unclassified (grey points). f) kNN was used to
+assign membership to the remaining tumors.
+
+### Survival
+
 In order to understand how the diagnosis of cancer changes the expectation for
-longevity of patients, we first modeled the longevity of the overall
-population of the U.S. based on CDC life tables. Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes accuracy in the right tail of the distribution. We modeled
-baseline longevity using a risk rate function:
+longevity of patients, we first modeled the longevity of the overall population
+of the U.S. based on CDC life tables. Most TCGA cases are diagnosed late in
+life, so we fit a model that emphasizes accuracy in the right tail of the
+distribution. We modeled baseline longevity using a risk rate function:
 
@@
 and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
-is an indicator for whether diagnosis has occurred or not (before diagnosis $D(t) = 0$, 
-after $D(t) = 1$). The model assumes that the risk of death
-is $R_{20}$ at age 20 years and increases exponentially with a
-constant rate $a$ up to the age at diagnosis. Risk of death assumes a new rate $ka$
-thereafter. Though the model extends to earlier ages, relatively few
-patients were under age 20.
+is an indicator for whether diagnosis has occurred or not (before diagnosis
+$D(t) = 0$, after $D(t) = 1$). The model assumes that the risk of death is
+$R_{20}$ at age 20 years and increases exponentially with a constant rate $a$ up
+to the age at diagnosis. Risk of death assumes a new rate $ka$ thereafter.
+Though the model extends to earlier ages, relatively few patients were under age
+20.
 
@@
 tissue specific multiplicative modifying factor. This allows the overall
-magnitude of $k$ to be primarily determined by the $k_{tis}$ value whose prior is
-significantly less constrained so that there is a wider range of risk across
+magnitude of $k$ to be primarily determined by the $k_{tis}$ value whose prior
+is significantly less constrained so that there is a wider range of risk across
 tissue types.
 
-The $k$ in our model can be taken as, roughly, the aging rate relative
-to the baseline rates. Thus, a value of $k=2$ means that a given
-cancer causes you to experience the same risk in time $dt$ as a
-non-cancer patient would experience at the same age, in time $2dt$.
+The $k$ in our model can be taken as, roughly, the aging rate relative to the
+baseline rates. Thus, a value of $k=2$ means that a given cancer causes you to
+experience the same risk in time $dt$ as a non-cancer patient would experience
+at the same age, in time $2dt$.
 
-Using this model, we split the data by Male/Female sex, and by White, _vs._ Black
-race as these were the main well-identified categories available to us in the
-dataset. Sex is known to be a risk factor for death, with males dying at
+Using this model, we split the data by Male/Female sex, and by White, _vs._
+Black race as these were the main well-identified categories available to us in
+the dataset. Sex is known to be a risk factor for death, with males dying at
 slightly higher rates for all ages of interest in our dataset. Black _vs._ White
@@
 
-Using the model described above we fit the data using Stan via the rstan package
-in R. Posterior inferences are shown in **Figure 7**.
-
-**supplementary figure S5** gives a comparison on log-scale between
+**supplementary figure SM2** gives a comparison on log-scale between
 the post-fit survival curves and the CDC data it is based on.
@@
 under-powered given the uneven representation of stage across cancer types.
+
+### References

### 2020-08-24 text/survival.Rmd

@@
 apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. Kaplan-Meier plots reveal large differences between
-cancer types as expected, and some modest differences between our classes (not
-shown). These analyses are profoundly limited by confounding factors of age and
-stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
-name the obvious ones. To explore these ideas, we tested models of survival
-using public longevity data from the CDC, accounting for age, gender and
-ethnicity. Our model assumes that at birth each person has a small initial
-probability per unit time to die, and that this probability grows exponentially
-in time at some baseline rate that matches the observed CDC data. The rate of
-growth in risk is assumed to jump to a higher level at diagnosis, which we take
-as a proxy for onset. This represents an accelerated aging type model. To fit
-the model, we constrain the baseline risks using the available CDC life tables.
-We then constrain the cancer specific multiplier using the longevity data within
-the cancer tumor dataset. Our model estimates the effect of cancer type and
-class-specific cancer effects independently, resulting in a cancer and class
-specific estimate of the aging rate multiplier.
+types or across all cancers. These analyses are profoundly limited by
+confounding factors of age and stage at diagnosis, sex, ethnicity, and
+tissue-specific disease progression, to name the obvious ones. To explore these
+ideas, we used Bayesian inference to test models of survival using public
+longevity data from the CDC, accounting for age, gender and ethnicity. Our model
+estimates the effect of cancer type and class-specific cancer effects
+independently, resulting in a cancer and class specific estimate of the aging
+rate multiplier.
 

### 2020-08-24 text/manuscript.Rmd
minor edit to main doc

--- 2020-08-24 text/manuscript.Rmd
@@
 other models are available from [our
-repositories](https://github.com/dennishazelett/TTmanu) in the following
-locations on the distributed version control site GitHub.
+repositories](https://github.com/dennishazelett/TTmanu) on the distributed
+version control site, GitHub.
 

### 2020-08-24 code/figureSM1.Rmd
Update chunk code name in SM1

--- 2020-08-21 code/figure2.Rmd
@@
-```{r figure-2-plot, fig.width = 9}
+```{r figure-SM1-plot, fig.width = 9}
 load("../data/figure1B_data.rda") ## Load in data processed in figure1.Rmd

### 2020-08-24 text/manuscript.Rmd
typos in table 1 legend
Update chunk code name in SM1

--- 2020-08-24 text/manuscript.Rmd
@@
 
-```{r, child='figure2.Rmd'}
+```{r, child='../code/figure2.Rmd'}
 ```
@@
 
-**Table 1: Cluster-specific enriched pathways (effect size >= 0.30) in metastatsis.**
+**Table 1: Cluster-specific enriched pathways (effect size ≥ 0.30) in metastasis.**
 

### 2020-08-24 code/figure3.Rmd

@@
 ## Sankey plot
-ggplot(sankey.data,
+sankey.plot <- ggplot(sankey.data,
        aes(x = type, stratum = value, alluvium = sample_id,
@@
   theme(legend.position = "none",
-        axis.text.x = element_text(size = 10),
+        axis.text.x = element_blank(),
         axis.text.y = element_blank(),
@@
         axis.line.x = element_blank(),
-        axis.line.y = element_blank()) +
-  scale_x_discrete(labels= c("Cluster", "Cancer Type"))
+        axis.line.y = element_blank(),
+        plot.margin = unit(c(-0.4, -3, -1, -2.5), "cm"),
+        panel.background = element_rect(fill = "transparent"))
 ```
 
-```{r figure-4-plot, fig.width = 9}
-### Figure A - Cancer Type
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
-par(mar = c(2,2,0,0)) ## setting plot margins for the following plots
+```{r, fig.width= 6, fig.height= 6}
+# start new page
+plot.new()
 
-selectedcancer <- c("BRCA", "COAD", "HNSC", "PAAD") ## selecting the 4 cancer types we're interested in
-A_colors <- c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4') ## Color scheme for cancer type
-names(A_colors) <- selectedcancer ## Assigning cancertype a color
+# setup layout
+gl <- grid.layout(nrow=8, ncol=3)
 
-with(umap_cancertype %>% 
-       mutate(A_color = case_when(project_code %in% selectedcancer ~ project_code, ## Create a new col named A color
-                                  TRUE ~ "Other")) %>% ## Convert all cancer type that isn't the selected cancers to "Other"
-       filter(A_color != "Other"), ## Remove "Other" type
-     scatter3D(plot_y, -plot_z, plot_x, 
-               bg = A_colors[A_color], pch = 21, cex = 0.8, lwd = 0.2,
-               theta = 0, phi = 65, scale = F,
-               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-               colvar = NULL))
-## Axis labels
-legend("top", inset = c(0), legend=rev(names(A_colors[1:2])), col=rev(A_colors[1:2]), pch = 20, bty = "n", horiz = T, cex = 1.2)
-legend("top", inset = c(0.10), legend=rev(names(A_colors[3:4])), col=rev(A_colors[3:4]), pch = 20, bty = "n", horiz = T, cex = 1.2)
-title(xlab="UMAP 2", line=0)
-mtext("A", side=3, line=-2.05, cex=1.2, adj=0)
+## Setup viewports - How figures will be laid out
+vp.1 <- viewport(layout.pos.col=1:2, layout.pos.row=1:8)
+vp.2 <- viewport(layout.pos.col=3, layout.pos.row=1:2) 
+vp.3 <- viewport(layout.pos.col=3, layout.pos.row=3:4) 
+vp.4 <- viewport(layout.pos.col=3, layout.pos.row=5:6) 
+vp.5 <- viewport(layout.pos.col=3, layout.pos.row=7:8)
 
-### Figure B - BRCA subtype
+# init layout
+pushViewport(viewport(layout=gl))
+
+## vp.1 - Figure A - sankey plot
+pushViewport(vp.1) ## Enter vp.1 viewpoint
+print(sankey.plot, newpage = FALSE) ## Plot ggplot
+mtext("A", side=3, line=2.6, cex=1.2, adj=-0.17)
+popViewport() ## Exit vp.1 viewpoint
+
+## vp.2 - Figure B - BRCA subtype
+pushViewport(vp.2)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
 brca_subtype <- umap_cancertype %>% filter(project_code == "BRCA") %>% ## Filter for selected cancer type
@@
                theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
@@
 ## Axis labels
-legend("top", inset = c(0), legend=names(B_color[1:2]), col=B_color[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-legend("top", inset = c(0.10), legend=names(B_color[3:4]), col=B_color[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
-legend("top", inset = c(0.20), legend=names(B_color[5:6]), col=B_color[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
-title(xlab="UMAP 2", line=0)
-mtext("B", side=3, line=-2.05, cex=1.2, adj=0)
+legend("top", inset = c(0), legend=names(B_color[1:2]), col=B_color[1:2], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.07), legend=names(B_color[3:4]), col=B_color[3:4], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.14), legend=names(B_color[5:6]), col=B_color[5:6], pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("B", side=3, line=-1.2, cex=1.2, adj=0)
 
-### Figure C - BRCA Somatic Mutations
-load("../data/gdc_gene_exp_adj.Rda")
+popViewport() ## exit
+
+## vp.3 - Figure C - BRCA somatic mutations
+pushViewport(vp.3)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
+load("../data/gdc_gene_exp_adj.Rda") ## load data
 load("../data/gene_data.Rda")
+
 brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
@@
 ## Axis labels
-legend("top", inset = c(0), legend=names(C_colors[c(1:2)]), col=C_colors[c(1:2)], pch = 20, bty = "n", horiz = T, cex = 1.2)
-legend("top", inset = c(0.10), legend=names(C_colors[c(3)]), col=C_colors[3], pch = 20, bty = "n", horiz = T, cex = 1.2)
-title(xlab="UMAP 2", line=0)
-mtext("C", side=3, line=-2.05, cex=1.2, adj=0)
+legend("top", inset = c(0), legend=names(C_colors[c(1:2)]), col=C_colors[c(1:2)], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.07), legend=names(C_colors[c(3)]), col=C_colors[3], pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("C", side=3, line=-1.2, cex=1.2, adj=0)
 
-### Figure D - Stage
+popViewport()
+
+## vp.4 - Figure D - Stage
+pushViewport(vp.4)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
 D_colors <- c('darkgoldenrod3','dodgerblue3','yellow3','tomato2') ## Color scheme
@@
 ## Axis labels
-legend("top", inset = c(0), legend=names(D_colors[1:2]), col=D_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-legend("top", inset = c(0.10), legend=names(D_colors[3:4]), col=D_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
-title(xlab="UMAP 2", line=0)
-mtext("D", side=3, line=-2.05, cex=1.2, adj=0)
+legend("top", inset = c(0), legend=names(D_colors[1:2]), col=D_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.07), legend=names(D_colors[3:4]), col=D_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("D", side=3, line=-1.2, cex=1.2, adj=0)
 
-### Figure E - Metastasis
+popViewport()
+
+## vp.5 - Figure E - Metastasis
+pushViewport(vp.5)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
 E_colors <- c('#a6cee3','#1f78b4') ## Color scheme
@@
 ## Axis labels
-legend("top", inset = c(0), legend=names(E_colors), col=E_colors, pch = 20, bty = "n", horiz = T, cex = 1.2)
-title(xlab="UMAP 2", line=0)
-mtext("E", side=3, line=-2.05, cex=1.2, adj=0)
+legend("top", inset = c(0), legend=names(E_colors), col=E_colors, pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("E", side=3, line=-1.2, cex=1.2, adj=0)
 
-plot.new()
+popViewport()
 ```

### 2020-08-24 code/figure2.Rmd

@@
 ```{r figure-2-plot, fig.width = 9}
-load("../data/figure1B_data.rda") ## Load in data processed in figure1.Rmd
-load("../data/optimal_minpts_score.rda")
-umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
 load("../data/kelly.colours.rda")
-load("../data/ditto.colours.rda")
+cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
+names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
 
-opar <- par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
-par(mar = c(4,6,2,0) + 0.1) ## setting plot margins for the following plots
-## Figure 2A: Explained variance of eigenvalues
-with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = expl_var / 100, ## plotting the explained variance of observed
-                                                 type = "l", yaxt = "n", bty = "l",
-                                                 col = "#998ec3", 
-                                                 xlab = "", ylab = ""))
-with(randomize %>% filter(eigenvalue < 50), lines(x = eigenvalue, y = perm / 100, ## plotting the explained variance of permuted
-                                                  col = "#f1a340"))
-yticks_val <- scales::breaks_pretty(n=5)(randomize %>% filter(eigenvalue < 50) %>% pull(expl_var) / 100)
-axis(2, at=yticks_val, lab=scales::percent(yticks_val), las = 1) ## Axis labels
-title(ylab = "explained variance", line = 3.5)
-title(xlab = "MCA Eigenvalues", line = 2)
-legend("topright", legend=c("Eigenvalues", "Permuted Eigenvalues"),
-       col=c("#998ec3", "#f1a340"), lty=1, cex=0.8)
-mtext("A", side=3, line=0, cex=1.2, adj=-0.35) ## plot tag
+theta <- c(0, 90, 270)
+phi <- c(65, 15, 15)
 
+opar <- par(mfrow = c(1,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
 par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
-## Figure 2C: MCA projected down to 3 dimensions (UMAP)
+
+for(n in 1) {
   with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "white", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
                                   xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                   colvar = NULL))
-title(xlab="UMAP 2", line=0)
-mtext("C", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
-
-par(mar = c(4,6,2,0) + 0.1) ## setting plot margins for the following plots
-## Figure 2D: Optimization of HDBscan
-with(sample_df, plot(x = minPts, y = value, 
-                     col = "grey", pch = 20, yaxt = "n",
-                     xlab = "", ylab = "", bty = "l",
-                     las = 1, cex = 0.3))
-with(score_df, lines(x = minPts, y = score,
-                     col = "black", las = 1))
-abline(v = optimal_minPts, col = "red") ## Red vertical line to show optimal minimum point for HDBscan
-title(ylab = "score", line = 3.5)
-title(xlab = "Minimum Cluster Size", line = 2)
-axis(2, at=c(0.25, 0.35, 0.45, 0.55), las = 1)
-mtext("D", side=3, line=0, cex=1.2, adj=-0.27) ## plot tag
-
-## Figure 2B: pvalue of eigenvalue 
-with(randomize %>% filter(eigenvalue < 50), plot(x = eigenvalue, y = pvalue,  ## line plot
-                                                 type = "l", bty = "l",
-                                                 col = "black", las = 1,
-                                                 xlab = "", ylab = ""))
-with(randomize %>% filter(eigenvalue < 50), points(x = eigenvalue, y = pvalue, pch = 20, ## add dots to line plot
-                                                   col = "black"))
-
-## Determining the optimal mca dimension
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>% ## filtering for pvalue < 0.05
-  filter(eigenvalue == max(eigenvalue)) %>% ## filtering for the maximum eigenvalue
-  pull(eigenvalue)
-
-abline(v = optimal_mca_dimensions, col = "red") ## Red vertical line to show optimal MCA dimension
-title(main = paste0("Significant dimensions = ", optimal_mca_dimensions), adj = 0, font = 3, cex.main = .8) ## Axis labels
-title(ylab = "P-value", line = 3.5)
-title(xlab = "MCA Eigenvalues", line = 2)
-mtext("B", side=3, line=0, cex=1.2, adj=-0.35) ## plot tag
-
-par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
-## Figure 2E: HDBscan
-cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
-names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
+  title(xlab="UMAP 2", line=-5.2)
+  for(p in 2:length(phi)) {
     with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = cluster.colours[as.character(clust)], pch = 21, cex = 0.8, lwd = 0.2, # clust is hdbscan result
-                                theta = 0, phi = 65, scale = F,
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, 
+                                theta = theta[p], phi = phi[p], scale = T,
                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                 colvar = NULL))
-title(xlab="UMAP 2", line=0)
-mtext("E", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
-
-## Figure 2F: kNN 
-with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
-                                theta = 0, phi = 65, scale = F,
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-mtext("F", side=3, line=-2.05, cex=1.2, adj=0) ## plot tag
+  }
+}
 

### 2020-08-24 text/survival.Rmd
change some text to better describe the bayesian model, and change colors to avoid colorblindness problems in the kmcurves

@@
 distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
-selected group with correct age distribution for the given cancer (blue) to the
-model predicted survival for the given average k (red) and the actual survival
-for the patients in our dataset (CHOOSE COLOR HERE, NOT GREEN). As can be seen,
-Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all
-very deadly. The modeled results fit the general trend for the actual patients.
-Thyroid Cancer (THCA) on the other hand has very little if any effect on life
-expectancy, in fact there may be a slight benefit which may be associated with a
-bias in diagnosis towards patients who are more health conscious, or higher
-socioeconomic status, with more access to care compared to random population
-members.
+selected group with correct age distribution for the given cancer (turqoise) to
+the model predicted survival for the given average k (blue) and the actual
+survival for the patients in our dataset (orange). As can
+be seen, Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD)
+are all very deadly. The modeled results fit the general trend for the actual
+patients.  Thyroid Cancer (THCA) on the other hand has very little if any effect
+on our estimates of life expectancy, in fact there may be a slight benefit which
+may be associated with a bias in diagnosis towards patients who are more health
+conscious, or higher socioeconomic status, with more access to care compared to
+random population members.
 
@@
 (Thyroid: THCA). Cancers with posterior probability for relative risk of less
-than 1 such as THCA may be those where diagnosis is associated with patients who
-have relatively healthy behaviors, are seen regularly by doctors, and comply
-with health recommendations.
+than 1 such as THCA should be interpreted carefully. This Bayesian model is a
+model for a state of information. The information that a person is diagnosed
+with cancer may lead us to expect that they will live a shorter time than the
+general population of matched age (k > 1), or a longer time than the general
+population of matched age (k < 1). One mechanism for a shorter result is that
+the cancer is aggressive and we can expect it to rapidly injure the body,
+causing death. One explanation for longer results may be that the cancer is
+relatively mild, and therefore diagnosis is potentially an informational signal
+that the patient is health conscious, with the comparison group having more
+people whose cancers go undiagnosed. It's important to note that the diagnosis
+can increase our expectation of life relative to the comparison group, even if
+it decreases the expectation of life of the individual relative to the
+counterfactual where they did not have cancer.
 

### 2020-08-24 code/figureSM2.Rmd
Restore code that generates a comparison between CDC and model fit for survival data

```
{r SM2}

load("../code/survival/survmodel.stansave")

sampsum = summary(samps)
sampsum

asamps = as.array(samps)

r20f = mean(asamps[,,"r20[1]"])
r20m = mean(asamps[,,"r20[2]"])
agrf = mean(asamps[,,"agerate[1]"])
agrm = mean(asamps[,,"agerate[2]"])

#sprintf("r20f mean = %f",r20f)
#sprintf("r20m mean = %f",r20m)
#sprintf("agrf mean = %f", agrf)
#sprintf("agrm mean = %f", agrm)

ourlccdf = function(t,age,agerate,k,r20){
    r20 = r20 * 1e-5
    agerate = agerate*.05
    return(-r20*((365*exp((agerate*k*t)/365+age*agerate-20*agerate))/(agerate*k)
        -(365*exp(age*agerate-20*agerate))/(agerate*k)));
}

ccdfsimpmkr = function(r20,agert){
    function(age){
        exp(ourlccdf(age*365,0,agert,1,r20))
    }
}

mccdf = ccdfsimpmkr(r20m,agrm)
fccdf = ccdfsimpmkr(r20f,agrf)

mortpl <- ggplot(menmort) + 
    geom_line(aes(yrnum,nsurv/100000),col="blue",data=menmort) +
    geom_line(aes(yrnum,nsurv/100000),col="red",data=wommort)+
    stat_function(fun=function(x) mccdf(x),col="green")+
    stat_function(fun=function(x) fccdf(x),col="orange")+labs(title="1-cdf(age) for Men and Women (CDC Data vs  model fit)",x="Age (yrs)",y="1-cdf") +
    theme_minimal()

print(mortpl)

```

### 2020-08-24 code/figureSM2.Rmd
load the mortality data to generate the supplemental figure

@@
-```
-{r SM2}
+```{r SM2}
+
+## mortality data from CDC:
+namesmort = c("agerng","pdead","nsurv","ndie","pyrs","pyrsabove","expectedlife")
+menmort = read_csv("../data/survival/lifetables/Table02.csv",skip=2)
+## drop the final row, it's just a text info statement
+menmort = menmort[1:NROW(menmort)-1,] 
+wommort = read_csv("../data/survival/lifetables/Table03.csv",skip=2)
+wommort = wommort[1:NROW(wommort)-1,]
+
+names(menmort) = namesmort
+names(wommort) = namesmort
+yrs=seq(0,100,by=1);
+menmort$yrnum = yrs
+wommort$yrnum = yrs
 

### 2020-08-24 text/manuscript.Rmd
update fig 3 & 5 legends, typos, survival text

@@
 **Figure 1: Molecular pathway profiles of tumor samples with one or more
-mutations.** Each of 377 selected Reactome pathways (columns) is classified as
-disrupted if one or more genes is mutated in the tumor sample (rows) where red
-and grey represents the pathway is mutated and not, respectively. Tumors are
+mutations.** Each of 377 selected Reactome pathways (rows) is classified as
+disrupted if one or more genes is mutated in the tumor sample (columns) where
+red and grey represents the pathway is mutated and not, respectively. Tumors are
 grouped by tissue of origin using standardized abbreviations from the TCGA
@@
 
+#### Figure 2
 ```{r, child='../code/figure2.Rmd'}
 ```
+
+**Figure 2: Clustering of tumor samples** Different rotational perspectives of
+the same MCA-based UMAP projection in three-dimensional space. Each dot
+corresponds to a tumor sample. The same colors indicate the tumor's cluster
+identity throughout this manuscript. 1=yellow, 2=lilac, 3=orange, 4=light blue,
+5=crimson, 6=beige, 7=grey, 8=dark green, 9=pink, 10=dark blue.
 
@@
 understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types (with some notable
-exceptions) were not heavily biased in one class, and all well-represented
-cancer types had tumors in every class (see **figure 3A** and full tumor
-profiles in **supplementary figure S1**, see also interactive media from
-**supplementary file 1 (html)**), suggesting that, in principle at least, our
-pathway-disruptions identify clusters of molecular pathology largely independent
-of tissue-of-origin. As an example of one type of cancer that does have a biased
-pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
-class 8 (**figure 3A** and **supplementary figure S1**). But even PAAD comprises
-tumors from the nine remaining classes, meaning that patients suffering these
-tumors have potentially different underlying molecular pathologies.
+predominant classes. To our surprise, most cancer types were not heavily biased
+in one class, and all well-represented cancer types had tumors in every class
+(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
+also interactive media from **supplementary file 1 (html)**), suggesting that,
+in principle at least, our pathway-disruptions identify clusters of molecular
+pathology largely independent of tissue-of-origin. As an example of one type of
+cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
+was predominantly found in class 8 (**figure 3A** and **supplementary figure
+S1**). But even PAAD comprises tumors from the nine remaining classes, meaning
+that patients suffering these tumors have potentially different underlying
+molecular pathologies.
 
@@
 **Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
-a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
+A) Sankey plot showing correspondence between cancer type and cluster identity.
+B) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
 COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
-Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the
-UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of
-tumor stage onto the UMAP, regardless of cancer type. e) Projection of
+Pancreatic adenocarcinoma. C) Projection of breast cancer subtypes onto the
+UMAP. D) Projection of BRCA1/2 somatic mutation onto the UMAP. E) Projection of
+tumor stage onto the UMAP, regardless of cancer type. E) Projection of
 metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
@@
 have common pathway disruption profiles, and the UMAP which features a series of
-lobe-like structures on a common backbone of tumor samples (**figure 2C**) could
+lobe-like structures on a common backbone of tumor samples (**figure 2**) could
 in principle also reflect progression through a series of stages. In support of
@@
 **Figure 5: Pan-cancer enrichment of pathway disruptions.** 
-a) Waterfall plot; columns represent tumors. Each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. The top 100 most mutated pathways are
-shown. Red signifies pathway disrupted pathways, grey signifies not disrupted.
-b) Heatmap shows the relative enrichment of each pathway (columns) within each
-numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
-purple, depletion). c) Proportion of mutated samples in the each significant
-pathway (columns; union set of pathways with effect size >= 0.30 in each
+A) Heatmap shows the relative enrichment of each pathway (rows) within each
+numbered cluster (columns). Effect size is displayed as colors representing
+percent enrichment. B) Proportion of mutated samples in the each significant
+pathway (columns; union set of pathways with effect size ≥ 0.30 in each
 cluster) within each numbered cluster (rows).
@@
 
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-```{r child='../text/fig_legend.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
 ## References

### 2020-08-24 text/survival.Rmd
update fig 3 & 5 legends, typos, survival text

@@
 apart from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. These analyses are profoundly limited by
-confounding factors of age and stage at diagnosis, sex, ethnicity, and
-tissue-specific disease progression, to name the obvious ones. To explore these
-ideas, we used Bayesian inference to test models of survival using public
-longevity data from the CDC, accounting for age, gender and ethnicity. Our model
-estimates the effect of cancer type and class-specific cancer effects
-independently, resulting in a cancer and class specific estimate of the aging
-rate multiplier.
+types or across all cancers. These analyses are limited by confounding factors
+of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
+progression. To explore these ideas, we used Bayesian inference to test models
+of survival using public longevity data from the CDC, accounting for age, gender
+and ethnicity. Our model estimates the effect of cancer type and
+cluster-specific cancer effects independently, resulting in a cancer and
+cluster-specific estimate of the aging rate multiplier, k.
 
@@
 have *different* survival curves. The model takes this into account and the
-aggressiveness of the cancer can be estimated without the confounding aspect of
-age.
+aggressiveness of the cancer can be estimated accurately without the confounding
+effect of age.
 
@@
 **Figure 7: Comparison of different cancers, simulated and actual results
-results.** A) To understand the risk model, we plot survival of four simulated
-cohorts. Compare survival of a group consisting of randomly selected 30 year old
+results.** To illustrate the risk model, we plot survival of four simulated
+cohorts. A) Compare survival of a group consisting of randomly selected 30 year old
 women (light blue) to an equal sized group of 30 year old women with cancer, k =
-2 (light orange). To understand how the mix of ages alters the survival curve,
+2 (light orange). To illustrate how the mix of ages alters the survival curve,
 we compare a randomly selected group of people with ages uniformly distributed
@@
 the model predicted survival for the given average k (blue) and the actual
-survival for the patients in our dataset (orange). As can
-be seen, Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD)
-are all very deadly. The modeled results fit the general trend for the actual
-patients.  Thyroid Cancer (THCA) on the other hand has very little if any effect
-on our estimates of life expectancy, in fact there may be a slight benefit which
-may be associated with a bias in diagnosis towards patients who are more health
+survival for the patients in our dataset (orange). As can be seen, Stomach
+Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
+deadly. The modeled results fit the general trend for the actual patients.
+Thyroid Cancer (THCA) on the other hand has very little if any effect on our
+estimates of life expectancy, in fact there may be a slight benefit which may be
+associated with a bias in diagnosis towards patients who are more health
 conscious, or higher socioeconomic status, with more access to care compared to

### 2020-08-24 text/supplemental_methods.Rmd
Restore code that generates a comparison between CDC and model fit for survival data

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE, message = FALSE)
 library(foreach)
@@
 ### Survival
+```{r child='../code/figureSM2.Rmd'}
+```
 

### 2020-08-24 code/figure2.Rmd

@@
-```{r figure-2-plot, fig.width = 9}
+```{r figure-2-plot, fig.width = 9, echo = FALSE}
 umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
@@
 opar <- par(mfrow = c(1,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
-par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
+par(mar = c(0,2,0,0), bg = NA) ## setting plot margins for the following plots
 
@@
                                 colvar = NULL))
+        if(p == 2) {
+      mtext("Cluster", side=3, line=-3.5, cex=1.2, adj=0.5)
+      legend("top", inset = c(0.10), legend=names(cluster.colours[2:6]), col=cluster.colours[2:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
+      legend("top", inset = c(0.15), legend=names(cluster.colours[7:11]), col=cluster.colours[7:11], pch = 20, bty = "n", horiz = T, cex = 1.2)
+    }
   }

### 2020-08-24 text/manuscript.Rmd

--- 2020-08-24 text/manuscript.Rmd
@@
 
+#### Figure 2
 ```{r, child='../code/figure2.Rmd'}
 ```
+
+**Figure 2: Dimensionality reduction of tumor samples using pathway aberrations data 
+identifies ten well-defined clusters.** Different perspectives of the same data in 
+three-dimensional space are shown using UMAP. Each dot corresponds to a tumor sample. 
+Colors indicate the tumor's cluster identity.
 
@@
 **Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
-a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
-COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
-Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the
-UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of
-tumor stage onto the UMAP, regardless of cancer type. e) Projection of
+a) Sankey diagram (flow is left to right) of the tumor type make up in a cluster. Cluster 
+membership is on the left. Cancer type is on the right. b) Projection of breast cancer
+subtypes onto the UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) 
+Projection of tumor stage onto the UMAP, regardless of cancer type. e) Projection of
 metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
@@
 
-In order to visualize pathway disruption enrichment across all cancers
-(pan-cancer), we created waterfall plots, replacing genes with pathways
-(**figure 5A**). A complete list of pathway disruptions with percent mutated
-samples and top genes is provided in **supplementary table ST1**. As expected,
+In order to identify pathway disruption enrichment across all cancers
+(pan-cancer), we created a list of pathway disruptions with percent mutated
+samples and top genes (**supplementary table ST1**). As expected,
 these analyses reveal the broad importance of many well known pathways that are
@@
 **Figure 5: Pan-cancer enrichment of pathway disruptions.** 
-a) Waterfall plot; columns represent tumors. Each row represents a pathway. The
-pathways are ranked such that the pathway with the most disruptions across all
-7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
-top pathway disruptions successively. The top 100 most mutated pathways are
-shown. Red signifies pathway disrupted pathways, grey signifies not disrupted.
-b) Heatmap shows the relative enrichment of each pathway (columns) within each
+a) Heatmap shows the relative enrichment of each pathway (columns) within each
 numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
-purple, depletion). c) Proportion of mutated samples in the each significant
+purple, depletion). b) Proportion of mutated samples in the each significant
 pathway (columns; union set of pathways with effect size >= 0.30 in each
@@
 
-Wish wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
+We wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
 reading of this manuscript, and Ivetth Corona for early discussions and
@@
 
-```{r child='../text/fig_legend.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
 ## References

### 2020-08-24 code/figure1v2.Rmd
minor edits

@@
+---
+output:
+  word_document: default
+  html_document: default
+---
 ```{r, fig.width = 9, fig.height= 4}

### 2020-08-24 text/manuscript.Rmd
minor edits

--- 2020-08-24 text/manuscript.Rmd
@@
 
-**Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
+**Figure 3: Pathway-based clustering independent of tissue-of-origin** 
 A) Sankey plot showing correspondence between cancer type and cluster identity.

### 2020-08-24 text/supplemental_methods.Rmd
moved survival figure to where the figure legend is in the supplement. Updated text

--- 2020-08-24 text/supplemental_methods.Rmd
@@
 
-### Survival
-
 In order to understand how the diagnosis of cancer changes the expectation for
@@
 
-**supplementary figure SM2** gives a comparison on log-scale between
-the post-fit survival curves and the CDC data it is based on.
+### Survival
+```{r child='../code/figureSM2.Rmd'}
+```
+
+**supplementary figure SM2** gives a comparison between the post-fit
+survival curves for men and women vs the CDC data it is based on. This
+comparison does not distinguish on ethnicity for simplicity in
+visualization.
 

### 2020-08-24 text/survival.Rmd
replace all instances of 'class' with 'cluster' in survival texts; move SM2 to after first reference.

@@
 ---
-If our molecular pathway disruption classes represent distinct biological states
+If our molecular pathway disruption clusters represent distinct biological states
 apart from tissue of origin, they may have different prognoses within cancer
@@
 **Figure 7: Comparison of different cancers, simulated and actual results
-results.** To illustrate the risk model, we plot survival of four simulated
-cohorts. A) Compare survival of a group consisting of randomly selected 30 year old
-women (light blue) to an equal sized group of 30 year old women with cancer, k =
+results.** To illustrate the risk model, we compare survival of four simulated
+cohorts. A) A group consisting of randomly selected 30 year old
+women (light blue) vs. an equal sized group of 30 year old women with cancer, k =
 2 (light orange). To illustrate how the mix of ages alters the survival curve,
@@
 
-The degree to which a given cancer accelerates aging can be determined by the
-$k_{tis}$ multiplier (**Figure 8**) multiplied by the class specific modifier
+The degree to which a given cancer accelerates mortality can be determined by
+the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier
 (**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
@@
 
-**Figure 9 Class specific $k_{class}$ values shown for each cancer type.** The
-overall $k$ value is the product of $k_{tis} \times k_{class}$. The class
-specific value shows the relative aggressiveness of that particular class within
-the tissue. For example class 1, 2, and 6 Glioblastoma (GBM) are apparently much
-less deadly than other types of GBM. A similar situation can be seen with
-Pancreatic Cancer (PAAD) where 1, 6, and 8 are less deadly compared to classes
-7, 9, 10. For some cancer types there are few data points and little ability to
-estimate values with precision (DLBC and KICH for example).
+**Figure 9 Cluster specific $k_{cl}$ values for each cancer type.** The overall
+$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
+represents the relative aggressiveness of each cluster within the cancer type.
+For example clusters 1, 2, and 6 Glioblastoma (GBM) are apparently much less
+deadly than other types of GBM. A similar situation can be seen with Pancreatic
+Cancer (PAAD) where 1, 6, and 8 are less deadly compared to clusters 7, 9, 10.
+For some cancer types there are few data points and little ability to estimate
+values with precision (DLBC and KICH for example).
 
@@
 
-Our estimates of tissue-specific class effects, in contrast, ranged from less
-than 1 up to about 3 or 4, reflecting that some classes are either less deadly
-or more deadly than other classes within each cancer types (**Figure 9**). A
-class specific rate of 1 represents typical rate for this tissue type. For
+Our estimates of tissue-specific cluster effects, in contrast, ranged from less
+than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
+or more deadly than other clusters within each cancer type (**Figure 9**). A
+cluster-specific rate of 1 represents typical rate for this tissue type. For
 several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
@@
 mortalities in the data to make an estimate (as expected for PRAD and THCA) or
-two few samples, period. We did not observe class-specific trends that held true
+two few samples, period. We did not observe cluster-specific trends that held true
 across cancer types, which could result from different cancers having different
 standards of care for example. In support of this interpretation, we also tested
-a factored model which considered both cancer type and each class independent of
+a factored model which considered both cancer type and each cluster independent of
 cancer type (not shown). Though we were able to successfully fit this model, it
-is a special case of the more general model where class and tissue can
-independently affect longevity, and there is no reason to believe that class
+is a special case of the more general model where cluster and tissue can
+independently affect longevity, and there is no reason to believe that cluster
 specific effects would necessarily remain constant across tissue types given how

### 2020-08-24 text/supplemental_methods.Rmd
replace all instances of 'class' with 'cluster' in survival texts; move SM2 to after first reference.
update more class-->cluster and revert mortality to aging in survival text

--- 2020-08-24 text/supplemental_methods.Rmd
@@
 cluster that capture the maximum number of tumor samples, measured by
-probability of membership of ≥ 5% in at least one class. Thus, we created a
+probability of membership of ≥ 5% in at least one cluster. Thus, we created a
 score metric as the fraction of classified tumors with max probability < 5% in
-one class and chose a cluster size of 92 to minimize the score function
+one cluster and chose a cluster size of 92 to minimize the score function
 (**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
@@
 lets the _k_ most similar tumors "vote" as to the identity of the query tumor
-sample based on their classification labels. We set _k_ to be the square root of
+sample based on their cluster labels. We set _k_ to be the square root of
 the number of tumor samples (87). Using this method, we assigned cluster
@@
 ### Survival
-```{r child='../code/figureSM2.Rmd'}
-```
 
@@
 the background risk rates, and the $k$ value, which is a function of both cancer
-tissue type, and the cancer genomic class.
+tissue type, and the cancer cluster identity.
 
@@
 Black patient tumors, we use ethnicity only to identify the baseline risk
-function, while the tissue and class specific multipliers are not ethnicity
-specific.
+function, while the tissue- and cluster-specific multipliers are not ethnicity
+specific. There were too few tumors from other races to obtain good inference so
+we excluded them from our analysis.
 
@@
 the post-fit survival curves and the CDC data it is based on.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+```{r child='../code/figureSM2.Rmd'}
+```
+
+**SM2 Legend here**
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 

### 2020-08-25 code/figureSM2.Rmd
Fix figureSM2 code chunk

--- 2020-08-24 code/figureSM2.Rmd
@@
-```
-{r SM2}
+```{r SM2}
+## mortality data from CDC:
+namesmort = c("agerng","pdead","nsurv","ndie","pyrs","pyrsabove","expectedlife")
+menmort = read_csv("../data/survival/lifetables/Table02.csv",skip=2)
+## drop the final row, it's just a text info statement
+menmort = menmort[1:NROW(menmort)-1,] 
+wommort = read_csv("../data/survival/lifetables/Table03.csv",skip=2)
+wommort = wommort[1:NROW(wommort)-1,]
+
+names(menmort) = namesmort
+names(wommort) = namesmort
+yrs=seq(0,100,by=1);
+menmort$yrnum = yrs
+wommort$yrnum = yrs
 
@@
 sampsum = summary(samps)
-sampsum
 
@@
 mortpl <- ggplot(menmort) + 
-    geom_line(aes(yrnum,nsurv/100000),col="blue",data=menmort) +
-    geom_line(aes(yrnum,nsurv/100000),col="red",data=wommort)+
-    stat_function(fun=function(x) mccdf(x),col="green")+
-    stat_function(fun=function(x) fccdf(x),col="orange")+labs(title="1-cdf(age) for Men and Women (CDC Data vs  model fit)",x="Age (yrs)",y="1-cdf") +
+    geom_line(aes(yrnum,nsurv/100000),col="#0072B2",data=menmort) +
+    geom_line(aes(yrnum,nsurv/100000),col="#D55E00",data=wommort)+
+    stat_function(fun=function(x) mccdf(x),col="#009E73")+
+    stat_function(fun=function(x) fccdf(x),col="#E69F00")+labs(title="1-cdf(age) for Men and Women (CDC Data vs  model fit)",x="Age (yrs)",y="1-cdf") +
     theme_minimal()
@@
 print(mortpl)
-
 ```

### 2020-08-25 code/figure9.Rmd
Update axis title to figure 8&9, add legend for SM2

@@
 		facet_wrap(. ~ tiss,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
-		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients")
+		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients") +
+	  labs(x = expression(paste("k"["cl"])), y = "Cluster Identity")
 }

### 2020-08-25 text/abstract.Rmd

@@
 molecular signatures remain largely indistinguishable from tissue-of-origin,
-thus obscuring opportunities for novel mechanistic insights and potential
-treatment vulnerabilities that hold true across different cancer types. To
+presenting a major barrier for precision health and individualized medicine. To
 address this challenge, we reduced mutation data to disruptions in a select set
@@
 cellular phenotypes. Moreover, driver gene mutations involved in replication
-fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally distributed
-among our clusters, contrary to the expectation that such genes are avatars of
-molecular subtype. We also investigated pathways that distinguish between
-metastatic and non-metastatic tumors, and found that most meaningful differences
-are cluster-specific. Some metastasis enriched pathways from one cluster are
-cluster-specific pathways from non-metastatic tumors of another cluster,
-suggesting phenotypic convergence, and potentially highlighting key pathways to
-focus on for prevention of cluster-specific relapse. Taken as a whole, our
-observations suggest that common driver genes combine with tissue-specific
-disruptions in complementary sets of tumor-promoting pathways to produce a
-limited number of distinct molecular phenotypes. Eventually these signatures
-converge on one or more end-stage disease phenotypes. Thus, we present a
-coherent view of global tumor biology, and explain how common cellular
-dysfunction might arise from tissue-specific mutations.
+fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally
+distributed among our clusters, contrary to the expectation that such genes are
+avatars of molecular subtype. We also investigated pathways that distinguish
+between metastatic and non-metastatic tumors, and found that most meaningful
+differences are cluster-specific. Some metastasis enriched pathways from one
+cluster are cluster-specific pathways from non-metastatic tumors of another
+cluster, suggesting phenotypic convergence, and potentially highlighting key
+pathways to focus on for prevention of cluster-specific relapse. Taken as a
+whole, our observations suggest that common driver genes combine with
+tissue-specific disruptions in tumor-promoting pathways to produce a limited
+number of distinct molecular phenotypes. Eventually these signatures converge on
+one or more end-stage disease phenotypes. Thus, we present a coherent view of
+global tumor biology, and explain how common cellular dysfunction might arise
+from tissue-specific mutations.

### 2020-08-25 text/survival.Rmd

@@
 
-The degree to which a given cancer accelerates mortality can be determined by
+The degree to which a given cancer accelerates aging can be determined by
 the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier

### 2020-08-25 code/supplemental_v2.Rmd
Minor update to supplemental
Remove st2 from supplemental pdf

@@
 
-**Supplementary Table 1. Cluster-specific enriched pathways (effect size >= 0.30).**
-```{r enrich-pathway-in-clust, fig.width= 9}
+**Supplementary Table 2. Cluster-specific enriched pathways (effect size >= 0.30).**
+```{r enrich-pathway-in-clust, fig.width= 9, eval = FALSE}
 ## docx friendly
@@
 
-st1 <- moderate_enrich_prob %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
+table_st2 <- moderate_enrich_prob %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
   dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
@@
 ```
-
-\newpage
-
-![**Supplemental figure S6.
-Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively.
-The orange and green curves show the model fits._]

### 2020-08-25 text/manuscript.Rmd
Remove st2 from supplemental pdf

@@
 We investigated this dataset further using multiple correspondence analysis
-(MCA), and visually summarized the analysis with UMAP (**figure 2**). We used
+(MCA) [@JSSv025i01], and visually summarized the analysis with UMAP (**figure 2**). We used
 the resulting UMAP graph coordinates to perform density based clustering with
@@
 A) Sankey plot showing correspondence between cancer type and cluster identity.
-B) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
-COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
-Pancreatic adenocarcinoma. C) Projection of breast cancer subtypes onto the
-UMAP. D) Projection of BRCA1/2 somatic mutation onto the UMAP. E) Projection of
-tumor stage onto the UMAP, regardless of cancer type. E) Projection of
-metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
-metastatic tumors.
+B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
+somatic mutation onto the UMAP. D) Projection of tumor stage onto the UMAP,
+regardless of cancer type. E) Projection of metastatic status onto the UMAP.
+Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.
 
@@
 (**figure 5B**). Using this approach, we identified 25 pathways highly enriched
-(enrichment score $> 0.3$, see methods) in cluster 2, 9 pathways enriched in
+(enrichment score > 0.3, see methods) in cluster 2, 9 pathways enriched in
 cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10 pathways in

### 2020-08-25 text/manuscript.Rmd

@@
 We investigated this dataset further using multiple correspondence analysis
-(MCA) [@JSSv025i01], and visually summarized the analysis with UMAP (**figure 2**). We used
-the resulting UMAP graph coordinates to perform density based clustering with
-HDBSCAN, which resulted in identification of 10 well-defined clusters capturing
-about 80% of the tumor samples. To capture the remaining samples into one of
-these 10 clusters we used kNN (see **Supplementary Methods** for details on all
-clustering algorithms).
+(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
+2**) [@McInnes2018]. We used the resulting UMAP graph coordinates to perform
+density based clustering with HDBSCAN [@hdbscan], which resulted in
+identification of 10 well-defined clusters capturing about 80% of the tumor
+samples. To capture the remaining samples into one of these 10 clusters we used
+kNN (see **Supplementary Methods** for details on all clustering algorithms).
 

### 2020-08-25 text/supplemental_methods.Rmd

@@
 remained ≤ 0.05 (see cutoff in **figure SM1B**). Then we performed a UMAP
-analysis, both in order to summarize the MCA graphically, and as a preprocessing
-step to boost the performance of density based clustering. The resulting map was
-notable for its lobed structure, with several reproducible projections
-regardless of random seed setting. A representative version of this 3D UMAP is
-shown in **figure SM1C**, rotated to enhance the visibility of the major features.
-Following this spatial mapping we attempted to define groupings of similar
-tumors within the spatial map using HDBSCAN, which performs hierarchical
-clustering and provides metrics of cluster stability and probabilities of
-cluster membership for each node. However, HDBSCAN is sensitive to several
-parameters; key for our analysis are the minimum number of tumor samples in a
-cluster that capture the maximum number of tumor samples, measured by
-probability of membership of ≥ 5% in at least one cluster. Thus, we created a
-score metric as the fraction of classified tumors with max probability < 5% in
-one cluster and chose a cluster size of 92 to minimize the score function
-(**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
+analysis [@McInnes2018], both in order to summarize the MCA graphically, and as
+a preprocessing step to boost the performance of density based clustering. The
+resulting map was notable for its lobed structure, with several reproducible
+projections regardless of random seed setting. A representative version of this
+3D UMAP is shown in **figure SM1C**, rotated to enhance the visibility of the
+major features. Following this spatial mapping we attempted to define groupings
+of similar tumors within the spatial map using HDBSCAN, which performs
+hierarchical clustering and provides metrics of cluster stability and
+probabilities of cluster membership for each node [@hdbscan]. However, HDBSCAN
+is sensitive to several parameters; key for our analysis are the minimum number
+of tumor samples in a cluster that capture the maximum number of tumor samples,
+measured by probability of membership of ≥ 5% in at least one cluster. Thus, we
+created a score metric as the fraction of classified tumors with max probability
+< 5% in one cluster and chose a cluster size of 92 to minimize the score
+function (**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
 high-density clusters which we then projected onto the UMAP (**figure SM1E**).
@@
 lets the _k_ most similar tumors "vote" as to the identity of the query tumor
-sample based on their cluster labels. We set _k_ to be the square root of
-the number of tumor samples (87). Using this method, we assigned cluster
-membership to the remaining tumors (**figure SM1F**).
+sample based on their cluster labels. We set _k_ to be the square root of the
+number of tumor samples (87). Using this method, we assigned cluster membership
+to the remaining tumors (**figure SM1F**).
 

### 2020-08-25 code/supplemental_v2.Rmd
Add code to save tables in single excel file

@@
   library(knitr),
-  library(kableExtra)
+  library(kableExtra),
+  library(xlsx)
 ))
@@
 #   fontsize(size = 8, part = "body")
-
-write.xlsx(table_st1, file = "../data/table_st1.xlsx", row.names = FALSE)
 ```
 
-**Supplementary Table 2. Cluster-specific enriched pathways (effect size >= 0.30).**
-```{r enrich-pathway-in-clust, fig.width= 9, eval = FALSE}
+```{r st2, fig.width= 9, eval = FALSE}
+##**Supplementary Table 2. Cluster-specific enriched pathways (effect size >= 0.30).**
 ## docx friendly
@@
 
-st1$log10pvalue <- round(-log10((st1$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-st1$Enrichment_Score <- round(st1$Enrichment_Score, digits = 2)
-st1$Signif <- NULL ## we don't need this anymore
+table_st2$log10pvalue <- round(-log10((table_st2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
+table_st2$Enrichment_Score <- round(table_st2$Enrichment_Score, digits = 2)
+table_st2$Signif <- NULL ## we don't need this anymore
 
@@
 #
-# knitr::knit_print(st1)
-names(st1) <- c("Pathway", "Cluster", "Effect Size", "-log(p)")
- st1 %>%  knitr::kable(format = "latex", longtable = TRUE, align = c("l", rep("c",3))) %>%
-   kable_styling(latex_options=c("scale_down", "repeat_header"), full_width = FALSE, font_size = 12) %>%
-   column_spec(1, color = "black", width = "4in")
+
+# names(st1) <- c("Pathway", "Cluster", "Effect Size", "-log(p)")
+#  st1 %>%  knitr::kable(format = "latex", longtable = TRUE, align = c("l", rep("c",3))) %>%
+#    kable_styling(latex_options=c("scale_down", "repeat_header"), full_width = FALSE, font_size = 12) %>%
+#    column_spec(1, color = "black", width = "4in")
+
+## Saving st1 and st2 in excel
+
+write.xlsx(table_st1, file="supplemental.table.xlsx", sheetName="ST1", row.names=FALSE)
+write.xlsx(table_st2, file="supplemental.table.xlsx", sheetName="ST2", append=TRUE, row.names=FALSE)
 ```

### 2020-08-25 code/figureSM2.Rmd

@@
 
+mytheme = theme(
+#    axis.text.x = element_blank(),
+#    axis.text.y = element_blank(),
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+#    axis.ticks = element_blank(),
+#    axis.line = element_blank(),
+    strip.text.y.right = element_text(angle = 0),
+    strip.background = element_blank(),
+    legend.key.height = unit(0.4, "lines"),
+    legend.title = element_blank(),
+    #legend.position = "none",
+    title = element_text(size=10)
+  )
+
 mortpl <- ggplot(menmort) + 
-    geom_line(aes(yrnum,nsurv/100000),col="#0072B2",data=menmort) +
-    geom_line(aes(yrnum,nsurv/100000),col="#D55E00",data=wommort)+
-    stat_function(fun=function(x) mccdf(x),col="#009E73")+
-    stat_function(fun=function(x) fccdf(x),col="#E69F00")+labs(title="1-cdf(age) for Men and Women (CDC Data vs  model fit)",x="Age (yrs)",y="1-cdf") +
-    theme_minimal()
+    geom_line(aes(yrnum,nsurv/100000, color = "CDC men"),data=menmort) +
+    geom_line(aes(yrnum,nsurv/100000, color ="CDC women"),data=wommort) +
+    stat_function(fun=function(x) mccdf(x), aes(color = "Model fit men")) +
+    stat_function(fun=function(x) fccdf(x), aes(color = "Model fit women")) + 
+    labs(title="1-cdf(age) for Men and Women (CDC Data vs model fit)",x="Age (yrs)",y="1-cdf", color = "") +
+    theme_classic() +
+    theme(axis.title.x = element_text(size = 10),
+          axis.title.y = element_text(size = 10),
+          title = element_text(size = 10),
+          legend.position = c(0.85, 0.9),
+          legend.text = element_text(size = 8),
+          legend.key.size = unit(0.4,"cm")) +
+    scale_color_manual(values = c("CDC men" = "#0072B2", "CDC women" = "#D55E00", 
+                                  "Model fit men" = "#009E73", "Model fit women" = "#E69F00"))
 

### 2020-08-25 code/survival/survfig.Rmd

@@
 	dfktis$cancname = as.factor(sapply(dfktis$parname,function(x) tisnames[x]))
-	ggplot(dfktis,aes_(quote(value),quote(cancname))) + geom_density_ridges()+ theme_minimal() + scale_fill_brewer(palette="Paired") + xlim(0,22) + labs(x="ktis multiplier",y="Cancer Type",title="Tissue Specific Effects")
+	ggplot(dfktis,aes_(quote(value),quote(cancname))) + geom_density_ridges()+ theme_minimal() + scale_fill_brewer(palette="Paired") + xlim(0,22) + labs(x=expression(paste("k"["tis"], " multiplier")),y="Cancer Type",title="Tissue Specific Effects")
 }

### 2020-08-25 text/supplemental_methods.Rmd
Fix placement of SM2 in supp meths

@@
 
-### Survival
-```{r child='../code/figureSM2.Rmd'}
-```
-
-**supplementary figure SM2** gives a comparison between the post-fit
+**Supplementary figure SM2** gives a comparison between the post-fit
 survival curves for men and women vs the CDC data it is based on. This
@@
 
+#### Figure SM2
+
 ```{r child='../code/figureSM2.Rmd'}
@@
 
-**SM2 Legend here**
+**Supplementary Figure SM2** Comparison between CDC data and ODE model fit for
+males and females of European descent.
 

### 2020-08-25 text/abstract.Rmd
fix passive voice in abstract

@@
 samples into ten clusters distinct from their tissues-of-origin and largely free
-of bias from mutational burden or clinical stage. We showed that these tumor
+of bias from mutational burden or clinical stage. We show that these tumor
 classes vary in prognosis by modeling relative risk of death in different
@@
 distributed among our clusters, contrary to the expectation that such genes are
-avatars of molecular subtype. We also investigated pathways that distinguish
-between metastatic and non-metastatic tumors, and found that most meaningful
+avatars of molecular subtype. We also investigate pathways that distinguish
+between metastatic and non-metastatic tumors, and show that most meaningful
 differences are cluster-specific. Some metastasis enriched pathways from one

### 2020-08-25 text/introduction.Rmd
edit for clarity

@@
 expression profiles, DNA methylation, and copy number variations from the same
-tissue samples. To take advantage of these tools several large consortia,
+tissue samples. To take advantage of these tools, several large consortia,
 including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
-genomes (PCAWG), sequenced large numbers of tumors and collected data from
+genomes (PCAWG) sequenced large numbers of tumors and collected data from
 multiple assays that could be analyzed together, with the goal of integration
@@
 [@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
-mutations in cancer driver genes are commonplace in somatic tissues and
+mutations in cancer driver genes are commonplace in healthy tissue and
 correlate with age and environmental exposures [@mutations-healthy-tissues].
@@
 approaches address the sparseness of mutations, allowing genes to be influenced
-by the aggregate of mutations in the nearest network neighbors
+by mutations in the nearest network neighbors
 [@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
@@
 @mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
-highlighting some potentially fundamental disease phenotypes at the pathway
-level that cross cancers with different origins at the cellular and tissue
+highlighting fundamental disease phenotypes at the pathway
+level across cancers with different origins at the cellular and tissue
 level.

### 2020-08-25 text/manuscript.Rmd
edit for clarity

@@
 samples. To capture the remaining samples into one of these 10 clusters we used
-kNN (see **Supplementary Methods** for details on all clustering algorithms).
+kNN (see **Supplementary Methods** for details on clustering methods).
 
@@
 (see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
-also interactive media from **supplementary file 1 (html)**), suggesting that,
+also interactive media from **supplementary file 1**), suggesting that,
 in principle at least, our pathway-disruptions identify clusters of molecular
@@
 molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based classes, therefore providing support for the classes as
+with our pathway-based clusters, therefore providing support for the clusters as
 proxies for molecular pathology sub-typing. To our surprise, we find a similar
@@
 categories of tumor, we did not observe any bias among the stages to specific
-pathway disruption classes (**figure 3D**), suggesting that stage is not an
-determining factor in pathway disruption classes.
+pathway disruption clusters (**figure 3D**), suggesting that stage is not a
+contributing factor to cluster identity.
 
@@
 **Figure 4: Somatic mutation frequencies for tumors in each class** 
-Column uses a logarithmic axis to represent mutation count. Row represents the
-tumor samples. Each dot corresponds to the sample mutation count for tumors in
-each class, ordered by mutation counts. The median mutation count in each
-cluster is indicated by the horizontal line.
+Vertical axis shows log mutation count, horizontal axis is cluster identity.
+Each dot represents an individual tumor sample, ranked lowest to highest by
+mutation count. The median mutation count in each cluster is indicated by the
+horizontal line.
 
@@
 
+To discover what pathways are most important for clustering, we calculated
+percent enrichment for each pathway _within cluster_ relative to all other
+clusters combined (see methods) and ranked pathways for each cluster from
+highest to lowest enrichment. We visualized the enrichment as a heatmap
+(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
+(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
+enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
+pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
+pathways in cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and
+9 had several pathways in common. To explore the specific pathways marking each
+cluster, we projected disruptions for each of the 377 pathways onto the UMAP
+(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
+pathways including RNA and protein biosynthesis (**supplementary figure S4**).
+Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
+and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
+and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
+extracellular, intracellular, and immune-related signaling pathways (see
+**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
+pathway enrichment levels of the three, having mutations in hedgehog signaling,
+"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
+cycle" and "apoptosis" among others.
+
 ```{r, results='asis'}
@@
 
-To discover what pathways are most important for clustering, we calculated
-percent enrichment for each pathway _within cluster_ relative to all other
-clusters combined (see methods) and ranked pathways for each cluster from
-highest to lowest enrichment. We visualized the enrichment as a heatmap
-(**figure 5B**). Using this approach, we identified 25 pathways highly enriched
-(enrichment score > 0.3, see methods) in cluster 2, 9 pathways enriched in
-cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10 pathways in
-cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10 pathways in
-cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and 9 had
-several pathways in common. To explore the specific pathways marking each
-cluster, we projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
-pathways including RNA and protein biosynthesis (**supplementary figure S4**).
-Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
-and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
-and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
-extracellular, intracellular, and immune-related signaling pathways (see
-**figure 5C** and **supplementary figure S4**). Cluster 2 had the highest
-pathway enrichment levels of the three, having mutations in hedgehog signaling,
-"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
-cycle" and "apoptosis" among others.
-
 Prior efforts to extract common signatures from pan-cancer datasets met with
@@
 also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
-enriched for both _PIK3CA_ and _KRAS_. This result suggests that tumors in this
-cluster could be derivative of either cluster 7 or 8. Clusters 3 and 5, defined
-by enrichment in metabolic pathways, had mutations in ribosomal proteins and
-nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible
-for nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
+enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
+in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
+complex, respectively. Cluster 4 had mutations in genes responsible for
+nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
 involved in protein degradation. We also observed that genes that were enriched
-for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster
-6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
+for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster 6,
+but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
 clusters 3 and 8). Next, we investigated the proportion of samples per cancer
@@
 Each cluster was compared against the background (all other clusters) to find
-deferentially mutated genes. Significant genes (pvalue < 0.01) were selected and
-limited to the top ten results for each clusters. Heatmap; Column represents the
-cancer type. Row represents the significant genes as described above. The
-heatmap shows the proportion of samples per cluster, per cancer type that is
-mutated for each of the significant genes. Depleted significant genes (left).
-Enriched significant genes (right).
+differentially mutated genes. Significant genes (pvalue < 0.01) were selected
+and limited to the top ten results for each clusters. Heatmap; Column represents
+the cancer type. Row represents the significant genes as described above. The
+heatmap shows the proportion of samples per cluster and cancer type that are
+mutated for each gene. Depleted significant genes (left). Enriched significant
+genes (right).
 
@@
 Following the same logic we used to investigate cluster-specific enrichment of
-pathways (**figure 5B**) we compared metastatic _vs._ non-metastatic
-tumors, as we find that metastatic tumors are distributed across all ten
-clusters ( _e.g._ **figure 3E**). Using all non-metastatic tumors as background,
-we found very low levels of enrichment (≤ 10%) in a handful of pathways. We
-reasoned that the individual clusters might be too different to detect global
-metastasis enrichment signal given the small sample size (n = 215 metastatic
-tumor samples).
+pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
+metastatic tumors are distributed across all ten clusters ( _e.g._ **figure
+3E**). Using all non-metastatic tumors as background, we found very low levels
+of enrichment (< 10%) in a handful of pathways. We reasoned that the individual
+clusters might be too different to detect global metastasis enrichment signal
+given the small sample size (n = 215 metastatic tumor samples).
 
 Therefore, we calculated cluster-specific enrichment in metastatic tumors and
-found a total of 31 enriched pathways across all clusters (**table 1**). A
-number of enrichments were found in multiple clusters represented by pathways
-that were already shown to be enriched in a neighboring cluster. For example,
-“Signaling by _PTK6_” is enriched in non-metastatic samples of cluster 8 (see
-**supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
-metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$, **table 1**). This is also
-true of “Erythropoietin activates _RAS_”, which is enriched in non-metastatic
-tumors of cluster 8 (**supplementary figure S4**) and also in metastatic tumors
-of clusters 7 and 9. Cluster 4 metastatic samples were enriched for “Fc epsilon
-receptor (FCERI) signaling”, a key neutrophil pathway, which is also specific to
-clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions specific to
-metastases tend toward complementary mutations that are otherwise specific to
+found a total of 31 enriched pathways (significant with enrichment score ≥ 0.3)
+across all clusters (**table 1**). A number of enrichments were found in
+multiple clusters represented by pathways that were already shown to be enriched
+in a neighboring cluster. For example, “Signaling by _PTK6_” is enriched in
+non-metastatic samples of cluster 8 (see **supplementary figure S4**) but not in
+7 and 9. This pathway is enriched in metastatic tumors of clusters 7 and 9 ($p <
+10^{-3}$, **table 1**). This is also true of “Erythropoietin activates _RAS_”,
+which is enriched in non-metastatic tumors of cluster 8 (**supplementary figure
+S4**) and also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic
+samples were enriched for “Fc epsilon receptor (FCERI) signaling”, a key
+neutrophil pathway, which is also specific to clusters 2, 7 and 8 non-metastatic
+tumors. Thus, metastatic-enriched pathways from one cluster are often enriched in 
 non-metastatic tumors of other clusters.

### 2020-08-25 text/survival.Rmd
edit for clarity

@@
 ---
-If our molecular pathway disruption clusters represent distinct biological states
-apart from tissue of origin, they may have different prognoses within cancer
+If our molecular pathway disruption clusters represent biological states
+distinct from tissue of origin, they may have different prognoses within cancer
 types or across all cancers. These analyses are limited by confounding factors
@@
 progression. To explore these ideas, we used Bayesian inference to test models
-of survival using public longevity data from the CDC, accounting for age, gender
-and ethnicity. Our model estimates the effect of cancer type and
-cluster-specific cancer effects independently, resulting in a cancer and
-cluster-specific estimate of the aging rate multiplier, k.
+(see Methods and **Supplemental Methods** for greater detail) of survival using
+public longevity data from the CDC, accounting for age, gender and ethnicity.
+Our model estimates the effect of cancer type and cluster-specific cancer
+effects independently, resulting in a cancer and cluster-specific estimate of
+the aging rate multiplier, $k$.
 
@@
 The degree to which a given cancer accelerates aging can be determined by
-the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier
+the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier $k_{cl}$
 (**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
@@
 in relative risk of age-adjusted death, the remaining cancer types ranged from
-about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
-close to the age predicted purely from their demographics independent of the
-information that they have the given cancer.
+about 1 to 5 in magnitude.
 
@@
 showing the probability density for the posterior value of the $k_{tis}$
-multiplier. These values reflect the general deadliness of the cancer type
-across all clusters. GBM, PAAD, and SKCM stand out as having values in the 10 to
-20 range with a wide degree of uncertainty. Other cancers such as Prostate
-(PRAD) and Thyroid (THCA) show values down near or even below 1 suggesting that
-these populations may have healthier lifestyles and more access to care. Most
-cancer types fall in the range 2-7 as might be expected.
+multiplier. Larger values correspond to decreased age-independent life
+expectancy.
 
@@
 mortalities in the data to make an estimate (as expected for PRAD and THCA) or
-two few samples, period. We did not observe cluster-specific trends that held true
-across cancer types, which could result from different cancers having different
-standards of care for example. In support of this interpretation, we also tested
-a factored model which considered both cancer type and each cluster independent of
-cancer type (not shown). Though we were able to successfully fit this model, it
-is a special case of the more general model where cluster and tissue can
-independently affect longevity, and there is no reason to believe that cluster
-specific effects would necessarily remain constant across tissue types given how
-widely the patients vary across tissue types as well as the different drug and
-surgical treatments typically used for each tissue.
+two few samples, period. We did not observe cluster-specific trends that held
+true across cancer types, which could result from different cancers having
+different standards of care for example. In support of this interpretation, we
+also tested a factored model which considered both cancer type and each cluster
+independent of cancer type (not shown). Though we were able to successfully fit
+this model, it is a special case of the more general model where cluster and
+tissue independently affect longevity, and there is no reason to believe that
+cluster specific effects would necessarily remain constant across tissue types
+given how widely the patients vary across tissue types, drug and surgical
+treatments.

### 2020-08-25 text/manuscript.Rmd
added url for the CDC data

@@
 
-We used Bayesian inference to explore the relative impact of cancer diagnosis on
-survival at the time of diagnosis. Our model assumes that at birth each person
-has a small initial probability per unit time to die, and that this probability
-grows exponentially in time at some baseline rate that matches the observed CDC
-data. The rate of growth in risk is assumed to jump to a higher level at
-diagnosis, which we take as a proxy for onset. This represents an accelerated
-aging type model. To fit the model, we constrain the baseline risks using the
-available CDC life tables. We then constrain the cancer specific multiplier
-using the longevity data within the cancer tumor dataset. Our model estimates
-the effect of cancer type and class-specific cancer effects independently,
-resulting in a cancer and class specific estimate of the aging rate multiplier.
-A detailed description of the CDC longevity model and likelihood is provided in
-the **Supplementary methods**.
+We used Bayesian inference to explore the relative impact of cancer
+diagnosis on survival at the time of diagnosis. Our model assumes that
+at birth each person has a small initial probability per unit time to
+die, and that this probability grows exponentially in time at some
+baseline rate that matches the observed CDC data. The rate of growth
+in risk is assumed to jump to a higher level at diagnosis, which we
+take as a proxy for onset. This represents an accelerated aging type
+model. To fit the model, we constrain the baseline risks using the
+available CDC life tables
+(ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We
+also constrain the cancer specific multiplier using the longevity data
+within the cancer tumor dataset. Our model estimates the effect of
+cancer type and class-specific cancer effects independently, resulting
+in a cancer and class specific estimate of the aging rate multiplier.
+A detailed description of the CDC longevity model and likelihood is
+provided in the **Supplementary methods**.
 

### 2020-08-25 text/discussion.Rmd
Minor text edits

@@
 genes with tissue-specific pathway disruptions result in common categories when
-viewed at the level of the pathway knowledge-base. Top cancer driver genes (e.g.
-PIK3C and TP53) are found in most of the clusters, in spite of the fact that
+viewed at the level of the pathway knowledge-base. Top cancer driver genes (_e.g.
+PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the fact that
 they contribute to many cluster-specific pathways. This can only be explained as
@@
 and are frequently the result of aberrations involving common driver genes such
-as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
+as BRCA1/2, MMR genes, mitotic checkpoints, cohesion complexes, _etc._
 Cluster-specific

### 2020-08-25 text/introduction.Rmd
Minor text edits

@@
 tissue-of-origin. In a second study involving 33 cancer types and a much greater
-number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify
+number of tumors, the authors identify
 28 clusters that could be further subdivided into organ specific groups,

### 2020-08-25 text/manuscript.Rmd

@@
 don't observe an obvious trend in the overall mutation burden across the
-backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
+backbone of the UMAP. Nonetheless, to test the hypothesis that
 the molecular-pathway disruption clusters represent advancement through stages,
@@
 pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
-pathways in cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and
+pathways in cluster 9, and 0 pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
 9 had several pathways in common. To explore the specific pathways marking each

### 2020-08-25 text/supplemental_methods.Rmd
add page break after SM1 in supp methods

@@
 **Figure SM1: Method for tumor classification using pathway aberration data** 
-a) MCA analysis was performed to determine the eigenvalue with the most
+A) MCA analysis was performed to determine the eigenvalue with the most
 explanatory power from the categorical data. Average of 100 permutations of the
-data was used as a baseline. b) The p-value of each eigenvalue was plotted and
-the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
+data was used as a baseline. B) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
 projected down to 3 dimensions for the purpose of density based clustering (only
-the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
+the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
 used to perform hierarchical density based clustering tuned to minimize the
-fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
+fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
 the UMAP projection; tumors are colored by cluster membership (see legend). This
-results in some tumors remaining unclassified (grey points). f) kNN was used to
+results in some tumors remaining unclassified (grey points). F) kNN was used to
 assign membership to the remaining tumors.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 
@@
 
-Using this model, we split the data by Male/Female sex, and by White, _vs._
+Using this model, we split the data by Male/Female sex, and by White _vs._
 Black race as these were the main well-identified categories available to us in

### 2020-08-25 code/supplemental_file_shiny/app.R
Add shiny app.R

@@ -1,0 +1,225 @@
+## library
+suppressPackageStartupMessages(c(
+    library(shiny),
+    library(tidyverse),
+    library(SummarizedExperiment),
+    library(plot3D),
+    library(data.table),
+    library(rgl),
+    library(htmltools),
+    library(dplyr),
+    library(rgl),
+    library(htmltools),
+    library(car),
+    library(grid),
+    library(gridBase)
+))
+
+## Load data
+umap_cancertype <- fread("/home/p3nguyen/TTmanu/data/umap_3d_coors.tsv")
+load("/home/p3nguyen/TTmanu/data/kelly.colours.rda")
+load("/home/p3nguyen/TTmanu/data/ditto.colours.rda")
+load("/home/p3nguyen/TTmanu/data/tumor_samples.Rda")
+load("/home/p3nguyen/TTmanu/data/supplemental_gene.Rda")
+load("/home/p3nguyen/TTmanu/data/gdc_reactome_path_analysis.Rda")
+
+cluster.colours <- c(kelly.colours[c(3:12)], "grey") ## Color scheme
+names(cluster.colours) <- c(1:11) ## Assign cluster to color
+
+## Generating selection inputs
+cancer_type <- tumor_samples %>% .$project_code %>% unique %>% ## get cancer type info
+    .[order(.)] ## Order alphabetically
+
+genename <- colnames(supplemental_gene) ## get gene names
+genename <- genename[order(genename)] ## order alphabetically
+
+pathwayname <- rownames(gdc_reactome_path_analysis) ## get pathway names
+pathwayname <- pathwayname[order(pathwayname)] ## order alphabetically
+
+## User interface setup
+ui <- (fluidPage(
+    ## App title
+    titlePanel("Interactive Supplemental File"),
+    ## Interactive settings
+    playwidgetOutput("player"),
+    
+    ## Slider inputs
+    sidebarPanel(
+        selectInput("cancer_type", label = "Cancer type", ## Cancer type
+                    choices = cancer_type, selected = "BRCA"),
+        selectInput("genename", label = "Gene:", ## Gene 
+                    choices = genename, selected = "TP53"),
+        selectInput("pathwayname", label = "Pathway", ## Pathway
+                    choices = pathwayname, selected = "Cellular response to hypoxia")
+        ),
+    
+    mainPanel(
+        tabsetPanel(type = "tabs",
[…]

### 2020-08-26 code/supplemental_file_shiny/app.R
Adds depenency to shinyWidgets to get select all/ select none

@@ -1,6 +1,7 @@
 ## library
 suppressPackageStartupMessages(c(
     library(shiny),
+    library(shinyWidgets),
     library(tidyverse),
     library(SummarizedExperiment),
     library(plot3D),
@@ -50,8 +51,9 @@ ui <- (fluidPage(
                     tabPanel(title = "Pan-cancer", rglwidgetOutput("pancan",  width = 500, height = 500)),
                     tabPanel(title = "Cancer Type",
                              
-                             checkboxGroupInput("cancer_type", label = "Cancer type:", ## Cancer type
-                                         choices = cancer_type, selected = "BRCA", inline = TRUE),
+                             pickerInput("cancer_type", label = "Cancer type:", ## Cancer type
+                                         choices = cancer_type, selected = "BRCA", multiple = TRUE,
+                                         options = list(`actions-box` = TRUE)),
                              plotOutput("cancertype")),
                     tabPanel(title = "Gene", 
                              selectInput("genename", label = "Gene:", ## Gene 

### 2020-08-26 text/manuscript.Rmd
Add supplemental file text ref and hyperlink

@@
 (MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
-2**) [@McInnes2018]. We used the resulting UMAP graph coordinates to perform
+2** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
+We used the resulting UMAP graph coordinates to perform
 density based clustering with HDBSCAN [@hdbscan], which resulted in
@@
 was predominantly found in class 8 (**figure 3A** and **supplementary figure
-S1**). But even PAAD comprises tumors from the nine remaining classes, meaning
+S1** and **supplementary file 1**). But even PAAD comprises tumors from the nine remaining classes, meaning
 that patients suffering these tumors have potentially different underlying
@@
 different prognoses and standards of care. We did not observe any exclusive
-segregation by pathway for these subtype annotations (**Figure 3B** and
-**supplemental file 1**). We also projected histological subtype data for the
-remaining cancers (see **supplementary figure S2** for the full set of
-projections); we find that the subtypes, though often biased towards one or more
-classes, are almost never exclusive. We interpret these data in aggregate to
-mean that our pathway disruption classes do not correspond to previously
-identified molecular subtypes within the parent cancer type.
+segregation by pathway for these subtype annotations (**Figure 3B**). We also 
+projected histological subtype data for the remaining cancers (see 
+**supplementary figure S2** for the full set of projections and see interactive
+media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the subtypes, though often 
+biased towards one or more classes, are almost never exclusive. We interpret 
+these data in aggregate to mean that our pathway disruption classes do not 
+correspond to previously identified molecular subtypes within the parent cancer type.
 
@@
 mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 3C**). We made
-similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_,
-_MLH3_, _PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these
-genes except for _APC_ exhibited any remarkable specificity with respect to
-cluster assignment (**supplementary figure S3**). To look at other risk factors
+segregation of these mutations into specific clusters (**figure 3C** and see 
+interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar projections 
+for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ 
+and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except 
+for _APC_ exhibited any remarkable specificity with respect to cluster assignment 
+(**supplementary figure S3**). To look at other risk factors
 such as maintenance of DNA methylation levels and telomere length, we projected
@@
 cluster, we projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
-pathways including RNA and protein biosynthesis (**supplementary figure S4**).
-Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
+(**supplementary figure S4** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). 
+Clusters 3 and 5 were distinguished by metabolic pathways including RNA and 
+protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was 
+distinguished by mutations affecting regulation of DNA
 and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
@@
 _PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
-_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
+_KRAS_ panels of **Supplementary figure S5** and see interactive media from 
+[**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly, cluster 9, which
 also shared multiple enrichment in signaling pathways with clusters 7 and 8, is

### 2020-08-26 text/abstract.Rmd

@@
 Cancer is a complex disease involving dysfunctional cellular metabolism, other
-basic biochemical processes, and the microenvironment. However, in spite of some
+basic biochemical processes, and the microenvironment. However, despite some
 generally agreed upon unifying principles [@weinberg2000; @hallmarks_II],

### 2020-08-26 text/abstract.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

@@
-Cancer is a complex disease involving dysfunctional cellular metabolism, other
-basic biochemical processes, and the microenvironment. However, despite some
-generally agreed upon unifying principles [@weinberg2000; @hallmarks_II],
-molecular signatures remain largely indistinguishable from tissue-of-origin,
-presenting a major barrier for precision health and individualized medicine. To
-address this challenge, we reduced mutation data to disruptions in a select set
-of pathways relevant to basic cell biology, from DNA replication to cellular
-communication. Using dimensionality reduction techniques, we assigned tumor
-samples into ten clusters distinct from their tissues-of-origin and largely free
-of bias from mutational burden or clinical stage. We show that these tumor
-classes vary in prognosis by modeling relative risk of death in different
-classes and cancer types. We identify cluster-specific mutations in different
-tissues, demonstrating that tissue-specific signatures contribute to common
-cellular phenotypes. Moreover, driver gene mutations involved in replication
-fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally
-distributed among our clusters, contrary to the expectation that such genes are
-avatars of molecular subtype. We also investigate pathways that distinguish
-between metastatic and non-metastatic tumors, and show that most meaningful
-differences are cluster-specific. Some metastasis enriched pathways from one
-cluster are cluster-specific pathways from non-metastatic tumors of another
-cluster, suggesting phenotypic convergence, and potentially highlighting key
-pathways to focus on for prevention of cluster-specific relapse. Taken as a
-whole, our observations suggest that common driver genes combine with
-tissue-specific disruptions in tumor-promoting pathways to produce a limited
-number of distinct molecular phenotypes. Eventually these signatures converge on
-one or more end-stage disease phenotypes. Thus, we present a coherent view of
-global tumor biology, and explain how common cellular dysfunction might arise
-from tissue-specific mutations.
+Cancer is a complex disease involving disrupted cellular metabolism, basic
+biochemical processes, and the microenvironment. However, despite some generally
+agreed upon unifying principles [@weinberg2000; @hallmarks_II], molecular
+signatures remain largely indistinguishable from tissue-of-origin, presenting a
+major barrier for precision health and individualized medicine. To address this
+challenge, we reduce mutation data to disruptions in a select set of pathways
+relevant to basic cell biology, from DNA replication to cellular communication.
+Using dimensionality reduction techniques, we assign tumor samples into ten
+clusters distinct from tissue-of-origin and largely free of bias from mutational
+burden or clinical stage. We show that the clusters vary in prognosis by
+modeling relative risk of death by cancer type and cluster. We identify
+cluster-specific mutations in different tissues, demonstrating that
+tissue-specific signatures contribute to common cellular phenotypes. Moreover,
+germline risk genes involved in replication fidelity and genome stability are
+equally distributed among clusters, contrary to the expectation that such genes
+are avatars of molecular subtype. We investigate metastatic and non-metastatic
+pathways, and show that most differences are cluster-specific. Some metastatic
+pathways from one cluster are cluster-specific pathways from non-metastatic
+tumors of another cluster, suggesting phenotypic convergence. Taken as a whole,
+our observations suggest that common driver genes combine with tissue-specific
+disruptions in tumor-promoting pathways to produce a limited number of distinct
+molecular phenotypes. Thus, we present a coherent view of global tumor biology,
+and explain how common cellular dysfunction might arise from tissue-specific
+mutations.

### 2020-08-26 text/discussion.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

@@
 protein-coding genes is comparable with the number of tumor samples, even with
-multiple mutations per sample. And that is not accounting for the fact that
+multiple mutations per sample. This also does not account for the fact that
 ~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
@@
 methods [@PATHNET] in that we chose to focus on the pathway (defined in the
-methods) as the unit of disruption instead of the gene, employing the conceit
-that individual mutations may be sufficient to disrupt or misdirect pathway
-activity (as is often the case with driver mutations). Rather than evaluating
-the burden of mutations on each pathway within or across tumors, which can be
-biased by the number of genes or the coverage of pathway-linked genes in the
-genome, we scored each pathway in binary fashion, allowing one mutation or
-multiple mutations to count equally for each sample.
+methods) as the unit of disruption instead of the gene, where individual
+mutations may be sufficient to disrupt or misdirect pathway activity (as is
+often the case with driver mutations). Rather than evaluating the burden of
+mutations on each pathway within or across tumors, which can be biased by the
+number of genes or the coverage of pathway-linked genes in the genome, we scored
+each pathway in binary fashion, allowing one mutation or multiple mutations to
+count equally for each sample.
 
@@
 cancer-promoting mutations likely occur at such low frequencies as to be
-indistinguishable from background. Previous attempts to capture these
-associations using Bayesian networks have yielded important new insights into
-this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that would not
-have been possible with traditional analysis, supporting the case that
-pathway-centric approaches have the potential to increase statistical power.
+indistinguishable from background.
 
@@
 clinical care for diagnostic markers remains elusive. Doubtless some of this is
-due to bad luck, as tumors can metastasize and remain dormant years before they
-are detected at distal sites, or resistant clones may have already arisen at
-undetectable levels [@hallmarks_II], but our analysis suggests the possibility
-of identifying more informative molecular, histological or cellular subtypes
-that could form a basis for future separation of patients into different
-precision treatment regimens.
+due to chance events, as tumors can metastasize and remain dormant years before
+they are detected at distal sites, or resistant clones may have already arisen
+at undetectable levels [@hallmarks_II], but our analysis suggests the
+possibility of identifying more informative molecular, histological or cellular
+subtypes that could form a basis for future stratification of patients into
+different precision treatment regimens.
 

### 2020-08-26 text/introduction.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

@@
 
-In this study, we attempt to address these questions by focusing on discrete
-molecular pathways as the unit of disruption in cancer. Our approach strips all
-cell-type-specific information from the mutation data and equates gene-level
-mutations to cell-biological pathway disruptions. We use this heuristic to
-evaluate all cancers and show that, surprisingly, tumors that exhibit
-tissue-specific gene mutation patterns nonetheless fall into common categories
-of pathway disruption having unique prognoses in each cancer type.
+In this study, we attempt to understand how tissue-specific gene disruptions
+create common cancer phenotypes by focusing on discrete molecular pathways as
+the unit of disruption. Our approach strips all cell-type-specific information
+from the mutation data and equates gene-level mutations to cell-biological
+pathway disruptions. We use this heuristic to evaluate all cancers and show
+that, surprisingly, tumors that exhibit tissue-specific gene mutation patterns
+nonetheless fall into common categories of pathway disruption having unique
+prognoses in each cancer type.

### 2020-08-26 text/manuscript.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

--- 2020-08-25 text/manuscript.Rmd
@@
 title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and
-  Dennis J. Hazelett$^{1,3}$
+author: Peter T. Nguyen$^{1,\dag}$, Simon G. Coetzee$^{1,\dag}$, Daniel L. Lakeland$^{2}$, and
+  Dennis J. Hazelett$^{1,3,*}$
 bibliography: references.bib
@@
 Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
+$\dag$ These authors contributed equally.
+$*$ To whom correspondence should be addressed.
 
@@
 and biochemical pathways, excluding gene sets that correspond to miscellaneous
-categories ( _e.g._ “transcription factors”) or disease associations ( _e.g._
+categories (\it{e.g.} “transcription factors”) or disease associations (\it{e.g.}
 “mutated in colon cancer”) and filtered our gene list on membership in these
@@
 
-To avoid bias toward larger pathways ( _i.e._ pathways with more member genes),
+To avoid bias toward larger pathways (\it{i.e.} pathways with more member genes),
 we counted pathways as disrupted if one or more member genes were mutated. We do
@@
 binarized by pathway are likely to be noisy for several reasons. First, point
-mutations can be deleterious (attenuating or hypomorphic) or activating
-(neomorphic or hypermorphic) in genes, and these can in turn act functionally as
-either activators or repressors. For this study we assume a significant fraction
-of these mutations are generically disruptive to normal pathway activity if they
-are activating or deactivating or affect a gene that one might classify as a
-tumor suppressor or oncogene. Second, we know that low-expressed genes and
-non-expressed genes accumulate mutations at a higher rate due to transcription
-coupled repair [@kandothMutationalLandscapeSignificance2013;
+mutations can be deleterious (attenuating, hypomorphic or antimorphic) or
+activating (neomorphic or hypermorphic) in genes, and these can in turn act
+functionally as either oncogenes or tumor suppressors. For this study we assume
+a significant fraction of these mutations are generically disruptive to normal
+pathway activity since it is impossible to know the tumor promoting effects of
+all mutations, including rarely studied genes. Second, we know that
+low-expressed genes and non-expressed genes accumulate mutations at a higher
+rate due to transcription coupled repair
+[@kandothMutationalLandscapeSignificance2013;
 @kimPancancerAnalysisSomatic2018;
@@
 metastasis. Given these assumptions, it is possible that more advanced tumors
-have common pathway disruption profiles, and the UMAP which features a series of
+have common pathway disruption profiles. The UMAP, which features a series of
 lobe-like structures on a common backbone of tumor samples (**figure 2**) could
-in principle also reflect progression through a series of stages. In support of
-this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
-the fewest point mutations and culminates in a cluster (class 10) which has
-nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
-don't observe an obvious trend in the overall mutation burden across the
-backbone of the UMAP. Nonetheless, to test the hypothesis that
-the molecular-pathway disruption clusters represent advancement through stages,
-we projected staging data onto the UMAP. Similar to tissue of origin and other
-categories of tumor, we did not observe any bias among the stages to specific
-pathway disruption clusters (**figure 3D**), suggesting that stage is not a
-contributing factor to cluster identity.
+in principle reflect progression through a series of stages. The backbone starts
+with a cluster of tumors (class 1) that has the fewest point mutations and
+culminates in a cluster (class 10) which has nearly every pathway disrupted
+(**figure 4**). However, outside of class 10 we don't observe an obvious trend
+in the overall mutation burden across the backbone of the UMAP. Nonetheless, to
+test the hypothesis that the molecular-pathway disruption clusters represent
+advancement through stages, we projected staging data onto the UMAP. Similar to
+tissue of origin and other categories of tumor, we did not observe any bias
+among the stages to specific pathway disruption clusters (**figure 3D**),
+suggesting that stage is not a contributing factor to cluster identity.
 
@@
 highest to lowest enrichment. We visualized the enrichment as a heatmap
-(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
-(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
-enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
-pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
-pathways in cluster 9, and 0 pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
+(**figure 5A**). Using this approach, we identified twenty-five pathways highly
+enriched (enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2,
+nine pathways enriched in cluster 3, twenty-four pathways in cluster 4, ten
+pathways in cluster 5, ten pathways in cluster 6, twenty-four pathways in
+cluster 7, twenty-four pathways in cluster 8, ten pathways in cluster 9, and no
+pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
 9 had several pathways in common. To explore the specific pathways marking each
@@
 the question, are tumor phenotypes driven entirely by common driver genes, or by
-“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
+“silent” tissue-specific effectors (\it{i.e.} too few samples to detect above
 statistical significance thresholds), or a combination of both? To answer this
@@
 involved in protein degradation. We also observed that genes that were enriched
-for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster 6,
+for one cluster are depleted from others (\it{i.e.} is enriched in cluster 6,
 but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
@@
 rate for these genes within samples that belong to a cluster, the mutation rate
-is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
+is heterogeneous across tumors by tissue origin (\it{e.g.} In cluster 4, CESC was
 enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
@@
 pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
-metastatic tumors are distributed across all ten clusters ( _e.g._ **figure
+metastatic tumors are distributed across all ten clusters (\it{e.g.} **figure
 3E**). Using all non-metastatic tumors as background, we found very low levels
@@
 criteria to define basic cellular processes and biochemical pathways: (1)
-human-derived pathways ( _“HSA”_) (2) limited to grandchild node for each parent
-pathway ( _e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
+human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
+parent pathway (\it{e.g.} ‘Beta-catenin independent WNT signaling’ in ‘Signal
 Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
@@
 excluded pathways have been shown to play an important role in cancer, they are
-highly specialized ( _e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
-most of the excluded pathways, a neutral version pathway of the pathway exists (
-_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577 Ensembl
-IDs from the TCGA dataset to the highly selected Reactome pathways. This
+highly specialized (\it{e.g.} "PI3K/AKT Signaling in Cancer"). Additionally, for
+most of the excluded pathways, a neutral version pathway of the pathway exists
+(\it{e.g.} "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
+Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
 operation produced a lookup table that consisted of 377 pathways mapped to 8,940
@@
 We filtered likely erroneous mutations due to transcription coupled repair. Our
-approach was to determine the status of each gene ( _i.e._ expressed or not
+approach was to determine the status of each gene (\it{i.e.} expressed or not
 expressed) in each tissue type in order to exclude gene mutations low expressed
@@
 
-We wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
-reading of this manuscript, and Ivetth Corona for early discussions and
-feedback.
+We wish to thank David Van Valen, Kate Lawrenson, Simon Knott, and Megan
+Hitchens for critical reading of this manuscript, and Ivetth Corona for early
+discussions and feedback.
 

### 2020-08-26 text/survival.Rmd
incorporate Megan Hitchens' comments, fix spaces after parens

@@
 or more deadly than other clusters within each cancer type (**Figure 9**). A
-cluster-specific rate of 1 represents typical rate for this tissue type. For
-several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
+cluster-specific rate of 1 represents the typical rate for this tissue type. For
+several cancers (\it{e.g.} PRAD, kidney chromophobe (KICH), diffuse large B-cell
 lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely

### 2020-08-26 text/discussion.Rmd
fix italics *again*

@@
 This could be addressed with application of ever-increasingly sophisticated
-filtering of likely passenger mutations [ _e.g._ @spliceAI;
+filtering of likely passenger mutations [_e.g._ @spliceAI;
 @deep-learning-clinical-variants]. In the future, we hope to incorporate these

### 2020-08-26 text/manuscript.Rmd
fix italics *again*

@@
 and biochemical pathways, excluding gene sets that correspond to miscellaneous
-categories (\it{e.g.} “transcription factors”) or disease associations (\it{e.g.}
+categories (_e.g._ “transcription factors”) or disease associations (_e.g._
 “mutated in colon cancer”) and filtered our gene list on membership in these
@@
 
-To avoid bias toward larger pathways (\it{i.e.} pathways with more member genes),
+To avoid bias toward larger pathways (_i.e._ pathways with more member genes),
 we counted pathways as disrupted if one or more member genes were mutated. We do
@@
 the question, are tumor phenotypes driven entirely by common driver genes, or by
-“silent” tissue-specific effectors (\it{i.e.} too few samples to detect above
+“silent” tissue-specific effectors (_i.e._ too few samples to detect above
 statistical significance thresholds), or a combination of both? To answer this
@@
 involved in protein degradation. We also observed that genes that were enriched
-for one cluster are depleted from others (\it{i.e.} is enriched in cluster 6,
+for one cluster are depleted from others (_i.e._ is enriched in cluster 6,
 but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
@@
 rate for these genes within samples that belong to a cluster, the mutation rate
-is heterogeneous across tumors by tissue origin (\it{e.g.} In cluster 4, CESC was
+is heterogeneous across tumors by tissue origin (_e.g._ In cluster 4, CESC was
 enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
@@
 pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
-metastatic tumors are distributed across all ten clusters (\it{e.g.} **figure
+metastatic tumors are distributed across all ten clusters (_e.g._ **figure
 3E**). Using all non-metastatic tumors as background, we found very low levels
@@
 human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
-parent pathway (\it{e.g.} ‘Beta-catenin independent WNT signaling’ in ‘Signal
+parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
 Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
@@
 excluded pathways have been shown to play an important role in cancer, they are
-highly specialized (\it{e.g.} "PI3K/AKT Signaling in Cancer"). Additionally, for
+highly specialized (_e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
 most of the excluded pathways, a neutral version pathway of the pathway exists
-(\it{e.g.} "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
+(_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
 Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
@@
 We filtered likely erroneous mutations due to transcription coupled repair. Our
-approach was to determine the status of each gene (\it{i.e.} expressed or not
+approach was to determine the status of each gene (_i.e._ expressed or not
 expressed) in each tissue type in order to exclude gene mutations low expressed

### 2020-08-26 text/supplemental_methods.Rmd
fix italics *again*

@@
 for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened aggressively (
-_e.g._ colorectal and prostate cancers) while others are diagnosed typically
-after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
+confounded with cancer type because some cancers are screened aggressively (_e.g._ colorectal and prostate cancers) while others are diagnosed typically
+after they become problematic for the patient's lifestyle (_e.g._ ovarian and
 pancreatic cancers). Secondly, such a model specification would suffer from

### 2020-08-26 text/survival.Rmd
fix italics *again*

@@
 cluster-specific rate of 1 represents the typical rate for this tissue type. For
-several cancers (\it{e.g.} PRAD, kidney chromophobe (KICH), diffuse large B-cell
+several cancers (_e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
 lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely

### 2020-08-26 text/manuscript.Rmd
last minit typos and fixes (DH)

@@
 corresponds to a tumor sample. The same colors indicate the tumor's cluster
-identity throughout this manuscript. 1=yellow, 2=lilac, 3=orange, 4=light blue,
-5=crimson, 6=beige, 7=grey, 8=dark green, 9=pink, 10=dark blue.
+identity throughout this manuscript.
 
@@
 Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include (among others)
-loss of differentiation, cell-cell contacts, epithelial to mesenchyme
-transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To
-determine whether any of our clusters correspond to an especially advanced stage
-of cancer across tissue types, we projected the metastasis data onto our UMAP,
-and surprisingly we observed an even distribution of metastases across classes
-(**figure 3E**). This final observation suggests that our pathway-disruption
-classification is dependent on particular combinations of gene mutations
-affecting different pathways that can each give rise to advanced stages of
-disease and metastasis, regardless and independent of overall mutational burden
-(see Discussion for further analysis).
+certain phenotypic criteria are met. These phenotypes include loss of
+differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
+system evasion and tissue invasiveness [@hallmarks_II]. To determine whether any
+of our clusters correspond to an especially advanced stage of cancer across
+tissue types, we projected the metastasis data onto our UMAP, and surprisingly
+we observed an even distribution of metastases across classes (**figure 3E**).
+This final observation suggests that our pathway-disruption classification is
+dependent on particular combinations of gene mutations affecting different
+pathways that can each give rise to advanced stages of disease and metastasis,
+regardless and independent of overall mutational burden.
 

### 2020-08-26 code/figure9.Rmd
Reveresed Factor Levels Figure 9 1->10 now

@@
 	dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
+	dfclas$class <- fct_rev(dfclas$class)
 	ggplot(dfclas,aes_(quote(value),quote(class), fill = quote(class))) + geom_density_ridges() +

### 2020-08-26 text/discussion.Rmd

@@
 methods) as the unit of disruption instead of the gene, where individual
-mutations may be sufficient to disrupt or misdirect pathway activity (as is
-often the case with driver mutations). Rather than evaluating the burden of
-mutations on each pathway within or across tumors, which can be biased by the
-number of genes or the coverage of pathway-linked genes in the genome, we scored
-each pathway in binary fashion, allowing one mutation or multiple mutations to
-count equally for each sample.
+mutations may be sufficient to disrupt or misdirect pathway activity. Rather
+than evaluating the burden of mutations on each pathway within or across tumors,
+which can be biased by the number of genes or the coverage of pathway-linked
+genes in the genome, we scored each pathway in binary fashion, allowing one
+mutation or multiple mutations to count equally for each sample.
 

### 2020-08-27 code/supplemental_v2.Rmd
Update app.R to have cancer type selection in gene and path, fix scale of last 2 plots in figs4

@@
 individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Figure 2C of the main text**.
+in **Figure 2 of the main text**.
 
@@
 **Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Figure 2C of the main text**.
+cancers projected onto the tumors of the UMAP from **Figure 2 of the main text**.
 
@@
 **Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
-genes involved in genome stability projected on the UMAP from **Figure 2C of the main text**.
+genes involved in genome stability projected on the UMAP from **Figure 2 of the main text**.
 
@@
 s4plot(x = data)
+```
 
+```{r, fig.width = 6.7}
 par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
+par(mar = c(2,2,0,0), cex.lab = 0.8)
 for(n in pathorder[376:377]) { 
@@
                                 colvar = NULL))
-title(xlab="UMAP 2", line=0)
+title(xlab="UMAP 2", line=-1.7)
 path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
@@
 path2 <- gsub("NA", "", path2)
-legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
-legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
+legend("top", inset = c(0.11), legend=path1, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.16), legend=path2, bty = "n", horiz = T, cex = 0.8)
 }
 ```
+
 **Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
 pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
-(column) from **Figure 2C of the main text**.
+(column) from **Figure 2 of the main text**.
 
@@
 ```{r supplemental-s5-plot, fig.width=9}
-# umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv")
-# 
-# umap_cancertype$clust_knn <- as.character(umap_cancertype$clust_knn)
-# 
-# clusters178 <- umap_cancertype %>% 
-#        mutate(color = case_when(clust_knn %in% c("1", "7", "8") ~ clust_knn,
-#                                   TRUE ~ "Other"))
-# 
-# I_outlines <- c("black", "black", "black", "#00000000")
-# names(I_outlines) <- c("1", "7", "8", "Other")
-# 
-# I_colors <- c("gold2", "gray51", "springgreen4", "#00000011")
-# names(I_colors) <- c("1", "7", "8", "Other")
-# 
-# load("../data/s5_data.Rda")
-# genes <- as.character(unique(s5_data$variable))
-# 
-# par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-# par(mar = c(2,2,0,0))
-# plot.new()
-# with(clusters178, 
-#      scatter3D(plot_y, -plot_z, plot_x, 
-#                                 bg = I_colors[color], col = I_outlines[color],
-#                                 pch = 21, cex = 0.8, lwd = 0.2,
-#                                 theta = 0, phi = 65, scale = F,
-#                                 xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-#                                 colvar = NULL))
-# legend("top", inset = c(0.05), legend=names(I_colors[c(1:3)]), col=I_colors[c(1:3)], pch = 20, bty = "n", horiz = T, cex = 1.2)
-# title(xlab="UMAP 2", line=0)
-# plot.new()
-
 load("../data/s5_data.Rda")
@@
 **Supplementary Figure S5.
-Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 2C of the main text**.
+Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 2 of the main text**.
 

### 2020-08-28 text/discussion.Rmd
cut all text to conform to Cell's 45k character limit

@@
 protein-coding genes is comparable with the number of tumor samples, even with
-multiple mutations per sample. This also does not account for the fact that
-~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
-elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
-@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
-@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
-@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
+multiple mutations per sample. We sought to
 simplify the problem by employing a "knowledge-base driven analysis"
@@
 methods) as the unit of disruption instead of the gene, where individual
-mutations may be sufficient to disrupt or misdirect pathway activity. Rather
-than evaluating the burden of mutations on each pathway within or across tumors,
-which can be biased by the number of genes or the coverage of pathway-linked
-genes in the genome, we scored each pathway in binary fashion, allowing one
-mutation or multiple mutations to count equally for each sample.
+mutations may be sufficient to alter pathway activity.
 
-To our knowledge, this approach has not previously been attempted in spite of
+To our knowledge, this approach has not previously been attempted despite
 its relative simplicity. We limited our analysis to mutations with likely
@@
 [@PCAWG-MAIN] but represents a reasonable attempt to account for the same
-biases. We also filtered genes on the basis of pathway membership in biochemical
+biases. We also restricted our analysis to biochemical
 pathways, excluding curated gene sets related to diseases, syndromes, or classes
@@
 potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
-redundancy and exclude biologically unrelated collections of genes. Our
-pathway-centric analysis allows us to circumvent the problem that some
-cancer-promoting mutations likely occur at such low frequencies as to be
-indistinguishable from background.
+redundancy and exclude biologically unrelated collections of genes.
 
 The hypothesis that cancer is the result of dysfunction from a limited number of
-basic cellular processes common to all living human cells was introduced and
+basic cellular processes common to eukaryotic cells was introduced and
 later expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
-@hallmarks_II]. In this view, all tumors regardless of tissue of origin acquire
-six essential phenotypes; sustaining proliferative signaling, evading growth
-suppressors, replicative immortality, evading apoptosis, supportive
-angiogenesis, tissue invasion and metastasis [@weinberg2000]. Advancements in
-the field over the ensuing decade led to the inclusion of new tumor promoting
-processes, namely deregulation of cellular energy cycle, tumor promoting
-inflammation, evasion of the immune system, genome instability, and on top of
-these acquired traits a new appreciation for the concept of the tumor
-microenvironment and its contribution to the evolution and promotion of
-metastatic disease [@hallmarks_II].
-
-An alternative hypothesis is that every tumor belongs to one of a large number
+@hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of a large number
 of syndromes which are unique to each tissue-of-origin, that share some
 mechanisms and treatment strategies. Recent publication of TCGA consortium
-papers present a view largely, and surprisingly, consistent with this hypothesis
+papers present a view largely, and surprisingly, consistent with this latter hypothesis
 [@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
@@
 instead support an interpretation of publicly available data in which all tumors
-manifest one of a limited number of metastable phenotypes resulting from
-disruptions in basic pathways as represented by our clusters.
+manifest one of a limited number of phenotypes resulting from
+disruptions of basic pathways.
 
 We attempted to account for our clusters in terms of more trivial explanations.
-For example, it could be that the clusters we identified are consistent with a
-signature of overall disease progression, such that successive, adjacent
-clusters on the UMAP projection exhibit increasing specificity. We identified no
-such trend in the number of mutations, the relative staging or metastasis; and
-each cluster instead was associated with unique combinations of pathways.
+For example, it could be that the clusters we identified are consistent with
+disease progression. We were unable to identify any such trend in the number of
+mutations, the relative staging or metastasis; and each cluster instead was
+associated with unique combinations of pathways.
 
-We find that most tumors segregate into one of ten biologically distinct
-clusters, regardless of the cancer type. Some cancer types are unevenly
-distributed, though we could not identify any cancers that were exclusive to a
-single tumor cluster. Only cancer types with the fewest samples were found to be
+Some cancer types are unevenly
+distributed among the clusters, though we could not identify any cancers that were exclusive to a
+single cluster. Only cancer types with the fewest samples were found to be
 absent from one or more clusters at all. Somewhat surprisingly to us, this
 finding extends to histological subtypes of breast, head and neck cancers,
-leukemias, _etc._ This result implies that many, if not most, histological
+leukemias, _etc._ This result implies that histological
 subtypes could reflect differences in cell-of-origin, rather than fundamental
-differences in proximal cause. The four major subtypes of breast cancer
+differences in cancer phenotype. The four major subtypes of breast cancer
 correspond to histological and molecular expression profiles that define them
@@
 are distributed across clusters. We interpret these data to mean that inherited
-cell-of-origin molecular profiles may predispose certain precursor cell types
+cell-of-origin signatures could predispose certain precursor cells
 within the breast epithelium to forming tumors of one cluster or another but are
-not determinative; this view is compatible with the previously stated hypothesis
+not determinative. This view is compatible with the previously stated hypothesis
 but opens the way for a more granular view of individual tumors. It would be
@@
 of the unique combination of driver genes that are disrupted in each tumor, and
-we speculate that many of these are relatively tissue-specific in terms of their
-sensitivity to mutation. Consistent with our observations, @moonlight showed how
-tissue- and cancer-specific driver genes and druggable targets of pathway moduli
-can be discovered by machine learning on integrated multi-omics datasets.
-Marrying these approaches in future work will greatly enhance our understanding
-of tumor biology.
+we speculate that many of these are sensitive to mutation in different tissues.
 
@@
 clusters, 1 and 10, for which we did not find many distinctive associations with
-pathways. Cluster 1 had a relatively low proportion of mutated pathways across
-the board, although it is broadly enriched in many of the same tumor-promoting
-pathways common to the other groups. This likely does not reflect a true
-difference in stages of tumor evolution, as our data show clearly that this
-cluster is as likely to contain stage IV metastatic tumors as it is to contain
-those of stage I. This cluster likely represents a group of tumors with
-aberrations in methylation, key deletions, or other structural variants.
-Consistent with this, kidney chromophobe and thyroid cancers have high
-proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and
-are heavily skewed to cluster 1 membership. Likewise, cluster 10 represents a
-group of hyper-mutated tumors that harbor so many mutations that virtually no
-pathway is unaffected. It seems likely that a significant fraction of the
-“mutant” samples for each pathway are burdened with excess passenger mutations.
-This could be addressed with application of ever-increasingly sophisticated
-filtering of likely passenger mutations [_e.g._ @spliceAI;
+pathways. Cluster 1 had a relatively low proportion of mutated pathways,
+although it is broadly enriched in many of the same tumor-promoting pathways
+common to the other groups. In addition, our data show clearly that this cluster
+is as likely to contain stage IV metastatic tumors as it is to contain those of
+stage I. This cluster likely represents a group of tumors with aberrations in
+methylation, copy number, or other structural variants. Consistent with this,
+kidney chromophobe and thyroid cancers have high proportions of structural
+variations _vs._ other variant types [@PCAWG-MAIN] and are heavily skewed to
+cluster 1 membership. Likewise, cluster 10 represents a group of hyper-mutated
+tumors that harbor so many mutations that virtually no pathway is unaffected. It
+seems likely that a significant fraction of the “mutant” samples for each
+pathway are burdened with excess passenger mutations. This could be addressed
+with more sophisticated filtering of likely passenger mutations [_e.g._
+@spliceAI;
 @deep-learning-clinical-variants]. In the future, we hope to incorporate these
@@
 of our pathway disruption clusters exhibit cancer-type specific effects on
-expected survival. Contrary to our initial expectation, we found that models
-assuming different survival based on cluster $\times$ cancer type fit better than
-models in which they are independent factors. However, considering that within
-each cancer type there are different clinical standards of care, and even within
-classes of drugs the preferred treatment can very between cancers, it makes
-sense that we observe tissue-specific cluster effects. Contrast the situation
-with ovarian _vs._ breast cancer, which are both hormonally driven cancers, for
-example. Ovarian cancer has but one main treatment axis, platinum, whereas
-breast cancer patients have a variety of treatment regimens based on molecular
-subtype and other factors. Standard of care for one cancer and cluster could
-produce different outcomes than the same cluster of a different cancer type.
-Unfortunately, given the diversity of drug classes and treatments (including
-non-chemical treatments such as surgery and palliative care), we lack sufficient
-power to explore these variables in the TCGA data. It is our hope that future
-studies will help to distinguish between treatment-specific effects on survival
-given different pathway disruption clusters.
+expected survival. However, considering that within each cancer type there are
+different clinical standards of care, and even within classes of drugs the
+preferred treatment can very between cancers, it makes sense that we observe
+tissue-specific cluster effects. Contrast the situation with ovarian _vs._
+breast cancer, which are both hormonally driven cancers, for example. Ovarian
+cancer has but one main treatment axis, platinum, whereas breast cancer patients
+have a variety of treatment regimens based on molecular subtype and other
+factors. Unfortunately, given the diversity of drug classes and treatments, we
+lack sufficient power to explore these variables in the TCGA data. It is our
+hope that future studies will help to distinguish between treatment-specific
+effects on survival given different pathway disruption clusters.
 
@@
 that can be broadly thought of as general cancer promoting, cluster-specific
-mutations and metastasis. General cancer promoting processes must include
-factors that relate to genome stability and immortality, as "enabling
-characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
-disrupted in all or most of the defined clusters
-and are frequently the result of aberrations involving common driver genes such
-as BRCA1/2, MMR genes, mitotic checkpoints, cohesion complexes, _etc._
-Cluster-specific
+mutations and metastasis. General cancer promoting processes include genome
+stability and immortality, as "enabling characteristics" of the cancer phenotype
+[@hallmarks_II]. Such pathways are disrupted in most clusters and are frequently
+the result of aberrations involving common driver genes such as BRCA1/2, MMR
+genes, mitotic checkpoints, cohesion complexes, _etc._ Cluster-specific
 evolution must involve the acquisition of disruptions to pathways that may
-individually be harmful ( _e.g._ highly proliferative cells are more likely to
+individually be harmful (_e.g._ highly proliferative cells are more likely to
 senesce) but together produce more specialized cancer phenotype and increased
-fitness. Importantly, our
-observations do not imply the order in which these mutations should accumulate.
-This could be addressed in a future study by integrating our analysis with
-evolutionary analysis of clonality, drawing inferences from variant allele
+fitness. Importantly, our observations do not imply the order in which these
+mutations should accumulate. This could be addressed in a future study by
+evolutionary analysis of clonality, drawing inference from variant allele
 frequencies as in @PCAWG-EVO. However, since many of the genes in the
 non-cluster-specific pathways involve the known driver genes, it is reasonable
-to hypothesize that these mutations promote or enable the subsequent acquisition
-of cluster-specific defects via random mutation and natural selection, thus
+to surmise that these mutations promote or enable acquisition of
+cluster-specific defects via random mutation and natural selection, thus
 producing the clusters we observed. In support of this, the pan-cancer analysis
@@
 are known to be responsible for discrete mutation signatures such as APOBEC,
-BRCA1 and BRCA2 produce mutational hotspots reflecting the varying selective
+BRCA1 and BRCA2 produce mutational hotspots reflecting varying selective
 pressures in different tissues [@PCAWG-MAIN].
@@
 
-We report that pathway enrichment in metastatic tumors across all clusters
-(pooled) yielded generally lower effect sizes and larger p-values than the
-cluster specific analysis, suggesting that signal is diluted when the samples
-are analyzed together, and supporting the view that metastasis has
-cluster-specific requirements. Since the number of metastatic samples is
-relatively low, this part of our analysis is likely underpowered and subject to
-expanded analysis with larger cohorts. The fact that most metastatic enrichment
-is cluster-specific and has a tendency to overlap with cluster-specific pathways
-from non-metastatic tumors of neighboring clusters suggests a mechanism of
-complementarity, in which newly acquired mutations result in similar clusters
-converging on one or more deadly phenotypes that capture most of the critical
-features of end-stage cancer. Thus, even at relatively low power, our analysis
-of metastasis further supports the pathway-centric approach as a fruitful way to
-uncover biologically meaningful differences between tumors against the noisy
-backdrop of tissue specific profiles. As a major caveat to this, we acknowledge
-that many therapies _do_ target highly-specific driver genes, markers, and
-signaling pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader
-context of the genetic background and pathway vulnerability of tumors containing
-such markers may aid in creating smarter combination therapies. We submit that
-when we discover the requirements of each cluster with respect to pathway
-disruptions and metastasis we may be able to target them therapeutically and
-prevent further adaptation.
+We report that enrichment in metastatic tumors across all clusters yielded
+generally lower effect sizes and larger p-values than the cluster specific
+analysis, suggesting that signal is diluted when clusters are pooled, and
+supporting the view that metastasis has cluster-specific requirements. Since the
+number of metastatic samples is relatively low, this part of our analysis is
+likely underpowered and subject to expanded analysis with larger cohorts. The
+fact that most metastatic enrichment is cluster-specific and has a tendency to
+overlap with cluster-specific pathways from non-metastatic tumors of neighboring
+clusters suggests that newly acquired mutations result in similar clusters
+converging on one or more deadly phenotypes with critical features of end-stage
+cancer. Thus, even at relatively low power, our analysis of metastasis uncovered
+differences between tumors against the noisy backdrop of tissue specific
+profiles. As a major caveat to our pan-cancer analysis, we acknowledge that many
+therapies _do_ target highly-specific driver genes, markers, and signaling
+pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader context of
+the genetic background and pathway vulnerability of tumors containing such
+markers may aid in creating smarter combination therapies. We submit that when
+we discover the requirements of each cluster with respect to pathway disruptions
+and metastasis we may be able to target them therapeutically and prevent further
+adaptation.

### 2020-08-28 text/introduction.Rmd
cut all text to conform to Cell's 45k character limit

@@
 Recent advances in technology have greatly expanded our view into the mechanisms
-of cancer at the systems biology level. Array and next-generation sequencing
-technologies have made it possible to assay germline and somatic mutations,
-expression profiles, DNA methylation, and copy number variations from the same
-tissue samples. To take advantage of these tools, several large consortia,
-including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
-genomes (PCAWG) sequenced large numbers of tumors and collected data from
-multiple assays that could be analyzed together, with the goal of integration
-and increased understanding of the mechanisms of cancer.
+of cancer at the systems biology level. Next-generation sequencing technologies
+have made it possible to assay germline and somatic mutations, expression
+profiles, DNA methylation, and copy number variations from the same tissue. To
+take advantage of these tools, several large consortia, including the cancer
+genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) sequenced
+large numbers of tumors multiple assays to analyze together, with the goal of
+integration and increased understanding of the mechanisms of cancer.
 
@@
 @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
-when genes are mutated at a greater rate than expected due to chance. It has
-been estimated that fewer than five mutations in key oncogenes and/or tumor
+genes mutated at a greater rate than expected due to chance. It has been
+estimated that fewer than five mutations in key oncogenes and/or tumor
 suppressors would be sufficient to transform a normal cell to a cancerous state
-[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
-mutations in cancer driver genes are commonplace in healthy tissue and
-correlate with age and environmental exposures [@mutations-healthy-tissues].
+[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Mutations in cancer driver
+genes are commonplace in healthy tissue and correlate with age and environmental
+exposures [@mutations-healthy-tissues].
 
 Other studies provide a comprehensive view of mutations, gene expression and
-genomic signatures, with the express goal of understanding the common themes of
-all cancer independent of tissue of origin. Understanding cancer as a disease of
-the cell has long been a goal of the field as characterized in essays by
-@weinberg2000; @hallmarks_II. The first of these studies in the genomic era
-identified 11 subtypes from 12 cancer types, using integrative analysis with
-co-equal weighting of gene expression, methylation, copy number and proteomics
-data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that
-tissue-of-origin is the predominant driving factor, though a significant
-proportion of tumors (~10%) could be reclassified independent of
-tissue-of-origin. In a second study involving 33 cancer types and a much greater
-number of tumors, the authors identify
-28 clusters that could be further subdivided into organ specific groups,
-including pan-gastrointestinal, pan-gynecological, pan-squamous,
-pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].
+genomic signatures, with the goal of understanding common themes of all cancer
+independent of tissue of origin. Understanding cancer as a disease of the cell
+has long been a goal of the field as characterized in essays by
+@weinberg2000; @hallmarks_II. The first of these studies in genomics identified
+11 subtypes from 12 cancer types, using integrative analysis with co-equal
+weighting of gene expression, methylation, copy number and proteomics data
+[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
+tissue-of-origin is the predominant driving factor, though ~10% could be
+reclassified independent of tissue-of-origin. In a second study involving 33
+cancer types and a much greater number of tumors, the authors identified 28
+clusters that could be further subdivided into organ specific groups, including
+pan-gastrointestinal, pan-gynecological, pan-squamous,
+pan-gynecological/squamous and pan-kidney
+[@hoadleyCellofOriginPatternsDominate2018].
 
-More recently, attempts have been made to consider pan-cancer mutations and
-other data as disruptions of normal pathway activity. @PCAWG-PATH carried out
-near-exhaustive analysis of pathway enrichment using whole genome data. Network
-approaches address the sparseness of mutations, allowing genes to be influenced
-by mutations in the nearest network neighbors
-[@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
-powerful especially when applied to a small number (on the order of hundreds) of
-tumors. One of the challenges in the field remains that beyond a couple hundred
-common driver genes and mutations, there may be thousands of moderate effect
-genes that occur at such low frequency--given tissue diversity--and low sample
-numbers that they are impossible to detect using positive selection theory. Some
-groups have attempted to address this challenge by applying machine learning to
-cancer data sets to discover groups of functionally related genes as they
-interact with larger pathways and networks [@kimPancancerAnalysisSomatic2018;
+More recently, pan-cancer mutations and other data have been looked at as
+disruptions of normal pathway activity. @PCAWG-PATH carried out exhaustive
+analysis of pathway enrichment using whole genome data. Network approaches
+address the sparseness of mutations, allowing genes to be influenced by
+mutations in nearest network neighbors [@hornNetSigNetworkbasedDiscovery2018;
+@Iranzo-Network]. These methods are powerful especially when applied to smaller
+numbers of tumors. One of the challenges in the field remains that there may be
+thousands of moderate effect genes that occur at such low frequency--given
+tissue diversity and low sample number--that they are impossible to detect using
+positive selection theory. Some groups have attempted to address this challenge
+by applying machine learning to cancer data to discover groups of
+functionally related genes as they interact with larger pathways and networks
+[@kimPancancerAnalysisSomatic2018;
 @mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
-highlighting fundamental disease phenotypes at the pathway
-level across cancers with different origins at the cellular and tissue
-level.
+highlighting fundamental disease phenotypes at the pathway level across cancers
+with different origins at the cellular and tissue level.
 

### 2020-08-28 text/manuscript.Rmd
cut all text to conform to Cell's 45k character limit

@@
 survival and staging. We chose to analyze somatic mutations in exome sequencing
-data because the affected target gene is known unambiguously. Therefore, we
+because the affected target gene is known unambiguously. Therefore, we
 selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
@@
 
-To avoid bias toward larger pathways (_i.e._ pathways with more member genes),
-we counted pathways as disrupted if one or more member genes were mutated. We do
-not attempt to calculate enrichment for mutations within a pathway. Data
-binarized by pathway are likely to be noisy for several reasons. First, point
-mutations can be deleterious (attenuating, hypomorphic or antimorphic) or
-activating (neomorphic or hypermorphic) in genes, and these can in turn act
-functionally as either oncogenes or tumor suppressors. For this study we assume
-a significant fraction of these mutations are generically disruptive to normal
-pathway activity since it is impossible to know the tumor promoting effects of
-all mutations, including rarely studied genes. Second, we know that
-low-expressed genes and non-expressed genes accumulate mutations at a higher
-rate due to transcription coupled repair
+To avoid bias toward larger pathways with more genes, we counted them as
+disrupted if one or more member genes were mutated. We do not attempt to
+calculate enrichment for mutations within a pathway. Data binarized by pathway
+are likely to be noisy for several reasons. First, point mutations can be
+deleterious (attenuating, hypomorphic or antimorphic) or activating (neomorphic
+or hypermorphic) in genes, and these can in turn act functionally as either
+oncogenes or tumor suppressors. For this study we assume a significant fraction
+of these mutations are generically disruptive to normal pathway activity since
+it is impossible to know the tumor promoting effects of all mutations, including
+rarely studied genes. Second, we know that low- and non-expressed genes
+accumulate mutations at a higher rate due to transcription coupled repair
 [@kandothMutationalLandscapeSignificance2013;
@@
 @Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
-issue, we identified genes with low expression in each type of cancer and
-eliminated them for that cancer type only (see methods for details). Highly
-expressed genes could also have high mutation rates owing to transcription
-induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that such
-mechanisms result in cell-type-specific biases that might be biologically
-meaningful for predisposition to different classes of cancer however, and
-therefore chose not to exclude these genes from our analysis. After selecting
-our pathways and genes, we then compiled a matrix of the pathways, assigning a
-Boolean value of 1 to each pathway with one or more genes mutated and 0 for all
-others (**figure 1**).
+issue, we identified low expressed genes in each type of cancer and eliminated
+them for that cancer type only (see methods for details). Highly expressed genes
+could also have high mutation rates owing to transcription induced mutagenesis
+[@parkGenomicEvidenceElevated2012]. This phenomenon would likely result in
+cell-type-specific biases that might be biologically meaningful for
+predisposition to different classes of cancer however, and therefore chose not
+to exclude these genes from our analysis. After selecting our pathways and
+genes, we then compiled a matrix of the pathways, assigning a Boolean value of 1
+to each pathway with one or more genes mutated and 0 for all others (**figure
+1**).
 
@@
 mutations.** Each of 377 selected Reactome pathways (rows) is classified as
-disrupted if one or more genes is mutated in the tumor sample (columns) where
-red and grey represents the pathway is mutated and not, respectively. Tumors are
-grouped by tissue of origin using standardized abbreviations from the TCGA
-project.
+disrupted if one or more genes is mutated in the tumor sample (columns). Red
+cells denote pathway disruption. Tumors types reflect standardized abbreviations
+from the TCGA project.
 
@@
 Having defined tumors in terms of their pathway disruption profile, we sought to
-understand whether different cancer types segregated into one or more
-predominant classes. To our surprise, most cancer types were not heavily biased
-in one class, and all well-represented cancer types had tumors in every class
-(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
-also interactive media from **supplementary file 1**), suggesting that,
-in principle at least, our pathway-disruptions identify clusters of molecular
-pathology largely independent of tissue-of-origin. As an example of one type of
-cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
-was predominantly found in class 8 (**figure 3A** and **supplementary figure
-S1** and **supplementary file 1**). But even PAAD comprises tumors from the nine remaining classes, meaning
-that patients suffering these tumors have potentially different underlying
+understand whether different cancer types segregate into one or more predominant
+clusters. To our surprise, most cancer types were not heavily biased in one
+cluster, and all well-represented cancer types had tumors in every cluster (see
+**figure 3A** and full tumor profiles in **supplementary figure S1**, see also
+interactive media from **supplementary file 1**), suggesting that our
+pathway-disruptions identify clusters of molecular pathology largely independent
+of tissue-of-origin. As an example of one type of cancer that does have a biased
+pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
+cluster 8 (**figure 3A** and **supplementary figure S1** and **supplementary
+file 1**). But even PAAD comprises tumors from the nine remaining classes,
+meaning that patients with these tumors have potentially different underlying
 molecular pathologies.
@@
 Many cancers have molecular or histological subtypes defined based on gene
-expression or other -omics profiles or pathology lab results. These subtypes
-often have different standards of care owing to different overall drug
-sensitivity (or other factors). If the histological subtypes represent true
-molecular phenotypes, one predicts that histological subtypes should segregate
-with our pathway-based clusters, therefore providing support for the clusters as
-proxies for molecular pathology sub-typing. To our surprise, we find a similar
-result to the previous analysis of cancer types projected onto the UMAP of
-pathway disruptions. To illustrate this, we projected annotations for each of
-the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
-positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
-among the most heavily studied molecular subtypes in cancer, which each have
-different prognoses and standards of care. We did not observe any exclusive
-segregation by pathway for these subtype annotations (**Figure 3B**). We also 
-projected histological subtype data for the remaining cancers (see 
-**supplementary figure S2** for the full set of projections and see interactive
-media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the subtypes, though often 
-biased towards one or more classes, are almost never exclusive. We interpret 
-these data in aggregate to mean that our pathway disruption classes do not 
-correspond to previously identified molecular subtypes within the parent cancer type.
+expression, pathology or other -omics profiles. These subtypes often have
+different standards of care owing to different drug sensitivity (or other
+factors). If the histological subtypes represent true molecular phenotypes, one
+predicts that they should segregate with our pathway-based clusters, therefore
+providing support for the clusters as proxies for molecular pathology
+sub-typing. To our surprise, we find a similar result to the previous analysis
+of cancer types projected onto the UMAP of pathway disruptions. To illustrate
+this, we projected annotations for each of the breast cancer subtypes, composed
+of Triple-negative/Basal-like, Her2 positive, normal-like, and luminal A and B
+subtypes onto the UMAP. These are among the most heavily studied molecular
+subtypes in cancer, each with different prognoses and standards of care. We did
+not observe any exclusive segregation by pathway for these subtypes (**Figure
+3B**). We also projected histological subtype data for the remaining cancers
+(see **supplementary figure S2** and online [**supplemental file
+1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the
+subtypes, though often biased to one or more clusters, are almost never
+exclusive. We interpret these data in aggregate to mean that the clusters do not
+correspond to previously identified molecular subtypes within the parent cancer
+types.
 
@@
 There are several well-known familial cancer-causing mutations that have been
-interrogated extensively for differences in basic biology, survival and
-treatment outcomes. However, the functions of these genes are related to risk
-factors such as genome stability generally, proof-reading and DNA damage repair,
-and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
-break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
-these genes confer elevated risk for breast, prostate and ovarian cancers. The
-mechanism of risk is thought to involve loss of heterozygosity resulting in loss
-of the wildtype, functional allele [@brca-loh], so we projected the somatic
-mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP, but did not observe
-segregation of these mutations into specific clusters (**figure 3C** and see 
-interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar projections 
-for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ 
-and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except 
-for _APC_ exhibited any remarkable specificity with respect to cluster assignment 
-(**supplementary figure S3**). To look at other risk factors
-such as maintenance of DNA methylation levels and telomere length, we projected
-somatic mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and
-_DKC1_, and observed similar lack of segregation by cluster (**supplementary
-figure S3**).
+studied for differences in basic biology, survival and treatment outcomes. The
+functions of these genes are related to risk factors such as genome stability,
+proof-reading & DNA damage repair, and telomere length. _BRCA1/2_ genes for
+example are key for DNA double-stranded break repair
+[@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in these genes
+confer elevated risk for breast, prostate and ovarian cancers. The mechanism of
+risk is thought to involve loss of heterozygosity resulting in loss of the
+wildtype, functional allele [@brca-loh], so we projected the somatic mutations
+for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe segregation of
+these mutations into specific clusters (**figure 3C** and see interactive media
+from [**supplemental file
+1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar
+projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_,
+_PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
+except for _APC_ exhibited any remarkable specificity with respect to cluster
+assignment (**supplementary figure S3**). To look at other risk factors such as
+maintenance of DNA methylation levels and telomere length, we projected somatic
+mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
+observed similar lack of segregation by cluster (**supplementary figure S3**).
 
@@
 is used clinically as a proxy for advancement toward a more deadly state and
-metastasis. Given these assumptions, it is possible that more advanced tumors
-have common pathway disruption profiles. The UMAP, which features a series of
+metastasis. Given these criteria, it is possible that more advanced tumors have
+common pathway disruption profiles. The UMAP, which features a series of
 lobe-like structures on a common backbone of tumor samples (**figure 2**) could
@@
 culminates in a cluster (class 10) which has nearly every pathway disrupted
-(**figure 4**). However, outside of class 10 we don't observe an obvious trend
-in the overall mutation burden across the backbone of the UMAP. Nonetheless, to
-test the hypothesis that the molecular-pathway disruption clusters represent
+(**figure 4**). However, outside of class 10 we don't observe a trend in the
+overall mutation burden across the backbone of the UMAP. Nonetheless, to test
+the hypothesis that the molecular-pathway disruption clusters represent
 advancement through stages, we projected staging data onto the UMAP. Similar to
 tissue of origin and other categories of tumor, we did not observe any bias
-among the stages to specific pathway disruption clusters (**figure 3D**),
-suggesting that stage is not a contributing factor to cluster identity.
+among the stages to specific clusters (**figure 3D**), suggesting that stage is
+not a contributing factor to cluster identity.
 
@@
 
-**Figure 4: Somatic mutation frequencies for tumors in each class** 
+**Figure 4: Somatic mutation frequencies for each cluster** 
 Vertical axis shows log mutation count, horizontal axis is cluster identity.
@@
 Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met. These phenotypes include loss of
+certain phenotypic criteria are met: loss of
 differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
@@
 of our clusters correspond to an especially advanced stage of cancer across
-tissue types, we projected the metastasis data onto our UMAP, and surprisingly
-we observed an even distribution of metastases across classes (**figure 3E**).
-This final observation suggests that our pathway-disruption classification is
+tissue types, we projected the metastases onto our UMAP, and surprisingly
+we observed even distribution of the samples across clusters (**figure 3E**).
+This final observation suggests that our pathway-disruption clustering is
 dependent on particular combinations of gene mutations affecting different
@@
 
-In order to identify pathway disruption enrichment across all cancers
-(pan-cancer), we created a list of pathway disruptions with percent mutated
-samples and top genes (**supplementary table ST1**). As expected,
-these analyses reveal the broad importance of many well known pathways that are
-disrupted in cancer, including "PIP3 activates Akt signaling" (77% of samples),
-"MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
-samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
-samples), _etc._
+In order to identify pathway enrichment across all cancers (pan-cancer), we
+created a list of pathway disruptions with percent mutated samples and top genes
+(**supplementary table ST1**). As expected, these analyses reveal the broad
+importance of many well known pathways that are disrupted in cancer, including
+"PIP3 activates Akt signaling" (77% of samples), "MAP1K/MAP3K signaling" (70% of
+samples), "Mitotic G2-G2/M phases" (67% of samples), "Cellular senescence" (64%
+of samples), "G2/M Checkpoints" (62% of samples), _etc._
 
 To discover what pathways are most important for clustering, we calculated
-percent enrichment for each pathway _within cluster_ relative to all other
-clusters combined (see methods) and ranked pathways for each cluster from
-highest to lowest enrichment. We visualized the enrichment as a heatmap
-(**figure 5A**). Using this approach, we identified twenty-five pathways highly
-enriched (enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2,
-nine pathways enriched in cluster 3, twenty-four pathways in cluster 4, ten
-pathways in cluster 5, ten pathways in cluster 6, twenty-four pathways in
-cluster 7, twenty-four pathways in cluster 8, ten pathways in cluster 9, and no
-pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
-9 had several pathways in common. To explore the specific pathways marking each
-cluster, we projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). 
-Clusters 3 and 5 were distinguished by metabolic pathways including RNA and 
-protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was 
-distinguished by mutations affecting regulation of DNA
-and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
-and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
-extracellular, intracellular, and immune-related signaling pathways (see
-**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
-pathway enrichment levels of the three, having mutations in hedgehog signaling,
-"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
-cycle" and "apoptosis" among others.
+percent enrichment _within cluster_ relative to all other clusters (see methods)
+and ranked pathways from highest to lowest enrichment. We visualized the
+enrichment as a heatmap (**figure 5A**). Using this approach, we identified
+about fourteen pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see
+methods) (**supplementary table ST2**). Clusters 7, 8, and 9 had several
+pathways in common. To explore the specific pathways marking each cluster, we
+projected disruptions for each of the 377 pathways onto the UMAP
+(**supplementary figure S4** and see interactive media from [**supplemental file
+1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
+were distinguished by metabolic pathways including RNA and protein biosynthesis
+(**supplementary figure S4**). Similarly, cluster 4 was distinguished by
+mutations affecting regulation of DNA and histone methylation ("DNA
+methylation", "PRC2 methylates histones and DNA", and "Nucleosome assembly").
+Clusters 7-9 have in common mutations in extracellular, intracellular, and
+immune-related signaling pathways (see **figure 5B** and **supplementary figure
+S4**). Cluster 2 had the highest pathway enrichment levels of the three, having
+mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
+hypoxia", "regulation of cell cycle" and "apoptosis" among others.
 
@@
 
-Prior efforts to extract common signatures from pan-cancer datasets met with
-difficulty in distinguishing tumor samples from their tissue-specific -omics
-data signatures. Given our pathway-disruption based clustering, this raises
-the question, are tumor phenotypes driven entirely by common driver genes, or by
+Prior efforts to extract signatures from pan-cancer datasets met with difficulty
+in distinguishing tumor samples from their tissue-specific -omics data
+signatures. Given our pathway-disruption based clustering, this raises the
+question, are tumor phenotypes driven entirely by common driver genes, or by
 “silent” tissue-specific effectors (_i.e._ too few samples to detect above
@@
 _KRAS_ panels of **Supplementary figure S5** and see interactive media from
-[**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly, cluster 9, which
-also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
-enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
-in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
-complex, respectively. Cluster 4 had mutations in genes responsible for
-nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
-involved in protein degradation. We also observed that genes that were enriched
-for one cluster are depleted from others (_i.e._ is enriched in cluster 6,
-but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
-clusters 3 and 8). Next, we investigated the proportion of samples per cancer
-type for the significant genes within a cluster (**Figure 6**; heatmap).
-Surprisingly, cluster-specific tumors were not predominated by one or more
-highly mutated genes across all cancers. Instead, when observing the mutation
-rate for these genes within samples that belong to a cluster, the mutation rate
-is heterogeneous across tumors by tissue origin (_e.g._ In cluster 4, CESC was
-enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
-_HIST1H2AC_). Even among the top most mutationally enriched genes within
-clusters there is no global pattern, indicating that our clusters are not driven
-by individual genes, but rather networks as a whole. Taken together, our data
-identifies a framework of cancer type-specific mutations associated with
-specific clusters.
+[**supplemental file
+1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
+cluster 9, which also shared multiple enrichment in signaling pathways with
+clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
+defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
+and nuclear pore complex, respectively. Cluster 4 had mutations in genes
+responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
+subunit genes involved in protein degradation. We also observed that genes that
+were enriched for one cluster are depleted from others (_i.e._ is enriched in
+cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
+depleted in clusters 3 and 8). Next, we investigated the proportion of samples
+per cancer type for the significant genes within a cluster (**Figure 6**;
+heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
+more highly mutated genes across all cancers. Instead, when observing the
+mutation rate for these genes within samples that belong to a cluster, the
+mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
+4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
+enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
+there is no global pattern, indicating that our clusters are not driven by
+individual genes, but rather networks as a whole. Taken together, our data
+identify a framework of cancer type-specific mutations associated with specific
+clusters.
 
@@
 Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
-95% confidence interval. Row represents the significant genes from each cluster.
-Each cluster was compared against the background (all other clusters) to find
-differentially mutated genes. Significant genes (pvalue < 0.01) were selected
-and limited to the top ten results for each clusters. Heatmap; Column represents
-the cancer type. Row represents the significant genes as described above. The
-heatmap shows the proportion of samples per cluster and cancer type that are
-mutated for each gene. Depleted significant genes (left). Enriched significant
-genes (right).
+95% confidence interval. Rows: significant genes from each cluster. Each cluster
+was compared against the background (all other clusters) to find differentially
+mutated genes. Significant genes (pvalue < 0.01) were selected and limited to
+the top ten results for each cluster. Heatmap; Columns: cancer type. Rows:
+significant genes. The heatmap shows proportion of samples per cluster and
+cancer type mutated for each gene. Depleted significant genes (left). Enriched
+significant genes (right).
 
@@
 
-Following the same logic we used to investigate cluster-specific enrichment of
-pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
-metastatic tumors are distributed across all ten clusters (_e.g._ **figure
-3E**). Using all non-metastatic tumors as background, we found very low levels
-of enrichment (< 10%) in a handful of pathways. We reasoned that the individual
-clusters might be too different to detect global metastasis enrichment signal
-given the small sample size (n = 215 metastatic tumor samples).
+Since metastatic tumors are distributed across all ten clusters, we first
+compared them with non-metastatic tumors following the same logic we used to
+investigate cluster specific enrichment. Using all non-metastatic tumors as
+background, we found very low levels of enrichment (< 10%) in a handful of
+pathways. We reasoned that the individual clusters might be too different to
+detect global metastasis enrichment signal given the small sample size (n = 215
+metastatic tumor samples).
 
@@
 found a total of 31 enriched pathways (significant with enrichment score ≥ 0.3)
-across all clusters (**table 1**). A number of enrichments were found in
-multiple clusters represented by pathways that were already shown to be enriched
-in a neighboring cluster. For example, “Signaling by _PTK6_” is enriched in
-non-metastatic samples of cluster 8 (see **supplementary figure S4**) but not in
-7 and 9. This pathway is enriched in metastatic tumors of clusters 7 and 9 ($p <
-10^{-3}$, **table 1**). This is also true of “Erythropoietin activates _RAS_”,
-which is enriched in non-metastatic tumors of cluster 8 (**supplementary figure
-S4**) and also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic
-samples were enriched for “Fc epsilon receptor (FCERI) signaling”, a key
-neutrophil pathway, which is also specific to clusters 2, 7 and 8 non-metastatic
-tumors. Thus, metastatic-enriched pathways from one cluster are often enriched in 
+across all clusters (**table 1**). A number of enrichments represented pathways
+that were already shown to be enriched in non-metastatic samples of other
+clusters. For example, “Signaling by _PTK6_” is enriched in non-metastatic
+samples of cluster 8 (see **supplementary figure S4**) but not in 7 and 9. This
+pathway is enriched in metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$,
+**table 1**). This is also true of “Erythropoietin activates _RAS_”, which is
+enriched in non-metastatic tumors of cluster 8 (**supplementary figure S4**) and
+also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic samples were
+enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathway,
+which is also specific to clusters 2, 7 and 8 non-metastatic tumors. Thus,
+metastatic-enriched pathways from one cluster are often enriched in
 non-metastatic tumors of other clusters.
@@
 
-All code for producing the analyses and figures herein are included in this
-fully reproducible manuscript in R markdown format. R markdown files and all
-other models are available from [our
+All code and data for producing the analyses and figures herein are included in
+this fully reproducible manuscript in R markdown format. R markdown files and
+all other models are available from [our
 repositories](https://github.com/dennishazelett/TTmanu) on the distributed
-version control site, GitHub.
-
-### Selection of pathways
-
-To understand the molecular mechanism of cancer at a pathway level, we used
-Reactome (https://reactome.org/), a knowledge-based pathway database. The
-mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
-complexes to top pathways were downloaded from
-https://reactome.org/download-data. Using these data, we imposed pathway
-criteria to define basic cellular processes and biochemical pathways: (1)
-human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
-parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
-Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
-"Muscle contraction", and "Reproduction" or pathway names that include any of
-the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
-"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
-excluded pathways have been shown to play an important role in cancer, they are
-highly specialized (_e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
-most of the excluded pathways, a neutral version pathway of the pathway exists
-(_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
-Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
-operation produced a lookup table that consisted of 377 pathways mapped to 8,940
-genes.
-
-### Filtering genes
-
-We filtered likely erroneous mutations due to transcription coupled repair. Our
-approach was to determine the status of each gene (_i.e._ expressed or not
-expressed) in each tissue type in order to exclude gene mutations low expressed
-enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
-effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
-the genes and tumor samples that were not included in our analysis, grouped the
-tumor samples by tissue type and computed the mean expression value for each
-gene. A minimum threshold of 10 transcripts per million was set for expressed
-genes based on an inflection point observed when plotting the mean expression
-values of genes ranked by expression in each tissue type. Genes that did not
-meet this threshold were considered not expressed. This operation produced a
-lookup table for gene expression status in each tumor sample for 18,127 genes.
-
-### Class and stage-specific Enrichment Calculations
-
-We calculated the enrichment of pathways in one set of tumors as the relative
-fraction of tumors (with estimated uncertainty) with a mutated gene in the
-pathway to all tumors, inclusive of the category of interest. To do this
-operation, we use the beta distribution to permute a posterior distribution on
-the fraction of tumors with a pathway mutated for each category based on the
-observed set, and compared this to the posterior obtained from the full set of
-tumors (all tumors) as the distribution of differences between all permuted
-samples. We considered a pathway enriched if the 95% range of credible
-differences thus obtained excludes 0 and the mean credible difference was
-greater than or equal to 30% enrichment, which excludes a large number of small
-differences that are not likely to be biologically relevant.
-
-### Models of survival based on population-specific longevity data
-
-We used Bayesian inference to explore the relative impact of cancer
-diagnosis on survival at the time of diagnosis. Our model assumes that
-at birth each person has a small initial probability per unit time to
-die, and that this probability grows exponentially in time at some
-baseline rate that matches the observed CDC data. The rate of growth
-in risk is assumed to jump to a higher level at diagnosis, which we
-take as a proxy for onset. This represents an accelerated aging type
-model. To fit the model, we constrain the baseline risks using the
-available CDC life tables
-(ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We
-also constrain the cancer specific multiplier using the longevity data
-within the cancer tumor dataset. Our model estimates the effect of
-cancer type and class-specific cancer effects independently, resulting
-in a cancer and class specific estimate of the aging rate multiplier.
-A detailed description of the CDC longevity model and likelihood is
-provided in the **Supplementary methods**.
+version control site, GitHub. Extended descriptions of clustering, models,
+formulae, pathway selection, and gene filtering are provided in supplementary
+methods.
 

### 2020-08-28 text/survival.Rmd
cut all text to conform to Cell's 45k character limit

@@
 ---
-If our molecular pathway disruption clusters represent biological states
+If our clusters represent biological states
 distinct from tissue of origin, they may have different prognoses within cancer
@@
 progression. To explore these ideas, we used Bayesian inference to test models
-(see Methods and **Supplemental Methods** for greater detail) of survival using
-public longevity data from the CDC, accounting for age, gender and ethnicity.
+of survival using
+public longevity data from the CDC.
 Our model estimates the effect of cancer type and cluster-specific cancer
@@
 
-To understand the model, we show in **Figure 7A** simulated survival time curves
+To illustrate the model, in **Figure 7A** we simulate survival time curves
 for four related groups. We start with a cohort of women aged 30 and show their
@@
 results.** To illustrate the risk model, we compare survival of four simulated
-cohorts. A) A group consisting of randomly selected 30 year old
-women (light blue) vs. an equal sized group of 30 year old women with cancer, k =
-2 (light orange). To illustrate how the mix of ages alters the survival curve,
-we compare a randomly selected group of people with ages uniformly distributed
-from 30 to 70 (dark blue) and a randomly selected group with uniformly
-distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
-selected group with correct age distribution for the given cancer (turqoise) to
-the model predicted survival for the given average k (blue) and the actual
-survival for the patients in our dataset (orange). As can be seen, Stomach
-Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
-deadly. The modeled results fit the general trend for the actual patients.
-Thyroid Cancer (THCA) on the other hand has very little if any effect on our
-estimates of life expectancy, in fact there may be a slight benefit which may be
-associated with a bias in diagnosis towards patients who are more health
-conscious, or higher socioeconomic status, with more access to care compared to
-random population members.
+cohorts. A) A group consisting of randomly selected 30 year old women (light
+blue) vs. an equal sized group of 30 year old women with cancer, k = 2 (light
+orange). To illustrate how the mix of ages alters the survival curve, we compare
+a randomly selected group of people with ages uniformly distributed from 30 to
+70 (dark blue) and a randomly selected group with uniformly distributed ages and
+cancer k = 2 (dark orange). B,C,D,E) We compare a randomly selected group with
+correct age distribution for the given cancer (turqoise) to the model predicted
+survival for the given average k (blue) and the actual survival for the patients
+in our dataset (orange). Differences with actual data are likely due to a
+mixture of cluster identities which we exclude from these simulations for
+simplicity.
 
@@
 We found that cancer types, as expected, have a range of prognoses relative to
-the general population. For example in **Figure 7B-E** we see three particularly
-deadly cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one
-cancer where diagnosis apparently decreases risk relative to background
-(Thyroid: THCA). Cancers with posterior probability for relative risk of less
-than 1 such as THCA should be interpreted carefully. This Bayesian model is a
-model for a state of information. The information that a person is diagnosed
-with cancer may lead us to expect that they will live a shorter time than the
-general population of matched age (k > 1), or a longer time than the general
-population of matched age (k < 1). One mechanism for a shorter result is that
-the cancer is aggressive and we can expect it to rapidly injure the body,
-causing death. One explanation for longer results may be that the cancer is
-relatively mild, and therefore diagnosis is potentially an informational signal
-that the patient is health conscious, with the comparison group having more
-people whose cancers go undiagnosed. It's important to note that the diagnosis
-can increase our expectation of life relative to the comparison group, even if
-it decreases the expectation of life of the individual relative to the
-counterfactual where they did not have cancer.
+the general population. In **Figure 7B-E** we see three particularly deadly
+cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
+where diagnosis apparently decreases risk relative to background (Thyroid:
+THCA). Cancers with posterior probability for relative risk of less than 1 such
+as THCA should be interpreted carefully. This Bayesian model is a model for a
+state of information. The information that a person is diagnosed with cancer may
+lead us to expect that they will live a shorter time than the general population
+of matched age (k > 1), or a longer time than the general population of matched
+age (k < 1). A shorter prognosis could result from the cancer being aggressive,
+injuring the body and causing death. By contrast, a longer prognosis may result
+from the cancer being relatively mild, and therefore diagnosis could be an
+informational signal that the patient is health conscious, with the comparison
+group having more people whose cancers go undiagnosed. It is important to note
+therefore that the diagnosis can increase our expectation of life relative to
+the comparison group, even if it decreases the expectation of life of the
+individual relative to the counterfactual where they did not have cancer.
 

### 2020-08-28 code/figure1v2.Rmd
Combine fig1+2, 3+4, 7+8

--- 2020-08-21 code/figure1v2.Rmd
@@
+---
+output:
+  word_document: default
+  html_document: default
+editor_options: 
+  chunk_output_type: console
+---
 ```{r, fig.width = 9, fig.height= 4}
@@
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,-10,0,-10))
+        legend.box.margin=margin(-10,-10,0,-10),
+        plot.margin = unit(c(0.6, -3, 0, 0), "cm"),
+        panel.background = element_rect(fill = "transparent"))
 
@@
 fig1 <- as.ggplot(fig1) ## turn back fig into 
-fig1
 ```
+
+```{r fig.width = 9, fig.height= 6}
+# start new page
+plot.new()
+
+# setup layout
+gl <- grid.layout(nrow=2, ncol=3)
+
+## Setup viewports - How figures will be laid out
+vp.1 <- viewport(layout.pos.col=1:3, layout.pos.row=1)
+vp.2 <- viewport(layout.pos.col=1, layout.pos.row=2) 
+vp.3 <- viewport(layout.pos.col=2, layout.pos.row=2) 
+vp.4 <- viewport(layout.pos.col=3, layout.pos.row=2)
+
+# init layout
+pushViewport(viewport(layout=gl))
+
+## vp.1 - Figure A - sankey plot
+pushViewport(vp.1) ## Enter vp.1 viewpoint
+print(fig1, newpage = FALSE) ## Plot ggplot
+mtext("A", side=3, line=3.15, cex=1.2, adj=-0.095)
+popViewport() ## Exit vp.1 viewpoint
+
+## Data for different perspectives of umap
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
+load("../data/kelly.colours.rda")
+cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
+names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
+
+theta <- c(0, 90, 270)
+phi <- c(65, 15, 15)
+## View 1
+pushViewport(vp.2)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+n <- 1
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
+                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                  colvar = NULL))
+title(xlab="UMAP 2", line=-1.5)
+mtext("B", side=3, line=-1.2, cex=1.2, adj=0)
+popViewport()
+## View 2
+pushViewport(vp.3)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+n <- 2
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
+                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                  colvar = NULL))
+popViewport()
+## View 3
+pushViewport(vp.4)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+n <- 3
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
+                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                  colvar = NULL))
+popViewport()
+
+```

### 2020-08-28 code/survival/survfig.Rmd
Combine fig1+2, 3+4, 7+8

@@
-```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=9, fig.width=9}
+```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=12, fig.width=9}
 
@@
     legend.title = element_blank(),
+    legend.position = c(0.7, 0.8),
     #legend.position = "none",
@@
 kmlayout = "
-AB
-AB
-AB
-CD
-CD
-CD
-EE
+AAFF
+AAFF
+AAFF
+BBFF
+BBFF
+BBFF
+CCFF
+CCFF
+CCFF
+DDFF
+DDFF
+DDFF
+EE##
 "
+
+BE_legend <- get_legend(thca[[1]] + theme(legend.direction = "horizontal"))
 kmlegend <- theme(legend.direction = "horizontal")
 
-print(
-  ((p1+mytheme)/((pstad[[1]]+mytheme + kmlegend)+(pgbm[[1]]+mytheme + kmlegend) +
-                   (ppaad[[1]]+mytheme + kmlegend) + (thca[[1]] + mytheme + kmlegend) + guide_area() + plot_layout(design = kmlayout, guides = "collect"))) + plot_annotation(tag_levels = "A")
-)
+  # ((p1+mytheme)/((pstad[[1]]+mytheme + kmlegend)+(pgbm[[1]]+mytheme + kmlegend) +
+  #                  (ppaad[[1]]+mytheme + kmlegend) + (thca[[1]] + mytheme) + (kmlegend) + (pcanc+mytheme) + guide_area() + plot_layout(design = kmlayout, guides = "collect"))) + plot_annotation(tag_levels = "A")
+layout = "
+AAAA
+AAAA
+AAAA
+AAAA
+AAAA
+AAAA
+BBFF
+BBFF
+CCFF
+CCFF
+DDFF
+DDFF
+EEFF
+EEFF
+"
+
+turn.off.legend <- theme(legend.position = "none")
+
+(p1+mytheme) + (pstad[[1]]+mytheme) + (pgbm[[1]]+mytheme + turn.off.legend) + (ppaad[[1]]+mytheme + turn.off.legend) + (thca[[1]] + mytheme + turn.off.legend) + (pcanc+mytheme) + plot_annotation(tag_levels = "A") + plot_layout(design = layout)
+
 ```

### 2020-08-28 text/supplemental_methods.Rmd

@@
 
+### Selection of pathways
+
+To understand the molecular mechanism of cancer at a pathway level, we used
+Reactome (https://reactome.org/), a knowledge-based pathway database. The
+mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
+complexes to top pathways were downloaded from
+https://reactome.org/download-data. Using these data, we imposed pathway
+criteria to define basic cellular processes and biochemical pathways: (1)
+human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
+parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
+Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
+"Muscle contraction", and "Reproduction" or pathway names that include any of
+the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
+"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
+excluded pathways have been shown to play an important role in cancer, they are
+highly specialized (_e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
+most of the excluded pathways, a neutral version pathway of the pathway exists
+(_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
+Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
+operation produced a lookup table that consisted of 377 pathways mapped to 8,940
+genes.
+
+### Filtering genes
+
+We filtered likely erroneous mutations due to transcription coupled repair. Our
+approach was to determine the status of each gene (_i.e._ expressed or not
+expressed) in each tissue type in order to exclude gene mutations low expressed
+enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
+effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
+the genes and tumor samples that were not included in our analysis, grouped the
+tumor samples by tissue type and computed the mean expression value for each
+gene. A minimum threshold of 10 transcripts per million was set for expressed
+genes based on an inflection point observed when plotting the mean expression
+values of genes ranked by expression in each tissue type. Genes that did not
+meet this threshold were considered not expressed. This operation produced a
+lookup table for gene expression status in each tumor sample for 18,127 genes.
+
+### Class and stage-specific Enrichment Calculations
+
+We calculated the enrichment of pathways in one set of tumors as the relative
+fraction of tumors (with estimated uncertainty) with a mutated gene in the
+pathway to all tumors, inclusive of the category of interest. To do this
+operation, we use the beta distribution to permute a posterior distribution on
+the fraction of tumors with a pathway mutated for each category based on the
+observed set, and compared this to the posterior obtained from the full set of
+tumors (all tumors) as the distribution of differences between all permuted
+samples. We considered a pathway enriched if the 95% range of credible
+differences thus obtained excludes 0 and the mean credible difference was
+greater than or equal to 30% enrichment, which excludes a large number of small
+differences that are not likely to be biologically relevant.
+
 ### Survival
 
-In order to understand how the diagnosis of cancer changes the expectation for
-longevity of patients, we first modeled the longevity of the overall population
-of the U.S. based on CDC life tables. Most TCGA cases are diagnosed late in
-life, so we fit a model that emphasizes accuracy in the right tail of the
-distribution. We modeled baseline longevity using a risk rate function:
+We used Bayesian inference to explore the relative impact of cancer diagnosis on
+survival at the time of diagnosis. Our model assumes that at birth each person
+has a small initial probability per unit time to die, and that this probability
+grows exponentially in time at some baseline rate that matches the observed CDC
+data. The rate of growth in risk is assumed to jump to a higher level at
+diagnosis, which we take as a proxy for onset. This represents an accelerated
+aging type model. To fit the model, we constrain the baseline risks using the
+available CDC life tables (ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We also constrain the cancer specific multiplier
+using the longevity data within the cancer tumor dataset. Our model estimates
+the effect of cancer type and class-specific cancer effects independently,
+resulting in a cancer and class specific estimate of the aging rate multiplier.
+
+Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes
+accuracy in the right tail of the distribution. We modeled baseline longevity
+using a risk rate function:
 

### 2020-08-28 text/manuscript.Rmd
Combine fig1+2, 3+4, 7+8
Update figure legend text

@@
 
-**Figure 1: Molecular pathway profiles of tumor samples with one or more
-mutations.** Each of 377 selected Reactome pathways (rows) is classified as
+**Figure 1: Clustering the molecular pathway profiles of tumor samples with one or more
+mutations.** A) Each of 377 selected Reactome pathways (rows) is classified as
 disrupted if one or more genes is mutated in the tumor sample (columns). Red
 cells denote pathway disruption. Tumors types reflect standardized abbreviations
-from the TCGA project.
+from the TCGA project. B) Different rotational perspectives of
+the same MCA-based UMAP projection in three-dimensional space. Each dot
+corresponds to a tumor sample. The same colors indicate the tumor's cluster
+identity throughout this manuscript.
 
@@
 kNN (see **Supplementary Methods** for details on clustering methods).
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 2
-```{r, child='../code/figure2.Rmd'}
-```
-
-**Figure 2: Clustering of tumor samples** Different rotational perspectives of
-the same MCA-based UMAP projection in three-dimensional space. Each dot
-corresponds to a tumor sample. The same colors indicate the tumor's cluster
-identity throughout this manuscript.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
 
@@
 regardless of cancer type. E) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.
+Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors. F) Somatic mutation 
+frequencies for each cluster. Vertical axis shows log mutation count, horizontal 
+axis is cluster identity. Each dot represents an individual tumor sample, ranked 
+lowest to highest by mutation count. The median mutation count in each cluster is 
+indicated by the horizontal line.
 
@@
 not a contributing factor to cluster identity.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 4
-```{r child = '../code/figure4.Rmd'}
-```
-
-**Figure 4: Somatic mutation frequencies for each cluster** 
-Vertical axis shows log mutation count, horizontal axis is cluster identity.
-Each dot represents an individual tumor sample, ranked lowest to highest by
-mutation count. The median mutation count in each cluster is indicated by the
-horizontal line.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
 
@@
 
-```{r, results='asis'}
-cat("\\newpage")
-```
-
 ### Pathway disruption clusters vary in short-term prognosis of survival

### 2020-08-28 text/survival.Rmd
Combine fig1+2, 3+4, 7+8
Update figure legend text

@@
----
-output:
-  word_document: default
-  html_document: default
----
 If our clusters represent biological states
@@
 mixture of cluster identities which we exclude from these simulations for
-simplicity.
+simplicity. F) Effects are compared between cancer type by
+showing the probability density for the posterior value of the $k_{tis}$
+multiplier. Larger values correspond to decreased age-independent life
+expectancy.
 
@@
 about 1 to 5 in magnitude.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 8
-```{r fig.width=3, message=FALSE, warning=FALSE}
-pcanc+mytheme
-```
-
-**Figure 8 Tissue specific effects** Effects are compared between cancer type by
-showing the probability density for the posterior value of the $k_{tis}$
-multiplier. Larger values correspond to decreased age-independent life
-expectancy.
 

### 2020-08-28 text/manuscript.Rmd
fix textual figure references; reflow text

@@
 (MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
-2** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
+1B** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
 We used the resulting UMAP graph coordinates to perform
@@
 cluster, and all well-represented cancer types had tumors in every cluster (see
-**figure 3A** and full tumor profiles in **supplementary figure S1**, see also
+**figure 2A** and full tumor profiles in **supplementary figure S1**, see also
 interactive media from **supplementary file 1**), suggesting that our
@@
 pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
-cluster 8 (**figure 3A** and **supplementary figure S1** and **supplementary
+cluster 8 (**figure 2A** and **supplementary figure S1** and **supplementary
 file 1**). But even PAAD comprises tumors from the nine remaining classes,
@@
 
-#### Figure 3
+#### Figure 2
 ```{r child = '../code/figure3.Rmd'}
@@
 
-**Figure 3: Pathway-based clustering independent of tissue-of-origin** 
+**Figure 2: Pathway-based clustering independent of tissue-of-origin** 
 A) Sankey plot showing correspondence between cancer type and cluster identity.
@@
 for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe segregation of
-these mutations into specific clusters (**figure 3C** and see interactive media
+these mutations into specific clusters (**figure 2C** and see interactive media
 from [**supplemental file
@@
 common pathway disruption profiles. The UMAP, which features a series of
-lobe-like structures on a common backbone of tumor samples (**figure 2**) could
+lobe-like structures on a common backbone of tumor samples (**figure 1B**) could
 in principle reflect progression through a series of stages. The backbone starts
@@
 culminates in a cluster (class 10) which has nearly every pathway disrupted
-(**figure 4**). However, outside of class 10 we don't observe a trend in the
+(**figure 2F**). However, outside of class 10 we don't observe a trend in the
 overall mutation burden across the backbone of the UMAP. Nonetheless, to test
@@
 tissue of origin and other categories of tumor, we did not observe any bias
-among the stages to specific clusters (**figure 3D**), suggesting that stage is
+among the stages to specific clusters (**figure 2D**), suggesting that stage is
 not a contributing factor to cluster identity.
@@
 tissue types, we projected the metastases onto our UMAP, and surprisingly
-we observed even distribution of the samples across clusters (**figure 3E**).
+we observed even distribution of the samples across clusters (**figure 2E**).
 This final observation suggests that our pathway-disruption clustering is
@@
 and ranked pathways from highest to lowest enrichment. We visualized the
-enrichment as a heatmap (**figure 5A**). Using this approach, we identified
+enrichment as a heatmap (**figure 3A**). Using this approach, we identified
 about fourteen pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see
@@
 Clusters 7-9 have in common mutations in extracellular, intracellular, and
-immune-related signaling pathways (see **figure 5B** and **supplementary figure
+immune-related signaling pathways (see **figure 3B** and **supplementary figure
 S4**). Cluster 2 had the highest pathway enrichment levels of the three, having
@@
 
-#### Figure 5
+#### Figure 3
 ```{r child = '../code/figure5.Rmd'}
@@
 
-**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
+**Figure 3: Pan-cancer enrichment of pathway disruptions.** 
 A) Heatmap shows the relative enrichment of each pathway (rows) within each
@@
 selected the top ten enriched and depleted genes (pvalue < 0.01) for each
-cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
+cluster (**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared
 multiple enrichment in signaling pathways, are largely driven by mutations in
@@
 depleted in clusters 3 and 8). Next, we investigated the proportion of samples
-per cancer type for the significant genes within a cluster (**Figure 6**;
+per cancer type for the significant genes within a cluster (**Figure 4**;
 heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
@@
 
-#### Figure 6
+#### Figure 4
 ```{r child = '../code/figure6.Rmd'}
@@
 
-**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
+**Figure 4: Gene level analysis reveals tissue-specific class signatures.** 
 Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a

### 2020-08-28 text/survival.Rmd
fix textual figure references; reflow text

@@
-If our clusters represent biological states
-distinct from tissue of origin, they may have different prognoses within cancer
-types or across all cancers. These analyses are limited by confounding factors
-of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
-progression. To explore these ideas, we used Bayesian inference to test models
-of survival using
-public longevity data from the CDC.
-Our model estimates the effect of cancer type and cluster-specific cancer
-effects independently, resulting in a cancer and cluster-specific estimate of
-the aging rate multiplier, $k$.
+If our clusters represent biological states distinct from tissue of origin, they
+may have different prognoses within cancer types or across all cancers. These
+analyses are limited by confounding factors of age and stage at diagnosis, sex,
+ethnicity, and tissue-specific disease progression. To explore these ideas, we
+used Bayesian inference to test models of survival using public longevity data
+from the CDC. Our model estimates the effect of cancer type and cluster-specific
+cancer effects independently, resulting in a cancer and cluster-specific
+estimate of the aging rate multiplier, $k$.
 
-To illustrate the model, in **Figure 7A** we simulate survival time curves
+To illustrate the model, in **Figure 5A** we simulate survival time curves
 for four related groups. We start with a cohort of women aged 30 and show their
@@
 
-#### Figure 7
+#### Figure 5
 ```{r child='../code/survival/survfig.Rmd'}
@@
 
-**Figure 7: Comparison of different cancers, simulated and actual results
+**Figure 5: Comparison of different cancers, simulated and actual results
 results.** To illustrate the risk model, we compare survival of four simulated
@@
 mixture of cluster identities which we exclude from these simulations for
-simplicity. F) Effects are compared between cancer type by
-showing the probability density for the posterior value of the $k_{tis}$
-multiplier. Larger values correspond to decreased age-independent life
-expectancy.
+simplicity. F) Effects are compared between cancer type by showing the
+probability density for the posterior value of the $k_{tis}$ multiplier. Larger
+values correspond to decreased age-independent life expectancy.
 
@@
 We found that cancer types, as expected, have a range of prognoses relative to
-the general population. In **Figure 7B-E** we see three particularly deadly
+the general population. In **Figure 5B-E** we see three particularly deadly
 cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
@@
 
-The degree to which a given cancer accelerates aging can be determined by
-the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier $k_{cl}$
-(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
+The degree to which a given cancer accelerates aging can be determined by the
+$k_{tis}$ multiplier (**Figure 5F**) multiplied by the cluster modifier $k_{cl}$
+(**Figure 6**). Looking at the tissue specific multiplier, the least deadly was
 prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
@@
 
-#### Figure 9
+#### Figure 6
 
@@
 
-**Figure 9 Cluster specific $k_{cl}$ values for each cancer type.** The overall
+**Figure 6 Cluster specific $k_{cl}$ values for each cancer type.** The overall
 $k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
@@
 than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
-or more deadly than other clusters within each cancer type (**Figure 9**). A
+or more deadly than other clusters within each cancer type (**Figure 6**). A
 cluster-specific rate of 1 represents the typical rate for this tissue type. For

### 2020-08-28 code/figure2.Rmd
Update text in manuscript and file names

--- 2020-08-24 code/figure2.Rmd
@@
-```{r figure-2-plot, fig.width = 9}
-umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
+```{r sankey-plot}
+umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
 load("../data/kelly.colours.rda")
-cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
-names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
+load("../data/ditto.colours.rda")
 
-theta <- c(0, 90, 270)
-phi <- c(65, 15, 15)
+sankey.data <- umap_cancertype %>% dplyr::select(sample_id, project_code, clust_knn) %>% ## select relevant columns
+  gather(., key = "type", value = "value", -sample_id) ## wide to long format
 
-opar <- par(mfrow = c(1,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
-par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots
+## Cancer names in alphabetical order
+cancerorder <- unique(umap_cancertype$project_code)
+cancerorder <- cancerorder[order(cancerorder)]
+## Cluster in numerical order
+clustorder <- unique(umap_cancertype$clust_knn)
+clustorder <- as.character(clustorder[order(clustorder)])
 
-for(n in 1) {
-  with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
-                                  theta = theta[n], phi = phi[n], scale = T,
+## Factor to force ordering
+sankey.data$type <- factor(sankey.data$type, levels = c( "clust_knn", "project_code"))
+sankey.data$value <- factor(sankey.data$value, levels= c(cancerorder, clustorder))
+
+## Color scheme for sankey plot
+sankey.colours <- c(ditto_colours[2:24],kelly.colours[3:12])
+names(sankey.colours) <- c(cancerorder, clustorder)
+
+## Sankey plot
+sankey.plot <- ggplot(sankey.data,
+       aes(x = type, stratum = value, alluvium = sample_id,
+           fill = value, label = value)) +
+  scale_fill_manual(values = sankey.colours) +
+  geom_flow(stat = "alluvium", lode.guidance = "frontback") +
+  geom_stratum(alpha = 0.5) +
+  geom_text(stat = "stratum", size = 3) +
+  theme_classic() +
+  xlab(label = "") +
+  theme(legend.position = "none",
+        axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line.x = element_blank(),
+        axis.line.y = element_blank(),
+        plot.margin = unit(c(-0.4, -3, -1, -2.5), "cm"),
+        panel.background = element_rect(fill = "transparent"))
+```
+
+```{r, message = FALSE}
+
+## Arranging mutation freq for each cluster
+mut_order <- list() ## store data in list 
+for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
+  mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
+    dplyr::select(sample_id, plot_x, plot_y, clust_knn, mut_count) %>% ## select relevant columns
+    arrange(mut_count) ## arrange mutation in increasing order
+}
+mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
+
+## Convert sample_id and clust_knn into factor
+mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
+mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10))
+
+## Determining the breaks positions for the x-axis
+previous_class <- 0 ## starting position 1
+position_of_previous_class <- 0 ## starting position 2
+xaxis_breaks <- list() ## list for the for loop
+for(n in 1) { ## loop through 1 to 10 clusters
+  class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
+  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+  previous_class <- class_spec ## storing class halfway mark
+  for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
+    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
+    class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
+    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
+    previous_class <- class_spec ## storing class halfway mark
+  }
+}
+xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
+xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
+
+total_n <- as.data.frame(table(mut_order$clust_knn)) ## getting total n in cluster
+
+mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count
+
+## Calculate median for each class
+mut_median <- mut_order %>% 
+  group_by(clust_knn) %>% 
+  dplyr::summarize(mut_med = median(log10mut))
+
+## Need to add 100 pseudo samples front and end of data so they don't touch the y-axis
+mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column
+
+begin_NA <- data.frame(sample_id = paste0("sample", 1:100), 
+                       clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
+                       log10mut = NA) ## put NA values so they don't get plotted
+end_NA <- data.frame(sample_id = paste0("sample", 101:200), ## repeat the step for the ends
+                       clust_knn = max(as.numeric(mut_order$clust_knn)),
+                       log10mut = NA)
+mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
+
+mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id) ## factor samples
+mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10)) ## factor clusters
+
+fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = clust_knn)) + 
+  geom_point(size = 0.25) + 
+  ylab("Log10(Mutation Count)") + 
+  xlab("") +
+  theme_classic() +
+  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") + ## facet by cluster
+  geom_hline(data= mut_median, aes(yintercept=mut_med)) + ## add median line
+  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
+  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[1]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
+  theme(axis.text.x = element_text(size = 10),
+        axis.text.y = element_text(size = 10),
+        axis.title.x = element_text(size = 10),
+        axis.title.y = element_text(size = 10),
+        axis.ticks.x = element_blank(),
+        axis.line.x = element_line(colour = "black", size = 0.5),
+        axis.line.x.top = element_line(colour = "black", size = 0.5),
+        panel.spacing = unit(0.25, "lines"),
+        strip.background = element_rect(size = 0.5),
+        strip.placement = "outside",
+        strip.text.x = element_text(size = 10),
+        legend.position = "none") +
+  scale_color_manual(values = kelly.colours[-c(1:2)])
+
+## code to keep only the bottom border line of the facet
+fig1G <- ggplotGrob(fig1G)
+
+lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), ## All lines off expect bottom line of facet
+                   gp=gpar(col="black", lwd=4))
+
+for (k in grep("strip-t",fig1G$layout$name)) { ## turn off 3 lines from the facet border text
+  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
+}
+
+fig1G <- as.ggplot(fig1G)
+```
+
+```{r, fig.width= 6, fig.height = 8}
+# start new page
+plot.new()
+
+# setup layout
+gl <- grid.layout(nrow=11, ncol=3)
+
+## Setup viewports - How figures will be laid out
+vp.1 <- viewport(layout.pos.col=1:2, layout.pos.row=1:8)
+vp.2 <- viewport(layout.pos.col=3, layout.pos.row=1:2) 
+vp.3 <- viewport(layout.pos.col=3, layout.pos.row=3:4) 
+vp.4 <- viewport(layout.pos.col=3, layout.pos.row=5:6) 
+vp.5 <- viewport(layout.pos.col=3, layout.pos.row=7:8)
+vp.6 <- viewport(layout.pos.col=1:3, layout.pos.row=9:11)
+
+# init layout
+pushViewport(viewport(layout=gl))
+
+## vp.1 - Figure A - sankey plot
+pushViewport(vp.1) ## Enter vp.1 viewpoint
+print(sankey.plot, newpage = FALSE) ## Plot ggplot
+mtext("A", side=3, line=2.6, cex=1, adj=-0.17)
+popViewport() ## Exit vp.1 viewpoint
+
+## vp.2 - Figure B - BRCA subtype
+pushViewport(vp.2)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
+brca_subtype <- umap_cancertype %>% filter(project_code == "BRCA") %>% ## Filter for selected cancer type
+  dplyr::select(plot_x, plot_y, plot_z, Subtype_Selected) ## Select relevant column 
+brca_subtype[is.na(brca_subtype)] <- "NA" ## Convert NA to character "NA"
+
+## Color scheme for subtype
+B_name <- unique(brca_subtype$Subtype_Selected) ## What are the subtype names
+B_name <- B_name[order(B_name,na.last=T)] ## Order the names ABC and have NA last
+B_color <- ditto_colours[c(2:(length(B_name) + 1))] ## Selecting enough colors from ditto
+names(B_color) <- B_name ## Assigning subtypes to color
+
+with(brca_subtype,
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = B_color, pch = 21, cex = 0.8, lwd = 0.2, 
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                colvar = NULL))
-  title(xlab="UMAP 2", line=-5.2)
-  for(p in 2:length(phi)) {
-    with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, 
-                                theta = theta[p], phi = phi[p], scale = T,
+## Axis labels
+legend("top", inset = c(0), legend=names(B_color[1:2]), col=B_color[1:2], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.07), legend=names(B_color[3:4]), col=B_color[3:4], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.14), legend=names(B_color[5:6]), col=B_color[5:6], pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("B", side=3, line=-1.2, cex=1, adj=0)
+
+popViewport() ## exit
+
+## vp.3 - Figure C - BRCA somatic mutations
+pushViewport(vp.3)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
+load("../data/gdc_gene_exp_adj.Rda") ## load data
+load("../data/gene_data.Rda")
+
+brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
+brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## Include BRCA gene
+  rownames_to_column(var = "ensembl_gene_id") %>% ## Move ensembl id to column
+  left_join(., brca, by = "ensembl_gene_id") %>% ## left join with brca df 
+  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace row with gene symbol
+  dplyr::select(-ensembl_gene_id) %>% ## get rid of ensembl_id col
+  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
+  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
+  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1", ## Convert booleans to categorical variables
+                           BRCA1 == "0" ~ "NA"),
+         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
+                           BRCA2 == "0" ~ "NA")) %>%
+  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
+  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
+  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "plot_z", "project_code")], by = "sample_id") ## leftjoin by sample id
+
+C_colors <- c('#33a02c', '#e31a1c', '#ff7f00') ## Color scheme 
+names(C_colors) <- c("BRCA1", "BRCA2", "BRCA1,BRCA2") ## Assigning mutation type to color
+with(brca %>% filter(mut_type != "NA"), ## Filter samples that do not have a mutation in BRCA1 or BRCA2
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = C_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                colvar = NULL))
-  }
-}
+## Axis labels
+legend("top", inset = c(0), legend=names(C_colors[c(1:2)]), col=C_colors[c(1:2)], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.07), legend=names(C_colors[c(3)]), col=C_colors[3], pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("C", side=3, line=-1.2, cex=1, adj=0)
 
+popViewport()
+
+## vp.4 - Figure D - Stage
+pushViewport(vp.4)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
+D_colors <- c('darkgoldenrod3','dodgerblue3','yellow3','tomato2') ## Color scheme
+names(D_colors) <- c("Stage I", "Stage II", "Stage III", "Stage IV") ## Assign stage to color
+with(umap_cancertype %>% filter(ajcc_pathologic_tumor_stage != "Not Available"), ## Filter samples that do not have stage info
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = D_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=names(D_colors[1:2]), col=D_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.07), legend=names(D_colors[3:4]), col=D_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("D", side=3, line=-1.2, cex=1, adj=0)
+
+popViewport()
+
+## vp.5 - Figure E - Metastasis
+pushViewport(vp.5)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+
+E_colors <- c('#a6cee3','#1f78b4') ## Color scheme
+names(E_colors) <- c("M0", "M1") ## Assign metastasis to color
+with(umap_cancertype %>% filter(ajcc_metastasis_pathologic_pm %in% c("M0", "M1")), ## Only M0 and M1 tumor samples
+     scatter3D(plot_y, -plot_z, plot_x, 
+               bg = E_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               theta = 0, phi = 65, scale = F,
+               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+               colvar = NULL))
+## Axis labels
+legend("top", inset = c(0), legend=names(E_colors), col=E_colors, pch = 20, bty = "n", horiz = T, cex = 0.8)
+title(xlab="UMAP 2", line=-0.25)
+mtext("E", side=3, line=-1.2, cex=1, adj=0)
+
+popViewport()
+
+## vp.1 - Figure A - sankey plot
+pushViewport(vp.6) ## Enter vp.1 viewpoint
+print(fig1G, newpage = FALSE) ## Plot ggplot
+mtext("F", side=3, line=-9.8, cex=1, adj=-2.4)
+popViewport() ## Exit vp.1 viewpoint
 ```

### 2020-08-28 code/figure3.Rmd
Update text in manuscript and file names

--- 2020-07-21 code/figure3.Rmd
@@
-```{r group-pathway-enrichment-data-generation, eval=FALSE}
-tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
-tumor_clusters <- tumor_clusters %>% as.data.frame() %>% dplyr::select(sample_id, clust_knn) ## select the relevant columns
+---
+output: html_document
+editor_options: 
+  chunk_output_type: console
+---
+```{r figure5-data-generation, eval = FALSE}
+### Figure A - Pathway enrichment by cluster
 
 ## Getting the total number of tumor samples in each cluster
-n_tumor_in_clust <- as.data.frame(table(tumor_clusters$clust_knn)) %>% ## table the number of tumors in each cluster
+n_tumor_in_clust <- as.data.frame(table(umap_cancertype$clust_knn)) %>% ## table the number of tumors in each cluster
   'colnames<-' (c("clust", "tot_n_clust")) %>% ## move cluster number to rownames
@@
 for(n in clust_group) {
-  print(n)
-  sample_id <- tumor_clusters %>% filter(clust_knn == n) %>% ## filter for tumor class
-    .[,"sample_id"] ## get sample_ids
+  sample_id <- umap_cancertype %>% filter(clust_knn == n) %>% ## filter for tumor class
+    .[,.$sample_id] ## get sample_ids
   path_mutation <- gdc_reactome_path_analysis %>% .[,names(.) %in% sample_id] %>% ## getting the tumor samples from path analysis
@@
 }
-cluster_path_count <- do.call(cbind, cluster_path_count) ## Combine the list together
+cluster_path_count <- as.data.frame(do.call(cbind, cluster_path_count)) ## Combine the list together
 
 global_path_mut <- cluster_path_count %>% rowSums() %>% as.data.frame() %>% ## global pathway mutation count
-  .[rep(1:ncol(.), each = max(tumor_clusters$clust_knn))] %>% ## replicate the column 0 times for the 10 tumor classes
+  .[rep(1:ncol(.), each = max(umap_cancertype$clust_knn))] %>% ## replicate the column 0 times for the 10 tumor classes
   'colnames<-' (clust_group) %>% ## rename column with tumor class
@@
                                             )
+
+## Calculating the foreground and background mutation rate for each cluster
 fg_mut_rate <- list()
@@
 for(n in 1:max(as.numeric(unique(enrich_prob_uncertainity_in_bg$clust)))) {
-  print(n)
-  df <- subset(enrich_prob_uncertainity_in_bg, clust == n)
-  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust_x
-  fg_numerator <- sum(df$fg_mut)
-  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1])
-  bg_numerator <- sum(df$bg_mut)
-  fg_mut_rate[[n]] <- fg_numerator / fg_denominator
+  df <- subset(enrich_prob_uncertainity_in_bg, clust == n) ## Only include pathways in selected cluster
+  fg_denominator <- nrow(df) * (df$fg_mut[1] + df$fg_not_mut[1]) ## n_path * tot_n_clust
+  fg_numerator <- sum(df$fg_mut) ## sum of pathway mutation in selected cluster
+  bg_denominator <- nrow(df) * (df$bg_mut[1] + df$bg_not_mut[1]) ## n_path * tot_n_clust
+  bg_numerator <- sum(df$bg_mut) ## sum of no pathway mutation in selectedcluster
+  fg_mut_rate[[n]] <- fg_numerator / fg_denominator ## calculate mutation rate and store in list
   bg_mut_rate[[n]] <- bg_numerator / bg_denominator
@@
 enrich_prob_uncertainity_in_bg <- enrich_prob_uncertainity_in_bg %>% left_join(., fg_mut_rate, by = "clust") %>% 
-  left_join(., bg_mut_rate, by = "clust") %>%
-  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>%
+  left_join(., bg_mut_rate, by = "clust") %>% ## add fg and bg mut rate to enrich df
+  mutate(rel_odds = .$fg_mut_rate / .$bg_mut_rate) %>% ## calculating relative odds ratio
   mutate(denominator = (.$bg_mut * .$rel_odds) + (.$bg_not_mut / .$rel_odds)) %>%
@@
 
+## beta distribution for 1e5 simulations for foreground and background
 moderate_fg <- apply(enrich_prob_uncertainity_in_bg[,c("fg_mut", "fg_not_mut")], MARGIN = 1, function(x) {
@@
 mean_prob <- apply(prob, MARGIN = 2, function(x) {
-  mean(x)
+  mean(x) ## calculating the mean probability
 })
@@
 
-diff_fg_bg <- moderate_fg - moderate_bg ## fg - bg
+diff_fg_bg <- moderate_fg - moderate_bg ## Calculating the difference between fg and bg for the next function 
 rm(moderate_fg, moderate_bg)
@@
 quant_diff <- t(apply(diff_fg_bg, MARGIN = 2, function(x) {
-  quantile(x, c(CI, 0.5, 1-CI))
+  quantile(x, c(CI, 0.5, 1-CI)) ## calculating the 1st quantile, median, and 3rd quantile
 }))
@@
                                    clust = enrich_prob_uncertainity_in_bg$clust,
-                                   probability = mean_prob,
-                                   signif = sig_fg_bg[,1],
-                                   difference = quant_diff[,2],
-                                   lower = quant_diff[,1],
-                                   upper = quant_diff[,3])
+                                   probability = mean_prob, ## probabilty of pathway mutation
+                                   signif = sig_fg_bg[,1], ## how many times fg-bg is less than 0
+                                   difference = quant_diff[,2], ## effect size
+                                   lower = quant_diff[,1], ## 1st quantile
+                                   upper = quant_diff[,3]) ## 3rd quantile
 rm(enrich_prob_uncertainity_in_bg, mean_prob, quant_diff, sig_fg_bg, diff_fg_bg)
 gc()
-save(moderate_enrich_prob, file = "../data/moderate_enrich_prob.Rda")
-load("../data/moderate_enrich_prob.Rda")
+save(moderate_enrich_prob, file = "../data/moderate_enrich_prob.Rda") ## save result
 
-diff_heatmap <- moderate_enrich_prob[,c("path", "clust", "difference")]
-diff_heatmap <- spread(diff_heatmap, key = "clust", value = "difference")
-diff_heatmap <- diff_heatmap %>% remove_rownames() %>% column_to_rownames(var="path")
+## Processing data for effect size heatmap
+diff_heatmap <- moderate_enrich_prob %>% dplyr::select(path, clust, difference) %>% ## select relevant column
+  spread(., key = "clust", value = "difference") %>% ## long to wide format
+  remove_rownames() %>% column_to_rownames(var = "path") ## pathway names to rownames
 
+## Determining pathway order for heatmap
 effect_size <- pheatmap(diff_heatmap,
-         show_rownames = F,
-         show_colnames = T,
          cluster_cols = T,
          cluster_rows = T,
-         cellwidth = 10,
-         cellheight = 0.76,
-         color = rev(c('#7f3b08','#b35806','#e08214','#fdb863','#fee0b6','white', '#d8daeb','#b2abd2','#8073ac','#542788','#2d004b')),
-         breaks = c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)),
          clustering_method = "ward.D2")
+## Extracting row order from pheatmap
+effect_row_order <- effect_size$tree_row$order
+effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames()
 
-effect_row_order <- effect_size$tree_row$order
-effect_row_order <- diff_heatmap %>% .[effect_row_order,] %>% rownames() ## getting row order of pheatmap
+## Revert df back to long format
+effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>%  ## pathway names to column
+  gather(., key = "cluster", value = "effect_size", -path_name)  ## wide to long
 
-effect_heatmap <- diff_heatmap %>% rownames_to_column(var = "path_name") %>% 
-  gather(., key = "cluster", value = "effect_size", -path_name)
+## Factor column to force order when generating plot
+effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))  ## pathway name ordering
+effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:max(as.numeric(effect_heatmap$cluster)))) ## cluster ordering
+effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2)) ## Creating breaks for heatmap legend (color scale)
 
-effect_heatmap$path_name <- factor(effect_heatmap$path_name, levels = c(rev(effect_row_order)))
-effect_heatmap$cluster <- factor(effect_heatmap$cluster, levels = c(1:max(as.numeric(effect_heatmap$cluster))))
-effect_segments <- c(seq(from = -1, -0.2, 0.2), -0.0001, 0.0001, seq(from = 0.2, to = 1, by = 0.2))
-
-effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when(
+effect_heatmap <- effect_heatmap %>% mutate(discrete_color = case_when( ## create discrete variable using continuous
   effect_size > effect_segments[1] & effect_size <= effect_segments[2] ~ "1",
@@
 ))
-effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11))
+effect_heatmap$discrete_color <- factor(effect_heatmap$discrete_color, levels = c(1:11)) ## Factor order the discrete color
 
 save(effect_heatmap, file = "../data/effect_heatmap.Rda")
-rm(list = ls(all.names = T))
-gc()
+
+### Figure C - barplot panel
+
+enrich_pathways <- moderate_enrich_prob %>%
+  group_by(clust) %>% 
+  filter(difference > 0.3 & signif <= 2500) %>% ## Filter pathways that have an effect size > 0.30 with a 95% CI (signif 2500/1e5)
+  mutate(pathway = make_clean_names(path)) %>% 
+  rename(clust_knn = "clust") %>%  ## rename clust column to clust_knn
+  select(clust_knn, pathway) %>% ## select relevant columns
+  mutate(clust_knn = as.numeric(clust_knn)) %>%
+  mutate(specific = TRUE) ## Add "specific" column with values TRUE
+
+pathway_counts <- umap_cancertype %>%
+  left_join(t(gdc_reactome_path_analysis) %>% ## combine umap_cancertype and gdc_path_analysis by sample_id
+              as.data.frame() %>% 
+              clean_names() %>% 
+              rownames_to_column("sample_id")) %>% 
+  group_by(clust_knn) %>% 
+  summarise(total = n(), across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, sum)) %>% ## Calculate the total_n_clust and total pathway mutation across cluster
+  mutate(across(intrinsic_pathway_for_apoptosis:signaling_by_the_b_cell_receptor_bcr, ~ .x / total)) %>% ## Divide sum mutation of cluster / total_n_clust to get percent mutation
+  select(-total) %>% ## Remove total column
+  pivot_longer(-clust_knn, names_to = "pathway", values_to = "proportion") %>% ## wide to long format
+  group_by(clust_knn) %>% 
+  filter(pathway %in% enrich_pathways$pathway) %>% ## only include pathways that are enriched
+  left_join(enrich_pathways %>% select(clust_knn, pathway, specific), by = c("clust_knn", "pathway"))
+
+pathway_counts[is.na(pathway_counts$specific), ]$specific <- FALSE ## Clusters not enriched for a pathway (NA) convert to FALSE
+
+## Determine the order of pathways on the barplot
+pathway_mat <- pathway_counts %>%
+  select(-specific) %>%
+  pivot_wider(names_from = pathway, values_from = proportion) %>% ## long to wide format
+  ungroup() %>%
+  select(-clust_knn) %>%
+  as.matrix()
+
+pathway_dist <- dist(t(pathway_mat)) ## Computing distance matrix of all values
+pathway_clust <- as.hclust(rev(as.dendrogram(seriate(pathway_dist, method = "OLO_ward")[[1]]))) ## cluster pathways based on distance matrix
+pathway_counts$pathway <- factor(pathway_counts$pathway, levels = colnames(pathway_mat)[pathway_clust$order]) ## force pathway order by factoring
+
+bp.cluster.colors <- c(kelly.colours[3:12], "gray15") ## Color scheme for plot
+names(bp.cluster.colors) <- 1:11 ## Cluster 11 is
+
+pathway_counts1 <- pathway_counts %>%
+  mutate(color = case_when(specific ~ clust_knn, TRUE ~ 11)) %>% ## Color: specific == FALSE is 11; specific == TRUE is its cluster
+  mutate(color = factor(color)) 
+save(pathway_counts, bp.cluster.colors, file = "barplot_pathways_fig2b.rda")
 ```
 
-```{r group-pathway-enrichment heatmap, fig.width= 5, fig.height= 10, fig.align= 'center', fig.cap="**Figure 3.
-Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns).
-The annotation at left indicates the Reactome pathway group at the second level (check this)."}
+```{r figure-5-plot, fig.align= 'center', fig.height= 7, fig.width = 7}
+### Figure A - Effect size heatmap
 load("../data/effect_heatmap.Rda") ## load data
 
-effect_colors = c("1" = "#2d004b", "2" = "#542788", "3" = "#8073ac", "4" = "#b2abd2", "5" = "#d8daeb", "6" = "white", "7" = "#fee0b6", "8" = "#fdb863", "9" = "#e08214","10" = "#b35806","11" ="#7f3b08") ## color scheme
+## Function to duplicate discrete axis labels
+guide_axis_label_trans <- function(label_trans = identity, ...) {
+  axis_guide <- guide_axis(...)
+  axis_guide$label_trans <- rlang::as_function(label_trans)
+  class(axis_guide) <- c("guide_axis_trans", class(axis_guide))
+  axis_guide
+}
 
-fig3 <- ggplot(data = effect_heatmap, aes(x=cluster, y=path_name)) + 
-  geom_tile(size = 0.6, aes(fill = discrete_color)) +
-  scale_fill_manual(values = effect_colors) +
-  ylab("377 Reactome Pathways") +
-  xlab("Cluster") +
-  theme(axis.text.x = element_text(size = 10),
+guide_train.guide_axis_trans <- function(x, ...) {
+  trained <- NextMethod()
+  trained$key$.label <- x$label_trans(trained$key$.label)
+  trained
+}
+
+fig5A <- ggplot(data = effect_heatmap, aes(x = cluster, y = path_name)) +
+  geom_tile(size = 1, aes(fill = effect_size)) +
+  scale_fill_fermenter( ## custom color scale
+    type = "div",
+    palette = "PuOr",
+    n.breaks = 12,
+    limits = c(-1, 1), ## Set scale range
+    name = "Effect Size",
+    labels = c(-1, "", 0.5, "", "", 0, "", "", 0.5, "", 1)
+  ) +
+  #facet_wrap(cluster~., ncol = 1, strip.position = "right", scales = "free_y") + ## facet by cluster
+  facet_grid(.~cluster, scales = "free", space = "free", switch = "x") +
+  guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
+  ylab("377 pathways, ward.D2 clustered") + xlab("Clusters") + 
+  scale_x_discrete(position = "bottom") + ## move x labels to top
+  guides(y.sec = guide_axis_label_trans(~ paste(.x))) + ## Duplicate axis.text on right side (cluster)
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank(), ## turn off y-axis text
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    axis.ticks = element_blank(),
+    strip.text.y = element_text(size = 10, hjust = .5),
+    strip.text.y.right = element_text(angle = 0), ## Fix orientation of dupped right y-axis text
+    strip.background = element_blank(),
+    panel.background = element_blank(),
+    panel.spacing = unit(0, "lines"),
+    ## legend parameters
+    legend.position = "top",
+    legend.key.height = unit(0.6, "lines"),
+    #legend.key.width = unit(0.6, "lines"),
+    legend.text = element_text(size = 7),
+    legend.title = element_text(size = 7),
+    legend.margin=margin(0,0,0,0),
+    legend.box.margin=margin(-10,-10,-5,-10)
+  )
+
+### Figure B - Effect size heatmap
+load("../data/barplot_pathways_fig2b.rda") ## load data
+fig5B <- pathway_counts %>% ggplot(aes(x = pathway, y = proportion, fill = color)) +
+  geom_col() +
+  facet_wrap(~clust_knn, ncol = 1, strip.position = "right") + ## facet by cluster and move facet label to the right
+  ylab("Proportion of mutated samples") +
+  xlab("Significant pathways") +
+  theme_classic() +
+  scale_fill_manual(values = bp.cluster.colors) +
+  theme(
+    axis.text.x = element_blank(),
     axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
     axis.ticks = element_blank(),
-        legend.position = "none")
-fig3_legend <- ggplot(data = data.frame(x = 1:11, y = 1), aes(x = x, y = y, fill = as.factor(x))) +
-  geom_bar(stat = "identity", width = 1) +
-  theme_classic() +
-  scale_fill_manual(values = effect_colors) + 
-  scale_x_continuous(labels= c("-1", "-0.5","0", "0.5", "1")) +
-  theme(legend.position = "none",
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.line = element_blank())
+    axis.line = element_blank(),
+    strip.text.y.right = element_text(angle = 0),
+    strip.background = element_blank(),
+    legend.position = "none"
+  )
 
-fig3_legend <- plot_grid(NULL, fig3_legend, NULL, ncol = 3)
+## Layout for plots using patchwork
+layout <- '
+AAABBB
+AAABBB
+AAABBB
+'
 
-plot_grid(fig3, fig3_legend, ncol = 1, rel_heights = c(15, 1))
+fig5A + fig5B  + plot_layout(design = layout) + plot_annotation(tag_levels = "A")
 ```

### 2020-08-28 code/supplemental_v2.Rmd
Update text in manuscript and file names

@@
 individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Figure 2 of the main text**.
+in **Figure 1B of the main text**.
 
@@
 **Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Figure 2 of the main text**.
+cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**.
 
@@
 **Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
-genes involved in genome stability projected on the UMAP from **Figure 2 of the main text**.
+genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**.
 
@@
 pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
-(column) from **Figure 2 of the main text**.
+(column) from **Figure 1B of the main text**.
 
@@
 **Supplementary Figure S5.
-Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 2 of the main text**.
+Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 1B of the main text**.
 

### 2020-08-28 text/discussion.Rmd
fix textual figure references; reflow text
Update text in manuscript and file names

@@
 The hypothesis that cancer is the result of dysfunction from a limited number of
-basic cellular processes common to eukaryotic cells was introduced and
-later expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
-@hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of a large number
-of syndromes which are unique to each tissue-of-origin, that share some
-mechanisms and treatment strategies. Recent publication of TCGA consortium
-papers present a view largely, and surprisingly, consistent with this latter hypothesis
-[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
-Perhaps owing to the intractable complexity of genomics, proteomics, and
-patient metadata in all its forms, the inescapable conclusion thus far is that
-tissue-of-origin remains the most important driver of tumor characteristics at
-every scale and by every measure. Our observations contrast with this view, and
-instead support an interpretation of publicly available data in which all tumors
-manifest one of a limited number of phenotypes resulting from
-disruptions of basic pathways.
+basic cellular processes common to eukaryotic cells was introduced and later
+expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
+@hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of
+a large number of syndromes which are unique to each tissue-of-origin, that
+share some mechanisms and treatment strategies. Recent publication of TCGA
+consortium papers present a view largely, and surprisingly, consistent with this
+latter hypothesis [@hoadleyMultiplatformAnalysis2014;
+@hoadleyCellofOriginPatternsDominate2018]. Perhaps owing to the intractable
+complexity of genomics, proteomics, and patient metadata in all its forms, the
+inescapable conclusion thus far is that tissue-of-origin remains the most
+important driver of tumor characteristics at every scale and by every measure.
+Our observations contrast with this view, and instead support an interpretation
+of publicly available data in which all tumors manifest one of a limited number
+of phenotypes resulting from disruptions of basic pathways.
 
@@
 
-Some cancer types are unevenly
-distributed among the clusters, though we could not identify any cancers that were exclusive to a
-single cluster. Only cancer types with the fewest samples were found to be
-absent from one or more clusters at all. Somewhat surprisingly to us, this
-finding extends to histological subtypes of breast, head and neck cancers,
-leukemias, _etc._ This result implies that histological
-subtypes could reflect differences in cell-of-origin, rather than fundamental
-differences in cancer phenotype. The four major subtypes of breast cancer
-correspond to histological and molecular expression profiles that define them
-and how they respond to experimental stimuli [@breast-subtypes2]. It has been
-hypothesized that differences in the molecular regulators of development in the
-precursor cell types present in the breast epithelium drive the various
-histological phenotypes [@breast-subtypes; @breast-subtypes3]. Consistent with
-this view we found that breast tumor samples of the Luminal A subtype were
-heavily biased toward membership in clusters 1 and 7, and basal tumor samples
-were biased toward cluster 6. However, both subtypes also contained samples in
-every other cluster (without exception), and Luminal B and Her2 positive samples
-are distributed across clusters. We interpret these data to mean that inherited
-cell-of-origin signatures could predispose certain precursor cells
-within the breast epithelium to forming tumors of one cluster or another but are
-not determinative. This view is compatible with the previously stated hypothesis
-but opens the way for a more granular view of individual tumors. It would be
-surprising if we did not observe bias for some cancers and subtypes amongst our
-classes, since some treatment regimens have greater efficacy for patients of a
-given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
-for some tumors being treatment-refractory in spite of receiving the standard of
-clinical care for diagnostic markers remains elusive. Doubtless some of this is
-due to chance events, as tumors can metastasize and remain dormant years before
-they are detected at distal sites, or resistant clones may have already arisen
-at undetectable levels [@hallmarks_II], but our analysis suggests the
-possibility of identifying more informative molecular, histological or cellular
-subtypes that could form a basis for future stratification of patients into
-different precision treatment regimens.
+Some cancer types are unevenly distributed among the clusters, though we could
+not identify any cancers that were exclusive to a single cluster. Only cancer
+types with the fewest samples were found to be absent from one or more clusters
+at all. Somewhat surprisingly to us, this finding extends to histological
+subtypes of breast, head and neck cancers, leukemias, _etc._ This result implies
+that histological subtypes could reflect differences in cell-of-origin, rather
+than fundamental differences in cancer phenotype. The four major subtypes of
+breast cancer correspond to histological and molecular expression profiles that
+define them and how they respond to experimental stimuli [@breast-subtypes2]. It
+has been hypothesized that differences in the molecular regulators of
+development in the precursor cell types present in the breast epithelium drive
+the various histological phenotypes [@breast-subtypes; @breast-subtypes3].
+Consistent with this view we found that breast tumor samples of the Luminal A
+subtype were heavily biased toward membership in clusters 1 and 7, and basal
+tumor samples were biased toward cluster 6. However, both subtypes also
+contained samples in every other cluster (without exception), and Luminal B and
+Her2 positive samples are distributed across clusters. We interpret these data
+to mean that inherited cell-of-origin signatures could predispose certain
+precursor cells within the breast epithelium to forming tumors of one cluster or
+another but are not determinative. This view is compatible with the previously
+stated hypothesis but opens the way for a more granular view of individual
+tumors. It would be surprising if we did not observe bias for some cancers and
+subtypes amongst our classes, since some treatment regimens have greater
+efficacy for patients of a given cancer or histological subtype
+[@breast-subtypes2]. Nonetheless, the basis for some tumors being
+treatment-refractory in spite of receiving the standard of clinical care for
+diagnostic markers remains elusive. Doubtless some of this is due to chance
+events, as tumors can metastasize and remain dormant years before they are
+detected at distal sites, or resistant clones may have already arisen at
+undetectable levels [@hallmarks_II], but our analysis suggests the possibility
+of identifying more informative molecular, histological or cellular subtypes
+that could form a basis for future stratification of patients into different
+precision treatment regimens.
 
@@
 with more sophisticated filtering of likely passenger mutations [_e.g._
-@spliceAI;
-@deep-learning-clinical-variants]. In the future, we hope to incorporate these
-other data into a comprehensive pathway-centric analysis as we have done here
-for point mutations and indels.
+@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
+incorporate these other data into a comprehensive pathway-centric analysis as we
+have done here for point mutations and indels.
 
@@
 greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
-are known to be responsible for discrete mutation signatures such as APOBEC,
-BRCA1 and BRCA2 produce mutational hotspots reflecting varying selective
+are known to be responsible for discrete mutation signatures such as _APOBEC_,
+_BRCA1_ and _BRCA2_ produce mutational hotspots reflecting varying selective
 pressures in different tissues [@PCAWG-MAIN].
@@
 therapies _do_ target highly-specific driver genes, markers, and signaling
-pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader context of
+pathways ( _e.g. TP53_, _EGFR_ or _HER2_), but understanding the broader context of
 the genetic background and pathway vulnerability of tumors containing such

### 2020-08-28 text/manuscript.Rmd
Update text in manuscript and file names

@@
 to each pathway with one or more genes mutated and 0 for all others (**figure
-1**).
+1A**).
 
@@
 #### Figure 1
-```{r child='../code/figure1v2.Rmd'}
+```{r child='../code/figure1.Rmd'}
 ```
@@
 #### Figure 2
-```{r child = '../code/figure3.Rmd'}
+```{r child = '../code/figure2.Rmd'}
 ```
@@
 subtypes in cancer, each with different prognoses and standards of care. We did
-not observe any exclusive segregation by pathway for these subtypes (**Figure
-3B**). We also projected histological subtype data for the remaining cancers
+not observe any exclusive segregation by pathway for these subtypes (**figure
+2B**). We also projected histological subtype data for the remaining cancers
 (see **supplementary figure S2** and online [**supplemental file
@@
 #### Figure 3
-```{r child = '../code/figure5.Rmd'}
+```{r child = '../code/figure3.Rmd'}
 ```
@@
 selected the top ten enriched and depleted genes (pvalue < 0.01) for each
-cluster (**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared
+cluster (**figure 4**; odds ratios plot). Clusters 7 and 8, which shared
 multiple enrichment in signaling pathways, are largely driven by mutations in
@@
 depleted in clusters 3 and 8). Next, we investigated the proportion of samples
-per cancer type for the significant genes within a cluster (**Figure 4**;
+per cancer type for the significant genes within a cluster (**figure 4**;
 heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
@@
 #### Figure 4
-```{r child = '../code/figure6.Rmd'}
+```{r child = '../code/figure4.Rmd'}
 ```

### 2020-08-28 text/survival.Rmd
Update text in manuscript and file names

@@
 
-To illustrate the model, in **Figure 5A** we simulate survival time curves
-for four related groups. We start with a cohort of women aged 30 and show their
+To illustrate the model, we simulate survival time curves
+for four related groups (**figure 5A**. We start with a cohort of women aged 30 and show their
 expected survival (light blue). By setting their cancer rate multiplier to $k=2$
@@
 #### Figure 5
-```{r child='../code/survival/survfig.Rmd'}
+```{r child='../code/survival/figure5.Rmd'}
 ```
@@
 We found that cancer types, as expected, have a range of prognoses relative to
-the general population. In **Figure 5B-E** we see three particularly deadly
+the general population. In **figure 5B-E** we see three particularly deadly
 cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
@@
 The degree to which a given cancer accelerates aging can be determined by the
-$k_{tis}$ multiplier (**Figure 5F**) multiplied by the cluster modifier $k_{cl}$
-(**Figure 6**). Looking at the tissue specific multiplier, the least deadly was
+$k_{tis}$ multiplier (**figure 5F**) multiplied by the cluster modifier $k_{cl}$
+(**figure 6**). Looking at the tissue specific multiplier, the least deadly was
 prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
@@
 
-```{r child='../code/figure9.Rmd'}
+```{r child='../code/figure6.Rmd'}
 ```
@@
 than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
-or more deadly than other clusters within each cancer type (**Figure 6**). A
+or more deadly than other clusters within each cancer type (**figure 6**). A
 cluster-specific rate of 1 represents the typical rate for this tissue type. For

### 2020-08-28 text/supplemental_methods.Rmd
Added graphs for priors for the ktis and kclass, as well as some text

--- 2020-08-26 text/supplemental_methods.Rmd
@@
 author: "Peter T. Nguyen, Simon G. Coetzee, Daniel L. Lakeland, and Dennis J. Hazelett"
-date: "8/24/2020"
 bibliography: references.bib
@@
 
+### Figure SM3
+```{r}
+
+pritis = ggplot(data.frame(x=0:30),aes(x)) + stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area",fill="#666699")+
+coord_cartesian(xlim=c(0,20),ylim=c(0,1))+labs(title="Prior for ktis",x="ktis")
+pricl = ggplot(data.frame(x=0:30),aes(x)) + stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area",fill="#666699")+
+coord_cartesian(xlim=c(0,20),ylim=c(0,1))+labs(title="Prior for kclass",x="kclass")
+print(pritis+pricl)
+
+```
+
+Priors for the tissue and class coefficients. The tissue coefficient
+is chosen to admit parameter values across a range from near 0 to 20+
+in the high density region, reflecting the wide variety of severities
+across tissue type. Within a given tissue, the class k parameter
+multiplies the tissue k and is given a peak density at 1, and a
+credible range from near 0 to 5, reflecting our expectation of a
+moderate amount of variation across the classes.
+
 As an aside, we did not explicitly account for stage at diagnosis in our model

### 2020-08-28 code/figure6.Rmd
Add SM3 to supplemental_methods

@@
-```{r forest-heatmap-data-generation, eval = FALSE}
+```{r, message = FALSE, fig.width = 9, fig.height= 9}
 load("../data/kelly.colours.rda")
+cluster.colours <- c("grey", kelly.colours[c(3:12)])
+names(cluster.colours) <- c(0:10)
 
-cluster.colors <- kelly.colours[-c(1)]
-names(cluster.colors) <- 0:20
-cluster_colors <- cluster.colors[2:11]
-names(cluster_colors) <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
+## Getting n_race for white and black
+cancerdat <- read_csv("../data/survival/clinical_plus_cluster.csv")
+cancer.clusttype <- cancerdat %>% dplyr::select(type, race, clust_knn) %>% ## select relevant columns
+  filter(race %in% c("WHITE", "BLACK OR AFRICAN AMERICAN")) ## filter black and white only
+cancer.race <- as.data.frame(table(cancer.clusttype$type)) ## table cancer type
+cancer.race$facetname <- paste0(cancer.race$Var1, " (N = ", cancer.race$Freq, (")")) ## Create a new col with type and freq together
+cancer.race$Freq <- NULL ## Remove Freq Column
+names(cancer.race) <- c("tiss", "facetname") ## rename cols for left join in kclassplot below
 
-### zcat simple_somatic_mutation.open.tsv.gz | grep -Fw -f mutation_classes.txt > simple_somatic_mutation.open.selmuts.tsv
-### cat <(zcat simple_somatic_mutation.open.tsv.gz | head -1) simple_somatic_mutation.open.selmuts.tsv | gzip -c - >simple_somatic_mutation.open.selmuts.tsv.gz
-all_muts <- icgcSimpleMutationToMAF("../data/simple_somatic_mutation.open.selmuts.tsv.gz", removeDuplicatedVariants = FALSE, addHugoSymbol = TRUE)
-all_muts <- distinct(.data = all_muts,
-                     Chromosome, Start_Position, End_Position, 
-                     Reference_Allele, Tumor_Seq_Allele2, consequence_type, 
-                     mutation_type, ens_id, icgc_donor_id, 
-                     submitted_sample_id, .keep_all = TRUE) %>% 
-  rename(icgc_sample_id = Tumor_Sample_Barcode) %>% 
-  mutate(Tumor_Sample_Barcode = stringr::str_trunc(submitted_sample_id, 16, ellipsis = ""))
-
-# vep.export <- all_muts %>% select(Chromosome, Start_Position, End_Position, Reference_Allele, Tumor_Seq_Allele2, Strand, icgc_mutation_id)
-# vep.export <- vep.export %>% arrange(Chromosome, Start_Position)
-# vep.export <- distinct(vep.export)
-# vep.export <- with(vep.export,
-#                    paste(Chromosome, Start_Position, End_Position,
-#                          paste(Reference_Allele, Tumor_Seq_Allele2, sep = "/"),
-#                          Strand, icgc_mutation_id, sep = "\t"))
-# writeLines(vep.export, con = "../data/vep.export.tsv")
-# vep settings availible in "../data/VEP_settings.png"
-
-clinical_plus_cluster <- readr::read_csv("../data/clinical_plus_cluster.csv")
-patient_ids <- clinical_plus_cluster$patient_id
-solid_tumor_studies <- unique(clinical_plus_cluster$type)
-tumor_clusters <- readr::read_tsv("../data/umap_3d_coors.tsv", guess_max = 10000)
-
-clinical_plus_cluster <- clinical_plus_cluster %>% rename(Tumor_Sample_Barcode = sample_id)
-clinical_plus_cluster <- left_join(clinical_plus_cluster,
-                                   tumor_clusters %>% 
-                                     select(starts_with(c("plot", "cluster", "clust_knn", "sample_id"))),
-                                   by = c("dna_mutation_id" = "sample_id"))
-all_muts <- all_muts %>% filter(Tumor_Sample_Barcode %in% clinical_plus_cluster$Tumor_Sample_Barcode)
-
-vep_output <- readr::read_tsv("../data/vep_output.basic.txt", na = c("", "NA", "-"), guess_max = 10000)
-
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", 
-                  "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant",
-                  "stop_lost", "start_lost", "stop_gained")
-selected_mut_alt <- c("3'UTR", "5'UTR", 
-                      "Frame_Shift_Del", "Frame_Shift_Ins",
-                      "Missense_Mutation", 
-                      "Splice_Site", "Translation_Start_Site", 
-                      "Nonsense_Mutation", "Nonstop_Mutation")
-
-vep_output <- vep_output %>% 
-  rename(icgc_mutation_id = "#Uploaded_variation")
-
-all_muts <- all_muts %>% 
-  left_join(vep_output, 
-            by = c("consequence_type" = "Consequence", 
-                   "icgc_mutation_id" = "icgc_mutation_id",
-                   "ens_id" = "Gene"))
-
-clinical_plus_cluster$clust_knn_char <- factor(clinical_plus_cluster$clust_knn, 
-                                               levels = c(1:10), 
-                                               labels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10"))
-
-all_muts <- left_join(all_muts, clinical_plus_cluster %>% select(Tumor_Sample_Barcode, clust_knn, clust_knn_char, type) %>% distinct())
-
-for( cluster_c in levels(clinical_plus_cluster$clust_knn_char) ) {
-  c_muts <- read.maf(maf = all_muts %>% filter(clust_knn_char == cluster_c),
-                      clinicalData = clinical_plus_cluster %>% filter(clust_knn_char == cluster_c),
-                      vc_nonSyn = selected_mut_alt, 
-                      removeDuplicatedVariants = FALSE)
-  other_muts <- read.maf(maf = all_muts %>% filter(clust_knn_char != cluster_c),
-                         clinicalData = clinical_plus_cluster %>% filter(clust_knn_char != cluster_c),
-                         vc_nonSyn = selected_mut_alt, 
-                         removeDuplicatedVariants = FALSE)
-
-  c_compare <- mafCompare( c_muts,
-                           other_muts )
-  c_compare <- c_compare$results %>% 
-    as_tibble() %>% 
-    filter(adjPval < 0.01) %>%
-    arrange(or) %>% 
-    mutate(Hugo_Symbol = factor(Hugo_Symbol, levels = Hugo_Symbol)) %>%
-    mutate(ci.up = log(ci.up), ci.low = log(ci.low), or = log(or)) %>%
-    mutate(cluster = cluster_c)
-
-  if (cluster_c == "01") {
-    all_compare <- c_compare
-  } else {
-    all_compare <- bind_rows(all_compare, c_compare)
-  }
+## Getting n_cluster within cancer.type
+cancer.nclustintype <- as.data.frame(table(cancer.clusttype$type, cancer.clusttype$clust_knn))
+cancer.nclustintype <- cancer.nclustintype %>% mutate(alpha = case_when(Freq < 10 ~ "0.05",
+                                                                        Freq >= 10 & Freq < 25 ~ "0.5",
+                                                                        Freq >= 25 ~ "1.0"))
+textlist <- list()
+cancertype <- "BRCA"
+for(n in cancertype) {
+  df <- cancer.nclustintype %>% filter(Var1 == n)
+  cancer.race$lab <- as.character(paste(df[1, "alpha"], "\n",df[2, "alpha"], "\n", df[3, "alpha"], "\n",df[4, "alpha"], "\n",df[5, "alpha"], "\n",df[6, "alpha"], "\n",df[7, "alpha"], "\n",df[8, "alpha"], "\n",df[9, "alpha"], "\n",df[10, "alpha"]))
 }
 
-save(all_compare, all_muts, clinical_plus_cluster, cluster_colors, file = "../data/forest.heatmap.data.rda")
+textlist <- as.data.frame(do.call(rbind, textlist))
+
+tisnames <- levels(cancerdat$type)
+
+  ## Processing data
+	kclnames = grep("k\\[",names(dfsamps), value=TRUE)
+	dfclas = dfsamps[,kclnames]
+	dfclas = stack(dfclas)
+	names(dfclas) = c("value","coefname")
+	dfclas$class = as.factor(as.numeric(gsub("k\\[(.*),(.*)\\]","\\2",dfclas$coefname)))
+	dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
+	dfclas$class <- fct_rev(dfclas$class)
+
+	## 
+	dfclas <- left_join(dfclas, cancer.race, by = "tiss")
+
+	## Plotting
+	ggplot(dfclas,aes_(x = quote(value),y = quote(class), fill = quote(class))) + geom_density_ridges() +
+		facet_wrap(. ~ facetname,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
+		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients") +
+	  labs(x = expression(paste("k"["cl"])), y = "Cluster Identity") +
+	  geom_text(data = dfclas, aes(label = lab), x = 2.5, y = 6) +
+	  mytheme + turn.off.legend +scale_fill_manual(values = cluster.colours)
 ```
-
-```{r forest-heatmap-data-plot, fig.align='center', fig.width= 9, fig.height= 8, message=FALSE}
-#fig.width= 16, fig.height= 20,
-load("../data/forest.heatmap.data.rda")
-
-down_compare_plot <- all_compare %>%
-  group_by(cluster) %>%
-  filter(or < 0) %>%
-  filter(adjPval < 0.01) %>%
-  arrange(-or) %>%
-  filter(is.finite(or)) %>%
-  slice_max(order_by = or, n = 10) %>%
-  arrange(cluster, or, .by.group = TRUE)
-down_compare_plot$Hugo_Symbol <- with(down_compare_plot,
-                                      factor(as.character(Hugo_Symbol),
-                                             levels = unique(as.character(Hugo_Symbol))))
-
-up_compare_plot <- all_compare %>%
-  group_by(cluster) %>%
-  filter(or > 0) %>%
-  arrange(-or) %>%
-  filter(is.finite(or)) %>%
-  slice_max(order_by = or, n = 10) %>%
-  arrange(cluster, or, .by.group = TRUE)
-up_compare_plot$Hugo_Symbol <- with(up_compare_plot,
-                                    factor(as.character(Hugo_Symbol),
-                                           levels = unique(as.character(Hugo_Symbol))))
-
-up_p <- ggplot(up_compare_plot, 
-               aes(x = or,
-                   y = Hugo_Symbol,
-                   color = cluster,
-                   group = cluster)) +
-  geom_vline(xintercept = 0, linetype = 2) +
-  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
-  scale_color_manual(values = cluster_colors) +
-  theme_classic() +
-  xlab("log odds ratio") +
-  xlim(c(0, 9)) +
-  ylab("Gene") +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  ggtitle("Enriched") +
-  theme(
-    panel.grid.major.y = element_line(colour = "#f0f0f0"),
-    axis.title.y = element_blank()
-  )
-
-dn_p <- ggplot(down_compare_plot, 
-               aes(x = or, 
-                   y = Hugo_Symbol, 
-                   color = cluster, 
-                   group = cluster)) +
-  geom_vline(xintercept = 0, linetype = 2) +
-  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
-  scale_color_manual(values = cluster_colors, guide = guide_none()) +
-  theme_classic() +
-  xlab("log odds ratio") +
-  xlim(c(-9, 0)) +
-  ylab("Gene") +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  ggtitle("Depleted") +
-  theme(panel.grid.major.y = element_line(colour = "#f0f0f0"))
-
-tdat_up <- all_muts %>%
-  filter(Hugo_Symbol %in% up_compare_plot$Hugo_Symbol) %>%
-  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
-  select(type, clust_knn_char, Hugo_Symbol) %>%
-  table()
-tdat_up <- left_join(
-  as.data.frame(tdat_up),
-  clinical_plus_cluster %>%
-    select(Tumor_Sample_Barcode, type, clust_knn_char) %>%
-    distinct() %>%
-    group_by(type, clust_knn_char) %>%
-    summarise(ntumors = n())) %>%
-  mutate(prop = Freq / ntumors) %>%
-  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
-  select(type, clust_knn_char, Hugo_Symbol, prop) %>%
-  rename(cluster = clust_knn_char)
-tdat_up$Hugo_Symbol <- factor(tdat_up$Hugo_Symbol, levels = levels(up_compare_plot$Hugo_Symbol))
-tdat_up <- up_compare_plot %>% left_join(tdat_up)
-
-tdat_up_p <- tdat_up %>% ggplot(aes(type, Hugo_Symbol, fill = prop)) +
-  geom_tile() +
-  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
-  xlab("Tumor Type") +
-  ylab("Gene Symbol") +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
-  theme_classic() +
-  theme(
-    strip.text = element_blank(),
-    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
-    axis.text.y = element_blank(),
-    axis.title.y = element_blank(),
-    panel.grid.major.y = element_line(colour = "#f0f0f0")
-  )
-
-tdat_dn <- all_muts %>%
-  filter(Hugo_Symbol %in% down_compare_plot$Hugo_Symbol) %>%
-  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
-  select(type, clust_knn_char, Hugo_Symbol) %>%
-  table()
-tdat_dn <- left_join(
-  as.data.frame(tdat_dn),
-  clinical_plus_cluster %>%
-    select(Tumor_Sample_Barcode, type, clust_knn_char) %>%
-    distinct() %>%
-    group_by(type, clust_knn_char) %>%
-    summarise(ntumors = n())) %>%
-  mutate(prop = Freq / ntumors) %>%
-  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
-  select(type, clust_knn_char, Hugo_Symbol, prop) %>%
-  rename(cluster = clust_knn_char)
-tdat_dn$Hugo_Symbol <- factor(tdat_dn$Hugo_Symbol, levels = levels(down_compare_plot$Hugo_Symbol))
-tdat_dn <- down_compare_plot %>% left_join(tdat_dn)
-
-tdat_dn_p <- tdat_dn %>% ggplot(aes(type, Hugo_Symbol, fill = prop)) +
-  geom_tile() +
-  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
-  xlab("Tumor Type") +
-  ylab("Gene Symbol") +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
-  theme_classic() +
-  theme(
-    strip.text = element_blank(),
-    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
-    axis.text.y = element_blank(),
-    axis.title.y = element_blank(),
-    panel.grid.major.y = element_line(colour = "#f0f0f0")
-  )
-
-layout <- "ABBCDD"
-
-dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
-```

### 2020-08-28 text/supplemental_methods.Rmd
Add SM3 to supplemental_methods

--- 2020-08-28 text/supplemental_methods.Rmd
@@
 
+#### Figure SM3
+
+```{r sm3, fig.height=3}
+pritis = ggplot(data.frame(x=0:30),aes(x)) + stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area",fill="#666699")+
+  labs(title=expression(paste("Prior for k"["tis"])),x=expression(paste("k"["tis"])), y = "Probability Density") +
+  theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
+  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) + theme(title = element_text(size = 10))
+
+pricl = ggplot(data.frame(x=0:30),aes(x)) + stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area",fill="#666699")+
+  labs(title=expression(paste("Prior for k"["class"])),x=expression(paste("k"["cl"])), y = "Probability Density") +
+  theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
+  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) + theme(title = element_text(size = 10))
+print(pritis+pricl)
+```
+
+**Supplementary Figure SM3: Priors for the tissue and class coefficients** The tissue coefficient
+is chosen to admit parameter values across a range from near 0 to 20+
+in the high density region, reflecting the wide variety of severities
+across tissue type. Within a given tissue, the class k parameter
+multiplies the tissue k and is given a peak density at 1, and a
+credible range from near 0 to 5, reflecting our expectation of a
+moderate amount of variation across the classes.
+
 As an aside, we did not explicitly account for stage at diagnosis in our model

### 2020-08-29 code/figure6.Rmd
Add alpha values to figure6, Add SM3 to methods

@@
 ## Getting n_race for white and black
-cancerdat <- read_csv("../data/survival/clinical_plus_cluster.csv")
-cancer.clusttype <- cancerdat %>% dplyr::select(type, race, clust_knn) %>% ## select relevant columns
+cancerinfo <- read_csv("../data/survival/clinical_plus_cluster.csv")
+cancer.clusttype <- cancerinfo %>% dplyr::select(type, race, clust_knn) %>% ## select relevant columns
   filter(race %in% c("WHITE", "BLACK OR AFRICAN AMERICAN")) ## filter black and white only
@@
 
-## Getting n_cluster within cancer.type
-cancer.nclustintype <- as.data.frame(table(cancer.clusttype$type, cancer.clusttype$clust_knn))
-cancer.nclustintype <- cancer.nclustintype %>% mutate(alpha = case_when(Freq < 10 ~ "0.05",
+## Getting n_cluster within cancer.type to calculate alpha
+cancer.nclustintype <- as.data.frame(table(cancer.clusttype$type, cancer.clusttype$clust_knn)) ## Determine n_cancer in each cluster
+cancer.nclustintype <- cancer.nclustintype %>% mutate(alpha = case_when(Freq < 10 ~ 0.1, ## Create alpha column for geom_ridges
+                                                                        Freq >= 10 & Freq < 25 ~ 0.5,
+                                                                        Freq >= 25 ~ 1.0),
+                                                      alphatext = case_when(Freq < 10 ~ "0.1", ## Create alpha text for labeling
                                                                         Freq >= 10 & Freq < 25 ~ "0.5",
                                                                         Freq >= 25 ~ "1.0"))
-textlist <- list()
-cancertype <- "BRCA"
-for(n in cancertype) {
-  df <- cancer.nclustintype %>% filter(Var1 == n)
-  cancer.race$lab <- as.character(paste(df[1, "alpha"], "\n",df[2, "alpha"], "\n", df[3, "alpha"], "\n",df[4, "alpha"], "\n",df[5, "alpha"], "\n",df[6, "alpha"], "\n",df[7, "alpha"], "\n",df[8, "alpha"], "\n",df[9, "alpha"], "\n",df[10, "alpha"]))
-}
 
-textlist <- as.data.frame(do.call(rbind, textlist))
+cancer.alpha <- cancer.nclustintype %>% dplyr::select(Var1, Var2, alphatext) %>%
+  'colnames<-' (c("tiss", "class","alphatext"))
 
-tisnames <- levels(cancerdat$type)
+tisnames <- levels(cancerdat$type) ## getting cancer type names
+
+# Code to generate the text label for each alpha
+# textlist <- list() ## create list for for loop
+# for(n in tisnames) {
+#   df <- cancer.nclustintype %>% filter(Var1 == n) ## Filter for specific cancer type
+#   textlist[[n]] <- as.character(paste(df[1, "alphatext"], "\n",df[2, "alphatext"], "\n", df[3, "alphatext"], "\n",df[4, "alphatext"], "\n",df[5, "alphatext"], "\n",df[6, "alphatext"], "\n",df[7, "alphatext"], "\n",df[8, "alphatext"], "\n",df[9, "alphatext"], "\n",df[10, "alphatext"])) ## collect all the alpha values and insert new line after each alpha
+# }
+# textlist <- as.data.frame(do.call(rbind, textlist)) ## rbind list
+# textlist <- textlist %>% 'colnames<-' ("lab") %>% ## rename col to lab
+#   rownames_to_column(var = "tiss") ## cancer type to column for left join
 
@@
 
-	## 
-	dfclas <- left_join(dfclas, cancer.race, by = "tiss")
+## Left joining data created above
+dfclas <- left_join(dfclas, cancer.race, by = "tiss") ## add facet names
+
+# ## Code for adding alpha level text
+# 	dfclas <- left_join(dfclas, textlist, by = "tiss")
+# 	for(n in tisnames) {
+# 	  dfclas[which(dfclas$tiss == n)[-1], "lab"] <- NA ## Remove the extra 29999 labels that will be generated on the ggplot
+# 	}
+
+## Code for setting alpha in geom_density ridges
+alpha.setup <- c() ## create empty vector for loop
+	for(n in tisnames) { ##loop through the cancer type in alphabetical order
+	  df <- cancer.nclustintype %>% filter(Var1 == n) %>% ## Filter for specific cancer type	
+	    arrange(desc(Var2)) ## Arrange in desc order cause thats how the alpha are set from 10 <- 1
+	  for(p in 1:nrow(df)) { ## loop through the clusters starting from 10 --> 1
+	    alpha.setup <- c(alpha.setup, rep(df[p, "alpha"], 512)) ## Get the alpha value for each cluster in each cancer type and rep 512 (necessary for geom ridges)
+	  }
+	}
 
 	## Plotting
-	ggplot(dfclas,aes_(x = quote(value),y = quote(class), fill = quote(class))) + geom_density_ridges() +
+	ggplot(dfclas,aes_(x = quote(value),y = quote(class), fill = quote(class))) + geom_density_ridges(alpha = alpha.setup) +
 		facet_wrap(. ~ facetname,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
@@
 	  labs(x = expression(paste("k"["cl"])), y = "Cluster Identity") +
-	  geom_text(data = dfclas, aes(label = lab), x = 2.5, y = 6) +
-	  mytheme + turn.off.legend +scale_fill_manual(values = cluster.colours)
+	  #geom_text(data = dfclas, aes(label = lab), x = 2.5, y = 6) +
+	  mytheme + theme(plot.title= element_text(hjust = 0.5))+ turn.off.legend +
+	  scale_fill_manual(values = cluster.colours)
 ```

### 2020-08-29 text/supplemental_methods.Rmd
Add alpha values to figure6, Add SM3 to methods

@@
 
-**Supplementary Figure SM3: Priors for the tissue and class coefficients** The tissue coefficient
+**Supplementary Figure SM3: Priors for the tissue and class coefficients.** The tissue coefficient
 is chosen to admit parameter values across a range from near 0 to 20+

### 2020-08-30 text/supplemental_methods.Rmd
Combine ktis and kcl into single plot with alpha set to 0.5

@@
 
-```{r sm3, fig.height=3}
-pritis = ggplot(data.frame(x=0:30),aes(x)) + stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area",fill="#666699")+
-  labs(title=expression(paste("Prior for k"["tis"])),x=expression(paste("k"["tis"])), y = "Probability Density") +
+```{r sm3, fig.height=3, fig.width=3}
+ggplot(data.frame(x=0:30),aes(x)) + 
+  stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area", aes(fill = "ktis"))+ 
+  stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area", aes(fill = "kcl"))+ 
+  labs(title=expression(paste("Prior for k"["tis"], "and k"["cl"])),x= "k value", y = "Probability Density") +
   theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
-  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) + theme(title = element_text(size = 10))
-
-pricl = ggplot(data.frame(x=0:30),aes(x)) + stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area",fill="#666699")+
-  labs(title=expression(paste("Prior for k"["class"])),x=expression(paste("k"["cl"])), y = "Probability Density") +
-  theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
-  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) + theme(title = element_text(size = 10))
-print(pritis+pricl)
+  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) +
+  scale_fill_manual(values = c(alpha("#E69F00", 0.5),alpha("#0072B2", 0.5)),
+                    labels = c(expression(paste("k"["cl"])), expression(paste("k"["tis"])))) +
+  theme(plot.title = element_text(size = 12, hjust = 0.5),
+        panel.grid.major = element_line(size = 0.75),
+        panel.grid.minor = element_line(),
+        legend.position = c(0.8, 0.8),
+        legend.spacing.y = unit(0, "mm"),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 8),
+        legend.key.size = unit(0.75,"line"),
+        legend.background = element_blank(),
+        legend.box.background = element_rect(colour = "black"))
 ```
@@
 moderate amount of variation across the classes.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 

### 2020-08-31 text/supplemental_methods.Rmd
place figures in supplementary methods

@@
 Though the model extends to earlier ages, relatively few patients were under age
-20.
+20 (**supplementary figure SM2**).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Figure SM2
+
+```{r child='../code/figureSM2.Rmd'}
+```
+
+**Supplementary Figure SM2** Comparison between CDC data and ODE model fit for
+males and females of European descent. This comparison does not distinguish on
+ethnicity for simplicity in visualization.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 
@@
 is significantly less constrained so that there is a wider range of risk across
-tissue types.
-
-The $k$ in our model can be taken as, roughly, the aging rate relative to the
-baseline rates. Thus, a value of $k=2$ means that a given cancer causes you to
-experience the same risk in time $dt$ as a non-cancer patient would experience
-at the same age, in time $2dt$.
-
-Using this model, we split the data by Male/Female sex, and by White _vs._
-Black race as these were the main well-identified categories available to us in
-the dataset. Sex is known to be a risk factor for death, with males dying at
-slightly higher rates for all ages of interest in our dataset. Black _vs._ White
-race is an important category due to differences in access to care and other
-socioeconomic and education related factors. However, due to the low number of
-Black patient tumors, we use ethnicity only to identify the baseline risk
-function, while the tissue- and cluster-specific multipliers are not ethnicity
-specific. There were too few tumors from other races to obtain good inference so
-we excluded them from our analysis.
-
-**Supplementary figure SM2** gives a comparison between the post-fit
-survival curves for men and women vs the CDC data it is based on. This
-comparison does not distinguish on ethnicity for simplicity in
-visualization.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure SM2
-
-```{r child='../code/figureSM2.Rmd'}
-```
-
-**Supplementary Figure SM2** Comparison between CDC data and ODE model fit for
-males and females of European descent.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
+tissue types (**supplementary figure SM3**).
 
@@
 
+The $k$ in our model can be taken as, roughly, the aging rate relative to the
+baseline rates. Thus, a value of $k=2$ means that a given cancer causes you to
+experience the same risk in time $dt$ as a non-cancer patient would experience
+at the same age, in time $2dt$.
+
+Using this model, we split the data by Male/Female sex, and by White _vs._
+Black race as these were the main well-identified categories available to us in
+the dataset. Sex is known to be a risk factor for death, with males dying at
+slightly higher rates for all ages of interest in our dataset. Black _vs._ White
+race is an important category due to differences in access to care and other
+socioeconomic and education related factors. However, due to the low number of
+Black patient tumors, we use ethnicity only to identify the baseline risk
+function, while the tissue- and cluster-specific multipliers are not ethnicity
+specific. There were too few tumors from other races to obtain good inference so
+we excluded them from our analysis.
+
 As an aside, we did not explicitly account for stage at diagnosis in our model

### 2020-08-31 text/survival.Rmd
final edits Dennis

@@
 To illustrate the model, we simulate survival time curves
-for four related groups (**figure 5A**. We start with a cohort of women aged 30 and show their
+for four related groups (**figure 5A**). We start with a cohort of women aged 30 and show their
 expected survival (light blue). By setting their cancer rate multiplier to $k=2$
@@
 
-**Figure 5: Comparison of different cancers, simulated and actual results
-results.** To illustrate the risk model, we compare survival of four simulated
-cohorts. A) A group consisting of randomly selected 30 year old women (light
-blue) vs. an equal sized group of 30 year old women with cancer, k = 2 (light
-orange). To illustrate how the mix of ages alters the survival curve, we compare
-a randomly selected group of people with ages uniformly distributed from 30 to
-70 (dark blue) and a randomly selected group with uniformly distributed ages and
-cancer k = 2 (dark orange). B,C,D,E) We compare a randomly selected group with
+**Figure 5: Comparison of different cancers, simulated and actual results.** To
+illustrate the risk model, we compare survival of four simulated cohorts. A) A
+group consisting of randomly selected 30 year old women (light blue) vs. an
+equal sized group of 30 year old women with cancer, $k = 2$ (light orange).
+Compare a randomly selected group of people with ages uniformly distributed from
+30 to 70 (dark blue) and a randomly selected group with uniformly distributed
+ages and cancer $k = 2$ (dark orange). B,C,D,E) Compare random group with
 correct age distribution for the given cancer (turqoise) to the model predicted
@@
 represents the relative aggressiveness of each cluster within the cancer type.
-For example clusters 1, 2, and 6 Glioblastoma (GBM) are apparently much less
-deadly than other types of GBM. A similar situation can be seen with Pancreatic
-Cancer (PAAD) where 1, 6, and 8 are less deadly compared to clusters 7, 9, 10.
-For some cancer types there are few data points and little ability to estimate
-values with precision (DLBC and KICH for example).
+Alpha transparency (α) is set to reflect varying confidence in the posterior
+distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
+≥ 25 α = 1.0.
 

### 2020-08-31 code/figure1.Rmd
Add legend for fig1B

@@
-```{r data-set-creation, eval=FALSE}
-## Data acquired from GDC
-tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
-## Somatic mutations focused in analysis
-selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
-## Filtering tcga_ssm for our selected mutations
-tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
-## Selecting column that we are interested in
-tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
-                   start = tcga_ssm$chromosome_start,
-                   end = tcga_ssm$chromosome_end,
-                   ref_allele = tcga_ssm$reference_genome_allele,
-                   alt_allele = tcga_ssm$mutated_to_allele,
-                   type = tcga_ssm$consequence_type,
-                   mut_type = tcga_ssm$mutation_type,
-                   ensembl_id = tcga_ssm$gene_affected,
-                   donor_id = tcga_ssm$icgc_donor_id,
-                   sample_id = tcga_ssm$submitted_sample_id,
-                   stringsAsFactors = F)
-## Removing duplicate rows
-tcga_ssm <- tcga_ssm %>% distinct()
-## Recategorizing the mutation type
-tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
-tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
-                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
-                                                 type == "splice_acceptor_variant" ~ "splice_variant",
-                                                 type == "splice_donor_variant" ~ "splice_variant",
-                                                 type == "start_lost" ~ "nonsense_variant",
-                                                 type == "stop_gained" ~ "nonsense_variant",
-                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
-                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
-                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
-                                                 type == "missense_variant" ~ "missense_variant",
-                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
-  dplyr::select(-mut_type) ## rm mut_type col
+```{r, fig.width = 9, fig.height= 4}
+load("../data/cancer_disrupt_heatmap.Rda") ## Load  in data processed in figure1.Rmd
 
-## Preparing sample metadata ##
+fig1 <- ggplot(data = cancer_disrupt_heatmap, aes(x=sample_id, y=path_name, fill = membership)) + 
+  geom_tile(size = 0.6) +
+  scale_fill_manual(values = c("grey", "firebrick3"),
+                    labels = c("Not Mutated", "Mutated")) +
+  facet_grid(.~type, scales = "free", space = "free", switch = "x") + ## facet by cancer type; switch: move facet labels top to bottom
+  xlab("") + ylab("377 Reactome Pathways") + labs(fill = "") +
+  guides(fill = guide_legend(override.aes = list(size = 2))) + ## make fill legend shape smaller
+  theme(axis.text.x = element_blank(), ## turn off x axis text
+        axis.text.y = element_blank(), ## turn off y axis text
+        axis.title.x = element_text(size = 12),
+        axis.title.y = element_text(size = 12),
+        axis.ticks = element_blank(), ## remove tick marks
+        strip.text.x = element_text(angle = 90), ## rotate x-axis
+        strip.background = element_blank(),
+        legend.position = "bottom", ## legend parameters
+        legend.key.height = unit(1, "lines"),
+        legend.text = element_text(size = 8),
+        legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(-10,-10,0,-10),
+        plot.margin = unit(c(0.6, -3, 0, 0), "cm"),
+        panel.background = element_rect(fill = "transparent"))
 
-complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples
+## fixing the text cutoff in heatmap
+ fig1 <- ggplotGrob(fig1)
 
-tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t")
-tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% 
-  .[.$submitted_sample_id %in% complete_sample,] %>%
-  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## 01 is primary solid tumor
-  filter(type == "01") ## filtering 01 samples
-tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
-  filter(study == "PCAWG") ## selecting dup samples that are PCAWG
-tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
-  rbind(., tcga_dup_id) ## rbind the fixed dups
-
-tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t")
-tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in
-
-sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id"))
-
-## Preparing gene metadata ##
-
-# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
-# ensembl_2_entrez_full_list <- bitr(genelist, 
-#                                    fromType = "ENSEMBL", 
-#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
-load("../data/ensembl_2_entrez_full_list.Rda")
-tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,]
-
-## Now we only have 19067 genes
-complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id)))
-names(complete_ensembl) <- "ensembl_id"
-
-selected_mut <- unique(as.character(tcga_ssm$type))
-## For every mutation type.. select them.. and create a tsv file 
-for(selected_mut in selected_mut) {
-  print(selected_mut)
-  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
-    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
-    mutate(mutation = 1) %>% ## create a new column with 1's
-    distinct() %>% ## remove duplicated rows
-    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
-    remove_rownames() %>% ## move ensembl_id column to rownames
-    column_to_rownames(var = "ensembl_id") %>% 
-    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
-    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
-  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
-  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
-  names(missing_gene) <- names(df) ## insert sample_id names into col names
-  df <- rbind(df, missing_gene) ## rbind df and missing_gene
-  rm(missing_gene) ## rm data
-  gc()
-  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
-  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
-  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
-  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
-  df <- cbind(df, missing_sample) ## rbind df and missing_gene
-  rm(missing_sample) ## rm data
-  gc()
-  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
-  assign(paste0(selected_mut), value = df) ## create df in r environment
-  rm(df)
-  gc()
+ for(i in which(grepl("strip-b", fig1$layout$name))){ ## grep all text strips at the bottom of plot 
+   fig1$grobs[[i]]$layout$clip <- "off" ## turn off clipping of text
  }
 
-# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
- # genelist <- rownames(missense_variant)
- # gene_data <- getBM(
- #   attributes = c(
- #     'ensembl_gene_id',
- #     'external_gene_name',
- #     'gene_biotype',
- #     'chromosome_name', 
- #     'start_position',
- #     'end_position', 
- #     'strand'),
- #   filters = 'ensembl_gene_id',
- #   values = genelist,
- #   mart = ensembl)
- # save(gene_data, file = "gene_data.Rda")
-load("../data/gene_data.Rda")
-######################
-
-missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
-UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
-UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
-frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
-frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
-splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
-nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
-nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]
-
-## Matching df to the ensembl_id orders in gene_data
-match_ensembl <- function(x) {
-  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),]
-}
-
-missense_variant <- match_ensembl(missense_variant)
-UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
-UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
-frameshdel_variant <- match_ensembl(frameshdel_variant)
-frameshins_variant <- match_ensembl(frameshins_variant)
-splice_variant <- match_ensembl(splice_variant)
-nonstop_variant <- match_ensembl(nonstop_variant)
-nonsense_variant <- match_ensembl(nonsense_variant)
-
-genes <- GRanges(
-  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
-  ranges = IRanges(
-    start = gene_data$start_position,
-    end = gene_data$end_position),
-  strand = gene_data$strand,
-  mcols = list(
-    ensembl_gene_id = gene_data$ensembl_gene_name,
-    hgnc_symbol = gene_data$hgnc_symbol))
-
-gdc_tumor_coding_complete <- SummarizedExperiment(
-  assays = list(
-    missense   = missense_variant,
-    splice     = splice_variant,
-    frameshdel = frameshdel_variant,
-    frameshins = frameshins_variant,
-    utr3       = UTR_3_prime_variant,
-    utr5       = UTR_5_prime_variant,
-    nonsense   = nonsense_variant,
-    nonstop    = nonstop_variant),
-  rowRanges = genes,
-  colData = sample_sheet)
-save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
-rm(list = ls(all.names = T))
-gc()
+fig1 <- as.ggplot(fig1) ## turn back fig into 
 ```
 
-```{r reactome-pathway-creation, eval=FALSE}
-load("../data/gene_data.Rda")
-## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
-main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")
+```{r fig.width = 9, fig.height= 6}
+# start new page
+plot.new()
 
-main2first <- main2first %>% .[,2:3] %>%
-  'colnames<-' (c("first_level", "main_level")) ## rename colnames
+# setup layout
+gl <- grid.layout(nrow=2, ncol=3)
 
-## Filtering out master pathways from first2second_level
-first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
-first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
-  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
-  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level
+## Setup viewports - How figures will be laid out
+vp.1 <- viewport(layout.pos.col=1:3, layout.pos.row=1)
+vp.2 <- viewport(layout.pos.col=1, layout.pos.row=2) 
+vp.3 <- viewport(layout.pos.col=2, layout.pos.row=2) 
+vp.4 <- viewport(layout.pos.col=3, layout.pos.row=2)
 
-ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
-ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
-  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
-  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
-  .[.$PATH_ID %in% first2second$second_level,] %>% ## include only second level pathways
-  left_join(., main2first, by = c("PATH_ID" = "first_level")) %>% distinct() %>% 
-  left_join(., ensembl2reactome[,c(2,4)], by = c("main_level" = "V2")) %>% distinct()
+# init layout
+pushViewport(viewport(layout=gl))
 
-## Filtering pathways not relevant to our analysis by keywords
-filtered_ensembl2reactome <- ensembl2reactome %>% filter(!V4 %in% c("Disease", "Muscle contraction", "Reproduction")) ## Pathways in disease, muscle contraction and reproduction parent pathway were removed
+## vp.1 - Figure A - sankey plot
+pushViewport(vp.1) ## Enter vp.1 viewpoint
+print(fig1, newpage = FALSE) ## Plot ggplot
+mtext("A", side=3, line=3.15, cex=1.2, adj=-0.095)
+popViewport() ## Exit vp.1 viewpoint
 
-toMatch <- c("disease", "defect", "cancer", "oncogenic", "disorder", "infection", "mutant", "bacteria", "listeria", "loss") ## pathways with the following key terms were removed.
+## Data for different perspectives of umap
+umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
+load("../data/kelly.colours.rda")
+cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
+names(cluster.colours) <- c(0:10) ## assigning cluster to the colors
 
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% mutate(l_pathname = tolower(.$PATH_NAME)) %>% 
-  .[!grep(paste(toMatch,collapse="|"), .$l_pathname),]
+theta <- c(0, 90, 270)
+phi <- c(65, 15, 15)
+## View 1
+pushViewport(vp.2)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+n <- 1
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
+                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                  colvar = NULL))
+title(xlab="UMAP 2", line=-1.5)
+legend("top", inset = c(0.1), legend=names(cluster.colours[2:6]), col=cluster.colours[2:6], pch = 20, bty = "n", horiz = T, cex = 0.8)
+legend("top", inset = c(0.15), legend=names(cluster.colours[7:11]), col=cluster.colours[7:11], pch = 20, bty = "n", horiz = T, cex = 0.8)
+mtext("B", side=3, line=-1.2, cex=1.2, adj=0)
+popViewport()
+## View 2
+pushViewport(vp.3)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+n <- 2
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
+                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                  colvar = NULL))
 
-filtered_ensembl2reactome <- filtered_ensembl2reactome %>% .[which(!.$PATH_ID %in% c("R-HSA-212436", "R-HSA-388396")),] ## Removing Generic Transcription Pathway and GPCR downstream signalling as they are broad pathways
-
-gene2level2_lookup <- filtered_ensembl2reactome %>% dplyr::select(ENSEMBL, PATH_ID) %>% ## remove PATH_NAME col
-  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
-  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
-  distinct() %>% ## keep unique rows
-  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
-  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
-  replace(is.na(.), values = 0) ## replace NA's with 0's
-#save(gene2level2_lookup, file = "../data/gene2_level2_lookup.Rda")
-## Pathway names to level 2 lookup table
-pathname2level2 <- filtered_ensembl2reactome %>% dplyr::select(PATH_ID, PATH_NAME) %>% ## remove ENSEMBL col
-  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
-  distinct(.) ## keep unique rows
-#save(pathname2level2, file = "../data/pathname2level2.Rda")
-rm(list = ls(all.names = T))
-gc()
-```
-
-```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
-gtf_file <- "../data/gencode.v30.annotation.gtf"
-gencode_gtf <- rtracklayer::import(gtf_file)
-gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")])#; rm(gencode_gtf)
-## We only want transcript only 
-gene_info <- gene_info[gene_info$type == "transcript",]
-## We only want protein_coding only
-gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
-gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
-## Filtering for transcript_id, gene_id, gene_name columns
-g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
-## Condense df to only having unique data. 
-good.g2t <- as_tibble(unique(g2t))
-
-## df with transcript_id and entrez id                      
-gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
-head(gencode_entrez[1:10,])
-colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
-## Joining data that are joined together by transcript_id
-entrez_g2t <- left_join(good.g2t, gencode_entrez)
-head(entrez_g2t[1:10,])
-## Remove transcript_id column
-entrez_g2t$transcript_id <- NULL
-## Remove . and the numbers following from the ensembl id.
-entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
-## Removing data that does not have entrez id. 83091 out of 83647 had NA values
-table(is.na(entrez_g2t$entrez_id))
-entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
-entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
-## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
-table(duplicated(entrez_g2t$gene_id))
-## These are pseudogenes... We will remove them all.
-dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
-## We're removing the rows that are associated with this pseudogene
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
-## It's gone
-which(entrez_g2t$gene_id == "ENSG00000285607")
-## Are there any duplicate gene names? Yes! 14 of them
-table(duplicated(entrez_g2t$gene_name))
-dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
-entrez_g2t[entrez_g2t$gene_name == "TBCE",]
-## ENSG00000285053  is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
-entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
-## ENSG00000284741 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
-entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
-## ENSG00000285258 is not in the database anywhere
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
-entrez_g2t[entrez_g2t$gene_name == "MATR3",]
-## ENSG00000280987 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
-entrez_g2t[entrez_g2t$gene_name == "SOD2",]
-## ENSG00000285441 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
-entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
-## ENSG00000285292 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
-entrez_g2t[entrez_g2t$gene_name == "PINX1",]
-## ENSG00000258724 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
-entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
-## ENSG00000284024 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
-entrez_g2t[entrez_g2t$gene_name == "IGF2",]
-## ENSG00000284779 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
-entrez_g2t[entrez_g2t$gene_name == "ATF7",]
-## ENSG00000267281 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
-entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
-## ENSG00000284934 is a homolog
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
-entrez_g2t[entrez_g2t$gene_name == "GGT1",]
-## ENSG00000286070 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
-entrez_g2t[entrez_g2t$gene_name == "SCO2",]
-## ENSG00000130489 no longer exist
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
-entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
-## ENSG00000269226 is a readthrough
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]
-
-## Are there any dupped entrez id? 
-dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]
-
-entrez_g2t[entrez_g2t$entrez_id == "55486",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
-entrez_g2t[entrez_g2t$entrez_id == "55300",]
-## ENSG00000281028 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
-entrez_g2t[entrez_g2t$entrez_id == "653082",]
-## ENSG00000286094 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
-entrez_g2t[entrez_g2t$entrez_id == "8622",]
-## ENSG00000284762 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
-entrez_g2t[entrez_g2t$entrez_id == "221468",]
-## ENSG00000286105 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
-entrez_g2t[entrez_g2t$entrez_id == "1124",]
-## ENSG00000285162 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
-entrez_g2t[entrez_g2t$entrez_id == "102723849",]
-## ENSG00000283765 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
-entrez_g2t[entrez_g2t$entrez_id == "883",]
-## ENSG00000286112 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
-entrez_g2t[entrez_g2t$entrez_id == "22891",]
-## ENSG00000261130 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
-entrez_g2t[entrez_g2t$entrez_id == "54816",]
-## ENSG00000285253 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
-entrez_g2t[entrez_g2t$entrez_id == "23370",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
-entrez_g2t[entrez_g2t$entrez_id == "284391",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
-entrez_g2t[entrez_g2t$entrez_id == "440519",]
-## ENSG00000283201 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
-entrez_g2t[entrez_g2t$entrez_id == "149951",]
-## ENSG00000285382 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
-entrez_g2t[entrez_g2t$entrez_id == "388820",]
-## ENSG00000268861 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
-entrez_g2t[entrez_g2t$entrez_id == "8471",]
-## ENSG00000286098 not found on ncbi
-entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]
-
-## manual gene removal complete
-
-load("../data/TCGA.RNA.Rda")
-
-## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_specimen_id) %>%
-  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))
-
-TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>%
-  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>% 
-  .[rownames(.) %in% entrez_g2t$entrez_id,] %>%
-  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
-  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis
-
-g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),]
-
-express_no_express_lookup <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
-    mean(x)
-  })) ## take the mean for each gene across the tumor samples
-  names(mean_gdc) <- i
-  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
-  rm(df, gdc, mean_gdc)
-  gc
-}
-express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
-express_no_express_lookup[is.na(express_no_express_lookup)] <- 0
-
-## 
-gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
-  rownames(.)
-g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset
-
-express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ]
-express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to >10 for gene expression
-express_no_express_lookup[express_no_express_lookup >10] <- 1
-
-## convert entrez id to ensembl_id
-express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
-  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with
-
-## Booleanize mutation count for each tumor sample
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) 
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
-gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1
-
-gdc_gene_exp_adj <- list()
-for(i in unique(tumor_samples$project_code)) {
-  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
-  df <- tumor_samples[df,] ## select only those rows
-  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
-  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
-  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
-  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
-  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
-    exp_data * x
-  })) ## Function to multiply lookup to the data for the specific cancer type
-  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
-  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
-  rm(df, gdc, exp_data, gdc_adj)
-  gc()
-}
-## Combining list into a dataframe.
-gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
-names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
-rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
-gc()
-save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
-```
-
-```{r genes-to-pathways-transformation, eval=FALSE}
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene2_level2_lookup.Rda")
-sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
-complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways
-
-df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
-df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
-df_gene <- as.data.frame(df_gene) ## turn matrix back to df
-
-df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
-  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
-  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-mutation) ## Remove column named mutation
-
-path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
-path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
-path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df
-
-path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
-  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
-  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
-  dplyr::select(-path_member)
-
-df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
-  na.omit(.) ## Omit NA's that were generated due to no matches
-
-## getting the genes involved in our analysis
-genes_in_pathanalysis <- unique(df_lookup$gene)
-write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data
-
-df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches
-
-df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
-  remove_rownames() %>%
-  column_to_rownames(var = "Var2") ## path_id to rownames
-
-## Insert missing tumor samples due to no mutations
-df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
-mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
-names(mx) <- df_missing ## Missing tumor sample names in col names
-rownames(mx) <- rownames(df_lookup) ## Path names in rownames
-df_lookup <- cbind(df_lookup, mx) ## cbind data
-
-## Insert missing pathway due to no mutations
-path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
-mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
-names(mx) <- names(df_lookup) ## Insert tumor sample names
-rownames(mx) <- path_missing ## Insert missing path names
-df_lookup <- rbind(df_lookup, mx) ## cbind data
-
-pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
-gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
-  'rownames<-' (pathway_order) %>%
-  rownames_to_column(var = "PATH_ID") %>%
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  dplyr::select(-PATH_ID) %>% 
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME")
-rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
-gc()
-save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
-```
-
-```{r tumor-clusters, eval = FALSE}
-####################
-## Figure 1A data ##
-####################
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gene2_level2_lookup.Rda")
-
-path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis))))
-path_daisy <- as.dist(as.matrix(path_daisy))
-
-path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
-path_daisy_row <- as.dist(as.matrix(path_daisy_row))
-
-pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
-         show_rownames = F,
-         show_colnames = F,
-         cellheight = 1,
-         cellwidth = 0.5,
-         clustering_distance_cols = path_daisy,
-         clustering_distance_rows = path_daisy_row,
-         cluster_cols = T,
-         cluster_rows = T,
-         cutree_cols = 19,
-         annotation_legend = F,
-         legend_breaks = c(0, 0.25, 0.75, 1),
-         legend_labels = c("","Not Mut", "Mut", ""),
-         color = c("grey", "firebrick3"),
-         clustering_method = "ward.D2",
-         legend = T)
-dev.off()
-pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering
-df <- rownames(gdc_reactome_path_analysis)
-pathway_dendro <- df[pathway_dendro]
-
-tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
-  as.data.frame(.) %>%
-  dplyr::select(project_code, submitted_sample_id, submitted_specimen_id) %>%  
-  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code))
-
-cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type
-
-cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info
-
-for(n in cancer_type) {
-  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
-    filter(project_code == n) %>% ## filter for specific cancer type
-    .[, "sample_id"] ## retain only sample_id
-  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
-  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
-                          show_rownames = F,
-                          show_colnames = F,
-                          cellheight = 1,
-                          cellwidth = 0.5,
-                          cluster_cols = T,
-                          cluster_rows = F,
-                          annotation_legend = F,
-                          color = c("grey", "firebrick3"),
-                          clustering_method = "single", ## ordering is single linkage
-                          legend = T)
-  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
-  cancer_path <- names(cancer_path) ## getting the sample_id original order 
-  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
-  rm(cancer_info, cancer_path)
-  gc()
-}
-dev.off()
-
-for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
-  print(n)
-  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
-  names(cancer_order) <- "sample_id"
-  cancer_order$type <- n
-  for(p in cancer_type[-1]) {
-    print(p)
-    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
-    names(df1) <- "sample_id"
-    df1$type <- p
-    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
-    rm(df1)
-    gc()
-  }
-}
-
-cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering
-
-cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
-  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
-cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories
-
-## set levels based on clustering algorithm above
-cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
-cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))
-
-cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
-save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")
-
-####################
-## Figure 1B data ##
-####################
-gdc_path <- gdc_reactome_path_analysis %>%
-  t() %>%
-  as.data.frame() %>%
-  rownames_to_column(var = "sample") %>% ## move path_names to column
-  .[match(tumor_samples$submitted_sample_id, .$sample), ]
-
-gdc_path_mca <- left_join(gdc_path, tumor_samples[, c("project_code", "submitted_sample_id")],
-  by = c("sample" = "submitted_sample_id")) %>%
-  column_to_rownames(var = "sample")
-
-gdc_path_sample_names <- rownames(gdc_path_mca) ## Getting sample id order
-gdc_path_tumor_types <- gdc_path_mca$project_code ## Getting cancer type order
-
-gdc_path_mca$project_code <- NULL ## Remove project_code column from df
-
-## create a categorical variable from the boolean variable
-pathway_list <- make_clean_names(colnames(gdc_path_mca))
-pathway_count <- length(pathway_list)
-gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
-  factor(x, levels = c(0, 1), labels = c("no", "yes"))
-}))
-colnames(gdc_path_mca) <- pathway_list
-gdc_path_mca$project_code <- gdc_path_tumor_types
-rownames(gdc_path_mca) <- gdc_path_sample_names
-
-## perform mca
-mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)
-
-expl_var <- mca_res$eig[, 2]
-expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)
-
-set.seed(4242)
-for (k in seq(100)) {
-  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
-  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
-  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
-}
-
-randomize <- tibble(
-  eigenvalue = 1:length(expl_var),
-  expl_var = expl_var,
-  perm = colMeans(expl_var_perm),
-  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
-)
-
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-mca_res <- MCA(
-  X = gdc_path_mca[, pathway_list],
-  ncp = optimal_mca_dimensions, graph = FALSE
-)
-
-save(randomize, mca_res, file = "figure1B_data.rda")
-
-####################
-## Figure 1C data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-rng <- RNGseq(max_n_pt * n_iter, 42)
-r <- rng[[3649]]
-
-rngtools::setRNG(r)
-gdc.umap <- umap(mca_res$ind$coord, 
-                 n_components = 3,
-                 n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
-                 n_epochs = 1000,
-                 n_threads = 1)
-gdc_umap_coors <- as.data.frame(gdc.umap) %>%
-  rename(plot_x = V1, plot_y = V2, plot_z = V3)
-rownames(gdc_umap_coors) <- rownames(gdc_path_mca)
-
-####################
-## Figure 1D data ##
-####################
-n_iter <- 40
-max_n_pt <- 200
-score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)
-
-n_tumors <- nrow(mca_res$ind$coord)
-
-cl <- makeCluster(8)
-registerDoParallel(cl)
-
-for (i in 3:max_n_pt) {
-  print(paste0("MIN POINTS = ", i))
-  score_iter <- foreach(1:n_iter,
-    .combine = c,
-    .export = c("mca_res"),
-    .packages = c("dbscan", "uwot"),
-    r = rng[(i - 1) * n_iter + 1:n_iter]
-  ) %dopar% {
-    rngtools::setRNG(r)
-    umap_init_mca <- umap(mca_res$ind$coord, 
-                          min_dist = 1e-6, 
-                          n_components = 3, 
-                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
-                          n_epochs = 1000, 
-                          n_threads = 1)
-    res <- hdbscan(umap_init_mca, minPts = i)
-    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
-    return(score_iter_temp)
-  }
-  score_mca[i - 2, ] <- score_iter
-}
-
-score <- rowMeans(score_mca)
-score_df <- tibble(minPts = 3:max_n_pt, score = score)
-
-sample_df <- score_mca %>%
-  as.data.frame() %>%
-  as_tibble() %>%
-  add_column(minPts = 3:max_n_pt, .before = 1) %>%
-  pivot_longer(-minPts) %>%
-  group_by(minPts)
-
-optimal_minPts <- score_df %>%
-  filter(score == min(score)) %>%
-  pull(minPts)
-
-save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")
-
-####################
-## Figure 1E data ##
-####################
-rngtools::setRNG(r)
-
-cluster_umap <- umap(mca_res$ind$coord,
-  min_dist = 1e-6,
-  n_components = 3,
-  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
-  n_epochs = 1000,
-  n_threads = 1
-)
-colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
-hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
-umap_3d_coors <- gdc_umap_coors %>%
-  bind_cols(as.data.frame(cluster_umap)) %>%
-  mutate(
-    n_pathways_mut = rowSums(gdc_path[, -1]),
-    clust = as.factor(hdb_cluster_assignment$cluster),
-    member_prob = hdb_cluster_assignment$membership_prob,
-    sample_id = rownames(gdc_umap_coors)
-  )
-## clust 0 is unclassified cluster
-
-####################
-## Figure 1F data ##
-####################
-## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
-umap_3d_coors$clust_knn <- umap_3d_coors$clust
-umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
-  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
-  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
-  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
-)
-
-####################
-## Figure 1G data ##
-####################
-missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
-splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
-frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
-frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
-utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
-utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
-nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
-nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)
-
-mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
-rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
-gc()
-mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
-  colSums() %>% as.data.frame() %>% ## sum the mutation count
-  'colnames<-' (c("mut_count")) %>% ## rename colname
-  rownames_to_column(var = "sample_id") ## move rownames to column
-
-umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id") 
-rm(mutation_count)
-
-####################
-## Figure 1H data ##
-####################
-umap_3d_coors <- left_join(umap_3d_coors, tumor_samples[,c("project_code", "submitted_sample_id")], by = c("sample_id" = "submitted_sample_id"))
-
-####################
-## Figure 1I data ##
-####################
-## getting subtype information on our cancer type
-subtypes <- PanCancerAtlas_subtypes() 
-tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
-subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
-  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
-  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
-  distinct() %>% ## only unique columns
-  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
-  filter(Subtype_Selected != "Normal") %>%
-  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups
-
-#save(subtypes, file = "subtypes.Rda")
-load("../data/subtypes.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
-  dplyr::select(-pan.samplesID)
-
-##########################
-## Figure 1K and L data ##
-##########################
-## getting tumor_metastasis, staging
-tcga_clinical_tumor_staging <- list()
-for(n in cancer_type) {
-  print(n)
-  cancername <- paste0("TCGA-",n)
-  query <- GDCquery(project = cancername,
-                  data.category = "Clinical",
-                  data.type = "Clinical Supplement",
-                  data.format = "BCR Biotab",
-                  file.type = "patient")
-  GDCdownload(query)
-  clinical <- GDCprepare(query)[[1]]
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
-    cat("Yep, ajcc info in there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) {
-    cat("Nope, ajcc stage not there, but pathologic scores is!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-  }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
-    cat("Nope, ajcc info not there!\n")
-    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
-    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
-    }
-  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
-  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
-  tcga_clinical_tumor_staging[[n]] <- clinical
-}
-
-tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
-tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
-missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 ))
-
-tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
-  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
-  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
-  rbind(tcga_clinical_tumor_staging, .) ## rbind
-
-save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")
-
-umap_3d_coors <- umap_3d_coors %>%
-  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
-  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
-  dplyr::select(-bcr_patient_barcode)
-
-M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
-M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
-NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")
-
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
-umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"
-
-stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
-stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
-stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
-stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
-unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")
-
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
-umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
-umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"
-
-write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
-```
-
-```{r fig-1a-1l, warning=FALSE, fig.height=12, fig.width=11, message = FALSE}
-#fig.height=15, fig.width=16,
-load("../data/cancer_disrupt_heatmap.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-
-kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
-                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
-                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
-                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")
-
-fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
-  geom_tile(size = 0.6) +
-  scale_fill_manual(values = c("grey", "firebrick3")) +
-  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
-scale_y_discrete(position = "left") +
-  ylab("") +
-  xlab("377 Reactome Pathways") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        strip.text.y = element_text(size = 12, hjust = 0),
-        strip.text.y.left = element_text(angle = 0),
-        strip.background = element_blank(),
-        legend.position = "none")
-
-## fixing the text cutoff in heatmap
- fig1Ag <- ggplotGrob(fig1A)
-
- for(i in which(grepl("strip-l", fig1Ag$layout$name))){
-   fig1Ag$grobs[[i]]$layout$clip <- "off"
- }
-
- fig1A <- as.ggplot(fig1Ag)
- rm(fig1Ag)
-
-##figure 1b
-load("../data/figure1B_data.rda")
-optimal_mca_dimensions <- randomize %>%
-  filter(pvalue < 0.05) %>%
-  filter(eigenvalue == max(eigenvalue)) %>%
-  pull(eigenvalue)
-
-permut_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
-  geom_line(aes(colour = "#998ec3")) +
-  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
-  scale_color_identity(guide = guide_legend(title = ""), 
-                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
-  scale_y_continuous(labels = scales::percent) +
-  xlab("MCA Eigenvalues") +
-  ylab("explained variance") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        legend.position = c(0.8, 0.8))
-
-pvalue_p <- randomize %>%
-  filter(eigenvalue < 50) %>%
-  ggplot(aes(x = eigenvalue, y = pvalue)) +
-  geom_line() +
-  geom_point() +
-  geom_vline(xintercept = optimal_mca_dimensions, color = "red") +
-  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_mca_dimensions)) +
-  xlab("MCA Eigenvalues") +
-  ylab("P-value") +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-fig1B <- permut_p /
-  pvalue_p
-
-fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
-                aes(y=plot_x, x=plot_y)) +
-  geom_point(size = 0.25) +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  scale_color_viridis_c() +
-  labs(tag = "C") +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-load("../data/optimal_minpts_score.rda")
-fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
-  geom_point(data = sample_df, color = "grey", aes(y = value), size = 0.25) +
-  geom_point(size = 0.25) + 
-  geom_line() +
-  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
-  scale_y_log10() +
-  xlab("Minimum Cluster Size") +
-  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
-  theme_classic() +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12))
-
-## umap hdb_scan
-fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(y = plot_x, x = plot_y, color = as.factor(clust))) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## umap knn_clust 
-fig1F <- ggplot(subset(umap_cancertype), aes(y = plot_x, x = plot_y, color = as.factor(clust_knn))) +
-  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
-  guides(color = F) +
-  theme_classic() +
-  xlab("UMAP 1") +
-  ylab("UMAP 2") +
-  coord_fixed() +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank())
-
-## mutation burden plot
-mut_order <- list() ## store data in list 
-for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
-  message(n)
-  mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
-    dplyr::select(sample_id, plot_x, plot_y, clust_knn, mut_count) %>% ## select relevant columns
-    arrange(mut_count) ## arrange mutation in increasing order
-}
-mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list
-
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
-mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10))
-
-## Need the breaks positions for the x-axis
-previous_class <- 0 ## starting position 1
-position_of_previous_class <- 0 ## starting position 2
-xaxis_breaks <- list() ## list for for loop
-for(n in 1) { ## loop through 1 to 12 clusters
-  message(n)
-  class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
-  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-  previous_class <- class_spec ## storing class halfway mark
-  for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
-    message(n)
-    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
-    class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
-    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
-    previous_class <- class_spec ## storing class halfway mark
-  }
-}
-xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
-xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions
-
-total_n <- as.data.frame(table(mut_order$clust_knn)) ## getting total n in cluster
-
-mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count
-
-mut_median <- mut_order %>% 
-  group_by(clust_knn) %>% 
-  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class
-
-## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
-mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column
-
-begin_NA <- data.frame(sample_id = paste0("sample", 1:100), 
-                       clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
-                       log10mut = NA) ## put NA values so they don't get plotted
-end_NA <- data.frame(sample_id = paste0("sample", 101:200), ## repeat the step for the ends
-                       clust_knn = max(as.numeric(mut_order$clust_knn)),
-                       log10mut = NA)
-mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order
-
-mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id) ## factor samples
-mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10)) ## factor clusters
-
-fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = clust_knn)) + 
-  geom_point(size = 0.25) + 
-  ylab("Log10(Mutation Count)") + 
-  xlab("") +
-  theme_classic() +
-  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") +
-  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
-  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
-  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[1]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
-  theme(axis.text.x = element_text(size = 10),
-        axis.text.y = element_text(size = 10),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks.x = element_blank(),
-        axis.line.x = element_line(colour = "black", size = 0.5),
-        axis.line.x.top = element_line(colour = "black", size = 0.5),
-        panel.spacing = unit(1, "lines"),
-        strip.background = element_rect(size = 0.5),
-        strip.placement = "outside",
-        strip.text.x = element_text(size = 10),
-        legend.position = "none") +
-  scale_color_manual(values = kelly.colours[-c(1:2)])
-
-## code to keep only the bottom border line of the facet
-fig1G <- ggplotGrob(fig1G)
-lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
-                   gp=gpar(col="black", lwd=4))
-
-for (k in grep("strip-t",fig1G$layout$name)) {
-  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
-}
-
-fig1G <- as.ggplot(fig1G)
-
-## umap by cancer type
-fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(y = plot_x, x = plot_y, colour = project_code)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
-                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap of brca subtype
-fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(y = plot_x, x = plot_y, colour = Subtype_Selected)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) +
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap brca somatic mutations
-load("../data/gdc_gene_exp_adj.Rda")
-load("../data/gene_data.Rda")
-umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
-brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
-  rownames_to_column(var = "ensembl_gene_id") %>%
-  left_join(., brca, by = "ensembl_gene_id") %>% 
-  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
-  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
-  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
-  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
-  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
-                           BRCA1 == "0" ~ "NA"),
-         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
-                           BRCA2 == "0" ~ "NA")) %>%
-  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
-  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
-  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "project_code")], by = "sample_id") ## leftjoin by sample id
-
-fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(y = plot_x, x = plot_y, colour = mut_type)) +
-  geom_point(data = brca, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
-                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap stage
-fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(y = plot_x, x = plot_y, colour = ajcc_pathologic_tumor_stage)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(size = 0.25) + 
-  theme_classic() +
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
-                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-## umap metastasis
-fig1L <- ggplot(umap_cancertype, aes(y = plot_x, x = plot_y)) +
-  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
-  ylab("UMAP 2") +
-  xlab("UMAP 1") +
-  coord_fixed() +
-  labs(color = "") +
-  guides(colour = guide_legend(override.aes = list(size=3))) +
-  scale_color_manual(labels = c("M0", "M1"), 
-                     values = c('#a6cee3','#1f78b4')) +
-  theme_classic() +
-  theme(axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 12),
-        axis.title.y = element_text(size = 12),
-        axis.ticks = element_blank(),
-        legend.position = "bottom")
-
-empty_plot <- ggplot() + theme_void()
-
-layout <- '
-AAAABBBCCCDDD
-AAAABBBCCCDDD
-AAAABBBEEEFFF
-AAAABBBEEEFFF
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAGGGGGGGGG
-AAAAHHHIIIJJJ
-AAAAHHHIIIJJJ
-AAAAKKKLLL###
-AAAAKKKLLL###
-'
-
-fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')
+popViewport()
+## View 3
+pushViewport(vp.4)
+par(new=TRUE, fig=gridFIG())
+par(mar = c(0,0,0,0), mai=c(0.1,0.1,0.1,0.1), cex.lab = 0.8, cex = 0.8) ## setting plot parameters
+n <- 3
+with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
+                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
+                                  theta = theta[n], phi = phi[n], scale = T,
+                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
+                                  colvar = NULL))
+popViewport()
 

### 2020-08-31 text/manuscript.Rmd

@@
 “mutated in colon cancer”) and filtered our gene list on membership in these
-pathways (total of 8,940 genes; see methods for details).
+pathways (total of 8,940 genes).
 
@@
 issue, we identified low expressed genes in each type of cancer and eliminated
-them for that cancer type only (see methods for details). Highly expressed genes
-could also have high mutation rates owing to transcription induced mutagenesis
+them for that cancer type only. Highly expressed genes could also have high
+mutation rates owing to transcription induced mutagenesis
 [@parkGenomicEvidenceElevated2012]. This phenomenon would likely result in
@@
 subtypes, though often biased to one or more clusters, are almost never
-exclusive. We interpret these data in aggregate to mean that the clusters do not
-correspond to previously identified molecular subtypes within the parent cancer
-types.
+exclusive. We interpret these data to mean that the clusters do not correspond
+to previously identified molecular subtypes within the parent cancer types.
 
@@
 To discover what pathways are most important for clustering, we calculated
-percent enrichment _within cluster_ relative to all other clusters (see methods)
-and ranked pathways from highest to lowest enrichment. We visualized the
-enrichment as a heatmap (**figure 3A**). Using this approach, we identified
-about fourteen pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see
-methods) (**supplementary table ST2**). Clusters 7, 8, and 9 had several
-pathways in common. To explore the specific pathways marking each cluster, we
-projected disruptions for each of the 377 pathways onto the UMAP
-(**supplementary figure S4** and see interactive media from [**supplemental file
+percent enrichment _within cluster_ relative to all other clusters and ranked
+pathways from highest to lowest enrichment. We visualized the enrichment as a
+heatmap (**figure 3A**). Using this approach, we identified about fourteen
+pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see methods)
+(**supplementary table ST2**). Clusters 7, 8, and 9 had several pathways in
+common. To explore the specific pathways marking each cluster, we projected
+disruptions for each of the 377 pathways onto the UMAP (**supplementary figure
+S4** and see interactive media from [**supplemental file
 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5

### 2020-08-31 text/abstract.Rmd
reduce abstract from 250 to 150 words (Cell requirement)

@@
-Cancer is a complex disease involving disrupted cellular metabolism, basic
-biochemical processes, and the microenvironment. However, despite some generally
-agreed upon unifying principles [@weinberg2000; @hallmarks_II], molecular
+Cancer affects cellular metabolism, biochemical processes, and the
+microenvironment. However, despite some unifying principles, molecular
 signatures remain largely indistinguishable from tissue-of-origin, presenting a
-major barrier for precision health and individualized medicine. To address this
-challenge, we reduce mutation data to disruptions in a select set of pathways
-relevant to basic cell biology, from DNA replication to cellular communication.
-Using dimensionality reduction techniques, we assign tumor samples into ten
-clusters distinct from tissue-of-origin and largely free of bias from mutational
-burden or clinical stage. We show that the clusters vary in prognosis by
-modeling relative risk of death by cancer type and cluster. We identify
-cluster-specific mutations in different tissues, demonstrating that
-tissue-specific signatures contribute to common cellular phenotypes. Moreover,
-germline risk genes involved in replication fidelity and genome stability are
-equally distributed among clusters, contrary to the expectation that such genes
-are avatars of molecular subtype. We investigate metastatic and non-metastatic
-pathways, and show that most differences are cluster-specific. Some metastatic
-pathways from one cluster are cluster-specific pathways from non-metastatic
-tumors of another cluster, suggesting phenotypic convergence. Taken as a whole,
-our observations suggest that common driver genes combine with tissue-specific
-disruptions in tumor-promoting pathways to produce a limited number of distinct
-molecular phenotypes. Thus, we present a coherent view of global tumor biology,
-and explain how common cellular dysfunction might arise from tissue-specific
-mutations.
+major barrier for precision health and individualized medicine. To address this,
+we assign tumors into ten clusters based on pathways largely free of bias from
+tissue-of-origin, mutational burden or stage. Germline risk genes are equally
+distributed among clusters, contrary to the expectation that they determine
+molecular subtype. We show that clusters vary in prognosis, modeling risk of
+death by cancer type and cluster. We investigate metastasis, and show that most
+differences are cluster-specific and complement other clusters, suggesting
+convergence. Our observations suggest that driver genes combine with uncommon
+genes in tumor-promoting pathways to produce distinct molecular phenotypes.
+Thus, we present a coherent view of global tumor biology, and explain how common
+cellular dysfunction might arise from tissue-specific mutations.

### 2020-08-31 code/figure4.Rmd
Modify Figure 4 to generate 2 panel figure (combining enriched / depleted into a single forest plot + heatmap)

--- 2020-08-11 code/figure4.Rmd
@@
 
-up_p <- ggplot(up_compare_plot, 
+all_compare_plot <- bind_rows(down_compare_plot, up_compare_plot)
+all_compare_plot %<>% arrange(or, .by_group = TRUE) %>% mutate(Hugo_Symbol = as.character(Hugo_Symbol)) %>% ungroup() %>% mutate(Row_Index = sprintf("%3i", row_number()))
+Hugo_rownames <- all_compare_plot$Hugo_Symbol
+names(Hugo_rownames) <- as.character(all_compare_plot$Row_Index)
+all_p <- ggplot(all_compare_plot, 
                aes(x = or,
-                   y = Hugo_Symbol,
+                   y = Row_Index,down_compare_plot <- all_compare %>%
+  group_by(cluster) %>%
+  filter(or < 0) %>%
+  filter(adjPval < 0.01) %>%
+  arrange(-or) %>%
+  filter(is.finite(or)) %>%
+  slice_max(order_by = or, n = 10) %>%
+  arrange(cluster, or, .by.group = TRUE)
+down_compare_plot$Hugo_Symbol <- with(down_compare_plot,
+                                      factor(as.character(Hugo_Symbol),
+                                             levels = unique(as.character(Hugo_Symbol))))
+
+up_compare_plot <- all_compare %>%
+  group_by(cluster) %>%
+  filter(or > 0) %>%
+  arrange(-or) %>%
+  filter(is.finite(or)) %>%
+  slice_max(order_by = or, n = 10) %>%
+  arrange(cluster, or, .by.group = TRUE)
+up_compare_plot$Hugo_Symbol <- with(up_compare_plot,
+                                    factor(as.character(Hugo_Symbol),
+                                           levels = unique(as.character(Hugo_Symbol))))
+
+all_compare_plot <- bind_rows(down_compare_plot, up_compare_plot)
+all_compare_plot %<>% arrange(or, .by_group = TRUE) %>% mutate(Hugo_Symbol = as.character(Hugo_Symbol)) %>% ungroup() %>% mutate(Row_Index = sprintf("%3i", row_number()))
+Hugo_rownames <- all_compare_plot$Hugo_Symbol
+names(Hugo_rownames) <- as.character(all_compare_plot$Row_Index)
+all_p <- ggplot(all_compare_plot, 
+               aes(x = or,
+                   y = Row_Index,
                    color = cluster,
@@
   geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
+  scale_y_discrete(labels = Hugo_rownames) +
   scale_color_manual(values = cluster_colors) +
@@
   xlab("log odds ratio") +
-  xlim(c(0, 9)) +
   ylab("Gene") +
   facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  ggtitle("Enriched")
   theme(
@@
 
-dn_p <- ggplot(down_compare_plot, 
-               aes(x = or, 
-                   y = Hugo_Symbol, 
-                   color = cluster, 
-                   group = cluster)) +
-  geom_vline(xintercept = 0, linetype = 2) +
-  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
-  scale_color_manual(values = cluster_colors, guide = guide_none()) +
-  theme_classic() +
-  xlab("log odds ratio") +
-  xlim(c(-9, 0)) +
-  ylab("Gene") +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  ggtitle("Depleted") +
-  theme(panel.grid.major.y = element_line(colour = "#f0f0f0"))
-
-tdat_up <- all_muts %>%
-  filter(Hugo_Symbol %in% up_compare_plot$Hugo_Symbol) %>%
+tdat_all <- all_muts %>%
+  filter(Hugo_Symbol %in% all_compare_plot$Hugo_Symbol) %>%
   distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
@@
   table()
-tdat_up <- left_join(
-  as.data.frame(tdat_up),
+tdat_all <- left_join(
+  as.data.frame(tdat_all),
   clinical_plus_cluster %>%
@@
     summarise(ntumors = n())) %>%
-  mutate(prop = Freq / ntumors) %>%
-  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
-  select(type, clust_knn_char, Hugo_Symbol, prop) %>%
+  group_by(type, clust_knn_char) %>%
+  mutate(c_prop = Freq / ntumors) %>%
+  mutate(c_prop = ifelse(is.na(c_prop), 0, c_prop)) %>%
+  select(type, clust_knn_char, Hugo_Symbol, c_prop) %>%
   rename(cluster = clust_knn_char)
-tdat_up$Hugo_Symbol <- factor(tdat_up$Hugo_Symbol, levels = levels(up_compare_plot$Hugo_Symbol))
-tdat_up <- up_compare_plot %>% left_join(tdat_up)
-
-tdat_up_p <- tdat_up %>% ggplot(aes(type, Hugo_Symbol, fill = prop)) +
+tdat_all <- all_compare_plot %>% left_join(tdat_all)
+tdat_all_p <- tdat_all %>% ggplot(aes(type, Row_Index, fill = c_prop)) +
   geom_tile() +
@@
   ylab("Gene Symbol") +
+  scale_y_discrete(labels = Hugo_rownames) +
   facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
@@
 
-tdat_dn <- all_muts %>%
-  filter(Hugo_Symbol %in% down_compare_plot$Hugo_Symbol) %>%
+layout <- "ABB"
+all_p + tdat_all_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
+                   color = cluster,
+                   group = cluster)) +
+  geom_vline(xintercept = 0, linetype = 2) +
+  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
+  scale_y_discrete(labels = Hugo_rownames) +
+  scale_color_manual(values = cluster_colors) +
+  theme_classic() +
+  xlab("log odds ratio") +
+  ylab("Gene") +
+  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
+  theme(
+    panel.grid.major.y = element_line(colour = "#f0f0f0"),
+    axis.title.y = element_blank()
+  )
+
+tdat_all <- all_muts %>%
+  filter(Hugo_Symbol %in% all_compare_plot$Hugo_Symbol) %>%
   distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
@@
   table()
-tdat_dn <- left_join(
-  as.data.frame(tdat_dn),
+tdat_all <- left_join(
+  as.data.frame(tdat_all),
   clinical_plus_cluster %>%
@@
     summarise(ntumors = n())) %>%
-  mutate(prop = Freq / ntumors) %>%
-  mutate(prop = ifelse(is.na(prop), 0, prop)) %>%
-  select(type, clust_knn_char, Hugo_Symbol, prop) %>%
+  group_by(type, clust_knn_char) %>%
+  mutate(c_prop = Freq / ntumors) %>%
+  mutate(c_prop = ifelse(is.na(c_prop), 0, c_prop)) %>%
+  select(type, clust_knn_char, Hugo_Symbol, c_prop) %>%
   rename(cluster = clust_knn_char)
-tdat_dn$Hugo_Symbol <- factor(tdat_dn$Hugo_Symbol, levels = levels(down_compare_plot$Hugo_Symbol))
-tdat_dn <- down_compare_plot %>% left_join(tdat_dn)
-
-tdat_dn_p <- tdat_dn %>% ggplot(aes(type, Hugo_Symbol, fill = prop)) +
+tdat_all <- all_compare_plot %>% left_join(tdat_all)
+tdat_all_p <- tdat_all %>% ggplot(aes(type, Row_Index, fill = c_prop)) +
   geom_tile() +
@@
 
-layout <- "ABBCDD"
-
-dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
+layout <- "ABB"
+all_p + tdat_all_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
 ```

### 2020-08-31 text/manuscript.Rmd

@@
 
-To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
-Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
-(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
-given the relative completeness of the patient metadata, particularly for
-survival and staging. We chose to analyze somatic mutations in exome sequencing
-because the affected target gene is known unambiguously. Therefore, we
+To study cancer pathways we obtained a set of 7,607 solid tumors from The Cancer
+Genome Atlas (TCGA) through the [Genomic Data Commons (GDC)](gdc.cancer.gov)
+portal. TCGA data are most appropriate given the relative completeness of the
+patient metadata for survival and staging. We chose to analyze exome sequencing
+data because the affected target gene is known unambiguously. Therefore, we
 selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
-splicing mutations. In order to minimize bias from well-studied diseases and
-processes, we selected 377 [Reactome pathways](https://reactome.org/) (see
-supplemental table ST1) of interest corresponding to basic cellular processes
-and biochemical pathways, excluding gene sets that correspond to miscellaneous
-categories (_e.g._ “transcription factors”) or disease associations (_e.g._
-“mutated in colon cancer”) and filtered our gene list on membership in these
-pathways (total of 8,940 genes).
+splicing mutations. To minimize bias from well-studied diseases and processes,
+we selected 377 [Reactome pathways](https://reactome.org/) (see supplemental
+table ST1) of interest corresponding to basic cellular processes and biochemical
+pathways, excluding those corresponding to catalytic categories (_e.g._
+“transcription factors”) or disease associations (_e.g._ “mutated in colon
+cancer”) and filtered our gene list on membership in these pathways (total of
+8,940 genes).
 
-To avoid bias toward larger pathways with more genes, we counted them as
-disrupted if one or more member genes were mutated. We do not attempt to
-calculate enrichment for mutations within a pathway. Data binarized by pathway
-are likely to be noisy for several reasons. First, point mutations can be
-deleterious (attenuating, hypomorphic or antimorphic) or activating (neomorphic
-or hypermorphic) in genes, and these can in turn act functionally as either
-oncogenes or tumor suppressors. For this study we assume a significant fraction
-of these mutations are generically disruptive to normal pathway activity since
-it is impossible to know the tumor promoting effects of all mutations, including
-rarely studied genes. Second, we know that low- and non-expressed genes
+To avoid bias toward pathways with more genes, we counted disruptions if one or
+more member genes were mutated. We do not attempt to calculate enrichment for
+mutations within a pathway. Binarized pathways are likely noisy for a couple
+reasons. First, point mutations can be deleterious (attenuating, hypomorphic or
+antimorphic) or activating (neomorphic or hypermorphic) in genes, and these can
+in turn be oncogenes or tumor suppressors. For this study we assume a
+significant fraction of these mutations are generically disruptive to pathway
+activity since it is impossible to know the tumor promoting effects of all
+mutations, including rarely studied genes. Second, low- and non-expressed genes
 accumulate mutations at a higher rate due to transcription coupled repair
@@
 mutation rates owing to transcription induced mutagenesis
-[@parkGenomicEvidenceElevated2012]. This phenomenon would likely result in
-cell-type-specific biases that might be biologically meaningful for
-predisposition to different classes of cancer however, and therefore chose not
-to exclude these genes from our analysis. After selecting our pathways and
-genes, we then compiled a matrix of the pathways, assigning a Boolean value of 1
-to each pathway with one or more genes mutated and 0 for all others (**figure
-1A**).
+[@parkGenomicEvidenceElevated2012]. This phenomenon could result in
+cell-type-specific biases that might result in predisposition to different
+classes of cancer however, and we therefore did not exclude these genes from our
+analysis. After selecting pathways and genes, we compiled a matrix of pathways,
+assigning a Boolean value of 1 to each pathway with one or more genes mutated
+and 0 for all others (**figure 1A**).
 
@@
 
-**Figure 1: Clustering the molecular pathway profiles of tumor samples with one or more
-mutations.** A) Each of 377 selected Reactome pathways (rows) is classified as
-disrupted if one or more genes is mutated in the tumor sample (columns). Red
-cells denote pathway disruption. Tumors types reflect standardized abbreviations
-from the TCGA project. B) Different rotational perspectives of
-the same MCA-based UMAP projection in three-dimensional space. Each dot
-corresponds to a tumor sample. The same colors indicate the tumor's cluster
-identity throughout this manuscript.
+**Figure 1: Clustering pathways of tumor samples.** A) Each of 377 selected
+Reactome pathways (rows) is classified as disrupted if one or more genes is
+mutated in the tumor sample (columns). Red cells denote pathway disruption.
+Tumors types reflect standardized abbreviations from the TCGA project. B)
+Different rotational perspectives of the same MCA-based UMAP projection in
+three-dimensional space. Each dot corresponds to a tumor sample. The same colors
+indicate the tumor's cluster identity throughout this manuscript.
 
@@
 
-We investigated this dataset further using multiple correspondence analysis
-(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
-1B** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
-We used the resulting UMAP graph coordinates to perform
-density based clustering with HDBSCAN [@hdbscan], which resulted in
-identification of 10 well-defined clusters capturing about 80% of the tumor
-samples. To capture the remaining samples into one of these 10 clusters we used
-kNN (see **Supplementary Methods** for details on clustering methods).
+We investigated this dataset using multiple correspondence analysis (MCA)
+[@JSSv025i01], and visually summarized the analysis with UMAP (**figure 1B** and
+see interactive media from [**supplemental file
+1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
+used the resulting graph coordinates to perform density based clustering
+with HDBSCAN [@hdbscan], which resulted in identification of 10 well-defined
+clusters capturing about 80% of the tumor samples. To capture the remaining
+samples into one of these 10 clusters we used kNN (see **Supplementary Methods**
+for details on clustering methods).
 
@@
 
-Having defined tumors in terms of their pathway disruption profile, we sought to
+Having defined tumors in terms of pathway disruption profile, we sought to
 understand whether different cancer types segregate into one or more predominant
-clusters. To our surprise, most cancer types were not heavily biased in one
+clusters. To our surprise, most cancer types were _not_ heavily biased in one
 cluster, and all well-represented cancer types had tumors in every cluster (see
-**figure 2A** and full tumor profiles in **supplementary figure S1**, see also
-interactive media from **supplementary file 1**), suggesting that our
-pathway-disruptions identify clusters of molecular pathology largely independent
-of tissue-of-origin. As an example of one type of cancer that does have a biased
-pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
-cluster 8 (**figure 2A** and **supplementary figure S1** and **supplementary
-file 1**). But even PAAD comprises tumors from the nine remaining classes,
-meaning that patients with these tumors have potentially different underlying
-molecular pathologies.
+**figure 2A** and full tumor profiles in **supplementary figure S1**,
+interactive media **supplementary file 1**), suggesting that these clusters
+indicate molecular pathology largely independent of tissue-of-origin. As an
+example of one cancer that does have a biased pathway profile, pancreatic
+adenocarcinoma (PAAD) was predominantly found in cluster 8 (**figure 2A** and
+**supplementary figure S1** and **supplementary file 1**), but even PAAD
+comprises tumors from the nine remaining classes. Thus, patients with these
+tumors have potentially different underlying molecular pathologies.
 
@@
 **Figure 2: Pathway-based clustering independent of tissue-of-origin** 
-A) Sankey plot showing correspondence between cancer type and cluster identity.
+A) Sankey plot of correspondence between cancer type and cluster identity.
 B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
@@
 different standards of care owing to different drug sensitivity (or other
-factors). If the histological subtypes represent true molecular phenotypes, one
-predicts that they should segregate with our pathway-based clusters, therefore
-providing support for the clusters as proxies for molecular pathology
-sub-typing. To our surprise, we find a similar result to the previous analysis
-of cancer types projected onto the UMAP of pathway disruptions. To illustrate
-this, we projected annotations for each of the breast cancer subtypes, composed
-of Triple-negative/Basal-like, Her2 positive, normal-like, and luminal A and B
-subtypes onto the UMAP. These are among the most heavily studied molecular
-subtypes in cancer, each with different prognoses and standards of care. We did
-not observe any exclusive segregation by pathway for these subtypes (**figure
-2B**). We also projected histological subtype data for the remaining cancers
-(see **supplementary figure S2** and online [**supplemental file
-1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the
-subtypes, though often biased to one or more clusters, are almost never
+clinical trial data). If histological subtypes represent true molecular
+phenotypes, one predicts they should segregate with pathway-based clusters,
+supporting the clusters as proxies for molecular pathology sub-typing. To our
+surprise, we find a similar result to the previous analysis of cancer types. To
+illustrate this, we projected annotations for each of the breast cancer
+subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like,
+and luminal A and B subtypes onto the UMAP. These are among the most heavily
+studied molecular subtypes in cancer, each with different prognoses and
+standards of care. We did not observe exclusive segregation by cluster for these
+subtypes (**figure 2B**). We also projected histological subtype data for the
+remaining cancers (see **supplementary figure S2** and online [**supplemental
+file1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that
+the subtypes, though often biased to one or more clusters, are almost never
 exclusive. We interpret these data to mean that the clusters do not correspond
@@
 
-There are several well-known familial cancer-causing mutations that have been
-studied for differences in basic biology, survival and treatment outcomes. The
-functions of these genes are related to risk factors such as genome stability,
-proof-reading & DNA damage repair, and telomere length. _BRCA1/2_ genes for
-example are key for DNA double-stranded break repair
-[@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in these genes
-confer elevated risk for breast, prostate and ovarian cancers. The mechanism of
-risk is thought to involve loss of heterozygosity resulting in loss of the
-wildtype, functional allele [@brca-loh], so we projected the somatic mutations
-for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe segregation of
-these mutations into specific clusters (**figure 2C** and see interactive media
-from [**supplemental file
-1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar
-projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_,
-_PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
-except for _APC_ exhibited any remarkable specificity with respect to cluster
-assignment (**supplementary figure S3**). To look at other risk factors such as
-maintenance of DNA methylation levels and telomere length, we projected somatic
-mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
-observed similar lack of segregation by cluster (**supplementary figure S3**).
+There are several familial cancer-causing mutations studied for differences in
+basic biology, survival and treatment outcomes. The functions of these genes are
+related to risk factors such as genome stability, proof-reading & DNA damage
+repair, and telomere length. _BRCA1/2_ genes are key for DNA double-stranded
+break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
+these genes confer elevated risk for breast, prostate and ovarian cancers. The
+mechanism of risk is thought to involve loss of heterozygosity, resulting in
+loss of the wildtype, functional allele [@brca-loh], so we projected somatic
+mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe
+segregation of these mutations into specific clusters (**figure 2C** and see
+interactive media from [**supplemental file
+1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We also projected
+mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_;
+_BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except for _APC_
+exhibited remarkable specificity with respect to cluster assignment
+(**supplementary figure S3**). To look at other risk factors such as maintenance
+of DNA methylation levels and telomere length, we projected somatic mutations of
+the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and observed
+similar lack of cluster bias (**supplementary figure S3**).
 
@@
 
-Tumor staging is based on physico-pathological criteria, including tumor
-diameter, which can vary greatly in importance between different tissues. Stage
-is used clinically as a proxy for advancement toward a more deadly state and
-metastasis. Given these criteria, it is possible that more advanced tumors have
-common pathway disruption profiles. The UMAP, which features a series of
-lobe-like structures on a common backbone of tumor samples (**figure 1B**) could
-in principle reflect progression through a series of stages. The backbone starts
-with a cluster of tumors (class 1) that has the fewest point mutations and
-culminates in a cluster (class 10) which has nearly every pathway disrupted
-(**figure 2F**). However, outside of class 10 we don't observe a trend in the
-overall mutation burden across the backbone of the UMAP. Nonetheless, to test
-the hypothesis that the molecular-pathway disruption clusters represent
-advancement through stages, we projected staging data onto the UMAP. Similar to
-tissue of origin and other categories of tumor, we did not observe any bias
-among the stages to specific clusters (**figure 2D**), suggesting that stage is
-not a contributing factor to cluster identity.
+Tumor staging is based on pathological criteria, including tumor diameter, which
+can vary greatly in importance between different tissues. Stage is used
+clinically as a proxy for advancement toward a more deadly state and metastasis.
+Thus, it is possible that more advanced tumors have common pathway disruption
+profiles. The UMAP, which features a series of lobe-like structures on a common
+backbone of tumor samples could reflect progression through a series of stages.
+The backbone starts with a cluster of tumors (class 1) that has the fewest point
+mutations and culminates in a cluster (class 10) which has nearly every pathway
+disrupted (**figure 2F**). However, we don't observe a trend in the overall
+mutation burden across the backbone of the UMAP. Nonetheless, to test the
+hypothesis that the molecular-pathway disruption clusters represent advancement
+through stages, we projected staging data onto the UMAP. Similar to tissue of
+origin and other categories of tumor, we do not observe any bias among the
+stages to specific clusters (**figure 2D**), suggesting that stage does not
+contribute to cluster identity.
 
 Finally, as a measure of tumor advancement, metastasis is the condition in which
-certain phenotypic criteria are met: loss of
-differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
-system evasion and tissue invasiveness [@hallmarks_II]. To determine whether any
-of our clusters correspond to an especially advanced stage of cancer across
-tissue types, we projected the metastases onto our UMAP, and surprisingly
-we observed even distribution of the samples across clusters (**figure 2E**).
-This final observation suggests that our pathway-disruption clustering is
-dependent on particular combinations of gene mutations affecting different
-pathways that can each give rise to advanced stages of disease and metastasis,
-regardless and independent of overall mutational burden.
+certain phenotypic criteria are met: loss of differentiation, cell-cell
+contacts, epithelial to mesenchyme transition, immune system evasion and tissue
+invasiveness [@hallmarks_II]. To determine whether any clusters correspond to an
+especially advanced stage of cancer across tissue types, we projected the
+metastases onto our UMAP, and surprisingly we observed even distribution of the
+samples across clusters (**figure 2E**). This final observation suggests that
+our pathway-disruption clustering is dependent on particular combinations of
+gene mutations affecting different pathways that can each give rise to advanced
+stages of disease and metastasis, regardless and independent of overall
+mutational burden.
 
@@
 
-In order to identify pathway enrichment across all cancers (pan-cancer), we
-created a list of pathway disruptions with percent mutated samples and top genes
-(**supplementary table ST1**). As expected, these analyses reveal the broad
-importance of many well known pathways that are disrupted in cancer, including
-"PIP3 activates Akt signaling" (77% of samples), "MAP1K/MAP3K signaling" (70% of
-samples), "Mitotic G2-G2/M phases" (67% of samples), "Cellular senescence" (64%
-of samples), "G2/M Checkpoints" (62% of samples), _etc._
+In order to identify pathway enrichment across all cancers, we created a list of
+pathway disruptions with percent mutated samples and top genes (**supplementary
+table ST1**). As expected, these analyses reveal the broad importance of many
+well known pathways that are disrupted in cancer, including "PIP3 activates Akt
+signaling" (77% of samples), "MAP1K/MAP3K signaling" (70% of samples), "Mitotic
+G2-G2/M phases" (67% of samples), "Cellular senescence" (64% of samples), "G2/M
+Checkpoints" (62% of samples), _etc._
 
@@
 percent enrichment _within cluster_ relative to all other clusters and ranked
-pathways from highest to lowest enrichment. We visualized the enrichment as a
+pathways from highest to lowest enrichment. We visualized enrichment as a
 heatmap (**figure 3A**). Using this approach, we identified about fourteen
@@
 common. To explore the specific pathways marking each cluster, we projected
-disruptions for each of the 377 pathways onto the UMAP (**supplementary figure
-S4** and see interactive media from [**supplemental file
+disruptions for each of the 377 pathways onto the UMAP (**supplementary table
+ST2** and see interactive media from [**supplemental file
 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
 were distinguished by metabolic pathways including RNA and protein biosynthesis
-(**supplementary figure S4**). Similarly, cluster 4 was distinguished by
+(**supplementary table ST2**). Similarly, cluster 4 was distinguished by
 mutations affecting regulation of DNA and histone methylation ("DNA
@@
 Clusters 7-9 have in common mutations in extracellular, intracellular, and
-immune-related signaling pathways (see **figure 3B** and **supplementary figure
-S4**). Cluster 2 had the highest pathway enrichment levels of the three, having
+immune-related signaling pathways (see **figure 3B** and **supplementary table
+ST2**). Cluster 2 had the highest pathway enrichment levels of the three, having
 mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
@@
 
-**Figure 3: Pan-cancer enrichment of pathway disruptions.** 
-A) Heatmap shows the relative enrichment of each pathway (rows) within each
-numbered cluster (columns). Effect size is displayed as colors representing
-percent enrichment. B) Proportion of mutated samples in the each significant
-pathway (columns; union set of pathways with effect size ≥ 0.30 in each
-cluster) within each numbered cluster (rows).
+**Figure 3: Pan-cancer enrichment of pathway disruptions.** A) Heatmap shows
+relative enrichment of each pathway (rows) within numbered cluster (columns).
+Color represents effect size as percent enrichment. B) Proportion of mutated
+samples in the each significant pathway (columns; union set of pathways with
+effect size ≥ 0.30 in each cluster) within cluster (rows).
 
@@
 Prior efforts to extract signatures from pan-cancer datasets met with difficulty
-in distinguishing tumor samples from their tissue-specific -omics data
-signatures. Given our pathway-disruption based clustering, this raises the
-question, are tumor phenotypes driven entirely by common driver genes, or by
-“silent” tissue-specific effectors (_i.e._ too few samples to detect above
-statistical significance thresholds), or a combination of both? To answer this
-question, we compared top pathway genes for each cluster relative to the TCGA
-background to find differentially mutated genes. We ranked odds ratios and
-selected the top ten enriched and depleted genes (pvalue < 0.01) for each
-cluster (**figure 4**; odds ratios plot). Clusters 7 and 8, which shared
-multiple enrichment in signaling pathways, are largely driven by mutations in
-_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
-_KRAS_ panels of **Supplementary figure S5** and see interactive media from
-[**supplemental file
+in distinguishing tumor samples from tissue-specific -omics data signatures.
+Given our pathway-disruption based clustering, this raises the question, are
+tumor phenotypes driven by common driver genes, “silent” tissue-specific
+effectors (_i.e._ too few samples to detect above statistical significance
+thresholds), or a combination of both? To answer this question, we compared top
+pathway genes for each cluster relative to TCGA background to find
+differentially mutated genes. We ranked odds ratios and selected the top ten
+enriched and depleted genes (pvalue < 0.01) for each cluster (**figure 4**; odds
+ratios plot). Clusters 7 and 8, which shared multiple enrichment in signaling
+pathways, are largely driven by mutations in _PI3K_ and its orthologs and _Ras_
+genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary
+figure S4** and see interactive media from [**supplemental file
 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
@@
 defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
-and nuclear pore complex, respectively. Cluster 4 had mutations in genes
-responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
-subunit genes involved in protein degradation. We also observed that genes that
-were enriched for one cluster are depleted from others (_i.e._ is enriched in
-cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
-depleted in clusters 3 and 8). Next, we investigated the proportion of samples
-per cancer type for the significant genes within a cluster (**figure 4**;
-heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
-more highly mutated genes across all cancers. Instead, when observing the
-mutation rate for these genes within samples that belong to a cluster, the
-mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
-4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
-enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
-there is no global pattern, indicating that our clusters are not driven by
-individual genes, but rather networks as a whole. Taken together, our data
-identify a framework of cancer type-specific mutations associated with specific
-clusters.
+and nuclear pore complex, respectively. Cluster 4 had mutations nucleosome
+structural or subunit genes. Cluster 2 had mutations in proteasomal subunits
+involved in protein degradation. We also observed that genes enriched for one
+cluster are depleted from others (_i.e._ is enriched in cluster 6, but depleted
+in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in clusters 3 and
+8). Next, we investigated the proportion of samples per cancer type for the
+significant genes within a cluster (**figure 4**; heatmap). Surprisingly,
+clusters were not predominated by one or more highly mutated genes across all
+cancers. Instead, when observing the mutation rate for these genes within
+samples that belong to a cluster, the mutation rate is heterogeneous across
+tumors by tissue origin (_e.g._ In cluster 4, CESC was enriched for _H2AFX_, OV
+was enriched for _HIST1H2BD_, and UCEC was enriched for _HIST1H2AC_). Even among
+the top enriched genes within clusters there is no global pattern, indicating
+that our clusters are not driven by individual genes, but rather networks as a
+whole. Taken together, our data identify a framework of cancer type-specific
+mutations associated with specific clusters.
 
@@
 Since metastatic tumors are distributed across all ten clusters, we first
-compared them with non-metastatic tumors following the same logic we used to
-investigate cluster specific enrichment. Using all non-metastatic tumors as
+compared them with non-metastatic tumors following the logic we used to
+investigate cluster specific enrichment. Using non-metastatic tumors as
 background, we found very low levels of enrichment (< 10%) in a handful of
@@
 clusters. For example, “Signaling by _PTK6_” is enriched in non-metastatic
-samples of cluster 8 (see **supplementary figure S4**) but not in 7 and 9. This
+samples of cluster 8 (**supplementary table ST2**) but not in 7 and 9. This
 pathway is enriched in metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$,
 **table 1**). This is also true of “Erythropoietin activates _RAS_”, which is
-enriched in non-metastatic tumors of cluster 8 (**supplementary figure S4**) and
-also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic samples were
+enriched in non-metastatic tumors of cluster 8 (**supplementary table ST2**) and
+also in metastatic tumors of clusters 7 and 9. Cluster 4 metastases were
 enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathway,
 which is also specific to clusters 2, 7 and 8 non-metastatic tumors. Thus,
-metastatic-enriched pathways from one cluster are often enriched in
-non-metastatic tumors of other clusters.
+metastatis pathways from one cluster are often enriched in non-metastases of
+other clusters.
 

### 2020-08-31 text/abstract.Rmd
incorporate Tim's suggestions for abstract

@@
-Cancer affects cellular metabolism, biochemical processes, and the
-microenvironment. However, despite some unifying principles, molecular
-signatures remain largely indistinguishable from tissue-of-origin, presenting a
-major barrier for precision health and individualized medicine. To address this,
-we assign tumors into ten clusters based on pathways largely free of bias from
-tissue-of-origin, mutational burden or stage. Germline risk genes are equally
-distributed among clusters, contrary to the expectation that they determine
-molecular subtype. We show that clusters vary in prognosis, modeling risk of
-death by cancer type and cluster. We investigate metastasis, and show that most
-differences are cluster-specific and complement other clusters, suggesting
-convergence. Our observations suggest that driver genes combine with uncommon
-genes in tumor-promoting pathways to produce distinct molecular phenotypes.
-Thus, we present a coherent view of global tumor biology, and explain how common
-cellular dysfunction might arise from tissue-specific mutations.
+Cancer is an organism-level disease, impacting processes from cellular
+metabolism and the microenvironment to systemic immune response. Nevertheless,
+efforts to distinguish overarching mutational processes from interactions with
+the cell of origin for a tumor have seen limited success, presenting a barrier
+to individualized medicine. Here we present a novel, pathway-centric approach,
+extracting somatic mutational profiles within and between tissues, largely
+orthogonal to cell of origin, mutational burden, or stage. Known predisposition
+variants are equally distributed among clusters, and largely independent of
+molecular subtype. Prognosis and risk of death vary jointly by cancer type and
+cluster. Analysis of metastatic tumors reveals that differences are largely
+cluster-specific and complementary, implicating convergent mechanisms that
+combine familiar driver genes with diverse low-frequency lesions in
+tumor-promoting pathways, ultimately producing distinct molecular phenotypes.
+The results shed new light on the interplay between organism-level dysfunction
+and tissue-specific lesions.

### 2020-08-31 text/manuscript.Rmd
add Tim to acknowledgements

@@
 We wish to thank David Van Valen, Kate Lawrenson, Simon Knott, and Megan
-Hitchens for critical reading of this manuscript, and Ivetth Corona for early
+Hitchens, for critical comments on this manuscript, and Tim Triche Jr for
+suggested improvements to the preprint. We also thank Ivetth Corona for early
 discussions and feedback.

### 2020-08-31 code/figure4.Rmd
Fix error with figure 4

@@
 ```{r forest-heatmap-data-plot, fig.align='center', fig.width= 9, fig.height= 8, message=FALSE}
-#fig.width= 16, fig.height= 20,
 load("../data/forest.heatmap.data.rda")
@@
   slice_max(order_by = or, n = 10) %>%
-  arrange(cluster, or, .by.group = TRUE)
+  arrange(cluster, or, .by.group = TRUE)))
 down_compare_plot$Hugo_Symbol <- with(down_compare_plot,
@@
 all_p + tdat_all_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
-                   color = cluster,
-                   group = cluster)) +
-  geom_vline(xintercept = 0, linetype = 2) +
-  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
-  scale_y_discrete(labels = Hugo_rownames) +
-  scale_color_manual(values = cluster_colors) +
-  theme_classic() +
-  xlab("log odds ratio") +
-  ylab("Gene") +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y", switch = "y") +
-  theme(
-    panel.grid.major.y = element_line(colour = "#f0f0f0"),
-    axis.title.y = element_blank()
-  )
-
-tdat_all <- all_muts %>%
-  filter(Hugo_Symbol %in% all_compare_plot$Hugo_Symbol) %>%
-  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
-  select(type, clust_knn_char, Hugo_Symbol) %>%
-  table()
-tdat_all <- left_join(
-  as.data.frame(tdat_all),
-  clinical_plus_cluster %>%
-    select(Tumor_Sample_Barcode, type, clust_knn_char) %>%
-    distinct() %>%
-    group_by(type, clust_knn_char) %>%
-    summarise(ntumors = n())) %>%
-  group_by(type, clust_knn_char) %>%
-  mutate(c_prop = Freq / ntumors) %>%
-  mutate(c_prop = ifelse(is.na(c_prop), 0, c_prop)) %>%
-  select(type, clust_knn_char, Hugo_Symbol, c_prop) %>%
-  rename(cluster = clust_knn_char)
-tdat_all <- all_compare_plot %>% left_join(tdat_all)
-tdat_all_p <- tdat_all %>% ggplot(aes(type, Row_Index, fill = c_prop)) +
-  geom_tile() +
-  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
-  xlab("Tumor Type") +
-  ylab("Gene Symbol") +
-  scale_y_discrete(labels = Hugo_rownames) +
-  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
-  theme_classic() +
-  theme(
-    strip.text = element_blank(),
-    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
-    axis.text.y = element_blank(),
-    axis.title.y = element_blank(),
-    panel.grid.major.y = element_line(colour = "#f0f0f0")
-  )
-
-layout <- "ABB"
-all_p + tdat_all_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
 ```

### 2020-08-31 code/figure4.Rmd

@@
 ```{r forest-heatmap-data-plot, fig.align='center', fig.width= 9, fig.height= 8, message=FALSE}
+#fig.width= 16, fig.height= 20,
 load("../data/forest.heatmap.data.rda")
@@
   arrange(cluster, or, .by.group = TRUE)
-down_compare_plot$Hugo_Symbol <- with(down_compare_plot,
-                                      factor(as.character(Hugo_Symbol),
-                                             levels = unique(as.character(Hugo_Symbol))))
-
-up_compare_plot <- all_compare %>%
-  group_by(cluster) %>%
-  filter(or > 0) %>%
-  arrange(-or) %>%
-  filter(is.finite(or)) %>%
-  slice_max(order_by = or, n = 10) %>%
-  arrange(cluster, or, .by.group = TRUE)
-up_compare_plot$Hugo_Symbol <- with(up_compare_plot,
-                                    factor(as.character(Hugo_Symbol),
-                                           levels = unique(as.character(Hugo_Symbol))))
-
-all_compare_plot <- bind_rows(down_compare_plot, up_compare_plot)
-all_compare_plot %<>% arrange(or, .by_group = TRUE) %>% mutate(Hugo_Symbol = as.character(Hugo_Symbol)) %>% ungroup() %>% mutate(Row_Index = sprintf("%3i", row_number()))
-Hugo_rownames <- all_compare_plot$Hugo_Symbol
-names(Hugo_rownames) <- as.character(all_compare_plot$Row_Index)
-all_p <- ggplot(all_compare_plot, 
-               aes(x = or,
-                   y = Row_Index,down_compare_plot <- all_compare %>%
-  group_by(cluster) %>%
-  filter(or < 0) %>%
-  filter(adjPval < 0.01) %>%
-  arrange(-or) %>%
-  filter(is.finite(or)) %>%
-  slice_max(order_by = or, n = 10) %>%
-  arrange(cluster, or, .by.group = TRUE)))
 down_compare_plot$Hugo_Symbol <- with(down_compare_plot,

### 2020-09-12 code/figure1.Rmd
Fix fig1 heatmap clip off, fix fig2 metastasis color annotation, add url to shinyapp in manuscript and supplemental_methods

@@
         legend.box.margin=margin(-10,-10,0,-10),
-        plot.margin = unit(c(0.6, -3, 0, 0), "cm"),
+        plot.margin = unit(c(0.6, 0, 0, 0), "cm"),
         panel.background = element_rect(fill = "transparent"))

### 2020-09-12 code/figure2.Rmd
Fix fig1 heatmap clip off, fix fig2 metastasis color annotation, add url to shinyapp in manuscript and supplemental_methods

@@
      scatter3D(plot_y, -plot_z, plot_x, 
-               bg = B_color, pch = 21, cex = 0.8, lwd = 0.2, 
+               bg = B_color[as.character(Subtype_Selected)], pch = 21, cex = 0.8, lwd = 0.2, 
                theta = 0, phi = 65, scale = F,
@@
      scatter3D(plot_y, -plot_z, plot_x, 
-               bg = C_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               bg = C_colors[as.character(mut_type)], pch = 21, cex = 0.8, lwd = 0.2,
                theta = 0, phi = 65, scale = F,
@@
      scatter3D(plot_y, -plot_z, plot_x, 
-               bg = D_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               bg = D_colors[as.character(ajcc_pathologic_tumor_stage)], pch = 21, cex = 0.8, lwd = 0.2,
                theta = 0, phi = 65, scale = F,
@@
      scatter3D(plot_y, -plot_z, plot_x, 
-               bg = E_colors, pch = 21, cex = 0.8, lwd = 0.2,
+               bg = E_colors[as.character(ajcc_metastasis_pathologic_pm)], pch = 21, cex = 0.8, lwd = 0.2,
                theta = 0, phi = 65, scale = F,

### 2020-09-12 code/clustering.Rmd

```{r data-set-creation, eval=FALSE}
## Data acquired from GDC
tcga_ssm <- fread("../data/simple_somatic_mutation.open.tsv.gz", header = T, sep = "\t")
## Somatic mutations focused in analysis
selected_mut <- c("3_prime_UTR_variant", "5_prime_UTR_premature_start_codon_gain_variant", "5_prime_UTR_variant", "frameshift_variant", "missense_variant", "splice_acceptor_variant", "splice_donor_variant", "stop_lost", "start_lost", "stop_gained")
## Filtering tcga_ssm for our selected mutations
tcga_ssm <- tcga_ssm[tcga_ssm$consequence_type %in% selected_mut, ]
## Selecting column that we are interested in
tcga_ssm <- data.frame(chr = paste0("chr",tcga_ssm$chromosome),
                   start = tcga_ssm$chromosome_start,
                   end = tcga_ssm$chromosome_end,
                   ref_allele = tcga_ssm$reference_genome_allele,
                   alt_allele = tcga_ssm$mutated_to_allele,
                   type = tcga_ssm$consequence_type,
                   mut_type = tcga_ssm$mutation_type,
                   ensembl_id = tcga_ssm$gene_affected,
                   donor_id = tcga_ssm$icgc_donor_id,
                   sample_id = tcga_ssm$submitted_sample_id,
                   stringsAsFactors = F)
## Removing duplicate rows
tcga_ssm <- tcga_ssm %>% distinct()
## Recategorizing the mutation type
tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "type"] <- tcga_ssm[which(tcga_ssm$type == "frameshift_variant"), "mut_type"] 
tcga_ssm <- tcga_ssm %>% mutate(type = case_when(type == "insertion of <=200bp" ~ "frameshins_variant", ## relabel values
                                                 type == "deletion of <=200bp" ~ "frameshdel_variant",
                                                 type == "splice_acceptor_variant" ~ "splice_variant",
                                                 type == "splice_donor_variant" ~ "splice_variant",
                                                 type == "start_lost" ~ "nonsense_variant",
                                                 type == "stop_gained" ~ "nonsense_variant",
                                                 type == "5_prime_UTR_premature_start_codon_gain_variant" ~ "UTR_5_prime_variant",
                                                 type == "5_prime_UTR_variant" ~ "UTR_5_prime_variant",
                                                 type == "3_prime_UTR_variant" ~ "UTR_3_prime_variant",
                                                 type == "missense_variant" ~ "missense_variant",
                                                 type == "stop_lost" ~ "nonstop_variant")) %>% 
  dplyr::select(-mut_type) ## rm mut_type col

## Preparing sample metadata ##

complete_sample <- tcga_ssm %>% .[,"sample_id"] %>% unique() %>% .[grep(pattern= "TCGA-", x = .)] # complete TCGA tumor samples

tcga_sample <- fread("../data/sample.tsv.gz", header = T, sep = "\t") ## loading in sample metadata
tcga_sample <- tcga_sample %>% dplyr::select(project_code, submitted_sample_id, submitted_specimen_id, icgc_donor_id, submitted_donor_id, study) %>% ## selecting the relevant columns
  .[.$submitted_sample_id %in% complete_sample,] %>% ## Only include samples in complete_sample
  mutate(type = substring(.$submitted_sample_id, first = 14, last = 15)) %>% ## Getting "sample type" info from TCGA barcode
  filter(type == "01") ## filtering 01 samples (primary solid tumor)
tcga_dup_id <- tcga_sample %>% .[.$icgc_donor_id %in% .[which(duplicated(.$submitted_donor_id)),]$icgc_donor_id,] %>% ## what are the dup donors
  filter(study == "PCAWG") ## selecting dup samples that are from PCAWG
tcga_sample <- tcga_sample %>% .[!.$icgc_donor_id %in% tcga_dup_id$icgc_donor_id, ] %>% ## remove the dups from tcga_samples
  rbind(., tcga_dup_id) ## rbind the fixed dups

tcga_donor <- fread("../data/donor.tsv.gz", header = T, sep = "\t") ## loading in sample survival data
tcga_donor <- tcga_donor %>% dplyr::select(icgc_donor_id, project_code, submitted_donor_id, donor_sex, donor_vital_status, disease_status_last_followup, donor_age_at_diagnosis, donor_age_at_enrollment, donor_age_at_last_followup, donor_survival_time, donor_interval_of_last_followup) ## selecting columns that we are interested in

sample_sheet <- left_join(tcga_sample, tcga_donor, by = c("submitted_donor_id", "project_code", "icgc_donor_id")) ## combining the tcga sample and donor data

## Preparing gene metadata ##

# genelist <- unique(as.character(tcga_ssm$ensembl_id)) ## complete genelist
# ensembl_2_entrez_full_list <- bitr(genelist, 
#                                    fromType = "ENSEMBL", 
#                                    toType = c("ENTREZID"), OrgDb = org.Hs.eg.db)
load("../data/ensembl_2_entrez_full_list.Rda")
tcga_ssm <- tcga_ssm[tcga_ssm$ensembl_id %in% ensembl_2_entrez_full_list$ENSEMBL,] ## Only include genes in the gdc data that is found in the ensembl_2_entrez_full_lust

## Now we only have 19067 genes
complete_ensembl <- as.data.frame(unique(as.character(tcga_ssm$ensembl_id))) ## ensembl gene list
names(complete_ensembl) <- "ensembl_id"

selected_mut <- unique(as.character(tcga_ssm$type))
## For every mutation type.. select them.. and create a tsv file 
for(selected_mut in selected_mut) {
  print(selected_mut)
  df <-  tcga_ssm %>% filter(type == selected_mut) %>% ## filter data by selected_mut
    dplyr::select(sample_id, ensembl_id) %>% ## select column sample_id and ensembl_id
    mutate(mutation = 1) %>% ## create a new column with 1's
    distinct() %>% ## remove duplicated rows
    spread(., key = "sample_id", value = "mutation") %>% ## long to wide format
    remove_rownames() %>% ## move ensembl_id column to rownames
    column_to_rownames(var = "ensembl_id") %>% 
    replace(is.na(.), values = 0) %>% ## convert NA's to 0's
    .[, names(.) %in% sample_sheet$submitted_sample_id] ## rm any sample_id not included in complete_sample
  missing_gene <- as.data.frame(matrix(data = 0, nrow = length(complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)]), ncol = length(df)))
  rownames(missing_gene) <- complete_ensembl$ensembl_id[!complete_ensembl$ensembl_id %in% rownames(df)] ## insert missing ensembl_id into row names
  names(missing_gene) <- names(df) ## insert sample_id names into col names
  df <- rbind(df, missing_gene) ## rbind df and missing_gene
  rm(missing_gene) ## rm data
  gc()
  df <- df[match(complete_ensembl$ensembl_id, rownames(df)),] ## match df rownames with genelist
  missing_sample <- as.data.frame(matrix(data = 0, nrow = nrow(df), ncol = length(sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)])))
  rownames(missing_sample) <- rownames(df) ## insert ensembl_id into row names
  names(missing_sample) <- sample_sheet$submitted_sample_id[!sample_sheet$submitted_sample_id %in% names(df)] ## insert missing sample_id names into col names
  df <- cbind(df, missing_sample) ## rbind df and missing_gene
  rm(missing_sample) ## rm data
  gc()
  df <- df[, match(sample_sheet$submitted_sample_id, names(df))]
  assign(paste0(selected_mut), value = df) ## create df in r environment
  rm(df)
  gc()
  }

# ensembl <- useEnsembl(biomart="ensembl", dataset = "hsapiens_gene_ensembl")
 # genelist <- rownames(missense_variant)
 # gene_data <- getBM(
 #   attributes = c(
 #     'ensembl_gene_id',
 #     'external_gene_name',
 #     'gene_biotype',
 #     'chromosome_name', 
 #     'start_position',
 #     'end_position', 
 #     'strand'),
 #   filters = 'ensembl_gene_id',
 #   values = genelist,
 #   mart = ensembl)
 # save(gene_data, file = "gene_data.Rda")
load("../data/gene_data.Rda")
###################### Converting ensembl id to gene symbol

## Saniety check to make sure all the genes in the mutation data is found in the gene_data. If not, 
## then only include the genes in gene_data
missense_variant <- missense_variant[rownames(missense_variant) %in% gene_data$ensembl_gene_id,]
UTR_3_prime_variant <- UTR_3_prime_variant[rownames(UTR_3_prime_variant) %in% gene_data$ensembl_gene_id,]
UTR_5_prime_variant <- UTR_5_prime_variant[rownames(UTR_5_prime_variant) %in% gene_data$ensembl_gene_id,]
frameshdel_variant <- frameshdel_variant[rownames(frameshdel_variant) %in% gene_data$ensembl_gene_id,]
frameshins_variant <- frameshins_variant[rownames(frameshins_variant) %in% gene_data$ensembl_gene_id,]
splice_variant <- splice_variant[rownames(splice_variant) %in% gene_data$ensembl_gene_id,]
nonstop_variant <- nonstop_variant[rownames(nonstop_variant) %in% gene_data$ensembl_gene_id,]
nonsense_variant <- nonsense_variant[rownames(nonsense_variant) %in% gene_data$ensembl_gene_id,]

## Matching df to the ensembl_id orders in gene_data
match_ensembl <- function(x) {
  x <- x[match(gene_data$ensembl_gene_id, rownames(x)),] 
}

missense_variant <- match_ensembl(missense_variant)
UTR_3_prime_variant <- match_ensembl(UTR_3_prime_variant)
UTR_5_prime_variant <- match_ensembl(UTR_5_prime_variant)
frameshdel_variant <- match_ensembl(frameshdel_variant)
frameshins_variant <- match_ensembl(frameshins_variant)
splice_variant <- match_ensembl(splice_variant)
nonstop_variant <- match_ensembl(nonstop_variant)
nonsense_variant <- match_ensembl(nonsense_variant)

## Convert gene_data into GRanges for Summarized Experiment
genes <- GRanges(
  seqnames = paste("chr", gene_data$chromosome_name, sep=""),
  ranges = IRanges(
    start = gene_data$start_position,
    end = gene_data$end_position),
  strand = gene_data$strand,
  mcols = list(
    ensembl_gene_id = gene_data$ensembl_gene_name,
    hgnc_symbol = gene_data$hgnc_symbol))
## Create Summarized Experiment
gdc_tumor_coding_complete <- SummarizedExperiment(
  assays = list(
    missense   = missense_variant,
    splice     = splice_variant,
    frameshdel = frameshdel_variant,
    frameshins = frameshins_variant,
    utr3       = UTR_3_prime_variant,
    utr5       = UTR_5_prime_variant,
    nonsense   = nonsense_variant,
    nonstop    = nonstop_variant),
  rowRanges = genes,
  colData = sample_sheet)
save(gdc_tumor_coding_complete, file = "../data/gdc_tumor_coding_complete.Rdata")
rm(list = ls(all.names = T))
gc()
```

```{r reactome-pathway-creation, eval=FALSE}
load("../data/gene_data.Rda")
## Preparing level_0 to level_1 pathway conversion table. level_0 is the master pathway. level_1 is the parent pathway
main2first <- fread(file = "../data/reactome_main2first_level.txt", header = T, sep = "\t")

main2first <- main2first %>% .[,2:3] %>%
  'colnames<-' (c("first_level", "main_level")) ## rename colnames

## Filtering out master pathways from first2second_level
first2second <- fread(file = "../data/reactome_first2second_level.txt", header = F, sep = "\t")
first2second <- first2second %>% 'colnames<-' (c("first_level", "second_level")) %>% ## rename colnames
  .[grep(pattern = "-HSA-", .$first_level)] %>% ## keep human pathways
  .[!.$first_level %in% unique(main2first$main_level),] %>% ## remove main level pathways from first level
  .[!.$first_level %in% unique(.$second_level),] ## remove second level pathways from first level

ensembl2reactome <- fread(file = "../data/Ensembl2Reactome_All_Levels.txt", header = F, sep = "\t") ## ensembl to pathway list acquired from reactome database
ensembl2reactome <- ensembl2reactome %>% .[,c(1:2,4)] %>% ## select columns ensembl, pathway_id, pathway_name columns
  'colnames<-' (c("ENSEMBL", "PATH_ID", "PATH_NAME")) %>% ## rename colnames
  .[grep(pattern = "-HSA-", .$PATH_ID),] %>% ## keep human pathways
  .[.$PATH_ID %in% first2second$second_level,] %>% ## include only second level pathways
  left_join(., main2first, by = c("PATH_ID" = "first_level")) %>% distinct() %>% 
  left_join(., ensembl2reactome[,c(2,4)], by = c("main_level" = "V2")) %>% distinct()

## Filtering pathways not relevant to our analysis by keywords
filtered_ensembl2reactome <- ensembl2reactome %>% filter(!V4 %in% c("Disease", "Muscle contraction", "Reproduction")) ## Pathways in disease, muscle contraction and reproduction parent pathway were removed

toMatch <- c("disease", "defect", "cancer", "oncogenic", "disorder", "infection", "mutant", "bacteria", "listeria", "loss") ## pathways with the following key terms were removed.

filtered_ensembl2reactome <- filtered_ensembl2reactome %>% mutate(l_pathname = tolower(.$PATH_NAME)) %>% ## lower casing the pathway names because some pathways have disease vs Disease
  .[!grep(paste(toMatch,collapse="|"), .$l_pathname),] ## Removing pathways with the key terms in toMatch

filtered_ensembl2reactome <- filtered_ensembl2reactome %>% .[which(!.$PATH_ID %in% c("R-HSA-212436", "R-HSA-388396")),] ## Removing Generic Transcription Pathway and GPCR downstream signalling as they are broad pathways

gene2level2_lookup <- filtered_ensembl2reactome %>% dplyr::select(ENSEMBL, PATH_ID) %>% ## remove PATH_NAME col
  .[.$ENSEMBL %in% gene_data$ensembl_gene_id,] %>% ## only include genes that are in the gene_data 
  mutate(member = 1) %>% ## add a column for ensembl to pathway membership
  distinct() %>% ## keep unique rows
  spread(data = ., key = "PATH_ID", value = "member") %>% ## long to wide format
  column_to_rownames(var = "ENSEMBL") %>% ## ensembl_id as rownames
  replace(is.na(.), values = 0) ## replace NA's with 0's
#save(gene2level2_lookup, file = "../data/gene2_level2_lookup.Rda")
## Pathway names to level 2 lookup table
pathname2level2 <- filtered_ensembl2reactome %>% dplyr::select(PATH_ID, PATH_NAME) %>% ## remove ENSEMBL col
  .[.$PATH_ID %in% names(gene2level2_lookup),] %>% ## only include pathways in gene2level2_lookup
  distinct(.) ## keep unique rows
#save(pathname2level2, file = "../data/pathname2level2.Rda")
rm(list = ls(all.names = T))
gc()
```

```{r filter-tissue-specific-low-expressed-genes, eval=FALSE}
gtf_file <- "../data/gencode.v30.annotation.gtf" ## loading gene information downloaded from gencode
gencode_gtf <- rtracklayer::import(gtf_file) ## import data
gene_info <- as.data.frame(mcols(gencode_gtf)[, c("type", "gene_id", "transcript_id", "gene_type", "transcript_type", "gene_name")]) ## selecting relevant columns
## We only want transcript only 
gene_info <- gene_info[gene_info$type == "transcript",]
## We only want protein_coding only
gene_info <- gene_info[gene_info$gene_type == "protein_coding",]
gene_info <- gene_info[gene_info$transcript_type == "protein_coding",]
## Filtering for transcript_id, gene_id, gene_name columns
g2t <- gene_info %>% dplyr::filter(type == "transcript") %>% dplyr::select(transcript_id, gene_id, gene_name)
## Condense df to only having unique data. 
good.g2t <- as_tibble(unique(g2t))

## df with transcript_id and entrez id                      
gencode_entrez <- readr::read_tsv("../data/gencode.v30.metadata.EntrezGene.gz", col_names = FALSE)
colnames(gencode_entrez) <- c("transcript_id", "entrez_id")
## Joining data that are joined together by transcript_id
entrez_g2t <- left_join(good.g2t, gencode_entrez)
head(entrez_g2t[1:10,])
## Remove transcript_id column
entrez_g2t$transcript_id <- NULL
## Remove . and the numbers following from the ensembl id.
entrez_g2t$gene_id <- gsub("\\..*", "", entrez_g2t$gene_id)
## Removing data that does not have entrez id. 83091 out of 83647 had NA values
table(is.na(entrez_g2t$entrez_id))
entrez_g2t <- entrez_g2t[!is.na(entrez_g2t$entrez_id),]
entrez_g2t <- unique(entrez_g2t[, c("gene_id", "gene_name", "entrez_id")])
## Are there any duplicates in our df? Yes. Only one dupped ensembl id.
table(duplicated(entrez_g2t$gene_id))
## These are pseudogenes... We will remove them all.
dups_gene_id <- entrez_g2t[which(duplicated(entrez_g2t$gene_id)),]
## We're removing the rows that are associated with this pseudogene
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285607")),]
## It's gone
which(entrez_g2t$gene_id == "ENSG00000285607")
## Are there any duplicate gene names? Yes! 14 of them
table(duplicated(entrez_g2t$gene_name))
dups_gene_name <- entrez_g2t[which(duplicated(entrez_g2t$gene_name)),]
entrez_g2t[entrez_g2t$gene_name == "TBCE",]
## ENSG00000285053  is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285053")),]
entrez_g2t[entrez_g2t$gene_name == "PDE11A",]
## ENSG00000284741 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284741")),]
entrez_g2t[entrez_g2t$gene_name == "ATXN7",]
## ENSG00000285258 is not in the database anywhere
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285258")),]
entrez_g2t[entrez_g2t$gene_name == "MATR3",]
## ENSG00000280987 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000280987")),]
entrez_g2t[entrez_g2t$gene_name == "SOD2",]
## ENSG00000285441 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285441")),]
entrez_g2t[entrez_g2t$gene_name == "ABCF2",]
## ENSG00000285292 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285292")),]
entrez_g2t[entrez_g2t$gene_name == "PINX1",]
## ENSG00000258724 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000258724")),]
entrez_g2t[entrez_g2t$gene_name == "HSPA14",]
## ENSG00000284024 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284024")),]
entrez_g2t[entrez_g2t$gene_name == "IGF2",]
## ENSG00000284779 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284779")),]
entrez_g2t[entrez_g2t$gene_name == "ATF7",]
## ENSG00000267281 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000267281")),]
entrez_g2t[entrez_g2t$gene_name == "DIABLO",]
## ENSG00000284934 is a homolog
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284934")),]
entrez_g2t[entrez_g2t$gene_name == "GGT1",]
## ENSG00000286070 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286070")),]
entrez_g2t[entrez_g2t$gene_name == "SCO2",]
## ENSG00000130489 no longer exist
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000130489")),]
entrez_g2t[entrez_g2t$gene_name == "TMSB15B",]
## ENSG00000269226 is a readthrough
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000269226")),]

## Are there any dupped entrez id? 
dups_entrez <- entrez_g2t[which(duplicated(entrez_g2t$entrez_id)),]

entrez_g2t[entrez_g2t$entrez_id == "55486",]
## ENSG00000283765 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283765")),]
entrez_g2t[entrez_g2t$entrez_id == "55300",]
## ENSG00000281028 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000281028")),]
entrez_g2t[entrez_g2t$entrez_id == "653082",]
## ENSG00000286094 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286094")),]
entrez_g2t[entrez_g2t$entrez_id == "8622",]
## ENSG00000284762 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000284762")),]
entrez_g2t[entrez_g2t$entrez_id == "221468",]
## ENSG00000286105 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286105")),]
entrez_g2t[entrez_g2t$entrez_id == "1124",]
## ENSG00000285162 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285162")),]
entrez_g2t[entrez_g2t$entrez_id == "102723849",]
## ENSG00000283765 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000273520")),]
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000262461")),]
entrez_g2t[entrez_g2t$entrez_id == "883",]
## ENSG00000286112 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286112")),]
entrez_g2t[entrez_g2t$entrez_id == "22891",]
## ENSG00000261130 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285551")),]
entrez_g2t[entrez_g2t$entrez_id == "54816",]
## ENSG00000285253 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285253")),]
entrez_g2t[entrez_g2t$entrez_id == "23370",]
## ENSG00000268861 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000268861")),]
entrez_g2t[entrez_g2t$entrez_id == "284391",]
## ENSG00000286098 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000286098")),]
entrez_g2t[entrez_g2t$entrez_id == "440519",]
## ENSG00000283201 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000283201")),]
entrez_g2t[entrez_g2t$entrez_id == "149951",]
## ENSG00000285382 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000285382")),]
entrez_g2t[entrez_g2t$entrez_id == "388820",]
## ENSG00000268861 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000278961")),]
entrez_g2t[entrez_g2t$entrez_id == "8471",]
## ENSG00000286098 not found on ncbi
entrez_g2t <- entrez_g2t[-c(which(entrez_g2t$gene_id == "ENSG00000260548")),]

## manual gene removal complete

load("../data/TCGA.RNA.Rda") ## loading in mRNA expression data from UCSC xena browser

## Put entrez ids as rownames and shortening the TCGA ids since the last three portions of the TCGA barcode is coded to explain the test performed and etc. 
tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>% 
  as.data.frame(.) %>%
  dplyr::select(project_code, submitted_specimen_id) %>%
  mutate(submitted_specimen_id = substring(text = .$submitted_specimen_id, first = 1, last = 15))

TCGA.RNA <- TCGA.RNA %>% dplyr::select(-gene_symbol) %>% ## Don't need gene symbol info
  remove_rownames(.) %>% column_to_rownames(var = "gene_id") %>%  ## entrez id to rownames
  .[rownames(.) %in% entrez_g2t$entrez_id,] %>% ## Filter TCGA data to only include genes found in gencode (entrez_g2t)
  'colnames<-' (substring(text = names(.), first = 1, last = 15)) %>% ## only need the first 15 character of barcode
  .[, names(.) %in% tumor_samples$submitted_specimen_id] ## only include tumor samples in our analysis

g2e_list <- entrez_g2t[entrez_g2t$entrez_id %in% rownames(TCGA.RNA),] ## getting the list of genes in entrez_g2t thats in TCGA.RNA

express_no_express_lookup <- list()
for(i in unique(tumor_samples$project_code)) {
  df <- which(tumor_samples$project_code == i) ## which rows are the spec cancer types
  df <- tumor_samples[df,] ## select only those rows
  gdc <- TCGA.RNA[, names(TCGA.RNA) %in% df$submitted_specimen_id] ## filter the norm_gdc_gene_exp dataset with those tumor samples
  mean_gdc <- as.data.frame(apply(gdc, MARGIN = 1, function(x) {
    mean(x)
  })) ## take the mean for each gene across the tumor samples
  names(mean_gdc) <- i
  express_no_express_lookup[[i]] <- mean_gdc ## store data in list
  rm(df, gdc, mean_gdc)
  gc
}
express_no_express_lookup <- do.call(cbind, express_no_express_lookup) ## cbind list
express_no_express_lookup[is.na(express_no_express_lookup)] <- 0 ## NA values are 0

gdc_ensembl_id <- gdc_tumor_coding_complete %>% rowRanges(.) %>% as.data.frame(.) %>% ## getting the ensembl_ids in our gdc dataset
  rownames(.)
g2e_list <- g2e_list[g2e_list$gene_id %in% gdc_ensembl_id,] ## subsetting the g2e_list to only have ensembl_ids in our gdc dataset

express_no_express_lookup <- express_no_express_lookup[rownames(express_no_express_lookup) %in% g2e_list$entrez_id, ] ## Saniety check to only include genes in express lookup that are in g2e_list
express_no_express_lookup[express_no_express_lookup <= 10] <- 0 ## Set threshold to <=10 for not expressed
express_no_express_lookup[express_no_express_lookup >10] <- 1 ## >10 is expressed

## convert entrez id to ensembl_id
express_no_express_lookup <- express_no_express_lookup %>% .[match(g2e_list$entrez_id, rownames(.)),] %>% ## match the entrez_id express_no_express_lookup with g2e_list
  'rownames<-' (g2e_list$gene_id) ## replace entrez_id with

## Booleanize mutation count for each tumor sample
missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## Load in each mut type data
splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)

gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
gc()
gdc_mut_count_patient <- gdc_mut_count_patient[rownames(gdc_mut_count_patient) %in% rownames(express_no_express_lookup),] ## filter data with genes in the express_no_express_lookup
gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 0 is equal to 1

gdc_gene_exp_adj <- list()
for(i in unique(tumor_samples$project_code)) {
  df <- which(tumor_samples$project_code == i) ## which rows are my spec cancer types
  df <- tumor_samples[df,] ## select only those rows
  gdc <- gdc_mut_count_patient[, names(gdc_mut_count_patient) %in% rownames(df)] ## filter the norm_gdc_gene_exp dataset with those tumor samples
  exp_data <- as.data.frame(express_no_express_lookup[, i]) ## select correct lookup column for the cancer type
  rownames(exp_data) <- rownames(express_no_express_lookup) ## Insert rownames to exp_data
  gdc <- gdc[match(rownames(exp_data), rownames(gdc)),] ## Match ensembl_id of exp_data with gdc data
  gdc_adj <- as.data.frame(apply(gdc, MARGIN = 2, function(x) {
    exp_data * x
  })) ## Function to multiply lookup to the data for the specific cancer type
  names(gdc_adj) <- names(gdc) ## Rename gdc_adj with tumor samples names
  gdc_gene_exp_adj[[i]] <- gdc_adj ## Store data to list
  rm(df, gdc, exp_data, gdc_adj)
  gc()
}
## Combining list into a dataframe.
gdc_gene_exp_adj <- do.call(cbind, gdc_gene_exp_adj) ## cbind list
names(gdc_gene_exp_adj) <- gsub("^.*\\.", "", names(gdc_gene_exp_adj)) ## tumor sample names have cancer type name in it. Removing it using gsub
rm(gdc_mut_count_patient, express_no_express_lookup, TCGA.RNA, gdc_tumor_coding_complete, gencode_gtf, dups_entrez, dups_gene_id, dups_gene_name, g2t, good.g2t, gene_info, g2e_list, entrez_g2t, gencode_entrez)
gc()
save(gdc_gene_exp_adj, file = "gdc_gene_exp_adj.Rda")
```

```{r genes-to-pathways-transformation, eval=FALSE}
load("../data/gdc_gene_exp_adj.Rda")
load("../data/gene2_level2_lookup.Rda")
sample_list <- as.character(names(gdc_gene_exp_adj)) ## Create sample list
complete_path <- as.character(names(gene2level2_lookup)) ## Get names of pathways

df_gene <- as.data.frame(t(gdc_gene_exp_adj)) ## Transpose data so samples are rows and genes are columns
df_gene <- df_gene > 0 ## if 1 == TRUE if 0 == FALSE
df_gene <- as.data.frame(df_gene) ## turn matrix back to df

df_gene <- df_gene %>% rownames_to_column(var = "sample") %>% 
  gather(key = "gene", value = "mutation", -sample) %>% ## Wide to long format
  .[.$mutation,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
  dplyr::select(-mutation) ## Remove column named mutation

path_lookup <- as.data.frame(t(gene2level2_lookup)) ## Transpose data so paths are rows and genes are columns
path_lookup <- path_lookup > 0 ## if 1 == TRUE if 0 == FALSE
path_lookup <- as.data.frame(path_lookup) ## turn matrix back to df

path_lookup <- path_lookup %>% rownames_to_column(var = "path") %>%
  gather(key = "gene", value = "path_member", -path) %>% ## Wide to long format
  .[.$path_member,] %>% ## Keep TRUE's (which are 1's); Remove FALSE's (0's)
  dplyr::select(-path_member)

df_lookup <- df_gene %>% left_join(., path_lookup, by = "gene") %>% ## Match df_gene and lookup dataframes
  na.omit(.) ## Omit NA's that were generated due to no matches

## getting the genes involved in our analysis
genes_in_pathanalysis <- unique(df_lookup$gene)
write.table(genes_in_pathanalysis, file = "genes_in_pathanalysis.tsv", quote = FALSE, sep = "\t", row.names =  FALSE) ## save data

df_lookup <- as.data.frame(table(df_lookup$sample, df_lookup$path)) ## Generate 1's for the matches

df_lookup <- df_lookup %>% spread(key = "Var1", value = "Freq") %>% ## Long to wide format
  remove_rownames() %>%
  column_to_rownames(var = "Var2") ## path_id to rownames

## Insert missing tumor samples due to no mutations
df_missing <- sample_list[!(sample_list %in% names(df_lookup))] 
mx <- as.data.frame(matrix(data = 0, nrow = nrow(df_lookup), ncol = length(df_missing))) ## Create mx to store data
names(mx) <- df_missing ## Missing tumor sample names in col names
rownames(mx) <- rownames(df_lookup) ## Path names in rownames
df_lookup <- cbind(df_lookup, mx) ## cbind data

## Insert missing pathway due to no mutations
path_missing <- complete_path[!c(complete_path %in% rownames(df_lookup))]
mx <- as.data.frame(matrix(data = 0, nrow = length(path_missing), ncol = ncol(df_lookup))) ## Create mx to store data
names(mx) <- names(df_lookup) ## Insert tumor sample names
rownames(mx) <- path_missing ## Insert missing path names
df_lookup <- rbind(df_lookup, mx) ## cbind data

pathway_order <- rownames(df_lookup) ## need to perserve the order of pathway for the next operation
gdc_reactome_path_analysis <- df_lookup %>% mutate_if(is.numeric, ~1 * (. > 0)) %>% as.data.frame() %>% ## Mutate anything >0 to 1's (Booleanize data)
  'rownames<-' (pathway_order) %>% ## rename rows with pathway order names
  rownames_to_column(var = "PATH_ID") %>% ## move rownames to column
  left_join(., pathname2level2, by = "PATH_ID") %>% ## leftjoin pathname with the data
  dplyr::select(-PATH_ID) %>% ## get rid of the path id 
  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") ## move pathnameto rownames
rm(df_missing, path_missing, mx, df_gene, sample_list, complete_path, pathway_order, df_lookup)
gc()
save(gdc_reactome_path_analysis, file = "gdc_reactome_path_analysis.Rda")
```

```{r tumor-clusters, eval = FALSE}
###################
## Figure 1 data ##
###################
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene2_level2_lookup.Rda")

## Column distance metric for pheatmap
path_daisy <- daisy(t(gdc_reactome_path_analysis), type = list(symm  = c(1:col(gdc_reactome_path_analysis)))) 
path_daisy <- as.dist(as.matrix(path_daisy))
## Row distance metric for pheatmap
path_daisy_row <- daisy(gdc_reactome_path_analysis, type = list(symm  = c(1:nrow(gdc_reactome_path_analysis))))
path_daisy_row <- as.dist(as.matrix(path_daisy_row))

pathway_dendro <- pheatmap(gdc_reactome_path_analysis,
         show_rownames = F,
         show_colnames = F,
         cellheight = 1,
         cellwidth = 0.5,
         clustering_distance_cols = path_daisy,
         clustering_distance_rows = path_daisy_row,
         cluster_cols = T,
         cluster_rows = T,
         cutree_cols = 19,
         annotation_legend = F,
         legend_breaks = c(0, 0.25, 0.75, 1),
         legend_labels = c("","Not Mut", "Mut", ""),
         color = c("grey", "firebrick3"),
         clustering_method = "ward.D2", ## ward.D2 clustering
         legend = T)
dev.off()
pathway_dendro <- pathway_dendro$tree_row$order ## getting pathway row ordering (numerical values)
df <- rownames(gdc_reactome_path_analysis) ## getting pathway name from the pathway analysis
pathway_dendro <- df[pathway_dendro] ## Reorder the pathway names based on the ordering from pheatmap

tumor_samples <- gdc_tumor_coding_complete %>% colData(.) %>%  ## get sample_id data from gdc
  as.data.frame(.) %>%
  dplyr::select(project_code, submitted_sample_id, submitted_specimen_id) %>%  ## selecting relevant colums
  mutate(project_code = gsub(pattern = "\\-.*", replacement = "", x = .$project_code)) ## getting rid of the -US at the end of each cancer type

cancer_type <- tumor_samples %>% .$project_code %>% unique %>% .[order(.)] ## get unique cancer_type

cancer_dendro <- list() ## getting cancer type specific tumor sample clustering info

for(n in cancer_type) {
  cancer_info <- tumor_samples %>% rownames_to_column(var = "sample_id") %>% ## move sample_id to column
    filter(project_code == n) %>% ## filter for specific cancer type
    .[, "sample_id"] ## retain only sample_id
  cancer_path <- gdc_reactome_path_analysis %>% .[, names(.) %in% cancer_info] ## only including filtered sample_id
  cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
                          show_rownames = F,
                          show_colnames = F,
                          cellheight = 1,
                          cellwidth = 0.5,
                          cluster_cols = T,
                          cluster_rows = F,
                          annotation_legend = F,
                          color = c("grey", "firebrick3"),
                          clustering_method = "single", ## ordering is single linkage
                          legend = T)
  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
  cancer_path <- names(cancer_path) ## getting the sample_id original order 
  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
  rm(cancer_info, cancer_path)
  gc()
}
dev.off()

for(n in cancer_type[1]) { ## getting the order of clustering by cancer type by alphabetical order
  print(n)
  cancer_order <- as.data.frame(do.call(cbind, cancer_dendro[n])) ## read in first list
  names(cancer_order) <- "sample_id"
  cancer_order$type <- n
  for(p in cancer_type[-1]) {
    print(p)
    df1 <- as.data.frame(do.call(cbind, cancer_dendro[p])) ## read in rest of list in loop
    names(df1) <- "sample_id"
    df1$type <- p
    cancer_order <- rbind(cancer_order, df1) ## combine the two df and repeat the loop until end of cancer_type
    rm(df1)
    gc()
  }
}

cancer_dendro <- cancer_order %>% .[,"sample_id"] %>% as.character(.) ## sample_id ordering

cancer_disrupt_heatmap <- gdc_reactome_path_analysis %>% rownames_to_column(var = "path_name") %>%  ## move rownames to column
  gather(., key = "sample_id", value = "membership", -path_name) ## wide to long format
cancer_disrupt_heatmap$membership <- as.character(cancer_disrupt_heatmap$membership) ## turn boolean values into categories

## set levels based on clustering algorithm above
cancer_disrupt_heatmap$path_name <- factor(cancer_disrupt_heatmap$path_name, levels = c(rev(pathway_dendro)))
cancer_disrupt_heatmap$sample_id <- factor(cancer_disrupt_heatmap$sample_id, levels = c(cancer_dendro))

cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda")

########################
## Figure 2A + B data ##
########################
gdc_path <- gdc_reactome_path_analysis %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>% ## move path_names to column
  .[match(tumor_samples$submitted_sample_id, .$sample), ]

gdc_path_mca <- left_join(gdc_path, tumor_samples[, c("project_code", "submitted_sample_id")],
  by = c("sample" = "submitted_sample_id")) %>%
  column_to_rownames(var = "sample")

gdc_path_sample_names <- rownames(gdc_path_mca) ## Getting sample id order
gdc_path_tumor_types <- gdc_path_mca$project_code ## Getting cancer type order

gdc_path_mca$project_code <- NULL ## Remove project_code column from df

## create a categorical variable from the boolean variable
pathway_list <- make_clean_names(colnames(gdc_path_mca))
pathway_count <- length(pathway_list)
gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
  factor(x, levels = c(0, 1), labels = c("no", "yes")) ## converting boolean values to categorical 'yes' or 'no'
}))
colnames(gdc_path_mca) <- pathway_list
gdc_path_mca$project_code <- gdc_path_tumor_types
rownames(gdc_path_mca) <- gdc_path_sample_names

## perform mca
mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)

expl_var <- mca_res$eig[, 2]
expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100) ## creating a matrix for permutation data

set.seed(4242)
for (k in seq(100)) { ## Perform 100 permutation MCA analysis
  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample)) ## permute data
  MCA_perm <- MCA(X = gdc_perm, graph = FALSE) ## perform MCA analysis 
  expl_var_perm[k, ] <- MCA_perm$eig[, 2] ## Store data in mx. 
}

randomize <- tibble(
  eigenvalue = 1:length(expl_var),
  expl_var = expl_var, ## observed explained variance
  perm = colMeans(expl_var_perm), ## mean permuted explained variance
  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
)

## Determining the optimal mca dimension
optimal_mca_dimensions <- randomize %>%
  filter(pvalue < 0.05) %>% ## filtering for pvalue < 0.05
  filter(eigenvalue == max(eigenvalue)) %>% ## filtering for the maximum eigenvalue
  pull(eigenvalue)

mca_res <- MCA(
  X = gdc_path_mca[, pathway_list],
  ncp = optimal_mca_dimensions, graph = FALSE
)

save(randomize, mca_res, file = "figure1B_data.rda")

####################
## Figure 2C data ##
####################
n_iter <- 40
max_n_pt <- 200
rng <- RNGseq(max_n_pt * n_iter, 42)
r <- rng[[3649]]

rngtools::setRNG(r)
gdc.umap <- umap(mca_res$ind$coord, 
                 n_components = 3, ## reducing mca 36 dimension into 3 dimension
                 n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), ## 87; which is the sqrt of sample size 7,607
                 n_epochs = 1000, ## number of times to find clusters
                 n_threads = 1)
gdc_umap_coors <- as.data.frame(gdc.umap) %>%
  rename(plot_x = V1, plot_y = V2, plot_z = V3) # extract x, y, z coordinates from umap
rownames(gdc_umap_coors) <- rownames(gdc_path_mca)

####################
## Figure 2D data ## Optimization of HDBscan
####################
n_iter <- 40
max_n_pt <- 200
score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)

n_tumors <- nrow(mca_res$ind$coord)

cl <- makeCluster(8)
registerDoParallel(cl)

for (i in 3:max_n_pt) {
  print(paste0("MIN POINTS = ", i))
  score_iter <- foreach(1:n_iter,
    .combine = c,
    .export = c("mca_res"),
    .packages = c("dbscan", "uwot"),
    r = rng[(i - 1) * n_iter + 1:n_iter]
  ) %dopar% {
    rngtools::setRNG(r)
    umap_init_mca <- umap(mca_res$ind$coord, 
                          min_dist = 1e-6, 
                          n_components = 3, 
                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
                          n_epochs = 1000, 
                          n_threads = 1)
    res <- hdbscan(umap_init_mca, minPts = i)
    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
    return(score_iter_temp)
  }
  score_mca[i - 2, ] <- score_iter
}

score <- rowMeans(score_mca)
score_df <- tibble(minPts = 3:max_n_pt, score = score)

sample_df <- score_mca %>%
  as.data.frame() %>%
  as_tibble() %>%
  add_column(minPts = 3:max_n_pt, .before = 1) %>%
  pivot_longer(-minPts) %>%
  group_by(minPts)

optimal_minPts <- score_df %>%
  filter(score == min(score)) %>%
  pull(minPts)

save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")

####################
## Figure 2E data ## HDBscan umap to find clusters
####################
rngtools::setRNG(r)

cluster_umap <- umap(mca_res$ind$coord,
  min_dist = 1e-6,
  n_components = 3,
  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
  n_epochs = 1000,
  n_threads = 1
)
colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
umap_3d_coors <- gdc_umap_coors %>%
  bind_cols(as.data.frame(cluster_umap)) %>%
  mutate(
    n_pathways_mut = rowSums(gdc_path[, -1]),
    clust = as.factor(hdb_cluster_assignment$cluster),
    member_prob = hdb_cluster_assignment$membership_prob,
    sample_id = rownames(gdc_umap_coors)
  )
## clust 0 is unclassified cluster

####################
## Figure 2F data ## kNN clustering of unclassified clusters
####################
## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
umap_3d_coors$clust_knn <- umap_3d_coors$clust
umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
)

###################
## Figure 4 data ## Mutation count
###################
missense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$missense) ## reading in mutation data from SE
splice <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$splice)
frameshdel <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshdel)
frameshins <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$frameshins)
utr3 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr3)
utr5 <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$utr5)
nonstop <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonstop)
nonsense <- as.data.frame(SummarizedExperiment::assays(gdc_tumor_coding_complete)$nonsense)

mutation_count <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense)
gc()
mutation_count <-mutation_count %>% .[rownames(.) %in% rownames(gene2level2_lookup),] %>% ## filter data with genes in our pathway lookup table
  colSums() %>% as.data.frame() %>% ## sum the mutation count
  'colnames<-' (c("mut_count")) %>% ## rename colname
  rownames_to_column(var = "sample_id") ## move rownames to column

umap_3d_coors <- left_join(umap_3d_coors, mutation_count, by = "sample_id") 
rm(mutation_count)

####################
## Figure 3A data ## Cancer type projections on UMAP
####################
umap_3d_coors <- left_join(umap_3d_coors, tumor_samples[,c("project_code", "submitted_sample_id")], by = c("sample_id" = "submitted_sample_id"))

####################
## Figure 3B data ## Subtype projections on UMAP
####################
## getting subtype information on our cancer type
subtypes <- PanCancerAtlas_subtypes() 
tumor_subtypes <- substring(text = tumor_samples$submitted_specimen_id, first = 1, last = 12) ## sample_id for subsetting subtypes
subtypes <- subtypes %>% dplyr::select(pan.samplesID, Subtype_Selected) %>% ## select the relevant columns
  mutate(Subtype_Selected = gsub(pattern = ".*\\.", replacement = "", x = .$Subtype_Selected)) %>% ## remove the cancer acronym at the beginning
  mutate(pan.samplesID = substring(text = .$pan.samplesID, first = 1, last = 12)) %>% ## interested in 
  distinct() %>% ## only unique columns
  .[.$pan.samplesID %in% tumor_subtypes, ] %>% ## include sample_ids that are in our analysis
  filter(Subtype_Selected != "Normal") %>%
  .[!.$pan.samplesID %in% .[which(duplicated(.$pan.samplesID)),]$pan.samplesID, ] ## remove dups

#save(subtypes, file = "subtypes.Rda")
load("../data/subtypes.Rda")

umap_3d_coors <- umap_3d_coors %>%
  mutate(pan.samplesID = substring(text = .$sample_id, first = 1, last = 12)) %>%
  left_join(., subtypes, by = "pan.samplesID") %>% ## left_join subtype data to umap
  dplyr::select(-pan.samplesID)

##########################
## Figure 3D and E data ## Staging and metastasis projection onto UMAP
##########################
## getting tumor_metastasis, staging
tcga_clinical_tumor_staging <- list()
for(n in cancer_type) {
  print(n)
  cancername <- paste0("TCGA-",n)
  query <- GDCquery(project = cancername, ## downloading clinical data for each cancer type using TCGAbiolink
                  data.category = "Clinical",
                  data.type = "Clinical Supplement",
                  data.format = "BCR Biotab",
                  file.type = "patient")
  GDCdownload(query)
  clinical <- GDCprepare(query)[[1]]
  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) {
    cat("Yep, ajcc info in there!\n") ## If all relevant info is there. Do this.
    clinical <- clinical[-c(1:2), c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")]
  }
  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & "ajcc_tumor_pathologic_pt" %in% colnames(clinical)== T) { 
    cat("Nope, ajcc stage not there, but pathologic scores is!\n") ## If ajcc_pathologic_tumor_stage is not there. Do this.
    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")] ## Use clinical stage in place of ajcc pathologic tumor stage
    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm")
  }
  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & "clinical_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) & "pathologic_T" %in% colnames(clinical) == T) {
    cat("Nope, ajcc info not there!\n") ##If ajcc pathologic tumor stage andpathologic pt is not there. Do this
    clinical <- clinical[-c(1:2), c("bcr_patient_barcode", "clinical_stage", "pathologic_T", "pathologic_M")]
    names(clinical) <- c("bcr_patient_barcode","ajcc_pathologic_tumor_stage", "ajcc_tumor_pathologic_pt", "ajcc_metastasis_pathologic_pm") ## Use clinical stage in place of ajcc pathologic tumor stage, and pathologic T in place of ajcc tumor pathologic pt.
  }
  ## If ajcc pathologic tumor stage and clinical stage is not present, skip to next
  if(!"ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"clinical_stage" %in% colnames(clinical) == T) next
  ## If ajcc_pathologic tumor stage is present and not ajcc_tumor_pathologic pt, skip to next
  if("ajcc_pathologic_tumor_stage" %in% colnames(clinical) & !"ajcc_tumor_pathologic_pt" %in% colnames(clinical) == T) next
  tcga_clinical_tumor_staging[[n]] <- clinical ## store data in list
}

tcga_clinical_tumor_staging <- do.call(rbind, tcga_clinical_tumor_staging) ## rbind the list together
tcga_clinical_tumor_staging <- tcga_clinical_tumor_staging[tcga_clinical_tumor_staging$bcr_patient_barcode %in% tumor_subtypes,] ## We have 6633 out of 7607 tumor data
missing_sample <- as.data.frame(matrix(data = "NA", nrow = length(tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]), ncol = length(tcga_clinical_tumor_staging) -1 )) ## creating df of NA's for missing samples

tcga_clinical_tumor_staging <- missing_sample %>% 'colnames<-' (names(tcga_clinical_tumor_staging[-1])) %>% ## colnames based on tcga_clinical_tumor_staging
  mutate(bcr_patient_barcode = tumor_subtypes[!tumor_subtypes %in% tcga_clinical_tumor_staging$bcr_patient_barcode]) %>%
  .[,c(4,1,2,3)] %>% ## reorder columns for rbinding
  rbind(tcga_clinical_tumor_staging, .) ## rbind

save(tcga_clinical_tumor_staging, file = "tcga_clinical_tumor_staging.Rda")

umap_3d_coors <- umap_3d_coors %>%
  mutate(bcr_patient_barcode = substring(text = .$sample_id, first = 1, last = 12)) %>%
  left_join(., tcga_clinical_tumor_staging, by = "bcr_patient_barcode") %>% ## left_join clinical data to umap
  dplyr::select(-bcr_patient_barcode)

## Convert metastasis inputs into set categorical values
M0 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("cM0 (i+)", "M0")
M1 <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("M1", "M1a", "M1b", "M1c")
NA.value <- umap_3d_coors$ajcc_metastasis_pathologic_pm %in% c("[Discrepancy]", "[Not Applicable]", "[Not Available]", "MX")

umap_3d_coors$ajcc_metastasis_pathologic_pm[M0] <- "M0"
umap_3d_coors$ajcc_metastasis_pathologic_pm[M1] <- "M1"
umap_3d_coors$ajcc_metastasis_pathologic_pm[NA.value] <- "NA"

## Convert stage inputs into set categorical values
stage1 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage I", "Stage IA", "Stage IA1", "Stage IA2", "Stage IB", "Stage IB1", "Stage IB2", "Stage IC")
stage2 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage II", "Stage IIA", "Stage IIA1", "Stage IIA2", "Stage IIB", "Stage IIC")
stage3 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC", "Stage IIIC1", "Stage IIIC2")
stage4 <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC")
unknown <- umap_3d_coors$ajcc_pathologic_tumor_stage %in% c("Stage X", "[Discrepancy]", "[Not Applicable]", "I/II NOS", "NA", "[Not Available]")

umap_3d_coors$ajcc_pathologic_tumor_stage[stage1] <- "Stage I"
umap_3d_coors$ajcc_pathologic_tumor_stage[stage2] <- "Stage II"
umap_3d_coors$ajcc_pathologic_tumor_stage[stage3] <- "Stage III"
umap_3d_coors$ajcc_pathologic_tumor_stage[stage4] <- "Stage IV"
umap_3d_coors$ajcc_pathologic_tumor_stage[unknown] <- "Not Available"

write.table(x = umap_3d_coors, file = "umap_3d_coors.tsv", quote = FALSE, sep = "\t", row.names = FALSE) ## save the data
```

```{r fig-1a-1l, warning=FALSE, fig.height=12, fig.width=11, message = FALSE}
#fig.height=15, fig.width=16,
load("../data/cancer_disrupt_heatmap.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv")

kelly.colours <- c("gray95", "gray13", "gold2", "plum4", "darkorange1",
                   "lightskyblue2", "firebrick", "burlywood3", "gray51", "springgreen4",
                   "lightpink2", "deepskyblue4", "lightsalmon2", "mediumpurple4",
                   "orange", "maroon", "yellow3", "brown4", "yellow4", "sienna4", "chocolate", "gray19")

fig1A <- ggplot(data = cancer_disrupt_heatmap, aes(y=sample_id, x=path_name, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  facet_grid(type ~ ., scales = "free", space = "free", switch = "y") +
scale_y_discrete(position = "left") +
  ylab("") +
  xlab("377 Reactome Pathways") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        strip.text.y = element_text(size = 12, hjust = 0),
        strip.text.y.left = element_text(angle = 0),
        strip.background = element_blank(),
        legend.position = "none")

## fixing the text cutoff in heatmap
 fig1Ag <- ggplotGrob(fig1A)

 for(i in which(grepl("strip-l", fig1Ag$layout$name))){
   fig1Ag$grobs[[i]]$layout$clip <- "off"
 }

 fig1A <- as.ggplot(fig1Ag)
 rm(fig1Ag)

##figure 1b
load("../data/figure1B_data.rda")
optimal_mca_dimensions <- randomize %>%
  filter(pvalue < 0.05) %>%
  filter(eigenvalue == max(eigenvalue)) %>%
  pull(eigenvalue)

permut_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
  geom_line(aes(colour = "#998ec3")) +
  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
  scale_color_identity(guide = guide_legend(title = ""), 
                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
  scale_y_continuous(labels = scales::percent) +
  xlab("MCA Eigenvalues") +
  ylab("explained variance") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.position = c(0.8, 0.8))

pvalue_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = pvalue)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = optimal_mca_dimensions, color = "red") +
  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_mca_dimensions)) +
  xlab("MCA Eigenvalues") +
  ylab("P-value") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

fig1B <- permut_p /
  pvalue_p

fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
                aes(y=plot_x, x=plot_y)) +
  geom_point(size = 0.25) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed() +
  theme_classic() +
  scale_color_viridis_c() +
  labs(tag = "C") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

load("../data/optimal_minpts_score.rda")
fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
  geom_point(data = sample_df, color = "grey", aes(y = value), size = 0.25) +
  geom_point(size = 0.25) + 
  geom_line() +
  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
  scale_y_log10() +
  xlab("Minimum Cluster Size") +
  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

## umap hdb_scan
fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(y = plot_x, x = plot_y, color = as.factor(clust))) +
  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
  guides(color = F) +
  theme_classic() +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed() +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

## umap knn_clust 
fig1F <- ggplot(subset(umap_cancertype), aes(y = plot_x, x = plot_y, color = as.factor(clust_knn))) +
  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
  guides(color = F) +
  theme_classic() +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed() +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

## mutation burden plot
mut_order <- list() ## store data in list 
for(n in 1:max(as.numeric(as.character(umap_cancertype$clust_knn)))) { ## loop through all the clusters
  message(n)
  mut_order[[n]] <- umap_cancertype %>% filter(clust_knn == n) %>%  ## filter for specific class
    dplyr::select(sample_id, plot_x, plot_y, clust_knn, mut_count) %>% ## select relevant columns
    arrange(mut_count) ## arrange mutation in increasing order
}
mut_order <- mut_order %>% do.call(rbind, .) %>% as.data.frame() ## rbind list

mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id)
mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10))

## Need the breaks positions for the x-axis
previous_class <- 0 ## starting position 1
position_of_previous_class <- 0 ## starting position 2
xaxis_breaks <- list() ## list for for loop
for(n in 1) { ## loop through 1 to 12 clusters
  message(n)
  class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of class 
  xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
  previous_class <- class_spec ## storing class halfway mark
  for(n in 2:max(as.numeric(as.character(mut_order$clust_knn)))) {
    message(n)
    position_of_previous_class <- xaxis_breaks[[n-1]] ## getting the previous class position
    class_spec <- round(mut_order %>% filter(clust_knn == n) %>% nrow(.) / 2, digits = 0) ## calculate halfway mark of new class
    xaxis_breaks[[n]] <- class_spec + previous_class + position_of_previous_class ## adding up all variables
    previous_class <- class_spec ## storing class halfway mark
  }
}
xaxis_breaks <- as.data.frame(do.call(rbind, xaxis_breaks)) ## rbind list
xaxis_breaks <- mut_order[rownames(mut_order) %in% xaxis_breaks$V1,] ## find the tumor samples in these positions

total_n <- as.data.frame(table(mut_order$clust_knn)) ## getting total n in cluster

mut_order$log10mut <- log10(mut_order$mut_count + 1) ## mutation count log transformation + pseudo count

mut_median <- mut_order %>% 
  group_by(clust_knn) %>% 
  dplyr::summarize(mut_med = median(log10mut)) ## calculate median for each class

## Need to add 50 pseudo samples front and end of data so they don't touch the y-axis
mut_order <- mut_order[,c("sample_id", "clust_knn", "log10mut")] ## selecting relevant column

begin_NA <- data.frame(sample_id = paste0("sample", 1:100), 
                       clust_knn = min(as.numeric(mut_order$clust_knn)), ## put lowest cluster to prevent a new facet
                       log10mut = NA) ## put NA values so they don't get plotted
end_NA <- data.frame(sample_id = paste0("sample", 101:200), ## repeat the step for the ends
                       clust_knn = max(as.numeric(mut_order$clust_knn)),
                       log10mut = NA)
mut_order <- rbind(begin_NA, mut_order, end_NA) ## rbind the data in a specfic order

mut_order$sample_id <- factor(mut_order$sample_id, levels = mut_order$sample_id) ## factor samples
mut_order$clust_knn <- factor(mut_order$clust_knn, levels = c(1:10)) ## factor clusters

fig1G <- ggplot(mut_order, aes(x = sample_id, y = log10mut, color = clust_knn)) + 
  geom_point(size = 0.25) + 
  ylab("Log10(Mutation Count)") + 
  xlab("") +
  theme_classic() +
  facet_wrap(~ factor(clust_knn, levels = c(1:10)), ncol = 10, scales = "free_x") +
  geom_hline(data= mut_median, aes(yintercept=mut_med)) +
  scale_y_continuous(sec.axis = sec_axis(~.)) + ## dup y-axis
  scale_x_discrete(breaks = c(as.character(mut_order$sample_id[1]), as.character(xaxis_breaks$sample_id)), labels = c("n =" ,total_n$Freq)) +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks.x = element_blank(),
        axis.line.x = element_line(colour = "black", size = 0.5),
        axis.line.x.top = element_line(colour = "black", size = 0.5),
        panel.spacing = unit(1, "lines"),
        strip.background = element_rect(size = 0.5),
        strip.placement = "outside",
        strip.text.x = element_text(size = 10),
        legend.position = "none") +
  scale_color_manual(values = kelly.colours[-c(1:2)])

## code to keep only the bottom border line of the facet
fig1G <- ggplotGrob(fig1G)
lg <- linesGrob(x=unit(c(0,1),"npc"), y=unit(c(0,0),"npc"), 
                   gp=gpar(col="black", lwd=4))

for (k in grep("strip-t",fig1G$layout$name)) {
  fig1G$grobs[[k]]$grobs[[1]]$children[[1]] <- lg
}

fig1G <- as.ggplot(fig1G)

## umap by cancer type
fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(y = plot_x, x = plot_y, colour = project_code)) +
  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(color = "") +
  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap of brca subtype
fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(y = plot_x, x = plot_y, colour = Subtype_Selected)) +
  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(color = "") +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap brca somatic mutations
load("../data/gdc_gene_exp_adj.Rda")
load("../data/gene_data.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv")
brca <- gene_data %>% .[.$external_gene_name %in% c("BRCA1", "BRCA2"),] %>% dplyr::select(ensembl_gene_id, external_gene_name) ## Getting BRCA gene information
brca <- gdc_gene_exp_adj %>% .[rownames(.) %in% brca$ensembl_gene_id, ] %>% ## we only want brca gene
  rownames_to_column(var = "ensembl_gene_id") %>%
  left_join(., brca, by = "ensembl_gene_id") %>% 
  remove_rownames()  %>% column_to_rownames(var = "external_gene_name") %>% ## replace ensembl_id with gene_name
  dplyr::select(-ensembl_gene_id) %>% ## get ride of ensembl_id  col
  t() %>% as.data.frame() %>% ## correct orientation for umap_cancertype df
  rownames_to_column(var = "sample_id") %>% ## getting data ready for leftjoin 
  mutate(BRCA1 = case_when(BRCA1 == "1" ~ "BRCA1",
                           BRCA1 == "0" ~ "NA"),
         BRCA2 = case_when(BRCA2 == "1" ~ "BRCA2",
                           BRCA2 == "0" ~ "NA")) %>%
  mutate(mut_type = gsub('NA,|,NA', '', paste(.$BRCA1, .$BRCA2, sep = ","))) %>% ## combine mutation cause some patients can have both
  dplyr::select(-BRCA1, -BRCA2) %>% ## remove col we don't need
  left_join(., umap_cancertype[,c("sample_id", "plot_x", "plot_y", "project_code")], by = "sample_id") ## leftjoin by sample id

fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(y = plot_x, x = plot_y, colour = mut_type)) +
  geom_point(data = brca, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(color = "") +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap stage
fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(y = plot_x, x = plot_y, colour = ajcc_pathologic_tumor_stage)) +
  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(color = "") +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap metastasis
fig1L <- ggplot(umap_cancertype, aes(y = plot_x, x = plot_y)) +
  geom_point(data = umap_cancertype, aes(y = plot_x, x = plot_y), colour = "grey", size = 0.25) +
  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(y = plot_x, x = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(color = "") +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_color_manual(labels = c("M0", "M1"), 
                     values = c('#a6cee3','#1f78b4')) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

empty_plot <- ggplot() + theme_void()

layout <- '
AAAABBBCCCDDD
AAAABBBCCCDDD
AAAABBBEEEFFF
AAAABBBEEEFFF
AAAAGGGGGGGGG
AAAAGGGGGGGGG
AAAAGGGGGGGGG
AAAAHHHIIIJJJ
AAAAHHHIIIJJJ
AAAAKKKLLL###
AAAAKKKLLL###
'

fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout) + plot_annotation(tag_levels = 'A')

```

### 2020-09-12 code/figure4.Rmd

@@
 
-```{r forest-heatmap-data-plot, fig.align='center', fig.width= 9, fig.height= 8, message=FALSE}
-#fig.width= 16, fig.height= 20,
+```{r forest-heatmap-data-plot, fig.align='center', fig.width= 8, fig.height= 10.25, message=FALSE}
 load("../data/forest.heatmap.data.rda")
@@
   geom_vline(xintercept = 0, linetype = 2) +
-  geom_pointrange(aes(xmin = ci.low, xmax = ci.up)) +
+  geom_pointrange(aes(xmin = ci.low, xmax = ci.up, fatten = 1)) +
   scale_y_discrete(labels = Hugo_rownames) +
@@
     panel.grid.major.y = element_line(colour = "#f0f0f0"),
-    axis.title.y = element_blank()
+    axis.text.y = element_text(size = 8),
+    axis.title.y = element_blank(),
+    legend.text = element_text(size = 8),
+    legend.title = element_text(size = 8)
   )
@@
     strip.text = element_blank(),
-    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
+    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
     axis.text.y = element_blank(),
     axis.title.y = element_blank(),
+    legend.text = element_text(size = 8),
+    legend.title = element_text(size = 8),
     panel.grid.major.y = element_line(colour = "#f0f0f0")

### 2020-09-12 code/supplemental_v2.Rmd

@@
 
-```{r supplementary-figure-s1, fig.width = 9}
+\pagenumbering{gobble}
+
+```{r supplementary-figure-s1, fig.width = 9, fig.height = 10}
 umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
+load("../data/kelly.colours.rda")
+cluster.colours <- c(kelly.colours[c(3:12)]) ## Color scheme
+names(cluster.colours) <- c(1:10) ## Assign cluster to color
 
@@
 
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in cancer_type[1:6]) {
+par(mfrow = c(6,4), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(0,0,0,0))
+for(n in cancer_type[1:23]) {
 with(umap_cancertype %>% 
@@
      scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F, axes = FALSE,
                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
@@
                                 colvar = NULL))
-title(xlab="UMAP 2", line=0)
-legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
+legend("bottom", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2, adj = 0.25)
 }
+```
 
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
-par(mar = c(2,2,0,0))
-for(n in cancer_type[7:12]) {
-with(umap_cancertype %>%
-       filter(project_code == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-}
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in cancer_type[13:15]) {
-with(umap_cancertype %>%
-       filter(project_code == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-}
-
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in cancer_type[19:23]) {
-with(umap_cancertype %>%
-       filter(project_code == n),
-     scatter3D(plot_y, -plot_z, plot_x,
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-}
-
-```
 **Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
 individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Figure 1B of the main text**.
+in **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
 
@@
 
-```{r supplementary-figure-s2-data-generation, fig.width = 9}
-ditto_color <- c(
-  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
-  "#0072B2", "#D55E00", "#CC79A7", "#666666",
-  "#AD7700", "#1C91D4", "#007756", "#D5C711",
-  "#005685", "#A04700", "#B14380", "#4D4D4D",
-  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
-  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
-  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
-  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
-  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
-  "#00446B", "#803800", "#8D3666", "#3D3D3D")
+```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height = 10}
+load("../data/ditto.colours.rda")
 
+## Determining which cancer type has subtype information
 cancer_w_subtypes <- list()
@@
 
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
+par(mfrow = c(6,4), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(0,0,0,0))
 
-for(n in cancer_w_subtypes[1:6]) {
+for(n in cancer_w_subtypes[1:19]) {
   spec_cancer <- umap_cancertype %>% 
@@
   subtypes <- unique(spec_cancer$J_color)
-  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
+  J_colors <- c(ditto_colours[c(2:(length(subtypes) + 1))])#, 'grey')
   names(J_colors) <- subtypes
@@
                bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
-               theta = 0, phi = 65, scale = F,
+               theta = 0, phi = 65, scale = F, axes = FALSE,
                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
@@
                colvar = NULL))
-  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  title(xlab="UMAP 2", line=0)
+  legend("bottom", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2, adj = 0.25)
+  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1)
+  legend("top", inset = c(0.08), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1)
+  legend("top", inset = c(0.16), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1)
 }
+```
 
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-
-for(n in cancer_w_subtypes[7:12]) {
-  spec_cancer <- umap_cancertype %>% 
-    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
-                               TRUE ~ "Other")) %>%
-    filter(J_color != "Other")
-  if(nrow(spec_cancer) == 0) {next
-    }
-  subtypes <- unique(spec_cancer$J_color)
-  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
-  names(J_colors) <- subtypes
-
-  with(spec_cancer,
-     scatter3D(plot_y, -plot_z, plot_x, 
-               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
-               theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-               colvar = NULL))
-  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  title(xlab="UMAP 2", line=0)
-}
-
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-
-for(n in cancer_w_subtypes[13:15]) {
-  spec_cancer <- umap_cancertype %>% 
-    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
-                               TRUE ~ "Other")) %>%
-    filter(J_color != "Other")
-  if(nrow(spec_cancer) == 0) {next
-    }
-  subtypes <- unique(spec_cancer$J_color)
-  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
-  names(J_colors) <- subtypes
-
-  with(spec_cancer,
-     scatter3D(plot_y, -plot_z, plot_x, 
-               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
-               theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-               colvar = NULL))
-  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  title(xlab="UMAP 2", line=0)
-}
-
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-
-for(n in cancer_w_subtypes[16:19]) {
-  spec_cancer <- umap_cancertype %>% 
-    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
-                               TRUE ~ "Other")) %>%
-    filter(J_color != "Other")
-  if(nrow(spec_cancer) == 0) {next
-    }
-  subtypes <- unique(spec_cancer$J_color)
-  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
-  names(J_colors) <- subtypes
-
-  with(spec_cancer,
-     scatter3D(plot_y, -plot_z, plot_x, 
-               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
-               theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-               colvar = NULL))
-  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
-  title(xlab="UMAP 2", line=0)
-}
-
-# ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
-#   geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-#   geom_point(size = 0.25) +
-#   theme_classic() +
-#   coord_fixed() +
-#   xlab("UMAP 1") + ylab("UMAP 2") +
-#   theme(axis.text.x = element_blank(),
-#     axis.text.y = element_blank(),
-#     axis.title.x = element_blank(),
-#     axis.title.y = element_blank(),
-#     axis.ticks = element_blank(),
-#     axis.line = element_blank(),
-#     legend.position = "none") +
-#   facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
-```
 **Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**.
+cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's histological subtype identity.
 
@@
 
-# ##oncogene
-# oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
-# ##tumor_supressor
-# tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL", "APOBEC3B") ## list of selected tumor suppressor
-# 
-# selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data
-
 selected_mut <- gdc_mut_count_patient %>% filter(rownames(.) %in% selected_mut$ensembl_gene_id) %>% ## filter by ensembl ids
@@
 
-selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
+selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z, clust_knn) %>% ## select relevant columns
   left_join(., selected_mut, by = "sample_id") %>%
-  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z", "clust_knn")) %>%
   filter(value != 0)
@@
 
-```{r supplementary-figure-s3, fig.width = 9}
-# out.width = '77%'
+```{r supplementary-figure-s3, fig.width = 9, fig.height = 10}
 load("../data/selected_mut.Rda")
@@
 
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in genes[1:6]) {
+par(mfrow = c(6,4), mai = c(0.1, 0.1, 0.1, 0.1))
+par(mar = c(0,0,0,0))
+for(n in genes[1:15]) {
 with(selected_mut %>% 
@@
      scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-}
-
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
-par(mar = c(2,2,0,0))
-for(n in genes[7:12]) {
-with(selected_mut %>% 
-       filter(variable == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
-}
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in genes[13:15]) {
-with(selected_mut %>% 
-       filter(variable == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
+                                theta = 0, phi = 65, scale = F, axes = FALSE,
                                 xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
@@
 ```
+
 **Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
-genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**.
+genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
 
@@
 
-```{r supplementary-figure-s4-data-generation, eval = FALSE}
-load("../data/gdc_reactome_path_analysis.Rda") ## load data
-umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
-n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway
-
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway
-
-selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "sample_id") ## prepare pathway for left_join
-
-selected_path <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
-  left_join(., selected_path, by = "sample_id") %>%
-  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
-  filter(value != 0)
-
-save(selected_path, file = "../data/selected_path.Rda")
-
-# load("../data/selected_path.Rda") ## load data
-# load("../data/gdc_reactome_path_analysis.Rda")
-# umap_cancertype <- fread("../data/umap_3d_coors.tsv")
-# umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
-# 
-# n_path <- 377 ## number of pathways we want to see
-# 
-# toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway..
-needed for facet
-# 
-# ## creating layout to plot all 377 pathways
-# figs4_layout <- data.frame(start = seq(from = 1, to = length(toppath), by = 25),
-#                           end = c(seq(from = 25, to = length(toppath), by = 25), length(toppath)))
-# 
-# for(n in 1:nrow(figs4_layout)) {
-#   message(n)
-#   pathway_layout <- figs4_layout %>% .[n,] ## getting the layout for selecting pathways
-#   pathway_layout <- toppath[pathway_layout$start:pathway_layout$end] ## filtering those pathways from toppath
-#   filtered_path <- selected_path %>% filter(variable %in% pathway_layout) %>% ## filtering our data to only include those pathways
-#     mutate(variable = factor(variable, levels = c(pathway_layout)))
-#   ggplot_umap <- ggplot(filtered_path, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
-#     geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
-#     geom_point(size = 0.25) +
-#     theme_classic() +
-#     coord_fixed() +
-#     xlab("UMAP 1") + ylab("UMAP 2") +
-#     scale_colour_manual(values = rep("firebrick3", 25)) +
-#     theme(axis.text.x = element_blank(),
-#       axis.text.y = element_blank(),
-#       axis.title.x = element_blank(),
-#       axis.title.y = element_blank(),
-#       axis.ticks = element_blank(),
-#       axis.line = element_blank(),
-#       strip.text = element_text(size = 5),
-#       legend.position = "none") +
-#     facet_wrap(~ variable,
-#                ncol = 5) ## facet by pathways
-#   new_dfname <- paste0("ggplot_umap_", n)
-#   assign(paste0(new_dfname), value = ggplot_umap)
-# }
-# 
-# save(ggplot_umap_1,ggplot_umap_2,ggplot_umap_3,ggplot_umap_4,ggplot_umap_5,ggplot_umap_6,ggplot_umap_7,ggplot_umap_8,ggplot_umap_9,ggplot_umap_10,ggplot_umap_11,ggplot_umap_12,ggplot_umap_13,ggplot_umap_14,ggplot_umap_15,ggplot_umap_16, file = "../data/fig_s4_ggplot.Rda")
-```
-
-```{r figure-s4-plot, fig.width= 9}
-load("../data/selected_path.Rda")
-load("../data/effect_heatmap.Rda") ## load effect heatmap data
-pathorder <- as.character(levels(effect_heatmap$path_name)) ## getting col order
-
-startingpoint <- seq(1, 377, by = 15)
-startingpointlist <- list()
-for(n in startingpoint) {
-  start1 <- n
-  end1 <- start1 + 5
-  start2 <- end1 + 1
-  end2 <- start2 + 5
-  start3 <- end2 + 1
-  end3 <- start3 + 2
-  startingpointlist[[n]] <- data.frame(start1 = start1,
-                                       end1 = end1,
-                                       start2 = start2,
-                                       end2 = end2,
-                                       start3 = start3,
-                                       end3 = end3)
-}
-startingpointlist <- as.data.frame(do.call(rbind, startingpointlist))
-
-s4plot <- function(x) {
-  a <-  x[,"start1"]
-  b <- x[,"end1"]
-  c <- x[,"start2"]
-  d <- x[,"end2"]
-  e <- x[,"start3"]
-  f <- x[,"end3"]
-  path01<- pathorder[a:b]
-  path02<- pathorder[c:d]
-  path03<- pathorder[e:f]
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in path01) { #pathorder[x[,"start1"]:x[,"end1"]]
-  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
-  pathname <- pathname[,1] ## get the character vectors
-  with(selected_path %>% 
-       filter(variable == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
-path1 <- gsub("NA", "", path1)
-path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
-path2 <- gsub("NA", "", path2)
-legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
-legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
-}
-
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in pathorder[x[,"start2"]:x[,"end2"]]) {
-  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
-  pathname <- pathname[,1] ## get the character vectors
-  with(selected_path %>% 
-       filter(variable == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
-path1 <- gsub("NA", "", path1)
-path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
-path2 <- gsub("NA", "", path2)
-legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
-legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
-}
-
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0))
-for(n in pathorder[x[,"start3"]:x[,"end3"]]) {
-  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
-  pathname <- pathname[,1] ## get the character vectors
-  with(selected_path %>% 
-       filter(variable == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=0)
-path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
-path1 <- gsub("NA", "", path1)
-path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
-path2 <- gsub("NA", "", path2)
-legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
-legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
-}
-}
-
-data <- startingpointlist[1,]
-s4plot(x = data)
-data <- startingpointlist[2,]
-s4plot(x = data)
-data <- startingpointlist[3,]
-s4plot(x = data)
-data <- startingpointlist[4,]
-s4plot(x = data)
-data <- startingpointlist[5,]
-s4plot(x = data)
-data <- startingpointlist[6,]
-s4plot(x = data)
-data <- startingpointlist[7,]
-s4plot(x = data)
-data <- startingpointlist[8,]
-s4plot(x = data)
-data <- startingpointlist[9,]
-s4plot(x = data)
-data <- startingpointlist[10,]
-s4plot(x = data)
-data <- startingpointlist[11,]
-s4plot(x = data)
-data <- startingpointlist[12,]
-s4plot(x = data)
-data <- startingpointlist[13,]
-s4plot(x = data)
-data <- startingpointlist[14,]
-s4plot(x = data)
-data <- startingpointlist[15,]
-s4plot(x = data)
-data <- startingpointlist[16,]
-s4plot(x = data)
-data <- startingpointlist[17,]
-s4plot(x = data)
-data <- startingpointlist[18,]
-s4plot(x = data)
-data <- startingpointlist[19,]
-s4plot(x = data)
-data <- startingpointlist[20,]
-s4plot(x = data)
-data <- startingpointlist[21,]
-s4plot(x = data)
-data <- startingpointlist[22,]
-s4plot(x = data)
-data <- startingpointlist[23,]
-s4plot(x = data)
-data <- startingpointlist[24,]
-s4plot(x = data)
-data <- startingpointlist[25,]
-s4plot(x = data)
-```
-
-```{r, fig.width = 6.7}
-par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
-par(mar = c(2,2,0,0), cex.lab = 0.8)
-for(n in pathorder[376:377]) { 
-  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
-  pathname <- pathname[,1] ## get the character vectors
-  with(selected_path %>% 
-       filter(variable == n),
-     scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
-                                theta = 0, phi = 65, scale = F,
-               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
-                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
-                                colvar = NULL))
-title(xlab="UMAP 2", line=-1.7)
-path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
-path1 <- gsub("NA", "", path1)
-path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
-path2 <- gsub("NA", "", path2)
-legend("top", inset = c(0.11), legend=path1, bty = "n", horiz = T, cex = 0.8)
-legend("top", inset = c(0.16), legend=path2, bty = "n", horiz = T, cex = 0.8)
-}
-```
-
-**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
-pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
-(column) from **Figure 1B of the main text**.
-
-\newpage
-
-```{r, supplemental-s5-data-generation, eval = FALSE}
+```{r, supplemental-s4-data-generation, eval = FALSE}
 load("../data/gdc_tumor_coding_complete.Rdata")
@@
 
-s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
+s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z, clust_knn) %>% ## select relevant columns
   left_join(., selected_mut, by = "sample_id") %>%
-  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
+  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z", "clust_knn")) %>%
   filter(value != 0)
@@
 
-```{r supplemental-s5-plot, fig.width=9}
+```{r supplemental-s4-plot, fig.width=9}
 load("../data/s5_data.Rda")
@@
      scatter3D(plot_y, -plot_z, plot_x, 
-                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
+                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
                                 theta = 0, phi = 65, scale = F,
@@
 ```
-**Supplementary Figure S5.
+
+**Supplementary Figure S4.
 Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 1B of the main text**.
+The colors indicate the tumor's cluster identity.
 
 \newpage
+
 ```{r supplemental-table-st1-data-generation, eval = FALSE}

### 2020-09-12 text/discussion.Rmd

@@
 protein-coding genes is comparable with the number of tumor samples, even with
-multiple mutations per sample. We sought to
-simplify the problem by employing a "knowledge-base driven analysis"
-[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
-biochemical pathways, which we accomplished by translating gene-level mutations
-into pathway level disruptions. Our approach differs from previously described
-methods [@PATHNET] in that we chose to focus on the pathway (defined in the
-methods) as the unit of disruption instead of the gene, where individual
-mutations may be sufficient to alter pathway activity.
+multiple mutations per sample. We sought to simplify the problem by employing a
+"knowledge-base driven analysis" [@pathways-Atul-Butte], investigating cancer as
+a disease of basic cellular and biochemical pathways. We accomplished this by
+translating gene-level mutations into pathway level disruptions. Our approach
+differs from previously described methods [@PATHNET] in that we chose to focus
+on the pathway (defined in the methods) as the unit of disruption instead of the
+gene, where individual mutations may be sufficient to alter pathway activity.
 
-To our knowledge, this approach has not previously been attempted despite
-its relative simplicity. We limited our analysis to mutations with likely
-deleterious effects, including missense, nonsense, frame-shift, stop-loss,
-untranslated region and splice site mutations in genes that are actively
-expressed in each cancer type, thus avoiding bias from transcription coupled
-repair. Our method of filtration differs from the "rank-and-cut" method of
-[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
-biases. We also restricted our analysis to biochemical
-pathways, excluding curated gene sets related to diseases, syndromes, or classes
-of proteins with shared catalytic activity or conserved domains which are
-potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
-redundancy and exclude biologically unrelated collections of genes.
+To our knowledge, this approach has not previously been attempted despite its
+relative simplicity. We limited our analysis to mutations with likely
+deleterious effects in genes that are actively expressed in each cancer type,
+thus avoiding bias from transcription coupled repair. Our method of filtration
+differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
+reasonable attempt to account for the same biases. We also restricted our
+analysis to biochemical pathways, excluding curated gene sets related to
+diseases, syndromes, or classes of proteins with shared catalytic activity or
+conserved domains which are potentially problematic [@pathways-Atul-Butte]. We
+chose this approach to limit redundancy and exclude biologically unrelated
+collections of genes.
 
-The hypothesis that cancer is the result of dysfunction from a limited number of
-basic cellular processes common to eukaryotic cells was introduced and later
-expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
+The hypothesis that cancer results from dysfunction in basic cellular processes
+common to eukaryotic cells was introduced and later expanded on in a pair of
+essays by Hanahan and Weinberg [@weinberg2000;
 @hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of
@@
 We attempted to account for our clusters in terms of more trivial explanations.
-For example, it could be that the clusters we identified are consistent with
-disease progression. We were unable to identify any such trend in the number of
+For example, it could be that the clusters are consistent with disease
+progression. We were unable to identify any such trend in the number of
 mutations, the relative staging or metastasis; and each cluster instead was
-associated with unique combinations of pathways.
+associated with unique combinations of pathways. Some cancer types are unevenly
+distributed among the clusters, though we could not identify any cancers that
+were exclusive to a single cluster. Only cancer types with the fewest samples
+were found to be absent from one or more clusters at all.
 
-Some cancer types are unevenly distributed among the clusters, though we could
-not identify any cancers that were exclusive to a single cluster. Only cancer
-types with the fewest samples were found to be absent from one or more clusters
-at all. Somewhat surprisingly to us, this finding extends to histological
-subtypes of breast, head and neck cancers, leukemias, _etc._ This result implies
-that histological subtypes could reflect differences in cell-of-origin, rather
-than fundamental differences in cancer phenotype. The four major subtypes of
-breast cancer correspond to histological and molecular expression profiles that
-define them and how they respond to experimental stimuli [@breast-subtypes2]. It
-has been hypothesized that differences in the molecular regulators of
-development in the precursor cell types present in the breast epithelium drive
-the various histological phenotypes [@breast-subtypes; @breast-subtypes3].
-Consistent with this view we found that breast tumor samples of the Luminal A
-subtype were heavily biased toward membership in clusters 1 and 7, and basal
-tumor samples were biased toward cluster 6. However, both subtypes also
-contained samples in every other cluster (without exception), and Luminal B and
-Her2 positive samples are distributed across clusters. We interpret these data
-to mean that inherited cell-of-origin signatures could predispose certain
-precursor cells within the breast epithelium to forming tumors of one cluster or
-another but are not determinative. This view is compatible with the previously
-stated hypothesis but opens the way for a more granular view of individual
-tumors. It would be surprising if we did not observe bias for some cancers and
+Somewhat surprisingly to us, this finding extends to histological subtypes of
+breast, head and neck cancers, leukemias, _etc._ This result implies that
+histological subtypes could reflect differences in cell-of-origin, rather than
+fundamental differences in cancer phenotype. The four major subtypes of breast
+cancer correspond to histological and molecular expression profiles that define
+them and how they respond to experimental stimuli [@breast-subtypes2]. It has
+been hypothesized that differences in molecular regulators of development in
+precursor cell types present in breast epithelium drive histological phenotypes
+[@breast-subtypes; @breast-subtypes3]. Consistent with this view we found that
+breast tumor samples of the Luminal A subtype were heavily biased toward
+membership in clusters 1 and 7, and basal tumor samples were biased toward
+cluster 6. However, both subtypes also contained samples in every other cluster
+(without exception), and Luminal B and Her2 positive samples are distributed
+across clusters. Our interpretation of these data is that inherited
+cell-of-origin signatures could predispose certain precursor cells within the
+breast epithelium to forming tumors of one cluster or another but are not
+determinative. This view is compatible with the previously stated hypothesis but
+opens the way for a more granular view of individual tumors.
+
+It would be surprising if we did not observe bias for some cancers and
 subtypes amongst our classes, since some treatment regimens have greater
@@
 genes with tissue-specific pathway disruptions result in common categories when
-viewed at the level of the pathway knowledge-base. Top cancer driver genes (_e.g.
-PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the fact that
-they contribute to many cluster-specific pathways. This can only be explained as
-a result of other less common driver genes complementing the pathway disruption
-of the unique combination of driver genes that are disrupted in each tumor, and
-we speculate that many of these are sensitive to mutation in different tissues.
+viewed at the level of the pathway knowledge-base. Top cancer driver genes
+(_e.g. PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the
+fact that they contribute to many cluster-specific pathways. This can only be
+explained by less common genes complementing the unique combination of driver
+genes in each tumor, and we speculate that many of these less common genes are
+sensitive to increased mutation frequency in different tissues.
 
@@
 
-Finally, we must remark on the limitations of our work exemplified in two of the
-clusters, 1 and 10, for which we did not find many distinctive associations with
+Finally, we must remark on the limitations of our work exemplified in 
+clusters 1 and 10, for which we did not find many distinctive associations with
 pathways. Cluster 1 had a relatively low proportion of mutated pathways,
 although it is broadly enriched in many of the same tumor-promoting pathways
-common to the other groups. In addition, our data show clearly that this cluster
+common to other clusters. In addition, our data show clearly that this cluster
 is as likely to contain stage IV metastatic tumors as it is to contain those of
@@
 By modeling CDC longevity data as a baseline risk function we showed that each
-of our pathway disruption clusters exhibit cancer-type specific effects on
-expected survival. However, considering that within each cancer type there are
-different clinical standards of care, and even within classes of drugs the
-preferred treatment can very between cancers, it makes sense that we observe
-tissue-specific cluster effects. Contrast the situation with ovarian _vs._
-breast cancer, which are both hormonally driven cancers, for example. Ovarian
-cancer has but one main treatment axis, platinum, whereas breast cancer patients
-have a variety of treatment regimens based on molecular subtype and other
-factors. Unfortunately, given the diversity of drug classes and treatments, we
-lack sufficient power to explore these variables in the TCGA data. It is our
-hope that future studies will help to distinguish between treatment-specific
-effects on survival given different pathway disruption clusters.
+of cluster exhibits cancer-type specific effects on survival expectancy.
+However, considering that within each cancer type there are different clinical
+standards of care, and even within classes of drugs the preferred treatment can
+very between cancers, it makes sense that we observe tissue-specific cluster
+effects. Contrast ovarian _vs._ breast cancer, which are both hormonally driven
+cancers, for example. Ovarian cancer has but one main treatment axis, platinum,
+whereas breast cancer patients have a variety of treatment regimens based on
+molecular subtype and other factors. Unfortunately, given the diversity of drug
+classes and treatments, we lack sufficient power to explore these variables in
+the TCGA data. It is our hope that future studies will help to distinguish
+between treatment-specific effects on survival given different pathway
+disruption clusters.
 
@@
 
-Our findings imply that there are separate processes in the etiology of cancer
+Our findings imply separate processes in the etiology of cancer
 that can be broadly thought of as general cancer promoting, cluster-specific

### 2020-09-12 text/manuscript.Rmd

@@
 To study cancer pathways we obtained a set of 7,607 solid tumors from The Cancer
-Genome Atlas (TCGA) through the [Genomic Data Commons (GDC)](gdc.cancer.gov)
+Genome Atlas (TCGA) through the Genomic Data Commons (GDC)(gdc.cancer.gov)
 portal. TCGA data are most appropriate given the relative completeness of the
@@
 splicing mutations. To minimize bias from well-studied diseases and processes,
-we selected 377 [Reactome pathways](https://reactome.org/) (see supplemental
+we selected 377 Reactome pathways (reactome.org) (see supplemental
 table ST1) of interest corresponding to basic cellular processes and biochemical
@@
 [@JSSv025i01], and visually summarized the analysis with UMAP (**figure 1B** and
-see interactive media from [**supplemental file
-1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
+see interactive media from **supplemental file
+1** (junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
 used the resulting graph coordinates to perform density based clustering
@@
 subtypes (**figure 2B**). We also projected histological subtype data for the
-remaining cancers (see **supplementary figure S2** and online [**supplemental
-file1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that
+remaining cancers (see **supplementary figure S2** and see interactive media from 
+**supplemental file 1**); we find that
 the subtypes, though often biased to one or more clusters, are almost never
@@
 segregation of these mutations into specific clusters (**figure 2C** and see
-interactive media from [**supplemental file
-1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We also projected
+interactive media from **supplemental file 1**). We also projected
 mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_;
@@
 disruptions for each of the 377 pathways onto the UMAP (**supplementary table
-ST2** and see interactive media from [**supplemental file
-1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
+ST2** and see interactive media from **supplemental file
+1**). Clusters 3 and 5
 were distinguished by metabolic pathways including RNA and protein biosynthesis
@@
 genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary
-figure S4** and see interactive media from [**supplemental file
-1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
+figure S4** and see interactive media from **supplemental file 1**. Interestingly,
 cluster 9, which also shared multiple enrichment in signaling pathways with
@@
 this fully reproducible manuscript in R markdown format. R markdown files and
-all other models are available from [our
-repositories](https://github.com/dennishazelett/TTmanu) on the distributed
+all other models are available from our
+repositories (https://github.com/dennishazelett/TTmanu) on the distributed
 version control site, GitHub. Extended descriptions of clustering, models,
 formulae, pathway selection, and gene filtering are provided in supplementary
-methods.
+methods. An interactive media of the results is available on junkdnalab.shinyapps.io/PANCAN_supplemental/.
 

### 2020-09-12 text/supplemental_methods.Rmd
Fix fig1 heatmap clip off, fix fig2 metastasis color annotation, add url to shinyapp in manuscript and supplemental_methods

@@
 number of tumor samples (87). Using this method, we assigned cluster membership
-to the remaining tumors (**figure SM1F**).
+to the remaining tumors (**figure SM1F** and see interactive media from **supplemental file
+1** (junkdnalab.shinyapps.io/PANCAN_supplemental/)).
 
@@
 
-```{r, results='asis'}
-cat("\\newpage")
-```
-
 #### Figure SM2
@@
 tissue types (**supplementary figure SM3**).
+
+```{r, results='asis'}
+cat("\\newpage")
+```
 

### 2020-09-12 text/survival.Rmd

@@
 may have different prognoses within cancer types or across all cancers. These
-analyses are limited by confounding factors of age and stage at diagnosis, sex,
+analyses are limited by confounding factors of age, stage at diagnosis, sex,
 ethnicity, and tissue-specific disease progression. To explore these ideas, we
 used Bayesian inference to test models of survival using public longevity data
-from the CDC. Our model estimates the effect of cancer type and cluster-specific
+from the CDC. The estimates the effect of cancer type and cluster-specific
 cancer effects independently, resulting in a cancer and cluster-specific
@@
 
-To illustrate the model, we simulate survival time curves
-for four related groups (**figure 5A**). We start with a cohort of women aged 30 and show their
-expected survival (light blue). By setting their cancer rate multiplier to $k=2$
-we can simulate the effect of a moderately deadly cancer (yellow). Compare this
-to a randomly selected group of women ranging in age from 30 to 70 years old
-(dark blue). Immediately the survival curve changes due to the mixture of ages,
-without any malignancy. Adding a $k=2$ malignancy further reduces expected
-lifespan (red). We can see that for different tissues, simply due to the change
-in distribution of age-at-diagnosis, we should expect equally deadly cancers to
-have *different* survival curves. The model takes this into account and the
-aggressiveness of the cancer can be estimated accurately without the confounding
-effect of age.
+To illustrate the model, we simulate survival four related groups (**figure
+5A**). We start with a cohort of women aged 30 and show expected survival (light
+blue). By setting cancer rate multiplier to $k=2$ we simulate the effect of a
+moderately deadly cancer (yellow). Compare this to randomly selected women
+ranging from 30 to 70 years old (dark blue). Immediately, survival changes due
+to the mixture of ages, without malignancy. Adding a $k=2$ malignancy further
+reduces expected lifespan (red). Hence, due to the change in distribution of
+age-at-diagnosis, we expect equally deadly cancers to have *different* survival
+curves. The model takes this into account and cancer aggressiveness is estimated
+accurately without the confounding effect of age.
 
@@
 
-**Figure 5: Comparison of different cancers, simulated and actual results.** To
-illustrate the risk model, we compare survival of four simulated cohorts. A) A
-group consisting of randomly selected 30 year old women (light blue) vs. an
-equal sized group of 30 year old women with cancer, $k = 2$ (light orange).
-Compare a randomly selected group of people with ages uniformly distributed from
-30 to 70 (dark blue) and a randomly selected group with uniformly distributed
-ages and cancer $k = 2$ (dark orange). B,C,D,E) Compare random group with
-correct age distribution for the given cancer (turqoise) to the model predicted
-survival for the given average k (blue) and the actual survival for the patients
-in our dataset (orange). Differences with actual data are likely due to a
-mixture of cluster identities which we exclude from these simulations for
-simplicity. F) Effects are compared between cancer type by showing the
-probability density for the posterior value of the $k_{tis}$ multiplier. Larger
-values correspond to decreased age-independent life expectancy.
+**Figure 5: Comparison of different cancers, simulated and actual results.**
+Comparison of survival in four simulated cohorts. A) A group consisting of
+randomly selected 30 year old women, unaffected (light blue) _vs._ cancer, 
+($k = 2$, light orange), another group consisting of random 30-70 year-old
+women, unaffecteds (dark blue) and cancer $k = 2$ (dark orange). B,C,D,E)
+Compare random group with correct age distribution for the given cancer
+(turqoise) to the model predicted survival for the given average k (blue) and
+the actual survival for the patients in our dataset (orange). Differences with
+actual data are likely due to a mixture of cluster identities which we exclude
+from these simulations for simplicity. F) Effects are compared between cancer
+type by showing the probability density for the posterior value of the $k_{tis}$
+multiplier. Larger values correspond to decreased age-independent life
+expectancy.
 
@@
 cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
-where diagnosis apparently decreases risk relative to background (Thyroid:
-THCA). Cancers with posterior probability for relative risk of less than 1 such
-as THCA should be interpreted carefully. This Bayesian model is a model for a
-state of information. The information that a person is diagnosed with cancer may
-lead us to expect that they will live a shorter time than the general population
-of matched age (k > 1), or a longer time than the general population of matched
-age (k < 1). A shorter prognosis could result from the cancer being aggressive,
-injuring the body and causing death. By contrast, a longer prognosis may result
-from the cancer being relatively mild, and therefore diagnosis could be an
-informational signal that the patient is health conscious, with the comparison
-group having more people whose cancers go undiagnosed. It is important to note
-therefore that the diagnosis can increase our expectation of life relative to
-the comparison group, even if it decreases the expectation of life of the
-individual relative to the counterfactual where they did not have cancer.
+where diagnosis decreases risk relative to background (Thyroid: THCA). Cancers
+with posterior probability for relative risk of less than 1 such as THCA should
+be interpreted carefully. This Bayesian model is for a state of information. The
+information that a person is diagnosed with cancer may lead us to expect shorter
+survival than the general population of matched age (k > 1), or longer survival
+than the general population of matched age (k < 1). Shorter prognosis could
+result from cancer aggressiveness, injuring the body and causing death. By
+contrast, longer prognosis could result from the cancer being relatively mild,
+and therefore diagnosis could be an informational signal that the patient is
+health conscious, with the comparison group having more people whose cancers go
+undiagnosed. It is important to note therefore that the diagnosis can increase
+our expectation of life relative to the comparison group, even if it decreases
+the expectation of life of the individual relative to the counterfactual where
+they did not have cancer.
 
-The degree to which a given cancer accelerates aging can be determined by the
-$k_{tis}$ multiplier (**figure 5F**) multiplied by the cluster modifier $k_{cl}$
-(**figure 6**). Looking at the tissue specific multiplier, the least deadly was
-prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
-risk multiplier of between 15 and 22 relative to the background risk of death in
-the population. This undoubtedly is influenced by the fact that relatively young
+The degree to which cancer accelerates aging can be determined by the $k_{tis}$
+multiplier (**figure 5F**) multiplied by the cluster modifier $k_{cl}$ (**figure
+6**). Looking at the tissue specific multiplier, the least deadly was prostate
+cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a risk
+multiplier of between 15 and 22 relative to the background risk of death in the
+population. This undoubtedly is influenced by the fact that relatively young
 patients are affected by GBM and that it is extremely deadly. Among the
@@
 **Figure 6 Cluster specific $k_{cl}$ values for each cancer type.** The overall
-$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
-represents the relative aggressiveness of each cluster within the cancer type.
-Alpha transparency (α) is set to reflect varying confidence in the posterior
+$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific $k$
+represents relative aggressiveness of each cluster within cancer type. Alpha
+transparency (α) is set to reflect varying confidence in the posterior
 distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
@@
 different standards of care for example. In support of this interpretation, we
-also tested a factored model which considered both cancer type and each cluster
-independent of cancer type (not shown). Though we were able to successfully fit
-this model, it is a special case of the more general model where cluster and
-tissue independently affect longevity, and there is no reason to believe that
-cluster specific effects would necessarily remain constant across tissue types
-given how widely the patients vary across tissue types, drug and surgical
-treatments.
+also tested a factored model which considered both cancer type cluster
+independently. Though we were able to successfully fit this model, it is a
+special case of the more general model where cluster and tissue independently
+affect longevity, and there is no reason to believe that cluster specific
+effects would necessarily remain constant across tissue types given how widely
+the patients vary across tissue types, drug and surgical treatments.

### 2020-09-13 code/figure1.Rmd
Reduce panel spacing in fig1A

@@
         plot.margin = unit(c(0.6, 0, 0, 0), "cm"),
-        panel.background = element_rect(fill = "transparent"))
+        panel.background = element_rect(fill = "transparent"),
+        panel.spacing = unit(0.15, "lines"))
 

### 2020-09-13 code/figure4.Rmd
Reduce y-axis text size for fig4

@@
     panel.grid.major.y = element_line(colour = "#f0f0f0"),
-    axis.text.y = element_text(size = 8),
+    axis.text.y = element_text(size = 6.5),
     axis.title.y = element_blank(),

### 2020-09-14 code/supplemental_file_shiny/app.R
Add code for survival simulation for ktissue only - class specific simulation not setup yet

@@ -15,7 +15,8 @@ suppressPackageStartupMessages(c(
     library(htmltools),
     library(car),
     library(grid),
-    library(gridBase)
+    library(gridBase),
+    library(ggkm)
 ))
 
 ## Load data
@@ -72,7 +73,26 @@ ui <- (fluidPage(
                              pickerInput("cancer_type_pathway", label = "Cancer type:", ## Cancer type
                                          choices = cancer_type, selected = cancer_type, multiple = TRUE,
                                          options = list(`actions-box` = TRUE)),
-                             plotOutput("pathway")))
+                             plotOutput("pathway")),
+                    tabPanel(title = "Survival",
+                             pickerInput("sim.cancertype", label = "Cancer type:", ## Cancer type
+                                         choices = tisnames, selected = "BRCA", #multiple = TRUE,
+                                         options = list(`actions-box` = TRUE)),
+                             pickerInput("sim.cluster", label = "Cluster:", ## Cluster
+                                         choices = 1:10, selected = "1", #multiple = TRUE,
+                                         options = list(`actions-box` = TRUE)),
+                             pickerInput("sim.sex", label = "Sex:", ## Sex
+                                         choices = c("MALE", "FEMALE"), selected = "FEMALE"),
+                             pickerInput("sim.ethnic", label = "Ethnicity:", ## Ethnicity
+                                         choices = c("WHITE", "BLACK OR AFRICAN AMERICAN"), selected = "WHITE"),
+                             sliderInput("sim.range", "Age range:", ## Age range
+                                         min = 1, max = 100,
+                                         value = c(30,70)),
+                             sliderInput("sim.population", label = "Population Size:", ## Population size
+                                         min = 5, max = 500,
+                                         value = 100),
+                             plotOutput("survival"))
+                    )
         )
     )
 )
@@ -243,6 +263,194 @@ server <- (function(input, output, session) {
                        xlab = "UMAP 2", ylab = "UMAP 3", zlab = "UMAP 1",
                        colvar = NULL))
     })
+    
+    ## Survival plot code
+    output$survival <- renderPlot({
+        
+        ## Survival plot code
[…]

### 2020-09-14 code/supplemental_file_shiny/app.R
Add checkbox for control, update ggplot theme for survival

@@ -91,6 +91,8 @@ ui <- (fluidPage(
                              sliderInput("sim.population", label = "Population Size:", ## Population size
                                          min = 5, max = 500,
                                          value = 100),
+                             checkboxInput("show.sim.k.eq.1", label = "Show Control",
+                                           value = FALSE),
                              plotOutput("survival"))
                     )
         )
@@ -393,13 +395,43 @@ server <- (function(input, output, session) {
                                   names_to = "condition"
         )
         
+        if(input$show.sim.k.eq.1 == FALSE) {
+            df1 <- df[!df$condition %in% "rtime",]
+        } else {
+            df1 <- df
+        }
         # df <- df %>% mutate(status = case_when(condition != "acttime" ~ as.integer(status),
         #                                        condition == "acttime" ~ as.integer(actstatus)))
         
-        ggplot(df, aes(time = time, color = condition, status = status)) +
+        load("ditto.colours.rda")
+        
+        rtime.colour <- ditto_colours[2]
+        stime.colour <- ditto_colours[3]
+        
+        km.colours <- c(rtime.colour, stime.colour)
+        names(km.colours) <- c("rtime", "stime")
+        
+        ggplot(df1, aes(time = time, color = condition, status = status)) +
             geom_km() +
             theme_classic() +
-            theme(legend.position = c(.8, .75))
+            theme(legend.position = c(.8, .75)) +
+            coord_cartesian(xlim = c(0,80)) +
+            scale_color_manual(values = km.colours) +
+            labs(x = "Survival Time (years past diagnosis)",
+                 y = "Probabilty",
+                 colour = "",
+                 title = sprintf("Survival for %s simulated %s %s patients", sim.cancertype.selection, tolower(sim.ethnic.selection), tolower(sim.sex.selection)),
+                 subtitle = sprintf("Simulated Patient Population; k = %.1f", kmean)) +
+            theme(axis.text = element_text(size = 12),
+                  axis.title = element_text(size = 14))
+        
+        
+  #      labs(
+            #                        x = "Survival Time (years past diagnosis)",
+            #                        y = "Probability",
+            #                        title = sprintf("Survival for %s simulated white female patients", sim.cancertype.selection),
+            #                        subtitle = sprintf("Simulated Patient Population; k = %.1f", kmean)
+            #                    )
         
         # plotkmcomp = function(sim,ref,act,actstatus){
         #     data.frame(

### 2020-09-14 code/supplemental_file_shiny/app.R
Add cluster kmean to survival simulation

@@ -79,7 +79,7 @@ ui <- (fluidPage(
                                          choices = tisnames, selected = "BRCA", #multiple = TRUE,
                                          options = list(`actions-box` = TRUE)),
                              pickerInput("sim.cluster", label = "Cluster:", ## Cluster
-                                         choices = 1:10, selected = "1", #multiple = TRUE,
+                                         choices = c(as.character(1:10), "All"), selected = "All", #multiple = TRUE,
                                          options = list(`actions-box` = TRUE)),
                              pickerInput("sim.sex", label = "Sex:", ## Sex
                                          choices = c("MALE", "FEMALE"), selected = "FEMALE"),
@@ -323,7 +323,7 @@ server <- (function(input, output, session) {
         ## Reactive options
         sim.cancertype.selection <- as.character(input$sim.cancertype)
         #sim.cancertype.selection <- c("BRCA")
-        #sim.cluster.selection <- input$sim.cancertype
+        sim.cluster.selection <- input$sim.cluster
         sim.sex.selection <- input$sim.sex
         #sim.sex.selection <- "FEMALE"
         sim.ethnic.selection <- input$sim.ethnic
@@ -367,8 +367,14 @@ server <- (function(input, output, session) {
         load("dfclas.rmd")
         load("~/TTmanu/code/supplemental_file_shiny/dfclas.rmd")
         ## Getting the kclass mean for specified cluster
+        if(sim.cluster.selection == "All") {
         kcl <- dfclas[which(dfclas$tissnames %in% sim.cancertype.selection),]
         kmean.cl <- mean(kcl$value)
+        } else{
+            kcl <- dfclas[which(dfclas$tissnames %in% sim.cancertype.selection &
+                                dfclas$class %in% sim.cluster.selection),]
+            kmean.cl <- mean(kcl$value)
+        }
         
         ## Getting kmean value ##
         kmean <- kmean.tis * kmean.cl

### 2020-09-21 text/introduction.Rmd
add hotnet2 reference

@@
 address the sparseness of mutations, allowing genes to be influenced by
-mutations in nearest network neighbors [@hornNetSigNetworkbasedDiscovery2018;
+mutations in nearest network neighbors [@hotnet2;@hornNetSigNetworkbasedDiscovery2018;
 @Iranzo-Network]. These methods are powerful especially when applied to smaller

### 2020-10-14 text/introduction.Rmd
minor edits to introduction

@@
-Recent advances in technology have greatly expanded our view into the mechanisms
+Advances in technology have greatly expanded our view into the mechanisms
 of cancer at the systems biology level. Next-generation sequencing technologies
-have made it possible to assay germline and somatic mutations, expression
+have made it possible to study germline and somatic mutations, expression
 profiles, DNA methylation, and copy number variations from the same tissue. To
@@
 numbers of tumors. One of the challenges in the field remains that there may be
-thousands of moderate effect genes that occur at such low frequency--given
-tissue diversity and low sample number--that they are impossible to detect using
-positive selection theory. Some groups have attempted to address this challenge
-by applying machine learning to cancer data to discover groups of
-functionally related genes as they interact with larger pathways and networks
-[@kimPancancerAnalysisSomatic2018;
+thousands of moderate effect genes that occur at such low frequency that they
+are impossible to detect using positive selection theory. Some researchers have
+attempted to address this challenge by applying machine learning to cancer data
+to discover groups of functionally related genes as they interact with larger
+pathways and networks [@kimPancancerAnalysisSomatic2018;
 @mourikis-ml-pathways; @moonlight]. These studies have proven very effective at

### 2021-05-03 code/supplemental_v2.Rmd
Edit manuscript for iscb iscience submission

@@
 ---
+title: "Supplementary figures"
 output:
-  bookdown::pdf_book:
-    number_sections: FALSE
-    toc: FALSE
-    keep_tex: yes
+  word_document:
+    reference_docx: "../text/word-style-reference-ttmanu-iscb-iscience-v2.docx"
 always_allow_html: TRUE
@@
 
-**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
+**S1 Fig. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
 individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
+in **Fig 1B of the main text**. The colors indicate the tumor's cluster identity.
 
@@
 
-**Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's histological subtype identity.
+**S2 Fig. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
+cancers projected onto the tumors of the UMAP from **Fig 1B of the main text**. The colors indicate the tumor's histological subtype identity.
 
@@
 
-**Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
-genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
+**S3 Fig. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
+genes involved in genome stability projected on the UMAP from **Fig 1B of the main text**. The colors indicate the tumor's cluster identity.
 
@@
 
-**Supplementary Figure S4.
-Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 1B of the main text**.
+**S4 Fig.
+Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Fig 1B of the main text**.
 The colors indicate the tumor's cluster identity.

### 2021-05-03 text/discussion.Rmd
Edit manuscript for iscb iscience submission

@@
 
-### Incompleteness of the pathway disruption data
-
-Finally, we must remark on the limitations of our work exemplified in 
-clusters 1 and 10, for which we did not find many distinctive associations with
-pathways. Cluster 1 had a relatively low proportion of mutated pathways,
-although it is broadly enriched in many of the same tumor-promoting pathways
-common to other clusters. In addition, our data show clearly that this cluster
-is as likely to contain stage IV metastatic tumors as it is to contain those of
-stage I. This cluster likely represents a group of tumors with aberrations in
-methylation, copy number, or other structural variants. Consistent with this,
-kidney chromophobe and thyroid cancers have high proportions of structural
-variations _vs._ other variant types [@PCAWG-MAIN] and are heavily skewed to
-cluster 1 membership. Likewise, cluster 10 represents a group of hyper-mutated
-tumors that harbor so many mutations that virtually no pathway is unaffected. It
-seems likely that a significant fraction of the “mutant” samples for each
-pathway are burdened with excess passenger mutations. This could be addressed
-with more sophisticated filtering of likely passenger mutations [_e.g._
-@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
-incorporate these other data into a comprehensive pathway-centric analysis as we
-have done here for point mutations and indels.
-
 ### Estimates of survival reveal pathway-dependent differences

### 2021-05-03 text/introduction.Rmd
Edit manuscript for iscb iscience submission

@@
 genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) sequenced
-large numbers of tumors multiple assays to analyze together, with the goal of
+large numbers of tumors and collected data from multiple assays to analyze together, with the goal of
 integration and increased understanding of the mechanisms of cancer.
@@
 has long been a goal of the field as characterized in essays by
-@weinberg2000; @hallmarks_II. The first of these studies in genomics identified
+[@weinberg2000; @hallmarks_II]. The first of these studies in genomics identified
 11 subtypes from 12 cancer types, using integrative analysis with co-equal
@@
 More recently, pan-cancer mutations and other data have been looked at as
-disruptions of normal pathway activity. @PCAWG-PATH carried out exhaustive
+disruptions of normal pathway activity. Sanchez-Vega et al. [@PCAWG-PATH] carried out exhaustive
 analysis of pathway enrichment using whole genome data. Network approaches

### 2021-05-03 text/manuscript.Rmd
Edit manuscript for iscb iscience submission

@@
 title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: Peter T. Nguyen$^{1,\dag}$, Simon G. Coetzee$^{1,\dag}$, Daniel L. Lakeland$^{2}$, and
+author: Peter T. Nguyen$^{1,¶}$, Simon G. Coetzee$^{1,¶}$, Daniel L. Lakeland$^{2}$, and
   Dennis J. Hazelett$^{1,3,*}$
 bibliography: references.bib
+csl: plos-computational-biology.csl
 output:
   word_document:
-    reference_docx: "../text/word-style-reference-ttmanu.docx"
+    reference_docx: "../text/word-style-reference-ttmanu-iscb-iscience-v2.docx"
 always_allow_html: TRUE
@@
 ---
+###### Institutional affiliations: $^{1}$The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, CA, 90048, USA.
+$^{2}$Lakeland Applied Sciences LLC, Los Angeles, CA, 91001, USA.
+$^{3}$Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA, 90048, USA.
+$^{¶}$These authors contributed equally to this work.
+$^{*}$Correspondence: Dennis.Hazelett@cshs.org
 
-###### Institutional affiliations: 1.
-The Center for Bioinformatics and Functional Genomics, Cedars-Sinai Medical Center, Los Angeles, California.
-2.
-Lakeland Applied Sciences LLC, Los Angeles, CA.
-3.
-Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA.
-$\dag$ These authors contributed equally.
-$*$ To whom correspondence should be addressed.
+```{r, results='asis', echo = FALSE}
+#cat("\\newpage")
+```
 
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, dpi=300, warning = FALSE)
 suppressPackageStartupMessages(c(
@@
 
-## Abstract
+## Summary
 
@@
 splicing mutations. To minimize bias from well-studied diseases and processes,
-we selected 377 Reactome pathways (reactome.org) (see supplemental
-table ST1) of interest corresponding to basic cellular processes and biochemical
+we selected 377 Reactome pathways (reactome.org) (see S1 Table) of interest 
+corresponding to basic cellular processes and biochemical
 pathways, excluding those corresponding to catalytic categories (_e.g._
@@
 assigning a Boolean value of 1 to each pathway with one or more genes mutated
-and 0 for all others (**figure 1A**).
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-#### Figure 1
-```{r child='../code/figure1.Rmd'}
-```
-
-**Figure 1: Clustering pathways of tumor samples.** A) Each of 377 selected
-Reactome pathways (rows) is classified as disrupted if one or more genes is
-mutated in the tumor sample (columns). Red cells denote pathway disruption.
-Tumors types reflect standardized abbreviations from the TCGA project. B)
-Different rotational perspectives of the same MCA-based UMAP projection in
-three-dimensional space. Each dot corresponds to a tumor sample. The same colors
-indicate the tumor's cluster identity throughout this manuscript.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
+and 0 for all others (**Fig 1A**).
 
 We investigated this dataset using multiple correspondence analysis (MCA)
-[@JSSv025i01], and visually summarized the analysis with UMAP (**figure 1B** and
-see interactive media from **supplemental file
-1** (junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
+[@JSSv025i01], and visually summarized the analysis with UMAP (**Fig 1B** and
+see interactive media from **S1 File** (junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
 used the resulting graph coordinates to perform density based clustering
@@
 
-#### Independence from tissue-of-origin.
+### Independence from tissue-of-origin.
 
@@
 cluster, and all well-represented cancer types had tumors in every cluster (see
-**figure 2A** and full tumor profiles in **supplementary figure S1**,
-interactive media **supplementary file 1**), suggesting that these clusters
+**Fig 2A** and full tumor profiles in **S1 Fig**,
+interactive media **S1 File**), suggesting that these clusters
 indicate molecular pathology largely independent of tissue-of-origin. As an
 example of one cancer that does have a biased pathway profile, pancreatic
-adenocarcinoma (PAAD) was predominantly found in cluster 8 (**figure 2A** and
-**supplementary figure S1** and **supplementary file 1**), but even PAAD
+adenocarcinoma (PAAD) was predominantly found in cluster 8 (**Fig 2A** and
+**S1 Fig** and **S1 File**), but even PAAD
 comprises tumors from the nine remaining classes. Thus, patients with these
@@
 
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 2
-```{r child = '../code/figure2.Rmd'}
-```
-
-**Figure 2: Pathway-based clustering independent of tissue-of-origin** 
-A) Sankey plot of correspondence between cancer type and cluster identity.
-B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
-somatic mutation onto the UMAP. D) Projection of tumor stage onto the UMAP,
-regardless of cancer type. E) Projection of metastatic status onto the UMAP.
-Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors. F) Somatic mutation 
-frequencies for each cluster. Vertical axis shows log mutation count, horizontal 
-axis is cluster identity. Each dot represents an individual tumor sample, ranked 
-lowest to highest by mutation count. The median mutation count in each cluster is 
-indicated by the horizontal line.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Independence from molecular and histological subtype.
+### Independence from molecular and histological subtype.
 
@@
 standards of care. We did not observe exclusive segregation by cluster for these
-subtypes (**figure 2B**). We also projected histological subtype data for the
-remaining cancers (see **supplementary figure S2** and see interactive media from 
-**supplemental file 1**); we find that
+subtypes (**Fig 2B**). We also projected histological subtype data for the
+remaining cancers (see **S2 Fig** and see interactive media from 
+**S1 File**); we find that
 the subtypes, though often biased to one or more clusters, are almost never
@@
 
-#### Independence from drivers of genome instability.
+### Independence from drivers of genome instability.
 
@@
 mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe
-segregation of these mutations into specific clusters (**figure 2C** and see
-interactive media from **supplemental file 1**). We also projected
+segregation of these mutations into specific clusters (**Fig 2C** and see
+interactive media from **S1 File**). We also projected
 mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_;
@@
 exhibited remarkable specificity with respect to cluster assignment
-(**supplementary figure S3**). To look at other risk factors such as maintenance
+(**S3 Fig**). To look at other risk factors such as maintenance
 of DNA methylation levels and telomere length, we projected somatic mutations of
 the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and observed
-similar lack of cluster bias (**supplementary figure S3**).
+similar lack of cluster bias (**S3 Fig**).
 
-#### Independence of stage, mutation count and mutation profile.
+### Independence of stage, mutation count and mutation profile.
 
@@
 mutations and culminates in a cluster (class 10) which has nearly every pathway
-disrupted (**figure 2F**). However, we don't observe a trend in the overall
+disrupted (**Fig 2F**). However, we don't observe a trend in the overall
 mutation burden across the backbone of the UMAP. Nonetheless, to test the
@@
 origin and other categories of tumor, we do not observe any bias among the
-stages to specific clusters (**figure 2D**), suggesting that stage does not
+stages to specific clusters (**Fig 2D**), suggesting that stage does not
 contribute to cluster identity.
@@
 metastases onto our UMAP, and surprisingly we observed even distribution of the
-samples across clusters (**figure 2E**). This final observation suggests that
+samples across clusters (**Fig 2E**). This final observation suggests that
 our pathway-disruption clustering is dependent on particular combinations of
@@
 In order to identify pathway enrichment across all cancers, we created a list of
-pathway disruptions with percent mutated samples and top genes (**supplementary
-table ST1**). As expected, these analyses reveal the broad importance of many
+pathway disruptions with percent mutated samples and top genes (**S1 Table**). 
+As expected, these analyses reveal the broad importance of many
 well known pathways that are disrupted in cancer, including "PIP3 activates Akt
@@
 pathways from highest to lowest enrichment. We visualized enrichment as a
-heatmap (**figure 3A**). Using this approach, we identified about fourteen
+heatmap (**Fig 3A**). Using this approach, we identified about fourteen
 pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see methods)
-(**supplementary table ST2**). Clusters 7, 8, and 9 had several pathways in
+(**S2 Table**). Clusters 7, 8, and 9 had several pathways in
 common. To explore the specific pathways marking each cluster, we projected
@@
 were distinguished by metabolic pathways including RNA and protein biosynthesis
-(**supplementary table ST2**). Similarly, cluster 4 was distinguished by
+(**S2 Table**). Similarly, cluster 4 was distinguished by
 mutations affecting regulation of DNA and histone methylation ("DNA
@@
 Clusters 7-9 have in common mutations in extracellular, intracellular, and
-immune-related signaling pathways (see **figure 3B** and **supplementary table
-ST2**). Cluster 2 had the highest pathway enrichment levels of the three, having
+immune-related signaling pathways (see **Fig 3B** and **S2 Table**). 
+Cluster 2 had the highest pathway enrichment levels of the three, having
 mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
 hypoxia", "regulation of cell cycle" and "apoptosis" among others.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 3
-```{r child = '../code/figure3.Rmd'}
-```
-
-**Figure 3: Pan-cancer enrichment of pathway disruptions.** A) Heatmap shows
-relative enrichment of each pathway (rows) within numbered cluster (columns).
-Color represents effect size as percent enrichment. B) Proportion of mutated
-samples in the each significant pathway (columns; union set of pathways with
-effect size ≥ 0.30 in each cluster) within cluster (rows).
-
-```{r, results='asis'}
-cat("\\newpage")
-```
 
@@
 differentially mutated genes. We ranked odds ratios and selected the top ten
-enriched and depleted genes (pvalue < 0.01) for each cluster (**figure 4**; odds
+enriched and depleted genes (pvalue < 0.01) for each cluster (**Fig 4**; odds
 ratios plot). Clusters 7 and 8, which shared multiple enrichment in signaling
 pathways, are largely driven by mutations in _PI3K_ and its orthologs and _Ras_
-genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary
-figure S4** and see interactive media from **supplemental file 1**. Interestingly,
+genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **S4 Fig** and see 
+interactive media from **S1 File**. Interestingly,
 cluster 9, which also shared multiple enrichment in signaling pathways with
@@
 involved in protein degradation. We also observed that genes enriched for one
-cluster are depleted from others (_i.e._ is enriched in cluster 6, but depleted
+cluster are depleted from others (_i.e._ _TP53_ is enriched in cluster 6, but depleted
 in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in clusters 3 and
 8). Next, we investigated the proportion of samples per cancer type for the
-significant genes within a cluster (**figure 4**; heatmap). Surprisingly,
+significant genes within a cluster (**Fig 4**; heatmap). Surprisingly,
 clusters were not predominated by one or more highly mutated genes across all
@@
 mutations associated with specific clusters.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 4
-```{r child = '../code/figure4.Rmd'}
-```
-
-**Figure 4: Gene level analysis reveals tissue-specific class signatures.** 
-Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
-95% confidence interval. Rows: significant genes from each cluster. Each cluster
-was compared against the background (all other clusters) to find differentially
-mutated genes. Significant genes (pvalue < 0.01) were selected and limited to
-the top ten results for each cluster. Heatmap; Columns: cancer type. Rows:
-significant genes. The heatmap shows proportion of samples per cluster and
-cancer type mutated for each gene. Depleted significant genes (left). Enriched
-significant genes (right).
-
-```{r, results='asis'}
-cat("\\newpage")
-```
 
@@
 found a total of 31 enriched pathways (significant with enrichment score ≥ 0.3)
-across all clusters (**table 1**). A number of enrichments represented pathways
+across all clusters (**Table 1**). A number of enrichments represented pathways
 that were already shown to be enriched in non-metastatic samples of other
 clusters. For example, “Signaling by _PTK6_” is enriched in non-metastatic
-samples of cluster 8 (**supplementary table ST2**) but not in 7 and 9. This
+samples of cluster 8 (**S2 Table**) but not in 7 and 9. This
 pathway is enriched in metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$,
-**table 1**). This is also true of “Erythropoietin activates _RAS_”, which is
-enriched in non-metastatic tumors of cluster 8 (**supplementary table ST2**) and
+**Table 1**). This is also true of “Erythropoietin activates _RAS_”, which is
+enriched in non-metastatic tumors of cluster 8 (**S2 Table**) and
 also in metastatic tumors of clusters 7 and 9. Cluster 4 metastases were
@@
 which is also specific to clusters 2, 7 and 8 non-metastatic tumors. Thus,
-metastatis pathways from one cluster are often enriched in non-metastases of
+metastases pathways from one cluster are often enriched in non-metastases of
 other clusters.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Table 1
-```{r child = '../code/table1.Rmd'}
-```
-
-**Table 1: Cluster-specific enriched pathways (effect size ≥ 0.30) in metastasis.**
 
@@
 
-## Materials & Methods
+## Limitations of the Study
 
-All code and data for producing the analyses and figures herein are included in
-this fully reproducible manuscript in R markdown format. R markdown files and
-all other models are available from our
+We must remark on the limitations of our work exemplified in 
+clusters 1 and 10, for which we did not find many distinctive associations with
+pathways. Cluster 1 had a relatively low proportion of mutated pathways,
+although it is broadly enriched in many of the same tumor-promoting pathways
+common to other clusters. In addition, our data show clearly that this cluster
+is as likely to contain stage IV metastatic tumors as it is to contain those of
+stage I. This cluster likely represents a group of tumors with aberrations in
+methylation, copy number, or other structural variants. Consistent with this,
+kidney chromophobe and thyroid cancers have high proportions of structural
+variations _vs._ other variant types [@PCAWG-MAIN] and are heavily skewed to
+cluster 1 membership. Likewise, cluster 10 represents a group of hyper-mutated
+tumors that harbor so many mutations that virtually no pathway is unaffected. It
+seems likely that a significant fraction of the “mutant” samples for each
+pathway are burdened with excess passenger mutations. This could be addressed
+with more sophisticated filtering of likely passenger mutations [_e.g._
+@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
+incorporate these other data into a comprehensive pathway-centric analysis as we
+have done here for point mutations and indels.
+
+## Resource Availability
+
+Further information and requests for resources should be directed to and will
+be fulfilled by the Lead Contact, Dennis Hazelett (Dennis.Hazelett@cshs.org). Data
+generated in this study have been deposited to Zenodo (https://doi.org/10.5281/zenodo.4011291).
+All code for producing the analyses and figures herein are included in this fully reproducible 
+manuscript in R markdown format. R markdown files and all other models are available from our 
 repositories (https://github.com/dennishazelett/TTmanu) on the distributed
-version control site, GitHub. Extended descriptions of clustering, models,
+version control site, GitHub. An interactive media of the results is available on https://junkdnalab.shinyapps.io/PANCAN_supplemental/.
+
+## Methods
+
+Extended descriptions of clustering, models,
 formulae, pathway selection, and gene filtering are provided in supplementary
-methods. An interactive media of the results is available on junkdnalab.shinyapps.io/PANCAN_supplemental/.
+methods.
 
@@
 suggested improvements to the preprint. We also thank Ivetth Corona for early
-discussions and feedback.
+discussions and feedback. This work was supported by a grant from the 
+Cedars-Sinai Precision Health Initiative.
 
-## Declarations
+## Author Contributions
+P.N., S.G. and D.L. conducted the experiments. D.H. designed the experiments and wrote the paper.
 
-The authors have no conflicts of interest to declare. This work was supported by
-a grant from the Cedars-Sinai Precision Health Initiative.
+## Declaration of Interests
+
+The authors declare no competing interests.
 
 ## References
+
+<div id="refs"></div>
+
+## Figure and Table Legend
+**Fig 1. Clustering pathways of tumor samples.** A) Each of 377 selected
+Reactome pathways (rows) is classified as disrupted if one or more genes is
+mutated in the tumor sample (columns). Red cells denote pathway disruption.
+Tumors types reflect standardized abbreviations from the TCGA project. B)
+Different rotational perspectives of the same MCA-based UMAP projection in
+three-dimensional space. Each dot corresponds to a tumor sample. The same colors
+indicate the tumor's cluster identity throughout this manuscript.
+
+**Fig 2. Pathway-based clustering independent of tissue-of-origin** 
+A) Sankey plot of correspondence between cancer type and cluster identity.
+B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
+somatic mutation onto the UMAP. D) Projection of tumor stage onto the UMAP,
+regardless of cancer type. E) Projection of metastatic status onto the UMAP.
+Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors. F) Somatic mutation 
+frequencies for each cluster. Vertical axis shows log mutation count, horizontal 
+axis is cluster identity. Each dot represents an individual tumor sample, ranked 
+lowest to highest by mutation count. The median mutation count in each cluster is 
+indicated by the horizontal line.
+
+**Fig 3. Pan-cancer enrichment of pathway disruptions.** A) Heatmap shows
+relative enrichment of each pathway (rows) within numbered cluster (columns).
+Color represents effect size as percent enrichment. B) Proportion of mutated
+samples in the each significant pathway (columns; union set of pathways with
+effect size ≥ 0.30 in each cluster) within cluster (rows).
+
+**Fig 4. Gene level analysis reveals tissue-specific class signatures.** 
+Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
+95% confidence interval. Rows: significant genes from each cluster. Each cluster
+was compared against the background (all other clusters) to find differentially
+mutated genes. Significant genes (pvalue < 0.01) were selected and limited to
+the top ten results for each cluster. Heatmap; Columns: cancer type. Rows:
+significant genes. The heatmap shows proportion of samples per cluster and
+cancer type mutated for each gene. Depleted significant genes (left). Enriched
+significant genes (right).
+
+**Fig 5. Comparison of different cancers, simulated and actual results.**
+Comparison of survival in four simulated cohorts. A) A group consisting of
+randomly selected 30 year old women, unaffected (light blue) _vs._ cancer, 
+($k = 2$, light orange), another group consisting of random 30-70 year-old
+women, unaffecteds (dark blue) and cancer $k = 2$ (dark orange). B,C,D,E)
+Compare random group with correct age distribution for the given cancer
+(turqoise) to the model predicted survival for the given average k (blue) and
+the actual survival for the patients in our dataset (orange). Differences with
+actual data are likely due to a mixture of cluster identities which we exclude
+from these simulations for simplicity. F) Effects are compared between cancer
+type by showing the probability density for the posterior value of the $k_{tis}$
+multiplier. Larger values correspond to decreased age-independent life
+expectancy.
+
+**Fig 6. Cluster specific $k_{cl}$ values for each cancer type.** The overall
+$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific $k$
+represents relative aggressiveness of each cluster within cancer type. Alpha
+transparency (α) is set to reflect varying confidence in the posterior
+distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
+≥ 25 α = 1.0.
+
+**Table 1: Cluster-specific enriched pathways (effect size ≥ 0.30) in metastasis.**
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig 1
+```{r child='../code/figure1.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig 2
+```{r child = '../code/figure2.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig 3
+```{r child = '../code/figure3.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig 4
+```{r child = '../code/figure4.Rmd'}
+```
+
+#### Fig 5
+```{r child='../code/survival/figure5.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig 6
+
+```{r child='../code/figure6.Rmd'}
+```
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Table 1
+```{r child = '../code/table1.Rmd'}
+```

### 2021-05-03 text/supplemental_methods.Rmd
Edit manuscript for iscb iscience submission

--- 2020-08-31 text/supplemental_methods.Rmd
@@
 title: "Supplementary methods"
-author: "Peter T. Nguyen, Simon G. Coetzee, Daniel L. Lakeland, and Dennis J. Hazelett"
-date: "8/24/2020"
 bibliography: references.bib
+csl: plos-computational-biology.csl
 output:
   word_document:
-    reference_docx: "../text/word-style-reference-ttmanu.docx"
+    reference_docx: "../text/word-style-reference-ttmanu-iscb-iscience-v2.docx"
 always_allow_html: TRUE
@@
 
-## Supplementary Methods
-
 ### Clustering
@@
 explanatory power, using the average of 100 permutations of the data as baseline
-(**figure SM1A**). We then chose the maximum eigenvalue for which the p-value
-remained ≤ 0.05 (see cutoff in **figure SM1B**). Then we performed a UMAP
+(**SM1A Fig**). We then chose the maximum eigenvalue for which the p-value
+remained ≤ 0.05 (see cutoff in **SM1B Fig**). Then we performed a UMAP
 analysis [@McInnes2018], both in order to summarize the MCA graphically, and as
@@
 projections regardless of random seed setting. A representative version of this
-3D UMAP is shown in **figure SM1C**, rotated to enhance the visibility of the
+3D UMAP is shown in **SM1C Fig**, rotated to enhance the visibility of the
 major features. Following this spatial mapping we attempted to define groupings
@@
 < 5% in one cluster and chose a cluster size of 92 to minimize the score
-function (**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
-high-density clusters which we then projected onto the UMAP (**figure SM1E**).
+function (**SM1D Fig**). HDBSCAN with these settings resulted in ten distinct
+high-density clusters which we then projected onto the UMAP (**SM1E Fig**).
 This classified 6,038 out of 7,607 tumors but still resulted in a significant
@@
 number of tumor samples (87). Using this method, we assigned cluster membership
-to the remaining tumors (**figure SM1F**).
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure SM1
-```{r child='../code/figureSM1.Rmd'}
-```
-
-**Figure SM1: Method for tumor classification using pathway aberration data** 
-A) MCA analysis was performed to determine the eigenvalue with the most
-explanatory power from the categorical data. Average of 100 permutations of the
-data was used as a baseline. B) The p-value of each eigenvalue was plotted and
-the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
-projected down to 3 dimensions for the purpose of density based clustering (only
-the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
-used to perform hierarchical density based clustering tuned to minimize the
-fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
-the UMAP projection; tumors are colored by cluster membership (see legend). This
-results in some tumors remaining unclassified (grey points). F) kNN was used to
-assign membership to the remaining tumors.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
+to the remaining tumors (**SM1F Fig** and see interactive media from **S1 File** (junkdnalab.shinyapps.io/PANCAN_supplemental/)).
 
@@
 criteria to define basic cellular processes and biochemical pathways: (1)
-human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
+human-derived pathways ( _“HSA"_ ) (2) limited to grandchild node for each
 parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
@@
 approach was to determine the status of each gene (_i.e._ expressed or not
-expressed) in each tissue type in order to exclude gene mutations low expressed
-enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
+expressed) in each tissue type in order to exclude lowly expressed
+genes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
 effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
@@
 Though the model extends to earlier ages, relatively few patients were under age
-20 (**supplementary figure SM2**).
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure SM2
-
-```{r child='../code/figureSM2.Rmd'}
-```
-
-**Supplementary Figure SM2** Comparison between CDC data and ODE model fit for
-males and females of European descent. This comparison does not distinguish on
-ethnicity for simplicity in visualization.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
+20 (**SM2 Fig**).
 
@@
 is significantly less constrained so that there is a wider range of risk across
-tissue types (**supplementary figure SM3**).
-
-#### Figure SM3
-
-```{r sm3, fig.height=3, fig.width=3}
-ggplot(data.frame(x=0:30),aes(x)) + 
-  stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area", aes(fill = "ktis"))+ 
-  stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area", aes(fill = "kcl"))+ 
-  labs(title=expression(paste("Prior for k"["tis"], "and k"["cl"])),x= "k value", y = "Probability Density") +
-  theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
-  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) +
-  scale_fill_manual(values = c(alpha("#E69F00", 0.5),alpha("#0072B2", 0.5)),
-                    labels = c(expression(paste("k"["cl"])), expression(paste("k"["tis"])))) +
-  theme(plot.title = element_text(size = 12, hjust = 0.5),
-        panel.grid.major = element_line(size = 0.75),
-        panel.grid.minor = element_line(),
-        legend.position = c(0.8, 0.8),
-        legend.spacing.y = unit(0, "mm"),
-        legend.title = element_blank(),
-        legend.text = element_text(size = 8),
-        legend.key.size = unit(0.75,"line"),
-        legend.background = element_blank(),
-        legend.box.background = element_rect(colour = "black"))
-```
-
-**Supplementary Figure SM3: Priors for the tissue and class coefficients.** The tissue coefficient
-is chosen to admit parameter values across a range from near 0 to 20+
-in the high density region, reflecting the wide variety of severities
-across tissue type. Within a given tissue, the class k parameter
-multiplies the tissue k and is given a peak density at 1, and a
-credible range from near 0 to 5, reflecting our expectation of a
-moderate amount of variation across the classes.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
+tissue types (**SM3 Fig**).
 
@@
 
+```{r, results='asis'}
+cat("\\newpage")
+```
+
 ### References
+
+<div id="refs"></div>
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig SM1
+```{r child='../code/figureSM1.Rmd'}
+```
+
+**Fig SM1. Method for tumor classification using pathway aberration data** 
+A) MCA analysis was performed to determine the eigenvalue with the most
+explanatory power from the categorical data. Average of 100 permutations of the
+data was used as a baseline. B) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
+projected down to 3 dimensions for the purpose of density based clustering (only
+the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
+used to perform hierarchical density based clustering tuned to minimize the
+fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
+the UMAP projection; tumors are colored by cluster membership (see legend). This
+results in some tumors remaining unclassified (grey points). F) kNN was used to
+assign membership to the remaining tumors.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig SM2
+
+```{r child='../code/figureSM2.Rmd'}
+```
+
+**Fig SM2. Comparison between CDC data and ODE model fit for
+males and females of European descent.** This comparison does not distinguish on
+ethnicity for simplicity in visualization.
+
+```{r, results='asis'}
+cat("\\newpage")
+```
+
+#### Fig SM3
+
+```{r sm3, fig.height=3, fig.width=3}
+ggplot(data.frame(x=0:30),aes(x)) + 
+  stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area", aes(fill = "ktis"))+ 
+  stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area", aes(fill = "kcl"))+ 
+  labs(title=expression(paste("Prior for k"["tis"], "and k"["cl"])),x= "k value", y = "Probability Density") +
+  theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
+  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) +
+  scale_fill_manual(values = c(alpha("#E69F00", 0.5),alpha("#0072B2", 0.5)),
+                    labels = c(expression(paste("k"["cl"])), expression(paste("k"["tis"])))) +
+  theme(plot.title = element_text(size = 12, hjust = 0.5),
+        panel.grid.major = element_line(size = 0.75),
+        panel.grid.minor = element_line(),
+        legend.position = c(0.8, 0.8),
+        legend.spacing.y = unit(0, "mm"),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 8),
+        legend.key.size = unit(0.75,"line"),
+        legend.background = element_blank(),
+        legend.box.background = element_rect(colour = "black"))
+```
+
+**Fig SM3. Priors for the tissue and class coefficients.** The tissue coefficient
+is chosen to admit parameter values across a range from near 0 to 20+
+in the high density region, reflecting the wide variety of severities
+across tissue type. Within a given tissue, the class k parameter
+multiplies the tissue k and is given a peak density at 1, and a
+credible range from near 0 to 5, reflecting our expectation of a
+moderate amount of variation across the classes.

### 2021-05-03 text/survival.Rmd
Edit manuscript for iscb iscience submission

@@
 used Bayesian inference to test models of survival using public longevity data
-from the CDC. The estimates the effect of cancer type and cluster-specific
+from the CDC. The model estimates the effect of cancer type and cluster-specific
 cancer effects independently, resulting in a cancer and cluster-specific
@@
 
-To illustrate the model, we simulate survival four related groups (**figure
+To illustrate the model, we simulate survival for four related groups (**Fig
 5A**). We start with a cohort of women aged 30 and show expected survival (light
@@
 
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 5
-```{r child='../code/survival/figure5.Rmd'}
-```
-
-**Figure 5: Comparison of different cancers, simulated and actual results.**
-Comparison of survival in four simulated cohorts. A) A group consisting of
-randomly selected 30 year old women, unaffected (light blue) _vs._ cancer, 
-($k = 2$, light orange), another group consisting of random 30-70 year-old
-women, unaffecteds (dark blue) and cancer $k = 2$ (dark orange). B,C,D,E)
-Compare random group with correct age distribution for the given cancer
-(turqoise) to the model predicted survival for the given average k (blue) and
-the actual survival for the patients in our dataset (orange). Differences with
-actual data are likely due to a mixture of cluster identities which we exclude
-from these simulations for simplicity. F) Effects are compared between cancer
-type by showing the probability density for the posterior value of the $k_{tis}$
-multiplier. Larger values correspond to decreased age-independent life
-expectancy.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
 We found that cancer types, as expected, have a range of prognoses relative to
-the general population. In **figure 5B-E** we see three particularly deadly
+the general population. In **Fig 5B-E** we see three particularly deadly
 cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
@@
 The degree to which cancer accelerates aging can be determined by the $k_{tis}$
-multiplier (**figure 5F**) multiplied by the cluster modifier $k_{cl}$ (**figure
+multiplier (**Fig 5F**) multiplied by the cluster modifier $k_{cl}$ (**Fig
 6**). Looking at the tissue specific multiplier, the least deadly was prostate
@@
 
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Figure 6
-
-```{r child='../code/figure6.Rmd'}
-```
-
-**Figure 6 Cluster specific $k_{cl}$ values for each cancer type.** The overall
-$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific $k$
-represents relative aggressiveness of each cluster within cancer type. Alpha
-transparency (α) is set to reflect varying confidence in the posterior
-distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N
-≥ 25 α = 1.0.
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
 Our estimates of tissue-specific cluster effects, in contrast, ranged from less
 than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
-or more deadly than other clusters within each cancer type (**figure 6**). A
+or more deadly than other clusters within each cancer type (**Fig 6**). A
 cluster-specific rate of 1 represents the typical rate for this tissue type. For
@@
 mortalities in the data to make an estimate (as expected for PRAD and THCA) or
-two few samples, period. We did not observe cluster-specific trends that held
+too few samples, period. We did not observe cluster-specific trends that held
 true across cancer types, which could result from different cancers having

### 2021-06-28 text/survival.Rmd
modify survival.Rmd to account for our new effective age model and describe the new graphics for Fig 5,6

@@
 cancer effects independently, resulting in a cancer and cluster-specific
-estimate of the aging rate multiplier, $k$.
+estimate of an effective age function and the aging rate multiplier, $k$.
 
 To illustrate the model, we simulate survival for four related groups (**Fig
-5A**). We start with a cohort of women aged 30 and show expected survival (light
-blue). By setting cancer rate multiplier to $k=2$ we simulate the effect of a
-moderately deadly cancer (yellow). Compare this to randomly selected women
-ranging from 30 to 70 years old (dark blue). Immediately, survival changes due
-to the mixture of ages, without malignancy. Adding a $k=2$ malignancy further
-reduces expected lifespan (red). Hence, due to the change in distribution of
-age-at-diagnosis, we expect equally deadly cancers to have *different* survival
-curves. The model takes this into account and cancer aggressiveness is estimated
-accurately without the confounding effect of age.
+5A**). We start with a cohort of women aged 30 randomly selected from the US
+population, and show expected survival (light blue). We then use typical values
+from the posterior samples to set the effective age of that cohort to the
+effective age of a BRCA patient (yellow). Finally to illustrate how the model
+varies by effective age we add an additional artificial 10 years to the BRCA
+effective age (dark blue). Compare these results to a randomly selected group of
+women ranging from 30 to 70 years old (green). Immediately, survival changes due
+to the mixture of ages, without malignancy. By computing the effective ages for
+this mixture of patients, we see that malignancy further reduces expected
+lifespan (pink). Adding an additional 10 years to effective age would reduce the
+lifespan even further (dark orange). Hence, simply due to the change in
+distribution of age-at-diagnosis, we expect two equally deadly cancers to have
+*different* survival curves. The model takes this into account and cancer
+aggressiveness is estimated accurately without the confounding effect of age
+through the effective age and aging rate parameters.
 
 We found that cancer types, as expected, have a range of prognoses relative to
-the general population. In **Fig 5B-E** we see three particularly deadly
-cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
-where diagnosis decreases risk relative to background (Thyroid: THCA). Cancers
-with posterior probability for relative risk of less than 1 such as THCA should
-be interpreted carefully. This Bayesian model is for a state of information. The
-information that a person is diagnosed with cancer may lead us to expect shorter
-survival than the general population of matched age (k > 1), or longer survival
-than the general population of matched age (k < 1). Shorter prognosis could
-result from cancer aggressiveness, injuring the body and causing death. By
-contrast, longer prognosis could result from the cancer being relatively mild,
-and therefore diagnosis could be an informational signal that the patient is
-health conscious, with the comparison group having more people whose cancers go
-undiagnosed. It is important to note therefore that the diagnosis can increase
-our expectation of life relative to the comparison group, even if it decreases
-the expectation of life of the individual relative to the counterfactual where
-they did not have cancer.
+the general population. In **Fig 5B-D** we show how effective age baseline, and
+effective age rate (which modifies the baseline based on your actual age) as
+well as the ktis which accelerates or decelerates further aging, varies
+considerably from cancer to cancer. For example a 20 year old diagnosed with
+THCA or PRAD would be threated similarly to a 40 year old based on teh Effective
+Age parameter. It's noticeable that for every year beyond 20 in the PRAD
+diagnosis we add only an extra ~ 0.2 years (Effective Age Rate). The ktis
+parameter represents an acceleration (greater than 1) or deceleration (less) of
+the further aging beyond diagnosis. For example for BRCA patients each year past
+diagnosis adds about 1 year of risk, whereas for LGG patients it adds around 2,
+and for SARC patients it adds only 0.25 or so.
 
-The degree to which cancer accelerates aging can be determined by the $k_{tis}$
-multiplier (**Fig 5F**) multiplied by the cluster modifier $k_{cl}$ (**Fig
-6**). Looking at the tissue specific multiplier, the least deadly was prostate
-cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a risk
-multiplier of between 15 and 22 relative to the background risk of death in the
-population. This undoubtedly is influenced by the fact that relatively young
-patients are affected by GBM and that it is extremely deadly. Among the
-deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
-pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
-in relative risk of age-adjusted death, the remaining cancer types ranged from
-about 1 to 5 in magnitude.
+In **Fig 5E-H** we see three particularly deadly cancers (Glioblastoma:GBM,
+Pancreatic:PAAD, Ovarian:OV), and one cancer where diagnosis hardly changes risk
+relative to background (Thyroid: THCA). Cancers with posterior probability for
+relative risk of less than 1 should be interpreted carefully. This Bayesian
+model is for a state of information. The information that a person is diagnosed
+with cancer may lead us to expect shorter survival than the general population
+of matched age, or longer survival than the general population of matched
+age. Shorter prognosis could result from cancer aggressiveness, injuring the
+body and causing death. By contrast, longer prognosis could result from the
+cancer being relatively mild, and therefore diagnosis could be an informational
+signal that the patient is health conscious, with the comparison group having
+more people whose cancers and other health issues go undiagnosed. It is
+important to note therefore that the diagnosis can increase our expectation of
+life relative to the comparison group, even if it decreases the expectation of
+life of the individual relative to the counterfactual where they did not have
+cancer.
 
-Our estimates of tissue-specific cluster effects, in contrast, ranged from less
-than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
-or more deadly than other clusters within each cancer type (**Fig 6**). A
-cluster-specific rate of 1 represents the typical rate for this tissue type. For
-several cancers (_e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
-lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
-indistinguishable from the prior, reflecting that either there were too few
-mortalities in the data to make an estimate (as expected for PRAD and THCA) or
-too few samples, period. We did not observe cluster-specific trends that held
-true across cancer types, which could result from different cancers having
-different standards of care for example. In support of this interpretation, we
-also tested a factored model which considered both cancer type cluster
-independently. Though we were able to successfully fit this model, it is a
-special case of the more general model where cluster and tissue independently
-affect longevity, and there is no reason to believe that cluster specific
-effects would necessarily remain constant across tissue types given how widely
-the patients vary across tissue types, drug and surgical treatments.
+Our estimates of tissue-specific cluster effects were for the Effective Age
+parameter. These cluster effects were multiplied by the tissue specific
+coefficient to form the full effective age. Priors for this parameter were
+relatively peaked around 1 reflecting the purpose of this coefficient as a
+multiplicative perturbation (**Fig 6**). A cluster-specific rate of 1 represents
+the typical rate for this tissue type. For several cancers (_e.g._  PRAD, kidney
+chromophobe (KICH), diffuse large B-cell lymphoma (DLBC), thyroid cancer (THCA))
+the posterior estimates are largely indistinguishable from the prior, reflecting
+that either there were too few mortalities in the data to make an estimate (as
+expected for PRAD and THCA) or too few samples, period. We did not observe
+cluster-specific trends that held true across cancer types, which could result
+from different cancers having different standards of care for example. For some
+cancers however, there is evidence that certain forms are more or less
+aggressive. PAAD class 8 seems to raise your effective age somewhat relative to
+other forms, as does LGG group 7, whereas BRCA group 5, COAD group 8, and LGG
+group 5 all seem somewhat lower risk than the other classes within those
+tissues.

### 2021-06-28 text/supplemental_methods.Rmd
modify supplemental methods section to reflect the newer model

@@
 grows exponentially in time at some baseline rate that matches the observed CDC
-data. The rate of growth in risk is assumed to jump to a higher level at
-diagnosis, which we take as a proxy for onset. This represents an accelerated
-aging type model. To fit the model, we constrain the baseline risks using the
-available CDC life tables (ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We also constrain the cancer specific multiplier
-using the longevity data within the cancer tumor dataset. Our model estimates
-the effect of cancer type and class-specific cancer effects independently,
-resulting in a cancer and class specific estimate of the aging rate multiplier.
+data. We assume that at diagnosis this information will allow us to treat
+patients as if they were a different age, and with further time passing at a
+different rate, a self-similarity based model. The effective age at diagnosis is
+based on a baseline age for that tissue/class combination, plus a perturbation
+for each year over 20 years old at diagnosis. To fit the model, we constrain the
+baseline risks using the available CDC life tables
+(ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We also
+constrain the cancer specific parameters using the longevity data within the
+cancer tumor dataset. Our model estimates the effect of cancer type and
+class-specific cancer effects independently, resulting in a cancer and class
+specific estimate of the effective age.
 
@@
 accuracy in the right tail of the distribution. We modeled baseline longevity
-using a risk rate function:
+using a risk rate function. The risk rate of death at a time $t$ days after the
+$A$ birthday is given as:
 
@@
 
-Time is split into two components. $A$ going from 0 up to the age of the patient
-at diagnosis and then remaining fixed, whereas $t$ starts at 0 upon diagnosis
-and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
+Time is split into two components. $A$ represents the "effective age" of the
+patient at diagnosis, whereas $t$ starts at 0 upon diagnosis and represents the
+days since diagnosis, a field recorded in the dataset. $D(t)$
 is an indicator for whether diagnosis has occurred or not (before diagnosis
@@
 to the age at diagnosis. Risk of death assumes a new rate $ka$ thereafter.
-Though the model extends to earlier ages, relatively few patients were under age
-20 (**SM2 Fig**).
 
@@
 
+However, to model the risk of cancer effectively we found that it was necessary
+not to use the actual age $\A'$ but instead the "effective age" for that
+tissue/class type. Our effective age is calculated as:
+
+$$A = A_{0} A_{cl} + A_r (A'-20)$$ Where $A_0$ is the tissue specific effective age
+parameter (expressed in years), $A_{cl}$ is a dimensionless multiplier for the
+given class, and $A_r$ is the dimensionless age rate which determines how actual
+age $A'$ affects effective age A using a linear perturbation.
+
 The $R(T)$ function has two important parameters, the $a$ value which represents
-the background risk rates, and the $k$ value, which is a function of both cancer
-tissue type, and the cancer cluster identity.
+the background risk rates, and the $k$ value, which is a function of cancer
+tissue type.
 
-$$k = k_{tis} k_{cl}$$
+$$k = k_{tis}$$
 
-The product $k_{tis}k_{cl}$ is obviously symmetric between the two factors. To
-disambiguate the meaning of the two, the prior distribution of $k_{cl}$ has peak
+The product $A_0 A_{cl}$ is obviously symmetric between the two factors. To
+disambiguate the meaning of the two, the prior for $A_0$ ranges from about 40 to
+100 years, whereas the prior distribution of $A_{cl}$ has peak
 probability at 1 and a relatively narrow width, due to the fact that it is a
-tissue specific multiplicative modifying factor. This allows the overall
-magnitude of $k$ to be primarily determined by the $k_{tis}$ value whose prior
-is significantly less constrained so that there is a wider range of risk across
-tissue types (**SM3 Fig**).
-
-The $k$ in our model can be taken as, roughly, the aging rate relative to the
-baseline rates. Thus, a value of $k=2$ means that a given cancer causes you to
-experience the same risk in time $dt$ as a non-cancer patient would experience
-at the same age, in time $2dt$.
+tissue specific multiplicative modifying factor. To understand the roles of
+these priors refer to (**SM3 Fig**).
 
@@
 for the following reasons. In the tumor sample data, stage at diagnosis is
-confounded with cancer type because some cancers are screened aggressively (_e.g._ colorectal and prostate cancers) while others are diagnosed typically
+confounded with cancer type because some cancers are screened aggressively
+(_e.g._ colorectal and prostate cancers) while others are diagnosed typically
 after they become problematic for the patient's lifestyle (_e.g._ ovarian and
@@
 
-**Fig SM3. Priors for the tissue and class coefficients.** The tissue coefficient
-is chosen to admit parameter values across a range from near 0 to 20+
-in the high density region, reflecting the wide variety of severities
-across tissue type. Within a given tissue, the class k parameter
-multiplies the tissue k and is given a peak density at 1, and a
-credible range from near 0 to 5, reflecting our expectation of a
-moderate amount of variation across the classes.
+**Fig SM3. Priors for the tissue and class coefficients.** The tissue
+coefficient is chosen to admit ages between 40 and 100 years old as a baseline
+age. Within a given tissue, the class parameter multiplies the tissue value and
+is given a peak density at 1, and a credible range from near 0.6 to 1.4,
+reflecting our expectation of a moderate multiplicative perturbation of
+variation across the classes.

### 2021-06-28 text/supplemental_methods.Rmd
Update sm3 fig with effage tis and cl

@@
 ```{r sm3, fig.height=3, fig.width=3}
-ggplot(data.frame(x=0:30),aes(x)) + 
-  stat_function(fun=function(x) dgamma(x,1.2,.2),geom="area", aes(fill = "ktis"))+ 
-  stat_function(fun=function(x) dgamma(x,3.5,2.5),geom="area", aes(fill = "kcl"))+ 
-  labs(title=expression(paste("Prior for k"["tis"], "and k"["cl"])),x= "k value", y = "Probability Density") +
-  theme_classic() + scale_x_continuous(limits = c(0, 20), breaks = c(seq(0,20, by =5))) +
-  scale_y_continuous(limits = c(0,0.75), breaks = c(seq(0,0.75, by = 0.25))) +
-  scale_fill_manual(values = c(alpha("#E69F00", 0.5),alpha("#0072B2", 0.5)),
-                    labels = c(expression(paste("k"["cl"])), expression(paste("k"["tis"])))) +
+sm3.1 <- ggplot(data.frame(x=0:110),aes(x)) + 
+  stat_function(fun=function(x) dgamma(x,30.0,29.0/65),geom="area", fill = alpha("black", 0.5)) +
+  theme_classic() + 
+  scale_x_continuous(limits = c(30, 110), breaks = c(seq(30,110, by =10))) +
   theme(plot.title = element_text(size = 12, hjust = 0.5),
         panel.grid.major = element_line(size = 0.75),
-        panel.grid.minor = element_line(),
-        legend.position = c(0.8, 0.8),
-        legend.spacing.y = unit(0, "mm"),
-        legend.title = element_blank(),
-        legend.text = element_text(size = 8),
-        legend.key.size = unit(0.75,"line"),
-        legend.background = element_blank(),
-        legend.box.background = element_rect(colour = "black"))
+        panel.grid.minor = element_line()) +
+  labs(x= expression(paste("Prior for Effage"["tis"])), y = "Probability Density")
+
+sm3.2 <- ggplot(data.frame(x=0:3),aes(x)) + 
+  stat_function(fun=function(x) dgamma(x,50.0,49.0/1.0),geom="area", fill = alpha("black", 0.5)) +
+  theme_classic() + 
+  scale_x_continuous(limits = c(0.5, 1.6), breaks = c(seq(0.6, 1.5, by = 0.2))) +
+  theme(plot.title = element_text(size = 12, hjust = 0.5),
+        panel.grid.major = element_line(size = 0.75),
+        panel.grid.minor = element_line()) +
+  labs(x= expression(paste("Prior for Effage"["cl"])), y = "Probability Density")
+
+layout = "
+AABB
+AABB
+"
+
+sm3.1 + sm3.2 + plot_annotation(tag_levels = "A") + plot_layout(design = layout)
 ```
 
-**Fig SM3. Priors for the tissue and class coefficients.** The tissue
+**Fig SM3. Priors for the tissue and class coefficients.** A) The tissue
 coefficient is chosen to admit ages between 40 and 100 years old as a baseline
-age. Within a given tissue, the class parameter multiplies the tissue value and
+age. B) Within a given tissue, the class parameter multiplies the tissue value and
 is given a peak density at 1, and a credible range from near 0.6 to 1.4,

### 2021-06-29 text/supplemental_methods.Rmd
Adjust width and update xaxis (Effage -> A) for SM3 Fig. Fix A' not italicized in text.

@@
 However, to model the risk of cancer effectively we found that it was necessary
-not to use the actual age $\A'$ but instead the "effective age" for that
+not to use the actual age $A'$ but instead the "effective age" for that
 tissue/class type. Our effective age is calculated as:
@@
 
-```{r sm3, fig.height=3, fig.width=3}
+```{r sm3, fig.height=3, fig.width=6}
 sm3.1 <- ggplot(data.frame(x=0:110),aes(x)) + 
@@
         panel.grid.minor = element_line()) +
-  labs(x= expression(paste("Prior for Effage"["tis"])), y = "Probability Density")
+  labs(x= expression(paste("Prior for A"["0"])), y = "Probability Density")
 
@@
         panel.grid.minor = element_line()) +
-  labs(x= expression(paste("Prior for Effage"["cl"])), y = "Probability Density")
+  labs(x= expression(paste("Prior for A"["cl"])), y = "Probability Density")
 

### 2021-06-29 code/figure6.Rmd
Update fig5 and 6 with new model. Fix knit error with figure 3

@@
 cancerinfo <- read_csv("../data/survival/clinical_plus_cluster.csv")
-cancer.clusttype <- cancerinfo %>% dplyr::select(type, race, clust_knn) %>% ## select relevant columns
-  filter(race %in% c("WHITE", "BLACK OR AFRICAN AMERICAN")) ## filter black and white only
-cancer.race <- as.data.frame(table(cancer.clusttype$type)) ## table cancer type
-cancer.race$facetname <- paste0(cancer.race$Var1, " (N = ", cancer.race$Freq, (")")) ## Create a new col with type and freq together
-cancer.race$Freq <- NULL ## Remove Freq Column
-names(cancer.race) <- c("tiss", "facetname") ## rename cols for left join in kclassplot below
+cancer.clusttype <- cancerinfo %>% dplyr::select(type, race, clust_knn) ## select relevant columns
+n_cancer <- as.data.frame(table(cancer.clusttype$type)) ## table cancer type
+n_cancer$facetname <- paste0(n_cancer$Var1, " (N = ", n_cancer$Freq, (")")) ## Create a new col with type and freq together
+n_cancer$Freq <- NULL ## Remove Freq Column
+names(n_cancer) <- c("tiss", "facetname") ## rename cols for left join in kclassplot below
 
@@
 ## Processing data
-kclnames = grep("k\\[",names(dfsamps), value=TRUE)
-dfclas = dfsamps[,kclnames]
+load("../data/survival/survmodel_effcl_knn_062521.stansave")
+dfsamps <- as.data.frame(samps)
+clnames = grep("effage0\\[",names(dfsamps), value=TRUE)
+dfclas = dfsamps[,clnames]
 dfclas = stack(dfclas)
 names(dfclas) = c("value","coefname")
-dfclas$class = as.factor(as.numeric(gsub("k\\[(.*),(.*)\\]","\\2",dfclas$coefname)))
-dfclas$tiss = tisnames[as.numeric(gsub("k\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
+dfclas$class = as.factor(as.numeric(gsub("effage0\\[(.*),(.*)\\]","\\2",dfclas$coefname)))
+dfclas$tiss = tisnames[as.numeric(gsub("effage0\\[(.*),(.*)\\]","\\1",dfclas$coefname))]
 dfclas$class <- fct_rev(dfclas$class)
@@
 ## Left joining data created above
-dfclas <- left_join(dfclas, cancer.race, by = "tiss") ## add facet names
+dfclas <- left_join(dfclas, n_cancer, by = "tiss") ## add facet names
 
@@
 
+normal_lines <- data.frame(facetname = unique(dfclas$facetname),
+                         x0 = 1) ## Add dashed line at x = 1
+
 	## Plotting
@@
 		facet_wrap(. ~ facetname,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
-		xlim(0,3) + theme(legend.position = "none") + ggtitle("Class Specific Coefficients") +
-	  labs(x = expression(paste("k"["cl"])), y = "Cluster Identity") +
-	  #geom_text(data = dfclas, aes(label = lab), x = 2.5, y = 6) +
-	  mytheme + theme(plot.title= element_text(hjust = 0.5))+ turn.off.legend +
-	  scale_fill_manual(values = cluster.colours)
+	  theme(legend.position = "none") + ggtitle("Class Specific Coefficients") +
+	  labs(x = expression(paste("A"["cl"])), y = "Cluster Identity") +
+	  theme(plot.title= element_text(hjust = 0.5)) +
+	  scale_fill_manual(values = cluster.colours) +
+	  scale_x_continuous(breaks=seq(0.5,1.5, by = 0.5), limits = c(0.5,1.5)) +
+    geom_vline(aes(xintercept=1), linetype = "dashed", color = "red")
 ```

### 2021-06-29 code/survival/figure5.Rmd
Update fig5 and 6 with new model. Fix knit error with figure 3

```{r, survfig, echo=FALSE, warning=FALSE, error=TRUE,message=FALSE,fig.height=11, fig.width=9}

library(ggridges)
library(patchwork)

source("./kmcurve.R")

## create example data for a fixed 30 year old without vs with cancer (effective age)
## cancer

## Effective age calc for typical BRCA
effage0tiss <- 62
effage0 <- 1.1
effagert <- 0.35

patientsurvivalsbkd = replicate(1000,simdeath(r20wf,ageratewf,30,1))/365 ## Typical survival for 30yr with k = 1
## Stimulating survival for white female with cancer - effective age for 30 yr old
effage <- effage0tiss * effage0 + effagert * (30 - 20)
patientsurvivalseffk1 <- replicate(1000,simdeath(r20wf,ageratewf,effage,1))/365 ## Typical survival for 30yr with effage due to cancer k=1
patientsurvivalseff10k1 <- replicate(1000,simdeath(r20wf,ageratewf,effage+10,1))/365 ## effage + 10 but still k = 1
## Mixed ages
mixedages = runif(1000,30,70)
mixedsurvwfk1 = sapply(mixedages, function(x) simdeath(r20wf,ageratewf,x,1)/365) ## Typical survival for mixed ages
## Stimulating survival for white female with cancer - mixed age
effage.mixed <- effage0tiss * effage0 + effagert * (mixedages - 20)
mixedsurvwfeffk1 = sapply(effage.mixed, function(x) simdeath(r20wf,ageratewf,x,1)/365) ## Typical survival for mixed ages with effage due to k=1
mixedsurvwfeff10k1 <- sapply(effage.mixed+10, function(x) simdeath(r20wf,ageratewf,x,1)/365)

simdata = data.frame(
  r = patientsurvivalsbkd,
  ce = patientsurvivalseffk1,
  ce10 = patientsurvivalseff10k1,
  m = mixedsurvwfk1,
  me = mixedsurvwfeffk1,
  me10 = mixedsurvwfeff10k1,
  status = rep(1, length(patientsurvivalsbkd))
) %>%
  rownames_to_column(var = "index") %>%
  pivot_longer(-c(index, status), values_to = "time", names_to = "condition")
p1 <- ggplot(simdata, aes(time = time, color = condition, status = status)) +
  geom_km() +
  labs(x = "Survival Time (years past diagnosis)", y = "Probability") +
  scale_color_manual(
    values = c(r = "#56b4e9", ce = "#e69f00", ce10 = "#0072b2", m = "#009E73", me = "#CC79A7", me10 = "#D55E00")
  ) +
  theme_classic() +
  annotate("text", x = 44, y = .75, label = "30yo women", color = "#56b4e9", size = 3) +
  annotate("text", x = 27, y = .70, label = "30yo women cancer BRCA typ", color = "#e69f00", size = 3) +
  annotate("text", x = 25, y = .65, label = "30yo women cancer BRCA+10yr", color = "#0072b2", size = 3) +
  annotate("text", x = 42, y = .45, label = "Mixed age women", color = "#009E73", size = 3) +
  annotate("text", x = 29, y = .40, label = "Mixed age women BRCA typ", color = "#CC79A7", size = 3) +
  annotate("text", x = 27, y = .35, label = "Mixed age women BRCA+10yr", color = "#D55E00", size = 3) +
  theme(legend.position = "none")

load("./survmodel_effcl_knn_062821.stansave")
cancerdat = read_csv("../../data/survival/clinical_plus_cluster.csv");
cancerdat$type <- as.factor(cancerdat$type)
tisnames <- levels(cancerdat$type)
## tissue specific effective age

effage0tiss <- as.data.frame(samps) %>% select(starts_with("effage0tiss[")) %>%
  gather(data = ., key = "coefname", value = "effage0tiss") %>%
  mutate(tiss = tisnames[as.numeric(gsub("effage0tiss\\[(.*)\\]","\\1",.$coefname))]) ## Replace with tumor type

p2 <- ggplot(effage0tiss,aes(x = effage0tiss, y = tiss)) + geom_density_ridges() +
  theme_minimal() +
  scale_x_continuous(breaks=seq(20,100, by = 20), limits = c(20,100)) +
  labs(x = "Effective Age", y = "Cancer Type") +
  theme(legend.position = "none") +
  geom_vline(aes(xintercept=20), linetype = "dashed", color = "red")

## tissue specific effective age rate

effagert <- as.data.frame(samps) %>% select(starts_with("effagert")) %>%
  gather(data = ., key = "coefname", value = "effagert") %>%
  mutate(tiss = tisnames[as.numeric(gsub("effagert\\[(.*)\\]","\\1",.$coefname))]) ## Replace with tumor type

p3 <- ggplot(effagert,aes(x = effagert, y = tiss)) + geom_density_ridges() +
  theme_minimal() +
  labs(x = "Effective Age Rate", y = "") +
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        plot.title= element_text(hjust = 0.5)) +
  geom_vline(aes(xintercept=1), linetype = "dashed", color = "red")

ktis <- as.data.frame(samps) %>% select(starts_with("ktis")) %>%
  gather(data = ., key = "coefname", value = "ktis") %>%
  mutate(tiss = tisnames[as.numeric(gsub("ktis\\[(.*)\\]","\\1",.$coefname))]) ## Replace with tumor type

p4 <- ggplot(ktis,aes(x = ktis, y = tiss)) + geom_density_ridges() +
  theme_minimal() +
  scale_x_continuous(breaks=seq(-1,3, by = 1), limits = c(-0.25,3)) +
  labs(x = expression(paste("k"["tis"])), y = "") +
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        plot.title= element_text(hjust = 0.5)) +
  geom_vline(aes(xintercept=1), linetype = "dashed", color = "red")

## Plot survival by cancer type
asamps <- as.array(samps)

endpoints = sqldf("select case when death_days_to is NULL then last_contact_days_to else death_days_to end as finaltime, case when death_days_to is NULL then 0 else 1 end as finalstatus from cancerdat;")

cancerdat = cbind(cancerdat,endpoints);

cancerdat = cancerdat[!is.na(cancerdat$finaltime),]

cancertype <- c("GBM", "PAAD", "OV", "THCA")
for(ct in cancertype) {
  lev <- which(levels(cancerdat$type) == ct)[[1]] ## level of cancer type
  women <- cancerdat[cancerdat$type == ct & cancerdat$gender == "FEMALE",]
  agesw <- women$age

  ## Mean of various parameters
  k = round(mean(asamps[,,sprintf("ktis[%d]",lev)]), digits = 2)
  effage0tiss <- round(mean(asamps[,,sprintf("effage0tiss[%d]",lev)]), digits = 0)
  effagert <- round(mean(asamps[,,sprintf("effagert[%d]",lev)]), digits = 2)
  effage0 <- round(median(asamps[,,sprintf("effage0[%d,%d]",lev,1:10)]), digits = 2)

  ## Simulation of women no cancer
  survyrswfref = sapply(agesw,function(a) tryCatch(simdeath(r20wf,ageratewf,a,1)/365,error= function(e) {0}))
  ## Effective age due to cancer
  effage.agesw <- effage0tiss * effage0 + effagert * (agesw - 20)
  survyrswf = sapply(effage.agesw,function(a) tryCatch(simdeath(r20wf,ageratewf,a,kmean)/365,error= function(e) {0}))

  df.plot <- plotkmcomp(survyrswf, survyrswfref, women$finaltime / 365, women$finalstatus) +
          coord_cartesian(xlim = c(0, 80)) +
          labs(
              x = "Survival Time (years past diagnosis)",
              y = "Probability"
          ) +
          scale_color_manual(
              values = c(
                  rtime = "#66c2a5",
                  stime = "#8da0cb",
                  acttime = "#fc8d62"
              ),
              breaks = c("rtime", "stime", "acttime"),
              labels = c(
                  "Simulated General Population; k = 1",
                  "Simulated Patient Population",
                  "Actual Patient Population"
              )
          ) +
    annotate(geom = "text", x = 50, y = 1.0, label = paste0(ct, " simulated")) +
    theme(legend.position = "none")
  if(ct == "GBM") {
    df.plot <- df.plot + annotate(geom = "text", x = 40, y = 0.7, label = "Simulated General Population; k = 1", color = "#66c2a5", size = 3) +
      annotate(geom = "text", x = 32, y = 0.5, label = "Simulated Patient Population", color = "#8da0cb", size = 3) +
      annotate(geom = "text", x = 27, y = 0.3, label = "Actual Patient Population", color = "#fc8d62", size = 3)
  }
  assign(ct, df.plot, envir = .GlobalEnv)
}

layout = "
AAAEE
AAAEE
AAAEE
AAAFF
AAAFF
AAAFF
BCDGG
BCDGG
BCDGG
BCDHH
BCDHH
BCDHH
"

p1 + p2 + p3 + p4 + GBM + PAAD + OV + THCA + plot_spacer() + plot_annotation(tag_levels = "A") + plot_layout(design = layout)

```

### 2021-06-30 code/supplemental_file_shiny/app.R
Update survival tab with effective age model. Remove ethnicity option

@@ -1,7 +1,10 @@
 ## library
+#BiocManager::install(version = "3.12")
+
 suppressPackageStartupMessages(c(
     library(BiocManager),
     options(repos = BiocManager::repositories()),
+    library(devtools),
     library(shiny),
     library(shinyWidgets),
     library(tidyverse),
@@ -11,7 +14,6 @@ suppressPackageStartupMessages(c(
     library(rgl),
     library(htmltools),
     library(dplyr),
-    library(rgl),
     library(htmltools),
     library(grid),
     library(gridBase),
@@ -21,6 +23,10 @@ suppressPackageStartupMessages(c(
     library(autoimage)
 ))
 
+
+#devtools::install_github("michaelway/ggkm")
+#library(ggkm)
+
 ## Load data
 umap_cancertype <- fread("umap_3d_coors.tsv")
 load("kelly.colours.rda")
@@ -29,7 +35,8 @@ load("tumor_samples.Rda")
 load("supplemental_gene.Rda")
 load("gdc_reactome_path_analysis.Rda")
 load("cancerdat.rda")
-load("survmodel-ethbkd.stansave")
+load("survmodel_effcl_knn_062821.stansave")
+dfsamps <- as.data.frame(samps)
 load("dfclas.rda")
 
 cluster.colours <- c(kelly.colours[c(3:12)], "grey") ## Color scheme
@@ -97,9 +104,9 @@ ui <- (fluidPage(
                                          options = list(`actions-box` = TRUE)),
                              pickerInput("sim.sex", label = "Sex:", ## Sex
                                          choices = list("MALE", "FEMALE"), selected = "FEMALE"),
-                             pickerInput("sim.ethnic", label = "Ethnicity:", ## Ethnicity
-                                         choices = list("European descent" = "WHITE", 
-                                                        "African descent"= "BLACK OR AFRICAN AMERICAN"), selected = "European descent"),
+                             # pickerInput("sim.ethnic", label = "Ethnicity:", ## Ethnicity
+                             #             choices = list("European descent" = "WHITE", 
+                             #                            "African descent"= "BLACK OR AFRICAN AMERICAN"), selected = "European descent"),
                              sliderInput("sim.range", "Age range:", ## Age range
                                          min = 1, max = 100,
                                          value = c(30,70)),
@@ -372,11 +379,17 @@ server <- (function(input, output, session) {
[…]

### 2021-06-30 text/supplemental_methods.Rmd

@@
 
-$$A = A_{0} A_{cl} + A_r (A'-20)$$ Where $A_0$ is the tissue specific effective age
-parameter (expressed in years), $A_{cl}$ is a dimensionless multiplier for the
-given class, and $A_r$ is the dimensionless age rate which determines how actual
-age $A'$ affects effective age A using a linear perturbation.
+$$A = A_{0} A_{cl} + A_r (A'-20)$$ Where $A_0$ is the tissue specific effective
+age parameter (expressed in years), $A_{cl}$ is a dimensionless multiplier for
+the given class, and $A_r$ is the dimensionless age rate which determines how
+actual age $A'$ affects effective age A using a linear perturbation. For a
+non-cancer patient, $A_0=20$, $A_{cl}=1$, and $A_r=1$ are the relevant
+benchmarks.
 
@@
 The product $A_0 A_{cl}$ is obviously symmetric between the two factors. To
-disambiguate the meaning of the two, the prior for $A_0$ ranges from about 40 to
-100 years, whereas the prior distribution of $A_{cl}$ has peak
-probability at 1 and a relatively narrow width, due to the fact that it is a
-tissue specific multiplicative modifying factor. To understand the roles of
+disambiguate the meaning of the two, the prior for $A_0$ has high probability
+range from about 40 to 100 years, whereas the prior distribution of $A_{cl}$ has
+peak probability at 1 and a relatively narrow width, due to the fact that it is
+a tissue specific multiplicative modifying factor. To understand the roles of
 these priors refer to (**SM3 Fig**).
 
-Using this model, we split the data by Male/Female sex, and by White _vs._
-Black race as these were the main well-identified categories available to us in
-the dataset. Sex is known to be a risk factor for death, with males dying at
-slightly higher rates for all ages of interest in our dataset. Black _vs._ White
-race is an important category due to differences in access to care and other
-socioeconomic and education related factors. However, due to the low number of
-Black patient tumors, we use ethnicity only to identify the baseline risk
-function, while the tissue- and cluster-specific multipliers are not ethnicity
-specific. There were too few tumors from other races to obtain good inference so
-we excluded them from our analysis.
+Using this model, we split the data by Male/Female sex. Sex is known to be a
+risk factor for death, with males dying at slightly higher rates for all ages of
+interest in our dataset.
 

### 2021-06-30 text/survival.Rmd

@@
 effective age rate (which modifies the baseline based on your actual age) as
-well as the ktis which accelerates or decelerates further aging, varies
+well as the $k_{tis}$ which accelerates or decelerates further aging, varies
 considerably from cancer to cancer. For example a 20 year old diagnosed with
-THCA or PRAD would be threated similarly to a 40 year old based on teh Effective
+THCA or PRAD would be treated similarly to a 40 year old based on the Effective
 Age parameter. It's noticeable that for every year beyond 20 in the PRAD
-diagnosis we add only an extra ~ 0.2 years (Effective Age Rate). The ktis
+diagnosis we add only an extra ~ 0.2 years (Effective Age Rate). The $k_{tis}$
 parameter represents an acceleration (greater than 1) or deceleration (less) of

### 2021-07-01 code/supplemental_v2.Rmd
Add supplemental fig5. Model comparison with other classification. ecdf and error distribution

@@
   library(kableExtra),
-  library(xlsx)
+  library(xlsx),
+  library(patchwork)
 ))
@@
 
-```{r supplemental-table-st1-data-generation, eval = FALSE}
-load("../data/gdc_tumor_coding_complete.Rdata")
-load("../data/gdc_reactome_path_analysis.Rda")
-load("../data/gene2_level2_lookup.Rda")
-load("../data/pathname2level2.Rda")
-load("../data/gene_data.Rda")
-load("../data/gdc_gene_exp_adj.Rda")
+```{r supplemental-s5-plot-data, eval = FALSE}
+## Calculate error distribution
 
-## Parameters ##
-n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
-mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
-show_perc <- TRUE ## show perc for mutated gene
+## break error into bins for entropy calculation
+## Set number of bins - 1024 bins
+binsize <- 1024 - 2 ## -2 bins for the less than -3650 and greater than 3650 values
+evaluebreak <- c(-1e10,seq(from = -3650.00, to = 3650.00, by = 3650*2/binsize), 1e10)
+elabel <- list()
+counter <- 2
+for(i in evaluebreak[c(-1,-(length(evaluebreak)-1))]) {
+  if(i == -1e10) {
+    elabel[[counter]] <- paste0(i, "+")
+  } else if(i == 1e10) {
+    elabel[[counter]] <- paste0(i, "+")
+  } else {
+    elabel[[counter]] <- paste0(i, "-", round(evaluebreak[[counter+1]], digits = 2))
+  }
+  counter <- counter + 1
+}
+elabel <- as.character(elabel)
+elabel[1] <- "-3650+"
+elabel[1024] <- "3650+"
 
-toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways
+load("../data/deadpat_all.rda")
+deadpat <- deadpat %>% rownames_to_column(var = "stanid")
+## convert negative tdeath to 0
+deadpat[which(deadpat$tevent < 0), "tevent"] <- 0
+model <- c("knn", "icluster", "coca", "tis")
+tdeath.error <- list() ## store error values
+entropy <- list() 
+for(n in model) {
+  ## Load correct model
+  if(n == "knn") {
+    load("./survival/survmodel_effcl_knn_062821.stansave")
+  } else if(n == "icluster") {
+    load("./survival/survmodel_effcl_icluster_062821.stansave")
+  } else if (n == "coca") {
+    load("./survival/survmodel_effcl_coca_062821.stansave")
+  } else if (n == "tis") {
+    load("./survival/survmodel_effcl_tis_062821.stansave")
+  } else {
+    print("Model does not exist")
+  }
+  ## Extract tdeath for "adeath" indivduals
+  tdeath <- samps %>% as.data.frame(.) %>% .[, grep(pattern = "tdeath", x = names(.))]
+  names(tdeath) <- gsub("tdeath\\[(.*)\\]","\\1", names(tdeath)) ## update colname with stanid
+  tdeath <- tdeath %>% gather(data = ., key = "stanid", value = "predicted") %>%
+    left_join(x = ., y = deadpat[,c("stanid", "tevent", "eventtype")], by = "stanid") %>%
+    dplyr::filter(eventtype == "adeath")
+  ## Calculate error for each sample
+  tdeath$error <- tdeath$predicted - tdeath$tevent
+  tdeath.error[[n]] <- tdeath$error
+  ## Split error data into bins and tally up 
+  tdeath <- tdeath %>% mutate(ebin = cut(x = error,
+                                         breaks = evaluebreak,
+                                         right = T,
+                                         labels = elabel))
+  eprob <- as.data.frame(table(tdeath$ebin))
+  eprob <- eprob %>% dplyr::filter(Freq > 0) %>%
+    mutate(prob = Freq / nrow(tdeath))
+  ## Entropy calculation
+  entropy[[n]] <- round(-sum(eprob$prob*log(eprob$prob))/log(2), digit = 3)
+}
+knn <- data.frame(error = tdeath.error$knn,
+                  model = "knn")
+icluster <- data.frame(error = tdeath.error$icluster,
+                       model = "icluster")
+coca <- data.frame(error = tdeath.error$coca,
+                   model = "coca")
+tis <- data.frame(error = tdeath.error$tis,
+                   model = "tis")
+tdeath.error <- rbind(knn, icluster, coca, tis); rm(knn, icluster, coca, tis); gc()
 
-id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id) %>%
-  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
-  left_join(., pathname2level2, by = "PATH_ID") %>%
-  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
-  dplyr::select(-PATH_ID) %>%
-  t() %>% as.data.frame()
+## empirical cumulative distribution function
+## Model selection
+model <- c("knn", "icluster", "coca", "tis")
 
-gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
-  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
-  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
-  dplyr::select(-ensembl_gene_id)
+for(n in model) {
+   ## Load correct model
+  if(n == "knn") {
+    load("./survival/survmodel_effcl_knn_062821.stansave")
+  } else if(n == "icluster") {
+    load("./survival/survmodel_effcl_icluster_062821.stansave")
+  } else if (n == "coca") {
+    load("./survival/survmodel_effcl_coca_062821.stansave")
+  } else if (n == "tis") {
+    load("./survival/survmodel_effcl_tis_062821.stansave")
+  }
+  else {
+    print("Model does not exist")
+  }
+  dfs <- as.data.frame(samps) %>% 
+    select(starts_with("tdeath[")) %>%
+    stack(.) %>%
+    'colnames<-' (c("tdeath", "stanid")) %>%
+    .[sample(x = nrow(.), size = 1e5),]
+  assign(x = n, value = dfs)
+}
 
-table_st1 <- list() ## list for forloop
-
-for(n in toppath) {
-  message(n)
-  ## finding top mutated gene in pathway
-  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
-    'rownames<-' (rownames(id_to_names)) %>% 
-    filter(. == 1) %>% ## filter for genes that are in that pathway
-    rownames(.)
-  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
-      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
-    rownames_to_column(var = "gene") %>%
-      mutate(n_sample = length(gene_mutation),
-             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
-       arrange(desc(obs_count))
-  top_gene_in_path <- list()
-  for(p in mutated_gene_position) {
-    message(p)
-    if(show_perc == TRUE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
-    }
-    if(show_perc == FALSE) {
-    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
-    }
-  }
-  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
-   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
-  ## calculating the percent mutated samples
-  spec_path <- data.frame(Pathway = n,
-                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
-  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
-}
-table_st1 <- as.data.frame(do.call(rbind, table_st1))
-save(table_st1, file = "../data/table_st1.Rda")  
-rm(list = ls(all.names = T))  
-gc()
+save(entropy, tdeath.error, knn, icluster, coca, tis, file = "../data/s5_plot.rda")
 ```
 
-```{r supplemental-table-st1, eval = FALSE}
-load("../data/table_st1.Rda")
-## for html only
-# table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable() %>%
-#   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
-#   column_spec(1, bold = T, color = "black")
+```{r supplemental-s5-plot, fig.width=9}
+load("../data/s5_plot.rda")
+load("../data/deadpat_all.rda")
 
-## for word doc
-# table_st1 %>% flextable() %>% 
-#   set_caption(caption = "") %>%
-#   align(part = "all") %>% 
-#   width(., width = c(4,0.5,0.5,0.5,0.5)) %>% 
-#   fontsize(size = 8, part = "body")
+s5.1 <- ggplot(knn) + stat_ecdf(aes(x=tdeath),color="#00BFC4", size = 1) + 
+  stat_ecdf(aes(x=tevent),data=subset(deadpat,eventtype=="adeath"),col="#00BA38", size = 1) +
+  stat_ecdf(aes(x=tdeath),data=icluster, color="#7CAE00", size = 1) + 
+  stat_ecdf(aes(x=tdeath),data=coca, color="#F8766D", size = 1) + 
+  stat_ecdf(aes(x=tdeath),data=tis, color="#F564E3", size = 1) + 
+  coord_cartesian(xlim=c(0,365*5)) + 
+  labs(x = "Survival Time (years past diagnosis)", y = "Fn(Survival Time)") +
+  annotate(geom = "text", x = 250, y = 0.5, label = "Actual Death", color = "#00BA38", size = 3) +
+  theme_minimal()
+
+s5.2 <- ggplot(data = tdeath.error, aes(x = error, group = model, fill = model)) +
+  geom_density(alpha = 0.5) +
+  theme_minimal() +
+  scale_x_continuous(breaks=seq(-4000, 4000, by = 2000), limits = c(-3650,3650)) +
+  labs(x = "Error (Predicted - Actual)", y = "Density")#, 
+       #subtitle = paste0("Entropy - knn: ", entropy$knn, " ; icluster: ", entropy$icluster, "\ncoca: ", entropy$coca, " ; tis: ", entropy$tis))
+
+layout = "
+AABB
+AABB
+"
+
+s5.1 + s5.2 + plot_annotation(tag_levels = "A") + plot_layout(design = layout)
 ```
 
-```{r st2, fig.width= 9, eval = FALSE}
-##**Supplementary Table 2. Cluster-specific enriched pathways (effect size >= 0.30).**
-## docx friendly
-load("../data/moderate_enrich_prob.Rda")
-
-table_st2 <- moderate_enrich_prob %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
-  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col
-
-table_st2$log10pvalue <- round(-log10((table_st2$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
-table_st2$Enrichment_Score <- round(table_st2$Enrichment_Score, digits = 2)
-table_st2$Signif <- NULL ## we don't need this anymore
-
-# st1 <- st1 %>% flextable() %>% 
-#   set_caption(caption = "") %>%
-#   align(part = "all") %>% 
-#   width(., width = c(4,0.75,1,1)) %>% 
-#   fontsize(size = 10, part = "body")
-#
-
-# names(st1) <- c("Pathway", "Cluster", "Effect Size", "-log(p)")
-#  st1 %>%  knitr::kable(format = "latex", longtable = TRUE, align = c("l", rep("c",3))) %>%
-#    kable_styling(latex_options=c("scale_down", "repeat_header"), full_width = FALSE, font_size = 12) %>%
-#    column_spec(1, color = "black", width = "4in")
-
-## Saving st1 and st2 in excel
-
-write.xlsx(table_st1, file="supplemental.table.xlsx", sheetName="ST1", row.names=FALSE)
-write.xlsx(table_st2, file="supplemental.table.xlsx", sheetName="ST2", append=TRUE, row.names=FALSE)
-```
+**S5 Fig.
+Performance comparison of different classification model - knn, COCA, iCluster, tissue-of-origin (tis).** A) Empirical cumulative distribution function comparing time of death prediction of all classification models with actual time of death (dark green).
+B) Error distribution plot of predicted and actual time of death.

### 2021-07-01 text/discussion.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

@@
 were exclusive to a single cluster. Only cancer types with the fewest samples
-were found to be absent from one or more clusters at all.
+were found to be absent from one or more clusters at all. Additionally, we annotated 
+the clusters with patient attribute values (_e.g._ age of diagnosis, gender, and race) 
+and were unable to identify any trends.
 
@@
 fact that they contribute to many cluster-specific pathways. This can only be
-explained by less common genes complementing the unique combination of driver
-genes in each tumor, and we speculate that many of these less common genes are
-sensitive to increased mutation frequency in different tissues.
+explained by less commonly mutated genes complementing the unique combination of driver
+genes in each tumor, and we speculate that many of these less commonly mutated genes are
+sensitive to increased mutation frequency in different tissues. For example, the most 
+frequently mutated genes (_TP53_, _PIK3CA_ and _MUC16_) in cluster 2 were also frequently 
+mutated in several other clusters. Yet, pathway enrichment in cluster 2 was defined 
+by less commonly mutated genes in the proteasomal degradation pathway (_e.g. PSMA3_, 
+_PSMA1_ or _PSME4_). The ubiquitin proteasome system maintains cellular protein 
+homeostasis by regulating protein turnover. Pathway mutations found in cluster 2 
+(_e.g._ “B-catenin degradation”, “cellular response to hypoxia”, “regulation of cell cycle” 
+and _etc_) covered several aspects of the cancer hallmarks described by Hanahan 
+and Weinberg. These hallmarks included “sustaining proliferative signaling”, “evading 
+tumor suppressors” and “resisting cell death” [@Bhattacharjee2014]. Cluster 
+3 was defined by less commonly mutated genes related to ribosomal subunit. Pathway 
+enrichment for cluster 3 included dysregulation in “rRNA processing in the nucleus 
+and cytosol” and consequently, pathways involved in eukaryotic translation and 
+nonsense mediated decay. Mutations in the ribosomal subunit have been associated 
+with selective translation of mRNA to upregulate proliferation [@kampen2019]. 
+Pathway enrichment in cluster 3 consist of hallmarks relating to “sustaining proliferative 
+signaling”, “evading tumor suppressors”, “resisting cell death”, “deregulating 
+cellular energetics”, and “genome instability and mutations”. Cluster 4 was defined 
+by less commonly mutated genes related to nucleosome structure that resulted in 
+pathway enrichment in DNA methylation and DNA double-strand break repair pathway. 
+Cluster 4 was also enriched for “telomere maintenance”, a mechanism contributing to 
+cancer progression by counteracting telomere shortening. Mutations in the 
+nucleosome as found in cluster 4 have been proposed to hit all attributes of hallmarks of cancer. Cluster 
+7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks: 
+“sustaining proliferative signaling”, “evading tumor suppressors”, “resisting cell 
+death”, and “inducing angiogenesis”. These clusters were strongly defined by 
+mutations in _PI3K_ or _RAS_, respectively. Taken together, our clusters highlight 
+the similarities and differences at a gene and pathway level between and within 
+different cancer types – uncovering the molecular mechanism in cancer. In the future,
+we hope to explore the defining features of the clusters by connecting the pathway 
+mutations to cancer hallmarks.
 

### 2021-07-01 text/introduction.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

@@
 @jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
-genes mutated at a greater rate than expected due to chance. It has been
+genes mutated at a greater rate than expected due to chance. Dees et al. [@deesMuSiC2012]
+developed a pipline to separate driver mutations from passenger mutations. 
+Tamborero et al. [@tamboreroOncodriveCLUST2013] expanded on this idea to identify 
+tumor mutation biases towards specific regions of a protein sequence. Lawrence et 
+al. [@lawrenceDiscoverySaturationAnalysis2014] focused on identifying cancer driver 
+genes with intermediate mutation frequencies using tumor-normal pairs. It has been
 estimated that fewer than five mutations in key oncogenes and/or tumor
@@
 More recently, pan-cancer mutations and other data have been looked at as
-disruptions of normal pathway activity. Sanchez-Vega et al. [@PCAWG-PATH] carried out exhaustive
-analysis of pathway enrichment using whole genome data. Network approaches
+disruptions of normal pathway activity. Pathway-centric analysis of tumors
+provides additional benefits over gene-centric analysis: (1) It is significantly 
+less noisy because it aggregates molecular events across multiple genes in the 
+same pathway and (2) the identification of a potential casual mechanism is easier 
+to interpret as genetic aberrations are linked to molecular pathways. In a perspective, 
+Creixell et al. [@Creixell-pathwaynetworkanalysis] described approaches of pathway 
+and network analysis applicable to next-generation sequencing data. The most common 
+approach being fixed-gene enrichment analysis to identify over-represented pathways. 
+Sanchez-Vega et al. [@PCAWG-PATH] carried out an exhaustive
+analysis of pathway enrichment using whole genome data. Focusing on ten frequently 
+altered canonical signaling pathways in cancer, they explored the mutation frequencies 
+of coding regions of genes in these pathways for each cancer type. Network approaches
 address the sparseness of mutations, allowing genes to be influenced by

### 2021-07-01 text/manuscript.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

@@
 **Fig 5. Comparison of different cancers, simulated and actual results.**
-Comparison of survival in four simulated cohorts. A) A group consisting of
-randomly selected 30 year old women, unaffected (light blue) _vs._ cancer, 
-($k = 2$, light orange), another group consisting of random 30-70 year-old
-women, unaffecteds (dark blue) and cancer $k = 2$ (dark orange). B,C,D,E)
-Compare random group with correct age distribution for the given cancer
-(turqoise) to the model predicted survival for the given average k (blue) and
-the actual survival for the patients in our dataset (orange). Differences with
-actual data are likely due to a mixture of cluster identities which we exclude
-from these simulations for simplicity. F) Effects are compared between cancer
-type by showing the probability density for the posterior value of the $k_{tis}$
-multiplier. Larger values correspond to decreased age-independent life
-expectancy.
+Comparison of survival in six simulated cohorts. A) A group consisting of
+randomly selected 30 year old women, unaffected (light blue) _vs._ cancer 
+using the typical values from the posterior samples for a BRCA patient (yellow) 
+and an artificial 10 years added to the effective age for a BRCA patient 
+(dark blue), another group consisting of uniformly selected 30-70 year-old
+women at random, unaffected (green) and cancer without (pink) and with (dark orange) 
+the artificial 10 years. B) Posterior probability density of a 20 year old 
+accelerated aging by cancer type (effective age). Red dashed line represents
+age for unaffected individual, C) Posterior probability density of effective 
+age rate for every additional year past 20 year old by cancer type. Red 
+dashed line represents effective age rate for unaffected individual, D) 
+Posterior probability density of the $k_{tis}$ multiplier. Values greater 
+than 1 correspond to decreased age-independent life expectancy. Red dashed 
+line represents $k_{tis}$ for unaffected individual. E,F,G,H) Comparison 
+of our model to the actual survival for the patients in our dataset. We 
+compare random group with correct age distribution for the given cancer 
+(turqoise) to the model predicted survival for the given average effective 
+age (blue) and the actual survival for the patients in our dataset (orange). 
+Differences with actual data are likely due to a mixture of cluster 
+identities which we exclude from these simulations for simplicity.
 
-**Fig 6. Cluster specific $k_{cl}$ values for each cancer type.** The overall
-$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific $k$
-represents relative aggressiveness of each cluster within cancer type. Alpha
-transparency (α) is set to reflect varying confidence in the posterior
+**Fig 6. Cluster specific $A_{cl}$ values for each cancer type.** The overall
+effective age value is the product of $A_{0} x A_{cl} + A_{r} x (A^1 – 20)$. 
+The cluster-specific A represents relative aggressiveness of each cluster within 
+cancer type. Alpha transparency (α) is set to reflect varying confidence in the posterior
 distribution when the number of observations is N<10, α = 0.1; N<25, α = 0.5; N

### 2021-07-01 text/survival.Rmd
Update text in intro, discussion, and figure legend. Add reference to ref.bib

@@
 
-To illustrate the model, we simulate survival for four related groups (**Fig
+To illustrate the model, we simulate survival for six related groups (**Fig
 5A**). We start with a cohort of women aged 30 randomly selected from the US
@@
 varies by effective age we add an additional artificial 10 years to the BRCA
-effective age (dark blue). Compare these results to a randomly selected group of
-women ranging from 30 to 70 years old (green). Immediately, survival changes due
+effective age (dark blue). Compare these results to a uniformly selected group of
+women ranging from 30 to 70 years old at random (green). Immediately, survival changes due
 to the mixture of ages, without malignancy. By computing the effective ages for

### 2021-07-02 text/introduction.Rmd
Minor edits to introduction. Add section in results regarding S5 Fig

@@
 genes mutated at a greater rate than expected due to chance. Dees et al. [@deesMuSiC2012]
-developed a pipline to separate driver mutations from passenger mutations. 
+developed a pipeline to separate driver mutations from passenger mutations. 
 Tamborero et al. [@tamboreroOncodriveCLUST2013] expanded on this idea to identify 
@@
 independent of tissue of origin. Understanding cancer as a disease of the cell
-has long been a goal of the field as characterized in essays by
+has long been a goal of the field as characterized in essays by Hanahan and Weinberg
 [@weinberg2000; @hallmarks_II]. The first of these studies in genomics identified

### 2021-07-02 text/manuscript.Rmd
Minor edits to introduction. Add section in results regarding S5 Fig

@@
 women at random, unaffected (green) and cancer without (pink) and with (dark orange) 
-the artificial 10 years. B) Posterior probability density of a 20 year old 
-accelerated aging by cancer type (effective age). Red dashed line represents
+the artificial 10 years. B) Posterior probability density of the effective age 
+baseline ($A_{0}$) by cancer type. Red dashed line represents
 age for unaffected individual, C) Posterior probability density of effective 
-age rate for every additional year past 20 year old by cancer type. Red 
+age rate ($A_{r}$) by cancer type. Red 
 dashed line represents effective age rate for unaffected individual, D) 

### 2021-07-02 text/survival.Rmd
Minor edits to introduction. Add section in results regarding S5 Fig

@@
 tissues.
+
+To determine the significance of our pathway-based clustering, we evaluated the 
+performance in predicting time to death (tdeath) with other published classification 
+models. We downloaded cluster membership form two studies conducted within the TCGA 
+consortium [@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018]. 
+We generated tdeath predictions from uniform random p, by inverting the cumulative 
+distribution function of our death probability function. Using the actual tdeath 
+from TCGA as the reference, we show that our pathway-based clustering (referred to as 
+kNN) had better performance compared to COCA [@hoadleyMultiplatformAnalysis2014], 
+iCluster [@hoadleyCellofOriginPatternsDominate2018] and tissue-of-origin (tis) - 
+although not statistically significant (**S5A Fig**). Additionally, we computed the
+tdeath error using matched sample tdeath from the prediction model and actual (**S5B Fig**). 
+Again, all models had similar performance with kNN performing slightly better as 
+indicated by a smaller error entropy.

### 2021-07-02 code/figure6.Rmd
Update stanfit for knn in figure 5/6. Update survmodel lower limit for ktis to 0.05

@@
 ## Processing data
-load("../data/survival/survmodel_effcl_knn_062521.stansave")
+load("./survival/survmodel_effcl_knn_070121.stansave")
 dfsamps <- as.data.frame(samps)

### 2021-07-02 code/survival/figure5.Rmd
Update stanfit for knn in figure 5/6. Update survmodel lower limit for ktis to 0.05

@@
 
-load("./survmodel_effcl_knn_062821.stansave")
+load("./survmodel_effcl_knn_070121.stansave")
 cancerdat = read_csv("../../data/survival/clinical_plus_cluster.csv");

### 2021-07-02 text/survival.Rmd

--- 2021-07-01 text/survival.Rmd
@@
 tissues.
+
+To compare our pathway-based clustering to other published methods, we evaluated
+the performance in predicting time to death (tdeath) with other published
+classification models. We downloaded cluster membership from two studies
+conducted within the TCGA consortium [@hoadleyMultiplatformAnalysis2014;
+@hoadleyCellofOriginPatternsDominate2018].  We generated tdeath predictions from
+uniform random p, by inverting the cumulative distribution function of our death
+probability function. Using the actual tdeath from TCGA as the reference, we
+show that our pathway-based clustering (referred to as kNN) had slightly better
+performance compared to COCA [@hoadleyMultiplatformAnalysis2014], iCluster
+[@hoadleyCellofOriginPatternsDominate2018] and tissue-of-origin (tis). (**S5A
+Fig**). Additionally, we computed the entropy of the error distribution in
+predictions (**S5B Fig**).  All models had similar performance with kNN
+performing slightly better as indicated by a slightly smaller error entropy.

### 2021-07-02 code/supplemental_v2.Rmd
Update stanfit for all models in s5 fig

@@
   if(n == "knn") {
-    load("./survival/survmodel_effcl_knn_062821.stansave")
+    load("./survival/survmodel_effcl_knn_070121.stansave")
   } else if(n == "icluster") {
-    load("./survival/survmodel_effcl_icluster_062821.stansave")
+    load("./survival/survmodel_effcl_icluster_070121.stansave")
   } else if (n == "coca") {
-    load("./survival/survmodel_effcl_coca_062821.stansave")
+    load("./survival/survmodel_effcl_coca_070121.stansave")
   } else if (n == "tis") {
-    load("./survival/survmodel_effcl_tis_062821.stansave")
+    load("./survival/survmodel_effcl_tis_070121.stansave")
   } else {
@@
   if(n == "knn") {
-    load("./survival/survmodel_effcl_knn_062821.stansave")
+    load("./survival/survmodel_effcl_knn_070121.stansave")
   } else if(n == "icluster") {
-    load("./survival/survmodel_effcl_icluster_062821.stansave")
+    load("./survival/survmodel_effcl_icluster_070121.stansave")
   } else if (n == "coca") {
-    load("./survival/survmodel_effcl_coca_062821.stansave")
+    load("./survival/survmodel_effcl_coca_070121.stansave")
   } else if (n == "tis") {
-    load("./survival/survmodel_effcl_tis_062821.stansave")
+    load("./survival/survmodel_effcl_tis_070121.stansave")
   }
@@
   theme_minimal() +
-  scale_x_continuous(breaks=seq(-4000, 4000, by = 2000), limits = c(-3650,3650)) +
+  #scale_x_continuous(breaks=seq(-4000, 4000, by = 2000), limits = c(-3650,3650)) +
   labs(x = "Error (Predicted - Actual)", y = "Density")#, 

### 2021-07-02 text/manuscript.Rmd
Update limitation of the study with model comparison result

@@
 
+The outcome of our model comparison with other published classification models 
+in predicting tdeath showed kNN performing slightly better. However, for the 
+purpose of predicting lifetime, all models are practically equivalent. One 
+explanation for this is the cluster identity does not provide much new information 
+from tissue-of-origin. Most of the cluster-specific rate ($A_{cl}$) peaked around 1 for 
+all models – leaving the remaining variables ($A_{0}$, $A_{r}$, and $k_{tis}$), which are 
+tissue-specific, to compute tdeath. While our study focused on somatic mutations, 
+the incorporation of other types of omics data may help differentiate the classes 
+within cancer type. Additionally, more work in solidifying the pathway model into 
+cancer hallmarks could provide biologically meaningful distinction of the clusters.
+
 ## Resource Availability

### 2021-07-02 text/manuscript.Rmd
add some stuff about comparisons between models

@@
 in predicting tdeath showed kNN performing slightly better. However, for the
-purpose of predicting lifetime, all models are practically equivalent. One 
-explanation for this is the cluster identity does not provide much new information 
-from tissue-of-origin. Most of the cluster-specific rate ($A_{cl}$) peaked around 1 for 
-all models – leaving the remaining variables ($A_{0}$, $A_{r}$, and $k_{tis}$), which are 
-tissue-specific, to compute tdeath. While our study focused on somatic mutations, 
-the incorporation of other types of omics data may help differentiate the classes 
-within cancer type. Additionally, more work in solidifying the pathway model into 
-cancer hallmarks could provide biologically meaningful distinction of the clusters.
+purpose of predicting lifetime within this dataset, all models are practically
+equivalent. One possible explanation for this is the cluster identity does not
+provide much new information from tissue-of-origin. However another possibility
+is that because treatment is not specialized to cluster, the treatments are not
+taking advantage of potential differences between clusters which might
+dramatically change survival. Most of the cluster-specific rate ($A_{cl}$)
+peaked around 1 for all models – leaving the remaining variables ($A_{0}$,
+$A_{r}$, and $k_{tis}$), which are tissue-specific, to compute tdeath. While our
+study focused on somatic mutations, the incorporation of other types of omics
+data may help differentiate the classes within cancer type. Additionally, more
+work in solidifying the pathway model into cancer hallmarks could provide
+biologically meaningful distinction of the clusters. Finally, if treatments can
+be specialized to biological information encoded in cluster, perhaps this could
+drive dramatic improvements in survival among some clusters in the future.
 

### 2021-07-02 text/introduction.Rmd
Minor edits to text

@@
 Creixell et al. [@Creixell-pathwaynetworkanalysis] described approaches of pathway 
-and network analysis applicable to next-generation sequencing data. The most common 
+and network analysis applicable to next-generation sequencing data, the most common 
 approach being fixed-gene enrichment analysis to identify over-represented pathways. 

### 2021-07-02 text/survival.Rmd
Minor edits to text

@@
 parameter represents an acceleration (greater than 1) or deceleration (less) of
-the further aging beyond diagnosis. For example for BRCA patients each year past
+the further aging beyond diagnosis. For example, for BRCA patients each year past
 diagnosis adds about 1 year of risk, whereas for LGG patients it adds around 2,

### 2021-07-04 text/discussion.Rmd
Update text for cluster 7 and 8 in discussion

@@
 nucleosome as found in cluster 4 have been proposed to hit all attributes of hallmarks of cancer. Cluster 
-7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks: 
+7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks, 
+showing that some clusters share common driver mechanisms: 
 “sustaining proliferative signaling”, “evading tumor suppressors”, “resisting cell 
 death”, and “inducing angiogenesis”. These clusters were strongly defined by 
-mutations in _PI3K_ or _RAS_, respectively. Taken together, our clusters highlight 
+mutations in _PI3K_ or _RAS_, respectively, suggesting a shared etiology in the EGFR 
+pathway for which these genes represent alternative intracellular signaling mechanisms.Taken together, our clusters highlight 
 the similarities and differences at a gene and pathway level between and within 

### 2021-07-07 code/supplemental_v2.Rmd
Minor edits to manuscript. Formatting to iScience

@@
 
-**S1 Fig. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
+**Figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
 individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Fig 1B of the main text**. The colors indicate the tumor's cluster identity.
+in **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
 
@@
 
-**S2 Fig. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Fig 1B of the main text**. The colors indicate the tumor's histological subtype identity.
+**Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
+cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's histological subtype identity.
 
@@
 
-**S3 Fig. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
-genes involved in genome stability projected on the UMAP from **Fig 1B of the main text**. The colors indicate the tumor's cluster identity.
+**Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
+genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
 
@@
 
-**S4 Fig.
-Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Fig 1B of the main text**.
+**Figure S4.
+Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 1B of the main text**.
 The colors indicate the tumor's cluster identity.
@@
 
-**S5 Fig.
+**Figure S5.
 Performance comparison of different classification model - knn, COCA, iCluster, tissue-of-origin (tis).** A) Empirical cumulative distribution function comparing time of death prediction of all classification models with actual time of death (dark green).

### 2021-07-07 text/manuscript.Rmd
Minor edits to text
Minor edits to manuscript. Formatting to iScience

@@
 bibliography: references.bib
-csl: plos-computational-biology.csl
+csl: iscience.csl
 output:
@@
 splicing mutations. To minimize bias from well-studied diseases and processes,
-we selected 377 Reactome pathways (reactome.org) (see S1 Table) of interest 
+we selected 377 Reactome pathways (reactome.org) (see **Table S1**) of interest 
 corresponding to basic cellular processes and biochemical
@@
 assigning a Boolean value of 1 to each pathway with one or more genes mutated
-and 0 for all others (**Fig 1A**).
+and 0 for all others (**Figure 1A**).
 
 We investigated this dataset using multiple correspondence analysis (MCA)
-[@JSSv025i01], and visually summarized the analysis with UMAP (**Fig 1B** and
-see interactive media from **S1 File** (junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
+[@JSSv025i01], and visually summarized the analysis with UMAP (**Figure 1B** and
+see interactive media from **File S1** (junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
 used the resulting graph coordinates to perform density based clustering
@@
 cluster, and all well-represented cancer types had tumors in every cluster (see
-**Fig 2A** and full tumor profiles in **S1 Fig**,
-interactive media **S1 File**), suggesting that these clusters
+**Figure 2A** and full tumor profiles in **Figure S1**,
+interactive media **File S1**), suggesting that these clusters
 indicate molecular pathology largely independent of tissue-of-origin. As an
 example of one cancer that does have a biased pathway profile, pancreatic
-adenocarcinoma (PAAD) was predominantly found in cluster 8 (**Fig 2A** and
-**S1 Fig** and **S1 File**), but even PAAD
+adenocarcinoma (PAAD) was predominantly found in cluster 8 (**Figure 2A** and
+**Figure S1** and **File S1**), but even PAAD
 comprises tumors from the nine remaining classes. Thus, patients with these
@@
 standards of care. We did not observe exclusive segregation by cluster for these
-subtypes (**Fig 2B**). We also projected histological subtype data for the
-remaining cancers (see **S2 Fig** and see interactive media from 
-**S1 File**); we find that
+subtypes (**Figure 2B**). We also projected histological subtype data for the
+remaining cancers (see **Figure S2** and see interactive media from 
+**File S1**); we find that
 the subtypes, though often biased to one or more clusters, are almost never
@@
 mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe
-segregation of these mutations into specific clusters (**Fig 2C** and see
-interactive media from **S1 File**). We also projected
-mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_;
+segregation of these mutations into specific clusters (**Figure 2C** and see
+interactive media from **File S1**). We also projected
+mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_,
 _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except for _APC_
 exhibited remarkable specificity with respect to cluster assignment
-(**S3 Fig**). To look at other risk factors such as maintenance
+(**Figure S3**). To look at other risk factors such as maintenance
 of DNA methylation levels and telomere length, we projected somatic mutations of
 the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and observed
-similar lack of cluster bias (**S3 Fig**).
+similar lack of cluster bias (**Figure S3**).
 
@@
 mutations and culminates in a cluster (class 10) which has nearly every pathway
-disrupted (**Fig 2F**). However, we don't observe a trend in the overall
+disrupted (**Figure 2F**). However, we don't observe a trend in the overall
 mutation burden across the backbone of the UMAP. Nonetheless, to test the
@@
 origin and other categories of tumor, we do not observe any bias among the
-stages to specific clusters (**Fig 2D**), suggesting that stage does not
+stages to specific clusters (**Figure 2D**), suggesting that stage does not
 contribute to cluster identity.
@@
 metastases onto our UMAP, and surprisingly we observed even distribution of the
-samples across clusters (**Fig 2E**). This final observation suggests that
+samples across clusters (**Figure 2E**). This final observation suggests that
 our pathway-disruption clustering is dependent on particular combinations of
@@
 In order to identify pathway enrichment across all cancers, we created a list of
-pathway disruptions with percent mutated samples and top genes (**S1 Table**). 
+pathway disruptions with percent mutated samples and top genes (**Table S1**). 
 As expected, these analyses reveal the broad importance of many
@@
 pathways from highest to lowest enrichment. We visualized enrichment as a
-heatmap (**Fig 3A**). Using this approach, we identified about fourteen
+heatmap (**Figure 3A**). Using this approach, we identified about fourteen
 pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see methods)
-(**S2 Table**). Clusters 7, 8, and 9 had several pathways in
+(**Table S2**). Clusters 7, 8, and 9 had several pathways in
 common. To explore the specific pathways marking each cluster, we projected
-disruptions for each of the 377 pathways onto the UMAP (**supplementary table
-ST2** and see interactive media from **supplemental file
-1**). Clusters 3 and 5
+disruptions for each of the 377 pathways onto the UMAP (**Table S2** 
+and see interactive media from **File S1**). Clusters 3 and 5
 were distinguished by metabolic pathways including RNA and protein biosynthesis
-(**S2 Table**). Similarly, cluster 4 was distinguished by
+(**Table S2**). Similarly, cluster 4 was distinguished by
 mutations affecting regulation of DNA and histone methylation ("DNA
@@
 Clusters 7-9 have in common mutations in extracellular, intracellular, and
-immune-related signaling pathways (see **Fig 3B** and **S2 Table**). 
+immune-related signaling pathways (see **Figure 3B** and **Table S2**). 
 Cluster 2 had the highest pathway enrichment levels of the three, having
@@
 differentially mutated genes. We ranked odds ratios and selected the top ten
-enriched and depleted genes (pvalue < 0.01) for each cluster (**Fig 4**; odds
+enriched and depleted genes (pvalue < 0.01) for each cluster (**Figure 4**; odds
 ratios plot). Clusters 7 and 8, which shared multiple enrichment in signaling
 pathways, are largely driven by mutations in _PI3K_ and its orthologs and _Ras_
-genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **S4 Fig** and see 
-interactive media from **S1 File**. Interestingly,
+genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Figure S4** and see 
+interactive media from **File S1**. Interestingly,
 cluster 9, which also shared multiple enrichment in signaling pathways with
@@
 8). Next, we investigated the proportion of samples per cancer type for the
-significant genes within a cluster (**Fig 4**; heatmap). Surprisingly,
+significant genes within a cluster (**Figure 4**; heatmap). Surprisingly,
 clusters were not predominated by one or more highly mutated genes across all
@@
 clusters. For example, “Signaling by _PTK6_” is enriched in non-metastatic
-samples of cluster 8 (**S2 Table**) but not in 7 and 9. This
+samples of cluster 8 (**Table S2**) but not in 7 and 9. This
 pathway is enriched in metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$,
 **Table 1**). This is also true of “Erythropoietin activates _RAS_”, which is
-enriched in non-metastatic tumors of cluster 8 (**S2 Table**) and
+enriched in non-metastatic tumors of cluster 8 (**Table S2**) and
 also in metastatic tumors of clusters 7 and 9. Cluster 4 metastases were
@@
 
-## Resource Availability
-
-Further information and requests for resources should be directed to and will
-be fulfilled by the Lead Contact, Dennis Hazelett (Dennis.Hazelett@cshs.org). Data
-generated in this study have been deposited to Zenodo (https://doi.org/10.5281/zenodo.4011291).
-All code for producing the analyses and figures herein are included in this fully reproducible 
-manuscript in R markdown format. R markdown files and all other models are available from our 
-repositories (https://github.com/dennishazelett/TTmanu) on the distributed
-version control site, GitHub. An interactive media of the results is available on https://junkdnalab.shinyapps.io/PANCAN_supplemental/.
-
-## Methods
-
-Extended descriptions of clustering, models,
-formulae, pathway selection, and gene filtering are provided in supplementary
-methods.
+## Author Contributions
+P.N., S.G. and D.L. conducted the experiments. D.H. designed the experiments and wrote the paper.
 
@@
 
-## Author Contributions
-P.N., S.G. and D.L. conducted the experiments. D.H. designed the experiments and wrote the paper.
-
 ## Declaration of Interests
@@
 
-## References
-
-<div id="refs"></div>
-
 ## Figure and Table Legend
-**Fig 1. Clustering pathways of tumor samples.** A) Each of 377 selected
+**Figure 1. Clustering pathways of tumor samples.** A) Each of 377 selected
 Reactome pathways (rows) is classified as disrupted if one or more genes is
@@
 
-**Fig 2. Pathway-based clustering independent of tissue-of-origin** 
+**Figure 2. Pathway-based clustering independent of tissue-of-origin** 
 A) Sankey plot of correspondence between cancer type and cluster identity.
@@
 
-**Fig 3. Pan-cancer enrichment of pathway disruptions.** A) Heatmap shows
+**Figure 3. Pan-cancer enrichment of pathway disruptions.** A) Heatmap shows
 relative enrichment of each pathway (rows) within numbered cluster (columns).
@@
 
-**Fig 4. Gene level analysis reveals tissue-specific class signatures.** 
+**Figure 4. Gene level analysis reveals tissue-specific class signatures.** 
 Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
@@
 
-**Fig 5. Comparison of different cancers, simulated and actual results.**
+**Figure 5. Comparison of different cancers, simulated and actual results.**
 Comparison of survival in six simulated cohorts. A) A group consisting of
@@
 
-**Fig 6. Cluster specific $A_{cl}$ values for each cancer type.** The overall
+**Figure 6. Cluster specific $A_{cl}$ values for each cancer type.** The overall
 effective age value is the product of $A_{0} x A_{cl} + A_{r} x (A^1 – 20)$. 
@@
 **Table 1: Cluster-specific enriched pathways (effect size ≥ 0.30) in metastasis.**
+
+## Resource Availability
+
+Further information and requests for resources should be directed to and will
+be fulfilled by the Lead Contact, Dennis Hazelett (Dennis.Hazelett@cshs.org). Data
+generated in this study have been deposited to Zenodo (https://doi.org/10.5281/zenodo.4011291).
+All code for producing the analyses and figures herein are included in this fully reproducible 
+manuscript in R markdown format. R markdown files and all other models are available from our 
+repositories (https://github.com/dennishazelett/TTmanu) on the distributed
+version control site, GitHub. An interactive media of the results is available on https://junkdnalab.shinyapps.io/PANCAN_supplemental/.
+
+## Methods
+
+Extended descriptions of clustering, models,
+formulae, pathway selection, and gene filtering are provided in supplementary
+methods.
+
+## References
+
+<div id="refs"></div>
 

### 2021-07-20 code/figure6.Rmd
Edit manuscript based on editor comments

@@
 		facet_wrap(. ~ facetname,nrow=3) + theme_minimal() + scale_fill_brewer(palette = "Paired") +
-	  theme(legend.position = "none") + ggtitle("Class Specific Coefficients") +
+	  theme(legend.position = "none") +
 	  labs(x = expression(paste("A"["cl"])), y = "Cluster Identity") +

### 2021-07-20 code/supplemental_v2.Rmd
Edit manuscript based on editor comments

@@
 ---
-title: "Supplementary figures"
 output:
@@
 
-**Figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
+**Figure S1. Distribution of cancer type on UMAP clusters, Related to Figure 2.** Each panel shows the identity of 
 individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
@@
 
-**Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
+**Figure S2. Histological subtypes, Related to Figure 2.** Each panel shows the histological subtypes of one of the 23 
 cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's histological subtype identity.
@@
 
-**Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
+**Figure S3. Distribution of mismatch repair genes on UMAP clusters, Related to Figure 2.** Each panel shows one of 15 
 genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
@@
 **Figure S4.
-Distribution of oncogenes/tumor suppressors on UMAP clusters.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 1B of the main text**.
+Distribution of oncogenes/tumor suppressors on UMAP clusters, Related to Figure 2.** Each panel shows shows one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from **Figure 1B of the main text**.
 The colors indicate the tumor's cluster identity.
@@
 **Figure S5.
-Performance comparison of different classification model - knn, COCA, iCluster, tissue-of-origin (tis).** A) Empirical cumulative distribution function comparing time of death prediction of all classification models with actual time of death (dark green).
+Performance comparison of different classification model - knn, COCA, iCluster, tissue-of-origin (tis), Related to Figure 5.** A) Empirical cumulative distribution function comparing time of death prediction of all classification models with actual time of death (dark green).
 B) Error distribution plot of predicted and actual time of death.
+
+\newpage
+
+```{r child='../code/figureSM1.Rmd'}
+```
+
+**Figure S6. Method for tumor classification using pathway aberration data, Related to Figure 1.** 
+A) MCA analysis was performed to determine the eigenvalue with the most
+explanatory power from the categorical data. Average of 100 permutations of the
+data was used as a baseline. B) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
+projected down to 3 dimensions for the purpose of density based clustering (only
+the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
+used to perform hierarchical density based clustering tuned to minimize the
+fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
+the UMAP projection; tumors are colored by cluster membership (see legend). This
+results in some tumors remaining unclassified (grey points). F) kNN was used to
+assign membership to the remaining tumors.
+
+\newpage
+
+```{r child='../code/figureSM2.Rmd'}
+```
+
+**Figure S7. Comparison between CDC data and ODE model fit for
+males and females of European descent, Related to Figure 5.** This comparison does not distinguish on
+ethnicity for simplicity in visualization.
+
+\newpage
+
+```{r sm3, fig.height=3, fig.width=6}
+sm3.1 <- ggplot(data.frame(x=0:110),aes(x)) + 
+  stat_function(fun=function(x) dgamma(x,30.0,29.0/65),geom="area", fill = alpha("black", 0.5)) +
+  theme_classic() + 
+  scale_x_continuous(limits = c(30, 110), breaks = c(seq(30,110, by =10))) +
+  theme(plot.title = element_text(size = 12, hjust = 0.5),
+        panel.grid.major = element_line(size = 0.75),
+        panel.grid.minor = element_line()) +
+  labs(x= expression(paste("Prior for A"["0"])), y = "Probability Density")
+
+sm3.2 <- ggplot(data.frame(x=0:3),aes(x)) + 
+  stat_function(fun=function(x) dgamma(x,50.0,49.0/1.0),geom="area", fill = alpha("black", 0.5)) +
+  theme_classic() + 
+  scale_x_continuous(limits = c(0.5, 1.6), breaks = c(seq(0.6, 1.5, by = 0.2))) +
+  theme(plot.title = element_text(size = 12, hjust = 0.5),
+        panel.grid.major = element_line(size = 0.75),
+        panel.grid.minor = element_line()) +
+  labs(x= expression(paste("Prior for A"["cl"])), y = "Probability Density")
+
+layout = "
+AABB
+AABB
+"
+
+sm3.1 + sm3.2 + plot_annotation(tag_levels = "A") + plot_layout(design = layout)
+```
+
+**Figure S8. Priors for the tissue and class coefficients, Related to Figure 5 and 6.** A) The tissue
+coefficient is chosen to admit ages between 40 and 100 years old as a baseline
+age. B) Within a given tissue, the class parameter multiplies the tissue value and
+is given a peak density at 1, and a credible range from near 0.6 to 1.4,
+reflecting our expectation of a moderate multiplicative perturbation of
+variation across the classes.

### 2021-07-20 code/survival/effage_stanmodel.Rmd
Edit manuscript based on editor comments

```{r load-libraries}
library(sqldf)
library(tidyverse)
library(rstan)
library(readxl)
library(ggkm)
library(gridExtra)
library(PerformanceAnalytics)
library(patchwork)
```

```{r add-other-classes}
setwd("D:/repo/TTmanu/data/survival/")

cancerdat = read_csv("clinical_plus_cluster.csv");
treatdays = read_csv("tcga_clinical_data.csv")

#####################################
## Other TCGA paper classification ##
#####################################
## Hoadley et al. 2018 PMID: 29625048
icluster <- read_excel(path = "NIHMS957746-supplement-6.xlsx", skip=1)
icluster <- icluster %>% 'colnames<-' (c("sample_id", "icluster"))
## Hoadley et al. 2015 PMID: 25109877
coca <- read_excel(path = "mmc2.xlsx", skip=1)
coca <- coca %>% dplyr::select(Sample, COCA) %>% 'colnames<-' (c("sample_id", "coca"))

## Append additional classification model to cancer data
cancerdat <- cancerdat %>% mutate(sample_id = substring(text = sample_id, first = 1, last = 12)) %>% 
  left_join(x = ., y = icluster, by = "sample_id") %>% 
  left_join(x = ., y = coca, by = "sample_id")

## Assign unclassified tumor samples in icluster and coca to the next cluster assignment number
cancerdat[which(is.na(cancerdat$icluster)), "icluster"] <- max(cancerdat$icluster, na.rm = TRUE) + 1
cancerdat[which(is.na(cancerdat$coca)), "coca"] <- max(cancerdat$coca, na.rm = TRUE) + 1
```

```{r prep-for-stan}
setwd("D:/repo/TTmanu/data/survival/")
#####################
## Processing Data ##
#####################
deadpat <- sqldf("select patient_id,
                 age_at_initial_pathologic_diagnosis as age, 
                 gender,
                 race, 
                 death_days_to,type,
                 clust_knn as clust, 
                 clust as origclust, 
                 icluster, 
                 coca,
                 `ajcc_pathologic_tumor_stage.y` as stage, 
                 last_contact_days_to as lastcont from cancerdat where (death_days_to is not null or lastcont is not null) and age is not null and age > 9")

## Clean up data
deadpat$patient_id <- as.factor(deadpat$patient_id)
deadpat$type <- as.factor(deadpat$type)
deadpat$tevent <- deadpat$death_days_to
deadpat[is.na(deadpat$tevent),"tevent"] <- deadpat[is.na(deadpat$tevent),"lastcont"]
deadpat[is.na(deadpat$death_days_to),"eventtype"] <- "bfollow"
deadpat[!is.na(deadpat$death_days_to),"eventtype"] = "adeath"
deadpat$eventtype = as.factor(deadpat$eventtype)
deadpat$gender = as.factor(deadpat$gender)

#############################
## mortality data from CDC ##
#############################
namesmort = c("agerng","pdead","nsurv","ndie","pyrs","pyrsabove","expectedlife")
menmort = read_csv("lifetables/Table02.csv",skip=2)
## drop the final row, it's just a text info statement
menmort = menmort[1:NROW(menmort)-1,] 
wommort = read_csv("lifetables/Table03.csv",skip=2)
wommort = wommort[1:NROW(wommort)-1,]

names(menmort) = namesmort
names(wommort) = namesmort
yrs=seq(0,100,by=1);
menmort$yrnum = yrs
wommort$yrnum = yrs

## Store original cancer data in new object for further analysis
#save(deadpat, file = "D:/repo/TTmanu/data/survival/deadpat_all.rda")
load("../../data/survival/deadpat_all.rda")
standata <- deadpat

## convert negative tdeath to 0
standata[which(standata$tevent < 0), "tevent"] <- 0
## There are currently 3 types of classification model in this analysis:
## knn (our classification), icluster, and coca

model <- "tis"

if(model == "knn") {
  standata$clust <- as.factor(standata$clust)
  colnames(standata)[which(names(standata) == "clust")] <- "modelofinterest"
} else if(model == "icluster") {
  standata$icluster <- as.factor(standata$icluster)
  colnames(standata)[which(names(standata) == "icluster")] <- "modelofinterest"
} else if(model == "coca") {
  standata$coca <- as.factor(standata$coca)
  colnames(standata)[which(names(standata) == "coca")] <- "modelofinterest"
} else if(model == "tis") {
  standata$modelofinterest <- standata$type ## Convert tissue type to modelofinterest
  standata$type <- "none" ## No tissue category
  standata$type <- as.factor(standata$type) ## Factor the no tissue category
} else {
  print("Incorrect model selection")
}

if(model %in% c("knn", "icluster", "coca")) {
  standata <- list(Np=NROW(standata), 
                Nt=NROW(levels(standata$type)), 
                Nc=NROW(levels(standata$modelofinterest)),
                ##patient=as.numeric(deadpat$patient_id),
                age=standata$age,
                tissue=as.numeric(standata$type),
                menccdf=menmort$nsurv/100000,
                womccdf=wommort$nsurv/100000,
                gender=as.numeric(standata$gender),
                tclass=as.numeric(standata$modelofinterest),
                tevent = standata$tevent,
                eventtype=as.numeric(standata$eventtype))
}

if(model == "tis") {
  standata <- list(Np=NROW(standata), 
                Nt=NROW(levels(standata$modelofinterest)), 
                age=standata$age,
                tissue=as.numeric(standata$modelofinterest),
                menccdf=menmort$nsurv/100000,
                womccdf=wommort$nsurv/100000,
                gender=as.numeric(standata$gender),
                tclass=as.numeric(standata$modelofinterest),
                tevent = standata$tevent,
                eventtype=as.numeric(standata$eventtype))
}

### Fit the survival model with one rate per tissue*class
rm(smodel, start, start2, start3); gc()
if(model %in% c("knn", "icluster", "coca") == TRUE) {
  smodel = stan_model(file="../../code/survival/survmodel.stan")
} else if(model %in% c("tis", "icluster.kcl", "coca.kcl") == TRUE) {
  smodel = stan_model(file="../../code/survival/survmodel_tisonly.stan")
} else {
  print("Incorrect model selection")
}

## optimization for initialization doesn't always work... 
start = optimizing(smodel,data=standata);
start2 = optimizing(smodel,data=standata);
start3 = optimizing(smodel,data=standata);

if(model == "knn") {
  tryCatch(load("survmodel_effcl_knn_070121.stansave"),error=function(e){
    samps = sampling(smodel,data=standata,chains=4,warmup=1000,iter=2500,cores=4,control = list(max_treedepth=9)) #,init=list(start,start2,start3));
    save("samps",file="survmodel_effcl_knn_070121.stansave")
  })
} else if(model == "tis") {
  tryCatch(load("survmodel_effcl_tis_070121.stansave"),error=function(e){
    samps = sampling(smodel,data=standata,chains=4,warmup=1000,iter=2500,cores=4,control=list(max_treedepth=9)) #,init=list(start,start2,start3));
    save("samps",file="survmodel_effcl_tis_070121.stansave")
  })
} else if(model == "icluster") {
  tryCatch(load("survmodel_effcl_icluster_070121.stansave"),error=function(e){
    samps = sampling(smodel,data=standata,chains=4,warmup=1000,iter=2500,cores=4,control=list(max_treedepth=9)) #,init=list(start,start2,start3));
    save("samps",file="survmodel_effcl_icluster_070121.stansave")
  })
} else if(model == "coca") {
  tryCatch(load("survmodel_effcl_coca_070121.stansave"),error=function(e){
    samps = sampling(smodel,data=standata,chains=4,warmup=1000,iter=2500,cores=4,control=list(max_treedepth=9)) #,init=list(start,start2,start3));
    save("samps",file="survmodel_effcl_coca_070121.stansave")
  })
} else {
  print("Incorrect model selection")
}

```

### 2021-07-20 code/survival/figure5.Rmd
Edit manuscript based on editor comments

@@
   effage.agesw <- effage0tiss * effage0 + effagert * (agesw - 20)
-  survyrswf = sapply(effage.agesw,function(a) tryCatch(simdeath(r20wf,ageratewf,a,kmean)/365,error= function(e) {0}))
+  survyrswf = sapply(effage.agesw,function(a) tryCatch(simdeath(r20wf,ageratewf,a,k)/365,error= function(e) {0}))
 

### 2021-07-20 code/supplemental_file_shiny/app.R

@@ -35,7 +35,7 @@ load("tumor_samples.Rda")
 load("supplemental_gene.Rda")
 load("gdc_reactome_path_analysis.Rda")
 load("cancerdat.rda")
-load("survmodel_effcl_knn_062821.stansave")
+load("survmodel_effcl_knn_070121.stansave")
 dfsamps <- as.data.frame(samps)
 load("dfclas.rda")
 

### 2021-07-20 code/figure3.Rmd

@@
     name = "Effect Size",
-    labels = c(-1, "", 0.5, "", "", 0, "", "", 0.5, "", 1)
+    labels = c(-0.8, "", -0.4, "", 0, "", 0.4, "", 0.8)
   ) +

### 2021-07-20 code/figure4.Rmd

@@
   geom_tile() +
-  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
+  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within  \nCluster", limits = c(0, 1)) +
   xlab("Tumor Type") +

### 2021-07-20 text/abstract.Rmd

@@
 the cell of origin for a tumor have seen limited success, presenting a barrier
-to individualized medicine. Here we present a novel, pathway-centric approach,
+to individualized medicine. Here we present a pathway-centric approach,
 extracting somatic mutational profiles within and between tissues, largely

### 2021-07-20 text/discussion.Rmd

--- 2021-07-01 text/discussion.Rmd
@@
 nucleosome as found in cluster 4 have been proposed to hit all attributes of hallmarks of cancer. Cluster 
-7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks: 
+7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks, 
+showing that some clusters share common driver mechanisms: 
 “sustaining proliferative signaling”, “evading tumor suppressors”, “resisting cell 
 death”, and “inducing angiogenesis”. These clusters were strongly defined by 
-mutations in _PI3K_ or _RAS_, respectively. Taken together, our clusters highlight 
+mutations in _PI3K_ or _RAS_, respectively, suggesting a shared etiology in the EGFR 
+pathway for which these genes represent alternative intracellular signaling mechanisms. Taken together, our clusters highlight 
 the similarities and differences at a gene and pathway level between and within 

### 2021-07-20 text/introduction.Rmd

--- 2021-07-02 text/introduction.Rmd
@@
 to interpret as genetic aberrations are linked to molecular pathways. In a perspective, 
-Creixell et al. [@Creixell-pathwaynetworkanalysis] described approaches of pathway 
-and network analysis applicable to next-generation sequencing data. The most common 
+Creixell et al. [@PATHNET] described approaches of pathway 
+and network analysis applicable to next-generation sequencing data, the most common 
 approach being fixed-gene enrichment analysis to identify over-represented pathways. 

### 2021-07-20 text/manuscript.Rmd

@@
 title: "A molecular taxonomy of tumors independent of tissue-of-origin"
-author: Peter T. Nguyen$^{1,¶}$, Simon G. Coetzee$^{1,¶}$, Daniel L. Lakeland$^{2}$, and
-  Dennis J. Hazelett$^{1,3,*}$
+author: Peter T. Nguyen$^{1,4}$, Simon G. Coetzee$^{1,4}$, Daniel L. Lakeland$^{2}$, and
+  Dennis J. Hazelett$^{1,3,5,*}$
 bibliography: references.bib
+nocite: '@*'
 csl: iscience.csl
@@
 $^{3}$Samuel Oschin Comprehensive Cancer Institute, Cedars-Sinai Medical Center, Los Angeles, CA, 90048, USA.
-$^{¶}$These authors contributed equally to this work.
-$^{*}$Correspondence: Dennis.Hazelett@cshs.org
+$^{4}$These authors contributed equally.
+$^{5}$Lead Contact $^{*}$Correspondence: Dennis.Hazelett@cshs.org
 
@@
 clusters capturing about 80% of the tumor samples. To capture the remaining
-samples into one of these 10 clusters we used kNN (see **Supplementary Methods**
+samples into one of these 10 clusters we used kNN (see **STAR Methods**
 for details on clustering methods).
@@
 
-## Author Contributions
-P.N., S.G. and D.L. conducted the experiments. D.H. designed the experiments and wrote the paper.
-
 ## Acknowledgements
@@
 suggested improvements to the preprint. We also thank Ivetth Corona for early
-discussions and feedback. This work was supported by a grant from the 
-Cedars-Sinai Precision Health Initiative.
+discussions and feedback. Research reported in this publication was supported 
+in part by Graduate Research Education at Cedars-Sinai (P.N.), by a Cedars-Sinai 
+Precision Health award (D.H.), by institutional commitments to D.H., and by the
+Office of the Director, National Institutes of Health under Award Number 
+OT2OD026675 (D.H.).
+
+## Author Contributions
+P.N., S.G. and D.L. conducted the experiments. D.H. designed the experiments and wrote the paper.
 
@@
 A) Sankey plot of correspondence between cancer type and cluster identity.
-B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
+B) Projection of breast cancer subtypes onto the UMAP. See also Figure S1. C) Projection of BRCA1/2
 somatic mutation onto the UMAP. D) Projection of tumor stage onto the UMAP,
@@
 
-## Resource Availability
+## STAR Methods
 
-Further information and requests for resources should be directed to and will
-be fulfilled by the Lead Contact, Dennis Hazelett (Dennis.Hazelett@cshs.org). Data
-generated in this study have been deposited to Zenodo (https://doi.org/10.5281/zenodo.4011291).
-All code for producing the analyses and figures herein are included in this fully reproducible 
-manuscript in R markdown format. R markdown files and all other models are available from our 
-repositories (https://github.com/dennishazelett/TTmanu) on the distributed
-version control site, GitHub. An interactive media of the results is available on https://junkdnalab.shinyapps.io/PANCAN_supplemental/.
+### KEY RESOURCES TABLE
 
-## Methods
+### RESOURCES AVAILABILITY
+### Lead contact
+Requests for further information and resources should be directed to the Lead Contact, 
+Dennis J. Hazelett (Dennis.Hazelett@cshs.org).
 
-Extended descriptions of clustering, models,
-formulae, pathway selection, and gene filtering are provided in supplementary
-methods.
+### Materials availbility
+This study did not generate any new material.
+
+### Data and code availability
+* This paper analyzes existing, publicly available data. These datasets are listed 
+in the key resources table. Data generated in this study have been deposited at Zenodo 
+and are publicly available as of the date of publication. DOI is listed in the key 
+resources table.
+* All code for producing the analyses and figures herein are included in this fully 
+reproducible manuscript in R markdown format. R markdown files and all other models are 
+available from our repository on the distributed version control site, Github, listed 
+in the key resources table.
+* Any additional information required to reanalyze the data reported in this paper is 
+available from the lead contact upon request.
+
+### METHOD DETAILS
+
+### Selection of pathways
+
+To understand the molecular mechanism of cancer at a pathway level, we used
+Reactome (https://reactome.org/), a knowledge-based pathway database. The
+mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
+complexes to top pathways were downloaded from
+https://reactome.org/download-data. Using these data, we imposed pathway
+criteria to define basic cellular processes and biochemical pathways: (1)
+human-derived pathways ( _“HSA"_ ) (2) limited to grandchild node for each
+parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
+Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
+"Muscle contraction", and "Reproduction" or pathway names that include any of
+the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
+"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
+excluded pathways have been shown to play an important role in cancer, they are
+highly specialized (_e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
+most of the excluded pathways, a neutral version pathway of the pathway exists
+(_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
+Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
+operation produced a lookup table that consisted of 377 pathways mapped to 8,940
+genes.
+
+### Filtering genes
+
+We filtered likely erroneous mutations due to transcription coupled repair. Our
+approach was to determine the status of each gene (_i.e._ expressed or not
+expressed) in each tissue type in order to exclude lowly expressed
+genes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
+effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
+the genes and tumor samples that were not included in our analysis, grouped the
+tumor samples by tissue type and computed the mean expression value for each
+gene. A minimum threshold of 10 transcripts per million was set for expressed
+genes based on an inflection point observed when plotting the mean expression
+values of genes ranked by expression in each tissue type. Genes that did not
+meet this threshold were considered not expressed. This operation produced a
+lookup table for gene expression status in each tumor sample for 18,127 genes.
+
+### Clustering
+
+In order to classify tumors using this dataset, we used multiple correspondence
+analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
+containing useful information by selecting the eigenvalue with the most
+explanatory power, using the average of 100 permutations of the data as baseline
+(**Figure S6A**). We then chose the maximum eigenvalue for which the p-value
+remained ≤ 0.05 (see cutoff in **Figure S6B**). Then we performed a UMAP
+analysis [@McInnes2018], both in order to summarize the MCA graphically, and as
+a preprocessing step to boost the performance of density based clustering. The
+resulting map was notable for its lobed structure, with several reproducible
+projections regardless of random seed setting. A representative version of this
+3D UMAP is shown in **Figure S6C**, rotated to enhance the visibility of the
+major features. Following this spatial mapping we attempted to define groupings
+of similar tumors within the spatial map using HDBSCAN, which performs
+hierarchical clustering and provides metrics of cluster stability and
+probabilities of cluster membership for each node [@hdbscan]. However, HDBSCAN
+is sensitive to several parameters; key for our analysis are the minimum number
+of tumor samples in a cluster that capture the maximum number of tumor samples,
+measured by probability of membership of ≥ 5% in at least one cluster. Thus, we
+created a score metric as the fraction of classified tumors with max probability
+< 5% in one cluster and chose a cluster size of 92 to minimize the score
+function (**Figure S6D**). HDBSCAN with these settings resulted in ten distinct
+high-density clusters which we then projected onto the UMAP (**Figure S6E**).
+This classified 6,038 out of 7,607 tumors but still resulted in a significant
+fraction of unclassified tumor samples. Since we ultimately wish to be able to
+classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
+analysis, which computes a similarity metric to every tumor in the set and then
+lets the _k_ most similar tumors "vote" as to the identity of the query tumor
+sample based on their cluster labels. We set _k_ to be the square root of the
+number of tumor samples (87). Using this method, we assigned cluster membership
+to the remaining tumors (**Figure S6F** and see interactive media from **File S1** 
+(junkdnalab.shinyapps.io/PANCAN_supplemental/)).
+
+### Survival
+
+We used Bayesian inference to explore the relative impact of cancer diagnosis on
+survival at the time of diagnosis. Our model assumes that at birth each person
+has a small initial probability per unit time to die, and that this probability
+grows exponentially in time at some baseline rate that matches the observed CDC
+data. We assume that at diagnosis this information will allow us to treat
+patients as if they were a different age, and with further time passing at a
+different rate, a self-similarity based model. The effective age at diagnosis is
+based on a baseline age for that tissue/class combination, plus a perturbation
+for each year over 20 years old at diagnosis. To fit the model, we constrain the
+baseline risks using the available CDC life tables (**Figure S7**). We also
+constrain the cancer specific parameters using the longevity data within the
+cancer tumor dataset. Our model estimates the effect of cancer type and
+class-specific cancer effects independently, resulting in a cancer and class
+specific estimate of the effective age.
+
+Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes
+accuracy in the right tail of the distribution. We modeled baseline longevity
+using a risk rate function. The risk rate of death at a time $t$ days after the
+$A$ birthday is given as:
+
+$$R(t) = R_{20}\exp(a(A-20) + D(t) k a (\frac{t}{365}))$$
+
+Time is split into two components. $A$ represents the "effective age" of the
+patient at diagnosis, whereas $t$ starts at 0 upon diagnosis and represents the
+days since diagnosis, a field recorded in the dataset. $D(t)$
+is an indicator for whether diagnosis has occurred or not (before diagnosis
+$D(t) = 0$, after $D(t) = 1$). The model assumes that the risk of death is
+$R_{20}$ at age 20 years and increases exponentially with a constant rate $a$ up
+to the age at diagnosis. Risk of death assumes a new rate $ka$ thereafter.
+
+Given this risk per unit time, the probability of death at time $T+dT$ is the
+probability to survive to time $T$ which is $(1-P(T))$ and then die in the
+remaining interval which is $R(T)dT$
+
+$$dP = (1-P(T)) R(T) dT$$
+
+leading to the differential equation for the cumulative probability of death
+$P(T)$
+
+$$\frac{dP}{dT} = (1-P(T))R(T)$$
+
+The solution of this ODE is:
+
+$$P(T) = 1 - \exp(-\int_0^T R(t) dt)$$
+
+And the density of deaths per unit time is the derivative with respect to T:
+
+$$p(T) = R(T) \exp(-\int_0^T R(t)dt)$$
+
+However, to model the risk of cancer effectively we found that it was necessary
+not to use the actual age $A'$ but instead the "effective age" for that
+tissue/class type. Our effective age is calculated as:
+
+$$A = A_{0} A_{cl} + A_r (A'-20)$$ Where $A_0$ is the tissue specific effective
+age parameter (expressed in years), $A_{cl}$ is a dimensionless multiplier for
+the given class, and $A_r$ is the dimensionless age rate which determines how
+actual age $A'$ affects effective age A using a linear perturbation. For a
+non-cancer patient, $A_0=20$, $A_{cl}=1$, and $A_r=1$ are the relevant
+benchmarks.
+
+The $R(T)$ function has two important parameters, the $a$ value which represents
+the background risk rates, and the $k$ value, which is a function of cancer
+tissue type.
+
+$$k = k_{tis}$$
+
+The product $A_0 A_{cl}$ is obviously symmetric between the two factors. To
+disambiguate the meaning of the two, the prior for $A_0$ has high probability
+range from about 40 to 100 years, whereas the prior distribution of $A_{cl}$ has
+peak probability at 1 and a relatively narrow width, due to the fact that it is
+a tissue specific multiplicative modifying factor. To understand the roles of
+these priors refer to (**Figure S8**).
+
+Using this model, we split the data by Male/Female sex. Sex is known to be a
+risk factor for death, with males dying at slightly higher rates for all ages of
+interest in our dataset.
+
+As an aside, we did not explicitly account for stage at diagnosis in our model
+for the following reasons. In the tumor sample data, stage at diagnosis is
+confounded with cancer type because some cancers are screened aggressively
+(_e.g._ colorectal and prostate cancers) while others are diagnosed typically
+after they become problematic for the patient's lifestyle (_e.g._ ovarian and
+pancreatic cancers). Secondly, such a model specification would suffer from
+added noise because the staging data are not well standardized across cancer
+types, have different criteria, and because it is unclear what the relationship
+between stage and advancement of disease is (for example some sub-stage 4 tumors
+are metastatic). To compound this latter issue, our tumor set is vastly
+under-powered given the uneven representation of stage across cancer types.
+
+### QUANTIFICATION AND STATISTICAL ANALYSIS
+
+### Class and stage-specific Enrichment Calculations
+
+We calculated the enrichment of pathways in one set of tumors as the relative
+fraction of tumors (with estimated uncertainty) with a mutated gene in the
+pathway to all tumors, inclusive of the category of interest. To do this
+operation, we use the beta distribution to permute a posterior distribution on
+the fraction of tumors with a pathway mutated for each category based on the
+observed set, and compared this to the posterior obtained from the full set of
+tumors (all tumors) as the distribution of differences between all permuted
+samples. We considered a pathway enriched if the 95% range of credible
+differences thus obtained excludes 0 and the mean credible difference was
+greater than or equal to 30% enrichment, which excludes a large number of small
+differences that are not likely to be biologically relevant.
+
+## Supplemental Figure and Table Legends
+
+**Figure S1. Distribution of cancer type on UMAP clusters, Related to Figure 2.** Each panel shows the identity of 
+individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
+in **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
+
+**Figure S2. Histological subtypes, Related to Figure 2.** Each panel shows the histological subtypes of one of the 23 
+cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors 
+indicate the tumor's histological subtype identity.
+
+**Figure S3. Distribution of mismatch repair genes on UMAP clusters, Related to Figure 2.** Each panel shows one of 15 
+genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. 
+The colors indicate the tumor's cluster identity.
+
+**Figure S4. Distribution of oncogenes/tumor suppressors on UMAP clusters, Related to Figure 2.** Each panel shows 
+one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from 
+**Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
+
+**Figure S5. Performance comparison of different classification model - knn, COCA, iCluster, 
+tissue-of-origin (tis), Related to Figure 5.** A) Empirical cumulative distribution function comparing time of death 
+prediction of all classification models with actual time of death (dark green). B) Error 
+distribution plot of predicted and actual time of death.
+
+**Figure S6. Method for tumor classification using pathway aberration data, Related to Figure 1.** 
+A) MCA analysis was performed to determine the eigenvalue with the most
+explanatory power from the categorical data. Average of 100 permutations of the
+data was used as a baseline. B) The p-value of each eigenvalue was plotted and
+the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
+projected down to 3 dimensions for the purpose of density based clustering (only
+the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
+used to perform hierarchical density based clustering tuned to minimize the
+fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
+the UMAP projection; tumors are colored by cluster membership (see legend). This
+results in some tumors remaining unclassified (grey points). F) kNN was used to
+assign membership to the remaining tumors.
+
+**Figure S7. Comparison between CDC data and ODE model fit for
+males and females of European descent, Related to Figure 5.** This comparison does not distinguish on
+ethnicity for simplicity in visualization.
+
+**Figure S8. Priors for the tissue and class coefficients, Related to Figure 5 and 6.** A) The tissue
+coefficient is chosen to admit ages between 40 and 100 years old as a baseline
+age. B) Within a given tissue, the class parameter multiplies the tissue value and
+is given a peak density at 1, and a credible range from near 0.6 to 1.4,
+reflecting our expectation of a moderate multiplicative perturbation of
+variation across the classes.
+
+**Table S1. Ranked pan-cancer reactome pathway mutation, Related to Figure 1.** 
+Pathways are ranked from highest to lowest percent mutated. Gene 1-3 are the top 3 
+pathway genes mutated.
+
+**Table S2. Cluster-specific enriched pathways (effect size $\ge$ 0.30), Related to Figure 3.**
+
+**File S1. An interactive media of the results, Related to Figure 1-6.**
 
@@
 <div id="refs"></div>
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Fig 1
-```{r child='../code/figure1.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Fig 2
-```{r child = '../code/figure2.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Fig 3
-```{r child = '../code/figure3.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Fig 4
-```{r child = '../code/figure4.Rmd'}
-```
-
-#### Fig 5
-```{r child='../code/survival/figure5.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Fig 6
-
-```{r child='../code/figure6.Rmd'}
-```
-
-```{r, results='asis'}
-cat("\\newpage")
-```
-
-#### Table 1
-```{r child = '../code/table1.Rmd'}
-```

### 2021-07-20 text/supplemental_methods.Rmd

@@
 explanatory power, using the average of 100 permutations of the data as baseline
-(**SM1A Fig**). We then chose the maximum eigenvalue for which the p-value
-remained ≤ 0.05 (see cutoff in **SM1B Fig**). Then we performed a UMAP
+(**Figure SM1A**). We then chose the maximum eigenvalue for which the p-value
+remained ≤ 0.05 (see cutoff in **Figure SM1B**). Then we performed a UMAP
 analysis [@McInnes2018], both in order to summarize the MCA graphically, and as
@@
 projections regardless of random seed setting. A representative version of this
-3D UMAP is shown in **SM1C Fig**, rotated to enhance the visibility of the
+3D UMAP is shown in **Figure SM1C**, rotated to enhance the visibility of the
 major features. Following this spatial mapping we attempted to define groupings
@@
 < 5% in one cluster and chose a cluster size of 92 to minimize the score
-function (**SM1D Fig**). HDBSCAN with these settings resulted in ten distinct
-high-density clusters which we then projected onto the UMAP (**SM1E Fig**).
+function (**Figure SM1D**). HDBSCAN with these settings resulted in ten distinct
+high-density clusters which we then projected onto the UMAP (**Figure SM1E**).
 This classified 6,038 out of 7,607 tumors but still resulted in a significant
@@
 number of tumor samples (87). Using this method, we assigned cluster membership
-to the remaining tumors (**SM1F Fig** and see interactive media from **S1 File** (junkdnalab.shinyapps.io/PANCAN_supplemental/)).
+to the remaining tumors (**Figure SM1F** and see interactive media from **File S1** (junkdnalab.shinyapps.io/PANCAN_supplemental/)).
 
@@
 a tissue specific multiplicative modifying factor. To understand the roles of
-these priors refer to (**SM3 Fig**).
+these priors refer to (**Figure SM3**).
 
@@
 
-#### Fig SM1
+#### Figure SM1
 ```{r child='../code/figureSM1.Rmd'}
@@
 
-**Fig SM1. Method for tumor classification using pathway aberration data** 
+**Figure SM1. Method for tumor classification using pathway aberration data** 
 A) MCA analysis was performed to determine the eigenvalue with the most
@@
 
-#### Fig SM2
+#### Figure SM2
 
@@
 
-**Fig SM2. Comparison between CDC data and ODE model fit for
+**Figure SM2. Comparison between CDC data and ODE model fit for
 males and females of European descent.** This comparison does not distinguish on
@@
 
-#### Fig SM3
+#### Figure SM3
 
@@
 
-**Fig SM3. Priors for the tissue and class coefficients.** A) The tissue
+**Figure SM3. Priors for the tissue and class coefficients.** A) The tissue
 coefficient is chosen to admit ages between 40 and 100 years old as a baseline

### 2021-07-20 text/survival.Rmd

@@
 
-To illustrate the model, we simulate survival for six related groups (**Fig
+To illustrate the model, we simulate survival for six related groups (**Figure
 5A**). We start with a cohort of women aged 30 randomly selected from the US
@@
 We found that cancer types, as expected, have a range of prognoses relative to
-the general population. In **Fig 5B-D** we show how effective age baseline, and
+the general population. In **Figure 5B-D** we show how effective age baseline, and
 effective age rate (which modifies the baseline based on your actual age) as
@@
 
-In **Fig 5E-H** we see three particularly deadly cancers (Glioblastoma:GBM,
+In **Figure 5E-H** we see three particularly deadly cancers (Glioblastoma:GBM,
 Pancreatic:PAAD, Ovarian:OV), and one cancer where diagnosis hardly changes risk
@@
 relatively peaked around 1 reflecting the purpose of this coefficient as a
-multiplicative perturbation (**Fig 6**). A cluster-specific rate of 1 represents
+multiplicative perturbation (**Figure 6**). A cluster-specific rate of 1 represents
 the typical rate for this tissue type. For several cancers (_e.g._  PRAD, kidney
@@
 performance compared to COCA [@hoadleyMultiplatformAnalysis2014], iCluster
-[@hoadleyCellofOriginPatternsDominate2018] and tissue-of-origin (tis). (**S5A
-Fig**). Additionally, we computed the entropy of the error distribution in
-predictions (**S5B Fig**).  All models had similar performance with kNN
+[@hoadleyCellofOriginPatternsDominate2018] and tissue-of-origin (tis). (**
+Figure S5A**). Additionally, we computed the entropy of the error distribution in
+predictions (**Figure S5B**).  All models had similar performance with kNN
 performing slightly better as indicated by a slightly smaller error entropy.

### 2021-08-27 text/manuscript.Rmd
Remove supplemental figure/legend from main text

@@
 We wish to thank David Van Valen, Kate Lawrenson, Simon Knott, and Megan
-Hitchens, for critical comments on this manuscript, and Tim Triche Jr for
+Hitchins, for critical comments on this manuscript, and Tim Triche Jr for
 suggested improvements to the preprint. We also thank Ivetth Corona for early
-discussions and feedback. Research reported in this publication was supported 
+discussions and feedback. The graphical abstract was created by BioRender.com. 
+Research reported in this publication was supported 
 in part by Graduate Research Education at Cedars-Sinai (P.N.), by a Cedars-Sinai 
@@
 
-**Figure S1. Distribution of cancer type on UMAP clusters, Related to Figure 2.** Each panel shows the identity of 
-individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
-in **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
-
-**Figure S2. Histological subtypes, Related to Figure 2.** Each panel shows the histological subtypes of one of the 23 
-cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors 
-indicate the tumor's histological subtype identity.
-
-**Figure S3. Distribution of mismatch repair genes on UMAP clusters, Related to Figure 2.** Each panel shows one of 15 
-genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. 
-The colors indicate the tumor's cluster identity.
-
-**Figure S4. Distribution of oncogenes/tumor suppressors on UMAP clusters, Related to Figure 2.** Each panel shows 
-one of the twelve oncogenes/tumor suppressors projected onto the tumors of the UMAP from 
-**Figure 1B of the main text**. The colors indicate the tumor's cluster identity.
-
-**Figure S5. Performance comparison of different classification model - knn, COCA, iCluster, 
-tissue-of-origin (tis), Related to Figure 5.** A) Empirical cumulative distribution function comparing time of death 
-prediction of all classification models with actual time of death (dark green). B) Error 
-distribution plot of predicted and actual time of death.
-
-**Figure S6. Method for tumor classification using pathway aberration data, Related to Figure 1.** 
-A) MCA analysis was performed to determine the eigenvalue with the most
-explanatory power from the categorical data. Average of 100 permutations of the
-data was used as a baseline. B) The p-value of each eigenvalue was plotted and
-the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
-projected down to 3 dimensions for the purpose of density based clustering (only
-the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
-used to perform hierarchical density based clustering tuned to minimize the
-fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
-the UMAP projection; tumors are colored by cluster membership (see legend). This
-results in some tumors remaining unclassified (grey points). F) kNN was used to
-assign membership to the remaining tumors.
-
-**Figure S7. Comparison between CDC data and ODE model fit for
-males and females of European descent, Related to Figure 5.** This comparison does not distinguish on
-ethnicity for simplicity in visualization.
-
-**Figure S8. Priors for the tissue and class coefficients, Related to Figure 5 and 6.** A) The tissue
-coefficient is chosen to admit ages between 40 and 100 years old as a baseline
-age. B) Within a given tissue, the class parameter multiplies the tissue value and
-is given a peak density at 1, and a credible range from near 0.6 to 1.4,
-reflecting our expectation of a moderate multiplicative perturbation of
-variation across the classes.
-
 **Table S1. Ranked pan-cancer reactome pathway mutation, Related to Figure 1.** 
